### 2013-04-12 paper.tex
Initial import of the ISSR article

%%%%%%%%%%%%%%%%%%%% author.tex %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% sample root file for your "contribution" to a contributed volume
%
% Use this file as a template for your own input.
%
%%%%%%%%%%%%%%%% Springer %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% RECOMMENDED %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[graybox]{svmult}

% choose options for [] as required from the list
% in the Reference Guide

\usepackage{mathptmx}       % selects Times Roman as basic font
\usepackage{helvet}         % selects Helvetica as sans-serif font
\usepackage{courier}        % selects Courier as typewriter font
\usepackage{type1cm}        % activate if the above 3 fonts are
                            % not available on your system
%
\usepackage{makeidx}         % allows index generation
\usepackage{graphicx, subfigure}        % standard LaTeX graphics tool
                             % when including figure files
\usepackage{multicol}        % used for the two-column index
\usepackage[bottom]{footmisc}% places footnotes at page bottom

\usepackage{fixme}
\usepackage{lscape}

% see the list of further useful packages
% in the Reference Guide

\makeindex             % used for the subject index
                       % please use the style svind.ist with
                       % your makeindex program

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% Custom macros
\newcommand{\stmt}[1]{{\footnotesize $\langle$\stmttt#1\relax$\rangle$}}
\newcommand{\rawstmt}[1]{{\footnotesize \stmttt#1\relax}}
\def\stmttt#1 #2 #3\relax{{\tt#1} {\bf{\tt #2}} {\tt #3}}

\newcommand{\setstmt}[1]{{\footnotesize [\setstmttt#1\relax]}}
\def\setstmttt#1,#2\relax{\rawstmt{#1}, \rawstmt{#2}}

\newcommand{\concept}[1]{{\footnotesize \texttt{#1}}}

\newcommand{\ie}{{\textit{i.e.~}}}
\newcommand{\cf}{{\textit{cf~}}}
\newcommand{\eg}{{\textit{e.g.~}}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

\title*{When the robot considers the human... }
% Use \titlerunning{Short Title} for an abbreviated version of
% your contribution title if the original one is too long
\author{Rachid Alami, Mathieu Warnier, Julien Guitton, S\'{e}verin Lemaignan and Emrah Akin Sisbot}
\authorrunning{R. Alami, M. Warnier, J. Guitton, S. Lemaignan and E.A. Sisbot}
% Use \authorrunning{Short Title} for an abbreviated version of
% your contribution title if the original one is too long
\institute{ LAAS-CNRS\\ 
        Universit\'e de Toulouse, UPS, INSA, INP, ISAE, LAAS\\ 
         F-31077 Toulouse, France\\ \email{Firstname.Name@laas.fr}}
%
% Use the package "url.sty" to avoid
% problems with special characters
% used in your e-mail or web address
%
\maketitle

\abstract*{This paper addresses some key decisional issues that are
  necessary for a cognitive robot which shares space and tasks with a
  human. We adopt a constructive approach based on the identification
  and the effective implementation of individual and collaborative
  skills. The system is comprehensive since it aims at dealing with a
  complete set of abilities articulated so that the robot controller
  is effectively able to conduct in a flexible manner a collaborative
  task with a human partner. These abilities include geometric
  reasoning and situation assessment based essentially on
  perspective-taking and affordances, management and exploitation of
  each agent (human and robot) knowledge in a separate cognitive
  model, human-aware task planning and human and robot interleaved
  plan achievement.}

\abstract{This paper addresses some key decisional issues that are
  necessary for a cognitive robot which shares space and tasks with a
  human. We adopt a constructive approach based on the identification
  and the effective implementation of individual and collaborative
  skills. The system is comprehensive since it aims at dealing with a
  complete set of abilities articulated so that the robot controller
  is effectively able to conduct in a flexible manner a collaborative
  task with a human partner. These abilities include geometric
  reasoning and situation assessment based essentially on
  perspective-taking and affordances, management and exploitation of
  each agent (human and robot) knowledge in a separate cognitive
  model, human-aware task planning and human and robot interleaved
  plan achievement.}

\section{Introduction}

Human-robot interaction requires to equip the robot with explicit
reasoning on the human and on its own capacities to achieve its tasks
in a collaborative way with a human partner. This paper presents a
robot control system which has been especially designed for a
cognitive robot which shares space and task with a human. We have
adopted a constructive approach based on effective individual and
collaborative skills. The system is comprehensive since it aims at
dealing with a complete set of abilities articulated so that the robot
controller is effectively able to conduct in a flexible manner a
collaborative task with a human partner.

We illustrate below how we deal with a typical human-robot interactive
task achievement and what are the abilities we claim are
necessary. Section \S\ref{sec:problem} proposes a typical human-robot
interaction problem that can be solved by the proposed robot
controller. Section \S\ref{sec:soa} reviews related work and analyses
the context of our contribution.  Section \S\ref{sec:Framework}
provides an overview of the robot controller and introduces three
activities which are described in \S\ref{sec:situ}, \S\ref{sec:plan}
and \S\ref{sec:action}. Finally, Section \S\ref{sec:expes} presents an
effective run on a real robot in face to face interaction with a
person.

\section{A typical human-robot interaction problem}\label{sec:problem}

Let us consider a robot which is supposed to achieve interactive
object manipulation, fetch and carry tasks and similar tasks in a
domestic environment. The problem we are dealing with here is the
following. Given:
\begin {itemize}
\item a joint goal, which has been previously established and agreed
  upon (through a process which is out of the scope of this paper),
\item the current situation, acquired through perception or
  deduction from previous perceptions, including the state of the
  environment of the robot and of the human,
\end {itemize}
the robot controller computes an action to execute and who (the 
human or the robot, or both in case of a joint action) has to perform
it, and then controls or monitors its execution. The operation
continues until the goal is achieved, is declared unachievable or is
abandoned by the human.

To do so, the robot has to be equipped with a number of decisional, planning
and interpretation abilities where its human partner is taken
explicitly into account. It needs to be able:
\begin {itemize}
\item to build and maintain relevant robot and human beliefs
  (from the robot perspective) with respect to state of the world and the task,
\item to build and maintain iteratively shared (human-robot) plans, 
\item to refine and execute the actions it has to perform, and to monitor 
those achieved by its human partner.
\end {itemize}

Besides, we would like to build such abilities in a generic way, and
to provide several levels of parametrization allowing to adapt to
various environments, and various levels of involvement of the robot
ranging from teammate behavior to assistant or proactive helper.

\section{Related work and vision}\label{sec:soa}

The human presence brings new requirements for robot's abilities both
at the functional and at the deliberative levels~\cite{Klein2004}. The
topics involve motion~\cite{Kulic2007,Berg2004,Madhav2006},
navigation~\cite{Althaus2004,Sisbot2007}, manipulation~\cite{Kemp2007}
in presence of humans as well as perception of human
activities~\cite{Breazeal2001,Burger2008}. Also, when
interacting with humans, robots need to incorporate communication and
collaboration abilities. Several theories dealing with
collaboration~\cite{Cohen1991,Grosz1996,Clark1996} emphasize that
collaborative tasks have specific requirements compared to individual
ones, \eg, since the robot and the person share a common goal, they
have to agree on the manner to realize it, they must show their
commitment to the goal during execution, etc. Several robotic systems
have already been built based on these
theories~\cite{Rich1997,Sidner2005,Tambe2005a,Breazeal2003} and they
all have shown benefits of this approach. They have also shown how
difficult it is to manage turn-taking between communication partners
and to interleave task realization and communication in a generic
way. Finally, today only few
systems~\cite{Fong_2006,Breazeal2003,Sisbot2008} take humans into
account at all levels.

Perspective Taking is a human ability which allows one to put
him/herself in another person's point of view. Studied in
psychology literature~\cite{Flavell1992,Tversky1999}, this ability is
crucial when interacting with people by allowing one to reason on
others' understanding of the world in terms of visual perception, spatial
descriptions, affordances and beliefs, etc.
Therefore, in the last years these notions have been gradually
employed in Human-Robot Interaction.~\cite{breazeal2006} presents a
learning algorithm that takes into account information about a
teacher's visual perspective in order to learn a
task. ~\cite{Johnson2005} apply visual perspective taking for action
recognition between two robots.~\cite{Trafton2005} use both visual and
spatial perspective taking for finding out the referent indicated by a
human partner.

Spatial reasoning~\cite{OKeefe1999}, on the other hand, has been used
for natural language processing for applications such as direction
recognition ~\cite{Kollar10,Matuszek10} or language
grounding~\cite{Tellex10}.~\cite{Skubic2004} presented a spatial
reasoner integrated in a robot which computes symbolic positions of
objects.

\begin{figure}[htb]
\centering
\includegraphics[width=9cm]{figs/grounding_robot.pdf}
\caption{Robot reasoning about HRI and anticipation of human activities:
  sources of information are multi-modal dialogue, and observation of
  environment and human activity}
\label{fig:hri-dec}
\end{figure}

We envision HRI in a context where two agents (a human and a robot)
share a common space and exchange information through various
modalities. Our aim is to endow the robot with an explicit
consideration of the human and with the ability to manage its
interactions with him (Figure~\ref{fig:hri-dec}). This must be
considered at the architecture level as well as at the task/motion
planning and execution level.

We have devised a decisional framework for human-robot interactive
task achievement that is aimed to allow the robot not only to
accomplish its tasks but also to produce behaviors that support its
engagement vis-a-vis its human partner and to interpret human
behaviors and intentions. 
Together and in coherence with this framework, we have developed
and experimented various task planners and interaction schemes that
allow the robot to select and perform its tasks while taking into
account explicitly the human abilities as well as the constraints
imposed by the presence of humans, their needs and preferences.

\section{A Decisional framework}\label{sec:Framework}

Interaction happens as a consequence of an explicit request of the
human to satisfy a goal or because the robot finds itself in a
situation where it is useful if not mandatory. In both cases, the
robot has a goal to satisfy.  An important issue is the notion of
engagement, a process in which the robot will have to establish,
maintain and terminate a connection with a human partner. 
This covers goal establishment, selection of an incremental refinement
of the task that is intended to be achieved, and execution control
including monitoring, and even influencing, human task performance and
his/her commitment to the goal. The human involvement may range from a
direct participation to the task achievement, to a simple
``acceptance'' of robot activity in his/her close vicinity.

Our robot is controlled by a three layer
architecture~\cite{Alami1998}.  We present briefly its decisional
layer. The proposed decisional framework consists of several entities,
having each a specific role as illustrated by
Figure~\ref{architecture_fg}.  We describe how the robot is controlled
through an analysis of the three main activities performed by the
robot controller:

\begin{figure*}[thpb]
  \centering
  \includegraphics[width=1.0\textwidth]{./figs/supervisor-architecture-new.pdf} \\
  \caption {Architecture of the robot control system}
  \label{architecture_fg}
\end{figure*}

\begin {enumerate}
\item Situation assessment and context management 
\item Goals and plans management
\item Action refinement, execution and monitoring
\end {enumerate}

The next three sections describe the three robot controller
activities and how they make use of a number of key components in the
architecture:

\begin{itemize}
\item SPARK: Spatial Reasoning and Knowledge module \cite{Sisbot2011}
\item ORO: a knowledge management module \cite{Lemaignan2010}
\item HATP: a Human-Aware Task Planner \cite{Alili2008}
\item A set of Human aware motion, placement and manipulation planners
 \cite{Sisbot2008, Mainprice2011, Pandey2010}
\end{itemize}

Other decisional activities, such as situated dialog (\cite{Ros2010b,
  Lemaignan2011}, not presented here) have been developed that use the
same set of components.

\section{Situation assessment and context management}\label{sec:situ}

This activity involves the geometric and temporal reasoning component,
the symbolic facts and belief management component and the dedicated
robot controller activity (Figure~\ref{architecture_fg}).

We assume that perception provides in real-time the identity and the
position of objects when they are in the field of view of the sensors.
In our implemented examples, the robot is localised using a standard
horizontal laser-scanning based localisation system, the objects are
identified and localized using ARToolkit \cite{ARToolkit} and the
humans are tracked using a commercial motion capture system and a
Kinect device from Microsoft.

\subsection{Geometric and Temporal Reasoning component}\label{sub:gtrc}

The geometric reasoning component plays a central role in our
architecture. It is called SPARK (Spatial Reasoning and
Knowledge~\cite{Sisbot2011}) in the current implementation. It is
responsible for geometric information gathering and it embeds a number
of decisional activities linked to abstraction (symbolic facts
production) and inference based on geometric and temporal reasoning.
SPARK maintains all geometric positions and configurations of agents,
objects and furniture coming from perception and previous or {\it a
  priori} knowledge.

 %\subsubsection*{Symbolic facts production:} 
\vspace{0.3cm}
\noindent
\textbf{Symbolic facts production:} 
Geometric state of the world is abstracted in symbolic facts that can
be classified in three different categories.

\begin {itemize}
\item Relative positions of object and agents, e.g.  \stmt{GREY\_TAPE
    isOn TABLE}.

\item Perception and manipulation capacity and state of agents,
  e.g. \stmt{ROBOT looksAt GREY\_TAPE}, \stmt{GREY\_TAPE isVisibleBy
    HUMAN1}.

\item Motion status for object or agent parts, e.g.  \stmt{GREY\_TAPE
    isMoving true},\\ \stmt{ROBOT\_HEAD isTurning true}.
\end {itemize}

Reasoning about human perspective allow to compute facts such as:
\stmt{GREY\_TAPE isBehind HUMAN1}, \stmt{GREY\_TAPE isVisibleBy
  HUMAN1}.

Figure~\ref{fig::reach-ex} illustrates different situations for the
\textit{reachable} relation. In this case, the robot and its human partner
are placed face to face, in a table-top setup (Figure~\ref{fig::reach-ex}.1).
The robot first estimates if the small grey box is reachable to itself. This is
done by finding a collision free posture to reach the object
(Figure~\ref{fig::reach-ex}.2). Next the robot switches to the human's
perspective to estimate if the same object is reachable to the human as well.
In the last scene, the human moves towards his left, farther from the object
(Figure~\ref{fig::reach-ex}.4). The situation is then reevaluated. In this
occasion though, the reasoner cannot find a satisfactory posture for the human
to reach the box because he is too far from the target.

\begin{figure*}[!t]
	\centering
	\includegraphics[width=1.0\textwidth]{figs/reachex.jpg}	
	\caption{An example illustrating the \textit{reachable} relation. The relation is computed from the perspectives of both the robot and the human. The computed posture at each step is illustrated with a global view of the scene (top), and from a closest view (bottom).}
\label{fig::reach-ex}
\end{figure*}

\begin{figure}[ht!]
   \label{fig:sparkSubfigures}
   \begin{center}
%
       \subfigure[Initial state]{%
%           \label{fig:wuweiPhoto}
           \includegraphics[width=0.5\textwidth]{./figs/etat2-P1010769_brightened-v2.jpg}
       }%
       \subfigure[3d model view of initial state]{%
%          \label{fig:sparkScreenshot}
          \includegraphics[width=0.43\textwidth]{./figs/etat2_photo.png}
       }\\ %  ------- End of the first row ----------------------%
%
   \end{center}

   \caption{%
     In this situation, there are three tapes on the table. Two tapes
     are only reachable by the robot: the LOTR\_TAPE (black in the 3d
     model) and GREY\_TAPE. The third tape WALLE\_TAPE (white in the
     3d model) and the trashbin PINK\_TRASHBIN are only reachable by
     the human HUMAN1. All tapes are on the table TABLE.  }%

\end{figure}

The set of facts computed in the situation depicted by
Figure~\ref{fig:sparkSubfigures} is the following:
\begin{footnotesize}
%\begin{small}
\begin{verbatim}
                ROBOT                          HUMAN1
PINK_TRASHBIN isReachable false    PINK_TRASHBIN isReachable true 
WALLE_TAPE isReachable false       WALLE_TAPE isVisible true 
LOTR_TAPE isReachable true         LOTR_TAPE isReachable false 
GREY_TAPE isReachable true         GREY_TAPE isReachable false
WALLE_TAPE isVisible true          WALLE_TAPE isReachable true 
LOTR_TAPE isVisible true           LOTR_TAPE isVisible true 
GREY_TAPE isVisible true           GREY_TAPE isVisible true 
WALLE_TAPE isOn TABLE              WALLE_TAPE isOn TABLE 
LOTR_TAPE isOn TABLE               LOTR_TAPE isOn TABLE 
GREY_TAPE isOn TABLE               GREY_TAPE isOn TABLE 
\end{verbatim}
%\end{small}
\end{footnotesize}

%\newline\newline
%\subsubsection*{Hypotheses on objects states and positions}
\vspace{0.3cm}
\noindent
\textbf{Hypotheses on objects states and positions:}
It is sometimes difficult or even impossible to see and/or track an
object in certain states. This happens, for instance, when the object
has been put in a container, when it is in the robot gripper or in the
human hand, and more generally in any state in which it is hidden by
something else. Our robot has a model of the possible symbolic states
for an object (whether the object is on a furniture, in an agent hand,
in a container, etc.).  According to the robot perception of what has
happened since the object was last seen, the robot tries to maintain a
belief of the current possible symbolic states and their associated
probabilities for this object. Such information can be used to update
the beliefs using input from exploration, dialog, human visual
focus,\ldots
SPARK currently provides a simple implementation of such a functionality. The
only managed hypotheses are \emph{in container} and \emph{in agent hand}. We
can have only one hypothesis at the same time. Hypothesis validity is checked
geometrically in case of incoming perception values.

%\subsubsection*{Primitive action recognition}
\vspace{0.3cm}
\noindent {\bf Primitive action recognition:} 
Monitoring human activity is crucial to maintain a coherent state of
the world. Full human action and activity monitoring is a difficult
task that requires knowledge and reasoning both on high level facts
like goals, intentions and plans, as well as bottom-up data from agent
and object motions. Simple temporal and geometric reasoning on human
hand trajectories and potential objects placements can provide some
useful clues for high level human monitoring processes. We call this
temporal and geometric reasoning \emph{primitive action recognition}.

For example, a \emph{pick}, a \emph{throw} or a \emph{place} action
can be recognized by observing that an object on table and an empty
human hand are close to each other, or that the human hand holding an
object is close to a container, etc. Human hand position is either
directly perceived or inferred from its initial perceived trajectory.
We have a simple implementation of such a primitive action recognition
in SPARK that relies on monitoring human hand and its motion near
objects or above containers.

\subsection{Symbolic facts and beliefs management}

The facts produced by the geometric and temporal reasoning component
are stored in a central symbolic knowledge base, called ORO. Besides
acting as a facts database, the ORO platform~\cite{Lemaignan2010}
exposes several functions: operations on knowledge statements relying
on inference (through a continuous first-order logic classification
process), management of \emph{per-agent} symbolic models, and also
higher cognitive and human-robot interaction related functionalities
like categorization of sets of concepts, profiles of memory (that
enable the robot to ``forget'' about some facts), natural language
grounding~\cite{Lemaignan2011}\ldots.

ORO stores independent knowledge models (in our implementation, as
\emph{ontologies}) for each agent (the robot and the humans it
interacts with). The robot architecture components (like the executive
layer or the situation assessment component) can then store the
agents' beliefs in specific models.  Each of these models is
independent and logically consistent, enabling reasoning on different
perspectives of the world that would otherwise be considered as
globally inconsistent (for instance, an object can be visible for the
robot but not for the human. This object can have at the same time the
property \concept{isVisible \textbf{true}} and \concept{isVisible
  \textbf{false}} in two different models). This feature actually
allows us to consider the robot to be endowed with a simple
\emph{theory of mind}~\cite{Scassellati2002}: it can explicitly
model the belief state of its interactors.

ORO also provides an event mechanism that allows components to be
triggered when specific events occur. A component can
for instance subscribe to events of kind \setstmt{?agent isVisible
  true, ?agent type Human}. As soon as the perception layer detects a
human in the robot's field of view and accordingly updates the
knowledge base, the executive layer would be triggered back. The
event framework also takes advantage of the inference capabilities of
ORO. Thus an event can be indirectly triggered if its triggering
conditions can be inferred to be true.

\subsection{Situation Assessment and Context Management Controller}

Building, updating and maintaining a correct state of the world at
geometric and symbolic level is crucial to the capacity of the robot
to carry on successfully a multi-step interaction with a human. Tight
integration between the robot controller and the geometric and
temporal reasoning functions in SPARK and symbolic facts and beliefs
management in ORO is central.

The robot controller has access to the symbolic facts in ORO that are
automatically updated whenever object and agent positions are changed.
Robot controller can also access geometric perceived or inferred
positions of objects and geometric positions and postures of the human
that will be used to orient its cameras.  Building and updating the
state of the world first relies on perceiving objects. Robot
controller can use:

\begin {itemize}
\item Exploration policies: robot will exhaustively scan the table to
  see all what can be seen.

\item Search policies: robot will search an object until it is
  detected if possible, scanning all the table and looking in human
  hand.

\end {itemize}

Robot reasons on possible positions for non perceived
objects. These hypotheses could be updated using new input from dialogue, human
action and focus of attention. Currently, we manage at most one hypothesis per
object. This hypothesis is produced by robot controller through an inference on
robot or human action. In case of perception conflicts with low probability for
the current hypothesis, robot controller will break this hypothesis and delete
corresponding symbolic fact in the ontology.

Robot must reacts to change in the world not linked to robot
action to drive world update. Robot controller uses SPARK to monitor human hand
motion and primitive action recognition for \emph{pick} and \emph{throw}. As
mentioned above these primitive action recognition should be used with higher
level information on goals, intentions and plans and some exploration to
achieve complex human action and activity monitoring.  In the current
implementation, these primitive actions are over-optimistically interpreted as
the corresponding actions \emph{pick} object and \emph{throw} object.

\section{Goal and Plan Management}\label{sec:plan}

The Goal and Plan Management activity involves the human-aware
symbolic task planner component and the dedicated robot controller
activity (Figure~\ref{architecture_fg}).

\subsection{Symbolic Task Planning}

In order to devise how a given goal can be accomplished, the robot has
to elaborate a plan,~\textit{i.e.} a set of actions to be achieved by
the robot and its human partners.  This is the role of a HATP
\cite{Alili2008} (for Human Aware Task Planner).  HATP is based on a
Hierarchical Task Network (HTN) refinement which performs an iterative
task decomposition into sub-tasks until reaching atomic
actions~\cite{Nau2003}.  The planning domain defines a set of methods
describing how to decompose a task and can be seen as the Howto
knowledge of the robot.  HATP is able to produce plans for the robot's
actions as well as for the other participants (humans or robots). It
can be tuned by setting up different costs depending on the actions to
apply and by taking into account a set of constraints called social
rules. This tuning aims at adapting the robot's behavior according to
the desired level of cooperation of the robot.

%\subsubsection*{Agents and action streams:}
\vspace{0.3cm}
\noindent
\textbf{Agents and action streams:}
The robot plans not only for itself but also for the other agents. The
resulting plan, called ``shared plan'' is a set of actions that form
a stream for each agent involved in the goal achievement. Depending on
the context, some ``shared plans'' contain causal relations between
agents. For example, the second agent needs to wait for the success of
the first agent's action to be able to start its own action. When the
plan is performed, causal links induce synchronization between
agents. Figure~\ref{plan_hatp1} illustrates a plan with two streams.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.95\columnwidth]{./figs/first_plan.pdf}
  \caption{A plan produced by HATP with 2 streams}
  \label{plan_hatp1}
\end{figure}

%\subsubsection*{Action costs and social rules:}
\vspace{0.3cm}
\noindent
\textbf{Action costs and social rules:}
A cost and a duration function is associated to each action.
The duration function provides a duration interval for the action
achievement and is used, in one hand, to schedule the different
streams and, in the other hand, as an additional cost function.
In addition to these costs, HATP also takes into account a set of social
rules.  Social rules are constraints aiming at leading the plan
construction towards the best plan according to some human
preferences. The social rules we have defined so far deal with:

\begin{itemize}
\item undesirable state: to avoid a state in which the human could
  feel uncomfortable;
\item undesirable sequence: to eliminate sequences of actions that can
  be misinterpreted by the human;
\item effort balancing: to adjust the work effort of the agents;
\item wasted time: used to avoid long delays between the actions of
  the human partner;
\item intricate links: to limit dependencies between the actions of
  two or more agents.
\end{itemize}

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.95\columnwidth]{./figs/second_plan.pdf}
  \caption{A plan with the wasted time social rule}
  \label{plan_hatp2}
\end{figure}

Figure~\ref{plan_hatp2} illustrates an alternative plan to the previous 
one (Figure~\ref{plan_hatp1}) if the wasted time social rule is used.
The obtained shared plan is the best plan according to a global evaluation of
these multiple criteria.

%\subsubsection*{Several levels of cooperation:} 
\vspace{0.3cm}
\noindent
\textbf{Several levels of cooperation:} 
By tuning its costs
and adapting its social rules, HATP can be used to compute various
alternative plans. These plans can be categorized into several levels
of cooperation

\begin{itemize}
\item helping the human to achieve his goal by acting for him
\item sharing concrete resources by handing some objects
\item collaboration of the robot and the human by coordinating their
  actions towards a human-robot joint goal.
\end{itemize}

\subsection{ Goal and plan Controller}
Figure~\ref{goal_plans_fg} sums up the Goal and Plan Management as
implemented in the robot controller.  When an event announcing a new
goal is caught by the controller, the validity of this goal is tested:
does it corresponds to capabilities of agents? is it not already
achieved? Then, the goal is sent to HATP which produces a first plan.
A goal is considered achievable as long as the planner computes a
valid plan and it is not abandoned by the human.

Plan execution consists in the management of all the actions of the
plan. Human and robot are not acting both at the same time.  In case
of plan failure a new plan is requested and executed.

The management at the action level is done in three steps. First, the
action preconditions are tested over the current state of the
world. Then the action is executed and monitored (only monitored of it
corresponds to a human action).  Finally, the expected effects are
verified in order to acknowledge the action achievement.

Concerning the speech acts during plan management, the robot informs
its partners on the goal existence and status, plan existence and
status, ongoing plan action, plan action failure, failing facts for
action precondition or effect assessment.

\begin{figure}[thpb]
  \centering
  \includegraphics[width=0.8\columnwidth]{./figs/plan_management.pdf}
  \caption {Automaton for goal and plan management}
  \label{goal_plans_fg}
\end{figure}

\section{Action execution and monitoring}\label{sec:action}

The action execution and monitoring task involves the motion and task
planning, the effectors and the dedicated robot controller activity
(Figure~\ref{architecture_fg}).

%\subsubsection*{Human aware motion, placement and manipulation planning}
\vspace{0.3cm}
\noindent
\textbf{Human-aware motion, placement and manipulation planning:}
In our scenarios, actions are only object manipulation actions: Pick,
Place, Throw, Give. Motion and Task planning allows to compute final
object placement, grasp and arm motion trajectories taking into
account task specific constraints and human postures, abilities and
preferences: see \cite{Sisbot2008, Mainprice2011, Pandey2010} for
details.

%\subsubsection*{Action Execution and Monitoring Robot Controller}
\vspace{0.3cm}
\noindent
\textbf{Execution and Monitoring Robot Controller:} Depending on the
context and on the shared plan elaborated by HATP for a given goal,
the robot controller decides to execute an action or to ask its human
partner to do it.  Actions feasibility by the human or the robot are
regularly reconsidered based on the reachability / visibility
computation mechanisms.

Robot action execution is based on simple automatons  that translate
symbolic planning atomic actions into sequences of planned arm motions
and gripper commands to execute according to current state
of the 3-tuple (gripper, object, furniture). We have three states
according to whether the object is in gripper and if it is in gripper
whether it is on furniture.  These states are directly obtained from
the updated symbolic state of the world in the ontology.

For plan action monitoring, primitive actions recognition defined
above are used. A primitive action detection is interpreted as action
success if it is the expected one and failure otherwise. The robot
also reacts to the absence of activity.

\section{An illustrative example}\label{sec:expes}

We assume here that the robot (and the human) has been given the joint goal
``CLEAN TABLE''. For HATP, this means putting all tapes that are
currently on the table in the trashbin. Depending on the state of the
world and agent preferences, different plans are produced.

Figure~\ref{plan-etat2} shows a plan produced to clean the table based on the initial the initial state given in  \S~\ref{sub:gtrc}.

\begin{figure*}[thpb]
  \centering
  \includegraphics[width=1.0\textwidth]{./figs/plan3.pdf} \\
  \caption {A plan produced by HATP for clean the table based on the initial state given in \S~\ref{sub:gtrc}}
  \label{plan-etat2}
\end{figure*}

Let us now take a simpler example to illustrate a full run of the
system. We have only one tape on the table and it is is reachable only
by the robot while the throw position on top of the trashbin is
reachable only by the human.

% \begin{landscape}
% \begin{figure*}[thpb]
%   \centering
% %  \includegraphics[width=1.0\textwidth]{./figs/manip_run.pdf} \\
%   \includegraphics[width=17cm]{./figs/manip_run.pdf} \\
% \begin{center}
%   \caption {An example of a human-robot collaborative goal achievement}
% \end{center}
%   \label{manip_run_fg}
% \end{figure*}
% \end{landscape}

\begin{figure*}[thpb]
  \centering
\includegraphics[width=1.0\textwidth]{./figs/rotated-rachid_manip_run-v2.pdf} \\
%  \includegraphics[width=17cm]{./figs/manip_run-vertical.pdf} \\
\begin{center}
  \caption {An example of a human-robot collaborative goal achievement}
\end{center}
  \label{manip_run_fg}
\end{figure*}

%\subsection{successful execution}

Figure~\ref{manip_run_fg} illustrates the main processes occurring
during a multi-step human-robot collaborative goal achievement.  The
plan produced is quite straightforward and is shown in the third row
called ``Goal and Plan''. It consists in 4 successive actions
involving the robot and the human. Robot grasps the tape and then
places it on the table at a position where it is visible and reachable
for the human. Human then is asked to pick the tape and throw it in
the trashbin. The first row, named ``Cameras'', shows several
snapshots corresponding to various execution steps. Snapshot~1
presents the initial situation. Snapshots~2, 3, 4 and 5 give the state
after the successive achievement of the four actions in the plan. The
second row, named ``3D Model'', shows the display of SPARK at the same
instants. The fourth, row called ``Robot Speech Acts'', gives robot
speech acts produced along the execution to inform the human partner
about goal and plan creation and status and to verbalize the actions
that the human is asked to execute. The fifth row illustrates robot
knowledge on itself and on the objects. The sixth row illustrates the
robot knowledge about the human state. The seventh row gives ongoing
robot action with action preconditions and effects assessment as well
as motion execution tasks. Motion trajectory typology can be found
between the 3D Model views. The eighth row gives ongoing human action
with action preconditions and effects assessment and monitoring
activity.

\section{Conclusion and future work}

In this paper we have presented a decisional framework designed for
robots operating in a human environment. Our objective is to provide a
management of human-robot interaction that is an integral part of a
general robot control architecture.  This was done in order to provide
a principled way to deal with HRI scenario.  The framework is also
suitable for the development and experiment of task planners and
interaction schemes that explicitly consider human abilities and
preferences.

The design choices and the results presented here are still preliminary.
While the general scheme we propose might be difficult to implement in
a general sense, we believe that it is a reasonable challenge to
implement it in the case of a personal robot assistant essentially
devoted to fetch-and-carry, as well as for interactive manipulation
tasks and associated activities.\\

\vspace{0.3cm}
\noindent
{\bf Acknowledgements:}
This work has been conducted within the EU CHRIS project ({\tt
  \url{http://www.chrisfp7.eu/}}) funded by the E.C. Division FP7-IST
under Contract 215805. We also would like to thank warmly a number of LAAS
robotics team members and more particularly Xavier Broquere, Mokhtar
Gharbi, Wuwei He, Matthieu Herrb, Jim Mainprice, Raquel Ros, Daniel
Sidobre, Thierry Sim\'{e}on, Muhammad Ali.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\bibliographystyle{alpha}
\bibliographystyle{plain}
\bibliography{laas_hri}
%% \end{thebibliography}

\end{document}

\section{Discussion and future work}

The design choices and the results presented here is still preliminary. 
While the general scheme we propose might be difficult to implement in
a general sense, we believe that it is a reasonable challenge to
implement it in the case of a personal robot assistant essentially
devoted to fetch-and-carry, as well interactive manipulation tasks and
associated activities. The robot would operates in a known in-door
environment (acquired in a preliminary phase).

In such contexts, it is possible ....

Fetch-and-carry and object manipulation task need 3D geometric
planning.  One challenging problem would be to extend the approach
discussed above to the situation where a robot has to hand
an object to human. Indeed, there is a need to take into account
visibility and reach, in terms of kinematic constraints, of the human
partner (Figure~\ref{fig_handing}).

%\input{conclusion} %% Conclusion and Future Work
\section{Conclusion}\label{sec:Conclusion}

In this paper we have presented a decisional framework designed for
robots operating in a human environment. Our objective is to provide a
management of human interaction that is an integral part of a general
robot control architecture.  This was done in order to provide a
principled way to deal with HRI.

The framework is also suitable for the development and experiment of
task planners and interaction schemes that explicitly consider
human abilities and preferences.

Examples of such planners are presented that integrate various models
of human abilities and preferences.

\end{document}

### 2013-04-23 paper.tex
Updated several figures
Updated style for Journal of Artificial Intelligence

@@
-%%%%%%%%%%%%%%%%%%%% author.tex %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-% sample root file for your "contribution" to a contributed volume
-%
-% Use this file as a template for your own input.
-%
-%%%%%%%%%%%%%%%% Springer %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\documentclass{elsarticle}
 
-% RECOMMENDED %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\documentclass[graybox]{svmult}
-
-% choose options for [] as required from the list
-% in the Reference Guide
-
-\usepackage{mathptmx}       % selects Times Roman as basic font
-\usepackage{helvet}         % selects Helvetica as sans-serif font
-\usepackage{courier}        % selects Courier as typewriter font
-\usepackage{type1cm}        % activate if the above 3 fonts are
-                            % not available on your system
-%
-\usepackage{makeidx}         % allows index generation
 \usepackage{graphicx, subfigure}        % standard LaTeX graphics tool
                              % when including figure files
-\usepackage{multicol}        % used for the two-column index
-\usepackage[bottom]{footmisc}% places footnotes at page bottom
-
+\usepackage{url}
 \usepackage{fixme}
@@
 
-% see the list of further useful packages
-% in the Reference Guide
-
-\makeindex             % used for the subject index
-                       % please use the style svind.ist with
-                       % your makeindex program
+\journal{Journal of Artificial Intelligence}
 
@@
 \begin{document}
+\begin{frontmatter}
 
-\title*{When the robot considers the human... }
+\title{When the robot considers the human... }
 % Use \titlerunning{Short Title} for an abbreviated version of
@@
 \author{Rachid Alami, Mathieu Warnier, Julien Guitton, S\'{e}verin Lemaignan and Emrah Akin Sisbot}
-\authorrunning{R. Alami, M. Warnier, J. Guitton, S. Lemaignan and E.A. Sisbot}
 % Use \authorrunning{Short Title} for an abbreviated version of
 % your contribution title if the original one is too long
-\institute{ LAAS-CNRS\\ 
+\address{ LAAS-CNRS\\ 
         Universit\'e de Toulouse, UPS, INSA, INP, ISAE, LAAS\\ 
-         F-31077 Toulouse, France\\ \email{Firstname.Name@laas.fr}}
-%
-% Use the package "url.sty" to avoid
-% problems with special characters
-% used in your e-mail or web address
-%
-\maketitle
+         F-31077 Toulouse, France\\ \texttt{Firstname.Name@laas.fr}}
 
-\abstract*{This paper addresses some key decisional issues that are
+\begin{abstract}
+  This paper addresses some key decisional issues that are
   necessary for a cognitive robot which shares space and tasks with a
@@
   model, human-aware task planning and human and robot interleaved
-  plan achievement.}
+  plan achievement.
+\end{abstract}
 
-\abstract{This paper addresses some key decisional issues that are
-  necessary for a cognitive robot which shares space and tasks with a
-  human. We adopt a constructive approach based on the identification
-  and the effective implementation of individual and collaborative
-  skills. The system is comprehensive since it aims at dealing with a
-  complete set of abilities articulated so that the robot controller
-  is effectively able to conduct in a flexible manner a collaborative
-  task with a human partner. These abilities include geometric
-  reasoning and situation assessment based essentially on
-  perspective-taking and affordances, management and exploitation of
-  each agent (human and robot) knowledge in a separate cognitive
-  model, human-aware task planning and human and robot interleaved
-  plan achievement.}
+\begin{keyword}
+    %% keywords here, in the form: keyword \sep keyword
+
+    %% MSC codes here, in the form: \MSC code \sep code
+    %% or \MSC[2008] code \sep code (2000 is the default)
+
+\end{keyword}
+
+\end{frontmatter}
 
@@
   \centering
-  \includegraphics[width=1.0\textwidth]{./figs/supervisor-architecture-new.pdf} \\
+  \includegraphics[width=1.0\textwidth]{./figs/archi.pdf} \\
   \caption {Architecture of the robot control system}
@@
 
-% \begin{landscape}
-% \begin{figure*}[thpb]
-%   \centering
-% %  \includegraphics[width=1.0\textwidth]{./figs/manip_run.pdf} \\
-%   \includegraphics[width=17cm]{./figs/manip_run.pdf} \\
-% \begin{center}
-%   \caption {An example of a human-robot collaborative goal achievement}
-% \end{center}
-%   \label{manip_run_fg}
-% \end{figure*}
-% \end{landscape}
-
 \begin{figure*}[thpb]
   \centering
-\includegraphics[width=1.0\textwidth]{./figs/rotated-rachid_manip_run-v2.pdf} \\
-%  \includegraphics[width=17cm]{./figs/manip_run-vertical.pdf} \\
+    \includegraphics[width=1.0\textwidth]{./figs/manip_run.pdf} \\
     \begin{center}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\bibliographystyle{alpha}
-\bibliographystyle{plain}
+\bibliographystyle{elsarticle-num}
 \bibliography{laas_hri}
-%% \end{thebibliography}
 

### 2013-04-23 paper.tex
Enriched the 'LAAS deliberative layer' part

@@
 
-\section{A Decisional framework}\label{sec:Framework}
+\section{The LAAS deliberative layer}
+\label{sec:Framework}
+
+Fig.~\ref{fig|archi} give an overview of the connections between the
+deliberative components of our architecture~\cite{Alami2011}.
+
+\begin{figure*}
+        \centering
+        \includegraphics[width=1.0\columnwidth]{figs/archi}
+        \caption{Overview of the LAAS deliberative layer. Knowledge is
+        centrally managed in an active \emph{semantic blackboard}, pictured
+        above with a thick border.}
+        \label{fig|archi}
+\end{figure*}
+
+This architecture moves away from standard layered approaches. Interactions
+between components are mostly bidirectional and, from the software components
+point of view, we do not introduce layers of abstraction (we do, however, have
+access to the lower level modules of the robot to execute
+actions~\cite{Alami1998}, but all cognition-related modules reside at the same
+level). This is especially visible for the dialogue input processing. This
+component does not simply act as an alternative perceptual input to the
+symbolic database, but also actively queries previously acquired knowledge to
+disambiguate and validate the newly created symbolic knowledge (see
+section~\ref{sect|com}).
+
+Our architecture relates but is to be distinguished from \emph{Beliefs,
+Desires, Intentions} (BDI) architectures. BDI architectures are primarily
+focused on \emph{practical reasoning}, \ie the process of deciding, step by
+step, which action to perform to reach a goal (as summarised by
+Woolridge~\cite{Woolridge1999}). The management of the interaction between
+knowledge (the beliefs) and task and plan representation and execution (the
+desires and the intentions) is central, and aims at selecting at each step the
+best subgoal. It becomes then an intention that the robot commits to.
+
+This interaction between knowledge and actions is also central to our approach
+(as for any cognitive system), but task representation and task execution is
+not seen as a monolithic, central function: it is one of the activities of the
+robot, actually split between communication components (that can acquire
+desires from interaction with agents, amongst other things) and an execution
+controller that may decide to take an incoming desire into account to create
+its own internal goals. The controller generates and controls intentions from
+these goals with the help of a symbolic task planner, that has also direct
+access to the knowledge base.
+
+The architecture is not only focused on this workload, and other
+activities are conducted without being explicitly considered as desires:
+assessment of the situation and the environment, dialogue (including
+performative dialogue that can possibly change the internal state of the robot,
+but does not lead to the creation of desires,  like question answering or
+statement assertion), various background monitoring and recognition tasks, etc.
+
+Regarding the anchoring question, this architecture is bidirectional. The
+components we described provide a \textit{bottom-up} grounding process:
+geometric reasoning and dialogue processing modules constantly build and push
+new symbolic contents about the world to the knowledge base where it becomes
+accessible to decisional layers. In parallel, the knowledge base relies on
+reasoning in a \textit{top-down} way to produce new facts that may trigger in
+return physical behaviours.
+
+\subsection*{Knowledge model}
+
+In our architecture (Fig.~\ref{fig|archi}), knowledge manipulation relies on a
+\emph{semantic blackboard}: one central server (the {\sc Oro}
+server~\cite{Lemaignan2010}) stores knowledge as it is produced by each of the
+deliberative components. It conversely exposes an API to query the knowledge
+base.
+
+Knowledge is represented as RDF triples in the OWL sub-language. Each
+time triples are added or removed from the knowledge base, a Description
+Logics reasoner ({\sc Pellet}\footnote{\url{http://clarkparsia.com/pellet/}})
+classifies the whole ontology and inserts all possible inferred triples.
+
+Relying on RDF triples and Description Logics has major advantages like the
+availability of numerous open-source libraries to manipulate the ontology,
+interoperability with several major on-line knowledge bases (like {\sc
+OpenCyc}, {\sc WordNet} or {\sc DBPedia}), open-world reasoning, and the formal
+guarantee of decidability (it is always possible to classify a Description
+Logics ontology).
+
+It has also notable limitations: RDF triples imply only binary predicates
+(\stmt{subject predicate object}), which constrains the expressiveness of the
+system or leads to cumbersome reifications. Alternative exists (like {\sc
+KnowRob}~\cite{Tenorth2009a}) that mixes RDF with more expressive logic
+languages like {\sc Prolog}, at the price, however, of other limitations, like
+closed-world reasoning or immutable T-Box. The classification
+performance is another issue: from our experience, with an ontology sized for a
+standard experiment (about 100 classes and 200 instances), classification
+typically take about 100ms, which becomes problematic during interactions.
+Besides, the performances are difficult to predict, since a seemingly
+inoffensive new statement may indirectly change radically the logical
+complexity of the whole knowledge model and lead to notable degradation of
+classification time.
+
+This knowledge model also largely exclude representation of continuous
+phenomena (like time) or uncertain phenomena. When required, these have to be
+managed by dedicated components.
+
+Amongst the features of {\sc Oro}, the system can dynamically creates
+new independent knowledge models (\ie independent instantiations of the
+ontology). This feature is used for instance to store separately the beliefs of
+each of the agents the robot interacts with (see section~\ref{sect|tom}).
+
+\subsection*{Interaction model}
 
@@
 
-Our robot is controlled by a three layer
-architecture~\cite{Alami1998}.  We present briefly its decisional
-layer. The proposed decisional framework consists of several entities,
-having each a specific role as illustrated by
-Figure~\ref{architecture_fg}.  We describe how the robot is controlled
-through an analysis of the three main activities performed by the
-robot controller:
-
-\begin{figure*}[thpb]
-  \centering
-  \includegraphics[width=1.0\textwidth]{./figs/archi.pdf} \\
-  \caption {Architecture of the robot control system}
-  \label{architecture_fg}
-\end{figure*}
+Next sections describe how the robot is controlled through an analysis of the
+three main activities performed by the robot controller:
 
@@
 
-The next three sections describe the three robot controller
-activities and how they make use of a number of key components in the
-architecture:
+These activities are caried out by different software composants:
 
 \begin{itemize}
-\item SPARK: Spatial Reasoning and Knowledge module \cite{Sisbot2011}
-\item ORO: a knowledge management module \cite{Lemaignan2010}
-\item HATP: a Human-Aware Task Planner \cite{Alili2008}
-\item A set of Human aware motion, placement and manipulation planners
- \cite{Sisbot2008, Mainprice2011, Pandey2010}
+
+    \item Situation assessment is managed by \textsc{Spark} (Spatial Reasoning
+        and Knowledge module)~\cite{Sisbot2011},
+
+    \item HATP (Human-Aware Task Planner)~\cite{Alili2008} is responsible for
+        symbolic task planning,
+
+    \item We finally also rely on a set of human-aware motion, placement and
+        manipulation planners, grouped under the \textsc{Move3D}
+        umbrella~\cite{Sisbot2008, Mainprice2011, Pandey2010}.
+
 \end{itemize}

### 2013-04-23 paper.tex
Complete biblio

@@
 
-Spatial reasoning~\cite{OKeefe1999}, on the other hand, has been used
+Spatial reasoning~\cite{O'Keefe1999}, on the other hand, has been used
 for natural language processing for applications such as direction
-recognition ~\cite{Kollar10,Matuszek10} or language
-grounding~\cite{Tellex10}.~\cite{Skubic2004} presented a spatial
+recognition ~\cite{Kollar2010,Matuszek2010} or language
+grounding~\cite{Tellex2010}.~\cite{Skubic2004} presented a spatial
 reasoner integrated in a robot which computes symbolic positions of
@@
 \centering
-\includegraphics[width=9cm]{figs/grounding_robot.pdf}
+\includegraphics[width=0.9\columnwidth]{figs/grounding_robot.pdf}
 \caption{Robot reasoning about HRI and anticipation of human activities:

### 2013-04-23 paper2.tex
Rebased the paper on Severin's IROS2013 paper

--- 2013-04-23 paper.tex
@@
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%2345678901234567890123456789012345678901234567890123456789012345678901234567890
+%        1         2         3         4         5         6         7         8
+
 \documentclass{elsarticle}
 
-\usepackage{graphicx, subfigure}        % standard LaTeX graphics tool
-                             % when including figure files
+\usepackage{cite}
+\usepackage{graphicx, subfigure}
+%\usepackage{hyperref}
 \usepackage{url}
-\usepackage{fixme}
-\usepackage{lscape}
+\graphicspath{{figs/}}
+\DeclareGraphicsExtensions{.pdf,.jpg,.png}
 
-\journal{Journal of Artificial Intelligence}
+\usepackage{ifthen}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\usepackage[utf8x]{inputenc}
+\usepackage[T1]{fontenc}
 
-%% Custom macros
-\newcommand{\stmt}[1]{{\footnotesize $\langle$\stmttt#1\relax$\rangle$}}
-\newcommand{\rawstmt}[1]{{\footnotesize \stmttt#1\relax}}
-\def\stmttt#1 #2 #3\relax{{\tt#1} {\bf{\tt #2}} {\tt #3}}
+\usepackage[draft]{fixme}
 
-\newcommand{\setstmt}[1]{{\footnotesize [\setstmttt#1\relax]}}
-\def\setstmttt#1,#2\relax{\rawstmt{#1}, \rawstmt{#2}}
+\usepackage{listings}
+\usepackage{alltt}
+\usepackage{fancyvrb}
+\usepackage{flushend} % equalize the column length on last page
 
-\newcommand{\concept}[1]{{\footnotesize \texttt{#1}}}
+\newcommand{\concept}[1]{{\small \texttt{#1}}}
 
-\newcommand{\ie}{{\textit{i.e.~}}}
-\newcommand{\cf}{{\textit{cf~}}}
-\newcommand{\eg}{{\textit{e.g.~}}}
+\newcommand{\stmt}[1]{{\footnotesize \tt $\langle$ #1\relax$\rangle$}}
+%\newcommand{\stmt}[1]{{\footnotesize $\langle\stmttt#1\relax\rangle$}}
+%\newcommand{\rawstmt}[1]{{\footnotesize \stmttt#1\relax}}
+%\def\stmttt#1 #2 #3\relax{
+%\texttt{\IfBeginWith{#1}{?}{\textbf{#1}}{#1} \IfBeginWith{#2}{?}{\textbf{#2}}{\emph{#2}} \ifthenelse {\equal{#3}{true} \OR \equal{#3}{false}}{\emph{#3}}{\IfBeginWith{#3}{?}{\textbf{#3}} {#3}}}}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%\newcommand{\setstmt}[1]{{\footnotesize [\setstmttt#1\relax]}}
+%\def\setstmttt#1,#2\relax{\stmttt#1\relax, \stmttt#2\relax}
+
+\newcommand{\ie}{{\textit{i.e.\ }}}
+\newcommand{\cf}{{\textit{cf\ }}}
+\newcommand{\eg}{{\textit{e.g.\ }}}
+
+\begin{frontmatter}
+
+\title{\LARGE \bf
+Explicit Knowledge and the Deliberative Layer: Lessons Learned
+}
+
+\author{Séverin Lemaignan and Rachid Alami}
+
+\address{
+CNRS, LAAS, 7 avenue du Colonel Roche, F-31400 Toulouse, France\\
+Univ de Toulouse, LAAS, F-31400 Toulouse, France\\
+{\tt severin.lemaignan@laas.fr}, {\tt rachid.alami@laas.fr}
+}
 
 \begin{document}
-\begin{frontmatter}
 
-\title{When the robot considers the human... }
-% Use \titlerunning{Short Title} for an abbreviated version of
-% your contribution title if the original one is too long
-\author{Rachid Alami, Mathieu Warnier, Julien Guitton, S\'{e}verin Lemaignan and Emrah Akin Sisbot}
-% Use \authorrunning{Short Title} for an abbreviated version of
-% your contribution title if the original one is too long
-\address{ LAAS-CNRS\\ 
-        Universit\'e de Toulouse, UPS, INSA, INP, ISAE, LAAS\\ 
-         F-31077 Toulouse, France\\ \texttt{Firstname.Name@laas.fr}}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{abstract}
 
-\begin{abstract}
-  This paper addresses some key decisional issues that are
-  necessary for a cognitive robot which shares space and tasks with a
-  human. We adopt a constructive approach based on the identification
-  and the effective implementation of individual and collaborative
-  skills. The system is comprehensive since it aims at dealing with a
-  complete set of abilities articulated so that the robot controller
-  is effectively able to conduct in a flexible manner a collaborative
-  task with a human partner. These abilities include geometric
-  reasoning and situation assessment based essentially on
-  perspective-taking and affordances, management and exploitation of
-  each agent (human and robot) knowledge in a separate cognitive
-  model, human-aware task planning and human and robot interleaved
-  plan achievement.
+This paper addresses some key decisional issues that are necessary for a
+cognitive robot which shares space and tasks with a human. We adopt a
+constructive approach based on the identification and the effective
+implementation of individual and collaborative skills. The system is
+comprehensive since it aims at dealing with a complete set of abilities
+articulated so that the robot controller is effectively able to conduct in a
+flexible manner a collaborative task with a human partner. These abilities
+include geometric reasoning and situation assessment based essentially on
+perspective-taking and affordances, management and exploitation of each agent
+(human and robot) knowledge in a separate cognitive model, human-aware task
+planning and human and robot interleaved plan achievement.
+
+This article presents our design choices, the articulations between the diverse
+deliberative components of the robot, and the strengths and weaknesses of our
+approach. We show that explicit knowledge management is not only a convenient
+tool from the software engineering point of view, but also pushes for a
+different, more \emph{semantic} way to address the decision-making issue in
+human-robot interactions.
+
 \end{abstract}
@@
 
-\section{Introduction}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{The challenge of human-robot interaction}
 
@@
 
-\section{A typical human-robot interaction problem}\label{sec:problem}
+\subsection{The typical human-robot interaction problem}\label{sec:problem}
 
@@
 
-\section{Related work and vision}\label{sec:soa}
+\begin{figure}[htb]
+\centering
+\includegraphics[width=0.9\columnwidth]{figs/grounding_robot.pdf}
+\caption{Robot reasoning about HRI and anticipation of human activities:
+  sources of information are multi-modal dialogue, and observation of
+  environment and human activity}
+\label{fig:hri-dec}
+\end{figure}
+
+We envision HRI in a context where two agents (a human and a robot)
+share a common space and exchange information through various
+modalities. Our aim is to endow the robot with an explicit
+consideration of the human and with the ability to manage its
+interactions with him (Figure~\ref{fig:hri-dec}). This must be
+considered at the architecture level as well as at the task/motion
+planning and execution level.
+
+We have devised a decisional framework for human-robot interactive
+task achievement that is aimed to allow the robot not only to
+accomplish its tasks but also to produce behaviors that support its
+engagement vis-a-vis its human partner and to interpret human
+behaviors and intentions. 
+Together and in coherence with this framework, we have developed
+and experimented various task planners and interaction schemes that
+allow the robot to select and perform its tasks while taking into
+account explicitly the human abilities as well as the constraints
+imposed by the presence of humans, their needs and preferences.
+
+\subsection{Towards the cognitive robot at LAAS}
+
+Natural interaction and cooperation are on the (dare we say,
+\emph{short-term}) agenda for the human-robot interaction community. They are
+keys to the broad class of \emph{interactive manipulation problems}: several
+agents agree on a (more or less implicit) joint goal that requires some sort of
+cooperation to be successfully achieved. This class of problems involves both
+dialogue and manipulation and they are often initially underspecified: they
+require iterative and interactive resolution.
+
+Over the last years we have focused our efforts on identifying the cognitive
+prerequisites of these challenges, and giving them experimental reality on the
+robots: what is required for sentences like ``Let's dress the table together''
+to be understood by the robot, correctly interpreted in the spatial and
+temporal context of the interaction, and eventually transformed into a set of
+actions.
+
+We have chosen to tackle the challenge from several ends: human-aware
+navigation and motion planning (\cite{Mainprice2011}), situation assessment
+coupled with motion planning (\cite{Mainprice2012}), projection of
+``mightabilities'' that anticipates what surrounding agents may do
+(\cite{Pandey2011}), and the design and deployment of a pervasive
+\emph{knowledge-centric} software architecture.
+
+The last item is the focus of this paper: what does ``knowledge-centric''
+architecture means, and what are its benefits and drawbacks.
+
+\subsection{Related work}\label{sec:soa}
 
@@
 
-\begin{figure}[htb]
-\centering
-\includegraphics[width=0.9\columnwidth]{figs/grounding_robot.pdf}
-\caption{Robot reasoning about HRI and anticipation of human activities:
-  sources of information are multi-modal dialogue, and observation of
-  environment and human activity}
-\label{fig:hri-dec}
-\end{figure}
+\subsection{Article overview}
 
-We envision HRI in a context where two agents (a human and a robot)
-share a common space and exchange information through various
-modalities. Our aim is to endow the robot with an explicit
-consideration of the human and with the ability to manage its
-interactions with him (Figure~\ref{fig:hri-dec}). This must be
-considered at the architecture level as well as at the task/motion
-planning and execution level.
+In the next section, we first introduce the deliberative layer of our robots at
+the functional level. It represents what we call a \emph{knowledge-oriented}
+architecture, built around a central knowledge management tool called the {\sc
+Oro} server.
 
-We have devised a decisional framework for human-robot interactive
-task achievement that is aimed to allow the robot not only to
-accomplish its tasks but also to produce behaviors that support its
-engagement vis-a-vis its human partner and to interpret human
-behaviors and intentions. 
-Together and in coherence with this framework, we have developed
-and experimented various task planners and interaction schemes that
-allow the robot to select and perform its tasks while taking into
-account explicitly the human abilities as well as the constraints
-imposed by the presence of humans, their needs and preferences.
+The following sections go over each of the facets of this deliberative layer:
+how knowledge is produced asynchronously from geometric reasoning, how the
+robot conducts grounded multi-modal interaction with humans, how knowledge is
+used by decision-making components like the robot controller or the task
+planner, and finally, how our knowledge-centric architecture enables the
+implementation of specific internal cognitive processes.
+
+We conclude the article with a discussion about the strengths and weaknesses of
+such a knowledge-oriented architecture.
 
 \section{The LAAS deliberative layer}
-\label{sec:Framework}
 
@@
         \centering
-        \includegraphics[width=1.0\columnwidth]{figs/archi}
+        \includegraphics[width=1.0\columnwidth]{archi}
         \caption{Overview of the LAAS deliberative layer. Knowledge is
@@
 point of view, we do not introduce layers of abstraction (we do, however, have
-access to the lower level modules of the robot to execute
-actions~\cite{Alami1998}, but all cognition-related modules reside at the same
-level). This is especially visible for the dialogue input processing. This
-component does not simply act as an alternative perceptual input to the
-symbolic database, but also actively queries previously acquired knowledge to
-disambiguate and validate the newly created symbolic knowledge (see
-section~\ref{sect|com}).
+access to the lower level modules of the robot to execute actions, but all
+cognition-related modules reside at the same level). This is especially visible
+for the dialogue input processing. This component does not simply act as an
+alternative perceptual input to the symbolic database, but also actively
+queries previously acquired knowledge to disambiguate and validate the newly
+created symbolic knowledge (see section~\ref{sect|com}).
 
@@
 
-\subsection*{Knowledge model}
+\subsection{Knowledge model}
 
@@
 
-\subsection*{Interaction model}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Situation Assessment and Context Management}
+\label{sect|sit-ass}
 
-Interaction happens as a consequence of an explicit request of the
-human to satisfy a goal or because the robot finds itself in a
-situation where it is useful if not mandatory. In both cases, the
-robot has a goal to satisfy.  An important issue is the notion of
-engagement, a process in which the robot will have to establish,
-maintain and terminate a connection with a human partner. 
-This covers goal establishment, selection of an incremental refinement
-of the task that is intended to be achieved, and execution control
-including monitoring, and even influencing, human task performance and
-his/her commitment to the goal. The human involvement may range from a
-direct participation to the task achievement, to a simple
-``acceptance'' of robot activity in his/her close vicinity.
+\begin{figure}
+        \centering
+        \includegraphics[width=\columnwidth]{spark-oro}
+    \caption{Functional overview of knowledge base ({\sc Oro} server, top part) and the geometric situation assessment module (\emph{SPARK}, bottom part)}
+        \label{fig|spark-oro}
+\end{figure}
 
-Next sections describe how the robot is controlled through an analysis of the
-three main activities performed by the robot controller:
+Anchoring perceptions in a symbolic model requires perception abilities and
+their symbolic interpretation. We rely on a dedicated geometric and temporal
+reasoning module (as named on Fig.~\ref{fig|archi}) called SPARK (\emph{SPAtial
+Reasoning \& Knowledge}~\cite{Sisbot2011}). It is a situation assessment
+reasoner that generates symbolic knowledge from the geometry of the
+environment with respect to relations between objects, robots and humans,
+also taking into account the different perspective that each agent has on
+the environment (Fig.~\ref{fig|spark-oro}).
 
-\begin {enumerate}
-\item Situation assessment and context management 
-\item Goals and plans management
-\item Action refinement, execution and monitoring
-\end {enumerate}
+SPARK is an \emph{amodal} geometric model of the environment that serves both
+as basis for the fusion of the perception modalities and as bridge with the
+symbolic layer. This geometric model is continuously updated at run-time by the
+robot based on its sensors. In our experiments, objects are identified and
+localised through 2D barcodes.  On the other hand, the human pose is tracked
+with a Kinect-like device (assisted by motion capture to accurately track the
+head motion, which is required to compute what the human is looking at).
 
-These activities are caried out by different software composants:
+\subsection{Situation Assessment}
 
-\begin{itemize}
+\subsubsection*{Symbolic facts production:}
 
-    \item Situation assessment is managed by \textsc{Spark} (Spatial Reasoning
-        and Knowledge module)~\cite{Sisbot2011},
-
-    \item HATP (Human-Aware Task Planner)~\cite{Alili2008} is responsible for
-        symbolic task planning,
-
-    \item We finally also rely on a set of human-aware motion, placement and
-        manipulation planners, grouped under the \textsc{Move3D}
-        umbrella~\cite{Sisbot2008, Mainprice2011, Pandey2010}.
-
-\end{itemize}
-
-Other decisional activities, such as situated dialog (\cite{Ros2010b,
-  Lemaignan2011}, not presented here) have been developed that use the
-same set of components.
-
-\section{Situation assessment and context management}\label{sec:situ}
-
-This activity involves the geometric and temporal reasoning component,
-the symbolic facts and belief management component and the dedicated
-robot controller activity (Figure~\ref{architecture_fg}).
-
-We assume that perception provides in real-time the identity and the
-position of objects when they are in the field of view of the sensors.
-In our implemented examples, the robot is localised using a standard
-horizontal laser-scanning based localisation system, the objects are
-identified and localized using ARToolkit \cite{ARToolkit} and the
-humans are tracked using a commercial motion capture system and a
-Kinect device from Microsoft.
-
-\subsection{Geometric and Temporal Reasoning component}\label{sub:gtrc}
-
-The geometric reasoning component plays a central role in our
-architecture. It is called SPARK (Spatial Reasoning and
-Knowledge~\cite{Sisbot2011}) in the current implementation. It is
-responsible for geometric information gathering and it embeds a number
-of decisional activities linked to abstraction (symbolic facts
-production) and inference based on geometric and temporal reasoning.
-SPARK maintains all geometric positions and configurations of agents,
-objects and furniture coming from perception and previous or {\it a
-  priori} knowledge.
-
- %\subsubsection*{Symbolic facts production:} 
-\vspace{0.3cm}
-\noindent
-\textbf{Symbolic facts production:} 
 Geometric state of the world is abstracted in symbolic facts that can
@@
 
-%\newline\newline
-%\subsubsection*{Hypotheses on objects states and positions}
-\vspace{0.3cm}
-\noindent
-\textbf{Hypotheses on objects states and positions:}
+\subsubsection*{Hypotheses on objects states and positions}
 It is sometimes difficult or even impossible to see and/or track an
@@
 
-%\subsubsection*{Primitive action recognition}
-\vspace{0.3cm}
-\noindent {\bf Primitive action recognition:} 
+\subsubsection*{Primitive action recognition}
 Monitoring human activity is crucial to maintain a coherent state of
@@
 objects or above containers.
+
+\subsection{Ego-centric and allo-centric frames}
+
+SPARK enables \emph{perspective-taking}: spatial relations between entities can
+be computed from different viewpoints, which let the robot build for each
+agents it interacts with a different, \emph{perspective-aware} symbolic model
+of the environment. These models are separately stored in the knowledge base.
+
+This allows us to deal with ambiguities that arises when one speaker refers to
+an object within a reference system (or changes the reference system, \ie
+switches perspective) without making the reference frame
+explicit~\cite{Breazeal2006, Ros2010}.
+
+As a result, the robot stores models on the environment either in the
+\emph{ego-centric} reference frame (from the robot perspective) or in
+addressee-centred (\ie, \emph{allo-centric}) frames.
 
@@
 triggered when specific events occur. A component can
-for instance subscribe to events of kind \setstmt{?agent isVisible
+for instance subscribe to events of kind \stmt{?agent isVisible
   true, ?agent type Human}. As soon as the perception layer detects a
@@
 
-\section{Goal and Plan Management}\label{sec:plan}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Communication}
+\label{sect|com}
 
-The Goal and Plan Management activity involves the human-aware
-symbolic task planner component and the dedicated robot controller
-activity (Figure~\ref{architecture_fg}).
+\subsection{Natural language grounding}
 
-\subsection{Symbolic Task Planning}
+Natural language processing is one of the fields of human-robot interaction
+where the introduction of the semantic layer has been most beneficial.
+We detail in~\cite{Lemaignan2011a} the techniques and the tool called
+\texttt{dialogs} that we have developed for English language parsing and
+grounding, along with verbalisation and (minimalist) dialogue management.
+
+Natural language input (in experiments, we rely on an Android-based interface,
+with Google speech recognition) is parsed into a grammatical structure, and
+each sentence atoms are resolved with the help of the ontology to ground
+concepts like objects (\ie when a user says ``pick the can'', resolve to which
+instance of \emph{can} the user is referring to) and actions. Sentences are
+sorted into questions, desires and statements, and processed accordingly.
+
+The system supports quantification (``give me \{a | the | some | all | any | ...\} can''),
+thematic roles (action-specific predicates that qualify the actions), interactive disambiguation (the robot asks
+questions when it needs more information), anaphora resolution (``give \emph{it} to me'') based on
+dialogue history. It also permits knowledge extension by learning new semantic
+structures (for instance, a sentence like ``learn that cats are animals'' is
+converted into \stmt{Cat subClassOf Animal}, interprets common temporal and place adverbs (like \emph{above} or \emph{tomorrow}) and translates to a certain extend \emph{states} (``I'm tired'') into
+\emph{experiences} (\stmt{HUMAN experiences state\_1, state\_1 hasFeature
+tired}).
+
+\subsection{Limits of disambiguation at semantic level}
+
+One prototypical example of semantic disambiguation has been given in
+\cite{Ros2010b} with the child game \emph{spygame}: two players are facing
+each other with a set of random objects in-between, one player mentally choose
+one object, and the other player has to guess the object by asking closed
+questions like \emph{Is your object small or large?} Based on the knowledge it
+has acquired, the robot is able to minimize the number of questions required to
+find the object.
+
+When playing this kind of game, however, the issue arises that the robot has no
+way to select which knowledge about the object is relevant in the interaction
+context. For instance, the knowledge base may store facts like \stmt{obj1 type
+ActiveConcept} (which internally means that this concept was mentioned in a
+discussion in the last few seconds): this information is not a relevant
+property of \concept{obj1} when trying to disambiguate concepts with humans.
+This distinction between \emph{internal knowledge} (meaningful to
+the system only) and \emph{common knowledge} (whose meaning is understood by
+all the interactors) has not been properly dealt with in our architecture.
+
+Besides, even knowledge that belongs the \emph{common knowledge} may not be
+appropriate in a given interaction context. For instance, the system may
+compute that at a given instant the human is looking at the object: \stmt{human
+looksAt obj1}. This property makes sense to both parties, but in the context of
+the \emph{spygame}, we would like to mainly use immanent properties, not
+volatile like a gaze. More research is required to identify relevant
+interaction contexts and knowledge classes attached to them.
+
+\subsection{Multi-modal communication}
+
+Because all components rely on the same RDF formalism to format their outputs,
+the different communication modalities (\emph{explicit} like verbal, deictic or
+based on gaze, or \emph{implicit} like postures) are presented in an
+homogeneous way.
+
+The dialogue grounding process makes use of them at two distinct levels.
+
+First, particular steps of the grounding process explicitly check for the
+presence and value of specific facts: for instance, when a set of instances
+match a category (the human says ``give me the bottle'' and the robot knows
+about three bottles), the module may decide (it actually depends on the
+quantifier preceding the class) to discard some of them based on their
+\emph{visibility} by the speaker (implicit communication context built on the
+human posture).
+
+Another example, when the human says ``this'', the robot checks if the human is
+currently pointing at some object. In that case, \emph{this} is replaced by the
+object focused on.
+
+Note that, while the system benefits from the complementary modalities, they
+are not all required. The dialogue system would for instance run equally well
+with only the verbal modality.
+
+The second level of integration of multi-modality is implicit: by computing
+symbolic properties from the geometry, richer descriptions and hence
+discrimination possibilities are available: for instance, if reachability is
+available, the robot may asks ``do you mean the bottle that is accessible to
+me?'' to discriminate between the three bottles. That way, procedures relying
+on discrimination transparently benefit from added modalities.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Robot control}
+\label{sect|ctrl}
+
+\subsection{Desires and experiences}
+
+Our robot execution controllers (OpenPRS-based {\sc Shary} or Python-based {\sc
+pyRobots}) have deep integration with the knowledge base. It serves as the primary source of information for semantic-aware decision-making.
+
+We split the interaction situations stemming from the situation assessment and
+communication components in two categories: \emph{desires} (performative act)
+and \emph{experiences} (assertive act).
+
+\emph{Desires} are typically human orders (``Give me that book''). The nature
+of the desired action (to pick, give, look, bring, show...), along with the
+action parametrization (what is acted on? who should perform the action? etc.)
+are extracted from the knowledge base, and either passed to a task planner
+(presented in the next section) or executed if the procedure is directly
+available.
+
+\emph{Experiences}, on the other hand, comprise of emotions, states and
+questions (when asking a question, we consider the human to be in an
+\emph{interrogative state}). When the knowledge base recognizes that an agent
+\emph{experiences} a particular emotion or state, the execution controller may
+decide to handle it, typically by trying to answer the question or using the
+emotional or physical state as a parameter for subsequent actions. As an example,
+when the speaker says ``I feel tired'', we change the motion planner
+parametrization to lower the effort the human need to provide for the following
+joint manipulation tasks. Note that this example has been implemented as a
+proof-of-concept. We have not yet tried to define a theoretical framework that
+would support action alteration based on the user's experienced states.
+
+\subsection{Event-driven control}
+
+The {\sc Oro} server proposes two paradigms to access its content: RPC-style
+queries (based on the standard SPARQL language) or events. In its simplest
+form, a module can register an event by passing through an event pattern, which
+is basically a partial statement like \stmt{? type Elephant}, and a callback.
+Every time a new instance of elephant appears in the knowledge base, the
+callback is triggered.
+
+This allows us to write reactive robot controllers with a high level of
+expressiveness: for instance, by subscribing to the event \stmt{human1 desires
+?action, ?action type Look, ?action hasGoal myself}, we could trigger a
+behaviour when the human expresses (through dialogue, gestures...) that he
+wants to look at the robot itself.
+
+The robot controller designer does not need to directly care about how this
+\emph{desire} is produced (this is delegated to perception modules), it can
+focuses on the semantic of the desire.
+
+Note also that we take advantage of the reasoning capabilities of the system:
+for example, the goal of the action (\stmt{action hasGoal myself}) may well not
+have been explicitly asserted, but inferred by the reasoner based on other
+assertions.
+
+\subsection{Action execution and monitoring}\label{sec:action}
+
+The action execution and monitoring task involves the motion and task
+planning, the effectors and the dedicated robot controller activity
+(Figure~\ref{architecture_fg}).
+
+\subsubsection*{Human aware motion, placement and manipulation planning}
+In our scenarios, actions are only object manipulation actions: Pick,
+Place, Throw, Give. Motion and Task planning allows to compute final
+object placement, grasp and arm motion trajectories taking into
+account task specific constraints and human postures, abilities and
+preferences: see \cite{Sisbot2008, Mainprice2011, Pandey2010} for
+details.
+
+\subsubsection*{Action Execution and Monitoring Robot Controller}
+context and on the shared plan elaborated by HATP for a given goal,
+the robot controller decides to execute an action or to ask its human
+partner to do it.  Actions feasibility by the human or the robot are
+regularly reconsidered based on the reachability / visibility
+computation mechanisms.
+
+Robot action execution is based on simple automatons  that translate
+symbolic planning atomic actions into sequences of planned arm motions
+and gripper commands to execute according to current state
+of the 3-tuple (gripper, object, furniture). We have three states
+according to whether the object is in gripper and if it is in gripper
+whether it is on furniture.  These states are directly obtained from
+the updated symbolic state of the world in the ontology.
+
+For plan action monitoring, primitive actions recognition defined
+above are used. A primitive action detection is interpreted as action
+success if it is the expected one and failure otherwise. The robot
+also reacts to the absence of activity.
+
+\section{Human-aware task planning}
+
+Our execution controllers rely on symbolic task planning to convert faraway
+desires into a succession of atomic actions. We use in our architecture the
+HATP planner (for Human Aware Task Planner~\cite{Alili2009}).  HATP is based on
+a Hierarchical Task Network (HTN) refinement, which performs an iterative task
+decomposition into sub-tasks until reaching atomic actions.  The \emph{planning
+domain} defines the set of methods describing how to decompose a task and can
+be seen as the procedural knowledge of the robot.
 
@@
 
-\subsection{ Goal and plan Controller}
-Figure~\ref{goal_plans_fg} sums up the Goal and Plan Management as
-implemented in the robot controller.  When an event announcing a new
-goal is caught by the controller, the validity of this goal is tested:
-does it corresponds to capabilities of agents? is it not already
-achieved? Then, the goal is sent to HATP which produces a first plan.
-A goal is considered achievable as long as the planner computes a
-valid plan and it is not abandoned by the human.
+Two important remarks: because HATP is a generic symbolic task planner, we have
+been able to design a planning domain at a semantic level which is close to the
+one used in the human-robot dialogue (the planner vocabulary contains concepts
+like \texttt{give}, \texttt{table}, \texttt{is on}...). Hence only a few
+ontology rules have been required to map both the knowledge extracted from the
+situation assessment and the statements originated from the verbal interaction
+to the planner domain.
 
-Plan execution consists in the management of all the actions of the
-plan. Human and robot are not acting both at the same time.  In case
-of plan failure a new plan is requested and executed.
+Second remark, after some research (see appendix B of~\cite{Lemaignan2012a} for
+a detailed discussion)  we have decided not to represent neither the planning
+domain nor the resulting plans in the knowledge base: the planning domain (with
+tasks pre- and postconditions) is stored in a specific format, outside of the
+knowledge repository, and the plans are directly communicated to the robot
+controller. From our experience, the OWL formalism is not a satisfying
+representation language for concepts that are both tightly bound to time
+representation and non-monotonic.
 
-The management at the action level is done in three steps. First, the
-action preconditions are tested over the current state of the
-world. Then the action is executed and monitored (only monitored of it
-corresponds to a human action).  Finally, the expected effects are
-verified in order to acknowledge the action achievement.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Internal cognitive processes}
+\label{sect|intern}
 
-Concerning the speech acts during plan management, the robot informs
-its partners on the goal existence and status, plan existence and
-status, ongoing plan action, plan action failure, failing facts for
-action precondition or effect assessment.
+\subsection{Theory of Mind}
+\label{sect|tom}
 
-\begin{figure}[thpb]
-  \centering
-  \includegraphics[width=0.8\columnwidth]{./figs/plan_management.pdf}
-  \caption {Automaton for goal and plan management}
-  \label{goal_plans_fg}
-\end{figure}
+Theory of Mind (originally defined in~\cite{Premack1978}) is the cognitive
+ability that a subject possesses to represent the mental state of another
+agent, possibly including knowledge that contradicts the subject own model: for
+example, a book can be at the same time \emph{visible} by myself, and \emph{not
+visible} by you.
 
-\section{Action execution and monitoring}\label{sec:action}
+This skill, which is essential to understand others' perspectives during
+interactions, appears in children around their third year. It supposes the ability to build, store and retrieve separate models of the knowledge of the interactors.
 
-The action execution and monitoring task involves the motion and task
-planning, the effectors and the dedicated robot controller activity
-(Figure~\ref{architecture_fg}).
+Our knowledge base implements such a mechanism: when the robots infers a new
+agent has been introduced in the knowledge base, it initializes a new,
+independent, ontology for this agent. All the ontologies that are created share
+the same common-sense knowledge, but rely on each agent's perspective for the
+actual instantiation: the robot (geometrically) computes that the book is in
+its own field of view, but not in the human one. The robot knowledge contains
+the fact \stmt{book isVisible true} while the human model contains \stmt{book
+isVisible false}.
 
-%\subsubsection*{Human aware motion, placement and manipulation planning}
-\vspace{0.3cm}
-\noindent
-\textbf{Human-aware motion, placement and manipulation planning:}
-In our scenarios, actions are only object manipulation actions: Pick,
-Place, Throw, Give. Motion and Task planning allows to compute final
-object placement, grasp and arm motion trajectories taking into
-account task specific constraints and human postures, abilities and
-preferences: see \cite{Sisbot2008, Mainprice2011, Pandey2010} for
-details.
+One classical application of this cognitive skills is the so-called
+\emph{False-Belief} experiment (also known as the \emph{Sally and Ann}
+experiment)~\cite{Leslie2000}: a child is asked to watch a scene where two
+people, A and B, manipulate objects. Then A leaves and B hides away one
+object. When A comes back, we ask the child ``where do you think A will
+look for the object?''. Before acquiring a theory of mind, children are not
+able to separate their own (true) model of the world (where they know that
+the object was hidden) from the model of A, which contains \emph{false
+beliefs} on the world (A still thinks the object is at its original
+position since he did not see B hiding it). Using seperate knowledge models
+in the knowledge base, we have been able to replicate this experience on
+our robots~\cite{Warnier2012a}.
 
-%\subsubsection*{Action Execution and Monitoring Robot Controller}
-\vspace{0.3cm}
-\noindent
-\textbf{Execution and Monitoring Robot Controller:} Depending on the
-context and on the shared plan elaborated by HATP for a given goal,
-the robot controller decides to execute an action or to ask its human
-partner to do it.  Actions feasibility by the human or the robot are
-regularly reconsidered based on the reachability / visibility
-computation mechanisms.
+\subsection{Working memory}
 
-Robot action execution is based on simple automatons  that translate
-symbolic planning atomic actions into sequences of planned arm motions
-and gripper commands to execute according to current state
-of the 3-tuple (gripper, object, furniture). We have three states
-according to whether the object is in gripper and if it is in gripper
-whether it is on furniture.  These states are directly obtained from
-the updated symbolic state of the world in the ontology.
+The {\sc Oro} server also feature a mechanism to mimic simple forms of
+biological memory.  When new statements are inserted in the knowledge base, a
+\emph{memory profile} is optionally attached to them.
 
-For plan action monitoring, primitive actions recognition defined
-above are used. A primitive action detection is interpreted as action
-success if it is the expected one and failure otherwise. The robot
-also reacts to the absence of activity.
+Three such profiles are predefined: {\tt short term}, {\tt episodic} and {\tt
+long term}. They each correspond to a different lifetime for the statements
+(respectively 10 seconds, 5 minutes and no time limit). After this duration,
+the statements are automatically removed from the knowledge base.
 
-\section{An illustrative example}\label{sec:expes}
+This approach is limited. In particular, \emph{episodic} memory primarily
+refers to the semantics of the statements (that is expected to be related to an
+event) and not to a specific life duration.
+
+We rely however on this short term memory for a particular use-case:
+\emph{active concept}. Some modules, like the natural language processor, use
+the {\tt short term} memory profile to mark for a few seconds important
+concepts that are currently manipulated by the robot. For example, if a human
+asks the robot: ``Give me all red objects'', the human, the \concept{Give}
+action, and each red objects that are found are successively marked as
+\emph{active concepts} by inserting statements like \stmt{human type
+ActiveConcept} in the short-term memory (which can be considered, in this case,
+to be a working memory). We use this feature to give a (visual) feedback to the
+users.
+
+\section{Experimental results}
 
@@
 
-\section{Conclusion and future work}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Discussion and conclusion}
+\label{sect|conclusion}
+
+\subsection{Knowledge-oriented architecture}
+
+Altogether, the components we have presented compose an architecture that we
+call \emph{knowledge-oriented}:
+
+\begin{itemize}
+
+    \item{Knowledge is explicitly stored in one central and consistent
+    repository of facts, accessible by all modules.}
+
+    \item{Knowledge is represented in a strict formalism (OWL statements) and
+    with a clearly defined vocabulary (stated in the common-sense ontology).}
+
+    \item{The first two points enable both a loosely-coupled architecture where
+    modules can very easily be removed or replaced by other ones as long as
+    they share the same semantics (modules are defined by the knowledge they
+    produce),}
+
+    \item{and a \emph{symbolic}, reactive, event-driven approach to robot control.
+    By managing events at the same level as the reasoner, we take full
+    advantage of the inference abilities of {\sc Oro} to trigger events whose
+    \texttt{true} conditions can be inferred.}
+
+    \item{Finally, this architecture allows for the combination of very
+    different knowledge modalities in a single homogeneous environment,
+    bringing mutual benefits to components. For instance, the dialogue
+    processing module can perfectly run without any geometric
+    perception, but its disambiguation routines can transparently
+    benefit from it when available (since richer symbolic descriptions of
+    objects are then available).}
+
+\end{itemize}
+
+We believe that this \emph{knowledge-oriented} approach has a strong potential
+not only to enable rich human-robot interaction, but also as a broader approach
+to information alignment and fusion in complex robotic systems.  The
+versatility of this paradigm could be illustrated by a simple imaginary
+scenario with a blind robot and a deaf robot. The blind robot does not see (no
+cameras or alike), but someone can verbally describe a scene to it. On the
+other hand, the deaf robot has a good vision system, but cannot process verbal
+input.  Without any changes to the software architecture that we described,
+control modules of both robots could equally perform the same tasks since all
+the knowledge is abstracted and centralised (note that to actually implement
+this imaginary situation, the blind robot would of course need \textit{a
+priori} 3D models of objects talked about to enable planning or pick and place
+actions, and the deaf robot would require at least some gesture interpretation
+to understand orders).
+
+\subsection{Conclusion}
 
@@
 devoted to fetch-and-carry, as well as for interactive manipulation
-tasks and associated activities.\\
+tasks and associated activities.
 
-\vspace{0.3cm}
-\noindent
-{\bf Acknowledgements:}
-This work has been conducted within the EU CHRIS project ({\tt
-  \url{http://www.chrisfp7.eu/}}) funded by the E.C. Division FP7-IST
-under Contract 215805. We also would like to thank warmly a number of LAAS
-robotics team members and more particularly Xavier Broquere, Mokhtar
-Gharbi, Wuwei He, Matthieu Herrb, Jim Mainprice, Raquel Ros, Daniel
-Sidobre, Thierry Sim\'{e}on, Muhammad Ali.
+\section*{Acknowledgment}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+This work has been supported by EU FP7 ``SAPHARI'' under grant agreement no. ICT-287513.
+
 \bibliographystyle{elsarticle-num}
@@
 \end{document}
-
-\section{Discussion and future work}
-
-The design choices and the results presented here is still preliminary. 
-While the general scheme we propose might be difficult to implement in
-a general sense, we believe that it is a reasonable challenge to
-implement it in the case of a personal robot assistant essentially
-devoted to fetch-and-carry, as well interactive manipulation tasks and
-associated activities. The robot would operates in a known in-door
-environment (acquired in a preliminary phase).
-
-In such contexts, it is possible ....
-
-Fetch-and-carry and object manipulation task need 3D geometric
-planning.  One challenging problem would be to extend the approach
-discussed above to the situation where a robot has to hand
-an object to human. Indeed, there is a need to take into account
-visibility and reach, in terms of kinematic constraints, of the human
-partner (Figure~\ref{fig_handing}).
-
-%\input{conclusion} %% Conclusion and Future Work
-\section{Conclusion}\label{sec:Conclusion}
-
-In this paper we have presented a decisional framework designed for
-robots operating in a human environment. Our objective is to provide a
-management of human interaction that is an integral part of a general
-robot control architecture.  This was done in order to provide a
-principled way to deal with HRI.
-
-The framework is also suitable for the development and experiment of
-task planners and interaction schemes that explicitly consider
-human abilities and preferences.
-
-Examples of such planners are presented that integrate various models
-of human abilities and preferences.
-
-\end{document}

### 2013-06-07 paper.tex
Renamed sections in original paper
Reorganized chapters according to meeting with Rachid

@@
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%2345678901234567890123456789012345678901234567890123456789012345678901234567890
+%        1         2         3         4         5         6         7         8
+
 \documentclass{elsarticle}
 
-\usepackage{graphicx, subfigure}        % standard LaTeX graphics tool
-                             % when including figure files
+\usepackage{cite}
+\usepackage{graphicx, subfigure}
+%\usepackage{hyperref}
 \usepackage{url}
-\usepackage{fixme}
-\usepackage{lscape}
+\graphicspath{{figs/}}
+\DeclareGraphicsExtensions{.pdf,.jpg,.png}
 
-\journal{Journal of Artificial Intelligence}
+\usepackage{ifthen}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\usepackage[utf8x]{inputenc}
+\usepackage[T1]{fontenc}
 
-%% Custom macros
-\newcommand{\stmt}[1]{{\footnotesize $\langle$\stmttt#1\relax$\rangle$}}
-\newcommand{\rawstmt}[1]{{\footnotesize \stmttt#1\relax}}
-\def\stmttt#1 #2 #3\relax{{\tt#1} {\bf{\tt #2}} {\tt #3}}
+\usepackage[draft]{fixme}
 
-\newcommand{\setstmt}[1]{{\footnotesize [\setstmttt#1\relax]}}
-\def\setstmttt#1,#2\relax{\rawstmt{#1}, \rawstmt{#2}}
+\usepackage{listings}
+\usepackage{alltt}
+\usepackage{fancyvrb}
+\usepackage{flushend} % equalize the column length on last page
 
-\newcommand{\concept}[1]{{\footnotesize \texttt{#1}}}
+\newcommand{\concept}[1]{{\small \texttt{#1}}}
 
-\newcommand{\ie}{{\textit{i.e.~}}}
-\newcommand{\cf}{{\textit{cf~}}}
-\newcommand{\eg}{{\textit{e.g.~}}}
+\newcommand{\stmt}[1]{{\footnotesize \tt $\langle$ #1\relax$\rangle$}}
+%\newcommand{\stmt}[1]{{\footnotesize $\langle\stmttt#1\relax\rangle$}}
+%\newcommand{\rawstmt}[1]{{\footnotesize \stmttt#1\relax}}
+%\def\stmttt#1 #2 #3\relax{
+%\texttt{\IfBeginWith{#1}{?}{\textbf{#1}}{#1} \IfBeginWith{#2}{?}{\textbf{#2}}{\emph{#2}} \ifthenelse {\equal{#3}{true} \OR \equal{#3}{false}}{\emph{#3}}{\IfBeginWith{#3}{?}{\textbf{#3}} {#3}}}}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%\newcommand{\setstmt}[1]{{\footnotesize [\setstmttt#1\relax]}}
+%\def\setstmttt#1,#2\relax{\stmttt#1\relax, \stmttt#2\relax}
+
+\newcommand{\ie}{{\textit{i.e.\ }}}
+\newcommand{\cf}{{\textit{cf\ }}}
+\newcommand{\eg}{{\textit{e.g.\ }}}
+
+\begin{frontmatter}
+
+\title{\LARGE \bf
+Human-Robot Interaction: A Deliberative Architecture
+}
+
+\author{Rachid Alami and Séverin Lemaignan}
+
+\address{
+CNRS, LAAS, 7 avenue du Colonel Roche, F-31400 Toulouse, France\\
+Univ de Toulouse, LAAS, F-31400 Toulouse, France\\
+{\tt severin.lemaignan@laas.fr}, {\tt rachid.alami@laas.fr}
+}
 
 \begin{document}
-\begin{frontmatter}
 
-\title{When the robot considers the human... }
-% Use \titlerunning{Short Title} for an abbreviated version of
-% your contribution title if the original one is too long
-\author{Rachid Alami, Mathieu Warnier, Julien Guitton, S\'{e}verin Lemaignan and Emrah Akin Sisbot}
-% Use \authorrunning{Short Title} for an abbreviated version of
-% your contribution title if the original one is too long
-\address{ LAAS-CNRS\\ 
-        Universit\'e de Toulouse, UPS, INSA, INP, ISAE, LAAS\\ 
-         F-31077 Toulouse, France\\ \texttt{Firstname.Name@laas.fr}}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{abstract}
 
-\begin{abstract}
-  This paper addresses some key decisional issues that are
-  necessary for a cognitive robot which shares space and tasks with a
-  human. We adopt a constructive approach based on the identification
-  and the effective implementation of individual and collaborative
-  skills. The system is comprehensive since it aims at dealing with a
-  complete set of abilities articulated so that the robot controller
-  is effectively able to conduct in a flexible manner a collaborative
-  task with a human partner. These abilities include geometric
-  reasoning and situation assessment based essentially on
-  perspective-taking and affordances, management and exploitation of
-  each agent (human and robot) knowledge in a separate cognitive
-  model, human-aware task planning and human and robot interleaved
-  plan achievement.
+This paper addresses some key decisional issues that are necessary for a
+cognitive robot which shares space and tasks with a human. We adopt a
+constructive approach based on the identification and the effective
+implementation of individual and collaborative skills. The system is
+comprehensive since it aims at dealing with a complete set of abilities
+articulated so that the robot controller is effectively able to conduct in a
+flexible manner a collaborative task with a human partner. These abilities
+include geometric reasoning and situation assessment based essentially on
+perspective-taking and affordances, management and exploitation of each agent
+(human and robot) knowledge in a separate cognitive model, human-aware task
+planning and human and robot interleaved plan achievement.
+
+This article presents our design choices, the articulations between the diverse
+deliberative components of the robot, and the strengths and weaknesses of our
+approach. We show that explicit knowledge management is not only a convenient
+tool from the software engineering point of view, but also pushes for a
+different, more \emph{semantic} way to address the decision-making issue in
+human-robot interactions.
+
 \end{abstract}
@@
 
-\section{Related work and vision}\label{sec:soa}
-
-The human presence brings new requirements for robot's abilities both
-at the functional and at the deliberative levels~\cite{Klein2004}. The
-topics involve motion~\cite{Kulic2007,Berg2004,Madhav2006},
-navigation~\cite{Althaus2004,Sisbot2007}, manipulation~\cite{Kemp2007}
-in presence of humans as well as perception of human
-activities~\cite{Breazeal2001,Burger2008}. Also, when
-interacting with humans, robots need to incorporate communication and
-collaboration abilities. Several theories dealing with
-collaboration~\cite{Cohen1991,Grosz1996,Clark1996} emphasize that
-collaborative tasks have specific requirements compared to individual
-ones, \eg, since the robot and the person share a common goal, they
-have to agree on the manner to realize it, they must show their
-commitment to the goal during execution, etc. Several robotic systems
-have already been built based on these
-theories~\cite{Rich1997,Sidner2005,Tambe2005a,Breazeal2003} and they
-all have shown benefits of this approach. They have also shown how
-difficult it is to manage turn-taking between communication partners
-and to interleave task realization and communication in a generic
-way. Finally, today only few
-systems~\cite{Fong_2006,Breazeal2003,Sisbot2008} take humans into
-account at all levels.
-
-Perspective Taking is a human ability which allows one to put
-him/herself in another person's point of view. Studied in
-psychology literature~\cite{Flavell1992,Tversky1999}, this ability is
-crucial when interacting with people by allowing one to reason on
-others' understanding of the world in terms of visual perception, spatial
-descriptions, affordances and beliefs, etc.
-Therefore, in the last years these notions have been gradually
-employed in Human-Robot Interaction.~\cite{breazeal2006} presents a
-learning algorithm that takes into account information about a
-teacher's visual perspective in order to learn a
-task. ~\cite{Johnson2005} apply visual perspective taking for action
-recognition between two robots.~\cite{Trafton2005} use both visual and
-spatial perspective taking for finding out the referent indicated by a
-human partner.
-
-Spatial reasoning~\cite{O'Keefe1999}, on the other hand, has been used
-for natural language processing for applications such as direction
-recognition ~\cite{Kollar2010,Matuszek2010} or language
-grounding~\cite{Tellex2010}.~\cite{Skubic2004} presented a spatial
-reasoner integrated in a robot which computes symbolic positions of
-objects.
-
 \begin{figure}[htb]
@@
 
-\section{The LAAS deliberative layer}
-\label{sec:Framework}
+\subsection{Our Contributions Towards the Cognitive Robot}
+
+Natural interaction and cooperation are on the (dare we say,
+\emph{short-term}) agenda for the human-robot interaction community. They are
+keys to the broad class of \emph{interactive manipulation problems}: several
+agents agree on a (more or less implicit) joint goal that requires some sort of
+cooperation to be successfully achieved. This class of problems involves both
+dialogue and manipulation and they are often initially underspecified: they
+require iterative and interactive resolution.
+
+Over the last years we have focused our efforts on identifying the cognitive
+prerequisites of these challenges, and giving them experimental reality on the
+robots: what is required for sentences like ``Let's dress the table together''
+to be understood by the robot, correctly interpreted in the spatial and
+temporal context of the interaction, and eventually transformed into a set of
+actions.
+
+We have chosen to tackle the challenge from several ends: human-aware
+navigation and motion planning (\cite{Mainprice2011}), situation assessment
+coupled with motion planning (\cite{Mainprice2012}), projection of
+``mightabilities'' that anticipates what surrounding agents may do
+(\cite{Pandey2011}), and the design and deployment of a pervasive
+\emph{knowledge-centric} software architecture.
+
+The last item is the focus of this paper: what does ``knowledge-centric''
+architecture means, and what are its benefits and drawbacks.
+
+\subsection{Article overview}
+
+In the next section, we first introduce the deliberative layer of our robots at
+the functional level. It represents what we call a \emph{knowledge-oriented}
+architecture, built around a central knowledge management tool called the {\sc
+Oro} server.
+
+The following sections go over each of the facets of this deliberative layer:
+how knowledge is produced asynchronously from geometric reasoning, how the
+robot conducts grounded multi-modal interaction with humans, how knowledge is
+used by decision-making components like the robot controller or the task
+planner, and finally, how our knowledge-centric architecture enables the
+implementation of specific internal cognitive processes.
+
+We conclude the article with a discussion about the strengths and weaknesses of
+such a knowledge-oriented architecture.
+
+\section{From Human-Robot Interaction to a Deliberative Architecture}
+
+\subsection{The Psycho-Cognitive Grounds of Human-Robot Interaction}
+
+....summary of the challenges that human-robot interaction presents to robotics.
+
+\subsection{The LAAS Deliberative Layer}
 
@@
 
-\subsection*{Knowledge model}
+\subsection{A Knowledge model}
 
@@
 
-\subsection*{Interaction model}
+\subsection{Internal cognitive processes}
+\label{sect|intern}
 
-Interaction happens as a consequence of an explicit request of the
-human to satisfy a goal or because the robot finds itself in a
-situation where it is useful if not mandatory. In both cases, the
-robot has a goal to satisfy.  An important issue is the notion of
-engagement, a process in which the robot will have to establish,
-maintain and terminate a connection with a human partner. 
-This covers goal establishment, selection of an incremental refinement
-of the task that is intended to be achieved, and execution control
-including monitoring, and even influencing, human task performance and
-his/her commitment to the goal. The human involvement may range from a
-direct participation to the task achievement, to a simple
-``acceptance'' of robot activity in his/her close vicinity.
+\subsubsection{Theory of Mind}
+\label{sect|tom}
 
-Next sections describe how the robot is controlled through an analysis of the
-three main activities performed by the robot controller:
+Theory of Mind (originally defined in~\cite{Premack1978}) is the cognitive
+ability that a subject possesses to represent the mental state of another
+agent, possibly including knowledge that contradicts the subject own model: for
+example, a book can be at the same time \emph{visible} by myself, and \emph{not
+visible} by you.
 
-\begin {enumerate}
-\item Situation assessment and context management 
-\item Goals and plans management
-\item Action refinement, execution and monitoring
-\end {enumerate}
+This skill, which is essential to understand others' perspectives during
+interactions, appears in children around their third year. It supposes the ability to build, store and retrieve separate models of the knowledge of the interactors.
 
-These activities are caried out by different software composants:
+Our knowledge base implements such a mechanism: when the robots infers a new
+agent has been introduced in the knowledge base, it initializes a new,
+independent, ontology for this agent. All the ontologies that are created share
+the same common-sense knowledge, but rely on each agent's perspective for the
+actual instantiation: the robot (geometrically) computes that the book is in
+its own field of view, but not in the human one. The robot knowledge contains
+the fact \stmt{book isVisible true} while the human model contains \stmt{book
+isVisible false}.
 
-\begin{itemize}
+One classical application of this cognitive skills is the so-called
+\emph{False-Belief} experiment (also known as the \emph{Sally and Ann}
+experiment)~\cite{Leslie2000}: a child is asked to watch a scene where two
+people, A and B, manipulate objects. Then A leaves and B hides away one
+object. When A comes back, we ask the child ``where do you think A will
+look for the object?''. Before acquiring a theory of mind, children are not
+able to separate their own (true) model of the world (where they know that
+the object was hidden) from the model of A, which contains \emph{false
+beliefs} on the world (A still thinks the object is at its original
+position since he did not see B hiding it). Using seperate knowledge models
+in the knowledge base, we have been able to replicate this experience on
+our robots~\cite{Warnier2012a}.
 
-    \item Situation assessment is managed by \textsc{Spark} (Spatial Reasoning
-        and Knowledge module)~\cite{Sisbot2011},
+\subsubsection{Working memory}
 
-    \item HATP (Human-Aware Task Planner)~\cite{Alili2008} is responsible for
-        symbolic task planning,
+The {\sc Oro} server also feature a mechanism to mimic simple forms of
+biological memory.  When new statements are inserted in the knowledge base, a
+\emph{memory profile} is optionally attached to them.
 
-    \item We finally also rely on a set of human-aware motion, placement and
-        manipulation planners, grouped under the \textsc{Move3D}
-        umbrella~\cite{Sisbot2008, Mainprice2011, Pandey2010}.
+Three such profiles are predefined: {\tt short term}, {\tt episodic} and {\tt
+long term}. They each correspond to a different lifetime for the statements
+(respectively 10 seconds, 5 minutes and no time limit). After this duration,
+the statements are automatically removed from the knowledge base.
 
-\end{itemize}
+This approach is limited. In particular, \emph{episodic} memory primarily
+refers to the semantics of the statements (that is expected to be related to an
+event) and not to a specific life duration.
 
-Other decisional activities, such as situated dialog (\cite{Ros2010b,
-  Lemaignan2011}, not presented here) have been developed that use the
-same set of components.
+We rely however on this short term memory for a particular use-case:
+\emph{active concept}. Some modules, like the natural language processor, use
+the {\tt short term} memory profile to mark for a few seconds important
+concepts that are currently manipulated by the robot. For example, if a human
+asks the robot: ``Give me all red objects'', the human, the \concept{Give}
+action, and each red objects that are found are successively marked as
+\emph{active concepts} by inserting statements like \stmt{human type
+ActiveConcept} in the short-term memory (which can be considered, in this case,
+to be a working memory). We use this feature to give a (visual) feedback to the
+users.
 
-\section{Situation assessment and context management}\label{sec:situ}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Implementing the Cognitive Skills}
 
-This activity involves the geometric and temporal reasoning component,
-the symbolic facts and belief management component and the dedicated
-robot controller activity (Figure~\ref{architecture_fg}).
+Previous section provided an overview of what we propose as a generic cognitive
+architecture for robots involved in interactions with other social agents.
 
-We assume that perception provides in real-time the identity and the
-position of objects when they are in the field of view of the sensors.
-In our implemented examples, the robot is localised using a standard
-horizontal laser-scanning based localisation system, the objects are
-identified and localized using ARToolkit \cite{ARToolkit} and the
-humans are tracked using a commercial motion capture system and a
-Kinect device from Microsoft.
+We now introduce each of the components and try to give insights on their main
+features.
 
-\subsection{Geometric and Temporal Reasoning component}\label{sub:gtrc}
+\subsection{Situation Assessment, Human Awareness, Perspective Taking}
+\label{sect|sit-ass}
 
-The geometric reasoning component plays a central role in our
-architecture. It is called SPARK (Spatial Reasoning and
-Knowledge~\cite{Sisbot2011}) in the current implementation. It is
-responsible for geometric information gathering and it embeds a number
-of decisional activities linked to abstraction (symbolic facts
-production) and inference based on geometric and temporal reasoning.
-SPARK maintains all geometric positions and configurations of agents,
-objects and furniture coming from perception and previous or {\it a
-  priori} knowledge.
+\begin{figure}
+        \centering
+        \includegraphics[width=\columnwidth]{spark-oro}
+    \caption{Functional overview of knowledge base ({\sc Oro} server, top part) and the geometric situation assessment module (\emph{SPARK}, bottom part)}
+        \label{fig|spark-oro}
+\end{figure}
 
- %\subsubsection*{Symbolic facts production:} 
-\vspace{0.3cm}
-\noindent
-\textbf{Symbolic facts production:} 
+Anchoring perceptions in a symbolic model requires perception abilities and
+their symbolic interpretation. We rely on a dedicated geometric and temporal
+reasoning module (as named on Fig.~\ref{fig|archi}) called SPARK (\emph{SPAtial
+Reasoning \& Knowledge}~\cite{Sisbot2011}). It is a situation assessment
+reasoner that generates symbolic knowledge from the geometry of the
+environment with respect to relations between objects, robots and humans,
+also taking into account the different perspective that each agent has on
+the environment (Fig.~\ref{fig|spark-oro}).
+
+SPARK is an \emph{amodal} geometric model of the environment that serves both
+as basis for the fusion of the perception modalities and as bridge with the
+symbolic layer. This geometric model is continuously updated at run-time by the
+robot based on its sensors. In our experiments, objects are identified and
+localised through 2D barcodes.  On the other hand, the human pose is tracked
+with a Kinect-like device (assisted by motion capture to accurately track the
+head motion, which is required to compute what the human is looking at).
+
+\subsubsection{Situation Assessment}
+
+\paragraph{Symbolic facts production:}
+
 Geometric state of the world is abstracted in symbolic facts that can
@@
 
-%\newline\newline
-%\subsubsection*{Hypotheses on objects states and positions}
-\vspace{0.3cm}
-\noindent
-\textbf{Hypotheses on objects states and positions:}
+\paragraph{Hypotheses on objects states and positions}
 It is sometimes difficult or even impossible to see and/or track an
@@
 
-%\subsubsection*{Primitive action recognition}
-\vspace{0.3cm}
-\noindent {\bf Primitive action recognition:} 
+\paragraph{Primitive action recognition}
 Monitoring human activity is crucial to maintain a coherent state of
@@
 
-\subsection{Symbolic facts and beliefs management}
+\subsubsection{Ego-centric and allo-centric frames}
+
+SPARK enables \emph{perspective-taking}: spatial relations between entities can
+be computed from different viewpoints, which let the robot build for each
+agents it interacts with a different, \emph{perspective-aware} symbolic model
+of the environment. These models are separately stored in the knowledge base.
+
+This allows us to deal with ambiguities that arises when one speaker refers to
+an object within a reference system (or changes the reference system, \ie
+switches perspective) without making the reference frame
+explicit~\cite{Breazeal2006, Ros2010}.
+
+As a result, the robot stores models on the environment either in the
+\emph{ego-centric} reference frame (from the robot perspective) or in
+addressee-centred (\ie, \emph{allo-centric}) frames.
+
+\subsubsection{Symbolic facts and beliefs management}
 
@@
 
-\subsection{Situation Assessment and Context Management Controller}
+\subsubsection{Context Management Policies}
 
@@
 
-\section{Goal and Plan Management}\label{sec:plan}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Multi-modal Communication}
+\label{sect|com}
 
-The Goal and Plan Management activity involves the human-aware
-symbolic task planner component and the dedicated robot controller
-activity (Figure~\ref{architecture_fg}).
+\subsubsection{Natural language grounding}
 
-\subsection{Symbolic Task Planning}
+Natural language processing is one of the fields of human-robot interaction
+where the introduction of the semantic layer has been most beneficial.
+We detail in~\cite{Lemaignan2011a} the techniques and the tool called
+\texttt{dialogs} that we have developed for English language parsing and
+grounding, along with verbalisation and (minimalist) dialogue management.
+
+Natural language input (in experiments, we rely on an Android-based interface,
+with Google speech recognition) is parsed into a grammatical structure, and
+each sentence atoms are resolved with the help of the ontology to ground
+concepts like objects (\ie when a user says ``pick the can'', resolve to which
+instance of \emph{can} the user is referring to) and actions. Sentences are
+sorted into questions, desires and statements, and processed accordingly.
+
+The system supports quantification (``give me \{a | the | some | all | any | ...\} can''),
+thematic roles (action-specific predicates that qualify the actions), interactive disambiguation (the robot asks
+questions when it needs more information), anaphora resolution (``give \emph{it} to me'') based on
+dialogue history. It also permits knowledge extension by learning new semantic
+structures (for instance, a sentence like ``learn that cats are animals'' is
+converted into \stmt{Cat subClassOf Animal}, interprets common temporal and place adverbs (like \emph{above} or \emph{tomorrow}) and translates to a certain extend \emph{states} (``I'm tired'') into
+\emph{experiences} (\stmt{HUMAN experiences state\_1, state\_1 hasFeature
+tired}).
+
+\subsubsection{Limits of disambiguation at semantic level}
+
+One prototypical example of semantic disambiguation has been given in
+\cite{Ros2010b} with the child game \emph{spygame}: two players are facing
+each other with a set of random objects in-between, one player mentally choose
+one object, and the other player has to guess the object by asking closed
+questions like \emph{Is your object small or large?} Based on the knowledge it
+has acquired, the robot is able to minimize the number of questions required to
+find the object.
+
+When playing this kind of game, however, the issue arises that the robot has no
+way to select which knowledge about the object is relevant in the interaction
+context. For instance, the knowledge base may store facts like \stmt{obj1 type
+ActiveConcept} (which internally means that this concept was mentioned in a
+discussion in the last few seconds): this information is not a relevant
+property of \concept{obj1} when trying to disambiguate concepts with humans.
+This distinction between \emph{internal knowledge} (meaningful to
+the system only) and \emph{common knowledge} (whose meaning is understood by
+all the interactors) has not been properly dealt with in our architecture.
+
+Besides, even knowledge that belongs the \emph{common knowledge} may not be
+appropriate in a given interaction context. For instance, the system may
+compute that at a given instant the human is looking at the object: \stmt{human
+looksAt obj1}. This property makes sense to both parties, but in the context of
+the \emph{spygame}, we would like to mainly use immanent properties, not
+volatile like a gaze. More research is required to identify relevant
+interaction contexts and knowledge classes attached to them.
+
+\subsubsection{Multi-modal communication}
+
+Because all components rely on the same RDF formalism to format their outputs,
+the different communication modalities (\emph{explicit} like verbal, deictic or
+based on gaze, or \emph{implicit} like postures) are presented in an
+homogeneous way.
+
+The dialogue grounding process makes use of them at two distinct levels.
+
+First, particular steps of the grounding process explicitly check for the
+presence and value of specific facts: for instance, when a set of instances
+match a category (the human says ``give me the bottle'' and the robot knows
+about three bottles), the module may decide (it actually depends on the
+quantifier preceding the class) to discard some of them based on their
+\emph{visibility} by the speaker (implicit communication context built on the
+human posture).
+
+Another example, when the human says ``this'', the robot checks if the human is
+currently pointing at some object. In that case, \emph{this} is replaced by the
+object focused on.
+
+Note that, while the system benefits from the complementary modalities, they
+are not all required. The dialogue system would for instance run equally well
+with only the verbal modality.
+
+The second level of integration of multi-modality is implicit: by computing
+symbolic properties from the geometry, richer descriptions and hence
+discrimination possibilities are available: for instance, if reachability is
+available, the robot may asks ``do you mean the bottle that is accessible to
+me?'' to discriminate between the three bottles. That way, procedures relying
+on discrimination transparently benefit from added modalities.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{Human-aware task planning}
+
+Our execution controllers rely on symbolic task planning to convert faraway
+desires into a succession of atomic actions. We use in our architecture the
+HATP planner (for Human Aware Task Planner~\cite{Alili2009}).  HATP is based on
+a Hierarchical Task Network (HTN) refinement, which performs an iterative task
+decomposition into sub-tasks until reaching atomic actions.  The \emph{planning
+domain} defines the set of methods describing how to decompose a task and can
+be seen as the procedural knowledge of the robot.
 
@@
 
-\subsection{ Goal and plan Controller}
-Figure~\ref{goal_plans_fg} sums up the Goal and Plan Management as
-implemented in the robot controller.  When an event announcing a new
-goal is caught by the controller, the validity of this goal is tested:
-does it corresponds to capabilities of agents? is it not already
-achieved? Then, the goal is sent to HATP which produces a first plan.
-A goal is considered achievable as long as the planner computes a
-valid plan and it is not abandoned by the human.
+Two important remarks: because HATP is a generic symbolic task planner, we have
+been able to design a planning domain at a semantic level which is close to the
+one used in the human-robot dialogue (the planner vocabulary contains concepts
+like \texttt{give}, \texttt{table}, \texttt{is on}...). Hence only a few
+ontology rules have been required to map both the knowledge extracted from the
+situation assessment and the statements originated from the verbal interaction
+to the planner domain.
 
-Plan execution consists in the management of all the actions of the
-plan. Human and robot are not acting both at the same time.  In case
-of plan failure a new plan is requested and executed.
+Second remark, after some research (see appendix B of~\cite{Lemaignan2012a} for
+a detailed discussion)  we have decided not to represent neither the planning
+domain nor the resulting plans in the knowledge base: the planning domain (with
+tasks pre- and postconditions) is stored in a specific format, outside of the
+knowledge repository, and the plans are directly communicated to the robot
+controller. From our experience, the OWL formalism is not a satisfying
+representation language for concepts that are both tightly bound to time
+representation and non-monotonic.
 
-The management at the action level is done in three steps. First, the
-action preconditions are tested over the current state of the
-world. Then the action is executed and monitored (only monitored of it
-corresponds to a human action).  Finally, the expected effects are
-verified in order to acknowledge the action achievement.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Robot control}
+\label{sect|ctrl}
 
-Concerning the speech acts during plan management, the robot informs
-its partners on the goal existence and status, plan existence and
-status, ongoing plan action, plan action failure, failing facts for
-action precondition or effect assessment.
+\subsubsection{Desires and experiences}
 
-\begin{figure}[thpb]
-  \centering
-  \includegraphics[width=0.8\columnwidth]{./figs/plan_management.pdf}
-  \caption {Automaton for goal and plan management}
-  \label{goal_plans_fg}
-\end{figure}
+Our robot execution controllers (OpenPRS-based {\sc Shary} or Python-based {\sc
+pyRobots}) have deep integration with the knowledge base. It serves as the primary source of information for semantic-aware decision-making.
 
-\section{Action execution and monitoring}\label{sec:action}
+We split the interaction situations stemming from the situation assessment and
+communication components in two categories: \emph{desires} (performative act)
+and \emph{experiences} (assertive act).
+
+\emph{Desires} are typically human orders (``Give me that book''). The nature
+of the desired action (to pick, give, look, bring, show...), along with the
+action parametrization (what is acted on? who should perform the action? etc.)
+are extracted from the knowledge base, and either passed to a task planner
+(presented in the next section) or executed if the procedure is directly
+available.
+
+\emph{Experiences}, on the other hand, comprise of emotions, states and
+questions (when asking a question, we consider the human to be in an
+\emph{interrogative state}). When the knowledge base recognizes that an agent
+\emph{experiences} a particular emotion or state, the execution controller may
+decide to handle it, typically by trying to answer the question or using the
+emotional or physical state as a parameter for subsequent actions. As an example,
+when the speaker says ``I feel tired'', we change the motion planner
+parametrization to lower the effort the human need to provide for the following
+joint manipulation tasks. Note that this example has been implemented as a
+proof-of-concept. We have not yet tried to define a theoretical framework that
+would support action alteration based on the user's experienced states.
+
+\subsubsection{Event-driven control}
+
+The {\sc Oro} server proposes two paradigms to access its content: RPC-style
+queries (based on the standard SPARQL language) or events. In its simplest
+form, a module can register an event by passing through an event pattern, which
+is basically a partial statement like \stmt{? type Elephant}, and a callback.
+Every time a new instance of elephant appears in the knowledge base, the
+callback is triggered.
+
+This allows us to write reactive robot controllers with a high level of
+expressiveness: for instance, by subscribing to the event \stmt{human1 desires
+?action, ?action type Look, ?action hasGoal myself}, we could trigger a
+behaviour when the human expresses (through dialogue, gestures...) that he
+wants to look at the robot itself.
+
+The robot controller designer does not need to directly care about how this
+\emph{desire} is produced (this is delegated to perception modules), it can
+focuses on the semantic of the desire.
+
+Note also that we take advantage of the reasoning capabilities of the system:
+for example, the goal of the action (\stmt{action hasGoal myself}) may well not
+have been explicitly asserted, but inferred by the reasoner based on other
+assertions.
+
+\subsubsection{Action execution and monitoring}\label{sec:action}
 
@@
 
-%\subsubsection*{Human aware motion, placement and manipulation planning}
-\vspace{0.3cm}
-\noindent
-\textbf{Human-aware motion, placement and manipulation planning:}
+\paragraph{Human aware motion, placement and manipulation planning}
 In our scenarios, actions are only object manipulation actions: Pick,
@@
 
-%\subsubsection*{Action Execution and Monitoring Robot Controller}
-\vspace{0.3cm}
-\noindent
-\textbf{Execution and Monitoring Robot Controller:} Depending on the
+\paragraph{Action Execution and Monitoring Robot Controller}
 context and on the shared plan elaborated by HATP for a given goal,
@@
 
-\section{An illustrative example}\label{sec:expes}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Experimental results}
 
@@
 
-\section{Conclusion and future work}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Perspectives}
+\label{sect|conclusion}
+
+\subsection{Contributions and Related work}\label{sec:soa}
+
+The human presence brings new requirements for robot's abilities both
+at the functional and at the deliberative levels~\cite{Klein2004}. The
+topics involve motion~\cite{Kulic2007,Berg2004,Madhav2006},
+navigation~\cite{Althaus2004,Sisbot2007}, manipulation~\cite{Kemp2007}
+in presence of humans as well as perception of human
+activities~\cite{Breazeal2001,Burger2008}. Also, when
+interacting with humans, robots need to incorporate communication and
+collaboration abilities. Several theories dealing with
+collaboration~\cite{Cohen1991,Grosz1996,Clark1996} emphasize that
+collaborative tasks have specific requirements compared to individual
+ones, \eg, since the robot and the person share a common goal, they
+have to agree on the manner to realize it, they must show their
+commitment to the goal during execution, etc. Several robotic systems
+have already been built based on these
+theories~\cite{Rich1997,Sidner2005,Tambe2005a,Breazeal2003} and they
+all have shown benefits of this approach. They have also shown how
+difficult it is to manage turn-taking between communication partners
+and to interleave task realization and communication in a generic
+way. Finally, today only few
+systems~\cite{Fong_2006,Breazeal2003,Sisbot2008} take humans into
+account at all levels.
+
+Perspective Taking is a human ability which allows one to put
+him/herself in another person's point of view. Studied in
+psychology literature~\cite{Flavell1992,Tversky1999}, this ability is
+crucial when interacting with people by allowing one to reason on
+others' understanding of the world in terms of visual perception, spatial
+descriptions, affordances and beliefs, etc.
+Therefore, in the last years these notions have been gradually
+employed in Human-Robot Interaction.~\cite{Breazeal2006} presents a
+learning algorithm that takes into account information about a
+teacher's visual perspective in order to learn a
+task. ~\cite{Johnson2005} apply visual perspective taking for action
+recognition between two robots.~\cite{Trafton2005} use both visual and
+spatial perspective taking for finding out the referent indicated by a
+human partner.
+
+Spatial reasoning~\cite{O'Keefe1999}, on the other hand, has been used
+for natural language processing for applications such as direction
+recognition ~\cite{Kollar2010,Matuszek2010} or language
+grounding~\cite{Tellex2010}.~\cite{Skubic2004} presented a spatial
+reasoner integrated in a robot which computes symbolic positions of
+objects.
+
+\subsection{Knowledge-oriented architecture}
+
+Altogether, the components we have presented compose an architecture that we
+call \emph{knowledge-oriented}:
+
+\begin{itemize}
+
+    \item{Knowledge is explicitly stored in one central and consistent
+    repository of facts, accessible by all modules.}
+
+    \item{Knowledge is represented in a strict formalism (OWL statements) and
+    with a clearly defined vocabulary (stated in the common-sense ontology).}
+
+    \item{The first two points enable both a loosely-coupled architecture where
+    modules can very easily be removed or replaced by other ones as long as
+    they share the same semantics (modules are defined by the knowledge they
+    produce),}
+
+    \item{and a \emph{symbolic}, reactive, event-driven approach to robot control.
+    By managing events at the same level as the reasoner, we take full
+    advantage of the inference abilities of {\sc Oro} to trigger events whose
+    \texttt{true} conditions can be inferred.}
+
+    \item{Finally, this architecture allows for the combination of very
+    different knowledge modalities in a single homogeneous environment,
+    bringing mutual benefits to components. For instance, the dialogue
+    processing module can perfectly run without any geometric
+    perception, but its disambiguation routines can transparently
+    benefit from it when available (since richer symbolic descriptions of
+    objects are then available).}
+
+\end{itemize}
+
+We believe that this \emph{knowledge-oriented} approach has a strong potential
+not only to enable rich human-robot interaction, but also as a broader approach
+to information alignment and fusion in complex robotic systems.  The
+versatility of this paradigm could be illustrated by a simple imaginary
+scenario with a blind robot and a deaf robot. The blind robot does not see (no
+cameras or alike), but someone can verbally describe a scene to it. On the
+other hand, the deaf robot has a good vision system, but cannot process verbal
+input.  Without any changes to the software architecture that we described,
+control modules of both robots could equally perform the same tasks since all
+the knowledge is abstracted and centralised (note that to actually implement
+this imaginary situation, the blind robot would of course need \textit{a
+priori} 3D models of objects talked about to enable planning or pick and place
+actions, and the deaf robot would require at least some gesture interpretation
+to understand orders).
+
+\subsection{The Next Steps}
 
@@
 devoted to fetch-and-carry, as well as for interactive manipulation
-tasks and associated activities.\\
+tasks and associated activities.
 
-\vspace{0.3cm}
-\noindent
-{\bf Acknowledgements:}
-This work has been conducted within the EU CHRIS project ({\tt
-  \url{http://www.chrisfp7.eu/}}) funded by the E.C. Division FP7-IST
-under Contract 215805. We also would like to thank warmly a number of LAAS
-robotics team members and more particularly Xavier Broquere, Mokhtar
-Gharbi, Wuwei He, Matthieu Herrb, Jim Mainprice, Raquel Ros, Daniel
-Sidobre, Thierry Sim\'{e}on, Muhammad Ali.
+\section*{Acknowledgment}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+This work has been supported by EU FP7 ``SAPHARI'' under grant agreement no. ICT-287513.
+
 \bibliographystyle{elsarticle-num}
+%\bibliographystyle{abbrv}
 \bibliography{laas_hri}

### 2013-07-11 paper.tex
Fixed many typos/style issues

@@
 
-\documentclass{elsarticle}
-
+\documentclass[preprint,12pt]{elsarticle}
+
+\usepackage[english]{babel}
 \usepackage{cite}
 \usepackage{graphicx, subfigure}
-%\usepackage{hyperref}
 \usepackage{url}
@@
 \DeclareGraphicsExtensions{.pdf,.jpg,.png}
-
-\usepackage{ifthen}
 
@@
 \newcommand{\stmt}[1]{{\footnotesize \tt $\langle$ #1\relax$\rangle$}}
-%\newcommand{\stmt}[1]{{\footnotesize $\langle\stmttt#1\relax\rangle$}}
-%\newcommand{\rawstmt}[1]{{\footnotesize \stmttt#1\relax}}
-%\def\stmttt#1 #2 #3\relax{
-%\texttt{\IfBeginWith{#1}{?}{\textbf{#1}}{#1} \IfBeginWith{#2}{?}{\textbf{#2}}{\emph{#2}} \ifthenelse {\equal{#3}{true} \OR \equal{#3}{false}}{\emph{#3}}{\IfBeginWith{#3}{?}{\textbf{#3}} {#3}}}}
-
-%\newcommand{\setstmt}[1]{{\footnotesize [\setstmttt#1\relax]}}
-%\def\setstmttt#1,#2\relax{\stmttt#1\relax, \stmttt#2\relax}
 
@@
 prerequisites of these challenges, and giving them experimental reality on the
-robots: what is required for sentences like ``Let's dress the table together''
+robots: what is required for sentences like ``Let's set the table together''
 to be understood by the robot, correctly interpreted in the spatial and
@@
 We have chosen to tackle the challenge from several ends: human-aware
-navigation and motion planning (\cite{Mainprice2011}), situation assessment
-coupled with motion planning (\cite{Mainprice2012}), projection of
-``mightabilities'' that anticipates what surrounding agents may do
-(\cite{Pandey2011}), and the design and deployment of a pervasive
-\emph{knowledge-centric} software architecture.
-
-The last item is the focus of this paper: what does ``knowledge-centric''
-architecture means, and what are its benefits and drawbacks.
+navigation and motion planning~\cite{Mainprice2011}, situation assessment
+coupled with motion planning~\cite{Mainprice2012}, projection of
+``mightabilities'' that anticipates what surrounding agents may
+do~\cite{Pandey2011}, and the design and deployment of a pervasive
+\emph{knowledge-oriented} software architecture.
+
+The last item is the focus of this paper: what ``knowledge-oriented''
+architecture means, and what its benefits and drawbacks are.
 
@@
 the functional level. It represents what we call a \emph{knowledge-oriented}
-architecture, built around a central knowledge management tool called the {\sc
-Oro} server.
-
-The following sections go over each of the facets of this deliberative layer:
+architecture, built around a central knowledge management tool called the
+OpenRobot Ontology ({\sc Oro}) server.
+
+The following sections present each of the facets of this deliberative layer:
 how knowledge is produced asynchronously from geometric reasoning, how the
@@
 used by decision-making components like the robot controller or the task
-planner, and finally, how our knowledge-centric architecture enables the
+planner, and finally, how our knowledge-oriented architecture enables the
 implementation of specific internal cognitive processes.
@@
 
-\subsection{The LAAS Deliberative Layer}
-
-Fig.~\ref{fig|archi} give an overview of the connections between the
+\subsection{The LAAS deliberative layer}
+
+Fig.~\ref{fig|archi} gives an overview of the connections between the
 deliberative components of our architecture~\cite{Alami2011}.
@@
 This architecture moves away from standard layered approaches. Interactions
-between components are mostly bidirectional and, from the software components
-point of view, we do not introduce layers of abstraction (we do, however, have
-access to the lower level modules of the robot to execute actions, but all
-cognition-related modules reside at the same level). This is especially visible
-for the dialogue input processing. This component does not simply act as an
-alternative perceptual input to the symbolic database, but also actively
-queries previously acquired knowledge to disambiguate and validate the newly
-created symbolic knowledge (see section~\ref{sect|com}).
-
-Our architecture relates but is to be distinguished from \emph{Beliefs,
-Desires, Intentions} (BDI) architectures. BDI architectures are primarily
-focused on \emph{practical reasoning}, \ie the process of deciding, step by
-step, which action to perform to reach a goal (as summarised by
-Woolridge~\cite{Woolridge1999}). The management of the interaction between
-knowledge (the beliefs) and task and plan representation and execution (the
-desires and the intentions) is central, and aims at selecting at each step the
-best subgoal. It becomes then an intention that the robot commits to.
+between components at the deliberative level are mostly bidirectional and we do
+not introduce layers of abstraction amongst software components\footnote{We do
+have lower-level modules to execute actions or manage sensors, but all
+cognition-related modules reside at the same level.}. This is nicely
+illustrated by the dialogue input processing. This component does not simply
+act as an alternative perceptual input to the symbolic database; it also
+actively queries previously acquired knowledge to disambiguate and validate the
+newly created symbolic knowledge (see section~\ref{sect|com}).
+
+Our architecture relates to \emph{Beliefs, Desires, Intentions} (BDI)
+architectures. BDI architectures are primarily focused on \emph{practical
+reasoning}, \ie the process of deciding, step by step, which action to perform
+to reach a goal (as summarized by Woolridge~\cite{Woolridge1999}). The
+management of the interaction between knowledge (the beliefs) and task and plan
+representation and execution (the desires and the intentions) is central, and
+aims at selecting at each step the best subgoal. It becomes then an intention
+that the robot commits to.
 
 This interaction between knowledge and actions is also central to our approach
-(as for any cognitive system), but task representation and task execution is
-not seen as a monolithic, central function: it is one of the activities of the
-robot, actually split between communication components (that can acquire
+(as for any cognitive system): it is one of the activities of the
+robot, actually shared between communication components (that can acquire
 desires from interaction with agents, amongst other things) and an execution
 controller that may decide to take an incoming desire into account to create
-its own internal goals. The controller generates and controls intentions from
-these goals with the help of a symbolic task planner, that has also direct
+its own internal goals. The controller generates and manages intentions from
+these goals with the help of a symbolic task planner, that also has direct
 access to the knowledge base.
 
-The architecture is not only focused on this workload, and other
-activities are conducted without being explicitly considered as desires:
-assessment of the situation and the environment, dialogue (including
+This activity is however not the backbone of our architecture. Other
+activities are conducted in parallel, without being explicitly considered as
+desires: assessment of the situation and the environment, dialogue (including
 performative dialogue that can possibly change the internal state of the robot,
@@
 accessible to decisional layers. In parallel, the knowledge base relies on
-reasoning in a \textit{top-down} way to produce new facts that may trigger in
-return physical behaviours.
+reasoning in a \textit{top-down} way to produce new facts that may in return
+trigger physical behaviours.
 
@@
 In our architecture (Fig.~\ref{fig|archi}), knowledge manipulation relies on a
-\emph{semantic blackboard}: one central server (the {\sc Oro}
+\emph{semantic blackboard}: a central server (the {\sc Oro}
 server~\cite{Lemaignan2010}) stores knowledge as it is produced by each of the
-deliberative components. It conversely exposes an API to query the knowledge
-base.
+deliberative components. It conversely exposes a {\tt json}-based RPC API to
+query the knowledge base~\cite{lemaignan2012kbapi}.
 
@@
 
-Relying on RDF triples and Description Logics has major advantages like the
-availability of numerous open-source libraries to manipulate the ontology,
+Relying on RDF triples and Description Logics has advantages such as the
+availability of numerous mature open-source libraries to manipulate the ontology,
 interoperability with several major on-line knowledge bases (like {\sc
@@
 
-It has also notable limitations: RDF triples imply only binary predicates
+It also has notable limitations, both fundamental (the suitability
+of Description Logics when reasoning on --typically non-monotonic-- commonsense
+knowledge is questionable) and practical: RDF triples imply only binary predicates
 (\stmt{subject predicate object}), which constrains the expressiveness of the
-system or leads to cumbersome reifications. Alternative exists (like {\sc
-KnowRob}~\cite{Tenorth2009a}) that mixes RDF with more expressive logic
+system or leads to cumbersome reifications. Alternatives exist (like {\sc
+KnowRob}~\cite{Tenorth2009a}) that mix RDF with more expressive logic
 languages like {\sc Prolog}, at the price, however, of other limitations, like
@@
 standard experiment (about 100 classes and 200 instances), classification
-typically take about 100ms, which becomes problematic during interactions.
+typically takes about 100ms, which becomes problematic during interactions.
 Besides, the performances are difficult to predict, since a seemingly
@@
 
-This knowledge model also largely exclude representation of continuous
-phenomena (like time) or uncertain phenomena. When required, these have to be
-managed by dedicated components.
-
-Amongst the features of {\sc Oro}, the system can dynamically creates
-new independent knowledge models (\ie independent instantiations of the
-ontology). This feature is used for instance to store separately the beliefs of
-each of the agents the robot interacts with (see section~\ref{sect|tom}).
+This knowledge model also largely excludes representation of continuous
+phenomena (like time) or uncertain phenomena. When required (for instance for
+action recognition), these are managed inside the corresponding components, and
+are not exposed at the semantic level.
+
+Tools to manipulate and reason over ontologies are readily available, mature,
+and already well accepted in the robotic community~\cite{Tenorth2009a,
+Lim2011}. While alternatives like \emph{Answer Set Programming} have also been
+successfully investigated in robotics~\cite{Chen2010,Erdem2012}, in particular
+to deal with non-monotonic reasoning, we did not actually hit any brick wall
+while working with OWL ontologies. We may reconsider this choice at a later
+stage, but until now it has proven an effective framework to quickly explore
+implementations of new cognitive abilities (for instance, it has been
+conceptually and technically easy to add support for independent knowledge
+models, one per the agents the robot interacts with --see
+section~\ref{sect|tom}).
+
+Besides, because ontologies and RDF statements are relatively simple concepts
+to grasp, it also effectively helped to grow awareness amongst colleagues on
+the significance of the ``semantic level'' when developing new components for
+the robot.
 
@@
 ability that a subject possesses to represent the mental state of another
-agent, possibly including knowledge that contradicts the subject own model: for
-example, a book can be at the same time \emph{visible} by myself, and \emph{not
-visible} by you.
-
-This skill, which is essential to understand others' perspectives during
-interactions, appears in children around their third year. It supposes the ability to build, store and retrieve separate models of the knowledge of the interactors.
+agent, possibly including knowledge that contradicts the subject's own model: for
+example, a book can be at the same time \emph{visible} for myself, and \emph{not
+visible} for you.
+
+Children develop this skill, which is essential to understand others' perspectives during
+interactions, around the age of three. It supposes the
+ability to build, store and retrieve separate models of the knowledge of the
+interactors.
 
@@
 
-One classical application of this cognitive skills is the so-called
+One classical application of this cognitive skill is the so-called
 \emph{False-Belief} experiment (also known as the \emph{Sally and Ann}
@@
 beliefs} on the world (A still thinks the object is at its original
-position since he did not see B hiding it). Using seperate knowledge models
-in the knowledge base, we have been able to replicate this experience on
+position since he did not see B hiding it). Using separate knowledge models
+in the knowledge base, we have been able to replicate this experience with
 our robots~\cite{Warnier2012a}.
@@
 
-The {\sc Oro} server also feature a mechanism to mimic simple forms of
+The {\sc Oro} server also features a mechanism to mimic minimalistic forms of
 biological memory.  When new statements are inserted in the knowledge base, a
@@
 We rely however on this short term memory for a particular use-case:
-\emph{active concept}. Some modules, like the natural language processor, use
+\emph{active concepts}. Some modules, like the natural language processor, use
 the {\tt short term} memory profile to mark for a few seconds important
@@
 action, and each red objects that are found are successively marked as
-\emph{active concepts} by inserting statements like \stmt{human type
+\emph{active concepts} by inserting statements such as \stmt{human type
 ActiveConcept} in the short-term memory (which can be considered, in this case,
-to be a working memory). We use this feature to give a (visual) feedback to the
-users.
+to be a working memory). We use this feature to trace certain knowledge-related
+processes.
 
@@
 their symbolic interpretation. We rely on a dedicated geometric and temporal
-reasoning module (as named on Fig.~\ref{fig|archi}) called SPARK (\emph{SPAtial
-Reasoning \& Knowledge}~\cite{Sisbot2011}). It is a situation assessment
-reasoner that generates symbolic knowledge from the geometry of the
-environment with respect to relations between objects, robots and humans,
-also taking into account the different perspective that each agent has on
-the environment (Fig.~\ref{fig|spark-oro}).
+reasoning module called SPARK (\emph{SPAtial Reasoning \&
+Knowledge}~\cite{Sisbot2011}). It is a situation assessment reasoner that
+generates symbolic knowledge from the geometry of the environment with respect
+to relations between objects, robots and humans (Fig.~\ref{fig|spark-oro}),
+also taking into account the different perspective that each agent has on the
+environment.
 
@@
 symbolic layer. This geometric model is continuously updated at run-time by the
-robot based on its sensors. In our experiments, objects are identified and
-localised through 2D barcodes.  On the other hand, the human pose is tracked
-with a Kinect-like device (assisted by motion capture to accurately track the
-head motion, which is required to compute what the human is looking at).
+robot based on its sensors (in our experiments, objects are identified and
+localised through 2D barcodes, while humans are tracked with Kinect-like
+devices, optionally assisted by motion capture to accurately track the head
+motion, which is required to compute what the human is looking at).
 
@@
 
-This allows us to deal with ambiguities that arises when one speaker refers to
+This allows us to deal with ambiguities that arise when one speaker refers to
 an object within a reference system (or changes the reference system, \ie
@@
 
-As a result, the robot stores models on the environment either in the
-\emph{ego-centric} reference frame (from the robot perspective) or in
-addressee-centred (\ie, \emph{allo-centric}) frames.
+As a result, the robot stores models of the environment either in the
+\emph{ego-centric} reference frame (from the robot perspective) or in the
+\emph{allo-centric} frame (addressee-centred).
 
@@
 Natural language processing is one of the fields of human-robot interaction
-where the introduction of the semantic layer has been most beneficial.
-We detail in~\cite{Lemaignan2011a} the techniques and the tool called
-\texttt{dialogs} that we have developed for English language parsing and
+for which the introduction of the semantic layer has been most beneficial.
+In~\cite{Lemaignan2011a} we detail the techniques and the tool called
+\texttt{dialogs} that we have developed for natural English language parsing and
 grounding, along with verbalisation and (minimalist) dialogue management.
@@
 with Google speech recognition) is parsed into a grammatical structure, and
-each sentence atoms are resolved with the help of the ontology to ground
+atoms of each sentence are resolved with the help of the ontology to ground
 concepts like objects (\ie when a user says ``pick the can'', resolve to which
@@
 
-The system supports quantification (``give me \{a | the | some | all | any | ...\} can''),
-thematic roles (action-specific predicates that qualify the actions), interactive disambiguation (the robot asks
-questions when it needs more information), anaphora resolution (``give \emph{it} to me'') based on
+The system supports quantification (``give me \{a | the | some | all | any |
+...\} can''), thematic roles (action-specific predicates that qualify the
+actions), interactive disambiguation (the robot asks questions when it needs
+more information), anaphora resolution (``give \emph{it} to me'') based on
 dialogue history. It also permits knowledge extension by learning new semantic
 structures (for instance, a sentence like ``learn that cats are animals'' is
-converted into \stmt{Cat subClassOf Animal}, interprets common temporal and place adverbs (like \emph{above} or \emph{tomorrow}) and translates to a certain extend \emph{states} (``I'm tired'') into
-\emph{experiences} (\stmt{HUMAN experiences state\_1, state\_1 hasFeature
-tired}).
+converted into \stmt{Cat subClassOf Animal}), interprets common temporal and
+place adverbs (like \emph{above} or \emph{tomorrow}) and translates to a
+certain extend \emph{states} (``I'm tired'') into \emph{experiences}
+(\stmt{HUMAN experiences state\_1, state\_1 hasFeature tired}).
 
@@
 
-Besides, even knowledge that belongs the \emph{common knowledge} may not be
+Besides, even knowledge that belongs to the \emph{common knowledge} may not be
 appropriate in a given interaction context. For instance, the system may
@@
 the different communication modalities (\emph{explicit} like verbal, deictic or
-based on gaze, or \emph{implicit} like postures) are presented in an
-homogeneous way.
-
-The dialogue grounding process makes use of them at two distinct levels.
+based on gaze, or \emph{implicit} like postures) are presented in a homogeneous
+way. The dialogue grounding process makes use of them at two distinct levels.
 
@@
 presence and value of specific facts: for instance, when a set of instances
-match a category (the human says ``give me the bottle'' and the robot knows
+matches a category (the human says ``give me the bottle'' and the robot knows
 about three bottles), the module may decide (it actually depends on the
 quantifier preceding the class) to discard some of them based on their
-\emph{visibility} by the speaker (implicit communication context built on the
+\emph{visibility} for the speaker (implicit communication context built on the
 human posture).
@@
 Note that, while the system benefits from the complementary modalities, they
-are not all required. The dialogue system would for instance run equally well
-with only the verbal modality.
+are not all required. The dialogue system would for instance happily run with
+only the verbal modality, at the cost of weaker interaction.
 
@@
 discrimination possibilities are available: for instance, if reachability is
-available, the robot may asks ``do you mean the bottle that is accessible to
+available, the robot may ask ``do you mean the bottle that is accessible to
 me?'' to discriminate between the three bottles. That way, procedures relying
@@
 decomposition into sub-tasks until reaching atomic actions. The \emph{planning
-domain} defines the set of methods describing how to decompose a task and can
-be seen as the procedural knowledge of the robot.
+domain} defines the set of methods describing how to decompose a task and
+represents the procedural knowledge of the robot.
 
@@
 Second remark, after some research (see appendix B of~\cite{Lemaignan2012a} for
-a detailed discussion)  we have decided not to represent neither the planning
+a detailed discussion)  we have decided to represent neither the planning
 domain nor the resulting plans in the knowledge base: the planning domain (with
-tasks pre- and postconditions) is stored in a specific format, outside of the
-knowledge repository, and the plans are directly communicated to the robot
-controller. From our experience, the OWL formalism is not a satisfying
-representation language for concepts that are both tightly bound to time
-representation and non-monotonic.
+task pre- and postconditions) is stored in a specific format, outside of the
+central declarative knowledge repository, and the plans are directly
+communicated to the robot controller. Thus, like many other cognitive
+architectures, we have independent declarative and procedural knowledge stores.
 
@@
 Our robot execution controllers (OpenPRS-based {\sc Shary} or Python-based {\sc
-pyRobots}) have deep integration with the knowledge base. It serves as the primary source of information for semantic-aware decision-making.
+pyRobots}) have deep integration with the knowledge base. It serves as the
+primary source of information for semantic-aware decision-making.
 
@@
 when the speaker says ``I feel tired'', we change the motion planner
-parametrization to lower the effort the human need to provide for the following
+parametrization to lower the effort the human needs to provide for the following
 joint manipulation tasks. Note that this example has been implemented as a
@@
 queries (based on the standard SPARQL language) or events. In its simplest
-form, a module can register an event by passing through an event pattern, which
-is basically a partial statement like \stmt{? type Elephant}, and a callback.
-Every time a new instance of elephant appears in the knowledge base, the
-callback is triggered.
+form, a module can subscribe to an event by passing through an event pattern
+(in its basic form, a partial statement like \stmt{? type Elephant}) and a
+callback.  Each time a new instance of elephant appears in the knowledge base,
+the callback is triggered.
 
@@
 The robot controller designer does not need to directly care about how this
-\emph{desire} is produced (this is delegated to perception modules), it can
-focuses on the semantic of the desire.
+\emph{desire} is produced (this is delegated to perception modules), he can
+focus on the semantic of the desire.
 
 Note also that we take advantage of the reasoning capabilities of the system:
-for example, the goal of the action (\stmt{action hasGoal myself}) may well not
-have been explicitly asserted, but inferred by the reasoner based on other
-assertions.
+for example, the goal of the action (\stmt{action hasGoal myself}) may not be
+explicitly asserted, but inferred by the reasoner based on other assertions.
 
@@
     \item{Knowledge is explicitly stored in one central and consistent
-    repository of facts, accessible by all modules.}
+    repository of facts, accessible for all modules.}
 
@@
 
-    \item{The first two points enable both a loosely-coupled architecture where
-    modules can very easily be removed or replaced by other ones as long as
+    \item{The first two points enable a loosely-coupled architecture where
+    modules can be removed or replaced easily by other ones as long as
     they share the same semantics (modules are defined by the knowledge they
@@
 
-    \item{and a \emph{symbolic}, reactive, event-driven approach to robot control.
+    \item{We adopt a \emph{symbolic}, reactive, event-driven approach to robot control.
     By managing events at the same level as the reasoner, we take full
@@
 
-We believe that this \emph{knowledge-oriented} approach has a strong potential
-not only to enable rich human-robot interaction, but also as a broader approach
-to information alignment and fusion in complex robotic systems.  The
+Those items are not new \emph{per se}. It seems however interesting to
+underline the shift of focus this implies during the design and integration
+phases. The
 versatility of this paradigm could be illustrated by a simple imaginary

### 2013-07-24 paper.tex
Added Mathieu and Aurelie to authors

@@
 
-\author{Rachid Alami and Séverin Lemaignan}
+\author{Rachid Alami, Séverin Lemaignan, Aurélie Clodic, Mathieu Warnier}
 
@@
 Univ de Toulouse, LAAS, F-31400 Toulouse, France\\
-{\tt severin.lemaignan@laas.fr}, {\tt rachid.alami@laas.fr}
+{\tt firstname.lastname@laas.fr}
 }

### 2013-07-24 paper.tex
Reinforce the AI orientation of the introduction

@@
 perspective-taking and affordances, management and exploitation of each agent
-(human and robot) knowledge in a separate cognitive model, human-aware task
-planning and human and robot interleaved plan achievement.
+(human and robot) knowledge in separate cognitive models, human-aware task
+planning, and human and robot interleaved plan achievement.
 
@@
     %% keywords here, in the form: keyword \sep keyword
-
+    human-robot interaction \sep cognitive robotics \sep perspective taking \sep knowledge representation and reasoning
     %% MSC codes here, in the form: \MSC code \sep code
@@
 
-Human-robot interaction requires to equip the robot with explicit
-reasoning on the human and on its own capacities to achieve its tasks
-in a collaborative way with a human partner. This paper presents a
-robot control system which has been especially designed for a
-cognitive robot which shares space and task with a human. We have
-adopted a constructive approach based on effective individual and
-collaborative skills. The system is comprehensive since it aims at
-dealing with a complete set of abilities articulated so that the robot
-controller is effectively able to conduct in a flexible manner a
-collaborative task with a human partner.
-
-We illustrate below how we deal with a typical human-robot interactive
-task achievement and what are the abilities we claim are
-necessary. Section \S\ref{sec:problem} proposes a typical human-robot
-interaction problem that can be solved by the proposed robot
-controller. Section \S\ref{sec:soa} reviews related work and analyses
-the context of our contribution.  Section \S\ref{sec:Framework}
-provides an overview of the robot controller and introduces three
-activities which are described in \S\ref{sec:situ}, \S\ref{sec:plan}
-and \S\ref{sec:action}. Finally, Section \S\ref{sec:expes} presents an
-effective run on a real robot in face to face interaction with a
-person.
+Human-robot interaction is a challenge for artificial intelligence. This field
+indeed lays at the crossroad of several other domains of AI and requires to
+tackle them in a holistic manner. Modeling humans and human cognition;
+acquiring, representing, manipulating in a tractable way abstract knowledge;
+reasoning on this knowledge to make decisions; instanciating those decisions
+into physical actions in coordination with humans... Many AI techniques are
+invited, from visual processing to symbolic reasoning, from task planing to
+\emph{theory of mind} building, from reactive control to action recognition and
+learning.
+
+We do not tackle all these issues. This article attempts however to organize
+them into a coherent challenge for artifical intelligence, and also explicit
+some of the trails that have been explored on our robots and that result in a
+deliberative, knowledge-focused, architecture for human-robot interaction.
+
+Human-robot interaction requires to equip the robot with explicit reasoning on
+the human and on its own capacities to achieve its tasks in a legible, if not
+collaborative, way with human partners. In the following pages, we present a
+robot control system which has been especially designed for a cognitive robot
+which shares space and task with a human. It has been designed by adopting a
+constructive approach based on effective individual and collaborative skills.
+The system is comprehensive since it aims\fxnote{it *is* comprehensive, or it *aims* at being comprehensive?} at dealing with a complete set of
+abilities articulated so that the robot controller is effectively able to
+conduct in a flexible manner a collaborative task with a human partner.
+
+We illustrate below how we deal with a typical human-robot interactive task
+achievement and what are the abilities we claim are necessary. Section
+\ref{sec:problem} proposes a typical human-robot interaction problem that can
+be solved by the proposed robot controller. Section
+\ref{sec:Framework} provides an overview of the robot controller and
+introduces three activities which are described in \ref{sec:situ},
+\ref{sec:plan} and \ref{sec:action}. Section \ref{sec:expes}
+presents the structure an actual run on a real robot in face to face interaction with a
+person. Finally, section~\ref{sec:soa} discusses and analyses the impact of
+each of our contribution in the light of existing literature and alternative
+approaches.
 
@@
 
-\begin{figure}[ht!]
+\begin{figure*}[ht!]
    \label{fig:sparkSubfigures}
@@
 
-\end{figure}
+\end{figure*}
 

### 2013-07-24 paper.tex
Added a number of fixmes + some text reorganization

@@
 
-\documentclass[preprint,12pt]{elsarticle}
+\documentclass[preprint,5p]{elsarticle}
+
+\usepackage[utf8x]{inputenc}
+\usepackage[T1]{fontenc}
 
@@
 
-\usepackage[utf8x]{inputenc}
-\usepackage[T1]{fontenc}
-
-\usepackage[draft]{fixme}
+\usepackage[draft, nomargin, marginclue, footnote]{fixme}
 
@@
 \usepackage{fancyvrb}
-\usepackage{flushend} % equalize the column length on last page
+
+%\usepackage[bookmarks]{hyperref}
 
@@
 
+\listoffixmes
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-\subsection{The typical human-robot interaction problem}\label{sec:problem}
-
-Let us consider a robot which is supposed to achieve interactive
-object manipulation, fetch and carry tasks and similar tasks in a
-domestic environment. The problem we are dealing with here is the
-following. Given:
-\begin {itemize}
-\item a joint goal, which has been previously established and agreed
-  upon (through a process which is out of the scope of this paper),
-\item the current situation, acquired through perception or
-  deduction from previous perceptions, including the state of the
-  environment of the robot and of the human,
-\end {itemize}
-the robot controller computes an action to execute and who (the 
-human or the robot, or both in case of a joint action) has to perform
-it, and then controls or monitors its execution. The operation
-continues until the goal is achieved, is declared unachievable or is
-abandoned by the human.
-
-To do so, the robot has to be equipped with a number of decisional, planning
-and interpretation abilities where its human partner is taken
-explicitly into account. It needs to be able:
-\begin {itemize}
-\item to build and maintain relevant robot and human beliefs
-  (from the robot perspective) with respect to state of the world and the task,
-\item to build and maintain iteratively shared (human-robot) plans, 
-\item to refine and execute the actions it has to perform, and to monitor 
-those achieved by its human partner.
-\end {itemize}
-
-Besides, we would like to build such abilities in a generic way, and
-to provide several levels of parametrization allowing to adapt to
-various environments, and various levels of involvement of the robot
-ranging from teammate behavior to assistant or proactive helper.
+\subsection{The human-robot interaction challenge}\label{sec:problem}
+
+\fxfatal{Develop/reorganize this section to organize the different problems of
+HRI into one coherent challenge for AI. See how to articulate it with section
+\ref{sec:psychocognitivegrounds}}
+
+We envision HRI in a context where two agents (a human and a robot)
+share a common space and exchange information through various
+modalities. Our aim is to endow the robot with an explicit
+consideration of the human and with the ability to manage its
+interactions with him (Figure~\ref{fig:hri-dec}). This must be
+considered at the architecture level as well as at the task/motion
+planning and execution level.
 
@@
 
-We envision HRI in a context where two agents (a human and a robot)
-share a common space and exchange information through various
-modalities. Our aim is to endow the robot with an explicit
-consideration of the human and with the ability to manage its
-interactions with him (Figure~\ref{fig:hri-dec}). This must be
-considered at the architecture level as well as at the task/motion
-planning and execution level.
+\paragraph{The Joint Action Challenge} Let us consider a robot which is
+supposed to achieve interactive object manipulation, fetch and carry tasks and
+similar tasks in a domestic environment. The problem we are dealing with here
+is the following. Given:
+
+\begin {itemize}
+\item a joint goal, which has been previously established and agreed
+  upon (through a process which is out of the scope of this paper),
+\item the current situation, acquired through perception or
+  deduction from previous perceptions, including the state of the
+  environment of the robot and of the human,
+\end {itemize}
+
+the robot controller computes an action to execute and who (the 
+human or the robot, or both in case of a joint action) has to perform
+it, and then controls or monitors its execution. The operation
+continues until the goal is achieved, is declared unachievable or is
+abandoned by the human.
+
+To do so, the robot has to be equipped with a number of decisional, planning
+and interpretation abilities where its human partner is taken
+explicitly into account. It needs to be able:
+
+\begin {itemize}
+\item to build and maintain relevant robot and human beliefs
+  (from the robot perspective) with respect to state of the world and the task,
+\item to build and maintain iteratively shared (human-robot) plans, 
+\item to refine and execute the actions it has to perform, and to monitor 
+those achieved by its human partner.
+\end {itemize}
+
+Besides, we would like to build such abilities in a generic way, and
+to provide several levels of parametrization allowing to adapt to
+various environments, and various levels of involvement of the robot
+ranging from teammate behavior to assistant or proactive helper.
 
@@
 
-Natural interaction and cooperation are on the (dare we say,
-\emph{short-term}) agenda for the human-robot interaction community. They are
-keys to the broad class of \emph{interactive manipulation problems}: several
-agents agree on a (more or less implicit) joint goal that requires some sort of
-cooperation to be successfully achieved. This class of problems involves both
-dialogue and manipulation and they are often initially underspecified: they
-require iterative and interactive resolution.
+The main scientific contributions of the article are summarized here, to guide
+the reader. We develop and discuss them, also in relation to other works, at
+the end of the article.
+
+\fxfatal{Rewrite section 'summary of contributions'}
 
@@
 do~\cite{Pandey2011}, and the design and deployment of a pervasive
-\emph{knowledge-oriented} software architecture.
-
-The last item is the focus of this paper: what ``knowledge-oriented''
-architecture means, and what its benefits and drawbacks are.
-
-\subsection{Article overview}
-
-In the next section, we first introduce the deliberative layer of our robots at
-the functional level. It represents what we call a \emph{knowledge-oriented}
-architecture, built around a central knowledge management tool called the
-OpenRobot Ontology ({\sc Oro}) server.
-
-The following sections present each of the facets of this deliberative layer:
-how knowledge is produced asynchronously from geometric reasoning, how the
-robot conducts grounded multi-modal interaction with humans, how knowledge is
-used by decision-making components like the robot controller or the task
-planner, and finally, how our knowledge-oriented architecture enables the
-implementation of specific internal cognitive processes.
-
-We conclude the article with a discussion about the strengths and weaknesses of
-such a knowledge-oriented architecture.
+\emph{knowledge-oriented} software architecture~\cite{Lemaignan2013}.
 
@@
 \subsection{The Psycho-Cognitive Grounds of Human-Robot Interaction}
-
-....summary of the challenges that human-robot interaction presents to robotics.
-
-\subsection{The LAAS deliberative layer}
+\label{sec:psychocognitivegrounds}
+
+\fxfatal{write section summary of the challenges that human-robot interaction presents to robotics.}
+
+\subsection{Building the Deliberative Layer}
+
+\fxfatal{Need to carefully introduce this section: how did we come up this this architecture?}
 
@@
 
+\fxfatal{Rewrite this section to level up AI aspects (Mathieu)}
+
 Building, updating and maintaining a correct state of the world at
@@
 (\stmt{HUMAN experiences state\_1, state\_1 hasFeature tired}).
+
+\fxwarning{Give more implementation details. Why did we go for a custom NLP? (Severin)}
 
@@
 
-%\subsubsection*{Agents and action streams:}
-\vspace{0.3cm}
-\noindent
-\textbf{Agents and action streams:}
+\paragraph{Agents and action streams}
 The robot plans not only for itself but also for the other agents. The
@@
 
-%\subsubsection*{Action costs and social rules:}
-\vspace{0.3cm}
-\noindent
-\textbf{Action costs and social rules:}
+\paragraph{Action costs and social rules}
 A cost and a duration function is associated to each action.
@@
 
-%\subsubsection*{Several levels of cooperation:} 
-\vspace{0.3cm}
-\noindent
-\textbf{Several levels of cooperation:} 
+\fxwarning{Relate HATP social rules to other AI techniques}
+
+\paragraph{Several levels of cooperation} 
 By tuning its costs
@@
 \section{Experimental results}
+
+\fxfatal{Most of the Experiment section needs to be written}
 
@@
 underline the shift of focus this implies during the design and integration
-phases. The
-versatility of this paradigm could be illustrated by a simple imaginary
-scenario with a blind robot and a deaf robot. The blind robot does not see (no
-cameras or alike), but someone can verbally describe a scene to it. On the
-other hand, the deaf robot has a good vision system, but cannot process verbal
-input.  Without any changes to the software architecture that we described,
-control modules of both robots could equally perform the same tasks since all
-the knowledge is abstracted and centralised (note that to actually implement
-this imaginary situation, the blind robot would of course need \textit{a
-priori} 3D models of objects talked about to enable planning or pick and place
-actions, and the deaf robot would require at least some gesture interpretation
-to understand orders).
+phases.
+
+\fxfatal{Insert here discussion of other major KRR systems for robots (Severin)}
 

### 2013-07-25 paper.tex
Minor tweak
Work on the related work/discussion section
Added many new references, reorganized the section

@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \begin{document}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \begin{abstract}
@@
 
-\subsection{Building the Deliberative Layer}
+\subsection{Building a human-aware deliberative layer}
 
@@
 
-\section{Perspectives}
+\section{Discussion: when AI enables HRI}
 \label{sect|conclusion}
 
-\subsection{Contributions and Related work}\label{sec:soa}
+This last section goes over again the main contributions we have presented
+so far: why are they relevant to robotics and artificial intelligence? How do
+they compare to similar approaches? How do we justify our technical choices and
+what remain to be developed/explored in the future?
+
+\subsection{Foundations of cognition for the robot}
+
+Because we build autonomous \emph{companion} robots, designed to live and
+interact in environments primarily conceived for humans, we endow our machines
+with a set of fundamental cognitive capabilities. The robots perform typically
+in dynamics, partially unknown environments, with a variety of places,
+artifacts and agents that may carry complex semantics.
+
+While our main focus is on the cognitive abilities required to interact with
+humans, \emph{situation assessment}, \emph{spatial reasoning}, \emph{knowledge
+representation and reasoning}, \emph{task planning} or \emph{high-level
+control} are mandatory cognitive foundations that we first have to mention.
+
+\paragraph{Situation assessment, spatial reasoning}
+
+Real service robots need to ground their decision-making processes in their
+perceptions, be it physical (\eg what a camera sees) or not (\eg remote
+knowledge bases, dialogue). We call \emph{physical situation assessment} the
+cognitive skill a robot exhibits when it assesses the nature and content of its
+surroundings and efficiently monitors its evolution.
+
+Numerous approaches exist, like amodal \emph{proxies}~\cite{Jacobsson2008},
+grounded amodal representations~\cite{Mavridis2006}, semantic maps
+(Figure~\ref{fig|semanticmap}, \cite{Nuechter2008, Galindo2008,Blodow2011}) or
+affordance-based planing and object classification~\cite{Lorken2008,
+Varadarajan2011}.
+
+Our approach, implemented in the {\sc Spark} module, is a grounded amodal
+representation of the robot environment, where different sensing modalities
+(vision-based fiducial markers tracking, RGBD human tracking, motion capture)
+are merged and where amodal spatial reasoning takes place (explanation
+generation for unexpected perceptions like a disappearing object).
+
+Spatial reasoning~\cite{O'Keefe1999} is a field in its own right, and has been
+used for natural language processing for applications such as direction
+recognition ~\cite{Kollar2010,Matuszek2010} or language
+grounding~\cite{Tellex2010}.~\cite{Skubic2004} presented a spatial reasoner
+integrated in a robot which computes symbolic positions of objects.
+
+One of the main limit of our current approach is the way semantics are
+conveyed: because we use 2D markers for artifacts, we skip most of the object
+extraction and recognition challenges. Each object has a unique identifier,
+which enable us to load a proper CAD model suitable for 3D spatial reasoning
+and prevent recognition mistakes in the knowledge base. While {\sc Spark}
+algorithms themselves are ignorant of the input sources, and would work equally
+well with a full object recognition stack, it is not currently supported.
+
+\fxwarning{spatial reasoning -> mention we do not support extensive temporal reasoning}
+
+\paragraph{Robot control}
+
+\fxwarning{Robot control related work -> what is the originality of SHARY? why is it relevant from an AI point of view? (Mathieu, Aurelie)}
+
+Beetz et al.~\cite{Beetz2010} proposes a cognitive architecture called
+\textsc{CRAM} (Cognitive Robot Abstract Machine) that integrates
+\textsc{KnowRob}~\cite{Tenorth2009a}, a knowledge processing framework based on
+Prolog. Its underlying storage is based on an OWL ontology, derived from
+\textsc{OpenCyc}. \textsc{CRAM} and \textsc{KnowRob} have been demonstrated on
+several real-world scenarios, where natural language recipes extracted from the
+Internet had to be translated into plans and executed in a kitchen environment,
+perceived and rebuilt on-line by the robots. While Prolog offers more flexible
+modeling (no constraints on the arity of predicate, where Description Logics as
+used in our work are limited to binary predicates), it is based on the closed
+world assumption (if something cannot be inferred to be true, it is inferred to
+be false) whereas we rely on the open world assumption, which is more realistic
+in real world scenarios. A probabilistic extension of \textsc{KnowRob}, called
+\textsc{ProbCog}~\cite{Jain2009} is also available. While in principle
+possible, currently the CRAM architecture does not provide explicit support for
+interacting with humans.
+
+\paragraph{Knowledge model}
+
+\subsection{Interacting with humans}
+
+\paragraph{Building a model of a human during the interaction} Perspective
+Taking is a human ability which allows one to put him/herself in another
+person's point of view. Studied in psychology
+literature~\cite{Flavell1992,Tversky1999}, this ability is crucial when
+interacting with people by allowing one to reason on others' understanding of
+the world in terms of visual perception, spatial descriptions, affordances and
+beliefs, etc.  Therefore, in the last years these notions have been gradually
+employed in Human-Robot Interaction.~\cite{Breazeal2006} presents a learning
+algorithm that takes into account information about a teacher's visual
+perspective in order to learn a task. ~\cite{Johnson2005} apply visual
+perspective taking for action recognition between two
+robots.~\cite{Trafton2005} use both visual and spatial perspective taking for
+finding out the referent indicated by a human partner.
+
+\fxwarning{What are our contribution regarding perspective taking? (Mathieu)}
+\fxwarning{Mention other source of knowledge about humans -- including social rules for instance}
+
+\paragraph{Reasoning with the humans}
+
+\fxwarning{Mention we represent the human knowledge as well (Severin)}
+
+\paragraph{Multi-modal communication}
+
+\subsection{Joint actions with humans}
+\label{sec:soa}
 
@@
 at the functional and at the deliberative levels~\cite{Klein2004}. The
-topics involve motion~\cite{Kulic2007,Berg2004,Madhav2006},
+topics involve motion~\cite{Kulic2007,Berg2004,Madhava2006},
 navigation~\cite{Althaus2004,Sisbot2007}, manipulation~\cite{Kemp2007}
@@
 theories~\cite{Rich1997,Sidner2005,Tambe2005a,Breazeal2003} and they
-all have shown benefits of this approach. They have also shown how
+all have shown benefits of this approach\fxwarning{which approach? how it is related to larger AI challenges?}. They have also shown how
 difficult it is to manage turn-taking between communication partners
 and to interleave task realization and communication in a generic
-way. Finally, today only few
-systems~\cite{Fong_2006,Breazeal2003,Sisbot2008} take humans into
-account at all levels.
-
-Perspective Taking is a human ability which allows one to put
-him/herself in another person's point of view. Studied in
-psychology literature~\cite{Flavell1992,Tversky1999}, this ability is
-crucial when interacting with people by allowing one to reason on
-others' understanding of the world in terms of visual perception, spatial
-descriptions, affordances and beliefs, etc.
-Therefore, in the last years these notions have been gradually
-employed in Human-Robot Interaction.~\cite{Breazeal2006} presents a
-learning algorithm that takes into account information about a
-teacher's visual perspective in order to learn a
-task. ~\cite{Johnson2005} apply visual perspective taking for action
-recognition between two robots.~\cite{Trafton2005} use both visual and
-spatial perspective taking for finding out the referent indicated by a
-human partner.
-
-Spatial reasoning~\cite{O'Keefe1999}, on the other hand, has been used
-for natural language processing for applications such as direction
-recognition ~\cite{Kollar2010,Matuszek2010} or language
-grounding~\cite{Tellex2010}.~\cite{Skubic2004} presented a spatial
-reasoner integrated in a robot which computes symbolic positions of
-objects.
+way\fxwarning{is it smthg we improve one?}. Finally, today only few
+systems~\cite{Fong2006,Breazeal2003,Sisbot2008} take humans into
+account at all levels\fxwarning{Old references, no new ones? we should also say what is new/better with our approach (Rachid)}.
 
@@
 underline the shift of focus this implies during the design and integration
-phases.
+phases: components of our deliberative layer are defined and glued together by
+the knowledge they produce and consume.
 

### 2013-07-25 paper.tex
Added section on Embodied Cognition

@@
 
+\fxwarning{robot control architecture for service robots: what else?}
+
 \paragraph{Knowledge model}
@@
 
+\subsection{Embodied Cognition}
+
+Robotics is traditionally regarded as the prototypical testbed for \emph{embodied}
+artificial intelligence, and this dimension is especially prevalent in
+human-robot interaction, where agents have to share a joint physical
+environment.
+
+It results in a tight coupling between the symbolic and the geometric realms:
+while AI at its origins was mostly a matter of symbolic models, it has been
+since recognised that not only the mind is not a purely abstract system,
+disconnected from the physical world, but even more, cognition fundamentally
+relies on its relation to the physical world (so-called \emph{embodied
+cognition}). Varela~\cite{Varela1992} is one of the main discoverer of these
+mechanisms, and coined the concept of \emph{enactivism} as the theoretical
+framework that study the links between cognition, embodiment and actions.
+
+The challenge of symbol grounding is tightly linked to this issue. It
+corresponds to the identification or creation, and then, maintenance of a link
+between the symbol (the syntactic form of knowledge the computer will
+manipulate) and its semantics, \ie its meaning, anchored in the world (the
+relations between the symbol, the referent of the symbol, and mediating minds
+is classically referred as the \emph{semantic triangle}, and has been
+extensively studied in linguistics). The issue of grounding is well known in
+cognitive science and is summarised by Harnard~\cite{Harnad1990} by this
+question: ``how the semantic interpretation of a formal symbol system can be
+made intrinsic to the system?''. This issue has a major practical importance in
+robotic: for a robot to be both endowed with a symbolic representational and
+reasoning system, and able to \emph{act} in the physical world, it must ground
+its knowledge.
+
+Grounding is implemented at different levels in our architecture. The main source
+of grounded symbolic knowledge is the situation assessment module, {\sc Spark}.
+Not only it builds symbolic facts from spatial reasoning, but it also relies on
+temporal reasoning to track the world state and build explanations to interpret
+unexpected perceptions (we mentioned above the example of a disappearing
+object).
+
+Because {\sc Spark} is also able to track humans, which enables in turn
+perspective-aware spatial reasoning, it produces grounded symbolic knowledge
+for interacting agents as well. This is a typical \emph{embodied} cognitive
+skill.
+
+Grounding also occurs during verbal and non-verbal communication. The {\sc
+Dialogs} module grounds new concepts introduced by the speaker by asking questions
+until it can attach the new concept to concepts already present in the
+knowledge pool (typically, if I ask the robot ``bring me a platypus'', the
+robot asks ``what is a platypus''. I may answer ``a mammal'' and the robot will
+ask ``what is a mammal?'' and so on until it anchors the new concepts to ones
+it already knows). Non verbal interaction (like gestures) is also partially
+grounded. When the user points at something, it can say ``give me this'' and
+the robot grounds \emph{this} to the pointed artifact.
+
+Because only objects marked with tags are currently recognized (in our typical
+lab experiments, it may be a dozen of them), the robot grounding mechanisms work
+in a small-sized closed world. This assumption obviously greatly facilitates
+the task. It remains to be seen if systematic marking of manipulated objects
+(for instance with RFID tags) is a realistic path. It may be the case if the
+robot is expected to be used in a well-defined, slowly changing environment.
+Real-time recognition of large libraries of objects is another path which may
+be viable soon~\cite{Dean2013Fast}.
+
 \subsection{The Next Steps}

### 2013-07-26 paper.tex
Added a table with the main experiments carried
Used as introduction to the 'experiements' section

@@
 in the knowledge base, we have been able to replicate this experience with
-our robots~\cite{Warnier2012a}.
+our robots~\cite{warnier2012when}.
 
@@
 
-\section{Experimental results}
+\section{Experimental outcomes}
+
+Our architecture has been deployed and tested in a number of experiments on
+several robotic platforms. Table~\ref{table|experiences} lists the most
+significant ones, with their main focuses and reference publication for the
+interested reader.
+
+\begin{table*}
+\begin{center}
+
+\begin{tabular}{llll}
+ \bf{Experiment} & Focus & \emph{Notes} & Reference \\
+\hline
+{\it Point \& Learn} (2010) & Interactive knowledge acquisition & & \cite{Lemaignan2010} \\
+{\it Spy Game} (2010) & Interactive object discrimination & & \cite{Ros2010b} \\
+{\it Moving to London} (2011) & Multi-modal interaction, perspective taking & \emph{immobile robot} & \cite{lemaignan2011what} \\
+{\it Roboscopie} (2011) & Theater, reflection on the future of HRI & \emph{semi-autonomous} & \cite{lemaignan2012roboscopie} \\
+{\it Cleaning the table} (2011) & Full integration & & \cite{Alami2011} \\
+{\it I'm in your shoes} (2012) & False beliefs & & \cite{warnier2012when} \\
+{\it Give me this} (2012) & Natural joint object manipulation & & \cite{gharbi2013natural} \\
+{\it Aperitif time} (2012) & Multi-modal interaction, perspective taking & & \cite{lemaignan2013talking} \\
+\hline
+
+\end{tabular}
+\end{center}
+\caption{Main experiments conducted with our cognitive architecture.}
+\label{table|experiences}
+\end{table*}
 

### 2013-07-26 paper.tex
Expand section on pyrobots. Give details on actions available

@@
 Our robot execution controllers (OpenPRS-based {\sc Shary} or Python-based {\sc
-pyRobots}) have deep integration with the knowledge base. It serves as the
+pyRobots}) rely on extensive integration with the knowledge base. It serves as the
 primary source of information for semantic-aware decision-making.
@@
 
-The action execution and monitoring task involves the motion and task
-planning, the effectors and the dedicated robot controller activity
-(Figure~\ref{architecture_fg}).
+Action execution and monitoring is the central activity of the robot executive
+controller, and involves interaction with most of the components presented on
+Figure~\ref{fig|archi}.
+
+Like most robotic architecture, actions are split into \emph{atomic actions}
+and more complex \emph{tasks}. Tasks are created either statically or dynamically.
+The {\sc pyRobots} controller, for instance, combines the actions listed in
+table~\ref{table|pyrobots_actions} to implement the high-level tasks it exposes: {\tt
+lookAt}, {\tt moveTo}, {\tt getFromHuman}, {\tt showObject}, {\tt giveObject},
+{\tt pickObject}, {\tt bringObject}, {\tt putObject}, {\tt hideObject}.
+
+Our other controller, {\sc Shary}, relies on the symbolic task planner to
+generate at runtime suitable tasks.
+
+\begin{table}
+\begin{center}
+\begin{tabular}{p{8cm}}
+\hline
+    {\bf Manipulation} \\
+     {\tt attachobject}, {\tt basicgive}, {\tt basictake}, {\tt close\_gripper}, {\tt configure\_grippers}, {\tt grab\_gripper}, {\tt handover}, {\tt hide}, {\tt open\_gripper}, {\tt pick}, {\tt put}, {\tt put\_accessible}, {\tt release\_gripper}, {\tt show}, {\tt take} \\
+\hline
+    {\bf Gaze control} \\
+     {\tt glance\_to}, {\tt look\_at}, {\tt sweep}, {\tt switch\_active\_stereo\_pair}, {\tt track}, {\tt cancel\_track} \\
+\hline
+    {\bf Navigation} \\
+     {\tt carry}, {\tt follow}, {\tt cancel\_follow}, {\tt goto}, {\tt moveclose}, {\tt waypoints} \\
+\hline
+    {\bf Local navigation} \\
+     {\tt dock}, {\tt rotate}, {\tt translate} \\
+\hline
+    {\bf Posture configuration} \\
+     {\tt extractpose}, {\tt idle}, {\tt manipose}, {\tt movearm}, {\tt rest}, {\tt setpose}, {\tt settorso}, {\tt tuckedpose} \\
+\hline
+\end{tabular}
+\end{center}
+\caption{Main {\sc pyRobots} \emph{actions}, sorted by categories. These
+actions are combined at runtime into higher-level \emph{tasks}.}
+
+\label{table|pyrobots_actions}
+\end{table}
 
 \paragraph{Human aware motion, placement and manipulation planning}
-In our scenarios, actions are only object manipulation actions: Pick,
-Place, Throw, Give. Motion and Task planning allows to compute final
-object placement, grasp and arm motion trajectories taking into
-account task specific constraints and human postures, abilities and
-preferences: see \cite{Sisbot2008, Mainprice2011, Pandey2010} for
-details.
+Manipulation and navigation actions listed in
+table~\ref{table|pyrobots_actions} rely on a collection of dedicated 3D motion
+planner to generate movements that are aware of surrounding humans. Robot
+placement and end-effectors trajectories are computed on-line to allow object
+manipulation that take into account task specific constraints and human
+postures, abilities and preferences. \cite{Sisbot2008, Mainprice2011,
+Pandey2010} detail these techniques.
 

### 2013-07-26 paper.tex
Tentative new title + abstract rewritten
The abstract is now more focused on AI

@@
 
+\usepackage{booktabs}
+%\usepackage{tabularx}
+
 \newcommand{\concept}[1]{{\small \texttt{#1}}}
@@
 \title{\LARGE \bf
-Human-Robot Interaction: A Deliberative Architecture
+Human-Robot Interaction: Tackling the AI Challenges
 }
@@
 
-This paper addresses some key decisional issues that are necessary for a
-cognitive robot which shares space and tasks with a human. We adopt a
+Human-Robot interaction is an area filled with challenges for artifical
+intelligence: dynamic, partially unknown environments that are not originally
+designed for autonomous machines; a large variety of situations and objects to
+deal with, with possibly complex semantics; physical interactions with humans
+that requires fine, low-latency control, representation and management of
+several mental models, good situation assessment skills...the list goes on.
+
+This article shed light on some key decisional issues that are to be tackled
+for a cognitive robot to share space and tasks with a human. We adopt a
 constructive approach based on the identification and the effective
-implementation of individual and collaborative skills. The system is
-comprehensive since it aims at dealing with a complete set of abilities
-articulated so that the robot controller is effectively able to conduct in a
-flexible manner a collaborative task with a human partner. These abilities
-include geometric reasoning and situation assessment based essentially on
-perspective-taking and affordances, management and exploitation of each agent
-(human and robot) knowledge in separate cognitive models, human-aware task
-planning, and human and robot interleaved plan achievement.
-
-This article presents our design choices, the articulations between the diverse
-deliberative components of the robot, and the strengths and weaknesses of our
-approach. We show that explicit knowledge management is not only a convenient
-tool from the software engineering point of view, but also pushes for a
-different, more \emph{semantic} way to address the decision-making issue in
-human-robot interactions.
+implementation of individual and collaborative skills. These cognitive
+abilities cover geometric reasoning and situation assessment based essentially
+on perspective-taking and affordances, management and exploitation of each
+agent (human and robot) knowledge in separate cognitive models, natural
+multi-modal communication, human-aware task planning, and human and robot
+interleaved plan achievement.
+
+We present our design choices, the articulations between the diverse
+deliberative components of the robot, experimental results, and eventually
+discuss the strengths and weaknesses of our approach. It appears that
+\emph{explicit} knowledge management, both symbolic and geometric, proves to be
+key as it pushes for a different, more \emph{semantic} way to address the
+decision-making issue in human-robot interactions.
 

### 2013-07-29 paper.tex
Extended a lot the section on knowledge model
Completed the state of art regarding knowledge management

@@
         \centering
-        \includegraphics[width=1.0\columnwidth]{archi}
+        \includegraphics[width=0.8\textwidth]{archi}
         \caption{Overview of the LAAS deliberative layer. Knowledge is
@@
 
-Relying on RDF triples and Description Logics has advantages such as the
-availability of numerous mature open-source libraries to manipulate the ontology,
-interoperability with several major on-line knowledge bases (like {\sc
-OpenCyc}, {\sc WordNet} or {\sc DBPedia}), open-world reasoning, and the formal
-guarantee of decidability (it is always possible to classify a Description
-Logics ontology).
-
-It also has notable limitations, both fundamental (the suitability
-of Description Logics when reasoning on --typically non-monotonic-- commonsense
-knowledge is questionable) and practical: RDF triples imply only binary predicates
-(\stmt{subject predicate object}), which constrains the expressiveness of the
-system or leads to cumbersome reifications. Alternatives exist (like {\sc
-KnowRob}~\cite{Tenorth2009a}) that mix RDF with more expressive logic
-languages like {\sc Prolog}, at the price, however, of other limitations, like
-closed-world reasoning or immutable T-Box. The classification
-performance is another issue: from our experience, with an ontology sized for a
-standard experiment (about 100 classes and 200 instances), classification
-typically takes about 100ms, which becomes problematic during interactions.
-Besides, the performances are difficult to predict, since a seemingly
-inoffensive new statement may indirectly change radically the logical
-complexity of the whole knowledge model and lead to notable degradation of
-classification time.
-
-This knowledge model also largely excludes representation of continuous
-phenomena (like time) or uncertain phenomena. When required (for instance for
-action recognition), these are managed inside the corresponding components, and
-are not exposed at the semantic level.
-
-Tools to manipulate and reason over ontologies are readily available, mature,
-and already well accepted in the robotic community~\cite{Tenorth2009a,
-Lim2011}. While alternatives like \emph{Answer Set Programming} have also been
-successfully investigated in robotics~\cite{Chen2010,Erdem2012}, in particular
-to deal with non-monotonic reasoning, we did not actually hit any brick wall
-while working with OWL ontologies. We may reconsider this choice at a later
-stage, but until now it has proven an effective framework to quickly explore
-implementations of new cognitive abilities (for instance, it has been
-conceptually and technically easy to add support for independent knowledge
-models, one per the agents the robot interacts with --see
-section~\ref{sect|tom}).
-
-Besides, because ontologies and RDF statements are relatively simple concepts
-to grasp, it also effectively helped to grow awareness amongst colleagues on
-the significance of the ``semantic level'' when developing new components for
-the robot.
+At any time, the knowledge available to the robot may come from three sources:
+Part of the knowledge is stored in an ontology (the {\sc OpenRobots Ontology}),
+loaded when the robot initializes. This static source covers the
+\emph{common-sense} knowledge of the robot, and optionally some
+scenario-specific knowledge (for instance about objects that are to be
+manipulated). The second part of the knowledge is acquired at runtime from
+perception, interaction, decision making.  We go over the processes it involves
+in the next section. Lastly, the third source of knowledge is the output of the
+reasoner.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\columnwidth]{top_tbox.pdf}
+
+    \caption{The upper part of the ORO common-sense TBox. All these concepts
+    belong to the {\sc OpenCyc} namespace.}
+
+    \label{fig|upper_tbox}
+\end{figure}
+
+\paragraph{The OpenRobots Ontology} The ORO common-sense ontology has been designed from two requirements: covering
+our experimental needs and conforming as much as possible to the {\sc OpenCyc}
+upper ontology.
+
+This lead to a bidirectional design process: from \emph{bottom-up} regarding
+the choices of concepts to model, \emph{top-down} regarding the upper part of the
+taxonomy. This upper part of the ontology is pictured on
+figure~\ref{fig|upper_tbox}. All the classes visible on this figure belong to the
+{\sc OpenCyc} namespace (the {\tt cyc:} prefix is omitted for clarity).
+
+By \emph{aligning} the upper part of the ontology on {\sc OpenCyc} (as other
+KRS, like {\sc KnowRob}~\cite{Tenorth2009a} or PEIS K\&R~\cite{Daoutis2009}, did)
+has multiple advantages. First the design of this part of the ontology is
+generally difficult: it pertains to abstract concepts whose mutual relations
+comes to philosophical debates. The upper taxonomy of {\sc OpenCyc} represents
+a relative consensus, at least within the semantic Web community. Then, because
+it is a well established project with numerous links to other on-line databases
+(like Wikipedia or WordNet), the reuse of important {\sc OpenCyc} concepts
+ensures to a certain extend that the knowledge stored by the robot can be
+shared or extended with well-defined semantics. A good example is the concept
+of \emph{Object}: In everyday conversation, an object is a relatively small
+physical thing, that can be typically manipulated. Normally, a human is not
+considered as an object. In {\sc Cyc}, an object has a more precise definition:
+it is something \emph{partially tangible}. That includes obviously the humans,
+and actually many other entities that would not be commonly said to be objects
+(the Earth for instance). Thus the importance of relying on well-defined and
+standard semantics to exchange informations between artificial systems.
+
+This figure~\ref{fig|upper_tbox} also illustrates the fundamental disjunction
+in the ORO model between \emph{temporal} and \emph{spatial} entities (formally,
+$(TemporalThing \sqcap SpatialThing)^{\mathcal{I}} = \emptyset$, with
+$\mathcal{I}$ the \emph{interpretation} of our model).
+
+The class \concept{purposeful action} is the superset of all the actions that
+are voluntarily performed by the robot (or another agent). Subclasses (like
+\concept{Give}, \concept{LookAt}, etc.) are not asserted in the common-sense
+ontology, but are added by the execution controller (in link with the symbolic
+task planner) and the natural language processor based on what is actually
+performable and/or understandable by the robot at run-time.
+
+The tree of figure~\ref{fig|upper_tbox} (this subset of the ontology is indeed
+a tree: this has however not to be the case in general, and, as a matter of
+fact, the TBox of the whole ORO common-sense ontology does not form a tree) is
+not equally developed at lower levels. We have already briefly mentioned the
+developments of the actions. The other important part is the descendants of the
+\concept{partially tangible thing} (what is commonly called an \emph{object}).
+Figure~\ref{fig|tangible_things_tbox} gives more details on this part of the
+ontology.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=0.6\columnwidth]{tangible_things_tbox.pdf}
+    \caption{TBox of the specialisations of \concept{PartiallyTangible}.}
+    \label{fig|tangible_things_tbox}
+\end{figure}
+
+This excerpt from the ontology makes the bottom-up design process visible: only
+few types of \emph{partially tangible} things appear, and only subclasses
+relevant to the context of service robotics in an human-like environment are
+present.
+
+Lastly, the ORO common-sense ontology contains several rules and class
+expressions that encode non-trivial inferences.
+
+The definition of the \concept{Bottle} is a case in point:
+
+\concept{Bottle} $\equiv$ \concept{Container} {\bf and} \concept{Tableware}
+{\bf that} (\concept{hasShape} {\bf value} \concept{cylinderShape} {\bf and}
+\concept{hasCharacteristicDimension} {\bf only} \concept{\em int[>= 0.1, <=
+0.3]})
+
+If a human informs the robot that a given object is indeed a bottle, the robot
+can then infer much more on this object. And if the human affirms that a car is
+a bottle, the robot may question this assertion because of the inconsistent
+size.
+
+We discuss in section~\ref{krs-discussion} the strengths and weaknesses of our
+knowledge framework.
 
@@
 
-\paragraph{Knowledge model}
+\subsection{Knowledge framework}
+\label{krs-discussion}
+
+Knowledge flows in our architecture through a \emph{semantic blackboard}, the
+{\sc Oro}-server. We rely on Description Logics (OWL) to represent and
+manipulate facts.
+
+Relying on RDF triples and Description Logics has decisive advantages such as
+the availability of numerous mature open-source libraries to manipulate the
+ontology, interoperability with several major on-line knowledge bases (besides
+{\sc OpenCyc}, {\sc WordNet}, {\sc DBPedia} or {\sc
+RoboEarth}~\cite{Waibel2011} are other example), open-world reasoning, and the
+formal guarantee of decidability (it is always possible to classify a
+Description Logics ontology).
+
+It also has notable limitations, both fundamental (the suitability
+of Description Logics when reasoning on --typically non-monotonic-- commonsense
+knowledge is questionable) and practical: RDF triples imply only binary predicates
+(\stmt{subject predicate object}), which constrains the expressiveness of the
+system or leads to cumbersome reifications. Alternatives exist (like {\sc
+KnowRob}~\cite{Tenorth2009a}) that mix RDF with more expressive logic
+languages like {\sc Prolog}, at the price, however, of other limitations, like
+closed-world reasoning or immutable T-Box. The classification
+performance is another issue: from our experience, with an ontology sized for a
+standard experiment (about 100 classes and 200 instances), classification
+typically takes about 100ms, which becomes problematic during interactions.
+Besides, the performances are difficult to predict, since a seemingly
+inoffensive new statement may indirectly change radically the logical
+complexity of the whole knowledge model and lead to notable degradation of
+classification time.
+
+This knowledge model also largely excludes representation of continuous
+phenomena (like time) or uncertain phenomena. When required (for instance for
+action recognition), these are managed inside the corresponding components, and
+are not exposed at the semantic level.
+
+Tools to manipulate and reason over ontologies are readily available, mature,
+and already well accepted in the robotic community~\cite{Tenorth2009a,
+Lim2011}. While alternatives like \emph{Answer Set Programming} have also been
+successfully investigated in robotics~\cite{Chen2010,Erdem2012}, in particular
+to deal with non-monotonic reasoning, we did not actually hit any brick wall
+while working with OWL ontologies. We may reconsider this choice at a later
+stage, but until now it has proven an effective framework to quickly explore
+implementations of new cognitive abilities (for instance, it has been
+conceptually and technically easy to add support for independent knowledge
+models, one per agent the robot interacts with --see section~\ref{sect|tom}).
+
+Besides, because ontologies and RDF statements are relatively simple concepts
+to grasp, it also effectively helped to grow awareness amongst colleagues on
+the significance of the ``semantic level'' when developing new components for
+the robot.
 

### 2013-08-09 paper.tex
Partial proofreading, light reogranization of chap 3

@@
 \newcommand{\eg}{{\textit{e.g.\ }}}
+\newcommand{\al}{{\textit{et al.\ }}}
 
@@
 
-SPARK is an \emph{amodal} geometric model of the environment that serves both
-as basis for the fusion of the perception modalities and as bridge with the
-symbolic layer. This geometric model is continuously updated at run-time by the
-robot based on its sensors (in our experiments, objects are identified and
-localised through 2D barcodes, while humans are tracked with Kinect-like
-devices, optionally assisted by motion capture to accurately track the head
-motion, which is required to compute what the human is looking at).
+SPARK is an \emph{amodal}~\cite{Mavridis2006} geometric model of the
+environment that serves both as basis for the fusion of the perception
+modalities and as bridge with the symbolic layer. This geometric model is
+continuously updated at run-time by the robot based on its sensors (in our
+experiments, objects are identified and localised through 2D barcodes, while
+humans are tracked with Kinect-like devices, optionally assisted by motion
+capture to accurately track the head motion, which is required to compute what
+the human is looking at).
+
+\fxnote{Add 2-3 lines to summarize the AI highlights of SPARK}
 
@@
 
-\paragraph{Symbolic facts production:}
+\paragraph{Symbolic facts production}
 
@@
 Reasoning about human perspective allow to compute facts such as:
-\stmt{GREY\_TAPE isBehind HUMAN1}, \stmt{GREY\_TAPE isVisibleBy
+\stmt{GREY\_TAPE isBehind HUMAN1}, \stmt{GREY\_TAPE leftOf
   HUMAN1}.
@@
 Figure~\ref{fig::reach-ex} illustrates different situations for the
-\textit{reachable} relation. In this case, the robot and its human partner
-are placed face to face, in a table-top setup (Figure~\ref{fig::reach-ex}.1).
-The robot first estimates if the small grey box is reachable to itself. This is
-done by finding a collision free posture to reach the object
-(Figure~\ref{fig::reach-ex}.2). Next the robot switches to the human's
-perspective to estimate if the same object is reachable to the human as well.
-In the last scene, the human moves towards his left, farther from the object
-(Figure~\ref{fig::reach-ex}.4). The situation is then reevaluated. In this
-occasion though, the reasoner cannot find a satisfactory posture for the human
-to reach the box because he is too far from the target.
+\textit{reachable} relation. The robot and its human partner are placed face to
+face (Figure~\ref{fig::reach-ex}.1).  The robot first estimates if the small
+grey box is reachable to itself. This is done by finding a collision free
+posture to reach the object (Figure~\ref{fig::reach-ex}.2). Next the robot
+switches to the human's perspective to estimate if the same object is reachable
+to the human as well.  In the last scene, the human moves towards his left,
+farther from the object (Figure~\ref{fig::reach-ex}.4). The situation is then
+reevaluated. In this occasion though, the reasoner cannot find a satisfactory
+posture for the human to reach the box because he is too far from the target.
 
@@
 	\centering
-	\includegraphics[width=1.0\textwidth]{figs/reachex.jpg}	
+    \subfigure[] { \includegraphics[width=0.2\textwidth]{figs/reachex-1.jpg} }
+    \subfigure[] {\includegraphics[width=0.2\textwidth]{figs/reachex-2.jpg} }
+    \subfigure[] {\includegraphics[width=0.2\textwidth]{figs/reachex-3.jpg} }
+    \subfigure[] {\includegraphics[width=0.2\textwidth]{figs/reachex-4.jpg} }
 	\caption{An example illustrating the \textit{reachable} relation.
@@
 \begin{figure*}[ht!]
-   \label{fig:sparkSubfigures}
    \begin{center}
@@
      the human HUMAN1. All tapes are on the table TABLE.  }%
+   \label{fig:sparkSubfigures}
 
@@
 the beliefs using input from exploration, dialog, human visual
-focus,\ldots
+focus,\ldots \fxnote{Paragraph need write. Tell what the module actually does, not what it could do}
+
 SPARK currently provides a simple implementation of such a functionality. The
@@
 
-\paragraph{Primitive action recognition}
-Monitoring human activity is crucial to maintain a coherent state of
-the world. Full human action and activity monitoring is a difficult
-task that requires knowledge and reasoning both on high level facts
-like goals, intentions and plans, as well as bottom-up data from agent
-and object motions. Simple temporal and geometric reasoning on human
-hand trajectories and potential objects placements can provide some
-useful clues for high level human monitoring processes. We call this
-temporal and geometric reasoning \emph{primitive action recognition}.
+\paragraph{Primitive action recognition} Monitoring human activity is crucial
+to maintain a coherent state of the world. Full human action and activity
+monitoring is a task that requires knowledge and reasoning both on high level
+facts like goals, intentions and plans, as well as bottom-up data from agent
+and object motions. Simple temporal and geometric reasoning on human hand
+trajectories and potential objects placements can provide some useful clues for
+high level human monitoring processes. We call this temporal and geometric
+reasoning \emph{primitive action recognition}.
 
@@
 object is close to a container, etc. Human hand position is either
-directly perceived or inferred from its initial perceived trajectory.
+directly perceived or inferred from its initial perceived trajectory\fxwarning{we really do that?}.
 We have a simple implementation of such a primitive action recognition
@@
 of the environment. These models are separately stored in the knowledge base.
+\fxwarning{Rephrase: already mentioned before}
 
@@
 As a result, the robot stores models of the environment either in the
-\emph{ego-centric} reference frame (from the robot perspective) or in the
-\emph{allo-centric} frame (addressee-centred).
+\emph{ego-centric} reference frame (from the robot perspective) or in 
+\emph{allo-centric} frames (addressee-centred).
 
 \subsubsection{Symbolic facts and beliefs management}
+\fxwarning{Do we need this section here?}
 
 The facts produced by the geometric and temporal reasoning component
-are stored in a central symbolic knowledge base, called ORO. Besides
-acting as a facts database, the ORO platform~\cite{Lemaignan2010}
-exposes several functions: operations on knowledge statements relying
-on inference (through a continuous first-order logic classification
-process), management of \emph{per-agent} symbolic models, and also
-higher cognitive and human-robot interaction related functionalities
-like categorization of sets of concepts, profiles of memory (that
-enable the robot to ``forget'' about some facts), natural language
-grounding~\cite{Lemaignan2011}\ldots.
+are stored in the central symbolic knowledge base.
 
@@
 
-ORO also provides an event mechanism that allows components to be
-triggered when specific events occur. A component can
-for instance subscribe to events of kind \stmt{?agent isVisible
-  true, ?agent type Human}. As soon as the perception layer detects a
-human in the robot's field of view and accordingly updates the
-knowledge base, the executive layer would be triggered back. The
-event framework also takes advantage of the inference capabilities of
-ORO. Thus an event can be indirectly triggered if its triggering
-conditions can be inferred to be true.
-
-\subsubsection{Context Management Policies}
-
-\fxfatal{Rewrite this section to level up AI aspects (Mathieu)}
-
-Building, updating and maintaining a correct state of the world at
-geometric and symbolic level is crucial to the capacity of the robot
-to carry on successfully a multi-step interaction with a human. Tight
-integration between the robot controller and the geometric and
-temporal reasoning functions in SPARK and symbolic facts and beliefs
-management in ORO is central.
-
-The robot controller has access to the symbolic facts in ORO that are
-automatically updated whenever object and agent positions are changed.
-Robot controller can also access geometric perceived or inferred
-positions of objects and geometric positions and postures of the human
-that will be used to orient its cameras.  Building and updating the
-state of the world first relies on perceiving objects. Robot
-controller can use:
-
-\begin {itemize}
-\item Exploration policies: robot will exhaustively scan the table to
-  see all what can be seen.
-
-\item Search policies: robot will search an object until it is
-  detected if possible, scanning all the table and looking in human
-  hand.
-
-\end {itemize}
-
-Robot reasons on possible positions for non perceived
-objects. These hypotheses could be updated using new input from dialogue, human
-action and focus of attention. Currently, we manage at most one hypothesis per
-object. This hypothesis is produced by robot controller through an inference on
-robot or human action. In case of perception conflicts with low probability for
-the current hypothesis, robot controller will break this hypothesis and delete
-corresponding symbolic fact in the ontology.
-
-Robot must reacts to change in the world not linked to robot
-action to drive world update. Robot controller uses SPARK to monitor human hand
-motion and primitive action recognition for \emph{pick} and \emph{throw}. As
-mentioned above these primitive action recognition should be used with higher
-level information on goals, intentions and plans and some exploration to
-achieve complex human action and activity monitoring.  In the current
-implementation, these primitive actions are over-optimistically interpreted as
-the corresponding actions \emph{pick} object and \emph{throw} object.
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 instance of \emph{can} the user is referring to) and actions. Sentences are
-sorted into questions, desires and statements, and processed accordingly.
+sorted into questions, desires and statements, and processed accordingly.\fxwarning{Severin: Develop this section}
 
@@
 structures (for instance, a sentence like ``learn that cats are animals'' is
-converted into \stmt{Cat subClassOf Animal}), interprets common temporal and
-place adverbs (like \emph{above} or \emph{tomorrow}) and translates to a
-certain extend \emph{states} (``I'm tired'') into \emph{experiences}
-(\stmt{HUMAN experiences state\_1, state\_1 hasFeature tired}).
+converted into \stmt{Cat subClassOf Animal} and added to th eknowledge pool),
+interprets common temporal and place adverbs (like \emph{above} or
+\emph{tomorrow}) and translates to a certain extend \emph{states} (``I'm
+tired'') into \emph{experiences} (\stmt{HUMAN experiences state\_1, state\_1
+hasFeature tired}).
 
@@
 First, particular steps of the grounding process explicitly check for the
-presence and value of specific facts: for instance, when a set of instances
+presence and value of specific facts: for instance, when several instances
 matches a category (the human says ``give me the bottle'' and the robot knows
@@
 Note that, while the system benefits from the complementary modalities, they
-are not all required. The dialogue system would for instance happily run with
-only the verbal modality, at the cost of weaker interaction.
+are not all required. The dialogue system can run with only the verbal
+modality, at the cost of a simpler interaction.
 
@@
 
-Our execution controllers rely on symbolic task planning to convert faraway
+Our execution controllers rely on symbolic task planning to convert long-term
 desires into a succession of atomic actions. We use in our architecture the
-HATP planner (for Human Aware Task Planner~\cite{Alili2009}).  HATP is based on
-a Hierarchical Task Network (HTN) refinement, which performs an iterative task
-decomposition into sub-tasks until reaching atomic actions. The \emph{planning
-domain} defines the set of methods describing how to decompose a task and
-represents the procedural knowledge of the robot.
-
-In order to devise how a given goal can be accomplished, the robot has
-to elaborate a plan,~\textit{i.e.} a set of actions to be achieved by
-the robot and its human partners.  This is the role of a HATP
-\cite{Alili2008} (for Human Aware Task Planner).  HATP is based on a
-Hierarchical Task Network (HTN) refinement which performs an iterative
-task decomposition into sub-tasks until reaching atomic
-actions~\cite{Nau2003}.  The planning domain defines a set of methods
-describing how to decompose a task and can be seen as the Howto
-knowledge of the robot.  HATP is able to produce plans for the robot's
-actions as well as for the other participants (humans or robots). It
-can be tuned by setting up different costs depending on the actions to
-apply and by taking into account a set of constraints called social
-rules. This tuning aims at adapting the robot's behavior according to
-the desired level of cooperation of the robot.
+HATP planner (for Human Aware Task Planner~\cite{Alili2008, Alili2009}).  HATP
+is based on a Hierarchical Task Network (HTN) refinement, which performs an
+iterative task decomposition into sub-tasks until reaching atomic
+actions\fxwarning{Rephrase: the definition is actually generic to HTN, not HATP
+specfic}. The \emph{planning domain} defines the set of methods describing how
+to decompose a task and represents the procedural knowledge of the robot. It is
+stored outside of the central knowledge base.
+
+HATP is able to produce plans for the robot's actions as well as for the other
+participants (humans or robots). It can be tuned by setting up different costs
+depending on the actions to apply and by taking into account a set of
+constraints called social rules. This tuning aims at adapting the robot's
+behavior according to the desired level of cooperation of the robot.
 
@@
 
+\paragraph{Context Management Policies} \fxfatal{Rewrite this section to level up AI aspects (Mathieu)}
+Building, updating and maintaining a correct state of the world at
+geometric and symbolic level is crucial to the capacity of the robot
+to carry on successfully a multi-step interaction with a human. Tight
+integration between the robot controller and the geometric and
+temporal reasoning functions in SPARK and symbolic facts and beliefs
+management in ORO is central.
+
+The robot controller has access to the symbolic facts in ORO that are
+automatically updated whenever object and agent positions are changed.
+Robot controller can also access geometric perceived or inferred
+positions of objects and geometric positions and postures of the human
+that will be used to orient its cameras.  Building and updating the
+state of the world first relies on perceiving objects. Robot
+controller can use:
+
+\begin {itemize}
+\item Exploration policies: robot will exhaustively scan the table to
+  see all what can be seen.
+
+\item Search policies: robot will search an object until it is
+  detected if possible, scanning all the table and looking in human
+  hand.
+
+\end {itemize}
+
+Robot reasons on possible positions for non perceived objects. These hypotheses
+could be updated\fxwarning{are they?} using new input from dialogue, human action and focus of
+attention. Currently, we manage at most one hypothesis per object. This
+hypothesis is produced by robot controller through an inference on robot or
+human action. In case of perception conflicts with low probability for the
+current hypothesis, robot controller will break this hypothesis and delete
+corresponding symbolic fact in the ontology.
+
+Robot must reacts to change in the world not linked to robot action to drive
+world update. Robot controller uses SPARK to monitor human hand motion and
+primitive action recognition for \emph{pick} and \emph{throw}. As mentioned
+above these primitive action recognition should be used with higher level
+information on goals, intentions and plans and some exploration to achieve
+complex human action and activity monitoring.  In the current implementation,
+these primitive actions are over-optimistically interpreted as the
+corresponding actions \emph{pick} object and \emph{throw} object.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2013-08-30 paper.tex
Completed the 'experiment' section
Since no input from LAAS, re-using 'Moving to London' scenario.
While here, corrected/fixed/simplified the 'Cleaning the table' description

@@
 several robotic platforms. Table~\ref{table|experiences} lists the most
-significant ones, with their main focuses and reference publication for the
+significant ones, with their main focuses and reference publications for the
 interested reader.
@@
 
-\fxfatal{Most of the Experiment section needs to be written}
-
-We assume here that the robot (and the human) has been given the joint goal
-``CLEAN TABLE''. For HATP, this means putting all tapes that are
-currently on the table in the trashbin. Depending on the state of the
-world and agent preferences, different plans are produced.
-
-Figure~\ref{plan-etat2} shows a plan produced to clean the table based on the initial the initial state given in  \S~\ref{sub:gtrc}.
+The two following sections present one such experiment (the {\it Moving to
+London} scenario), and the synopsis of another one (the {\it Cleaning the
+table} scenario).
+
+The first one is focused on knowledge representation and verbal interaction:
+the human asks for help to find and pack objects. In this experiment, robot
+actions are limited to simple motions (like head tracking or predefined
+pick-and-place).
+
+The second one involves the SHARY execution controller and the HATP symbolic
+task planner besides ORO and {\sc Dialogs}. In this scenario, the human and the
+robot try to cooperatively remove objects from a table. In this scenario,
+behaviours and motions are fully planned.
+
+\subsection{\emph{Moving to London} scenario}
+
+This first experiment is based on the following daily life situation: Tom and
+Jerry are moving to London, so they are packing things in boxes. The scenario
+takes places in the living-room, where Jido (our robot) is observing while they
+move things here and there. To assess the reasoning abilities of the robot they
+ask Jido for information (entered through keyboard). Ideally, the robot should
+also perform actions when required (e.g. hand an object when asking ``give
+me...''). However, since it is out of the scope of this work, we do not include
+any motion from the robot's side.
+
+Perception of objects is done through a tag-based system and humans are
+detected through motion capture. The robot knowledge base is pre-loaded with
+the \emph{ORO Commonsense Ontology}.  We next describe in detail two
+situations where we can follow the internal robot's reasoning and the
+interaction with the users.
+
+\paragraph{Implicit disambiguation through visual perspective taking}
+
+\begin{figure}[!ht]
+  \centering
+  \includegraphics[width=0.9\linewidth]{./figs/pt.jpg}
+\caption{Asking the robot to hand over a tape on the table.}
+  \label{fig|vpt}
+\end{figure}
+
+Tom enters the room while carrying a big box (Figure~\ref{fig|vpt}, page 1). He
+approaches the table and asks Jido to handle him the video tape: ``Jido, can
+you give me the video tape''. The \textsc{Dialogs} module queries the ontology to
+identify the object the human is referring to: \stmt{?obj type VideoTape}.
+
+There are two video tapes in the scene: one on the table, and another one
+inside the cardboard box. Thus, the knowledge base returns both: $\Rightarrow$
+\concept{?obj = [videoTape1, videoTape2]}.
+
+However, only one is visible for Tom (the one on the
+table). Thus, although there is an ambiguity from the robot's perspective
+(since it can see both video tapes), based on the perspective of its human
+partner it infers that Tom is referring to the video tape on the table, and not
+the one inside the box which is not visible from his view. Therefore,
+non-visible objects are removed obtaining: \concept{?obj =[videoTape1]}.
+
+Since only one object is available, the robot infers
+that the human refers to it and would eventually execute the command, \ie give
+it to the human. Alternatively, the robot could first verify with the human if
+that was the object being referred to or not before proceeding to execute the
+action. Table~\ref{table|ptbeliefs} lists the robot's beliefs about itself and
+its human partner involved in this situation.
+
+\begin{table}
+\begin{center}
+\begin{tabular}{l}
+\hline
+Robot's beliefs about itself (\emph{robot's model}):\\
+\hline
+  \hspace{0.7cm}\stmt{videoTape1 type VideoTape}\\
+  \hspace{0.7cm}\stmt{videoTape1 isOn table}\\
+  \hspace{0.7cm}\stmt{videoTape1 isVisible true}\\
+  \hspace{0.7cm}\stmt{videoTape2 type VideoTape}\\
+  \hspace{0.7cm}\stmt{videoTape2 isIn cardBoardBox}\\
+  \hspace{0.7cm}\stmt{videoTape2 isVisible true}\\
+\hline
+\hline
+Robot's beliefs about Tom (\emph{Tom's model}):\\
+\hline
+  \hspace{0.7cm}\stmt{videoTape1 type VideoTape}\\
+  \hspace{0.7cm}\stmt{videoTape1 isOn table}\\
+  \hspace{0.7cm}\stmt{videoTape1 isVisible true}\\
+  \hspace{0.7cm}\stmt{videoTape2 type VideoTape}\\
+  \hspace{0.7cm}\stmt{videoTape2 isIn cardBoardBox}\\
+  \hspace{0.7cm}\stmt{videoTape2 isVisible false}\\
+ \hline
+\end{tabular}
+\end{center}
+\caption{Robot's beliefs about itself and its human partner.}
+\label{table|ptbeliefs}
+\end{table}
+
+\paragraph{Explicit disambiguation through verbal interaction and gestures}
+
+\begin{figure}[!ht]
+  \centering
+  \includegraphics[width=0.9\linewidth]{./figs/inTheBox2.jpg}
+\caption{Jerry asks Jido for the content of the box by pointing at it.}
+  \label{fig|pointing}
+\end{figure}
+
+In this situation, Jerry enters the
+living room without knowing where Tom had placed the video tapes. So he first
+asks Jido: ``What's in the box?''. Before the robot can answer the question it
+has to figure out which box Jerry is talking about. Similar to the previous
+situation, there are two available boxes:
+
+\begin{center}
+\begin{tabular}{l}
+\stmt{?obj type box}\\
+\hspace{0.7cm}$\Rightarrow$ {\tt ?obj = [cardBoardBox, toolbox]}
+\end{tabular}
+\end{center}
+
+However both are visible and the cognitive ambiguity resolution cannot be
+applied. The only option is to ask Jerry which box he is referring to: ``Which
+box, the toolbox or the cardboard box?'' Jerry could now simply answer the
+question. Instead, he decides to point at it while indicating: ``This box''
+(Figure~\ref{fig|pointing}). The robot's perception identifies the {\tt
+cardBoardBox} as being pointed at and looked at by the human and updates the
+ontology with this new information using a rule available in the commonsense
+ontology ({\tt \textbf{pointsAt}(?ag, ?obj) $\land$ \textbf{looksAt}(?ag, ?obj) $\to$
+\textbf{focusesOn}(?ag, ?obj)}) The \textsc{Dialogs} module is then able to merge both
+sources of information, verbal (``this'') and gestural to distinguish the box
+Jerry refers to.
+
+\begin{center}
+\begin{tabular}{l}
+\stmt{Jerry pointsAt carboardBox}\\
+\stmt{Jerry looksAt carboardBox}\\
+$\to$ \stmt{Jerry focusesAt carboardBox}\\
+\hspace{0.7cm}$\Rightarrow$ {\tt ?obj = [cardBoardBox]}
+\end{tabular}
+\end{center}
+
+Finally, the \textsc{Dialogs} queries the ontology about the content of the box
+and the question can be answered: ``Jido-E''. Note that the object's label is
+used instead of its ID. This way we enhance interaction using familiar names
+given by the users.
+
+\begin{center}
+\begin{tabular}{l}
+\stmt{?obj isIn cardBoardBox}\\
+\hspace{0.7cm}$\Rightarrow$ \concept{?obj = videoTape2}\\
+\end{tabular}
+\end{center}
+
+At this point Jerry wants to know where the other tape is, and that is exactly
+what he asks Jido: ``And where is the other tape?''. In this occasion, the
+\textsc{Dialogs} module is able to interpret that Jerry is not referring to the
+video which they were just talking about, but to the other one:
+
+\begin{center}
+\begin{tabular}{l}
+\stmt{?obj type VideoTape}\\
+\stmt{?obj differentFrom videoTape2}\\
+\hspace{0.7cm}$\Rightarrow$ \concept{?obj = [videoTape1]}
+\end{tabular}
+\end{center}
+
+Since there is only one possible ``other'' video (there are only two videos in
+the scene), it can directly answer Jerry: ``The other tape is on the table and
+next to the toolbox.''
+
+\begin{center}
+\begin{tabular}{l}
+\stmt{videoTape1 isOn table}\\
+\stmt{videoTape1 isNextTo toolbox}
+\end{tabular}
+\end{center}
+
+\subsection{\emph{Cleaning the table} scenario}
+
+This experiment involved a more complex decisional layer where the ORO
+server was used in conjunction with the HATP symbolic task planner and the
+SHARY execution controller (the complete software architecture we deployed for
+this experiment is pictured on Figure~\ref{fig|archi},
+page~\pageref{fig|archi}).
+
+In this scenario (Figure~\ref{fig|cleantable-video}), a human and a robot
+cooperate to remove objects from a table. The robot produces symbolic plans for
+both itself and the human (Figure~\ref{plan-etat2})
+that allow the robot to (verbally) share the task with the human (like ``I take
+the green box and I put it in the trashbin, you take the black video tape and
+you throw it in the trashbin''). Plans are created based on perceived
+visibility and reachability of the objects, and the robot also monitors the
+human activities to track the advancement of the whole plan (this last aspect
+is presented in~\cite{warnier2012when}).
+
+\begin{figure}
+    \centering
+    \includegraphics[width=0.7\columnwidth]{./figs/cleantable.jpg}
+
+    \caption{Snapshot of the film of the ``Clean the Table'' scenario. The
+    physical situation, the SPARK model, and the current step of the plan can
+    be seen.}
+
+    \label{fig|cleantable-video}
+\end{figure}
 
@@
   \includegraphics[width=1.0\textwidth]{./figs/plan3.pdf} \\
-  \caption {A plan produced by HATP for clean the table based on the initial state given in \S~\ref{sub:gtrc}}
+  \caption {A plan produced by HATP to execute the high-level order ``clean the table''.
+Two streams of actions are generated: for the human (top) and for the robot(bottom).
+Synchronization points ensure the coordination.}
   \label{plan-etat2}
@@
 
-Let us now take a simpler example to illustrate a full run of the
-system. We have only one tape on the table and it is is reachable only
-by the robot while the throw position on top of the trashbin is
-reachable only by the human.
+Figure~\ref{fig|cleantable-timeline} presents a excerpt of the whole task and
+shows how the different components produce and use symbolic knowledge (note
+that the plan is actually computed {\it a priori}. It appears nevertheless on
+this diagram for better readability).
+
+This sequence of events depicts a run with only one tape on the table. The tape is
+reachable by the robot only, while the human is the one who can reach the
+trashbin to throw objects.
 
@@
     \begin{center}
-    \caption {An example of a human-robot collaborative goal achievement}
+        \caption {Sequence of events of a simplified version of the \emph{Cleaning the table} scenario.}
     \end{center}
-  \label{manip_run_fg}
+  \label{fig|cleantable-timeline}
 \end{figure*}
 
-%\subsection{successful execution}
-
-Figure~\ref{manip_run_fg} illustrates the main processes occurring
-during a multi-step human-robot collaborative goal achievement.  The
-plan produced is quite straightforward and is shown in the third row
-called ``Goal and Plan''. It consists in 4 successive actions
-involving the robot and the human. Robot grasps the tape and then
-places it on the table at a position where it is visible and reachable
-for the human. Human then is asked to pick the tape and throw it in
-the trashbin. The first row, named ``Cameras'', shows several
-snapshots corresponding to various execution steps. Snapshot~1
-presents the initial situation. Snapshots~2, 3, 4 and 5 give the state
-after the successive achievement of the four actions in the plan. The
-second row, named ``3D Model'', shows the display of SPARK at the same
-instants. The fourth, row called ``Robot Speech Acts'', gives robot
-speech acts produced along the execution to inform the human partner
-about goal and plan creation and status and to verbalize the actions
-that the human is asked to execute. The fifth row illustrates robot
-knowledge on itself and on the objects. The sixth row illustrates the
-robot knowledge about the human state. The seventh row gives ongoing
-robot action with action preconditions and effects assessment as well
-as motion execution tasks. Motion trajectory typology can be found
-between the 3D Model views. The eighth row gives ongoing human action
-with action preconditions and effects assessment and monitoring
-activity.
+The plan produced in that case by HATP is straightforward and shown in the
+third row. It consists in four successive actions involving the robot and the
+human. Robot grasps the tape and then places it on the table at a position
+where it is visible and reachable for the human. Human then is asked to pick
+the tape and throw it in the trashbin.
 
@@
 have already been built based on these
-theories~\cite{Rich1997,Sidner2005,Tambe2005a,Breazeal2003} and they
+theories~\cite{Rich1997,Sidner2005,Breazeal2003} and they
 all have shown benefits of this approach\fxwarning{which approach? how it is related to larger AI challenges?}. They have also shown how

### 2013-09-04 paper.tex
Fix path for figures
Required for Elsevier submission
Removed Aurelie and Mathieu from authors

@@
 
-\author{Rachid Alami, Séverin Lemaignan, Aurélie Clodic, Mathieu Warnier}
+\author{Séverin Lemaignan, Rachid Alami}
 
@@
 \centering
-\includegraphics[width=0.9\columnwidth]{figs/grounding_robot.pdf}
+\includegraphics[width=0.9\columnwidth]{grounding_robot.pdf}
 \caption{Robot reasoning about HRI and anticipation of human activities:
@@
 	\centering
-    \subfigure[] { \includegraphics[width=0.2\textwidth]{figs/reachex-1.jpg} }
-    \subfigure[] {\includegraphics[width=0.2\textwidth]{figs/reachex-2.jpg} }
-    \subfigure[] {\includegraphics[width=0.2\textwidth]{figs/reachex-3.jpg} }
-    \subfigure[] {\includegraphics[width=0.2\textwidth]{figs/reachex-4.jpg} }
+    \subfigure[] { \includegraphics[width=0.2\textwidth]{reachex-1.jpg} }
+    \subfigure[] {\includegraphics[width=0.2\textwidth]{reachex-2.jpg} }
+    \subfigure[] {\includegraphics[width=0.2\textwidth]{reachex-3.jpg} }
+    \subfigure[] {\includegraphics[width=0.2\textwidth]{reachex-4.jpg} }
 	\caption{An example illustrating the \textit{reachable} relation.
@@
 %           \label{fig:wuweiPhoto}
-           \includegraphics[width=0.5\textwidth]{./figs/etat2-P1010769_brightened-v2.jpg}
+           \includegraphics[width=0.5\textwidth]{etat2-P1010769_brightened-v2.jpg}
        }%
@@
 %          \label{fig:sparkScreenshot}
-          \includegraphics[width=0.43\textwidth]{./figs/etat2_photo.png}
+          \includegraphics[width=0.43\textwidth]{etat2_photo.png}
        }\\ %  ------- End of the first row ----------------------%
@@
   \centering
-  \includegraphics[width=0.95\columnwidth]{./figs/first_plan.pdf}
+  \includegraphics[width=0.95\columnwidth]{first_plan.pdf}
   \caption{A plan produced by HATP with 2 streams}
@@
   \centering
-  \includegraphics[width=0.95\columnwidth]{./figs/second_plan.pdf}
+  \includegraphics[width=0.95\columnwidth]{second_plan.pdf}
   \caption{A plan with the wasted time social rule}
@@
   \centering
-  \includegraphics[width=0.9\linewidth]{./figs/pt.jpg}
+  \includegraphics[width=0.9\linewidth]{pt.jpg}
 \caption{Asking the robot to hand over a tape on the table.}
@@
   \centering
-  \includegraphics[width=0.9\linewidth]{./figs/inTheBox2.jpg}
+  \includegraphics[width=0.9\linewidth]{inTheBox2.jpg}
 \caption{Jerry asks Jido for the content of the box by pointing at it.}
@@
     \centering
-    \includegraphics[width=0.7\columnwidth]{./figs/cleantable.jpg}
+    \includegraphics[width=0.7\columnwidth]{cleantable.jpg}
 
@@
   \centering
-  \includegraphics[width=1.0\textwidth]{./figs/plan3.pdf} \\
+  \includegraphics[width=1.0\textwidth]{plan3.pdf} \\
   \caption {A plan produced by HATP to execute the high-level order ``clean the table''.
@@
   \centering
-    \includegraphics[width=1.0\textwidth]{./figs/manip_run.pdf} \\
+    \includegraphics[width=1.0\textwidth]{manip_run.pdf} \\
     \begin{center}

### 2013-09-04 paper.tex
Rephrasing, fixing fixmes...

@@
         \includegraphics[width=\columnwidth]{spark-oro}
-    \caption{Functional overview of knowledge base ({\sc Oro} server, top part) and the geometric situation assessment module (\emph{SPARK}, bottom part)}
+    \caption{Functional overview of knowledge base ({\sc Oro} server, top part) and the geometric situation assessment module (\emph{{\sc Spark}}, bottom part)}
         \label{fig|spark-oro}
@@
 their symbolic interpretation. We rely on a dedicated geometric and temporal
-reasoning module called SPARK (\emph{SPAtial Reasoning \&
+reasoning module called {\sc Spark} (\emph{SPAtial Reasoning \&
 Knowledge}~\cite{Sisbot2011}). It is a situation assessment reasoner that
@@
 
-SPARK is an \emph{amodal}~\cite{Mavridis2006} geometric model of the
+{\sc Spark} is an \emph{amodal}~\cite{Mavridis2006} geometric model of the
 environment that serves both as basis for the fusion of the perception
@@
 
-\fxnote{Add 2-3 lines to summarize the AI highlights of SPARK}
+\fxnote{Add 2-3 lines to summarize the AI highlights of {\sc Spark}}
 
@@
 
-SPARK currently provides a simple implementation of such a functionality. The
+{\sc Spark} currently provides a simple implementation of such a functionality. The
 only managed hypotheses are \emph{in container} and \emph{in agent hand}. We
@@
 high level human monitoring processes. We call this temporal and geometric
-reasoning \emph{primitive action recognition}.
-
-For example, a \emph{pick}, a \emph{throw} or a \emph{place} action
-can be recognized by observing that an object on table and an empty
-human hand are close to each other, or that the human hand holding an
-object is close to a container, etc. Human hand position is either
-directly perceived or inferred from its initial perceived trajectory\fxwarning{we really do that?}.
-We have a simple implementation of such a primitive action recognition
-in SPARK that relies on monitoring human hand and its motion near
-objects or above containers.
+reasoning \emph{primitive action recognition}. Those primitive actions are
+assessed through monitoring situations like ``an empty hand is close to an
+object on a table'' (precursor for a \emph{pick}), or ``a hand holding an
+object is over a container'' (precursor for a \emph{throw}).
 
@@
 
-SPARK enables \emph{perspective-taking}: spatial relations between entities can
+{\sc Spark} enables \emph{perspective-taking}: spatial relations between entities can
 be computed from different viewpoints, which let the robot build for each
@@
 HATP planner (for Human Aware Task Planner~\cite{Alili2008, Alili2009}).  HATP
-is based on a Hierarchical Task Network (HTN) refinement, which performs an
-iterative task decomposition into sub-tasks until reaching atomic
-actions\fxwarning{Rephrase: the definition is actually generic to HTN, not HATP
-specfic}. The \emph{planning domain} defines the set of methods describing how
-to decompose a task and represents the procedural knowledge of the robot. It is
-stored outside of the central knowledge base.
-
-HATP is able to produce plans for the robot's actions as well as for the other
-participants (humans or robots). It can be tuned by setting up different costs
-depending on the actions to apply and by taking into account a set of
-constraints called social rules. This tuning aims at adapting the robot's
-behavior according to the desired level of cooperation of the robot.
+is a refinement of a Hierarchical Task Network (HTN). As an HTN, it performs an
+iterative task decomposition into sub-tasks until reaching atomic actions. The
+\emph{planning domain} defines the set of methods describing how to decompose a
+task and represents the procedural knowledge of the robot. It is stored outside
+of the central knowledge base, using a specific formalism.
+
+HATP originality resides in its ability to produce plans for the robot's
+actions \emph{as well as for the other participants} (humans or robots): the
+robot plans for itself, but also for its interactors, which effectively endows the robot
+with new interaction capabilities: It can partially predict the behaviour, and
+thus monitor the engagement, of its human partner; the planner generates
+synchronisation points (arrow going from the robot stram to the human stream on
+Figure~\ref{plan_hatp1}) between the agents; the controller can also verbalize
+the plans to explain its partner how a task may be shared~\cite{warnier2012when}.
+
+HATP can be tuned by setting up different costs depending on the actions to
+apply and by taking into account a set of constraints called \emph{social
+rules}. This tuning aims at adapting the robot's behavior according to the
+desired level of cooperation of the robot.
 
@@
 integration between the robot controller and the geometric and
-temporal reasoning functions in SPARK and symbolic facts and beliefs
+temporal reasoning functions in {\sc Spark} and symbolic facts and beliefs
 management in ORO is central.
@@
 Robot must reacts to change in the world not linked to robot action to drive
-world update. Robot controller uses SPARK to monitor human hand motion and
+world update. Robot controller uses {\sc Spark} to monitor human hand motion and
 primitive action recognition for \emph{pick} and \emph{throw}. As mentioned
@@
     \caption{Snapshot of the film of the ``Clean the Table'' scenario. The
-    physical situation, the SPARK model, and the current step of the plan can
+    physical situation, the {\sc Spark} model, and the current step of the plan can
     be seen.}
@@
 This last section goes over again the main contributions we have presented
-so far: why are they relevant to robotics and artificial intelligence? How do
-they compare to similar approaches? How do we justify our technical choices and
-what remain to be developed/explored in the future?
+so far: why they are relevant to robotics and artificial intelligence, how do
+they compare to similar approaches, how do we support our technical choices and
+what remains to be developed/explored in the future.
 
@@
 algorithms themselves are ignorant of the input sources, and would work equally
-well with a full object recognition stack, it is not currently supported.
-
-\fxwarning{spatial reasoning -> mention we do not support extensive temporal reasoning}
+well with a full object recognition stack, we have not tackled it yet.
+
+Besides, temporal reasoning (essential for accurate action recognition for
+instance) is not properly addressed in our stack. Temporal reasoning is used
+only locally, and does not allow for tracking of long or global events.
 
@@
 the robot.
+
+\fxfatal{Insert here discussion of other major KRR systems for robots (Severin)}
 
@@
 
-Altogether, the components we have presented compose an architecture that we
-call \emph{knowledge-oriented}:
+Altogether, the components we have presented compose a
+\emph{knowledge-oriented} architecture:
 
@@
 
-Those items are not new \emph{per se}. It seems however interesting to
+Those items are not new \emph{per se}. It is however interesting to
 underline the shift of focus this implies during the design and integration
-phases: components of our deliberative layer are defined and glued together by
+phases of robots: components of our deliberative layer are defined and glued together by
 the knowledge they produce and consume.
 
-\fxfatal{Insert here discussion of other major KRR systems for robots (Severin)}
+It reveals how the robotic community is ``catching up'' with semantic
+technologies. Human-robot interaction, because it supposes operations in
+environments with complex semantics, drives this evolution.
 
@@
 
-Robotics is traditionally regarded as the prototypical testbed for \emph{embodied}
+Robotics is traditionally regarded as the prototypical instance of \emph{embodied}
 artificial intelligence, and this dimension is especially prevalent in

### 2013-09-05 paper.tex
Rephrasing, rewrite...

@@
 
-This article shed light on some key decisional issues that are to be tackled
-for a cognitive robot to share space and tasks with a human. We adopt a
-constructive approach based on the identification and the effective
-implementation of individual and collaborative skills. These cognitive
-abilities cover geometric reasoning and situation assessment based essentially
-on perspective-taking and affordances, management and exploitation of each
-agent (human and robot) knowledge in separate cognitive models, natural
-multi-modal communication, human-aware task planning, and human and robot
-interleaved plan achievement.
+This article sheds light on some key decisional issues that are to be tackled
+for a cognitive robot to share space and tasks with a human, and present our
+take on these challenges. We adopt a constructive approach based on the
+identification and the effective implementation of individual and collaborative
+skills. These cognitive abilities cover geometric reasoning and situation
+assessment mainly based on perspective-taking and affordances, management and
+exploitation of each agent (human and robot) knowledge in separate cognitive
+models, natural multi-modal communication, human-aware task planning, and human
+and robot interleaved plan achievement.
 
@@
 indeed lays at the crossroad of several other domains of AI and requires to
-tackle them in a holistic manner. Modeling humans and human cognition;
+tackle them in a holistic manner: Modeling humans and human cognition;
 acquiring, representing, manipulating in a tractable way abstract knowledge;
-reasoning on this knowledge to make decisions; instanciating those decisions
-into physical actions in coordination with humans... Many AI techniques are
-invited, from visual processing to symbolic reasoning, from task planing to
-\emph{theory of mind} building, from reactive control to action recognition and
-learning.
+reasoning on this knowledge to make decisions; and eventually instanciating
+those decisions into physical actions in coordination with humans. Many AI
+techniques are invited, from visual processing to symbolic reasoning, from task
+planing to \emph{theory of mind} building, from reactive control to action
+recognition and learning.
 
@@
 them into a coherent challenge for artifical intelligence, and also explicit
-some of the trails that have been explored on our robots and that result in a
-deliberative, knowledge-focused, architecture for human-robot interaction.
-
-Human-robot interaction requires to equip the robot with explicit reasoning on
-the human and on its own capacities to achieve its tasks in a legible, if not
-collaborative, way with human partners. In the following pages, we present a
-robot control system which has been especially designed for a cognitive robot
-which shares space and task with a human. It has been designed by adopting a
-constructive approach based on effective individual and collaborative skills.
-The system is comprehensive since it aims\fxnote{it *is* comprehensive, or it *aims* at being comprehensive?} at dealing with a complete set of
-abilities articulated so that the robot controller is effectively able to
-conduct in a flexible manner a collaborative task with a human partner.
-
-We illustrate below how we deal with a typical human-robot interactive task
-achievement and what are the abilities we claim are necessary. Section
-\ref{sec:problem} proposes a typical human-robot interaction problem that can
-be solved by the proposed robot controller. Section
-\ref{sec:Framework} provides an overview of the robot controller and
-introduces three activities which are described in \ref{sec:situ},
-\ref{sec:plan} and \ref{sec:action}. Section \ref{sec:expes}
-presents the structure an actual run on a real robot in face to face interaction with a
-person. Finally, section~\ref{sec:soa} discusses and analyses the impact of
-each of our contribution in the light of existing literature and alternative
-approaches.
+some of the paths that have been investigated on our robots and that result in a
+deliberative, knowledge-oriented, architecture for human-robot interaction.
 
@@
 
-\fxfatal{Develop/reorganize this section to organize the different problems of
-HRI into one coherent challenge for AI. See how to articulate it with section
-\ref{sec:psychocognitivegrounds}}
+\begin{figure}[htb]
+\centering
+\includegraphics[width=0.9\columnwidth]{grounding_robot.pdf}
+\caption{Robot reasoning about HRI and anticipation of human activities.
+  Sources of information are multi-modal dialogue, and monitoring of the
+  environment and human activity. The robot must adapt on-line its behaviours
+  by merging computed plans with reactive control.}
+\label{fig:hri-dec}
+\end{figure}
+
+\fxfatal{Merge the following 3 paragraphs}
+Our frame is the robot which is
+expected to achieve interactive object manipulation, fetch and carry tasks and
+similar tasks in a domestic environment (Figure~\ref{fig:hri-dec}).
 
@@
 
-\begin{figure}[htb]
-\centering
-\includegraphics[width=0.9\columnwidth]{grounding_robot.pdf}
-\caption{Robot reasoning about HRI and anticipation of human activities:
-  sources of information are multi-modal dialogue, and observation of
-  environment and human activity}
-\label{fig:hri-dec}
-\end{figure}
-
-\paragraph{The Joint Action Challenge} Let us consider a robot which is
-supposed to achieve interactive object manipulation, fetch and carry tasks and
-similar tasks in a domestic environment. The problem we are dealing with here
+Over the last years we have focused our efforts on identifying the cognitive
+prerequisites of these challenges, and giving them experimental reality on the
+robots: what is required for sentences like ``Let's set the table together''
+to be understood by the robot, correctly interpreted in the spatial and
+temporal context of the interaction, and eventually transformed into a set of
+actions.
+
+The problem we are dealing with here
 is the following. Given:
@@
   environment of the robot and of the human,
+\item [\it a priori} common-sense knowledge,
 \end {itemize}
@@
 
-We have devised a decisional framework for human-robot interactive
-task achievement that is aimed to allow the robot not only to
-accomplish its tasks but also to produce behaviors that support its
-engagement vis-a-vis its human partner and to interpret human
-behaviors and intentions. 
-Together and in coherence with this framework, we have developed
-and experimented various task planners and interaction schemes that
-allow the robot to select and perform its tasks while taking into
-account explicitly the human abilities as well as the constraints
-imposed by the presence of humans, their needs and preferences.
-
-\subsection{Our Contributions Towards the Cognitive Robot}
+\subsection{Our contributions towards the cognitive robot}
 
@@
 
-Over the last years we have focused our efforts on identifying the cognitive
-prerequisites of these challenges, and giving them experimental reality on the
-robots: what is required for sentences like ``Let's set the table together''
-to be understood by the robot, correctly interpreted in the spatial and
-temporal context of the interaction, and eventually transformed into a set of
-actions.
-
-We have chosen to tackle the challenge from several ends: human-aware
-navigation and motion planning~\cite{Mainprice2011}, situation assessment
-coupled with motion planning~\cite{Mainprice2012}, projection of
-``mightabilities'' that anticipates what surrounding agents may
-do~\cite{Pandey2011}, and the design and deployment of a pervasive
-\emph{knowledge-oriented} software architecture~\cite{Lemaignan2013}.
-
-\section{From Human-Robot Interaction to a Deliberative Architecture}
-
-\subsection{The Psycho-Cognitive Grounds of Human-Robot Interaction}
-\label{sec:psychocognitivegrounds}
-
-\fxfatal{write section summary of the challenges that human-robot interaction presents to robotics.}
+\subsection{Article organization}
+
+The remaining of the article discuss the robotic architecture we have built to
+tackle the AI challenges of human-robot interaction, and how its relates to
+other approaches. We propose to organize this discussion in four sections.
+
+The next section presents the overall architecture and the knowledge model we
+have developed for our robots. Section~\ref{sec:impl} gives some insights on
+the each of the main components of the architecture, and try to highlight their
+significance for artificial intelligence. Section~\ref{sec:expe} presents two
+experiments that illustrate in a concrete way what can be currently achieved
+with our robots.  Section~\ref{sect|conclusion} finally summarizes our main
+contribution and relate them to other recent researches in the field.
+
+\section{From human-robot interaction to a deliberative architecture}
 
@@
 
-Fig.~\ref{fig|archi} gives an overview of the connections between the
+\fxfatal{Describe archi}
+
+Figure~\ref{fig|archi} gives an overview of the connections between the
 deliberative components of our architecture~\cite{Alami2011}.
@@
 
-This architecture moves away from standard layered approaches. Interactions
+This architecture moves away from standard layered approaches found in
+robotics~\cite{Gat1998three, Volpe2001CLARAty, Goldberg2002}. Interactions
 between components at the deliberative level are mostly bidirectional and we do
@@
 have lower-level modules to execute actions or manage sensors, but all
-cognition-related modules reside at the same level.}. This is nicely
-illustrated by the dialogue input processing. This component does not simply
-act as an alternative perceptual input to the symbolic database; it also
+cognition-related modules reside at the same level.}. Dialogue processing, for
+instance, illustrates this structure: This component does not simply act as an
+alternative perceptual input to the symbolic database; it also
 actively queries previously acquired knowledge to disambiguate and validate the
@@
 This interaction between knowledge and actions is also central to our approach
-(as for any cognitive system): it is one of the activities of the
-robot, actually shared between communication components (that can acquire
-desires from interaction with agents, amongst other things) and an execution
-controller that may decide to take an incoming desire into account to create
-its own internal goals. The controller generates and manages intentions from
-these goals with the help of a symbolic task planner, that also has direct
-access to the knowledge base.
+(as for any cognitive system): it is one amongst all the activities carried by
+the robot, actually performed by the communication components (that acquire
+desires from interaction with agents) and the execution controller that may
+decide to take an incoming desire into account to create its own internal
+goals. The controller generates and manages intentions from these goals with
+the help of a symbolic task planner, that also has direct access to the
+knowledge base.
 
@@
 performative dialogue that can possibly change the internal state of the robot,
-but does not lead to the creation of desires, like question answering or
-statement assertion), various background monitoring and recognition tasks, etc.
+but does not lead to the creation of desires, like statement assertion or
+question answering), various background monitoring and recognition tasks, etc.
 
@@
 
-In our architecture (Fig.~\ref{fig|archi}), knowledge manipulation relies on a
-\emph{semantic blackboard}: a central server (the {\sc Oro}
+In this architecture, knowledge manipulation relies on a
+semantic \emph{blackboard}: a central server (the {\sc Oro}
 server~\cite{Lemaignan2010}) stores knowledge as it is produced by each of the
-deliberative components. It conversely exposes a {\tt json}-based RPC API to
-query the knowledge base~\cite{lemaignan2012kbapi}.
+deliberative components. It conversely exposes a {\tt json}-based RPC
+API~\cite{lemaignan2012kbapi} to query the knowledge base.
 
@@
 At any time, the knowledge available to the robot may come from three sources:
-Part of the knowledge is stored in an ontology (the {\sc OpenRobots Ontology}),
+{\it a priori} knowledge is stored in an ontology (the {\sc OpenRobots Ontology}),
 loaded when the robot initializes. This static source covers the
@@
 perception, interaction, decision making.  We go over the processes it involves
-in the next section. Lastly, the third source of knowledge is the output of the
+in the next sections. Lastly, the third source of knowledge is the output of the
 reasoner.
@@
 taxonomy. This upper part of the ontology is pictured on
-figure~\ref{fig|upper_tbox}. All the classes visible on this figure belong to the
+Figure~\ref{fig|upper_tbox}. All the classes visible on this figure belong to the
 {\sc OpenCyc} namespace (the {\tt cyc:} prefix is omitted for clarity).
 
-By \emph{aligning} the upper part of the ontology on {\sc OpenCyc} (as other
-KRS, like {\sc KnowRob}~\cite{Tenorth2009a} or PEIS K\&R~\cite{Daoutis2009}, did)
+Aligning the upper part of the ontology on {\sc OpenCyc} (as done by other
+KRS, like {\sc KnowRob}~\cite{Tenorth2009a} or PEIS K\&R~\cite{Daoutis2009})
 has multiple advantages. First the design of this part of the ontology is
@@
 
-This figure~\ref{fig|upper_tbox} also illustrates the fundamental disjunction
+Figure~\ref{fig|upper_tbox} also illustrates the fundamental disjunction
 in the ORO model between \emph{temporal} and \emph{spatial} entities (formally,
@@
 
-The tree of figure~\ref{fig|upper_tbox} (this subset of the ontology is indeed
+The tree in Figure~\ref{fig|upper_tbox} (this subset of the ontology is indeed
 a tree: this has however not to be the case in general, and, as a matter of
@@
 relevant to the context of service robotics in an human-like environment are
-present.
+present. For performances and ease of use, we do not attempt to comprehensively
+cover the field. We opportunistically extend the common-sense knowledge when
+required by experiments.
 
@@
 The {\sc Oro} server also features a mechanism to mimic minimalistic forms of
-biological memory.  When new statements are inserted in the knowledge base, a
+memory families.  When new statements are inserted in the knowledge base, a
 \emph{memory profile} is optionally attached to them.
@@
 Three such profiles are predefined: {\tt short term}, {\tt episodic} and {\tt
-long term}. They each correspond to a different lifetime for the statements
+long term}. They are currently attached to different lifetime for the statements
 (respectively 10 seconds, 5 minutes and no time limit). After this duration,
@@
 
-This approach is limited. In particular, \emph{episodic} memory primarily
-refers to the semantics of the statements (that is expected to be related to an
+This approach is limited. In particular, \emph{episodic} memory should primarily
+refer to the semantics of the statements (that is expected to be related to an
 event) and not to a specific life duration.
 
-We rely however on this short term memory for a particular use-case:
-\emph{active concepts}. Some modules, like the natural language processor, use
-the {\tt short term} memory profile to mark for a few seconds important
-concepts that are currently manipulated by the robot. For example, if a human
-asks the robot: ``Give me all red objects'', the human, the \concept{Give}
-action, and each red objects that are found are successively marked as
-\emph{active concepts} by inserting statements such as \stmt{human type
-ActiveConcept} in the short-term memory (which can be considered, in this case,
-to be a working memory). We use this feature to trace certain knowledge-related
-processes.
+We rely however on this mechanism in certain cases: for instance, some modules
+like the natural language processor use the {\tt short term} memory profile to
+mark for a few seconds important concepts that are currently manipulated by the
+robot as \emph{active concepts}. For example, if a human asks the robot: ``Give
+me all red objects'', the human, the \concept{Give} action, and each red
+objects that are found are successively marked as \emph{active concepts} by
+inserting statements such as \stmt{human type ActiveConcept} in the short-term
+memory (which can be considered, in this case, to be a working memory). We use
+this feature to trace certain knowledge-related processes.
 
@@
 \section{Implementing the Cognitive Skills}
+\label{sec:impl}
 
@@
         \includegraphics[width=\columnwidth]{spark-oro}
-    \caption{Functional overview of knowledge base ({\sc Oro} server, top part) and the geometric situation assessment module (\emph{{\sc Spark}}, bottom part)}
+    \caption{Functional overview of knowledge base ({\sc Oro} server, top part)
+    and the geometric situation assessment module (\emph{{\sc Spark}}, bottom
+    part). {\sc Spark} computes symbolic relationships between objects and
+    agents, and fed them to the knowledge base.}
+
         \label{fig|spark-oro}
@@
 generates symbolic knowledge from the geometry of the environment with respect
-to relations between objects, robots and humans (Fig.~\ref{fig|spark-oro}),
+to relations between objects, robots and humans (Figure~\ref{fig|spark-oro}),
 also taking into account the different perspective that each agent has on the
@@
 \section{Experimental outcomes}
+\label{sec:expe}
 
@@
 
+\fxfatal{Rewrite conclusion}
+
+We have devised a decisional framework for human-robot interactive
+task achievement that is aimed to allow the robot not only to
+accomplish its tasks but also to produce behaviors that support its
+engagement vis-a-vis its human partner and to interpret human
+behaviors and intentions. 
+Together and in coherence with this framework, we have developed
+and experimented various task planners and interaction schemes that
+allow the robot to select and perform its tasks while taking into
+account explicitly the human abilities as well as the constraints
+imposed by the presence of humans, their needs and preferences.
+
 In this paper we have presented a decisional framework designed for

### 2013-09-05 paper.tex
Many rewrite, fix, typos,...

@@
 
-Human-Robot interaction is an area filled with challenges for artifical
+Human-Robot interaction is an area filled with challenges for artificial
 intelligence: dynamic, partially unknown environments that are not originally
@@
 acquiring, representing, manipulating in a tractable way abstract knowledge;
-reasoning on this knowledge to make decisions; and eventually instanciating
+reasoning on this knowledge to make decisions; and eventually instantiating
 those decisions into physical actions in coordination with humans. Many AI
@@
 We do not tackle all these issues. This article attempts however to organize
-them into a coherent challenge for artifical intelligence, and also explicit
+them into a coherent challenge for artificial intelligence, and also explicit
 some of the paths that have been investigated on our robots and that result in a
@@
   environment of the robot and of the human,
-\item [\it a priori} common-sense knowledge,
+\item {\it a priori} common-sense knowledge,
 \end {itemize}
@@
 
-Previous section provided an overview of what we propose as a generic cognitive
-architecture for robots involved in interactions with other social agents.
-
-We now introduce each of the components and try to give insights on their main
+Previous section provided an overview of our cognitive architecture for robots
+involved in interactions with other social agents, as well as the knowledge
+model we rely on.
+
+We now introduce each of the components and give insights on their main
 features.
@@
         \centering
-        \includegraphics[width=\columnwidth]{spark-oro}
+        \includegraphics[width=\columnwidth]{spark-oro.png}
     \caption{Functional overview of knowledge base ({\sc Oro} server, top part)
@@
     part). {\sc Spark} computes symbolic relationships between objects and
-    agents, and fed them to the knowledge base.}
+    agents, and feeds them to the knowledge base.}
 
@@
 generates symbolic knowledge from the geometry of the environment with respect
-to relations between objects, robots and humans (Figure~\ref{fig|spark-oro}),
+to relations between objects, robots and humans (Figures~\ref{fig|spark-oro} and~\ref{fig:sparkSubfigures}),
 also taking into account the different perspective that each agent has on the
@@
 continuously updated at run-time by the robot based on its sensors (in our
-experiments, objects are identified and localised through 2D barcodes, while
+experiments, objects are identified and localised through 2D fiducial markers, while
 humans are tracked with Kinect-like devices, optionally assisted by motion
@@
 
-\fxnote{Add 2-3 lines to summarize the AI highlights of {\sc Spark}}
-
-\subsubsection{Situation Assessment}
-
-\paragraph{Symbolic facts production}
-
-Geometric state of the world is abstracted in symbolic facts that can
-be classified in three different categories.
-
-\begin {itemize}
-\item Relative positions of object and agents, e.g.  \stmt{GREY\_TAPE
-    isOn TABLE}.
-
-\item Perception and manipulation capacity and state of agents,
-  e.g. \stmt{ROBOT looksAt GREY\_TAPE}, \stmt{GREY\_TAPE isVisibleBy
-    HUMAN1}.
-
-\item Motion status for object or agent parts, e.g.  \stmt{GREY\_TAPE
-    isMoving true},\\ \stmt{ROBOT\_HEAD isTurning true}.
-\end {itemize}
-
-Reasoning about human perspective allow to compute facts such as:
-\stmt{GREY\_TAPE isBehind HUMAN1}, \stmt{GREY\_TAPE leftOf
-  HUMAN1}.
-
-Figure~\ref{fig::reach-ex} illustrates different situations for the
-\textit{reachable} relation. The robot and its human partner are placed face to
-face (Figure~\ref{fig::reach-ex}.1).  The robot first estimates if the small
-grey box is reachable to itself. This is done by finding a collision free
-posture to reach the object (Figure~\ref{fig::reach-ex}.2). Next the robot
-switches to the human's perspective to estimate if the same object is reachable
-to the human as well.  In the last scene, the human moves towards his left,
-farther from the object (Figure~\ref{fig::reach-ex}.4). The situation is then
-reevaluated. In this occasion though, the reasoner cannot find a satisfactory
-posture for the human to reach the box because he is too far from the target.
-
-\begin{figure*}[!t]
-	\centering
-    \subfigure[] { \includegraphics[width=0.2\textwidth]{reachex-1.jpg} }
-    \subfigure[] {\includegraphics[width=0.2\textwidth]{reachex-2.jpg} }
-    \subfigure[] {\includegraphics[width=0.2\textwidth]{reachex-3.jpg} }
-    \subfigure[] {\includegraphics[width=0.2\textwidth]{reachex-4.jpg} }
-	\caption{An example illustrating the \textit{reachable} relation.
-The relation is computed from the perspectives of both the robot and the human.
-The computed posture at each step is illustrated with a global view of the scene (top), and from a closest view (bottom).}
-\label{fig::reach-ex}
-\end{figure*}
+From an artificial intelligence perspective, {\sc Spark} takes care of merging
+sensing modalities, it computes and grounds perceptions into symbolic
+knowledge (situation assessment), it is responsible temporal interpretation of
+situations and performs spatial perspective-taking.
 
@@
        }%
-       \subfigure[3d model view of initial state]{%
+       \subfigure[Corresponding 3D model view]{%
 %          \label{fig:sparkScreenshot}
@@
 
-   \caption{%
-     In this situation, there are three tapes on the table. Two tapes
-     are only reachable by the robot: the LOTR\_TAPE (black in the 3d
-     model) and GREY\_TAPE. The third tape WALLE\_TAPE (white in the
-     3d model) and the trashbin PINK\_TRASHBIN are only reachable by
-     the human HUMAN1. All tapes are on the table TABLE.  }%
+   \caption{% In this situation, there are three tapes on the table. Two tapes
+       are only reachable by the robot: the \concept{LOTR\_TAPE} (black in the 3D
+       model) and \concept{GREY\_TAPE}. The third tape \concept{WALLE\_TAPE} (white
+       in the 3D model) and the trashbin \concept{PINK\_TRASHBIN} are only
+       reachable by the human \concept{HUMAN1}. All tapes are on the table
+   \concept{TABLE}. The corresponding symbolic facts are listed in Table~\ref{table|beliefsfig7}.  }%
+
  \label{fig:sparkSubfigures}
@@
 
-The set of facts computed in the situation depicted by
-Figure~\ref{fig:sparkSubfigures} is the following:
-\begin{footnotesize}
-%\begin{small}
-\begin{verbatim}
-                ROBOT                          HUMAN1
-PINK_TRASHBIN isReachable false    PINK_TRASHBIN isReachable true 
-WALLE_TAPE isReachable false       WALLE_TAPE isVisible true 
-LOTR_TAPE isReachable true         LOTR_TAPE isReachable false 
-GREY_TAPE isReachable true         GREY_TAPE isReachable false
-WALLE_TAPE isVisible true          WALLE_TAPE isReachable true 
-LOTR_TAPE isVisible true           LOTR_TAPE isVisible true 
-GREY_TAPE isVisible true           GREY_TAPE isVisible true 
-WALLE_TAPE isOn TABLE              WALLE_TAPE isOn TABLE 
-LOTR_TAPE isOn TABLE               LOTR_TAPE isOn TABLE 
-GREY_TAPE isOn TABLE               GREY_TAPE isOn TABLE 
-\end{verbatim}
-%\end{small}
-\end{footnotesize}
+\begin{table}
+\begin{center}
+\begin{tabular}{l}
+\hline
+Robot's beliefs about itself (\emph{robot's model}):\\
+\hline
+  \hspace{0.7cm}\stmt{PINK\_TRASHBIN isReachable false}\\
+  \hspace{0.7cm}\stmt{WALLE\_TAPE isReachable {\bf false}}\\
+  \hspace{0.7cm}\stmt{LOTR\_TAPE isReachable {\bf true}}\\
+  \hspace{0.7cm}\stmt{GREY\_TAPE isReachable {\bf true}}\\
+  \hspace{0.7cm}\stmt{WALLE\_TAPE isVisible true}\\
+  \hspace{0.7cm}\stmt{LOTR\_TAPE isVisible true}\\
+  \hspace{0.7cm}\stmt{GREY\_TAPE isVisible true}\\
+  \hspace{0.7cm}\stmt{WALLE\_TAPE isOn TABLE}\\
+  \hspace{0.7cm}\stmt{LOTR\_TAPE isOn TABLE}\\
+  \hspace{0.7cm}\stmt{GREY\_TAPE isOn TABLE}\\
+\hline
+\hline
+Robot's beliefs about the human (\emph{human's model}):\\
+\hline
+  \hspace{0.7cm}\stmt{PINK\_TRASHBIN isReachable false}\\
+  \hspace{0.7cm}\stmt{WALLE\_TAPE isReachable {\bf true}}\\
+      \hspace{0.7cm}\stmt{LOTR\_TAPE isReachable {\bf false}}\\
+      \hspace{0.7cm}\stmt{GREY\_TAPE isReachable {\bf false}}\\
+  \hspace{0.7cm}\stmt{WALLE\_TAPE isVisible true}\\
+  \hspace{0.7cm}\stmt{LOTR\_TAPE isVisible true}\\
+  \hspace{0.7cm}\stmt{GREY\_TAPE isVisible true}\\
+  \hspace{0.7cm}\stmt{WALLE\_TAPE isOn TABLE}\\
+  \hspace{0.7cm}\stmt{LOTR\_TAPE isOn TABLE}\\
+  \hspace{0.7cm}\stmt{GREY\_TAPE isOn TABLE}\\ 
+ \hline
+\end{tabular}
+\end{center}
+\caption{Symbolic facts computed from the situation depicted in
+Figure~\ref{fig:sparkSubfigures}. Note how reachability differs for the two
+agents.}
+
+\label{table|beliefsfig7}
+\end{table}
+
+\subsubsection{Situation Assessment}
+
+\paragraph{Symbolic facts production}
+
+Geometric state of the world is abstracted in symbolic facts that can
+be classified in three different categories.
+
+\begin {itemize}
+\item Relative positions of object and agents, e.g.  \stmt{GREY\_TAPE
+    isOn TABLE}.
+
+\item Perception and manipulation capacity and state of agents,
+  e.g. \stmt{ROBOT looksAt GREY\_TAPE}, \stmt{GREY\_TAPE isVisibleBy
+    HUMAN1}.
+
+\item Motion status for object or agent parts, e.g.  \stmt{GREY\_TAPE
+    isMoving true},\\ \stmt{ROBOT\_HEAD isTurning true}.
+\end {itemize}
+
+Reasoning about human perspective allow to compute facts such as:
+\stmt{GREY\_TAPE isBehind HUMAN1}, \stmt{GREY\_TAPE leftOf
+  HUMAN1}.
+
+Figure~\ref{fig::reach-ex} illustrates different situations for the
+\textit{reachable} relation.
+
+\begin{figure*}[!t]
+	\centering
+    \subfigure[] { \includegraphics[width=0.2\textwidth]{reachex-1.png} }
+    \subfigure[] {\includegraphics[width=0.2\textwidth]{reachex-2.jpg} }
+    \subfigure[] {\includegraphics[width=0.2\textwidth]{reachex-3.jpg} }
+    \subfigure[] {\includegraphics[width=0.2\textwidth]{reachex-4.jpg} }
+    \caption{An example illustrating the \textit{reachable} relation. The
+        relation is computed from the perspectives of both the robot and the
+        human. The computed posture at each step is illustrated with a global
+        view of the scene (top), and from a closest view (bottom). The robot
+        and its human partner are placed face to face (a).  The robot first
+        estimates if the small grey box is reachable to itself. This is done by
+        finding a collision free posture to reach the object (b). Next the
+        robot switches to the human's perspective to estimate if the same
+        object is reachable to the human as well (c).  In the last scene, the
+        human moves towards his left, farther from the object (d). The
+        situation is then reevaluated. In this occasion though, the reasoner
+        cannot find a satisfactory posture for the human to reach the box
+        because he is too far from the target.  }
+
+\label{fig::reach-ex}
+\end{figure*}
 
@@
 object in certain states. This happens, for instance, when the object
-has been put in a container, when it is in the robot gripper or in the
-human hand, and more generally in any state in which it is hidden by
-something else. Our robot has a model of the possible symbolic states
-for an object (whether the object is on a furniture, in an agent hand,
-in a container, etc.).  According to the robot perception of what has
-happened since the object was last seen, the robot tries to maintain a
-belief of the current possible symbolic states and their associated
-probabilities for this object. Such information can be used to update
-the beliefs using input from exploration, dialog, human visual
-focus,\ldots \fxnote{Paragraph need write. Tell what the module actually does, not what it could do}
-
-{\sc Spark} currently provides a simple implementation of such a functionality. The
-only managed hypotheses are \emph{in container} and \emph{in agent hand}. We
-can have only one hypothesis at the same time. Hypothesis validity is checked
-geometrically in case of incoming perception values.
+is hidden in a container, when the robot holds it in its gripper, and more
+generally in any state in which it is hidden by something else. {\sc Spark} models the possible symbolic states of objects (whether the object is on a furniture, in an agent hand,
+in a container, etc.): According to the robot perception of what has
+happened since the object was last seen, the robot builds a
+belief of the current possible state with associated probability for the.
 
@@
 
-{\sc Spark} enables \emph{perspective-taking}: spatial relations between entities can
+{\sc Spark} also supports \emph{perspective-taking}: spatial relations between entities can
 be computed from different viewpoints, which let the robot build for each
@@
 of the environment. These models are separately stored in the knowledge base.
-\fxwarning{Rephrase: already mentioned before}
 
@@
 \emph{allo-centric} frames (addressee-centred).
-
-\subsubsection{Symbolic facts and beliefs management}
-\fxwarning{Do we need this section here?}
-
-The facts produced by the geometric and temporal reasoning component
-are stored in the central symbolic knowledge base.
-
-ORO stores independent knowledge models (in our implementation, as
-\emph{ontologies}) for each agent (the robot and the humans it
-interacts with). The robot architecture components (like the executive
-layer or the situation assessment component) can then store the
-agents' beliefs in specific models.  Each of these models is
-independent and logically consistent, enabling reasoning on different
-perspectives of the world that would otherwise be considered as
-globally inconsistent (for instance, an object can be visible for the
-robot but not for the human. This object can have at the same time the
-property \concept{isVisible \textbf{true}} and \concept{isVisible
-  \textbf{false}} in two different models). This feature actually
-allows us to consider the robot to be endowed with a simple
-\emph{theory of mind}~\cite{Scassellati2002}: it can explicitly
-model the belief state of its interactors.
 
@@
 atoms of each sentence are resolved with the help of the ontology to ground
-concepts like objects (\ie when a user says ``pick the can'', resolve to which
+concepts like objects (\ie when a user says ``pick the can'', it resolves to which
 instance of \emph{can} the user is referring to) and actions. Sentences are
-sorted into questions, desires and statements, and processed accordingly.\fxwarning{Severin: Develop this section}
+sorted into questions, desires and statements, and processed accordingly.
+Figure~\ref{dialogs|ex} gives a example of the processing of a simple order.
+
+\begin{figure}
+    \centering
+	\begin{tabular}{l|l}
+	\emph{Initial human knowledge} &
+	\emph{Human input}\\
+
+	\hline
+
+    	\stmt{book\_1 type Book} &
+	``Give me the book.'' \\
+
+    	\stmt{book\_1 hasColor blue} & \\
+	\vspace{0.5em}\\
+	\hline
+
+	\emph{Generated partial statements} &
+	\emph{Newly created statements}\\
+	\hline
+    	\stmt{?obj type Book} & 
+	\stmt{human desires sit\_1} \\
+
+	\hspace{0.2cm}$\Rightarrow$ \concept{?obj = book\_1}
+    	& \stmt{sit\_1 type Give} \\
+    	& \stmt{sit\_1 performedBy myself} \\
+    	& \stmt{sit\_1 actsOnObject book\_01} \\
+    	& \stmt{sit\_1 receivedBy human} \\
+	\end{tabular}
+
+    \caption{Processing an order: this simple example (no ambiguity), taken
+    from~\cite{Lemaignan2011a}, shows the thematic roles computed by {\sc
+Dialogs} for the order ``Give me the book''.}
+
+\label{dialogs|ex}
+\end{figure}
 
@@
 structures (for instance, a sentence like ``learn that cats are animals'' is
-converted into \stmt{Cat subClassOf Animal} and added to th eknowledge pool),
+converted into \stmt{Cat subClassOf Animal} and added to the knowledge base),
 interprets common temporal and place adverbs (like \emph{above} or
@@
 
-\fxwarning{Give more implementation details. Why did we go for a custom NLP? (Severin)}
-
-\subsubsection{Limits of disambiguation at semantic level}
-
-One prototypical example of semantic disambiguation has been given in
-\cite{Ros2010b} with the child game \emph{spygame}: two players are facing
-each other with a set of random objects in-between, one player mentally choose
-one object, and the other player has to guess the object by asking closed
-questions like \emph{Is your object small or large?} Based on the knowledge it
-has acquired, the robot is able to minimize the number of questions required to
-find the object.
-
-When playing this kind of game, however, the issue arises that the robot has no
-way to select which knowledge about the object is relevant in the interaction
-context. For instance, the knowledge base may store facts like \stmt{obj1 type
-ActiveConcept} (which internally means that this concept was mentioned in a
-discussion in the last few seconds): this information is not a relevant
-property of \concept{obj1} when trying to disambiguate concepts with humans.
-This distinction between \emph{internal knowledge} (meaningful to
-the system only) and \emph{common knowledge} (whose meaning is understood by
-all the interactors) has not been properly dealt with in our architecture.
-
-Besides, even knowledge that belongs to the \emph{common knowledge} may not be
-appropriate in a given interaction context. For instance, the system may
-compute that at a given instant the human is looking at the object: \stmt{human
-looksAt obj1}. This property makes sense to both parties, but in the context of
-the \emph{spygame}, we would like to mainly use immanent properties, not
-volatile like a gaze. More research is required to identify relevant
-interaction contexts and knowledge classes attached to them.
-
 \subsubsection{Multi-modal communication}
@@
 based on gaze, or \emph{implicit} like postures) are presented in a homogeneous
-way. The dialogue grounding process makes use of them at two distinct levels.
+way, as statements in the knowledge base. The dialogue grounding process makes
+use of them at two distinct levels to provide seamless multi-modal interaction.
 
@@
 thus monitor the engagement, of its human partner; the planner generates
-synchronisation points (arrow going from the robot stram to the human stream on
+synchronisation points (arrow going from the robot stream to the human stream on
 Figure~\ref{plan_hatp1}) between the agents; the controller can also verbalize
@@
 are extracted from the knowledge base, and either passed to a task planner
-(presented in the next section) or executed if the procedure is directly
+(presented in the previous section) or executed if the procedure is directly
 available.
@@
 questions (when asking a question, we consider the human to be in an
-\emph{interrogative state}). When the knowledge base recognizes that an agent
+\emph{interrogative state}). When the knowledge base states that an agent
 \emph{experiences} a particular emotion or state, the execution controller may
 decide to handle it, typically by trying to answer the question or using the
-emotional or physical state as a parameter for subsequent actions. As an example,
-when the speaker says ``I feel tired'', we change the motion planner
-parametrization to lower the effort the human needs to provide for the following
-joint manipulation tasks. Note that this example has been implemented as a
-proof-of-concept. We have not yet tried to define a theoretical framework that
-would support action alteration based on the user's experienced states.
+emotional or physical state as a parameter for subsequent actions. As an
+example, when the speaker says ``I feel tired'', we change the motion planner
+parametrization to lower the effort the human needs to provide for the
+following joint manipulation tasks. Note that this example has been implemented
+as a proof-of-concept. A broader framework that would support action alteration
+based on the user's experienced states is yet to be devised.
 
@@
 The {\sc Oro} server proposes two paradigms to access its content: RPC-style
-queries (based on the standard SPARQL language) or events. In its simplest
-form, a module can subscribe to an event by passing through an event pattern
-(in its basic form, a partial statement like \stmt{? type Elephant}) and a
-callback.  Each time a new instance of elephant appears in the knowledge base,
-the callback is triggered.
+queries (based on the standard SPARQL language) or events. A module can
+subscribe to an event by passing through an event pattern (in its simplest
+form, a partial statement like \stmt{? type Elephant}) and a callback.  Each
+time a new instance of elephant appears in the knowledge base, the callback is
+triggered.
 
@@
 Our other controller, {\sc Shary}, relies on the symbolic task planner to
-generate at runtime suitable tasks.
+generate at runtime suitable sets of atomic actions.
 
@@
 postures, abilities and preferences. \cite{Sisbot2008, Mainprice2011,
-Pandey2010} detail these techniques.
+Pandey2011} detail these techniques.
 
 \paragraph{Action Execution and Monitoring Robot Controller}
-context and on the shared plan elaborated by HATP for a given goal,
+Based on context and on the shared plan elaborated by HATP for a given goal,
 the robot controller decides to execute an action or to ask its human
@@
 
-For plan action monitoring, primitive actions recognition defined
-above are used. A primitive action detection is interpreted as action
-success if it is the expected one and failure otherwise. The robot
-also reacts to the absence of activity.
-
-\paragraph{Context Management Policies} \fxfatal{Rewrite this section to level up AI aspects (Mathieu)}
-Building, updating and maintaining a correct state of the world at
-geometric and symbolic level is crucial to the capacity of the robot
-to carry on successfully a multi-step interaction with a human. Tight
-integration between the robot controller and the geometric and
-temporal reasoning functions in {\sc Spark} and symbolic facts and beliefs
-management in ORO is central.
-
-The robot controller has access to the symbolic facts in ORO that are
-automatically updated whenever object and agent positions are changed.
-Robot controller can also access geometric perceived or inferred
-positions of objects and geometric positions and postures of the human
-that will be used to orient its cameras.  Building and updating the
-state of the world first relies on perceiving objects. Robot
-controller can use:
-
-\begin {itemize}
-\item Exploration policies: robot will exhaustively scan the table to
-  see all what can be seen.
-
-\item Search policies: robot will search an object until it is
-  detected if possible, scanning all the table and looking in human
-  hand.
-
-\end {itemize}
-
-Robot reasons on possible positions for non perceived objects. These hypotheses
-could be updated\fxwarning{are they?} using new input from dialogue, human action and focus of
-attention. Currently, we manage at most one hypothesis per object. This
-hypothesis is produced by robot controller through an inference on robot or
-human action. In case of perception conflicts with low probability for the
-current hypothesis, robot controller will break this hypothesis and delete
-corresponding symbolic fact in the ontology.
-
-Robot must reacts to change in the world not linked to robot action to drive
-world update. Robot controller uses {\sc Spark} to monitor human hand motion and
-primitive action recognition for \emph{pick} and \emph{throw}. As mentioned
-above these primitive action recognition should be used with higher level
-information on goals, intentions and plans and some exploration to achieve
-complex human action and activity monitoring.  In the current implementation,
-these primitive actions are over-optimistically interpreted as the
-corresponding actions \emph{pick} object and \emph{throw} object.
+For plan action monitoring, primitive actions recognition is used. A primitive
+action detection is interpreted as action success if it is the expected one and
+failure otherwise. The robot also reacts to the absence of activity.
 
@@
 task planner besides ORO and {\sc Dialogs}. In this scenario, the human and the
-robot try to cooperatively remove objects from a table. In this scenario,
-behaviours and motions are fully planned.
+robot try to cooperatively remove objects from a table. Behaviours and motions
+are fully planned.
 
@@
 move things here and there. To assess the reasoning abilities of the robot they
-ask Jido for information (entered through keyboard). Ideally, the robot should
-also perform actions when required (e.g. hand an object when asking ``give
-me...''). However, since it is out of the scope of this work, we do not include
-any motion from the robot's side.
+ask Jido for information. Ideally, the robot should
+also perform actions when required (\eg hand an object when asking ``give
+me...'').
 
@@
 detected through motion capture. The robot knowledge base is pre-loaded with
-the \emph{ORO Commonsense Ontology}.  We next describe in detail two
+the \emph{ORO Commonsense Ontology}.  We next describe two
 situations where we can follow the internal robot's reasoning and the
@@
 
-Tom enters the room while carrying a big box (Figure~\ref{fig|vpt}, page 1). He
-approaches the table and asks Jido to handle him the video tape: ``Jido, can
+Tom enters the room while carrying a big box (Figure~\ref{fig|vpt}). He
+approaches the table and asks Jido to hand him the video tape: ``Jido, can
 you give me the video tape''. The \textsc{Dialogs} module queries the ontology to
@@
 partner it infers that Tom is referring to the video tape on the table, and not
-the one inside the box which is not visible from his view. Therefore,
-non-visible objects are removed obtaining: \concept{?obj =[videoTape1]}.
-
-Since only one object is available, the robot infers
+the one inside the box which is not visible for him.
+
+Since only one object remains, the robot infers
 that the human refers to it and would eventually execute the command, \ie give
@@
 
-In this situation, Jerry enters the
-living room without knowing where Tom had placed the video tapes. So he first
-asks Jido: ``What's in the box?''. Before the robot can answer the question it
-has to figure out which box Jerry is talking about. Similar to the previous
-situation, there are two available boxes:
+In this situation, Jerry enters the living room without knowing where Tom had
+placed the video tapes (Figure~\ref{fig|pointing}). So he first asks Jido:
+``What's in the box?''. Before the robot can answer the question it has to
+figure out which box Jerry is talking about. Similar to the previous situation,
+there are two available boxes:
 
@@
 applied. The only option is to ask Jerry which box he is referring to: ``Which
-box, the toolbox or the cardboard box?'' Jerry could now simply answer the
+box, the toolbox or the cardboard box?'' Jerry could answer the
 question. Instead, he decides to point at it while indicating: ``This box''
@@
 \textbf{focusesOn}(?ag, ?obj)}) The \textsc{Dialogs} module is then able to merge both
-sources of information, verbal (``this'') and gestural to distinguish the box
+sources of information, verbal (``this'') and deictic to distinguish the box
 Jerry refers to.
@@
 At this point Jerry wants to know where the other tape is, and that is exactly
-what he asks Jido: ``And where is the other tape?''. In this occasion, the
+what he asks Jido: ``And where is the other tape?''. In this situation, the
 \textsc{Dialogs} module is able to interpret that Jerry is not referring to the
@@
 
-This experiment involved a more complex decisional layer where the ORO
-server was used in conjunction with the HATP symbolic task planner and the
-SHARY execution controller (the complete software architecture we deployed for
-this experiment is pictured on Figure~\ref{fig|archi},
+This experiment involved a more complex decisional layer where the ORO server
+was used in conjunction with the HATP symbolic task planner and the SHARY
+execution controller (the complete software architecture we deployed for this
+experiment is the one pictured on Figure~\ref{fig|archi},
 page~\pageref{fig|archi}).
@@
 visibility and reachability of the objects, and the robot also monitors the
-human activities to track the advancement of the whole plan (this last aspect
-is presented in~\cite{warnier2012when}).
+human activities to track the advancement of the whole plan.
 
@@
 
-Figure~\ref{fig|cleantable-timeline} presents a excerpt of the whole task and
-shows how the different components produce and use symbolic knowledge (note
-that the plan is actually computed {\it a priori}. It appears nevertheless on
-this diagram for better readability).
+Figure~\ref{fig|cleantable-timeline} presents a sample of the whole task and
+shows how the different components produce and use symbolic knowledge.
 
@@
     \includegraphics[width=1.0\textwidth]{manip_run.pdf} \\
-    \begin{center}
         \caption {Sequence of events of a simplified version of the \emph{Cleaning the table} scenario.}
-    \end{center}
   \label{fig|cleantable-timeline}
@@
 representation and reasoning}, \emph{task planning} or \emph{high-level
-control} are mandatory cognitive foundations that we first have to mention.
+control} are mandatory cognitive foundations to build upon.
 
@@
 knowledge bases, dialogue). We call \emph{physical situation assessment} the
-cognitive skill a robot exhibits when it assesses the nature and content of its
+cognitive skill that a robot exhibits when it assesses the nature and content of its
 surroundings and efficiently monitors its evolution.
@@
 grounded amodal representations~\cite{Mavridis2006}, semantic maps
-(Figure~\ref{fig|semanticmap}, \cite{Nuechter2008, Galindo2008,Blodow2011}) or
+(\cite{Nuechter2008, Galindo2008,Blodow2011}) or
 affordance-based planing and object classification~\cite{Lorken2008,
@@
 (vision-based fiducial markers tracking, RGBD human tracking, motion capture)
-are merged and where amodal spatial reasoning takes place (explanation
-generation for unexpected perceptions like a disappearing object).
+are merged and where amodal spatial reasoning takes place (for instance,
+explanation generation for unexpected perceptions like a disappearing object).
 
@@
 recognition ~\cite{Kollar2010,Matuszek2010} or language
-grounding~\cite{Tellex2010}.~\cite{Skubic2004} presented a spatial reasoner
+grounding~\cite{Tellex2010}.~\cite{Skubic2004} present a spatial reasoner
 integrated in a robot which computes symbolic positions of objects.
@@
 Internet had to be translated into plans and executed in a kitchen environment,
-perceived and rebuilt on-line by the robots. While Prolog offers more flexible
-modeling (no constraints on the arity of predicate, where Description Logics as
-used in our work are limited to binary predicates), it is based on the closed
-world assumption (if something cannot be inferred to be true, it is inferred to
-be false) whereas we rely on the open world assumption, which is more realistic
-in real world scenarios. A probabilistic extension of \textsc{KnowRob}, called
-\textsc{ProbCog}~\cite{Jain2009} is also available. While in principle
-possible, currently the CRAM architecture does not provide explicit support for
-interacting with humans.
+perceived and rebuilt on-line by the robots.
 
@@
 
-Relying on RDF triples and Description Logics has decisive advantages such as
-the availability of numerous mature open-source libraries to manipulate the
-ontology, interoperability with several major on-line knowledge bases (besides
-{\sc OpenCyc}, {\sc WordNet}, {\sc DBPedia} or {\sc
-RoboEarth}~\cite{Waibel2011} are other example), open-world reasoning, and the
-formal guarantee of decidability (it is always possible to classify a
-Description Logics ontology).
-
-It also has notable limitations, both fundamental (the suitability
-of Description Logics when reasoning on --typically non-monotonic-- commonsense
-knowledge is questionable) and practical: RDF triples imply only binary predicates
-(\stmt{subject predicate object}), which constrains the expressiveness of the
-system or leads to cumbersome reifications. Alternatives exist (like {\sc
-KnowRob}~\cite{Tenorth2009a}) that mix RDF with more expressive logic
-languages like {\sc Prolog}, at the price, however, of other limitations, like
-closed-world reasoning or immutable T-Box. The classification
-performance is another issue: from our experience, with an ontology sized for a
-standard experiment (about 100 classes and 200 instances), classification
-typically takes about 100ms, which becomes problematic during interactions.
-Besides, the performances are difficult to predict, since a seemingly
-inoffensive new statement may indirectly change radically the logical
-complexity of the whole knowledge model and lead to notable degradation of
-classification time.
+Relying on RDF triples and Description Logics has advantages such as good
+understanding of its trades-off, thanks to being widespread in the semantic Web
+community, the availability of mature libraries to manipulate the ontology,
+interoperability with several major on-line knowledge bases ({\sc
+OpenCyc}, {\sc WordNet}, {\sc DBPedia} or {\sc RoboEarth}~\cite{Waibel2011} are
+some examples), open-world reasoning, and the formal guarantee of decidability
+(it is always possible to classify a Description Logics ontology).
+
+It also has notable limitations, both fundamental (the suitability of
+Description Logics when reasoning on --typically non-monotonic-- commonsense
+knowledge is questionable) and practical: RDF triples imply only binary
+predicates, which constrains the expressiveness of the system or leads to
+cumbersome reifications. Alternatives exist (like {\sc
+KnowRob}~\cite{Tenorth2009a}) that mix RDF with more expressive logic languages
+like {\sc Prolog}, at the price, however, of other limitations, like
+closed-world reasoning or immutable T-Box. The classification performance is
+another issue: from our experience, with an ontology sized for a standard
+experiment (about 100 classes and 200 instances), classification typically
+takes about 100ms, which becomes problematic during interactions.  Besides, the
+performances are difficult to predict, since a seemingly inoffensive new
+statement may indirectly change abruptly the logical complexity of the whole
+knowledge model and lead to notable degradation of classification time.
 
@@
 phenomena (like time) or uncertain phenomena. When required (for instance for
-action recognition), these are managed inside the corresponding components, and
-are not exposed at the semantic level.
+action recognition), these are managed by dedicated components (like {\sc
+Spark}), and are not exposed at the semantic level.
 
@@
 and already well accepted in the robotic community~\cite{Tenorth2009a,
-Lim2011}. While alternatives like \emph{Answer Set Programming} have also been
+Lim2011}. Alternatives like \emph{Answer Set Programming} have also been
 successfully investigated in robotics~\cite{Chen2010,Erdem2012}, in particular
-to deal with non-monotonic reasoning, we did not actually hit any brick wall
+to deal with non-monotonic reasoning. However we did not actually hit any brick wall
 while working with OWL ontologies. We may reconsider this choice at a later
@@
 the world in terms of visual perception, spatial descriptions, affordances and
-beliefs, etc.  Therefore, in the last years these notions have been gradually
-employed in Human-Robot Interaction.~\cite{Breazeal2006} presents a learning
+beliefs, etc.  In the last years these notions have been gradually
+employed in HRI.~\cite{Breazeal2006} presents a learning
 algorithm that takes into account information about a teacher's visual
@@
 perspective taking for action recognition between two
-robots.~\cite{Trafton2005} use both visual and spatial perspective taking for
-finding out the referent indicated by a human partner.
+robots.~\cite{Trafton2005} use both visual and spatial perspective taking to
+find out the referent indicated by a human partner.
 
@@
 
-\paragraph{Multi-modal communication}
+\paragraph{Limits of disambiguation at semantic level}
+
+\fxfatal{Introduce the paragraph}
+
+One prototypical example of semantic disambiguation has been given in
+\cite{Ros2010b} with the child game \emph{spygame}: two players are facing
+each other with a set of random objects in-between, one player mentally choose
+one object, and the other player has to guess the object by asking closed
+questions like \emph{Is your object small or large?} Based on the knowledge it
+has acquired, the robot is able to minimize the number of questions required to
+find the object.
+
+When playing this kind of game, however, the issue arises that the robot has no
+way to select which knowledge about the object is relevant in the interaction
+context. For instance, the knowledge base may store facts like \stmt{obj1 type
+ActiveConcept} (which internally means that this concept was mentioned in a
+discussion in the last few seconds): this information is not a relevant
+property of \concept{obj1} when trying to disambiguate concepts with humans.
+This distinction between \emph{internal knowledge} (meaningful to
+the system only) and \emph{common knowledge} (whose meaning is understood by
+all the interactors) has not been properly dealt with in our architecture.
+
+Besides, even knowledge that belongs to the \emph{common knowledge} may not be
+appropriate in a given interaction context. For instance, the system may
+compute that at a given instant the human is looking at the object: \stmt{human
+looksAt obj1}. This property makes sense to both parties, but in the context of
+the \emph{spygame}, we would like to mainly use immanent properties, not
+volatile like a gaze. More research is required to identify relevant
+interaction contexts and knowledge classes attached to them.
 
@@
 
-\section*{Acknowledgment}
-
-This work has been supported by EU FP7 ``SAPHARI'' under grant agreement no. ICT-287513.
+\section*{Acknowledgments}
+
+Building such a robotic architecture is the work of many hands, and we would
+like to acknowledge here the worthwhile contributions of Aurélie Clodic, Mamoun
+Gharbi, Julien Guitton, Matthieu Herrb, Jim Mainprice, Amit Kumar Pandey and
+Mathieu Warnier.
+
+This work has been supported by EU FP7 ``SAPHARI'' under grant agreement no.
+ICT-287513.
 

### 2013-09-05 paper.tex
Tweaking and more tweaking

@@
 
-\documentclass[preprint,5p]{elsarticle}
+%\documentclass[preprint,5p]{elsarticle}
+\documentclass[final,3p,times]{elsarticle}
 
@@
 
-\usepackage[draft, nomargin, marginclue, footnote]{fixme}
+\usepackage[draft,nomargin, marginclue, footnote]{fixme}
 
@@
 
-\listoffixmes
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-\subsection{The human-robot interaction challenge}\label{sec:problem}
+We scope our work to the robot which is
+expected to achieve interactive object manipulation, fetch and carry tasks and
+similar tasks in a domestic environment (Figure~\ref{fig:hri-dec}).
+
+Two agents (a human and a robot) share a common space and exchange information
+through multiple modalities. The robot is expected to explicitly take into
+account the human and to manage its interactions with him.
+
+Namely, the robot must be able to recognize, understand and participate
+to communication situations, both explicit (\eg the human addresses verbally
+the robot) and implicit (\eg the human points to an object); the robot must be
+able to take part to joint actions, both proactively (by planning and proposing
+such an action to the human) and reactively; the robot must be able to move and
+act in a safe, efficient and legible way, taking into account social rules like
+proxemics.
 
@@
 
-\fxfatal{Merge the following 3 paragraphs}
-Our frame is the robot which is
-expected to achieve interactive object manipulation, fetch and carry tasks and
-similar tasks in a domestic environment (Figure~\ref{fig:hri-dec}).
-
-We envision HRI in a context where two agents (a human and a robot)
-share a common space and exchange information through various
-modalities. Our aim is to endow the robot with an explicit
-consideration of the human and with the ability to manage its
-interactions with him (Figure~\ref{fig:hri-dec}). This must be
-considered at the architecture level as well as at the task/motion
-planning and execution level.
-
-Over the last years we have focused our efforts on identifying the cognitive
-prerequisites of these challenges, and giving them experimental reality on the
-robots: what is required for sentences like ``Let's set the table together''
-to be understood by the robot, correctly interpreted in the spatial and
-temporal context of the interaction, and eventually transformed into a set of
-actions.
-
-The problem we are dealing with here
-is the following. Given:
+Those three items have broad implications regarding the required underlying
+cognitive skills.  A \emph{joint action}, for instance, can be described as
+follow. Given:
 
@@
 \item a joint goal, which has been previously established and agreed
-  upon (through a process which is out of the scope of this paper),
+  upon (typically through dialogue),
 \item the current situation, acquired through perception or
@@
 To do so, the robot has to be equipped with a number of decisional, planning
-and interpretation abilities where its human partner is taken
-explicitly into account. It needs to be able:
-
-\begin {itemize}
+and interpretation abilities where its human partner is taken explicitly into
+account. It needs to be able:
+
+\begin{itemize}
 \item to build and maintain relevant robot and human beliefs
@@
 those achieved by its human partner.
-\end {itemize}
+\end{itemize}
 
@@
 
-\subsection{Our contributions towards the cognitive robot}
-
-The main scientific contributions of the article are summarized here, to guide
-the reader. We develop and discuss them, also in relation to other works, at
-the end of the article.
-
-\fxfatal{Rewrite section 'summary of contributions'}
+We will see, in the following sections, how cognitive skills for realistic
+human-robot interaction can be implemented and organized together in a coherent
+architecture.
 
@@
 with our robots.  Section~\ref{sect|conclusion} finally summarizes our main
-contribution and relate them to other recent researches in the field.
+contributions and relate them to other recent researches in the field.
 
@@
 
-\fxfatal{Need to carefully introduce this section: how did we come up this this architecture?}
-
-\fxfatal{Describe archi}
+\fxwarning{Need to carefully introduce this section: how did we come up this this architecture? Describe archi}
+
+Over the last years, we have been slowly ramping up explicit knowledge
+representation and manipulation in the deliberative and executive layers of our
+robots. Ranging from situation assessment to symbolic task planning, from
+verbal interaction to event-driven execution control, we have built up a
+\emph{knowledge-oriented} architecture which is now used on a daily basis on
+our robots.
 
@@
 
-\fxwarning{Relate HATP social rules to other AI techniques}
-
 \paragraph{Several levels of cooperation} 
@@
 
-\fxfatal{Insert here discussion of other major KRR systems for robots (Severin)}
-
 \subsection{Interacting with humans}
@@
 
-\fxfatal{Introduce the paragraph}
+\fxwarning{Introduce the paragraph}
 
@@
 theories~\cite{Rich1997,Sidner2005,Breazeal2003} and they
-all have shown benefits of this approach\fxwarning{which approach? how it is related to larger AI challenges?}. They have also shown how
+all have shown benefits of this approach. They have also shown how
 difficult it is to manage turn-taking between communication partners
 and to interleave task realization and communication in a generic
-way\fxwarning{is it smthg we improve one?}. Finally, today only few
+way. Finally, today only few
 systems~\cite{Fong2006,Breazeal2003,Sisbot2008} take humans into
-account at all levels\fxwarning{Old references, no new ones? we should also say what is new/better with our approach (Rachid)}.
+account at all levels.
 
@@
 
-\fxfatal{Rewrite conclusion}
+\fxwarning{Rewrite conclusion}
 

### 2013-09-06 paper.tex
Cleaning headers (remove unused packages)

@@
 %\documentclass[preprint,5p]{elsarticle}
-\documentclass[final,3p,times]{elsarticle}
+\documentclass[preprint,3p,times]{elsarticle}
 
@@
 \usepackage{graphicx, subfigure}
-\usepackage{url}
-\graphicspath{{figs/}}
-\DeclareGraphicsExtensions{.pdf,.jpg,.png}
 
@@
 
-\usepackage{listings}
-\usepackage{alltt}
-\usepackage{fancyvrb}
-
-%\usepackage[bookmarks]{hyperref}
-
-\usepackage{booktabs}
-%\usepackage{tabularx}
-
 \newcommand{\concept}[1]{{\small \texttt{#1}}}
-
 \newcommand{\stmt}[1]{{\footnotesize \tt $\langle$ #1\relax$\rangle$}}
@@
 \newcommand{\al}{{\textit{et al.\ }}}
+
+\graphicspath{{figs/}}
+\DeclareGraphicsExtensions{.pdf,.jpg,.png}
 
@@
 time triples are added or removed from the knowledge base, a Description
-Logics reasoner ({\sc Pellet}\footnote{\url{http://clarkparsia.com/pellet/}})
+Logics reasoner ({\sc Pellet}\footnote{\tt http://clarkparsia.com/pellet/})
 classifies the whole ontology and inserts all possible inferred triples.

### 2013-09-06 paper.tex
Fixed my terrible compilation issue
Was only a question of correct ordering of 'frontmatter' and 'document'. My bad.
Modify coauthors + acknowledgments according to Rachid's wishes

@@
 \usepackage[english]{babel}
-\usepackage{cite}
 \usepackage{graphicx, subfigure}
@@
 
+\begin{document}
 \begin{frontmatter}
@@
 
-\author{Séverin Lemaignan, Rachid Alami}
+\author{Séverin Lemaignan, Mathieu Warnier, E. Akin Sisbot, Rachid Alami}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{document}
 \begin{abstract}
@@
 Building such a robotic architecture is the work of many hands, and we would
-like to acknowledge here the worthwhile contributions of Aurélie Clodic, Mamoun
-Gharbi, Julien Guitton, Matthieu Herrb, Jim Mainprice, Amit Kumar Pandey and
-Mathieu Warnier.
+like to acknowledge here the worthwhile contributions of Samir Alili, Aurélie
+Clodic, Raquel Ros Espinoza, Mamoun Gharbi, Julien Guitton, Matthieu Herrb, Jim
+Mainprice and Amit Kumar Pandey.
 

### 2013-09-06 paper.tex
Updated style of the 3rd HATP plan
Merged Rachid's comments

@@
 techniques are invited, from visual processing to symbolic reasoning, from task
-planing to \emph{theory of mind} building, from reactive control to action
+planning to \emph{theory of mind} building, from reactive control to action
 recognition and learning.
@@
 We do not tackle all these issues. This article attempts however to organize
-them into a coherent challenge for artificial intelligence, and also explicit
-some of the paths that have been investigated on our robots and that result in a
-deliberative, knowledge-oriented, architecture for human-robot interaction.
-
-We scope our work to the robot which is
-expected to achieve interactive object manipulation, fetch and carry tasks and
-similar tasks in a domestic environment (Figure~\ref{fig:hri-dec}).
-
-Two agents (a human and a robot) share a common space and exchange information
+them into a coherent challenge for artificial intelligence, and also to make
+explicit some of the paths that have been investigated on our robots and that
+result in a deliberative, knowledge-oriented, architecture for human-robot
+interaction.
+
+The class of interactions we consider are essentially multi-modal situated
+dialog and collaborative task achievement: a robot which is expected to achieve
+interactive object manipulation, fetch and carry tasks and similar tasks in a
+domestic environment, with human partners (Figure~\ref{fig:hri-dec}). Two
+agents (a human and a robot) share a common space and exchange information
 through multiple modalities. The robot is expected to explicitly take into
@@
 \centering
-\includegraphics[width=0.9\columnwidth]{grounding_robot.pdf}
+\includegraphics[width=0.6\columnwidth]{grounding_robot.pdf}
 \caption{Robot reasoning about HRI and anticipation of human activities.
@@
   environment and human activity. The robot must adapt on-line its behaviours
-  by merging computed plans with reactive control.}
+  by merging computed plans with reactive control. The robot explicitly reasons
+  on the fact that it is (or not) observed by the human. Reasoning and planning
+  take place at symbolic as well as geometric level and systematically take
+  into account both agents beliefs and activities (from the point of view of
+  the robot).}
+
 \label{fig:hri-dec}
@@
 
-The remaining of the article discuss the robotic architecture we have built to
+The remaining of the article discusses the robotic architecture we have built to
 tackle the AI challenges of human-robot interaction, and how its relates to
@@
 have developed for our robots. Section~\ref{sec:impl} gives some insights on
-the each of the main components of the architecture, and try to highlight their
+each of the main components of the architecture, and try to highlight their
 significance for artificial intelligence. Section~\ref{sec:expe} presents two
@@
         \centering
-        \includegraphics[width=0.8\textwidth]{archi}
+        \includegraphics[width=0.9\textwidth]{archi}
         \caption{Overview of the LAAS deliberative layer. Knowledge is
@@
 performative dialogue that can possibly change the internal state of the robot,
-but does not lead to the creation of desires, like statement assertion or
+but does not lead directly to the creation of desires, like statement assertion or
 question answering), various background monitoring and recognition tasks, etc.
@@
         \centering
-        \includegraphics[width=\columnwidth]{spark-oro.png}
+        \includegraphics[width=0.8\columnwidth]{spark-oro.png}
     \caption{Functional overview of knowledge base ({\sc Oro} server, top part)
@@
 
-   \caption{% In this situation, there are three tapes on the table. Two tapes
-       are only reachable by the robot: the \concept{LOTR\_TAPE} (black in the 3D
-       model) and \concept{GREY\_TAPE}. The third tape \concept{WALLE\_TAPE} (white
-       in the 3D model) and the trashbin \concept{PINK\_TRASHBIN} are only
-       reachable by the human \concept{HUMAN1}. All tapes are on the table
-   \concept{TABLE}. The corresponding symbolic facts are listed in Table~\ref{table|beliefsfig7}.  }%
+   \caption{ The scenario involves a several videotapes that have to be
+       manipulated. Other objects are used as supports and containers.  After
+       identification and localisation of the set of objects and acquisition of
+       the position and posture of the human partner, the robot is able to compute
+       that two tapes only are reachable by itself: the \concept{LOTR\_TAPE}
+       (black in the 3D model) and \concept{GREY\_TAPE}. The third tape
+       \concept{WALLE\_TAPE} (white in the 3D model) and the trashbin
+       \concept{PINK\_TRASHBIN} are only reachable by the human. The
+       corresponding symbolic facts are listed in Table~\ref{table|beliefsfig7}.
+   }%
 
@@
     \caption{Processing an order: this simple example (no ambiguity), taken
-    from~\cite{Lemaignan2011a}, shows the thematic roles computed by {\sc
-Dialogs} for the order ``Give me the book''.}
+    from~\cite{Lemaignan2011a}, shows the thematic roles computed by 
+    {\sc Dialogs} for the order ``Give me the book''.}
 
@@
   \includegraphics[width=0.95\columnwidth]{second_plan.pdf}
-  \caption{A plan with the wasted time social rule}
+  \caption{A plan with the wasted time social rule. {\tt PUTRV} stands for {\it Put it so it is both Reachable and Visible}.}
   \label{plan_hatp2}
@@
   \centering
-  \includegraphics[width=1.0\textwidth]{plan3.pdf} \\
+  \includegraphics[width=1.0\textwidth]{third_plan.pdf} \\
   \caption {A plan produced by HATP to execute the high-level order ``clean the table''.
@@
 (\cite{Nuechter2008, Galindo2008,Blodow2011}) or
-affordance-based planing and object classification~\cite{Lorken2008,
+affordance-based planning and object classification~\cite{Lorken2008,
 Varadarajan2011}.
@@
 \fxwarning{Mention other source of knowledge about humans -- including social rules for instance}
+
+\paragraph{Planning for cooperation} HRI offers new challenges to planning
+by adding a complementary constraint: the synthesis of plans that are not only
+feasible but legible (in terms of their intention) and compliant with human
+preferences. This constraint is often ill-defined, dynamic, context-dependent.
+
+Indeed, we posit that human-robot collaborative task achievement calls for more
+subtlety for a robot than autonomous independent task achievement. HATP, by
+co-planning for the robot and the human, and modeling social rules as planning
+parameters, is a first step in this direction.
 
@@
 
+We are planning to extend the scheme by taking into accoount situation where
+divergent belief may appear between the human and the robot. Preliminary
+results have been pusblished in~\cite{warnier2012when}.
+
+There is also extensive work to be done in order to refine the notion of "good
+shared plan" and "good/acceptable robt behaviour" in this context. There are
+obviously large avenues for learning and adaptation in this context.
+
 \section*{Acknowledgments}

### 2013-09-06 paper.tex
Submitted version?

@@
 
-\usepackage[draft,nomargin, marginclue, footnote]{fixme}
+\usepackage[nomargin, marginclue, footnote]{fixme}
+
+\usepackage{hyperref}
 
@@
   Sources of information are multi-modal dialogue, and monitoring of the
-  environment and human activity. The robot must adapt on-line its behaviours
+  environment and human activity. The robot must adapt on-line its behaviors
   by merging computed plans with reactive control. The robot explicitly reasons
@@
 
-\fxwarning{Need to carefully introduce this section: how did we come up this this architecture? Describe archi}
-
-Over the last years, we have been slowly ramping up explicit knowledge
-representation and manipulation in the deliberative and executive layers of our
-robots. Ranging from situation assessment to symbolic task planning, from
-verbal interaction to event-driven execution control, we have built up a
-\emph{knowledge-oriented} architecture which is now used on a daily basis on
-our robots.
-
-Figure~\ref{fig|archi} gives an overview of the connections between the
-deliberative components of our architecture~\cite{Alami2011}.
+Connecting multiple independent software modules in one coherent robotic
+architecture is not only a technical challenge, but also a design and
+architectural challenge.
+
+In particular, dealing with the rich semantics of natural interaction with
+humans has been one of our major focus. To this end, we have been deploying
+over the last years explicit knowledge representation and manipulation in the
+deliberative layer of the robots as a {\it lingua franca} between our
+components.
+
+Figure~\ref{fig|archi} gives an overview of resulting
+architecture~\cite{Alami2011}. An active knowledge base, conveniently thought
+as a \emph{semantic blackboard}, connects most of the modules together: the
+\emph{geometric reasoning} module (called {\sc Spark}) produces symbolic
+assertions describing the dynamics of the robot's environment. These logical
+statements are stored in the knowledge base, and reused by the language
+processing module (called {\sc Dialogs}), the symbolic task planner (HATP) and
+the execution controllers. The output of the language processing module, or the
+activities started by the robot controller are likewise stored as symbolic
+statements.
+
+Incoming statements may in turn trigger events in the system, which
+allow reactive behaviors. These mechanisms are illustrated in the coming
+sections.
 
@@
         \includegraphics[width=0.9\textwidth]{archi}
-        \caption{Overview of the LAAS deliberative layer. Knowledge is
+        \caption{Overview of the deliberative layer. Knowledge is
         centrally managed in an active \emph{semantic blackboard}, pictured
@@
 reasoning in a \textit{top-down} way to produce new facts that may in return
-trigger physical behaviours.
+trigger physical behaviors.
 
@@
 robot plans for itself, but also for its interactors, which effectively endows the robot
-with new interaction capabilities: It can partially predict the behaviour, and
+with new interaction capabilities: It can partially predict the behavior, and
 thus monitor the engagement, of its human partner; the planner generates
@@
 ?action, ?action type Look, ?action hasGoal myself}, we could trigger a
-behaviour when the human expresses (through dialogue, gestures...) that he
+behavior when the human expresses (through dialogue, gestures...) that he
 wants to look at the robot itself.
@@
 task planner besides ORO and {\sc Dialogs}. In this scenario, the human and the
-robot try to cooperatively remove objects from a table. Behaviours and motions
+robot try to cooperatively remove objects from a table. behaviors and motions
 are fully planned.
@@
 
-\fxwarning{Robot control related work -> what is the originality of SHARY? why is it relevant from an AI point of view? (Mathieu, Aurelie)}
-
-Beetz et al.~\cite{Beetz2010} proposes a cognitive architecture called
-\textsc{CRAM} (Cognitive Robot Abstract Machine) that integrates
+Cognitive control architectures for robotics is a research field in its own right.
+Our originality of our contribution relates to the interleaving of a
+semantic-aware architecture designed for autonomy with human-aware decisional
+components. To our knowledge, no other robotic architecture covers in depth
+those two perspectives.
+
+The Ke Jia project~\cite{Chen2010} is maybe one of the closest approach. It
+integrates on a mobile platform a knowledge representation language with
+natural language processing, task planning and motion planning. Default and
+non-monotonic reasoning has been especially researched within the Ke Jia
+project for symbolic task planing~\cite{Ji2011} and underspecified natural
+language processing.
+
+The Ke Jia robot has been demonstrated in several tasks involving human-robot
+interaction with natural language. These tasks include a task with multiple
+\emph{pick \& carry} that are globally optimised, naive physics reasoning via
+taught rules or more complex scenarios with the robot delivering drinks, taking
+into account changing and mutually exclusive preferences of users.
+
+Also notable, Beetz et al.~\cite{Beetz2010} proposes a cognitive architecture
+called \textsc{CRAM} (Cognitive Robot Abstract Machine) that integrates
 \textsc{KnowRob}~\cite{Tenorth2009a}, a knowledge processing framework based on
@@
 Internet had to be translated into plans and executed in a kitchen environment,
-perceived and rebuilt on-line by the robots.
-
-\fxwarning{robot control architecture for service robots: what else?}
+perceived and rebuilt on-line by the robots. This architecture has however not
+been deployed in scenarios involving human-robot interactions.
 
@@
 
-\fxwarning{What are our contribution regarding perspective taking? (Mathieu)}
-\fxwarning{Mention other source of knowledge about humans -- including social rules for instance}
+In our implementation, perspective taking is tightly connected to the symbolic
+knowledge models, and since our knowledge base allows for storage of one
+knowledge model per agent, we have been able to endow the robot with a simple
+theory of mind (as explained in section~\ref{sect|tom}): we explicitly model
+what the robot knows about its partners in a symbolic way. This knowledge is
+then re-used in different places, to correctly interpret what the human says,
+or to plan tasks that are actually doable for the human.
+
+The cognitive model that the robot builds for the agent it interacts with is
+still simple and mostly focused on geometric features (who sees what? What are
+our relative positions? etc.). Extending this knowledge with more subtle
+perceptions (emotional state for instance) remains to be done.
+
+We should also mention that modelling of human behaviors also appear in our
+architecture at the level of task and motion planning with a set of
+\emph{social rules}, including proxemics, action legibility or time
+optimization.
 
@@
 
-\paragraph{Reasoning with the humans}
-
-\fxwarning{Mention we represent the human knowledge as well (Severin)}
-
 \paragraph{Limits of disambiguation at semantic level}
 
-\fxwarning{Introduce the paragraph}
+Interaction with humans implies the ability to deal semantics: semantics of
+verbal interaction, semantics of gestures, etc.  As a consequence, it also
+implies to deal with semantic disambiguation.
 
@@
 
-\fxwarning{Rewrite conclusion}
-
-We have devised a decisional framework for human-robot interactive
-task achievement that is aimed to allow the robot not only to
-accomplish its tasks but also to produce behaviors that support its
-engagement vis-a-vis its human partner and to interpret human
-behaviors and intentions. 
-Together and in coherence with this framework, we have developed
-and experimented various task planners and interaction schemes that
-allow the robot to select and perform its tasks while taking into
-account explicitly the human abilities as well as the constraints
-imposed by the presence of humans, their needs and preferences.
-
-In this paper we have presented a decisional framework designed for
-robots operating in a human environment. Our objective is to provide a
-management of human-robot interaction that is an integral part of a
-general robot control architecture.  This was done in order to provide
-a principled way to deal with HRI scenario.  The framework is also
-suitable for the development and experiment of task planners and
-interaction schemes that explicitly consider human abilities and
-preferences.
+We have presented in this article a decisional framework for human-robot
+interactive task achievement that is aimed to allow the robot not only to
+accomplish its tasks but also to produce behaviors that support its engagement
+vis-a-vis its human partner and to interpret human behaviors and intentions.
+Together and in coherence with this framework, we have developed and
+experimented various task planners and interaction schemes that allow the robot
+to select and perform its tasks while taking into account explicitly the human
+abilities as well as the constraints imposed by the presence of humans, their
+needs and preferences.
 
@@
 
-We are planning to extend the scheme by taking into accoount situation where
-divergent belief may appear between the human and the robot. Preliminary
-results have been pusblished in~\cite{warnier2012when}.
-
-There is also extensive work to be done in order to refine the notion of "good
-shared plan" and "good/acceptable robt behaviour" in this context. There are
+One direction that we would like to further investigate is how to account for
+situations where divergent beliefs appear between the human and the robot.
+Preliminary results have been published in~\cite{warnier2012when}.
+
+There is also extensive work to be done in order to refine the notion of ``good
+shared plan" and ``good/acceptable robot behavior" in this context. There are
 obviously large avenues for learning and adaptation in this context.
+
+Another broader direction to head deals with \emph{contexts representation}.
+Contexts are currently often limited to the current spatial and temporal
+situation. Some projects model offer the possibility to jump in the past or to
+switch to another agent's perspective, but in current approaches, selecting a
+context always basically consists in retrieving a set of beliefs corresponding
+to a situation, and temporarily replacing the current beliefs by those other
+ones. This misses the fact that at a given moment, not one but many contexts
+co-exist at different scales. We do not want to retrieve one monolithic set of
+beliefs, but instead carefully craft a context from several \emph{atomic}
+contexts. Techniques for representation of overlapping pools of knowledge
+largely remain to be developed, as well as efficient algorithms to retrieve (or
+discard) such context-related pools of knowledge. This is a challenge not only for robotics, but more generally for artificial intelligence.
+
+The ability to explicitly manage contexts and context switches would endow the
+robot with a cognitive capability similar to what is known as
+\emph{context-dependent memory} in cognitive psychology. This is also related
+to Tulving's \emph{autonoetic consciousness}~\cite{Tulving1985a}: the ability
+to reflect upon its own past or future experiences.
+
+Much remain to be done to this regard, starting with a formal analysis of what
+are the relevant contexts for our robots.
+
+Human-Robot Interaction is and is going to remain a challenging field for
+artificial intelligence. We hope that this contribution helps with clarifying
+some of these challenges.
 

### 2014-04-22 paper.tex
'Small' corrections from reviewer 1
Includes anyway a few rewrite of sections

@@
 
-\usepackage[nomargin, marginclue, footnote]{fixme}
+\usepackage{paralist}
+
+\usepackage[draft, nomargin, marginclue, footnote]{fixme}
 
@@
 sensing modalities, it computes and grounds perceptions into symbolic
-knowledge (situation assessment), it is responsible temporal interpretation of
+knowledge (situation assessment), it is responsible for temporal interpretation of
 situations and performs spatial perspective-taking.
@@
 
-   \caption{ The scenario involves a several videotapes that have to be
-       manipulated. Other objects are used as supports and containers.  After
+   \caption{The toy scenario pictured here involves videotapes that have to be
+       manipulated, with other objects used as supports and containers.  After
        identification and localisation of the set of objects and acquisition of
-       the position and posture of the human partner, the robot is able to compute
-       that two tapes only are reachable by itself: the \concept{LOTR\_TAPE}
-       (black in the 3D model) and \concept{GREY\_TAPE}. The third tape
-       \concept{WALLE\_TAPE} (white in the 3D model) and the trashbin
-       \concept{PINK\_TRASHBIN} are only reachable by the human. The
-       corresponding symbolic facts are listed in Table~\ref{table|beliefsfig7}.
-   }%
+       the position and posture of the human partner, the robot is able to
+       compute that two tapes only are reachable by itself: the
+       \concept{LOTR\_TAPE} (black in the 3D model) and \concept{GREY\_TAPE}.
+       The third tape \concept{WALLE\_TAPE} (white in the 3D model) and the
+   trashbin \concept{PINK\_TRASHBIN} are only reachable by the human. The
+   corresponding symbolic facts are listed in Table~\ref{table|beliefsfig7}.  }%
 
@@
 
-Geometric state of the world is abstracted in symbolic facts that can
-be classified in three different categories.
-
-\begin {itemize}
-\item Relative positions of object and agents, e.g.  \stmt{GREY\_TAPE
-    isOn TABLE}.
-
-\item Perception and manipulation capacity and state of agents,
-  e.g. \stmt{ROBOT looksAt GREY\_TAPE}, \stmt{GREY\_TAPE isVisibleBy
-    HUMAN1}.
-
-\item Motion status for object or agent parts, e.g.  \stmt{GREY\_TAPE
-    isMoving true},\\ \stmt{ROBOT\_HEAD isTurning true}.
-\end {itemize}
-
-Reasoning about human perspective allow to compute facts such as:
-\stmt{GREY\_TAPE isBehind HUMAN1}, \stmt{GREY\_TAPE leftOf
-  HUMAN1}.
-
-Figure~\ref{fig::reach-ex} illustrates different situations for the
-\textit{reachable} relation.
+\begin{inparaenum}[\itshape 1\upshape)]
+Geometric state of the world is abstracted in symbolic facts that can be
+classified in three different categories: \item relative positions of object and
+agents, \eg  \stmt{GREY\_TAPE isOn TABLE}, \item perception and manipulation
+capacity and state of agents, \eg \stmt{ROBOT looksAt GREY\_TAPE},
+\stmt{GREY\_TAPE isVisibleBy HUMAN1}, \item motion status for object or
+agent parts, \eg \stmt{GREY\_TAPE isMoving true}, \stmt{ROBOT\_HEAD
+isTurning true}.
+\end{inparaenum}
+By reasoning about human perspective, it computes facts such as:
+\stmt{GREY\_TAPE isBehind HUMAN1}, \stmt{GREY\_TAPE leftOf HUMAN1}.
+
+As an illustration, Figure~\ref{fig::reach-ex} explains how the \emph{reachable}
+affordance is mutually computed for the human and the robot. The \emph{is visible},
+\emph{looks at}, \emph{points at} affordances are computed in similar ways, by
+virtually placing OpenGL cameras in different pose (human head pose, robot head
+pose, human index finger, etc.) and evaluating which objects are visible to the camera
+(hence taking into account possible occlusions).
 
@@
         and its human partner are placed face to face (a).  The robot first
-        estimates if the small grey box is reachable to itself. This is done by
-        finding a collision free posture to reach the object (b). Next the
-        robot switches to the human's perspective to estimate if the same
-        object is reachable to the human as well (c).  In the last scene, the
-        human moves towards his left, farther from the object (d). The
-        situation is then reevaluated. In this occasion though, the reasoner
-        cannot find a satisfactory posture for the human to reach the box
-        because he is too far from the target.  }
+        estimates if the small grey box is reachable to itself using an inverse
+        kinematic (IK) solver and collision checks to find a collision free
+        posture to reach the object (b). Next the robot switches to the human's
+        perspective to estimate if the same object is reachable to the human as
+        well (c).  In the last scene, the human moves towards his left, farther
+        from the object (d). The situation is then reevaluated. In this occasion
+        though, the reasoner cannot find a satisfactory posture for the human to
+    reach the box because he is too far from the target.  }
 
@@
 \paragraph{Hypotheses on objects states and positions}
-It is sometimes difficult or even impossible to see and/or track an
-object in certain states. This happens, for instance, when the object
-is hidden in a container, when the robot holds it in its gripper, and more
-generally in any state in which it is hidden by something else. {\sc Spark} models the possible symbolic states of objects (whether the object is on a furniture, in an agent hand,
-in a container, etc.): According to the robot perception of what has
-happened since the object was last seen, the robot builds a
-belief of the current possible state with associated probability for the.
-
-\paragraph{Primitive action recognition} Monitoring human activity is crucial
-to maintain a coherent state of the world. Full human action and activity
-monitoring is a task that requires knowledge and reasoning both on high level
-facts like goals, intentions and plans, as well as bottom-up data from agent
-and object motions. Simple temporal and geometric reasoning on human hand
-trajectories and potential objects placements can provide some useful clues for
-high level human monitoring processes. We call this temporal and geometric
-reasoning \emph{primitive action recognition}. Those primitive actions are
-assessed through monitoring situations like ``an empty hand is close to an
-object on a table'' (precursor for a \emph{pick}), or ``a hand holding an
-object is over a container'' (precursor for a \emph{throw}).
+
+It is sometimes difficult or even impossible to see and/or track an object in
+certain states. This happens, for instance, when the object is hidden in a
+container, when the robot holds it in its gripper, and more generally in any
+state in which it is hidden by something else. {\sc Spark} models the possible
+symbolic states of objects (whether the object is on a furniture, in an agent
+hand, in a container, etc.): According to the robot perception of what has
+happened since the object was last seen, the robot builds a belief of the
+current possible state with associated probability model.
+
+\paragraph{Primitive action recognition}
+
+Monitoring human activity is crucial for maintaining a coherent state of the
+world. Full human action and activity monitoring is a task that requires
+knowledge and reasoning both on high level facts like goals, intentions and
+plans, as well as bottom-up data from agent and object motions. Simple temporal
+and geometric reasoning on human hand trajectories and potential objects
+placements can provide some useful clues for high level human monitoring
+processes. We call this temporal and geometric reasoning \emph{primitive action
+recognition}. Those primitive actions are assessed through monitoring situations
+like ``an empty hand is close to an object on a table'' (precursor for a
+\emph{pick}), or ``a hand holding an object is over a container'' (precursor for
+a \emph{throw}). \emph{Spark} recognizes a handful \fxfatal{which ones?} of these
+primitives, that are relied on by the robot to track the state transitions in
+the human plans (as explained below, at section~\ref{hatp}).
 
@@
 presence and value of specific facts: for instance, when several instances
-matches a category (the human says ``give me the bottle'' and the robot knows
+match a category (the human says ``give me the bottle'' and the robot knows
 about three bottles), the module may decide (it actually depends on the
@@
 me?'' to discriminate between the three bottles. That way, procedures relying
-on discrimination transparently benefit from added modalities.
+on discrimination transparently benefit from added modalities (we kindly refer
+the interested readers to~\cite{Ros2010b} for a detailed account of our
+approach to interactive concept discrimination with the accompanying algorithms).
 
@@
 \subsection{Human-aware task planning}
+\label{hatp}
 
@@
 
-HATP originality resides in its ability to produce plans for the robot's
-actions \emph{as well as for the other participants} (humans or robots): the
-robot plans for itself, but also for its interactors, which effectively endows the robot
-with new interaction capabilities: It can partially predict the behavior, and
-thus monitor the engagement, of its human partner; the planner generates
-synchronisation points (arrow going from the robot stream to the human stream on
-Figure~\ref{plan_hatp1}) between the agents; the controller can also verbalize
-the plans to explain its partner how a task may be shared~\cite{warnier2012when}.
-
-HATP can be tuned by setting up different costs depending on the actions to
-apply and by taking into account a set of constraints called \emph{social
-rules}. This tuning aims at adapting the robot's behavior according to the
-desired level of cooperation of the robot.
-
 \paragraph{Agents and action streams}
-The robot plans not only for itself but also for the other agents. The
-resulting plan, called ``shared plan'' is a set of actions that form
-a stream for each agent involved in the goal achievement. Depending on
-the context, some ``shared plans'' contain causal relations between
-agents. For example, the second agent needs to wait for the success of
-the first agent's action to be able to start its own action. When the
-plan is performed, causal links induce synchronization between
-agents. Figure~\ref{plan_hatp1} illustrates a plan with two streams.
+
+HATP originality resides in its ability to produce plans for the robot's actions
+\emph{as well as for the other participants} (humans or robots), so-called
+\emph{shared plans}. This effectively enriches the robot's interaction
+capabilities by allowing for a form of \emph{human behaviour prediction}, which
+is used in turn by the robot execution controller to monitor the engagement of
+the human partner during joint tasks.
+
+The planner also generates synchronisation points between the agents.
+Figure~\ref{plan_hatp1} illustrates such a plan with two streams.  Here, the
+robot needs to wait for the success of the human agent's {\tt THROW} action
+(estimated through the fulfilment of the action post-conditions, typically
+$\neg$\stmt{GREY\_TAPE isOn TABLE}) to be able to start its own action.
+
+We also use shared plans to verbalize the sequence of actions and to explain
+the human partner how a task may be shared~\cite{warnier2012when}.
 
@@
 \paragraph{Action costs and social rules}
-A cost and a duration function is associated to each action.
-The duration function provides a duration interval for the action
-achievement and is used, in one hand, to schedule the different
-streams and, in the other hand, as an additional cost function.
-In addition to these costs, HATP also takes into account a set of social
-rules.  Social rules are constraints aiming at leading the plan
-construction towards the best plan according to some human
-preferences. The social rules we have defined so far deal with:
+
+A cost and a duration function is associated to each action.  The duration
+function provides a duration interval for the action achievement and is used, in
+one hand, to schedule the different streams and, in the other hand, as an
+additional cost function.
+
+In addition to these costs, HATP also takes into
+account a set of social rules.  Social rules are constraints aiming at leading
+the plan construction towards the best plan according to some human preferences.
+The social rules we have defined so far deal with:
 
@@
 
-Two important remarks: because HATP is a generic symbolic task planner, we have
-been able to design a planning domain at a semantic level which is close to the
-one used in the human-robot dialogue (the planner vocabulary contains concepts
-like \texttt{give}, \texttt{table}, \texttt{is on}...). Hence only a few
-ontology rules have been required to map both the knowledge extracted from the
-situation assessment and the statements originated from the verbal interaction
-to the planner domain.
-
-Second remark, after some research (see appendix B of~\cite{Lemaignan2012a} for
-a detailed discussion)  we have decided to represent neither the planning
-domain nor the resulting plans in the knowledge base: the planning domain (with
-task pre- and postconditions) is stored in a specific format, outside of the
-central declarative knowledge repository, and the plans are directly
-communicated to the robot controller. Thus, like many other cognitive
-architectures, we have independent declarative and procedural knowledge stores.
+Because HATP is a generic symbolic task planner, we have been able to design a
+planning domain at a semantic level which is close to the one used in the
+human-robot dialogue (the planner vocabulary contains concepts like
+\texttt{give}, \texttt{table}, \texttt{is on}...). Hence only a few ontology
+rules have been required to map both the knowledge extracted from the situation
+assessment and the statements originated from the verbal interaction to the
+planner domain.
+
+Also, we would like to mention that, after some research (see appendix B
+of~\cite{Lemaignan2012a} for a detailed discussion)  we have decided to
+represent neither the planning domain nor the resulting plans in the knowledge
+base: the planning domain (with task pre- and postconditions) is stored in a
+specific format, outside of the central declarative knowledge repository, and
+the plans are directly communicated to the robot controller. Thus, like other
+cognitive architectures, we have independent declarative and procedural
+knowledge stores.
 
@@
 
+While parts of the architecture (SPARK, ORO) have been deployed with external
+execution controllers (like {\sc Cram}~\cite{Beetz2010} or the BERT
+platform~\cite{Lallee2010b}, as reported in~\cite{Lemaignan2010}), we have also
+developed dedicated robot controllers which integrate the whole stack introduced
+on Figure~\ref{fig|archi}. {\sc Shary}~\cite{clodic2008shary} is the main one,
+written in the PRS~\cite{Ingrand1996} language. We also use the Python-based
+{\sc pyRobots} collection of tools that provide a large set of high-level
+actions and an event-based architecture well suited for prototyping.
+
+They both rely on extensive integration with the knowledge base that serves as the
+primary source of semantics for decision making.
+
 \subsubsection{Desires and experiences}
 
-Our robot execution controllers (OpenPRS-based {\sc Shary} or Python-based {\sc
-pyRobots}) rely on extensive integration with the knowledge base. It serves as the
-primary source of information for semantic-aware decision-making.
-
 We split the interaction situations stemming from the situation assessment and
-communication components in two categories: \emph{desires} (performative act)
-and \emph{experiences} (assertive act).
+communication components in two categories: \emph{desires} (related to
+\emph{performative acts} in Austin's classification of speech
+acts~\cite{Austin1962}) and \emph{experiences}.
 
@@
 example, when the speaker says ``I feel tired'', we change the motion planner
-parametrization to lower the effort the human needs to provide for the
-following joint manipulation tasks. Note that this example has been implemented
-as a proof-of-concept. A broader framework that would support action alteration
-based on the user's experienced states is yet to be devised.
+parametrization to lower the effort the human needs to provide for the following
+joint manipulation tasks.\footnote{Note that this very example has been
+implemented as a proof-of-concept. A broader framework that would support
+action alteration based on the user's experienced states remains to be worked on.}
 
@@
 {\it Spy Game} (2010) & Interactive object discrimination & & \cite{Ros2010b} \\
-{\it Moving to London} (2011) & Multi-modal interaction, perspective taking & \emph{immobile robot} & \cite{lemaignan2011what} \\
+{\it Interactive Grounding I} (2011) & Multi-modal interaction, perspective taking & \emph{immobile robot} & \cite{lemaignan2011what} \\
 {\it Roboscopie} (2011) & Theater, reflection on the future of HRI & \emph{semi-autonomous} & \cite{lemaignan2012roboscopie} \\
-{\it Cleaning the table} (2011) & Full integration & & \cite{Alami2011} \\
+{\it Cleaning the table} (2011) & Whole stack integration & & \cite{Alami2011} \\
 {\it I'm in your shoes} (2012) & False beliefs & & \cite{warnier2012when} \\
 {\it Give me this} (2012) & Natural joint object manipulation & & \cite{gharbi2013natural} \\
-{\it Aperitif time} (2012) & Multi-modal interaction, perspective taking & & \cite{lemaignan2013talking} \\
+{\it Interactive Grounding II} (2012) & Multi-modal interaction, perspective taking & & \cite{lemaignan2013talking} \\
 \hline
@@
 
-The two following sections present one such experiment (the {\it Moving to
-London} scenario), and the synopsis of another one (the {\it Cleaning the
+The two following sections present one such experiment (the {\it Interactive
+Grounding I} scenario), and the synopsis of another one (the {\it Cleaning the
 table} scenario).
@@
 
-The second one involves the SHARY execution controller and the HATP symbolic
+The second one involves the {\sc Shary} execution controller and the HATP symbolic
 task planner besides ORO and {\sc Dialogs}. In this scenario, the human and the
@@
 ontology ({\tt \textbf{pointsAt}(?ag, ?obj) $\land$ \textbf{looksAt}(?ag, ?obj) $\to$
-\textbf{focusesOn}(?ag, ?obj)}) The \textsc{Dialogs} module is then able to merge both
+\textbf{focusesOn}(?ag, ?obj)}). The \textsc{Dialogs} module is then able to merge both
 sources of information, verbal (``this'') and deictic to distinguish the box
-Jerry refers to.
-
+Jerry refers to:
 \begin{center}
@@
 used instead of its ID. This way we enhance interaction using familiar names
-given by the users.
+given by the users:
 
 \begin{center}
-\begin{tabular}{l}
 \stmt{?obj isIn cardBoardBox}\\
 \hspace{0.7cm}$\Rightarrow$ \concept{?obj = videoTape2}\\
-\end{tabular}
 \end{center}
@@
 This experiment involved a more complex decisional layer where the ORO server
-was used in conjunction with the HATP symbolic task planner and the SHARY
+was used in conjunction with the HATP symbolic task planner and the {\sc Shary}
 execution controller (the complete software architecture we deployed for this
@@
 The plan produced in that case by HATP is straightforward and shown in the
-third row. It consists in four successive actions involving the robot and the
+third row. It consists of four successive actions involving the robot and the
 human. Robot grasps the tape and then places it on the table at a position
-where it is visible and reachable for the human. Human then is asked to pick
+where it is visible and reachable for the human. The human is then asked to pick
 the tape and throw it in the trashbin.
@@
 
-\section{Discussion: when AI enables HRI}
+\section{Discussion: when Artificial Intelligence enables Human Robot
+Interaction}
 \label{sect|conclusion}
@@
 non-monotonic reasoning has been especially researched within the Ke Jia
-project for symbolic task planing~\cite{Ji2011} and underspecified natural
+project for symbolic task planning~\cite{Ji2011} and underspecified natural
 language processing.
@@
 
-Interaction with humans implies the ability to deal semantics: semantics of
+Interaction with humans implies the ability to deal with semantics: semantics of
 verbal interaction, semantics of gestures, etc.  As a consequence, it also

### 2014-04-22 paper.tex
'Small' corrections from reviewer 2

@@
 
-This lead to a bidirectional design process: from \emph{bottom-up} regarding
+This leads to a bidirectional design process: from \emph{bottom-up} regarding
 the choices of concepts to model, \emph{top-down} regarding the upper part of the
 taxonomy. This upper part of the ontology is pictured on
-Figure~\ref{fig|upper_tbox}. All the classes visible on this figure belong to the
+Figure~\ref{fig|upper_tbox}. All the classes visible in this figure belong to the
 {\sc OpenCyc} namespace (the {\tt cyc:} prefix is omitted for clarity).
@@
 
-The definition of the \concept{Bottle} is a case in point:
+The definition of the \concept{Bottle} is a typical example:
 
@@
 structures (for instance, a sentence like ``learn that cats are animals'' is
-converted into \stmt{Cat subClassOf Animal} and added to the knowledge base),
+converted into \stmt{Cat subClassOf Animal} and added to the knowledge base
+after checking for possible contradictions with existing knowledge),
 interprets common temporal and place adverbs (like \emph{above} or
@@
 tired'') into \emph{experiences} (\stmt{HUMAN experiences state\_1, state\_1
-hasFeature tired}).
+hasFeature tired}). We kindly refer the readers to~\cite{Lemaignan2011a} for a
+complete discussion of these capabilities and the related algorithms.
 
@@
 currently pointing at some object. In that case, \emph{this} is replaced by the
-object focused on.
-
-Note that, while the system benefits from the complementary modalities, they
-are not all required. The dialogue system can run with only the verbal
-modality, at the cost of a simpler interaction.
-
-The second level of integration of multi-modality is implicit: by computing
-symbolic properties from the geometry, richer descriptions and hence
-discrimination possibilities are available: for instance, if reachability is
-available, the robot may ask ``do you mean the bottle that is accessible to
-me?'' to discriminate between the three bottles. That way, procedures relying
-on discrimination transparently benefit from added modalities (we kindly refer
-the interested readers to~\cite{Ros2010b} for a detailed account of our
-approach to interactive concept discrimination with the accompanying algorithms).
+object focused on. Otherwise, the robot performs a lookup in the dialogue history
+to find a previous concept that the user could refer to (anaphora resolution).
+
+Note that, while the system benefits from the complementary modalities, they are
+not all required. The dialogue system can run with the verbal modality alone, at
+the cost of a simpler interaction: if the human says ``this'' without the robot
+tracking what the human points at, no \stmt{HUMAN pointsAt ...} fact is
+available in the knowledge base, and the robot fallbacks on the anaphora
+resolution step alone.
+
+The second level of integration of multi-modality is implicit. By continuously
+computing symbolic properties from the geometric model, richer descriptions and
+hence discrimination possibilities are available. For instance, the robot may
+compute that one bottle is next to a glass while another one stands alone, and
+these symbolic descriptions are transparently re-used in a dialogue context to
+generate unambiguous references to discriminate between similar objects: ``do
+you mean the bottle that is next to the glass?'' (we kindly refer the interested
+readers to~\cite{Ros2010b} for a detailed account of our approach to interactive
+concept discrimination, along with the related algorithms).
 
@@
 developed dedicated robot controllers which integrate the whole stack introduced
-on Figure~\ref{fig|archi}. {\sc Shary}~\cite{clodic2008shary} is the main one,
+in Figure~\ref{fig|archi}. {\sc Shary}~\cite{clodic2008shary} is the main one,
 written in the PRS~\cite{Ingrand1996} language. We also use the Python-based
@@
 execution controller (the complete software architecture we deployed for this
-experiment is the one pictured on Figure~\ref{fig|archi},
+experiment is the one pictured in Figure~\ref{fig|archi},
 page~\pageref{fig|archi}).

### 2014-04-22 rebuttal.tex
First steps of the rebuttal letter

\documentclass{letter}
\usepackage{letterbib}
\usepackage[utf8]{inputenc}
\usepackage{hyperref}
\usepackage{booktabs}

\signature{Séverin Lemaignan}
\begin{document}

\subsection{General Comments}

\emph{1. The novelty of this paper with respect to other papers by the same
authors, including one in IJSR 2011, has been questioned. The authors
should convincingly explain what are the novel contributions and the
added value of this work compared to the previous publications.}

We address this question with our reponse to reviewer 3, below.

\emph{2. The form and depth of the presentation are below the standards of an
AIJ article.  The reviewers made many suggestions to make the presentation
better structured, more detailed, and more focused.}

We detail below the steps that were taken to improve the overall quality of the
paper (note that minor typos/style issues reported by the reviewers have been
fixed directly in the manuscript, and are not discussed here anymore).

\subsection{Comments from Reviewer 1}

\begin{itemize}

    \item \emph{For longer journal articles, a citation style like 'apalike' that uses the
author names in the references would be more pleasant to read since it saves
reviewers from jumping back and forth all the time.

[EiC note: please use the standard AIJ referencing style -- we do not have the
liberty to change this -- thanks]}

As indicated by the Editor in Chief, we need to keep the current citation style.

    \item \emph{p10 l 49: How is this discrimination done?} The discrimination
        algorithms are detailed in \cite{Ros2010b} (Best Paper Award at
        ROMAN2010). We have added a reference to this paper.

    \item \emph{The example scenarios have very bold names for often rather
        simple things.} The reviewer is indeed right. We have mistakenly used internal
        \emph{code names} for the experiments' names. These have been reworded.
\end{itemize}

\subsection{Comments from Reviewer 2}

\begin{itemize}
    \item \emph{pg 10 (section 3.2.1)
            "The system supports ...". This is one example of a topic that would
            be really interesting to the readers and it is just hinted. In
            particular, learning rules like "cats are animals" maybe really
            nice, but are such rules compatible with the existing knowledge?
        Adding them as triples may work only in absence of contradictions} The
        various natural language grounding processes that are made possible
        within this semantic
        architecture would require a relatively long discussion to be adequately
        explained. They are besides already presented in~\cite{Lemaignan2011a}.
        We have added an explicit reference to this previous publication.

        To specifically answer the reviewer question on how possible
        contradictions are managed, we indeed check that new statements proposed
        by the human do not lead to contradictions before inserting them.

    \item \emph{pg 10 (section 3.2.1)

            "Note that, while the system benefits from the complementary
            modalities, they are not all
            required. The dialogue system can run with only the verbal modality,
            at the cost of a simpler interaction."
        Possibly true, but unjustified claim.} An example has been added to
        clarify this fall-back mechanism.

    \item \emph{"The second level of integration of multi-modality is implicit:
            by computing symbolic properties 
            from the geometry, richer descriptions and hence discrimination
            possibilities are available: 
            for instance, if reachability is available, the robot may ask "do
            you mean the bottle that is 
            accessible to me?" to discriminate between the three bottles. That
            way, procedures relying on 
            discrimination transparently benefit from added modalities." Concept
        remains unclear.} The paragraph has been rephrased and the example
        updated to clarify the concepts.
\end{itemize}

\end{document}

### 2014-04-28 paper.tex
lazy computation
Fix #3

@@
 \begin{keyword}
-    %% keywords here, in the form: keyword \sep keyword
     human-robot interaction \sep cognitive robotics \sep perspective taking \sep knowledge representation and reasoning
-    %% MSC codes here, in the form: \MSC code \sep code
-    %% or \MSC[2008] code \sep code (2000 is the default)
-
 \end{keyword}
@@
 reasoner.
+
+Contrary to other projects like KnowRob~\cite{Tenorth2009a} that have introduced
+the concept of \emph{computables} to lazily evaluate/acquire symbolic facts when
+needed, we have an \emph{explicit} approach where we greedly compute and assert
+symbolic statements (like spatial relations between objects, see
+Section~\ref{sect|sit-ass}). This design choice trades scalability for explicit
+reasoning: at any time, the full belief state is explicitely stated, and
+provides the reasoner with the largest possible inference domain. This is of
+special importance for an \emph{event-based} architecture (see
+Section~\ref{sect|ctrl}), where, by definition, we do not know beforehand what
+we are ``looking for'', and we prefer instead greedly infer as much as possible
+to build a belief state as comprehensive as possible.
 

### 2014-04-28 paper.tex
Do not mention TBOX immutable in Prolog
Fix #12

@@
 KnowRob}~\cite{Tenorth2009a}) that mix RDF with more expressive logic languages
-like {\sc Prolog}, at the price, however, of other limitations, like
-closed-world reasoning or immutable T-Box. The classification performance is
-another issue: from our experience, with an ontology sized for a standard
-experiment (about 100 classes and 200 instances), classification typically
-takes about 100ms, which becomes problematic during interactions.  Besides, the
-performances are difficult to predict, since a seemingly inoffensive new
-statement may indirectly change abruptly the logical complexity of the whole
-knowledge model and lead to notable degradation of classification time.
+like {\sc Prolog} with other limitations, like closed-world reasoning. The
+classification performance is another issue: from our experience, with an
+ontology sized for a standard experiment (about 100 classes and 200 instances),
+classification typically takes about 100ms, which becomes problematic during
+interactions.  Besides, the performances are difficult to predict, since a
+seemingly inoffensive new statement may indirectly change abruptly the logical
+complexity of the whole knowledge model and lead to notable degradation of
+classification time.
 

### 2014-04-28 paper.tex
...clarifying reasoning for desambiguation
~ Fix #22

@@
 
-Since only one object remains, the robot infers
-that the human refers to it and would eventually execute the command, \ie give
-it to the human. Alternatively, the robot could first verify with the human if
-that was the object being referred to or not before proceeding to execute the
-action. Table~\ref{table|ptbeliefs} lists the robot's beliefs about itself and
-its human partner involved in this situation.
-
-\begin{table}
+Since only one object remains (\ie the referent that \concept{VideoTape} stands
+for is not ambiguous anymore), the robot infers that the human refers to it and
+eventually execute the command, \ie give it to the human. Alternatively, the
+robot first asks the human for confirmation (``was \concept{videoTape1} the
+object you had in mind?'') before proceeding with the action.
+Table~\ref{table|ptbeliefs} lists the robot's beliefs about itself and its human
+partner involved in this situation.
+
+\begin{table}[ht!]
 \begin{center}
+    \subfigure{
 \begin{tabular}{l}
@@
 \hline
+\end{tabular}
+}\hspace{1em}%
+    \subfigure{
+\begin{tabular}{l}
 \hline
@@
 \end{tabular}
+}
 \end{center}

### 2014-05-01 paper.tex
Archi redraw in tikz
While here, updated some labels, improved caption

@@
 \usepackage{graphicx, subfigure}
+\usepackage[usenames,dvipsnames]{xcolor}
+\usepackage{tikz}
+\usetikzlibrary{positioning}
 
@@
 Incoming statements may in turn trigger events in the system, which
-allow reactive behaviors. These mechanisms are illustrated in the coming
+allow reactive behaviours. These mechanisms are illustrated in the coming
 sections.
@@
         \centering
-        \includegraphics[width=0.9\textwidth]{archi}
-        \caption{Overview of the deliberative layer. Knowledge is
-        centrally managed in an active \emph{semantic blackboard}, pictured
-        above with a thick border.}
+
+\resizebox{\textwidth}{!}{%
+
+\tikzset{subpart/.style={draw, font=\scriptsize, fill opacity=0.5, text opacity=1, fill=black!50}}
+
+\begin{tikzpicture}[
+    >=latex,
+    every edge/.style={draw, very thick},
+    skill/.style={draw, rounded corners, align=center, inner sep=5pt, fill=black!20},
+    label/.style={midway, align=center, font=\scriptsize, fill=white}]
+
+  %%% ORO
+  \node at (0,0)[skill, ultra thick, fill=LimeGreen!50] (oro) {Symbolic facts \\ and beliefs management};
+
+  %%% HATP
+  \node at (-6, 2)[skill, fill=PineGreen!50] (hatp) {Human-aware symbolic \\ task planning};
+
+  %%% DIALOGS
+  \node at (-6, -2) [skill, fill=BrickRed!50] (dialogs) {Dialogue processing};
+
+  %%% SPARK
+  \node at (4,-3)[skill, fill=Purple!50] (spark) {%
+      \begin{tikzpicture}
+        \node at (0,0) (geom) {Geometric \& Temporal Reasoning};
+        \node [subpart, below=0.2 of geom.south west, anchor=north west] (world-update) {Sensors fusion};
+        \node [subpart, right=0.2 of world-update] (geom-model) {Geometric model of the environment};
+        \node [subpart, right=0.2 of geom-model] (fact-prod) {Symbolic facts production};
+      \end{tikzpicture}
+    };
+
+  %%% MHP
+  \node at (9,0)[skill, fill=BurntOrange!50] (mhp) {Human-aware \\ Motion and Manipulation \\ Planning};
+
+  %%% SHARY
+  \node at (4,4)[skill, fill=RoyalBlue!50] (shary) {%
+      \begin{tikzpicture}
+        \node at (0,0) (exec) {Execution Controller};
+        \node [subpart, below=0.2 of exec.south west, anchor=north west] (plans) {Goal \& Plans \\ management};
+        \node [subpart, right=0.2 of plans] (sit-asses) {Situation assessment \\ and context management};
+        \node [subpart, right=0.2 of sit-asses] {Action instanciation, \\ execution and monitoring};
+      \end{tikzpicture}
+    };
+
+  %%% LOWLEVEL
+  \node [skill, below=of spark] (lowlevel) {%
+      \begin{tikzpicture}
+        \node at (0,0) (sensori) {Sensorimotor layer};
+        \node [subpart, below=0.2 of sensori.south west, anchor=north west, align=left] (perception) {{\bf Perception} \\ 2D markers, RGB-D, motion capture};
+        \node [subpart, align=right, right=0.2 of perception] {{\bf Actuation} \\ Head's pan-tilt unit, grippers, arms, wheels};
+      \end{tikzpicture}
+  };
+
+  %%% Separation between deliberative layer and sensori-motor layer
+  \draw[dotted, thick] (-8,-4.5) -- (12, -4.5);
+
+  %%% Relations between components
+  \path (shary.340) edge [<->, bend left] node[label] {motion plan \\ requests} (mhp);
+  \path (shary.west) edge [<->, bend right] node[label] {shared \\ plans} (hatp);
+  \path (hatp) edge [<->, bend right] node[label] {world model and \\ agents beliefs} (oro.170);
+  \path (dialogs) edge [<->, bend left] node[label] {natural language \\ grounding} (oro.190);
+  \path (spark.100) edge [->, bend right] node[label] {symbolic \\ facts} (oro);
+  \path (shary) edge [<->, bend left] node[label] {Events, \\ World model and \\ agents beliefs} (oro);
+  \path (shary) edge [<->, bend left] node[label] {action monitoring \\ and management of \\ position hypotheses} (spark);
+  \path (lowlevel) edge [->] (spark);
+  \path (lowlevel.east) edge [<-, bend right=80, looseness=1.2] (shary.east);
+
+\end{tikzpicture}
+}
+
+\caption{Overview of the architecture. The deliberative layer (on the top),
+composed of six main modules, interacts with the low-level sensori-motor
+layer (on the bottom). Knowledge is centrally managed in an active
+\emph{semantic blackboard}, pictured above with a thick border. The links
+between components depicted on the figure underline the central role of the
+knowledge base: most of the datastreams are actually symbolic statements exchanged
+through the semantic blackboard.}
+
     \label{fig|archi}

### 2014-05-01 paper.tex
Tikzified the ORO TBox

@@
     \centering
-    \includegraphics[width=\columnwidth]{top_tbox.pdf}
+
+\resizebox{\textwidth}{!}{%
+\begin{tikzpicture}[
+    >=latex,
+    every edge/.style={<-, draw, very thick},
+    every node/.style={draw, font=\tt, node distance=0.5, rounded corners,
+align=center, inner sep=5pt,fill=LimeGreen!50}]
+
+    \node[fill=Purple!50] (thing) {\textbf{thing}};
+    \node [fill=BurntOrange!50, node distance=1.8, below left=of thing](sthing) {spatial thing} edge (thing);
+        \node [fill=BurntOrange!50, above left=of sthing] {posture} edge (sthing);
+        \node [fill=BurntOrange!50, left=of sthing] {shape} edge (sthing);
+        \node [fill=BurntOrange!50, below=of sthing] (location) {location} edge (sthing);
+            \node [fill=BurntOrange!50, below right=of location] {zone} edge (location);
+            \node [fill=BurntOrange!50, below left=of location] (set) {spatial enduring thing} edge (location);
+                \node [fill=BurntOrange!50, below right=of set] {obstacle} edge (set);
+                \node [fill=BurntOrange!50, below left=of set] {opening} edge (set);
+                \node [fill=BurntOrange!50, below=1 of set] {partially tangible thing} edge (set);
+                \node [fill=BurntOrange!50, above left=of set] {place} edge (set);
+
+    \node [node distance=1, below right=of thing] (tthing) {temporal thing} edge (thing);
+        \node [below=of tthing] (tte) {thing with temporal extend} edge (tthing);
+            \node [below right=of tte] {time interval} edge (tte);
+            \node [below=1 of tte] (sit) {situation} edge (tte);
+                \node [below right=of sit] (evt) {event} edge (sit);
+                    \node [below right=of evt] (act) {action} edge (evt);
+                        \node [below=of act] {purposeful action} edge (act);
+                \node [below left=of sit] {static situation} edge (sit);
+
+\end{tikzpicture}
+}
 
@@
     \centering
-    \includegraphics[width=0.6\columnwidth]{tangible_things_tbox.pdf}
+
+\resizebox{0.7\textwidth}{!}{%
+\begin{tikzpicture}[
+    >=latex,
+    every edge/.style={<-, draw, very thick},
+    every node/.style={draw, font=\tt, node distance=0.5, rounded corners,
+align=center, inner sep=5pt,fill=BurntOrange!50}]
+
+    \node[fill=Purple!50] (thing) {\textbf{thing}};
+    \node [below=1.5 of thing] (ptt) {partially tangible thing} edge[dashed] (thing);
+        \node [left=of ptt] {physical support} edge (ptt);
+        \node [below left=of ptt] {visual mark} edge (ptt);
+        \node [below right=of ptt] {container} edge (ptt);
+        \node [right=of ptt] {fluid} edge (ptt);
+        \node [below=1.5 of ptt] (stt) {solid tangible thing} edge (ptt);
+            \node [below=1 of stt] {graspable object} edge (stt);
+            \node [below left=of stt] {body part} edge (stt);
+            \node [left=of stt] {furniture} edge (stt);
+            \node [below right=of stt] (ea) {embodied agent} edge (stt);
+                \node [below=of ea] {human} edge (ea);
+                \node [right=of ea] {robot} edge (ea);
+
+\end{tikzpicture}
+}
     \caption{TBox of the specialisations of \concept{PartiallyTangible}.}

### 2014-05-01 paper.tex
Rewrite of section I

@@
 
-\section{The challenge of human-robot interaction}
+\section{The Challenge of Human-Robot Interaction}
+
+\subsection{The Human-Robot Interaction Context}
 
 Human-robot interaction is a challenge for artificial intelligence. This field
-indeed lays at the crossroad of several other domains of AI and requires to
-tackle them in a holistic manner: Modeling humans and human cognition;
-acquiring, representing, manipulating in a tractable way abstract knowledge;
-reasoning on this knowledge to make decisions; and eventually instantiating
-those decisions into physical actions in coordination with humans. Many AI
-techniques are invited, from visual processing to symbolic reasoning, from task
-planning to \emph{theory of mind} building, from reactive control to action
-recognition and learning.
-
-We do not tackle all these issues. This article attempts however to organize
-them into a coherent challenge for artificial intelligence, and also to make
-explicit some of the paths that have been investigated on our robots and that
+lays at the crossroad of several domains of AI and requires to tackle them in a
+holistic manner: Modeling humans and human cognition; acquiring, representing,
+manipulating in a tractable way abstract knowledge at the human level; reasoning
+on this knowledge to make decisions; and eventually instantiating those
+decisions into physical actions both legible to and in coordination with humans.
+Many AI techniques are invited, from visual processing to symbolic reasoning,
+from task planning to \emph{theory of mind} building, from reactive control to
+action recognition and learning.
+
+We do not claim to address here the whole issue. This article attempts however to organize
+it into a coherent challenge for artificial intelligence, and to explain and
+illustrate some of the paths that we have investigated on our robots and that
 result in a deliberative, knowledge-oriented, architecture for human-robot
@@
 
-The class of interactions we consider are essentially multi-modal situated
-dialog and collaborative task achievement: a robot which is expected to achieve
-interactive object manipulation, fetch and carry tasks and similar tasks in a
-domestic environment, with human partners (Figure~\ref{fig:hri-dec}). Two
-agents (a human and a robot) share a common space and exchange information
-through multiple modalities. The robot is expected to explicitly take into
-account the human and to manage its interactions with him.
-
-Namely, the robot must be able to recognize, understand and participate
-to communication situations, both explicit (\eg the human addresses verbally
-the robot) and implicit (\eg the human points to an object); the robot must be
-able to take part to joint actions, both proactively (by planning and proposing
-such an action to the human) and reactively; the robot must be able to move and
-act in a safe, efficient and legible way, taking into account social rules like
-proxemics.
+We focus on a specific class of interactions: collaborative task achievement
+supported by multi-modal and situated communication. Figure~\ref{fig:hri-dec}
+illustrates this context: the two agents share a common space and exchange
+information through multiple modalities, and the robot is expected to achieve
+interactive object manipulation, fetch and carry tasks and other similar
+domestic tasks by taking into account, at every stages, the intentions, beliefs,
+perspectives, skills of the human partner.  Namely, the robot must be able to
+recognize, understand and participate to communication situations, both explicit
+(\eg the human addresses verbally the robot) and implicit (\eg the human points
+to an object); the robot must be able to take part to joint actions, both
+proactively (by planning and proposing resulting plans to the human) and
+reactively; the robot must be able to move and act in a safe, efficient and
+legible way, taking into account social rules like proxemics.
 
@@
 \centering
-\includegraphics[width=0.6\columnwidth]{grounding_robot.pdf}
-\caption{Robot reasoning about HRI and anticipation of human activities.
-  Sources of information are multi-modal dialogue, and monitoring of the
-  environment and human activity. The robot must adapt on-line its behaviors
-  by merging computed plans with reactive control. The robot explicitly reasons
-  on the fact that it is (or not) observed by the human. Reasoning and planning
-  take place at symbolic as well as geometric level and systematically take
-  into account both agents beliefs and activities (from the point of view of
-  the robot).}
+\begin{tikzpicture}
+    \node at (0,0) {\includegraphics[width=0.6\textwidth]{grounding_robot.png}};
+    \node[circle, draw] at (-3.5,2.5) {A};
+    \node[circle, draw] at (5.5,2) {D};
+    \node[circle, draw] at (5.2,-1) {C};
+    \node[circle, draw] at (-1,2) {B};
+\end{tikzpicture}
+
+\caption{The robot reasons and acts in domestic interaction scenarios.  The
+    sources of information are multi-modal dialogue {\it (A)}, and
+    perspective-aware monitoring of the environment and human activity {\it
+    (B)}. The robot must adapt on-line its behaviours by merging computed plans
+    with reactive control. The robot explicitly reasons on the fact that it is
+    (or not) observed by the human. Reasoning and planning take place at
+    symbolic {\it (C)} as well as geometric level and take into account agents
+    beliefs, perspectives and capabilities {\it (D)} as estimated by the robot.}
 
@@
 
-Those three items have broad implications regarding the required underlying
-cognitive skills.  A \emph{joint action}, for instance, can be described as
-follow. Given:
+\emph{Communication}, \emph{Joint Actions}, \emph{Human-Aware Execution}: these
+three challenges structure the whole field of human-robot interaction research,
+and can be analysed in terms of the cognitive skills that they require.
+
+A \emph{joint action}, for instance, builds from:
 
 \begin {itemize}
-\item a joint goal, which has been previously established and agreed
+    \item a joint \emph{goal}, which has been previously established and agreed
         upon (typically through dialogue),
-\item the current situation, acquired through perception or
-  deduction from previous perceptions, including the state of the
-  environment of the robot and of the human,
-\item {\it a priori} common-sense knowledge,
+
+    \item a physical environment, estimated through the robot's exteroceptive
+        sensing capabilities, and complemented by inferences drawn from previous
+        observations,
+
+    \item a belief state that includes {\it a priori} common-sense knowledge and
+        mental models of each of the interactors.
+
 \end {itemize}
 
-the robot controller computes an action to execute and who (the 
-human or the robot, or both in case of a joint action) has to perform
-it, and then controls or monitors its execution. The operation
-continues until the goal is achieved, is declared unachievable or is
-abandoned by the human.
-
-To do so, the robot has to be equipped with a number of decisional, planning
-and interpretation abilities where its human partner is taken explicitly into
-account. It needs to be able:
-
-\begin{itemize}
-\item to build and maintain relevant robot and human beliefs
-  (from the robot perspective) with respect to state of the world and the task,
-\item to build and maintain iteratively shared (human-robot) plans, 
-\item to refine and execute the actions it has to perform, and to monitor 
-those achieved by its human partner.
-\end{itemize}
-
-Besides, we would like to build such abilities in a generic way, and
-to provide several levels of parametrization allowing to adapt to
-various environments, and various levels of involvement of the robot
-ranging from teammate behavior to assistant or proactive helper.
-
-We will see, in the following sections, how cognitive skills for realistic
-human-robot interaction can be implemented and organized together in a coherent
-architecture.
+The robot controller then decides on an action to execute next and who has to
+perform it, from the human, the robot or both in case of a joint action, and
+eventually controls or monitors its execution. The operation continues until the
+goal is achieved, is declared unachievable or is abandoned by the human.
+
+\begin{inparaenum}
+This translates into several decisional, planning, representation skills that
+need to be available to the robot. It must be able \item to represent and
+manipulate symbolic belief states, \item to acquire and maintain them up-to-date
+with respect to state of the world and the task at hand, \item to build and
+maintain, and iteratively refine shared (human-robot) plans, \item to
+instantiate and execute the actions it has to perform, and conversely,
+to monitor those achieved by its human partner.
+\end{inparaenum}
+
+Such abilities must be designed and implemented in generic ways and
+provide several levels of parametrization such as they adapt to
+various environments, different tasks and variable levels of engagement of the
+robot, ranging from teammate behaviour to assistant or proactive helper.
+
+\subsection{The Key Issues}
+
+TDB
 
@@
 reasoning in a \textit{top-down} way to produce new facts that may in return
-trigger physical behaviors.
+trigger physical behaviours.
 
@@
 
+SPARK is designed to run and continously update the knowledge base (in our
+experiments, it runs at about 10Hz). At each steps, it re-computes spatial
+relationships and affordances for the whole scene, and send the delta (new
+relations and relations that do not hold anymore) to the knowledge base. This
+approach may raise scalability concerns (we did not however hit such issues in
+our constrainted scenarios), but greatly simplifies the management of the
+\emph{dynamics} of the knowledge (\emph{When do I discard outdated knowledge?
+When do I update it?}). Since it is equivalent to a reset of the reasoner
+domain, it also hides issues linked to non-monotonic reasoning
+(see~\cite{McCarthy2007} for a discussion on that question).
+
 \begin{figure*}[ht!]
@@
 \begin{center}
+    \subfigure{
 \begin{tabular}{l}
 \hline
-Robot's beliefs about itself (\emph{robot's model}):\\
+Robot's beliefs about itself (\emph{robot's model})\\
 \hline
@@
 \hline
-\hline
-Robot's beliefs about the human (\emph{human's model}):\\
+\end{tabular}
+}\hspace{1em}%
+\subfigure{
+\begin{tabular}{l}
+\hline
+Robot's beliefs about the human (\emph{human's model})\\
 \hline
@@
 ?action, ?action type Look, ?action hasGoal myself}, we could trigger a
-behavior when the human expresses (through dialogue, gestures...) that he
+behaviour when the human expresses (through dialogue, gestures...) that he
 wants to look at the robot itself.
@@
 task planner besides ORO and {\sc Dialogs}. In this scenario, the human and the
-robot try to cooperatively remove objects from a table. behaviors and motions
+robot try to cooperatively remove objects from a table. behaviours and motions
 are fully planned.
@@
 
-We should also mention that modelling of human behaviors also appear in our
+We should also mention that modelling of human behaviours also appear in our
 architecture at the level of task and motion planning with a set of
@@
 interactive task achievement that is aimed to allow the robot not only to
-accomplish its tasks but also to produce behaviors that support its engagement
-vis-a-vis its human partner and to interpret human behaviors and intentions.
+accomplish its tasks but also to produce behaviours that support its engagement
+vis-a-vis its human partner and to interpret human behaviours and intentions.
 Together and in coherence with this framework, we have developed and
@@
 There is also extensive work to be done in order to refine the notion of ``good
-shared plan" and ``good/acceptable robot behavior" in this context. There are
+shared plan" and ``good/acceptable robot behaviour" in this context. There are
 obviously large avenues for learning and adaptation in this context.

### 2014-05-01 paper.tex
Tikzified HATP plans

@@
 \newcommand{\al}{{\textit{et al.\ }}}
+
+\newcommand{\action}[3]{#1\\\textsf{\scriptsize #2,}\\\textsf{\scriptsize #3}}
 
@@
 Figure~\ref{plan_hatp1} illustrates such a plan with two streams.  Here, the
-robot needs to wait for the success of the human agent's {\tt THROW} action
+robot needs to wait for the success of the human agent's {\tt PUT} action
 (estimated through the fulfilment of the action post-conditions, typically
@@
     \centering
-  \includegraphics[width=0.95\columnwidth]{first_plan.pdf}
-  \caption{A plan produced by HATP with 2 streams}
+\resizebox{0.9\textwidth}{!}{%
+\begin{tikzpicture}[
+    >=latex,
+    robot/.style={fill=RoyalBlue!50},
+    every edge/.style={<-, draw, ultra thick},
+    every node/.style={ draw, 
+                        thick,  
+                        circle, 
+                        font=\sf,
+                        align=center,
+                        node distance=2cm,
+                        fill=BurntOrange!50, 
+                        minimum size=2cm, 
+                        inner sep=0.1cm}]
+
+    \node (h1) {TAKE\\{\tiny \it \sf GREY\_TAPE, TABLE}};
+    \node[right=of h1] (h2) {PUT\\{\tiny\it\sf GREY\_TAPE, BIN1 }} edge (h1);
+
+    \node[robot, below=1.5 of h1] (r1) {TAKE\\{\tiny \it \sf BLACK\_TAPE, TABLE }};
+    \node[robot, right=of r1] (r2) {PUT\\{\tiny \it \sf BLACK\_TAPE, BIN2 }} edge (r1);
+    \node[robot, right=of r2] (r3) {TAKE\\{\tiny \it \sf BOTTLE, TABLE }} edge (r2);
+    \node[robot, right=of r3] (r4) {PUTRV\\{\tiny \it \sf BOTTLE, TABLE}} edge (r3);
+    \node[robot, right=of r4] (r5) {TAKE\\{\tiny \it \sf BOOK, TABLE }} edge (r4);
+    \node[robot, right=of r5] (r6) {PUT\\{\tiny \it \sf BOOK, BIN2 }} edge (r5);
+
+    \node[right=9cm of h2] (h3) {TAKE\\{\tiny \it \sf BOTTLE, TABLE}} edge (h2) edge (r4);
+    \node[right=of h3] (h4) {PUT\\{\tiny \it \sf BOTTLE, BIN1}} edge (h3);
+
+\end{tikzpicture}
+}
+\caption{A plan produced by HATP with two action streams (here, human actions
+    are at the top while robot actions are at the bottom). The arrow between the
+    two streams represents a synchronization point ({\sf PUTRV} stands for {\it Put it so it is both
+    reachable and visible}).}
+
   \label{plan_hatp1}
@@
   \centering
-  \includegraphics[width=0.95\columnwidth]{second_plan.pdf}
-  \caption{A plan with the wasted time social rule. {\tt PUTRV} stands for {\it Put it so it is both Reachable and Visible}.}
+\resizebox{0.9\textwidth}{!}{%
+\begin{tikzpicture}[
+    >=latex,
+    robot/.style={fill=RoyalBlue!50},
+    every edge/.style={<-, draw, ultra thick},
+    every node/.style={ draw, 
+                        thick,  
+                        circle, 
+                        font=\sf,
+                        align=center,
+                        node distance=2cm,
+                        fill=BurntOrange!50, 
+                        minimum size=2cm, 
+                        inner sep=0.1cm}]
+
+    \node (h1) {TAKE\\{\tiny \it \sf GREY\_TAPE, TABLE}};
+    \node[right=of h1] (h2) {PUT\\{\tiny\it\sf GREY\_TAPE, BIN1 }} edge (h1);
+
+    \node[robot, below=1.5 of h1] (r3) {TAKE\\{\tiny \it \sf BOTTLE, TABLE }};
+    \node[robot, right=of r3] (r4) {PUTRV\\{\tiny \it \sf BOTTLE, TABLE}} edge (r3);
+    \node[robot, right=of r4] (r1) {TAKE\\{\tiny \it \sf BLACK\_TAPE, TABLE }} edge (r4);
+    \node[robot, right=of r1] (r2) {PUT\\{\tiny \it \sf BLACK\_TAPE, BIN2 }} edge (r1);
+    \node[robot, right=of r2] (r5) {TAKE\\{\tiny \it \sf BOOK, TABLE }} edge (r2);
+    \node[robot, right=of r5] (r6) {PUT\\{\tiny \it \sf BOOK, BIN2 }} edge (r5);
+
+    \node[right=of h2] (h3) {TAKE\\{\tiny \it \sf BOTTLE, TABLE}} edge (h2) edge (r4);
+    \node[right=of h3] (h4) {PUT\\{\tiny \it \sf BOTTLE, BIN1}} edge (h3);
+
+\end{tikzpicture}
+}
+
+\caption{An alternative plan where the \emph{wasted time} social rule is used to optimize the
+total duration of the task. }
+
   \label{plan_hatp2}
@@
   \centering
-  \includegraphics[width=1.0\textwidth]{third_plan.pdf} \\
-  \caption {A plan produced by HATP to execute the high-level order ``clean the table''.
-Two streams of actions are generated: for the human (top) and for the robot(bottom).
-Synchronization points ensure the coordination.}
+\resizebox{0.9\textwidth}{!}{%
+\begin{tikzpicture}[
+    >=latex,
+    robot/.style={fill=RoyalBlue!50},
+    every edge/.style={<-, draw, ultra thick},
+    every node/.style={ draw, 
+                        thick,  
+                        circle, 
+                        font=\sf,
+                        align=center,
+                        node distance=2cm,
+                        fill=BurntOrange!50, 
+                        minimum size=2cm, 
+                        inner sep=0.1cm}]
+
+    \node (h1) {\action{TAKE}{WALLE\_TAPE}{TABLE}};
+    \node[right=of h1] (h2) {\action{THROW}{WALLE\_TAPE}{TRASH\_1 }} edge (h1);
+
+    \node[robot, below=1.5 of h1] (r1) {\action{TAKE}{GREY\_TAPE}{TABLE }};
+    \node[robot, right=of r1] (r2) {\action{PUTRV}{GREY\_TAPE}{TABLE}} edge (r1);
+    \node[robot, right=of r2] (r3) {\action{TAKE}{LOTR\_TAPE}{TABLE }} edge (r2);
+    \node[robot, right=of r3] (r4) {\action{PUTRV}{LOTR\_TAPE}{TABLE }} edge (r3);
+
+    \node[right=of h2] (h3) {\action{TAKE}{GREY\_TAPE}{TABLE}} edge (h2) edge (r2);
+    \node[right=of h3] (h4) {\action{THROW}{GREY\_TAPE}{TRASH\_1}} edge (h3);
+    \node[right=of h4] (h5) {\action{TAKE}{LOTR\_TAPE}{TABLE}} edge (h4) edge (r4);
+    \node[right=of h5] (h6) {\action{THROW}{LOTR\_TAPE}{TRASH\_1}} edge (h5);
+
+\end{tikzpicture}
+}
+
+  \caption {A plan produced by HATP to execute the high-level order ``clean the
+  table''. Two streams of actions are generated: for the human (top) and for the
+  robot (bottom). Synchronization points ensure the coordination.}
+
   \label{plan-etat2}

### 2014-05-02 paper.tex
sans serif font for ontology excerpts

@@
     every edge/.style={<-, draw, very thick},
-    every node/.style={draw, font=\tt, node distance=0.5, rounded corners,
+    every node/.style={draw, font=\sf, node distance=0.5, rounded corners,
 align=center, inner sep=5pt,fill=LimeGreen!50}]
@@
     every edge/.style={<-, draw, very thick},
-    every node/.style={draw, font=\tt, node distance=0.5, rounded corners,
+    every node/.style={draw, font=\sf, node distance=0.5, rounded corners,
 align=center, inner sep=5pt,fill=BurntOrange!50}]

### 2014-05-02 paper.tex
Tikzified spark-oro diagram

@@
 \centering
-        \includegraphics[width=0.8\columnwidth]{spark-oro.png}
+\resizebox{0.8\textwidth}{!}{%
+\begin{tikzpicture}[
+        >=latex,
+        box/.style={draw,rectangle, dotted, minimum width=#1, minimum height=3.8cm},
+        box/.default={4.5cm},
+        every edge/.style={draw, ultra thick},
+        every node/.style={align=center}]
+
+    %\draw[step=1cm, draw=black!20] (-10,-5) grid (10,5);
+
+    \node[draw=none] at (0,0) {\includegraphics[width=0.8\columnwidth]{spark-oro-notext.pdf}};
+
+    \node[box=16cm, minimum height=5.7cm, solid, thick, rounded corners] at (0,1.8) (oro) {};
+    \node at (0,4.2) {\sc \Large oro-server};
+
+    \node[anchor=south] at (-4.6,2.4) {Description Logics \\ (Open World assumption)};
+    \node[box=3.7cm] at (-4.6,1.9) (owl) {};
+    \node[anchor=south] at (-1,2.4) {Multiple simultaneous \\ symbolic models};
+    \node[box=3.3cm] at (-1,1.9) (multi) {};
+    \node[anchor=south] at (2.1,2.4) {Bio-inspired \\ memory};
+    \node[box=2.7cm] at (2.1,1.9) (mem) {};
+    \node[anchor=south] at (5,2.4) {Discrimination \& \\ categorization};
+    \node[box=3cm] at (5,1.9) (disc) {};
+
+    \node at (-3,-0.6) {{\it OWL store:} OpenJena};
+    \node at (3,-0.6) {{\it Reasoner:} Pellet};
+
+    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+    \node[box=16cm, minimum height=3cm, dotted, thick, rounded corners] at (0,-3.75) (spark) {};
+    \path (spark.north) edge[->] node[align=left,anchor=west] {Symbolic\\facts} (oro.south);
+
+    \node[anchor=north] at (-4.5,-4.5) {Relative locations};
+    \node[anchor=north] at (-1.5,-4.5) {Spatial relations};
+    \node[anchor=north] at (1.5,-4.5) {Visibility};
+    \node[anchor=north] at (4.8,-4.5) {Reachability};
+
+\end{tikzpicture}
+}
     \caption{Functional overview of knowledge base ({\sc Oro} server, top part)

### 2014-05-02 paper.tex
Rewrite section 2

@@
 \usetikzlibrary{positioning}
+
+\usepackage{amsmath} % \text
 
@@
 
-\section{From human-robot interaction to a deliberative architecture}
-
-\subsection{Building a human-aware deliberative layer}
-
-Connecting multiple independent software modules in one coherent robotic
-architecture is not only a technical challenge, but also a design and
-architectural challenge.
-
-In particular, dealing with the rich semantics of natural interaction with
-humans has been one of our major focus. To this end, we have been deploying
-over the last years explicit knowledge representation and manipulation in the
-deliberative layer of the robots as a {\it lingua franca} between our
-components.
-
-Figure~\ref{fig|archi} gives an overview of resulting
-architecture~\cite{Alami2011}. An active knowledge base, conveniently thought
-as a \emph{semantic blackboard}, connects most of the modules together: the
-\emph{geometric reasoning} module (called {\sc Spark}) produces symbolic
-assertions describing the dynamics of the robot's environment. These logical
-statements are stored in the knowledge base, and reused by the language
-processing module (called {\sc Dialogs}), the symbolic task planner (HATP) and
-the execution controllers. The output of the language processing module, or the
-activities started by the robot controller are likewise stored as symbolic
-statements.
-
-Incoming statements may in turn trigger events in the system, which
-allow reactive behaviours. These mechanisms are illustrated in the coming
-sections.
-
-\begin{figure*}
+\section{Deliberative Architecture and Knowledge Model}
+
+\begin{figure*}[ht]
 \centering
@@
   %%% ORO
-  \node at (0,0)[skill, ultra thick, fill=LimeGreen!50] (oro) {Symbolic facts \\ and beliefs management};
+    \node at (0,0)[skill, ultra thick, fill=LimeGreen!50] (oro) {{\sc Oro} -- Symbolic facts \\ and beliefs management};
 
   %%% HATP
-  \node at (-6, 2)[skill, fill=PineGreen!50] (hatp) {Human-aware symbolic \\ task planning};
+  \node at (-6, 2)[skill, fill=PineGreen!50] (hatp) {HATP -- Human-aware \\ symbolic task planning};
 
   %%% DIALOGS
-  \node at (-6, -2) [skill, fill=BrickRed!50] (dialogs) {Dialogue processing};
+  \node at (-6, -2) [skill, fill=BrickRed!50] (dialogs) {{\sc Dialogs} \\ Dialogue processing};
 
@@
       \begin{tikzpicture}
-        \node at (0,0) (geom) {Geometric \& Temporal Reasoning};
+          \node at (0,0) (geom) {{\sc Spark} -- Geometric \& Temporal Reasoning};
         \node [subpart, below=0.2 of geom.south west, anchor=north west] (world-update) {Sensors fusion};
@@
   %%% MHP
-  \node at (9,0)[skill, fill=BurntOrange!50] (mhp) {Human-aware \\ Motion and Manipulation \\ Planning};
+    \node at (9,0)[skill, fill=BurntOrange!50] (mhp) {{\sc mhp} -- Human-aware \\ Motion and Manipulation \\ Planning};
 
@@
   \path (spark.100) edge [->, bend right] node[label] {symbolic \\ facts} (oro);
-  \path (shary) edge [<->, bend left] node[label] {Events, \\ World model and \\ agents beliefs} (oro);
+  \path (shary) edge [<->, bend left] node[label] {events, \\ world model and \\ agents beliefs} (oro);
   \path (shary) edge [<->, bend left] node[label] {action monitoring \\ and management of \\ position hypotheses} (spark);
@@
 
-This architecture moves away from standard layered approaches found in
-robotics~\cite{Gat1998three, Volpe2001CLARAty, Goldberg2002}. Interactions
-between components at the deliberative level are mostly bidirectional and we do
-not introduce layers of abstraction amongst software components\footnote{We do
-have lower-level modules to execute actions or manage sensors, but all
-cognition-related modules reside at the same level.}. Dialogue processing, for
-instance, illustrates this structure: This component does not simply act as an
-alternative perceptual input to the symbolic database; it also
-actively queries previously acquired knowledge to disambiguate and validate the
-newly created symbolic knowledge (see section~\ref{sect|com}).
-
-Our architecture relates to \emph{Beliefs, Desires, Intentions} (BDI)
-architectures. BDI architectures are primarily focused on \emph{practical
-reasoning}, \ie the process of deciding, step by step, which action to perform
-to reach a goal (as summarized by Woolridge~\cite{Woolridge1999}). The
-management of the interaction between knowledge (the beliefs) and task and plan
-representation and execution (the desires and the intentions) is central, and
-aims at selecting at each step the best subgoal. It becomes then an intention
-that the robot commits to.
-
-This interaction between knowledge and actions is also central to our approach
-(as for any cognitive system): it is one amongst all the activities carried by
-the robot, actually performed by the communication components (that acquire
-desires from interaction with agents) and the execution controller that may
-decide to take an incoming desire into account to create its own internal
-goals. The controller generates and manages intentions from these goals with
-the help of a symbolic task planner, that also has direct access to the
-knowledge base.
-
-This activity is however not the backbone of our architecture. Other
-activities are conducted in parallel, without being explicitly considered as
-desires: assessment of the situation and the environment, dialogue (including
-performative dialogue that can possibly change the internal state of the robot,
-but does not lead directly to the creation of desires, like statement assertion or
-question answering), various background monitoring and recognition tasks, etc.
-
-Regarding the anchoring question, this architecture is bidirectional. The
-components we described provide a \textit{bottom-up} grounding process:
-geometric reasoning and dialogue processing modules constantly build and push
-new symbolic contents about the world to the knowledge base where it becomes
-accessible to decisional layers. In parallel, the knowledge base relies on
-reasoning in a \textit{top-down} way to produce new facts that may in return
-trigger physical behaviours.
-
-\subsection{A Knowledge model}
-
-In this architecture, knowledge manipulation relies on a
-semantic \emph{blackboard}: a central server (the {\sc Oro}
-server~\cite{Lemaignan2010}) stores knowledge as it is produced by each of the
-deliberative components. It conversely exposes a {\tt json}-based RPC
-API~\cite{lemaignan2012kbapi} to query the knowledge base.
-
-Knowledge is represented as RDF triples in the OWL sub-language. Each
-time triples are added or removed from the knowledge base, a Description
-Logics reasoner ({\sc Pellet}\footnote{\tt http://clarkparsia.com/pellet/})
-classifies the whole ontology and inserts all possible inferred triples.
-
-At any time, the knowledge available to the robot may come from three sources:
-{\it a priori} knowledge is stored in an ontology (the {\sc OpenRobots Ontology}),
-loaded when the robot initializes. This static source covers the
+\subsection{Building a human-aware deliberative layer}
+
+Connecting multiple independent software modules in one coherent robotic
+architecture is not only a technical challenge, but also a design and
+architectural challenge.
+
+In particular, dealing with the rich semantics of natural interaction with
+humans has been one of our major focus. To this end, we have been deploying
+over the last years explicit knowledge representation and manipulation in the
+deliberative layer of the robots as a {\it lingua franca} between our
+components.
+
+Figure~\ref{fig|archi} gives an overview of our architecture. An active
+knowledge base ({\sc Oro}), conveniently thought as a \emph{semantic
+blackboard}, connects most of the modules together: the \emph{geometric
+reasoning} module (called {\sc Spark}) produces symbolic assertions (like
+\stmt{BOOK1 isOn TABLE}) describing the state and dynamics of the robot's
+environment. These logical statements are stored in the knowledge base, and
+reused by the language processing module ({\sc Dialogs}), the symbolic task
+planner (HATP) and the execution controller. The output of the language
+processing module, or the activities started by the robot controller are
+likewise stored as symbolic statements.
+
+For instance, when processing a sentence like ``give me another book'', the {\sc
+Dialogs} module sends at some point a request to the knowledge base that looks
+like {\tt \footnotesize find(?book type Book, ?book differentFrom
+BOOK1)}\footnote{We present a similar but complete example at
+section~\ref{moving-london}.} and may write back assertions like \stmt{HUMAN
+desires GIVE\_ACTION45, GIVE\_ACTION45 actsOn BOOK2} to {\sc Oro}. The HATP
+planner then uses the knowledge base to initialize the planning domains with
+similar requests ({\tt \footnotesize find(BOOK2 isAt ?location)}, etc.). The
+execution controller typically monitor conditions (by subscribing to events
+like: {\tt \footnotesize onNewMatch(HUMAN desires ?goal)}) and stores what the
+robot is currently doing (\stmt{myself currentlyPerforms GIVE\_ACTION45}).
+
+To some extend, this architecture moves away from classical layered approaches
+found in robotics~\cite{Gat1998three, Volpe2001CLARAty, Goldberg2002}.
+Interactions between components at the deliberative level are mostly
+bidirectional and we do not introduce layers of abstraction amongst software
+components\footnote{We do have lower-level modules to execute actions or manage
+sensors, but all cognition-related modules live in the same global deliberative
+layer.}. Dialogue processing, for instance, illustrates this structure. This
+component does not purely act as an alternative sensing modality that would be
+fed to a distinct deliberative component: on the contrary, it actively queries
+the knowledge base to interpret and disambiguate natural language, and in that
+regards, it is fully integrated to the deliberative process itself (we detail
+natural language processing at section~\ref{sect|com}).
+
+Our architecture however relates to \emph{Beliefs, Desires, Intentions} (BDI)
+architectures. As put by Woolridge~\cite{Woolridge1999}, BDI architectures are
+primarily focused on \emph{practical reasoning}, \ie the process of deciding,
+step by step, which action to perform to reach a goal. The management of the
+interaction between knowledge (the beliefs) and task and plan representation and
+execution (the desires and the intentions) is central, and aims at selecting at
+each step the best subgoal. It becomes then an intention that the robot commits
+to.
+
+As for any cognitive system, this fundamental interaction between knowledge and
+actions is central to our approach, and involves the dialogue component that
+acquires \emph{desires} from the other agents, the planner and the execution
+controller that may first decide to take or not into account an incoming desire
+as a \emph{goal}, and then generates and manages \emph{intentions} from these
+goals with the help of a symbolic task planner.
+
+Other activities are however conducted in parallel, without being explicitly
+triggered by \emph{desires} in the BDI sense, like situation assessment and
+monitoring or other forms of dialogue (including performative dialogue that can
+possibly change the internal state of the robot, but does not lead directly to
+the creation of \emph{desires}, like assertion of new facts or question
+answering).
+
+\subsection{Knowledge Model}
+
+In our architecture, knowledge manipulation relies on a central server (the {\sc
+Oro} server~\cite{Lemaignan2010}, Figure~\ref{fig|spark-oro} top) which stores
+knowledge as it is produced by each of the deliberative components. It
+conversely exposes a {\tt json}-based RPC API~\cite{lemaignan2012kbapi} to query
+the knowledge base.  We represent knowledge as RDF triples in the OWL
+sub-language. Each time triples are added or removed from the knowledge base, a
+Description Logics reasoner ({\sc Pellet}\footnote{\tt
+http://clarkparsia.com/pellet/}) classifies the whole ontology and inserts all
+possible inferred triples.
+
+This architecture design (a central knowledge base that essentially appears as a
+passive component to the rest of the system -- even though it actually actively
+processes the knowledge pool in the background, for instance with the reasoner)
+departs from other approaches like the CAST model~\cite{Hawes2007} where
+knowledge is represented as a diffuse resource, or the CRAM/KnowRob
+architecture~\cite{Beetz2010} where the knowledge base is an active hub that
+pro-actively queries perceptual components to acquire knowledge. We believe that
+our design leads to a good \emph{observability} (knowledge flows are explicit
+and easy to capture since they are centralized) as well as high modularity
+(modules communicate through a defined and unified API).
+
+At run-time, the knowledge available to the robot at a given time comes from
+three sources. {\it A priori} knowledge is stored in an ontology (the {\sc
+OpenRobots} ontology, discussed below) and loaded at startup. This static source covers the
 \emph{common-sense} knowledge of the robot, and optionally some
-scenario-specific knowledge (for instance about objects that are to be
-manipulated). The second part of the knowledge is acquired at runtime from
-perception, interaction, decision making.  We go over the processes it involves
-in the next sections. Lastly, the third source of knowledge is the output of the
-reasoner.
-
-Contrary to other projects like KnowRob~\cite{Tenorth2009a} that have introduced
+scenario-specific knowledge (for instance, about objects that are to be
+manipulated). The second part of the knowledge is acquired at run-time from
+perception, interaction and planning. The next sections detail these processes.
+Lastly, the third source of symbolic statements is the reasoner itself.
+
+Contrary to other projects like KnowRob~\cite{Tenorth2009a} that relies on
 the concept of \emph{computables} to lazily evaluate/acquire symbolic facts when
@@
 provides the reasoner with the largest possible inference domain. This is of
-special importance for an \emph{event-based} architecture (see
+special importance for an \emph{event-based} architecture like ours (see
 Section~\ref{sect|ctrl}), where, by definition, we do not know beforehand what
-we are ``looking for'', and we prefer instead greedly infer as much as possible
+we are ``looking for'', and we need instead greedly infer as much as possible
 to build a belief state as comprehensive as possible.
@@
 
-\resizebox{\textwidth}{!}{%
+\resizebox{0.8\textwidth}{!}{%
 \begin{tikzpicture}[
@@
 
-\paragraph{The OpenRobots Ontology} The ORO common-sense ontology has been designed from two requirements: covering
-our experimental needs and conforming as much as possible to the {\sc OpenCyc}
-upper ontology.
+\paragraph{The OpenRobots Ontology} The ORO common-sense ontology has been
+designed from two requirements: covering our experimental needs and conforming
+as much as possible to the {\sc OpenCyc} upper ontology.
 
 This leads to a bidirectional design process: from \emph{bottom-up} regarding
-the choices of concepts to model, \emph{top-down} regarding the upper part of the
-taxonomy. This upper part of the ontology is pictured on
-Figure~\ref{fig|upper_tbox}. All the classes visible in this figure belong to the
-{\sc OpenCyc} namespace (the {\tt cyc:} prefix is omitted for clarity).
+the choices of concepts to model, \emph{top-down} regarding the upper part of
+the taxonomy. This upper part of the ontology is pictured on
+Figure~\ref{fig|upper_tbox}. All the classes visible in this figure belong to
+the {\sc OpenCyc} namespace.
 
 Aligning the upper part of the ontology on {\sc OpenCyc} (as done by other
-KRS, like {\sc KnowRob}~\cite{Tenorth2009a} or PEIS K\&R~\cite{Daoutis2009})
-has multiple advantages. First the design of this part of the ontology is
-generally difficult: it pertains to abstract concepts whose mutual relations
-comes to philosophical debates. The upper taxonomy of {\sc OpenCyc} represents
-a relative consensus, at least within the semantic Web community. Then, because
-it is a well established project with numerous links to other on-line databases
-(like Wikipedia or WordNet), the reuse of important {\sc OpenCyc} concepts
-ensures to a certain extend that the knowledge stored by the robot can be
-shared or extended with well-defined semantics. A good example is the concept
-of \emph{Object}: In everyday conversation, an object is a relatively small
-physical thing, that can be typically manipulated. Normally, a human is not
-considered as an object. In {\sc Cyc}, an object has a more precise definition:
-it is something \emph{partially tangible}. That includes obviously the humans,
-and actually many other entities that would not be commonly said to be objects
-(the Earth for instance). Thus the importance of relying on well-defined and
-standard semantics to exchange informations between artificial systems.
+knowledge representation systems, like {\sc KnowRob}~\cite{Tenorth2009a} or PEIS
+K\&R~\cite{Daoutis2009}) has multiple advantages. First the design of this part
+of the ontology is generally difficult: it pertains to abstract concepts whose
+mutual relations comes to philosophical debates. The upper taxonomy of {\sc
+OpenCyc} represents a relative consensus, at least within the semantic Web
+community. Then, because it is a well established project with numerous links to
+other on-line databases (like Wikipedia or WordNet), the reuse of important {\sc
+OpenCyc} concepts ensures to a certain extend that the knowledge stored by the
+robot can be shared or extended with well-defined semantics. A good example is
+the concept of \emph{Object}: In everyday conversation, an object is a
+relatively small physical thing, that can be typically manipulated. Normally, a
+human is not considered as an object. In {\sc Cyc}, an object has a more precise
+definition: it is something \emph{partially tangible}. That includes obviously
+the humans, and actually many other entities that would not be commonly said to
+be objects (the Earth for instance). Thus the importance of relying on
+well-defined and standard semantics to exchange informations between artificial
+systems.
 
@@
 in the ORO model between \emph{temporal} and \emph{spatial} entities (formally,
-$(TemporalThing \sqcap SpatialThing)^{\mathcal{I}} = \emptyset$, with
+$(\text{TemporalThing} \sqcap \text{SpatialThing})^{\mathcal{I}} = \emptyset$, with
 $\mathcal{I}$ the \emph{interpretation} of our model).
 
-The class \concept{purposeful action} is the superset of all the actions that
-are voluntarily performed by the robot (or another agent). Subclasses (like
-\concept{Give}, \concept{LookAt}, etc.) are not asserted in the common-sense
-ontology, but are added by the execution controller (in link with the symbolic
-task planner) and the natural language processor based on what is actually
-performable and/or understandable by the robot at run-time.
-
-The tree in Figure~\ref{fig|upper_tbox} (this subset of the ontology is indeed
-a tree: this has however not to be the case in general, and, as a matter of
-fact, the TBox of the whole ORO common-sense ontology does not form a tree) is
-not equally developed at lower levels. We have already briefly mentioned the
-developments of the actions. The other important part is the descendants of the
-\concept{partially tangible thing} (what is commonly called an \emph{object}).
-Figure~\ref{fig|tangible_things_tbox} gives more details on this part of the
-ontology.
+The class \concept{PurposefulAction} is the superset of all the actions that are
+voluntarily performed by the robot (or another agent). Actual actions (\ie
+subclasses of \concept{PurposefulAction} like \concept{Give} or
+\concept{LookAt}) are not initially asserted in the common-sense ontology. They
+are instead added at run-time by the execution controller (in link with the
+symbolic task planner) and the natural language processor based on what is
+actually performable and/or understandable in the scenario context (\ie the
+current robot configuration and the actions required by the scenario).
+
+The tree in Figure~\ref{fig|upper_tbox}\footnote{This subset of the ontology is
+indeed a tree. This has however not to be the case in general, and, as a matter
+of fact, the TBox of the whole ORO common-sense ontology does not form a tree}
+is not equally developed in each directions at lower levels. The descendants of
+\concept{PartiallyTangibleThing} (\ie what we commonly call \emph{objects}), for
+instance, are pictured in Figure~\ref{fig|tangible_things_tbox}. This excerpt
+from the ontology makes the bottom-up design process clear: only few types of
+\emph{partially tangible} things appear, and those are only subclasses relevant
+to the context of service robotics in an human-like environment. For performance
+reasons as well as clarity, we do not attempt to proactively extend the
+conceptual coverage of the field. We instead opportunistically extend the
+common-sense knowledge when required by experiments.
 
@@
 
-\resizebox{0.7\textwidth}{!}{%
+\resizebox{0.5\textwidth}{!}{%
 \begin{tikzpicture}[
@@
 }
-    \caption{TBox of the specialisations of \concept{PartiallyTangible}.}
+    \caption{TBox of the specialisations of \concept{PartiallyTangibleThing}.}
     \label{fig|tangible_things_tbox}
@@
 
-This excerpt from the ontology makes the bottom-up design process visible: only
-few types of \emph{partially tangible} things appear, and only subclasses
-relevant to the context of service robotics in an human-like environment are
-present. For performances and ease of use, we do not attempt to comprehensively
-cover the field. We opportunistically extend the common-sense knowledge when
-required by experiments.
-
 Lastly, the ORO common-sense ontology contains several rules and class
-expressions that encode non-trivial inferences.
-
-The definition of the \concept{Bottle} is a typical example:
+expressions that encode non-trivial inferences. The definition of
+\concept{Bottle} as found in the ORO ontology is a typical example:
 
@@
 If a human informs the robot that a given object is indeed a bottle, the robot
-can then infer much more on this object. And if the human affirms that a car is
-a bottle, the robot may question this assertion because of the inconsistent
-size.
-
-We discuss in section~\ref{krs-discussion} the strengths and weaknesses of our
-knowledge framework.
-
-\subsection{Internal cognitive processes}
+can then derive a few more informations on this object. And, conversely, if the
+human affirms that a car is a bottle, the robot may question this assertion
+because of the inconsistent size (the {\sc Dialogs} module relies on such
+logical consistency checks when processing natural language inputs, both to
+ensure that the verbal input has been correctly acquired and to verify that what
+the human says does actually make sense).
+
+We will come back and discuss in section~\ref{krs-discussion} the strengths and
+weaknesses of this knowledge framework.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Implementing the Cognitive Skills}
+\label{sec:impl}
+
+Previous section provided an overview of our cognitive architecture for robots
+involved in interactions with other social agents, as well as the knowledge
+model we rely on.
+
+We now introduce each of the components and give insights on their main
+features.
+
+\subsection{Internal Cognitive Processes}
 \label{sect|intern}
+
+\subsubsection{Symbolic Anchoring}
+
+TDB
+
+Regarding the anchoring question, this architecture is bidirectional. The
+components we described provide a \textit{bottom-up} grounding process:
+geometric reasoning and dialogue processing modules constantly build and push
+new symbolic contents about the world to the knowledge base where it becomes
+accessible to decisional layers. In parallel, the knowledge base relies on
+reasoning in a \textit{top-down} way to produce new facts that may in return
+trigger physical behaviours.
 
@@
 this feature to trace certain knowledge-related processes.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Implementing the Cognitive Skills}
-\label{sec:impl}
-
-Previous section provided an overview of our cognitive architecture for robots
-involved in interactions with other social agents, as well as the knowledge
-model we rely on.
-
-We now introduce each of the components and give insights on their main
-features.
 
@@
 Our other controller, {\sc Shary}, relies on the symbolic task planner to
-generate at runtime suitable sets of atomic actions.
+generate at run-time suitable sets of atomic actions.
 
@@
 \caption{Main {\sc pyRobots} \emph{actions}, sorted by categories. These
-actions are combined at runtime into higher-level \emph{tasks}.}
+actions are combined at run-time into higher-level \emph{tasks}.}
 
@@
 \subsection{\emph{Moving to London} scenario}
+\label{moving-london}
 

### 2014-05-02 paper.tex
Re-organization sect 2 <-> sect 3
- added a sub section on reasoning
- added a dedicated sub section on symbol grounding
- moved 'internal cognitive skills' to sect 3

@@
 
+\subsection{Symbol Grounding}
+
+\emph{Grounding} (also called \emph{anchoring} when specifically referring to
+the building of links between \emph{percepts} and \emph{physical
+objects}~\cite{Coradeschi2003}) is the task consisting in building and
+maintaining a bi-directional link between sub-symbolic representations (sensors
+data, low-level actuation) and symbolic representations that can be manipulated
+and reasoned about~\cite{Harnad1990}. This represents an important cognitive
+skill, in particular in the human-robot interaction context: in that case, the
+link that the robot has to established between its percepts and its symbol must
+\emph{model} to a large extend the human representations in order to effectively
+support communication.
+
+\emph{Symbol grounding} relates the knowledge model to the perception and
+actuation capabilities of the robot, and our architecture adopts here a
+bidirectional approach.  The different components that we have described so far
+provide a \textit{bottom-up} grounding process: geometric reasoning and dialogue
+processing modules constantly build and push new symbolic contents about the
+world to the knowledge base where it becomes accessible to decisional layers. In
+parallel, the knowledge base relies on reasoning in a \textit{top-down} way to
+produce new facts, \ie new beliefs that may lead to actual decision making and
+physical actions.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Implementing the Cognitive Skills}
+\section{Cognitive Skills}
 \label{sec:impl}
 
-Previous section provided an overview of our cognitive architecture for robots
-involved in interactions with other social agents, as well as the knowledge
-model we rely on.
-
-We now introduce each of the components and give insights on their main
-features.
-
-\subsection{Internal Cognitive Processes}
+While the previous section provided an overview of our cognitive architecture,
+as well as its associated knowledge model, this section discuss each of its main
+``building blocks'' that we each consider to instantiate different cognitive
+skills. Since each of these components have multiple connections to the others,
+no obvious presentation order stands out, and the reader may have to rely on
+Figure~\ref{fig|archi} as a guide.
+
+We start by discussing the main \emph{internal} cognitive capabilities,
+implemented in the knowledge base itself, and then discuss successively {\sc
+Spark}, the situation assessment module, {\sc Dialogs}, the dialogue processor,
+HATP, the symbolic task planner, and finally the main features of our execution
+controllers.
+
+\subsection{Internal Cognitive Skills}
 \label{sect|intern}
 
-\subsubsection{Symbolic Anchoring}
-
-TDB
-
-Regarding the anchoring question, this architecture is bidirectional. The
-components we described provide a \textit{bottom-up} grounding process:
-geometric reasoning and dialogue processing modules constantly build and push
-new symbolic contents about the world to the knowledge base where it becomes
-accessible to decisional layers. In parallel, the knowledge base relies on
-reasoning in a \textit{top-down} way to produce new facts that may in return
-trigger physical behaviours.
+We call \emph{internal} those cognitive capabilities that are tightly bound to
+the knowledge model, and hence implemented directly inside {\sc Oro}, the knowledge base.
+We present here three of them, \emph{reasoning}, \emph{theory of mind} modeling
+and our (naive) approach to \emph{memory management}.
+
+\subsubsection{Reasoning}
+
+As mentioned in the previous section, we use the Pellet open-source reasoner to
+reason on the knowledge base. This enables several standard inference services:
+\emph{consistency checking}, \emph{concept satisfiability},
+\emph{classification} and \emph{realisation} (computation of the most specific
+classes that an individual belongs to). In case of logical inconsistency, the
+reasoner can also provide \emph{explanations} that are important for debugging.
+
+Besides those, {\sc Oro} server implements several algorithms to identify
+similarities and differences between concepts (classes or
+instances)~\cite{Ros2010b}: the \emph{Common Ancestors} algorithm, useful to
+determine the most specific class(es) that include a given set of individuals;
+the \emph{First Different Ancestors} algorithm that returns what can be
+intuitively understood as the most generic types that \emph{differentiate} two
+concepts; and \emph{clarification} and \emph{discrimination} algorithms that
+play a key role in the process of \emph{interactive grounding} of the semantics
+of concepts (we discuss this process at section~\ref{sect|com}). Clarification
+and discrimination algorithms are based on what we call \emph{descriptors}, \ie
+properties of individuals, either statically asserted in the common-sense
+ontology, acquired by the robot through perception or proactive questioning of
+the human partner, or derived from other reasoning algorithms like the
+\emph{Common Ancestors} and \emph{Different Ancestors}. The
+\emph{discrimination} algorithm consists then in looking for discriminants, \ie
+descriptors that allow a maximum discrimination among a set of individuals.
 
@@
 Theory of Mind (originally defined in~\cite{Premack1978}) is the cognitive
-ability that a subject possesses to represent the mental state of another
+ability that allows a subject to represent the mental state of another
 agent, possibly including knowledge that contradicts the subject's own model: for
 example, a book can be at the same time \emph{visible} for myself, and \emph{not
-visible} for you.
-
-Children develop this skill, which is essential to understand others' perspectives during
-interactions, around the age of three. It supposes the
-ability to build, store and retrieve separate models of the knowledge of the
-interactors.
-
-Our knowledge base implements such a mechanism: when the robots infers a new
-agent has been introduced in the knowledge base, it initializes a new,
+visible} for you. Children develop this skill, which is essential to understand others' perspectives during
+interactions, around the age of three.
+
+From a robotics point of view, it supposes the ability to build, store and
+retrieve separate models of the beliefs of the agents the robot interacts with.
+Our knowledge base implements such a mechanism: when the robots recognises that
+a new agent has been created in the knowledge base, it initializes a new,
 independent, ontology for this agent. All the ontologies that are created share
 the same common-sense knowledge, but rely on each agent's perspective for the
-actual instantiation: the robot (geometrically) computes that the book is in
-its own field of view, but not in the human one. The robot knowledge contains
-the fact \stmt{book isVisible true} while the human model contains \stmt{book
-isVisible false}.
+actual instantiation: the robot (geometrically) computes that the book is in its
+own field of view, but not in the human one. The robot knowledge contains the
+fact \stmt{book isVisible true} while the human model contains \stmt{book
+isVisible false} (this computation, called \emph{perspective taking}, is
+discussed in the next section).
 
@@
 mark for a few seconds important concepts that are currently manipulated by the
-robot as \emph{active concepts}. For example, if a human asks the robot: ``Give
+robot as \emph{active concepts}: If a human asks the robot ``Give
 me all red objects'', the human, the \concept{Give} action, and each red

### 2014-05-02 paper.tex
Polished the rendering of HATP plans

@@
 
-    \node (h1) {TAKE\\{\tiny \it \sf GREY\_TAPE, TABLE}};
-    \node[right=of h1] (h2) {PUT\\{\tiny\it\sf GREY\_TAPE, BIN1 }} edge (h1);
-
-    \node[robot, below=1.5 of h1] (r1) {TAKE\\{\tiny \it \sf BLACK\_TAPE, TABLE }};
-    \node[robot, right=of r1] (r2) {PUT\\{\tiny \it \sf BLACK\_TAPE, BIN2 }} edge (r1);
-    \node[robot, right=of r2] (r3) {TAKE\\{\tiny \it \sf BOTTLE, TABLE }} edge (r2);
-    \node[robot, right=of r3] (r4) {PUTRV\\{\tiny \it \sf BOTTLE, TABLE}} edge (r3);
-    \node[robot, right=of r4] (r5) {TAKE\\{\tiny \it \sf BOOK, TABLE }} edge (r4);
-    \node[robot, right=of r5] (r6) {PUT\\{\tiny \it \sf BOOK, BIN2 }} edge (r5);
-
-    \node[right=9cm of h2] (h3) {TAKE\\{\tiny \it \sf BOTTLE, TABLE}} edge (h2) edge (r4);
-    \node[right=of h3] (h4) {PUT\\{\tiny \it \sf BOTTLE, BIN1}} edge (h3);
+    \node[font=\rm\large, draw=none, fill=none, rotate=90] at (-2, 0) {Human\\actions};
+    \node[font=\rm\large, draw=none, fill=none, rotate=90] at (-2, -3.5) {Robot\\actions};
+
+    \node (h1) {\action{TAKE}{GREY\_TAPE}{TABLE}};
+    \node[right=of h1] (h2) {\action{PUT}{GREY\_TAPE}{BIN1 }} edge (h1);
+
+    \node[robot, below=1.5 of h1] (r1) {\action{TAKE}{BLACK\_TAPE}{TABLE }};
+    \node[robot, right=of r1] (r2) {\action{PUT}{BLACK\_TAPE}{BIN2 }} edge (r1);
+    \node[robot, right=of r2] (r3) {\action{TAKE}{BOTTLE}{TABLE }} edge (r2);
+    \node[robot, right=of r3] (r4) {\action{PUTRV}{BOTTLE}{TABLE}} edge (r3);
+    \node[robot, right=of r4] (r5) {\action{TAKE}{BOOK}{TABLE }} edge (r4);
+    \node[robot, right=of r5] (r6) {\action{PUT}{BOOK}{BIN2 }} edge (r5);
+
+    \node[right=9cm of h2] (h3) {\action{TAKE}{BOTTLE}{TABLE}} edge (h2) edge (r4);
+    \node[right=of h3] (h4) {\action{PUT}{BOTTLE}{BIN1}} edge (h3);
 
@@
 
-    \node (h1) {TAKE\\{\tiny \it \sf GREY\_TAPE, TABLE}};
-    \node[right=of h1] (h2) {PUT\\{\tiny\it\sf GREY\_TAPE, BIN1 }} edge (h1);
-
-    \node[robot, below=1.5 of h1] (r3) {TAKE\\{\tiny \it \sf BOTTLE, TABLE }};
-    \node[robot, right=of r3] (r4) {PUTRV\\{\tiny \it \sf BOTTLE, TABLE}} edge (r3);
-    \node[robot, right=of r4] (r1) {TAKE\\{\tiny \it \sf BLACK\_TAPE, TABLE }} edge (r4);
-    \node[robot, right=of r1] (r2) {PUT\\{\tiny \it \sf BLACK\_TAPE, BIN2 }} edge (r1);
-    \node[robot, right=of r2] (r5) {TAKE\\{\tiny \it \sf BOOK, TABLE }} edge (r2);
-    \node[robot, right=of r5] (r6) {PUT\\{\tiny \it \sf BOOK, BIN2 }} edge (r5);
-
-    \node[right=of h2] (h3) {TAKE\\{\tiny \it \sf BOTTLE, TABLE}} edge (h2) edge (r4);
-    \node[right=of h3] (h4) {PUT\\{\tiny \it \sf BOTTLE, BIN1}} edge (h3);
+    \node[font=\rm\large, draw=none, fill=none, rotate=90] at (-2, 0) {Human\\actions};
+    \node[font=\rm\large, draw=none, fill=none, rotate=90] at (-2, -3.5) {Robot\\actions};
+
+    \node (h1) {\action{TAKE}{GREY\_TAPE}{TABLE}};
+    \node[right=of h1] (h2) {\action{PUT}{GREY\_TAPE}{BIN1 }} edge (h1);
+
+    \node[robot, below=1.5 of h1] (r3) {\action{TAKE}{BOTTLE}{TABLE }};
+    \node[robot, right=of r3] (r4) {\action{PUTRV}{BOTTLE}{TABLE}} edge (r3);
+    \node[robot, right=of r4] (r1) {\action{TAKE}{BLACK\_TAPE}{TABLE }} edge (r4);
+    \node[robot, right=of r1] (r2) {\action{PUT}{BLACK\_TAPE}{BIN2 }} edge (r1);
+    \node[robot, right=of r2] (r5) {\action{TAKE}{BOOK}{TABLE }} edge (r2);
+    \node[robot, right=of r5] (r6) {\action{PUT}{BOOK}{BIN2 }} edge (r5);
+
+    \node[right=of h2] (h3) {\action{TAKE}{BOTTLE}{TABLE}} edge (h2) edge (r4);
+    \node[right=of h3] (h4) {\action{PUT}{BOTTLE}{BIN1}} edge (h3);
 
@@
                         inner sep=0.1cm}]
+
+    \node[font=\rm\large, draw=none, fill=none, rotate=90] at (-2, 0) {Human\\actions};
+    \node[font=\rm\large, draw=none, fill=none, rotate=90] at (-2, -3.5) {Robot\\actions};
 

### 2014-05-05 rebuttal.tex
Rebuttal letter: general comments and reviewer 3

@@
-\documentclass{letter}
-\usepackage{letterbib}
+\documentclass{article}
 \usepackage[utf8]{inputenc}
-\usepackage{hyperref}
-\usepackage{booktabs}
 
-\signature{Séverin Lemaignan}
 \begin{document}
 
-\subsection{General Comments}
+\section{General Comments}
 
 \emph{1. The novelty of this paper with respect to other papers by the same
-authors, including one in IJSR 2011, has been questioned. The authors
-should convincingly explain what are the novel contributions and the
-added value of this work compared to the previous publications.}
+authors, including one in IJSR 2011, has been questioned. The authors should
+convincingly explain what are the novel contributions and the added value of
+this work compared to the previous publications.}
 
@@
 
-We detail below the steps that were taken to improve the overall quality of the
-paper (note that minor typos/style issues reported by the reviewers have been
-fixed directly in the manuscript, and are not discussed here anymore).
+We would like first to thank the reviewers for their thoughtful review of our
+work. The manuscript underwent a major rewrite, impacting both the overall style
+(which should now be much more pleasant to read) and the whole organisation of
+the paper (following the reviewers' recommendations). We also dedicated time to
+redraw the figures in an improved academic style.
 
-\subsection{Comments from Reviewer 1}
+We address hereafter each of the specific concerns raised by the reviewers (note
+that minor typos/style issues have been fixed directly in the manuscript, and
+are not discussed here).
+
+\section{Comments from Reviewer 1}
 
@@
 author names in the references would be more pleasant to read since it saves
-reviewers from jumping back and forth all the time.
-
-[EiC note: please use the standard AIJ referencing style -- we do not have the
-liberty to change this -- thanks]}
+reviewers from jumping back and forth all the time.}
 
@@
 
-\subsection{Comments from Reviewer 2}
+\section{Comments from Reviewer 2}
 
@@
     \item \emph{pg 10 (section 3.2.1)
-
             "Note that, while the system benefits from the complementary
@@
 
+\section{Comments from Reviewer 3}
+
+The main concern of Reviewer 3 relates to the lack of originality of this
+contribution.
+
+The reviewer references two previous publications by the authors. \emph{Grounding
+the Interaction: Anchoring Situated Discourse in Everyday Human-Robot
+Interaction} has indeed been published in 2011 in the \emph{International Journal of
+Social Robotics}. However, the overlap with this article appears to be minimal. The
+discussion on natural language grounding, which is the focus of the previously
+published article, is only briefly summarized in the present article (section
+3.3) and as far as we can tell, only one picture, used as an example, as been
+reused.
+
+The second publication, \emph{When the robot considers the human...} is supposed
+to appear in the proceedings of the \emph{2011 International Symposium on
+Robotics Research}, but has yet to be actually pusblished.  While this article
+indeed presents our architecture in a way similar to our proposed contribution,
+the fact that it belongs to the proceedings of a conference and that it has
+actually not yet been published should make it mostly irrelevant.
+
+The reviewer 3 also expresses concerns regarding our proposal not \emph{``helping
+researchers to build upon each other's work''} because of the way we
+\emph{``presents our design choices''}. We understand this remark has being
+related to the comment by Reviewer 2 regarding us not presenting clearly
+enough the key issues of human-robot interaction and consequently, the
+challenges that HRI brings to artificial intelligence. As explained above, we
+have hopefully addressed this concern in the current version of the paper.
+
+Generally speaking, we would like to underline that, while we acknowledge that a
+lot of material has been already published on the different sub-systems of
+our architecture (as rightfully noted by Reviewer 1), this article give for the
+first time a complete picture of the system. This offers, in our opinion, a
+perspective on the state of the art in cognitive robotic architectures
+for human-robot interaction that may be of interest for the AI and robotics
+research communities.
+
 \end{document}

### 2014-05-05 paper.tex
Clarified even more how purposeful actions are added
Should close #18

@@
 The class \concept{PurposefulAction} is the superset of all the actions that are
-voluntarily performed by the robot (or another agent). Actual actions (\ie
+purposefully performed by the robot (or another agent). Actual actions (\ie
 subclasses of \concept{PurposefulAction} like \concept{Give} or
@@
 are instead added at run-time by the execution controller (in link with the
-symbolic task planner) and the natural language processor based on what is
-actually performable and/or understandable in the scenario context (\ie the
-current robot configuration and the actions required by the scenario).
+symbolic task planner) and the natural language processor based on what the
+robot is actually able to perform and/or to interpret in the current context
+(\ie the current robot configuration and the actions required by the scenario).
+The set of actions that the robot can interpret typically corresponds to the
+planing domain in use: the robot can only process the \concept{GoTo} action if
+the corresponding task description with its pre- and post-conditions is
+available to the symbolic task planner.
 

### 2014-05-05 paper.tex
Completed discussion on memory
Closes #20

@@
 \subsubsection{Working memory}
-
-The {\sc Oro} server also features a mechanism to mimic minimalistic forms of
-memory families.  When new statements are inserted in the knowledge base, a
-\emph{memory profile} is optionally attached to them.
-
-Three such profiles are predefined: {\tt short term}, {\tt episodic} and {\tt
-long term}. They are currently attached to different lifetime for the statements
-(respectively 10 seconds, 5 minutes and no time limit). After this duration,
-the statements are automatically removed from the knowledge base.
+\label{memory}
+
+Memory has been studied at length in the cognitive psychology and
+neuropsychology communities: very briefly, the idea of \emph{short-term} and
+\emph{long-term} memory is due to Atkinson and Shiffrin~\cite{Atkinson1968} ,
+Anderson~\cite{Anderson1976} splits memory into \emph{declarative} (explicit)
+and \emph{procedural} (implicit) memories, Tulving~\cite{Tulving1985} organises
+the concepts of \emph{procedural}, \emph{semantic} and \emph{episodic} memories
+into a hierarchy. Short-term memory is eventually refined with the concept of
+\emph{working memory} by Baddeley~\cite{Baddeley2010}
+
+It is worth emphasising that if memory is commonly associated to the process of
+forgetting facts after a variable amount of \emph{time}, it actually covers
+more mechanisms that are relevant to robotics, like selective remembering
+triggered by a specific context or reinforcement learning.
+
+The {\sc Oro} server features a mechanism to mimic only minimalistic forms of memory
+families.  When new statements are inserted in the knowledge base, a
+\emph{memory profile} is optionally attached to them.  Three such profiles are
+predefined: {\tt short term}, {\tt episodic} and {\tt long term}. They are
+currently attached to different lifetime for the statements (respectively 10
+seconds, 5 minutes and no time limit). After this duration, the statements are
+automatically removed from the knowledge base.
 
@@
 this feature to trace certain knowledge-related processes.
+
+In the larger field of cognitive architectures, the {\sc Soar}
+architecture~\cite{Lehman2006} is one of those that tries to reproduce a
+human-like memory organisation. The GLAIR cognitive
+architecture~\cite{Shapiro2009} also has a concept of long term/short term and
+episodic/semantic memories.
 

### 2014-05-05 paper.tex
Explicit the knowledge dynamics management
Closes #4

@@
 Oro} server~\cite{Lemaignan2010}, Figure~\ref{fig|spark-oro} top) which stores
-knowledge as it is produced by each of the deliberative components. It
-conversely exposes a {\tt json}-based RPC API~\cite{lemaignan2012kbapi} to query
-the knowledge base.  We represent knowledge as RDF triples in the OWL
-sub-language. Each time triples are added or removed from the knowledge base, a
-Description Logics reasoner ({\sc Pellet}\footnote{\tt
-http://clarkparsia.com/pellet/}) classifies the whole ontology and inserts all
-possible inferred triples.
+knowledge as it is produced by each of the deliberative components (the
+clients). It exposes a {\tt json}-based RPC API~\cite{lemaignan2012kbapi} to
+query the knowledge base. The clients of the {\sc Oro} server are in charge of
+managing themselves the knowledge (when to add, when to update, when to
+retract knowledge): no meta-semantics is carried over that would let the server
+manage these dynamics itself.\footnote{There is one exception: so-called
+\emph{memory profiles} let the server discard facts after a specific period
+of time. We present this experimental feature at section~\ref{memory}.} We
+represent knowledge as RDF triples in the OWL sub-language. Each time
+triples are added or removed from the knowledge base, a Description Logics
+reasoner ({\sc Pellet}\footnote{\tt http://clarkparsia.com/pellet/})
+classifies the whole ontology and inserts all possible inferred triples.
 

### 2014-05-05 rebuttal.tex
Added more content to the rebuttal letter

@@
 
+    \item \emph{p9: Instead of storing these models separately (having to
+    update each
+of them), wouldn't it be easier to compute them once they are needed, and
+to maintain a quasi-global world belief?} This is a design choice that trades
+off scalability for complete domain explicitation. We have clarified and
+justified this design choice in section 2.2 (\emph{Knowledge model}). Also, in
+the specific case of agent-dependent models, we want to represent
+false-beliefs situations (where one agent carries wrong beliefs on the world)
+that can typically not be computed ``on demand'' because they are observable
+only in specific situations.
+
     \item \emph{p10 l 49: How is this discrimination done?} The discrimination
@@
 
+    \item \emph{p15 While the production of such rules is one thing, managing the
+    dynamics of this knowledge and e.g. removing it once it's outdated or updating
+once it has changed are crucial. This should be explained in more detail.}
+The dynamics of the knowledge are managed by the clients (ie, the deliberative
+components like the situation assessment module, the dialogue processing module,
+etc.) of the knowledge base themselves. In particular, the situation assessment
+module recomputes at every step the whole set of spatial relations, and update
+accordingly the knowledge base. This has been clarified in both section 2.2
+(\emph{Knowledge model}) and section 3.2 (\emph{Situation Assessment, Human
+Awareness, Perspective Taking})
+
     \item \emph{The example scenarios have very bold names for often rather
@@
         \emph{code names} for the experiments' names. These have been reworded.
+
+    \item \emph{p 20: AFAIK KnowRob stores OWL as facts in Prolog, why
+        should the TBOX be immutable?} The reviewer is right insofar as there is
+        not limitations inherent to Prolog in that regard. We have removed the
+        reference to KnowRob TBOX, that originated in a discussion with KnowRob's author who told
+        us then that KnowRob could not update at run-time its TBox. Since we were
+        not able to find litterature confirming this fact, it is likely better
+        not to mention it.
+
 \end{itemize}
@@
             updated to clarify the concepts.
+
+    \item \emph{In the current form, Section 2.3 seems outside the scope of the
+            paper. The discussion on working memory is quite interesting, but
+            since it is still to be developed in detail it maybe a topic for
+            future perspectives. In that case it would be nice to read which AI
+            techniques could help tackling the issue.} The presentation of
+            ``internal cognitive skills'' has been extended, refined and integrated
+            to section 3. Specifically, several litterature references have been
+            added to the discussion on memory, both on the theoritical side and on
+            existing implementations.
+
+    \item \emph{``Subclasses (like Give, LookAt, etc.) are not asserted in the
+            common-sense ontology, but are added by the execution controller (in
+            link with the symbolic task planner) and the natural language
+            processor based on what is actually performable and/or
+            understandable by the robot at run-time'' This is very difficult to
+            understand. Provide more details and, possibly, an example} The paragraph
+            has been rephrased and illustrated to be easier to understand.
+
 \end{itemize}

### 2014-05-05 paper.tex
Moved the 'key issues' stub to the discussion
Clarify what 'cognitive skills' are + style of intro sec 3

@@
 robot, ranging from teammate behaviour to assistant or proactive helper.
-
-\subsection{The Key Issues}
-
-TDB
 
@@
 
-While the previous section provided an overview of our cognitive architecture,
-as well as its associated knowledge model, this section discuss each of its main
-``building blocks'' that we each consider to instantiate different cognitive
-skills. Since each of these components have multiple connections to the others,
-no obvious presentation order stands out, and the reader may have to rely on
-Figure~\ref{fig|archi} as a guide.
-
-We start by discussing the main \emph{internal} cognitive capabilities,
-implemented in the knowledge base itself, and then discuss successively {\sc
-Spark}, the situation assessment module, {\sc Dialogs}, the dialogue processor,
-HATP, the symbolic task planner, and finally the main features of our execution
+While the previous section provided an overview of our cognitive architecture
+as well as its associated knowledge model, this section discusses each of its main
+``building blocks''. Each of these components may have connections to several
+others, and we invite the reader to refer to Figure~\ref{fig|archi} as a guide
+if needed.
+
+\begin{inparaenum} In our context, we call \emph{cognitive skills} the
+deliberative behaviours that are \item \emph{stateful} both in terms of
+\emph{time} (keeping track of previous states is required for the component
+to perform adequately) and \emph{modalities} (for instance, \emph{situated}
+dialogue requires interleaving different communication modalities), \item
+manipulate \emph{explicit semantics}, typically by the mean of symbolic
+reasoning, \item operate at the \emph{human-level}, \ie are legible to the
+humans, typically by acting at similar levels of abstraction.\end{inparaenum}
+
+We start by discussing first the main \emph{internal} cognitive capabilities,
+implemented in the knowledge base itself, and then discuss successively the
+situation assessment module {\sc Spark}, the dialogue processor {\sc Dialogs},
+the symbolic task planner HATP, and finally the main features of our execution
 controllers.
@@
 We call \emph{internal} those cognitive capabilities that are tightly bound to
-the knowledge model, and hence implemented directly inside {\sc Oro}, the knowledge base.
-We present here three of them, \emph{reasoning}, \emph{theory of mind} modeling
+the knowledge model, and hence implemented directly within the {\sc Oro} server.
+We present here three of them: \emph{reasoning}, \emph{theory of mind} modeling
 and our (naive) approach to \emph{memory management}.
@@
 what remains to be developed/explored in the future.
+
+\subsection{The Key Issues}
+
+We would like first to restate the main challenges to artificial intelligence
+raised by human-robot interaction.
+
+\fxfatal{TDB}
 

### 2014-05-05 paper.tex
Start to rework discussion
Spread out some parts to other places in the paper,
Merged sections as suggested by reviewer 1, cf #15

@@
 
-    \caption{The upper part of the ORO common-sense TBox. All these concepts
+\caption{The upper part of the {\sc Oro} common-sense TBox. All these concepts
     belong to the {\sc OpenCyc} namespace.}
@@
 
-\paragraph{The OpenRobots Ontology} The ORO common-sense ontology has been
+\paragraph{The OpenRobots Ontology} The {\sc Oro} common-sense ontology has been
 designed from two requirements: covering our experimental needs and conforming
@@
 Figure~\ref{fig|upper_tbox} also illustrates the fundamental disjunction
-in the ORO model between \emph{temporal} and \emph{spatial} entities (formally,
+in the {\sc Oro} model between \emph{temporal} and \emph{spatial} entities (formally,
 $(\text{TemporalThing} \sqcap \text{SpatialThing})^{\mathcal{I}} = \emptyset$, with
@@
 indeed a tree. This has however not to be the case in general, and, as a matter
-of fact, the TBox of the whole ORO common-sense ontology does not form a tree}
+of fact, the TBox of the whole {\sc Oro} common-sense ontology does not form a tree}
 is not equally developed in each directions at lower levels. The descendants of
@@
 
-Lastly, the ORO common-sense ontology contains several rules and class
+Lastly, the {\sc Oro} common-sense ontology contains several rules and class
 expressions that encode non-trivial inferences. The definition of
-\concept{Bottle} as found in the ORO ontology is a typical example:
+\concept{Bottle} as found in the {\sc Oro} ontology is a typical example:
 
@@
 Anchoring perceptions in a symbolic model requires perception abilities and
-their symbolic interpretation. We rely on a dedicated geometric and temporal
-reasoning module called {\sc Spark} (\emph{SPAtial Reasoning \&
-Knowledge}~\cite{Sisbot2011}). It is a situation assessment reasoner that
-generates symbolic knowledge from the geometry of the environment with respect
-to relations between objects, robots and humans (Figures~\ref{fig|spark-oro} and~\ref{fig:sparkSubfigures}),
-also taking into account the different perspective that each agent has on the
-environment.
+their symbolic interpretation. We call \emph{physical situation assessment} the
+cognitive skill that a robot exhibits when it assesses the nature and content of its
+surroundings and efficiently monitors its evolution.
+
+Numerous approaches exist, like amodal \emph{proxies}~\cite{Jacobsson2008},
+grounded amodal representations~\cite{Mavridis2006}, semantic maps
+(\cite{Nuechter2008, Galindo2008,Blodow2011}) or affordance-based planning and
+object classification~\cite{Lorken2008, Varadarajan2011}.
+
+%Spatial reasoning~\cite{O'Keefe1999} is a field in its own right, and has been
+%used for natural language processing for applications such as direction
+%recognition ~\cite{Kollar2010,Matuszek2010} or language
+%grounding~\cite{Tellex2010}.~\cite{Skubic2004} present a spatial reasoner
+%integrated in a robot which computes symbolic positions of objects.
+
+We rely on a dedicated geometric and temporal reasoning module called {\sc
+Spark} (\emph{SPAtial Reasoning \& Knowledge}~\cite{Sisbot2011}). It is a
+situation assessment reasoner that generates symbolic knowledge from the
+geometry of the environment with respect to relations between objects,
+robots and humans (Figures~\ref{fig|spark-oro}
+and~\ref{fig:sparkSubfigures}), also taking into account the different
+perspective that each agent has on the environment.
 
@@
 
+%Our approach, implemented in the {\sc Spark} module, is a grounded amodal
+%representation of the robot environment, where different sensing modalities
+%(vision-based fiducial markers tracking, RGBD human tracking, motion capture)
+%are merged and where amodal spatial reasoning takes place (for instance,
+%explanation generation for unexpected perceptions like a disappearing object).
+
 From an artificial intelligence perspective, {\sc Spark} takes care of merging
@@
 
-SPARK is designed to run and continously update the knowledge base (in our
+{\sc Spark} is designed to run and continously update the knowledge base (in our
 experiments, it runs at about 10Hz). At each steps, it re-computes spatial
@@
 
-\paragraph{Symbolic facts production}
+\paragraph{Symbolic Facts Production}
 
@@
 
-\paragraph{Hypotheses on objects states and positions}
+\paragraph{Hypotheses on Objects States and Positions}
 
@@
 
-\paragraph{Primitive action recognition}
+\paragraph{Primitive Action Recognition}
 
@@
 
-\subsubsection{Ego-centric and allo-centric frames}
+\subsubsection{Ego-centric and Allo-centric Frames}
 
@@
 \emph{allo-centric} frames (addressee-centred).
+
+\subsubsection{Specific Limitations}
+
+One of the main limit of our current approach is the way semantics are
+conveyed: because we use 2D markers for artifacts, we skip most of the object
+extraction and recognition challenges. Each object has a unique identifier,
+which enable us to load a proper CAD model suitable for 3D spatial reasoning
+and prevent recognition mistakes in the knowledge base. While {\sc Spark}
+algorithms themselves are ignorant of the input sources, and would work equally
+well with a full object recognition stack, we do not have tackled it yet.
+
+Besides, temporal reasoning (essential for accurate action recognition for
+instance) is not properly addressed in our stack. Temporal reasoning is used
+only locally, and does not allow for tracking of long or global events.
 
@@
 
-While parts of the architecture (SPARK, ORO) have been deployed with external
+While parts of the architecture ({\sc Spark}, {\sc Oro}) have been deployed with external
 execution controllers (like {\sc Cram}~\cite{Beetz2010} or the BERT
@@
 actions and an event-based architecture well suited for prototyping.
-
-They both rely on extensive integration with the knowledge base that serves as the
+They both rely on deep integration with the knowledge base that serves as the
 primary source of semantics for decision making.
+
+The novelty of our contribution here is the interleaving of a semantic-aware
+architecture designed for autonomy with human-aware decisional components. To
+our knowledge, no other robotic architecture covers in depth those two
+perspectives.
+
+The Ke Jia project~\cite{Chen2010} is maybe one of the closest
+approach. It integrates on a mobile platform a knowledge representation language
+with natural language processing, task planning and motion planning. Default and
+non-monotonic reasoning has been especially researched within the Ke Jia project
+for symbolic task planning~\cite{Ji2011} and underspecified natural language
+processing. The Ke Jia robot has been demonstrated in several tasks involving human-robot
+interaction with natural language. These tasks include a task with multiple
+\emph{pick \& carry} that are globally optimised, naive physics reasoning via
+taught rules or more complex scenarios with the robot delivering drinks, taking
+into account changing and mutually exclusive preferences of users.
+
+Also notable, Beetz et al.~\cite{Beetz2010} proposes a cognitive architecture
+called \textsc{CRAM} (Cognitive Robot Abstract Machine) that deeply integrates
+with \textsc{KnowRob}~\cite{Tenorth2009a}, a knowledge processing framework
+based on Prolog. Its underlying storage is based on an OWL ontology, derived
+from \textsc{OpenCyc}. \textsc{CRAM} and \textsc{KnowRob} have been demonstrated
+on several real-world scenarios, where natural language recipes extracted from
+the Internet had to be translated into plans and executed in a kitchen
+environment, perceived and rebuilt on-line by the robots. This architecture has
+however not been deployed in many scenarios involving human-robot interactions.
 
@@
 The second one involves the {\sc Shary} execution controller and the HATP symbolic
-task planner besides ORO and {\sc Dialogs}. In this scenario, the human and the
+task planner besides {\sc Oro} and {\sc Dialogs}. In this scenario, the human and the
 robot try to cooperatively remove objects from a table. behaviours and motions
@@
 detected through motion capture. The robot knowledge base is pre-loaded with
-the \emph{ORO Commonsense Ontology}.  We next describe two
+the \emph{{\sc Oro} Commonsense Ontology}.  We next describe two
 situations where we can follow the internal robot's reasoning and the
@@
 
-This experiment involved a more complex decisional layer where the ORO server
+This experiment involved a more complex decisional layer where the {\sc Oro} server
 was used in conjunction with the HATP symbolic task planner and the {\sc Shary}
@@
 
-\section{Discussion: when Artificial Intelligence enables Human Robot
+\section{Discussion: when Artificial Intelligence enables Human-Robot
 Interaction}
@@
 
-This last section goes over again the main contributions we have presented
-so far: why they are relevant to robotics and artificial intelligence, how do
-they compare to similar approaches, how do we support our technical choices and
-what remains to be developed/explored in the future.
+This last section summarises our main contributions and tries to clarify the ``take home'' message: what are the challenges
+that human-robot interaction brings to artificial intelligence, and what is the
+current state-of-the-art in this field?
 
@@
 
-\subsection{Foundations of cognition for the robot}
-
-Because we build autonomous \emph{companion} robots, designed to live and
-interact in environments primarily conceived for humans, we endow our machines
-with a set of fundamental cognitive capabilities. The robots perform typically
-in dynamics, partially unknown environments, with a variety of places,
-artifacts and agents that may carry complex semantics.
-
-While our main focus is on the cognitive abilities required to interact with
-humans, \emph{situation assessment}, \emph{spatial reasoning}, \emph{knowledge
-representation and reasoning}, \emph{task planning} or \emph{high-level
-control} are mandatory cognitive foundations to build upon.
-
-\paragraph{Situation assessment, spatial reasoning}
-
-Real service robots need to ground their decision-making processes in their
-perceptions, be it physical (\eg what a camera sees) or not (\eg remote
-knowledge bases, dialogue). We call \emph{physical situation assessment} the
-cognitive skill that a robot exhibits when it assesses the nature and content of its
-surroundings and efficiently monitors its evolution.
-
-Numerous approaches exist, like amodal \emph{proxies}~\cite{Jacobsson2008},
-grounded amodal representations~\cite{Mavridis2006}, semantic maps
-(\cite{Nuechter2008, Galindo2008,Blodow2011}) or
-affordance-based planning and object classification~\cite{Lorken2008,
-Varadarajan2011}.
-
-Our approach, implemented in the {\sc Spark} module, is a grounded amodal
-representation of the robot environment, where different sensing modalities
-(vision-based fiducial markers tracking, RGBD human tracking, motion capture)
-are merged and where amodal spatial reasoning takes place (for instance,
-explanation generation for unexpected perceptions like a disappearing object).
-
-Spatial reasoning~\cite{O'Keefe1999} is a field in its own right, and has been
-used for natural language processing for applications such as direction
-recognition ~\cite{Kollar2010,Matuszek2010} or language
-grounding~\cite{Tellex2010}.~\cite{Skubic2004} present a spatial reasoner
-integrated in a robot which computes symbolic positions of objects.
-
-One of the main limit of our current approach is the way semantics are
-conveyed: because we use 2D markers for artifacts, we skip most of the object
-extraction and recognition challenges. Each object has a unique identifier,
-which enable us to load a proper CAD model suitable for 3D spatial reasoning
-and prevent recognition mistakes in the knowledge base. While {\sc Spark}
-algorithms themselves are ignorant of the input sources, and would work equally
-well with a full object recognition stack, we have not tackled it yet.
-
-Besides, temporal reasoning (essential for accurate action recognition for
-instance) is not properly addressed in our stack. Temporal reasoning is used
-only locally, and does not allow for tracking of long or global events.
-
-\paragraph{Robot control}
-
-Cognitive control architectures for robotics is a research field in its own right.
-Our originality of our contribution relates to the interleaving of a
-semantic-aware architecture designed for autonomy with human-aware decisional
-components. To our knowledge, no other robotic architecture covers in depth
-those two perspectives.
-
-The Ke Jia project~\cite{Chen2010} is maybe one of the closest approach. It
-integrates on a mobile platform a knowledge representation language with
-natural language processing, task planning and motion planning. Default and
-non-monotonic reasoning has been especially researched within the Ke Jia
-project for symbolic task planning~\cite{Ji2011} and underspecified natural
-language processing.
-
-The Ke Jia robot has been demonstrated in several tasks involving human-robot
-interaction with natural language. These tasks include a task with multiple
-\emph{pick \& carry} that are globally optimised, naive physics reasoning via
-taught rules or more complex scenarios with the robot delivering drinks, taking
-into account changing and mutually exclusive preferences of users.
-
-Also notable, Beetz et al.~\cite{Beetz2010} proposes a cognitive architecture
-called \textsc{CRAM} (Cognitive Robot Abstract Machine) that integrates
-\textsc{KnowRob}~\cite{Tenorth2009a}, a knowledge processing framework based on
-Prolog. Its underlying storage is based on an OWL ontology, derived from
-\textsc{OpenCyc}. \textsc{CRAM} and \textsc{KnowRob} have been demonstrated on
-several real-world scenarios, where natural language recipes extracted from the
-Internet had to be translated into plans and executed in a kitchen environment,
-perceived and rebuilt on-line by the robots. This architecture has however not
-been deployed in scenarios involving human-robot interactions.
-
-\subsection{Knowledge framework}
+\subsubsection{Embodied Cognition}
+
+Robotics is traditionally regarded as the prototypical instance of \emph{embodied}
+artificial intelligence, and this dimension is especially prevalent in
+human-robot interaction, where agents have to share a joint physical
+environment.
+
+It results in a tight coupling between the symbolic and the geometric realms:
+while AI at its origins was mostly a matter of symbolic models, it has been
+since recognised that not only the mind is not a purely abstract system,
+disconnected from the physical world, but even more, cognition fundamentally
+relies on its relation to the physical world (so-called \emph{embodied
+cognition}). Varela~\cite{Varela1992} is one of the main discoverer of these
+mechanisms, and coined the concept of \emph{enactivism} as the theoretical
+framework that study the links between cognition, embodiment and actions.
+
+The challenge of symbol grounding is tightly linked to this issue. It
+corresponds to the identification or creation, and then, maintenance of a link
+between the symbol (the syntactic form of knowledge the computer will
+manipulate) and its semantics, \ie its meaning, anchored in the world (the
+relations between the symbol, the referent of the symbol, and mediating minds
+is classically referred as the \emph{semantic triangle}, and has been
+extensively studied in linguistics). The issue of grounding is well known in
+cognitive science and is summarised by Harnard~\cite{Harnad1990} by this
+question: ``how the semantic interpretation of a formal symbol system can be
+made intrinsic to the system?''. This issue has a major practical importance in
+robotic: for a robot to be both endowed with a symbolic representational and
+reasoning system, and able to \emph{act} in the physical world, it must ground
+its knowledge.
+
+Grounding is implemented at different levels in our architecture. The main source
+of grounded symbolic knowledge is the situation assessment module, {\sc Spark}.
+Not only it builds symbolic facts from spatial reasoning, but it also relies on
+temporal reasoning to track the world state and build explanations to interpret
+unexpected perceptions (we mentioned above the example of a disappearing
+object).
+
+Because {\sc Spark} is also able to track humans, which enables in turn
+perspective-aware spatial reasoning, it produces grounded symbolic knowledge
+for interacting agents as well. This is a typical \emph{embodied} cognitive
+skill.
+
+Grounding also occurs during verbal and non-verbal communication. The {\sc
+Dialogs} module grounds new concepts introduced by the speaker by asking questions
+until it can attach the new concept to concepts already present in the
+knowledge pool (typically, if I ask the robot ``bring me a platypus'', the
+robot asks ``what is a platypus''. I may answer ``a mammal'' and the robot will
+ask ``what is a mammal?'' and so on until it anchors the new concepts to ones
+it already knows). Non verbal interaction (like gestures) is also partially
+grounded. When the user points at something, it can say ``give me this'' and
+the robot grounds \emph{this} to the pointed artifact.
+
+Because only objects marked with tags are currently recognized (in our typical
+lab experiments, it may be a dozen of them), the robot grounding mechanisms work
+in a small-sized closed world. This assumption obviously greatly facilitates
+the task. It remains to be seen if systematic marking of manipulated objects
+(for instance with RFID tags) is a realistic path. It may be the case if the
+robot is expected to be used in a well-defined, slowly changing environment.
+Real-time recognition of large libraries of objects is another path which may
+be viable soon~\cite{Dean2013Fast}.
+
+\subsection{Knowledge and Robotics}
 \label{krs-discussion}
 
+\subsubsection{RDF as a Formalism for Semantics}
+
 Knowledge flows in our architecture through a \emph{semantic blackboard}, the
-{\sc Oro}-server. We rely on Description Logics (OWL) to represent and
+{\sc Oro} server. We rely on Description Logics (OWL) to represent and
 manipulate facts.
@@
 
-Tools to manipulate and reason over ontologies are readily available, mature,
-and already well accepted in the robotic community~\cite{Tenorth2009a,
-Lim2011}. Alternatives like \emph{Answer Set Programming} have also been
+Alternatives like \emph{Answer Set Programming} have also been
 successfully investigated in robotics~\cite{Chen2010,Erdem2012}, in particular
@@
 
-\subsection{Interacting with humans}
-
-\paragraph{Building a model of a human during the interaction} Perspective
-Taking is a human ability which allows one to put him/herself in another
-person's point of view. Studied in psychology
+\subsubsection{Knowledge and Architecture}
+
+As a whole, the components that we have introduced so far build a
+knowledge-oriented architecture: Knowledge is explicitly stored in one central
+and consistent repository of facts, accessible for all modules. It relies on a
+\emph{strict formalism} (OWL statements), with a \emph{well defined vocabulary}
+(stated in the common-sense ontology). These first two points lead to a
+\emph{loosely-coupled architecture} where modules can be removed or replaced
+easily by other ones as long as they share the same semantics (modules are
+defined by the knowledge they produce).
+
+Also, we adopt a symbolic, reactive, event-driven approach to robot control. By
+managing events at the same level as the reasoner, we take full advantage of the
+inference abilities of {\sc Oro} to trigger events whose \texttt{true}
+conditions can be inferred.
+
+And finally, this architecture allows for the combination of very different
+knowledge modalities in a single homogeneous environment, bringing mutual
+benefits to components. For instance, the dialogue processing module can
+perfectly run without any geometric perception, but its disambiguation routines
+can transparently benefit from it when available (since richer symbolic
+descriptions of objects are then available).
+
+Those items are not new \emph{per se}. It is however interesting to underline
+the shift of focus this implies during the design and integration phases of
+robots: components of our deliberative layer are defined and glued together by
+the knowledge they produce and consume. Human-robot interaction, because it
+supposes operations at \emph{human-level} and in environments with complex
+semantics, acts here as a motivational force.
+
+\subsection{Bringing the Humans into the Equation}
+
+\subsubsection{Modelling the Humans}
+
+Perspective taking is this human ability which allows one to put him/herself in
+another person's point of view. Studied in psychology
 literature~\cite{Flavell1992,Tversky1999}, this ability is crucial when
@@
 the world in terms of visual perception, spatial descriptions, affordances and
-beliefs, etc.  In the last years these notions have been gradually
-employed in HRI.~\cite{Breazeal2006} presents a learning
-algorithm that takes into account information about a teacher's visual
-perspective in order to learn a task. ~\cite{Johnson2005} apply visual
-perspective taking for action recognition between two
-robots.~\cite{Trafton2005} use both visual and spatial perspective taking to
-find out the referent indicated by a human partner.
+beliefs, etc.  In the last years these notions have been gradually employed in
+HRI.~\cite{Breazeal2006} presents a learning algorithm that takes into account
+information about a teacher's visual perspective in order to learn a task.
+~\cite{Johnson2005} apply visual perspective taking for action recognition
+between two robots.~\cite{Trafton2005} use both visual and spatial perspective
+taking to find out the referent indicated by a human partner.
 
@@
 
-\paragraph{Planning for cooperation} HRI offers new challenges to planning
+\subsubsection{Acting in Presence of Humans}
+
+HRI offers new challenges to planning
 by adding a complementary constraint: the synthesis of plans that are not only
@@
 
-\paragraph{Limits of disambiguation at semantic level}
-
-Interaction with humans implies the ability to deal with semantics: semantics of
-verbal interaction, semantics of gestures, etc.  As a consequence, it also
-implies to deal with semantic disambiguation.
-
-One prototypical example of semantic disambiguation has been given in
-\cite{Ros2010b} with the child game \emph{spygame}: two players are facing
-each other with a set of random objects in-between, one player mentally choose
-one object, and the other player has to guess the object by asking closed
-questions like \emph{Is your object small or large?} Based on the knowledge it
-has acquired, the robot is able to minimize the number of questions required to
-find the object.
-
-When playing this kind of game, however, the issue arises that the robot has no
-way to select which knowledge about the object is relevant in the interaction
-context. For instance, the knowledge base may store facts like \stmt{obj1 type
-ActiveConcept} (which internally means that this concept was mentioned in a
-discussion in the last few seconds): this information is not a relevant
-property of \concept{obj1} when trying to disambiguate concepts with humans.
-This distinction between \emph{internal knowledge} (meaningful to
-the system only) and \emph{common knowledge} (whose meaning is understood by
-all the interactors) has not been properly dealt with in our architecture.
-
-Besides, even knowledge that belongs to the \emph{common knowledge} may not be
-appropriate in a given interaction context. For instance, the system may
-compute that at a given instant the human is looking at the object: \stmt{human
-looksAt obj1}. This property makes sense to both parties, but in the context of
-the \emph{spygame}, we would like to mainly use immanent properties, not
-volatile like a gaze. More research is required to identify relevant
-interaction contexts and knowledge classes attached to them.
-
-\subsection{Joint actions with humans}
-\label{sec:soa}
-
+\paragraph{Joint actions with humans}\label{sec:soa}
 The human presence brings new requirements for robot's abilities both
@@
 
-\subsection{Knowledge-oriented architecture}
-
-Altogether, the components we have presented compose a
-\emph{knowledge-oriented} architecture:
-
-\begin{itemize}
-
-    \item{Knowledge is explicitly stored in one central and consistent
-    repository of facts, accessible for all modules.}
-
-    \item{Knowledge is represented in a strict formalism (OWL statements) and
-    with a clearly defined vocabulary (stated in the common-sense ontology).}
-
-    \item{The first two points enable a loosely-coupled architecture where
-    modules can be removed or replaced easily by other ones as long as
-    they share the same semantics (modules are defined by the knowledge they
-    produce),}
-
-    \item{We adopt a \emph{symbolic}, reactive, event-driven approach to robot control.
-    By managing events at the same level as the reasoner, we take full
-    advantage of the inference abilities of {\sc Oro} to trigger events whose
-    \texttt{true} conditions can be inferred.}
-
-    \item{Finally, this architecture allows for the combination of very
-    different knowledge modalities in a single homogeneous environment,
-    bringing mutual benefits to components. For instance, the dialogue
-    processing module can perfectly run without any geometric
-    perception, but its disambiguation routines can transparently
-    benefit from it when available (since richer symbolic descriptions of
-    objects are then available).}
-
-\end{itemize}
-
-Those items are not new \emph{per se}. It is however interesting to
-underline the shift of focus this implies during the design and integration
-phases of robots: components of our deliberative layer are defined and glued together by
-the knowledge they produce and consume.
-
-It reveals how the robotic community is ``catching up'' with semantic
-technologies. Human-robot interaction, because it supposes operations in
-environments with complex semantics, drives this evolution.
-
-\subsection{Embodied Cognition}
-
-Robotics is traditionally regarded as the prototypical instance of \emph{embodied}
-artificial intelligence, and this dimension is especially prevalent in
-human-robot interaction, where agents have to share a joint physical
-environment.
-
-It results in a tight coupling between the symbolic and the geometric realms:
-while AI at its origins was mostly a matter of symbolic models, it has been
-since recognised that not only the mind is not a purely abstract system,
-disconnected from the physical world, but even more, cognition fundamentally
-relies on its relation to the physical world (so-called \emph{embodied
-cognition}). Varela~\cite{Varela1992} is one of the main discoverer of these
-mechanisms, and coined the concept of \emph{enactivism} as the theoretical
-framework that study the links between cognition, embodiment and actions.
-
-The challenge of symbol grounding is tightly linked to this issue. It
-corresponds to the identification or creation, and then, maintenance of a link
-between the symbol (the syntactic form of knowledge the computer will
-manipulate) and its semantics, \ie its meaning, anchored in the world (the
-relations between the symbol, the referent of the symbol, and mediating minds
-is classically referred as the \emph{semantic triangle}, and has been
-extensively studied in linguistics). The issue of grounding is well known in
-cognitive science and is summarised by Harnard~\cite{Harnad1990} by this
-question: ``how the semantic interpretation of a formal symbol system can be
-made intrinsic to the system?''. This issue has a major practical importance in
-robotic: for a robot to be both endowed with a symbolic representational and
-reasoning system, and able to \emph{act} in the physical world, it must ground
-its knowledge.
-
-Grounding is implemented at different levels in our architecture. The main source
-of grounded symbolic knowledge is the situation assessment module, {\sc Spark}.
-Not only it builds symbolic facts from spatial reasoning, but it also relies on
-temporal reasoning to track the world state and build explanations to interpret
-unexpected perceptions (we mentioned above the example of a disappearing
-object).
-
-Because {\sc Spark} is also able to track humans, which enables in turn
-perspective-aware spatial reasoning, it produces grounded symbolic knowledge
-for interacting agents as well. This is a typical \emph{embodied} cognitive
-skill.
-
-Grounding also occurs during verbal and non-verbal communication. The {\sc
-Dialogs} module grounds new concepts introduced by the speaker by asking questions
-until it can attach the new concept to concepts already present in the
-knowledge pool (typically, if I ask the robot ``bring me a platypus'', the
-robot asks ``what is a platypus''. I may answer ``a mammal'' and the robot will
-ask ``what is a mammal?'' and so on until it anchors the new concepts to ones
-it already knows). Non verbal interaction (like gestures) is also partially
-grounded. When the user points at something, it can say ``give me this'' and
-the robot grounds \emph{this} to the pointed artifact.
-
-Because only objects marked with tags are currently recognized (in our typical
-lab experiments, it may be a dozen of them), the robot grounding mechanisms work
-in a small-sized closed world. This assumption obviously greatly facilitates
-the task. It remains to be seen if systematic marking of manipulated objects
-(for instance with RFID tags) is a realistic path. It may be the case if the
-robot is expected to be used in a well-defined, slowly changing environment.
-Real-time recognition of large libraries of objects is another path which may
-be viable soon~\cite{Dean2013Fast}.
+\paragraph{Limits of Disambiguation at Semantic Level}
+
+Interaction with humans implies the ability to deal with semantics: semantics of
+verbal interaction, semantics of gestures, etc.  As a consequence, it also
+implies to deal with semantic disambiguation.
+
+One prototypical example of semantic disambiguation has been given in
+\cite{Ros2010b} with the child game \emph{spygame}: two players are facing
+each other with a set of random objects in-between, one player mentally choose
+one object, and the other player has to guess the object by asking closed
+questions like \emph{Is your object small or large?} Based on the knowledge it
+has acquired, the robot is able to minimize the number of questions required to
+find the object.
+
+When playing this kind of game, however, the issue arises that the robot has no
+way to select which knowledge about the object is relevant in the interaction
+context. For instance, the knowledge base may store facts like \stmt{obj1 type
+ActiveConcept} (which internally means that this concept was mentioned in a
+discussion in the last few seconds): this information is not a relevant
+property of \concept{obj1} when trying to disambiguate concepts with humans.
+This distinction between \emph{internal knowledge} (meaningful to
+the system only) and \emph{common knowledge} (whose meaning is understood by
+all the interactors) has not been properly dealt with in our architecture.
+
+Besides, even knowledge that belongs to the \emph{common knowledge} may not be
+appropriate in a given interaction context. For instance, the system may
+compute that at a given instant the human is looking at the object: \stmt{human
+looksAt obj1}. This property makes sense to both parties, but in the context of
+the \emph{spygame}, we would like to mainly use immanent properties, not
+volatile like a gaze. More research is required to identify relevant
+interaction contexts and knowledge classes attached to them.
 
@@
 
-We have presented in this article a decisional framework for human-robot
-interactive task achievement that is aimed to allow the robot not only to
-accomplish its tasks but also to produce behaviours that support its engagement
-vis-a-vis its human partner and to interpret human behaviours and intentions.
-Together and in coherence with this framework, we have developed and
-experimented various task planners and interaction schemes that allow the robot
-to select and perform its tasks while taking into account explicitly the human
-abilities as well as the constraints imposed by the presence of humans, their
-needs and preferences.
+%We have presented in this article a decisional framework for human-robot
+%interactive task achievement that is aimed to allow the robot not only to
+%accomplish its tasks but also to produce behaviours that support its engagement
+%vis-a-vis its human partner and to interpret human behaviours and intentions.
+%Together and in coherence with this framework, we have developed and
+%experimented various task planners and interaction schemes that allow the robot
+%to select and perform its tasks while taking into account explicitly the human
+%abilities as well as the constraints imposed by the presence of humans, their
+%needs and preferences.
 
@@
 There is also extensive work to be done in order to refine the notion of ``good
-shared plan" and ``good/acceptable robot behaviour" in this context. There are
+shared plan'' and ``good/acceptable robot behaviour'' in this context. There are
 obviously large avenues for learning and adaptation in this context.
@@
 largely remain to be developed, as well as efficient algorithms to retrieve (or
-discard) such context-related pools of knowledge. This is a challenge not only for robotics, but more generally for artificial intelligence.
+discard) such context-related pools of knowledge. This is a challenge not only
+for robotics, but more generally for artificial intelligence.
 

### 2014-05-05 paper.tex
Adjustment to spark-oro figure

@@
 
-    \node[draw=none] at (0,0) {\includegraphics[width=0.8\columnwidth]{spark-oro-notext.pdf}};
+    \node[draw=none] at (0,-1) {\includegraphics[width=0.8\columnwidth]{spark-oro-notext.pdf}};
 
@@
 
-    \node[box=16cm, minimum height=3cm, dotted, thick, rounded corners] at (0,-3.75) (spark) {};
+    \node[box=16cm, minimum height=3.5cm, dotted, thick, rounded corners] at (0,-3.75) (spark) {};
     \path (spark.north) edge[->] node[align=left,anchor=west] {Symbolic\\facts} (oro.south);
 
-    \node[anchor=north] at (-4.5,-4.5) {Relative locations};
-    \node[anchor=north] at (-1.5,-4.5) {Spatial relations};
-    \node[anchor=north] at (1.5,-4.5) {Visibility};
-    \node[anchor=north] at (4.8,-4.5) {Reachability};
+    \node[anchor=north] at (-5.5,-4.5) {a) Agent-independent\\locations};
+    \node[anchor=north] at (-1.8,-4.5) {b) Agent-dependent\\placements};
+    \node[anchor=north] at (1.5,-4.5) {c) Visibility};
+    \node[anchor=north] at (5.3,-4.5) {d) Reachability};
 

### 2014-05-05 paper.tex
Major rewrite of the SPARK section
Based on my PhD work

@@
 
-\subsection{Situation Assessment, Human Awareness, Perspective Taking}
+\subsection{Acquiring and Anchoring Knowledge in the Physical World}
 \label{sect|sit-ass}
@@
 
-{\sc Spark} is designed to run and continously update the knowledge base (in our
+{\sc Spark} is designed to run and continuously update the knowledge base (in our
 experiments, it runs at about 10Hz). At each steps, it re-computes spatial
@@
 approach may raise scalability concerns (we did not however hit such issues in
-our constrainted scenarios), but greatly simplifies the management of the
+our constrained scenarios), but greatly simplifies the management of the
 \emph{dynamics} of the knowledge (\emph{When do I discard outdated knowledge?
@@
 %
-       \subfigure[Initial state]{%
+       \subfigure[Physical setting]{%
 %           \label{fig:wuweiPhoto}
@@
 
-   \caption{The toy scenario pictured here involves videotapes that have to be
-       manipulated, with other objects used as supports and containers.  After
-       identification and localisation of the set of objects and acquisition of
-       the position and posture of the human partner, the robot is able to
-       compute that two tapes only are reachable by itself: the
-       \concept{LOTR\_TAPE} (black in the 3D model) and \concept{GREY\_TAPE}.
-       The third tape \concept{WALLE\_TAPE} (white in the 3D model) and the
-   trashbin \concept{PINK\_TRASHBIN} are only reachable by the human. The
-   corresponding symbolic facts are listed in Table~\ref{table|beliefsfig7}.  }%
+   \caption{Toy scenario involving videotapes that have to be manipulated, with
+       other objects used as supports and containers.  After identification and
+       localisation of the set of objects and acquisition of the position and
+       posture of the human partner, the robot is able to compute that two tapes
+       only are reachable by itself: the black and grey (in the 3D model) tapes.
+       The third tape and the pink box are only reachable by the human. This
+       situation leads to the symbolic model presented in
+       Table~\ref{table|beliefsfig7}.  }%
 
@@
 \hline
-  \hspace{0.7cm}\stmt{PINK\_TRASHBIN isReachable false}\\
-  \hspace{0.7cm}\stmt{WALLE\_TAPE isReachable {\bf false}}\\
-  \hspace{0.7cm}\stmt{LOTR\_TAPE isReachable {\bf true}}\\
+  \hspace{0.7cm}\stmt{PINK\_BOX isReachable false}\\
+  \hspace{0.7cm}\stmt{WHITE\_TAPE isReachable {\bf false}}\\
+  \hspace{0.7cm}\stmt{BLACK\_TAPE isReachable {\bf true}}\\
   \hspace{0.7cm}\stmt{GREY\_TAPE isReachable {\bf true}}\\
-  \hspace{0.7cm}\stmt{WALLE\_TAPE isVisible true}\\
-  \hspace{0.7cm}\stmt{LOTR\_TAPE isVisible true}\\
+  \hspace{0.7cm}\stmt{WHITE\_TAPE isVisible true}\\
+  \hspace{0.7cm}\stmt{BLACK\_TAPE isVisible true}\\
   \hspace{0.7cm}\stmt{GREY\_TAPE isVisible true}\\
-  \hspace{0.7cm}\stmt{WALLE\_TAPE isOn TABLE}\\
-  \hspace{0.7cm}\stmt{LOTR\_TAPE isOn TABLE}\\
+  \hspace{0.7cm}\stmt{WHITE\_TAPE isOn TABLE}\\
+  \hspace{0.7cm}\stmt{BLACK\_TAPE isOn TABLE}\\
   \hspace{0.7cm}\stmt{GREY\_TAPE isOn TABLE}\\
@@
 \hline
-  \hspace{0.7cm}\stmt{PINK\_TRASHBIN isReachable false}\\
-  \hspace{0.7cm}\stmt{WALLE\_TAPE isReachable {\bf true}}\\
-      \hspace{0.7cm}\stmt{LOTR\_TAPE isReachable {\bf false}}\\
+  \hspace{0.7cm}\stmt{PINK\_BOX isReachable false}\\
+  \hspace{0.7cm}\stmt{WHITE\_TAPE isReachable {\bf true}}\\
+      \hspace{0.7cm}\stmt{BLACK\_TAPE isReachable {\bf false}}\\
       \hspace{0.7cm}\stmt{GREY\_TAPE isReachable {\bf false}}\\
-  \hspace{0.7cm}\stmt{WALLE\_TAPE isVisible true}\\
-  \hspace{0.7cm}\stmt{LOTR\_TAPE isVisible true}\\
+  \hspace{0.7cm}\stmt{WHITE\_TAPE isVisible true}\\
+  \hspace{0.7cm}\stmt{BLACK\_TAPE isVisible true}\\
   \hspace{0.7cm}\stmt{GREY\_TAPE isVisible true}\\
-  \hspace{0.7cm}\stmt{WALLE\_TAPE isOn TABLE}\\
-  \hspace{0.7cm}\stmt{LOTR\_TAPE isOn TABLE}\\
+  \hspace{0.7cm}\stmt{WHITE\_TAPE isOn TABLE}\\
+  \hspace{0.7cm}\stmt{BLACK\_TAPE isOn TABLE}\\
   \hspace{0.7cm}\stmt{GREY\_TAPE isOn TABLE}\\ 
@@
 
-\subsubsection{Situation Assessment}
-
-\paragraph{Symbolic Facts Production}
-
-\begin{inparaenum}[\itshape 1\upshape)]
-Geometric state of the world is abstracted in symbolic facts that can be
-classified in three different categories: \item relative positions of object and
-agents, \eg  \stmt{GREY\_TAPE isOn TABLE}, \item perception and manipulation
-capacity and state of agents, \eg \stmt{ROBOT looksAt GREY\_TAPE},
-\stmt{GREY\_TAPE isVisibleBy HUMAN1}, \item motion status for object or
-agent parts, \eg \stmt{GREY\_TAPE isMoving true}, \stmt{ROBOT\_HEAD
-isTurning true}.
-\end{inparaenum}
-By reasoning about human perspective, it computes facts such as:
-\stmt{GREY\_TAPE isBehind HUMAN1}, \stmt{GREY\_TAPE leftOf HUMAN1}.
-
-As an illustration, Figure~\ref{fig::reach-ex} explains how the \emph{reachable}
-affordance is mutually computed for the human and the robot. The \emph{is visible},
-\emph{looks at}, \emph{points at} affordances are computed in similar ways, by
-virtually placing OpenGL cameras in different pose (human head pose, robot head
-pose, human index finger, etc.) and evaluating which objects are visible to the camera
-(hence taking into account possible occlusions).
-
-\begin{figure*}[!t]
+\subsubsection{Building an Agent-aware Symbolic Model of the Environment}
+\label{sect|situ}
+
+\paragraph{On Perspective Taking} Visual perspective taking refers to the
+ability for visually perceiving the environment from other's point of view.
+This ability allows us to identify the objects in situations where the visual
+perception of one person differs from the other one. In developmental
+psychology, one typical example consists of two similar objects in a room (\eg.
+two balls) where both are visible for the child, but only one is visible for
+the adult. Thus, when the adult asks the child to hand over ``the ball'', the
+child is able to correctly identify which ball the adult is referring to (\ie
+the one visible from the adult point of view), without asking~\cite{Moll2006}.
+
+Besides, in order to compute a visual perspective, the actual visibility alone
+is not enough. We include not only what the other person sees in a given
+moment, but also what he \emph{can} see with a minimal effort (moving the eyes
+or the head). To model the potential visibility of an object we compute a
+visibility ratio while turning the head of the agent model towards the object.
+
+Spatial perspective taking refers to the qualitative spatial location of
+objects (or agents) with respect to a frame (\eg \emph{the keys on my left}).
+Based on this frame of reference, the description of an object
+varies~\cite{Marin2008}. Humans mix perspectives frequently during interaction.
+This is more effective than maintaining a consistent one, either because the
+(cognitive) cost of switching is lower than remaining with the same
+perspective, or if the cost is about the same, because the spatial situation
+may be more easily described from one perspective rather than
+another~\cite{Tversky1999}. Ambiguities arise when one speaker refers to an
+object within a reference system (or changes the reference system, \ie switches
+perspective) without informing his/her partner about it~\cite{Breazeal2006,
+Ros2010}. For example, the speaker could ask for the ``keys on the left''.
+Since no reference system has been given, the listener would not know where
+exactly to look.  However, asking for ``the keys on your left'' gives enough
+information to the listener to understand where the speaker is referring to. On
+the contrary, when using an exact, unambiguous term of reference to describe a
+location (\eg. ``go north'') no ambiguity arises.
+
+In {\sc Spark}, we use two types of the frames of reference: \emph{egocentric}
+(from the robot perspective) and \emph{allo-centric}, \ie addressee-centred
+(from the human perspective).
+
+\paragraph{Symbolic Locations}
+
+Humans commonly refer to the positions of objects with symbolic descriptors
+(like \emph{on}, \emph{next to}...) instead of precise, absolute positions.
+These type of descriptors have been studied in the context of language
+grounding \cite{O'Keefe1999,Matuszek2010,Regier2001,Kelleher2006,Blisard2005}.
+In {\sc Spark} we focus on agent-independent symbolic locations and agent-dependent,
+relative locations.
+
+\paragraph{Agent-Independent Locations}
+
+We can refer to object locations with respect to other objects in the
+environment, such as \emph{above, next to, in}, etc. {\sc Spark} computes
+three main relations based on the bounding box and centre of mass of the
+objects (Figure~\ref{fig|spark-oro}\emph{a}):
+
+\begin{itemize}
+	\item \concept{isOn}: computes if an object $O_1$ is on another object $O_2$ by
+	evaluating the center of mass of $O_1$ according to the bounding box of $O_2$.
+
+	\item \concept{isIn}: evaluates if an object $O_1$ is inside another object
+	$O_2$ based on their bounding boxes $BB_{O_1}$ and $BB_{O_2}$.
+
+	\item \concept{isNextTo}: indicates whether an object $O_1$ is next to another
+	object $O_2$. We cannot use a simple distance threshold to determine if two
+	objects are next to each other since the relation is highly dependent on the
+	dimensions of the objects. For instance, the maximum distance between large
+	objects (\eg two houses) to consider them as being next to each other is much
+	larger than the maximum distance we would consider for two small objects (\eg
+	two bottles). Thus, the relation between the dimensions and the distances of
+	the objects are taken into account.
+
+\end{itemize}
+
+{\sc Spark} also compute symbolic facts related to agent independent world dynamics.
+The predicate \concept{isMoving} states, for each tracked entity, whether it is
+currently moving or not.
+
+\paragraph{Agent-Dependent Placements}
+
+While in previous section we listed several \emph{absolute} location predicate,
+many topological relations are directly dependent from the observation point.
+
+The predicate \concept{hasRelativePosition} represents spatial locations
+between agents and objects that are agent dependent.  We compute these spatial
+locations by dividing the space around the referent (an agent) into $n$ regions
+based on arbitrary angle values relative to the referent orientation
+(Figure~\ref{fig|spark-oro}\emph{b}).  For
+example, for $n = 4$ we would have the space divided into \emph{front, left,
+right} and \emph{back}. Additionally, two proximity values, \emph{near} and
+\emph{far}, are also considered. The number of regions and proximity values can
+be chosen depending on the context where the interaction takes place.
+
+Through perspective taking, {\sc Spark} computes for each agent a symbolic
+description of the relative positioning of objects in the environment.
+
+\subsubsection{Building a Model of Agents}
+\label{sect|grounding_agents}
+
+Building a grounded symbolic model of the physical environment does not suffice
+in general to fully ground the human-robot interaction, and a model of the
+current capabilities of the agents surrounding the robot is also required.
+
+There are a number of common properties for a robot and a human related to
+their capabilities in a given situation: they can both reach, grasp, look at,
+point at, etc.: we group them in the \emph{Agent} category, defined as entities
+that can act in the environment and manipulate it.
+
+{\sc Spark} computes the following capabilities from the perspectives of each agent:
+
+\begin{itemize}
+
+\item \emph{Sees}: An important ability to know about an agent is to predict
+\emph{What can it see?}, \ie what is within its field of view (FOV). Being able
+to compute this information, the robot can reuse it for instance to infers
+which object a human is searching for (the one that is not currently visible,
+otherwise the user would not be searching for it).  In
+figure~\ref{fig|spark-oro}\emph{c} the field of view of a person is
+illustrated with a grey cone (broader one). While he is able to see the two
+small boxes on the table in front of him, the big box on his right is out of
+his FOV, and therefore, he is not able to see it.
+
+\item \emph{Looks At}: this relation corresponds to what the agent is focused
+on, \ie where its focus of attention is directed. This model is based on a
+narrower field of view, the field of attention (FOA). 
+Figure~\ref{fig::sparkRepresentations}\emph{c}
+shows the field of attention of a person with a green cone (narrower one). In
+this example only the grey box satisfies the \concept{looksAt} relation.
+
+\item \emph{Points At}: verifies whether an object is pointed at by an agent.
+    This relation is computed by placing a virtual camera on the hand, aligned
+    with the forearm. \concept{pointsAt} is particularly useful during
+    interaction when one of the agents is referring to an object saying ``this''
+    or ``that'' while pointing at it.
+
+If a larger object occludes a smaller one while an agent is pointing at them, the
+outcome of the evaluation will result only in one relation, \ie \stmt{agent\_01
+pointsAt object\_01} since the small one is not visible to the agent.  On the
+contrary, if the small object is in front of the big one, then both objects
+will satisfy the relation, which may generate an ambiguity (which object the
+agent refers to?) that is let to be solved by other discrimination algorithms.
+
+To make recognition more robust, these three first capabilities are filtered
+with an hysteresis function at the geometric level.
+
+\item \emph{Reachable}: it allows the robot to estimate the agent's capability
+to reach an object, which is fundamental for task planning. For example, if the
+user asks the robot to give him/her an object, the robot must compute a transfer
+point where the user is able to get the object afterwards. 
+Figure~\ref{fig::reach-ex} explains how addressee-dependent reachability is
+computed, based on Generalized Inverse Kinematics and collision detection.
+
+\end{itemize}
+
+Table~\ref{facts} lists all the symbolic facts that are computed by {\sc Spark},
+along with the type of the statements' subjects and objects.
+
+\renewcommand{\concept}[1]{{\scriptsize \texttt{#1}}}
+\begin{table}[h]
     \centering
-    \subfigure[] { \includegraphics[width=0.2\textwidth]{reachex-1.png} }
-    \subfigure[] {\includegraphics[width=0.2\textwidth]{reachex-2.jpg} }
-    \subfigure[] {\includegraphics[width=0.2\textwidth]{reachex-3.jpg} }
-    \subfigure[] {\includegraphics[width=0.2\textwidth]{reachex-4.jpg} }
-    \caption{An example illustrating the \textit{reachable} relation. The
-        relation is computed from the perspectives of both the robot and the
-        human. The computed posture at each step is illustrated with a global
-        view of the scene (top), and from a closest view (bottom). The robot
-        and its human partner are placed face to face (a).  The robot first
-        estimates if the small grey box is reachable to itself using an inverse
-        kinematic (IK) solver and collision checks to find a collision free
-        posture to reach the object (b). Next the robot switches to the human's
-        perspective to estimate if the same object is reachable to the human as
-        well (c).  In the last scene, the human moves towards his left, farther
-        from the object (d). The situation is then reevaluated. In this occasion
-        though, the reasoner cannot find a satisfactory posture for the human to
-    reach the box because he is too far from the target.  }
-
-\label{fig::reach-ex}
-\end{figure*}
+    \begin{tabular}{p{1.5cm}lp{2cm}p{3.7cm}}
+    \textbf{Subject} & \textbf{Predicate} & \textbf{Object} & \emph{Notes} \\ 
+    \hline
+	 \concept{Location} & \concept{isAt} $\equiv$ \concept{cyc:objectFoundInLocation}  &  \concept{Location} & \\ 
+	 &  $\rightarrow$ \concept{isOn} $\equiv$ \concept{cyc:above\_Touching}  &  & \\ 
+	 &  $\rightarrow$ \concept{isIn}  &  & \\ 
+	 &  $\rightarrow$ \concept{isNextTo}  & &  \\ 
+	 \concept{Location}  & \concept{isAbove} $\equiv$ \concept{cyc:above-Generally}  &  \concept{Location}  &  inverse of \concept{isBelow} \par \concept{isOn} $\Rightarrow$ \concept{isAbove}\\ 
+	 \concept{Location}  & \concept{isBelow}  & \concept{Location}  &  inverse
+	of \concept{isAbove} \\
+	\hline
+	 \concept{Location}  & \concept{hasRelativePosition}  & \concept{Location} & \\ 
+	 & 	$\rightarrow$ \concept{behind} $\equiv$ \concept{cyc:behind-Generally}  &  & inverse of \concept{inFrontOf}  \\ 
+	 &  $\rightarrow$ \concept{inFrontOf} $\equiv$ \concept{cyc:inFrontOf-Generally}  & 	 & 	 inverse of \concept{behind}  \\ 
+	 &  $\rightarrow$ \concept{leftOf}  &  &  inverse of \concept{rightOf} \\ 
+	 &  $\rightarrow$ \concept{rightOf}  & 	 & 	 inverse of \concept{leftOf}  \\ 
+	 \concept{Object}  & \concept{cyc:farFrom}  &  \concept{Agent} & \\ 
+	 \concept{Object}  & \concept{cyc:near}  &  \concept{Agent} & \\
+
+		\hline
+		 \concept{Agent}  & \concept{looksAt}  & \concept{SpatialThing} \\
+		 \concept{Agent}  & \concept{sees}  &  \concept{SpatialThing}  &    \\ 
+		 \concept{SpatialThing}  & \concept{isInFieldOfView}  &  \concept{xsd:boolean}  & via inference: \par \tiny \stmt{myself sees *} $\Leftrightarrow$ \stmt{* isInFieldOfView true} \\ 
+		 \concept{Agent}  & \concept{pointsAt} $\equiv$ \concept{cyc:pointingToward}  & \concept{SpatialThing} \\ 
+		 \concept{Agent}  & \concept{focusesOn}  &  \concept{SpatialThing}  &  via inference: \par \tiny \concept{looksAt} $\wedge$ \concept{pointsAt} $\Rightarrow$ \concept{focusesOn} \\
+		\concept{Agent} & \concept{seesWithHeadMovement} &  \concept{SpatialThing} \\
+		\concept{Agent} & \concept{reaches} &  \concept{Object} \\
+
+	\end{tabular}
+
+    \caption{List of statements describing agent-independent spatial
+        relationships between objects (top), agent-dependent placements (middle)
+        and attentional states and abilities of agents (bottom).
+        ``$\rightarrow$'' indicates sub-properties. When existing, the
+        equivalent predicate in the {\sc OpenCyc} standard (prefix \concept{cyc:})
+        is specified. Note that some relationships are not computed by {\sc Spark}, but
+        are instead inferred by the reasoner.}
+
+	\label{facts}
+\end{table}
+\renewcommand{\concept}[1]{{\small \texttt{#1}}}
+
+%\subsubsection{Situation Assessment}
+%
+%\paragraph{Symbolic Facts Production} 
+%
+%\begin{inparaenum}[\itshape 1\upshape)]
+%Geometric state of the world is abstracted in symbolic facts that can be
+%classified in three different categories: \item relative positions of object and
+%agents, \eg  \stmt{GREY\_TAPE isOn TABLE}, \item perception and manipulation
+%capacity and state of agents, \eg \stmt{ROBOT looksAt GREY\_TAPE},
+%\stmt{GREY\_TAPE isVisibleBy HUMAN1}, \item motion status for object or
+%agent parts, \eg \stmt{GREY\_TAPE isMoving true}, \stmt{ROBOT\_HEAD
+%isTurning true}.
+%\end{inparaenum}
+%By reasoning about human perspective, it computes facts such as:
+%\stmt{GREY\_TAPE isBehind HUMAN1}, \stmt{GREY\_TAPE leftOf HUMAN1}.
+%
+%As an illustration, Figure~\ref{fig::reach-ex} explains how the \emph{reachable}
+%affordance is mutually computed for the human and the robot. The \emph{is visible},
+%\emph{looks at}, \emph{points at} affordances are computed in similar ways, by
+%virtually placing OpenGL cameras in different pose (human head pose, robot head
+%pose, human index finger, etc.) and evaluating which objects are visible to the camera
+%(hence taking into account possible occlusions).
+%
+%\begin{figure*}[!t]
+%	\centering
+%    \subfigure[] { \includegraphics[width=0.2\textwidth]{reachex-1.png} }
+%    \subfigure[] {\includegraphics[width=0.2\textwidth]{reachex-2.jpg} }
+%    \subfigure[] {\includegraphics[width=0.2\textwidth]{reachex-3.jpg} }
+%    \subfigure[] {\includegraphics[width=0.2\textwidth]{reachex-4.jpg} }
+%    \caption{An example illustrating the \textit{reachable} relation. The
+%        relation is computed from the perspectives of both the robot and the
+%        human. The computed posture at each step is illustrated with a global
+%        view of the scene (top), and from a closest view (bottom). The robot
+%        and its human partner are placed face to face (a).  The robot first
+%        estimates if the small grey box is reachable to itself using an inverse
+%        kinematic (IK) solver and collision checks to find a collision free
+%        posture to reach the object (b). Next the robot switches to the human's
+%        perspective to estimate if the same object is reachable to the human as
+%        well (c).  In the last scene, the human moves towards his left, farther
+%        from the object (d). The situation is then reevaluated. In this occasion
+%        though, the reasoner cannot find a satisfactory posture for the human to
+%    reach the box because he is too far from the target.  }
+%
+%\label{fig::reach-ex}
+%\end{figure*}
+%
 
@@
 
+\fxfatal{Rephrase}
 It is sometimes difficult or even impossible to see and/or track an object in
@@
 
-\paragraph{Primitive Action Recognition}
+\subsubsection{Primitive Action Recognition}
 
@@
 
-\subsubsection{Ego-centric and Allo-centric Frames}
-
-{\sc Spark} also supports \emph{perspective-taking}: spatial relations between entities can
-be computed from different viewpoints, which let the robot build for each
-agents it interacts with a different, \emph{perspective-aware} symbolic model
-of the environment. These models are separately stored in the knowledge base.
-
-This allows us to deal with ambiguities that arise when one speaker refers to
-an object within a reference system (or changes the reference system, \ie
-switches perspective) without making the reference frame
-explicit~\cite{Breazeal2006, Ros2010}.
-
-As a result, the robot stores models of the environment either in the
-\emph{ego-centric} reference frame (from the robot perspective) or in 
-\emph{allo-centric} frames (addressee-centred).
-
-\subsubsection{Specific Limitations}
-
-One of the main limit of our current approach is the way semantics are
+%\subsubsection{Ego-centric and Allo-centric Frames}
+%
+%{\sc Spark} also supports \emph{perspective-taking}: spatial relations between entities can
+%be computed from different viewpoints, which let the robot build for each
+%agents it interacts with a different, \emph{perspective-aware} symbolic model
+%of the environment. These models are separately stored in the knowledge base.
+%
+%This allows us to deal with ambiguities that arise when one speaker refers to
+%an object within a reference system (or changes the reference system, \ie
+%switches perspective) without making the reference frame
+%explicit~\cite{Breazeal2006, Ros2010}.
+%
+%As a result, the robot stores models of the environment either in the
+%\emph{ego-centric} reference frame (from the robot perspective) or in 
+%\emph{allo-centric} frames (addressee-centred).
+
+\subsubsection{Limitations}
+
+The main weakness of our current implementation relates to the way semantics are
 conveyed: because we use 2D markers for artifacts, we skip most of the object

### 2014-05-06 paper.tex
US->GB, typos, capitalization

@@
 lays at the crossroad of several domains of AI and requires to tackle them in a
-holistic manner: Modeling humans and human cognition; acquiring, representing,
+holistic manner: Modelling humans and human cognition; acquiring, representing,
 manipulating in a tractable way abstract knowledge at the human level; reasoning
@@
 
-We do not claim to address here the whole issue. This article attempts however to organize
+We do not claim to address here the whole issue. This article attempts however to organise
 it into a coherent challenge for artificial intelligence, and to explain and
@@
 perspectives, skills of the human partner.  Namely, the robot must be able to
-recognize, understand and participate to communication situations, both explicit
+recognise, understand and participate to communication situations, both explicit
 (\eg the human addresses verbally the robot) and implicit (\eg the human points
@@
 tackle the AI challenges of human-robot interaction, and how its relates to
-other approaches. We propose to organize this discussion in four sections.
+other approaches. We propose to organise this discussion in four sections.
 
@@
 experiments that illustrate in a concrete way what can be currently achieved
-with our robots.  Section~\ref{sect|conclusion} finally summarizes our main
+with our robots.  Section~\ref{sect|conclusion} finally summarises our main
 contributions and relate them to other recent researches in the field.
@@
         \node [subpart, right=0.2 of plans] (sit-asses) {Situation assessment \\ and context management};
-        \node [subpart, right=0.2 of sit-asses] {Action instanciation, \\ execution and monitoring};
+        \node [subpart, right=0.2 of sit-asses] {Action instantiation, \\ execution and monitoring};
       \end{tikzpicture}
@@
 desires GIVE\_ACTION45, GIVE\_ACTION45 actsOn BOOK2} to {\sc Oro}. The HATP
-planner then uses the knowledge base to initialize the planning domains with
+planner then uses the knowledge base to initialise the planning domains with
 similar requests ({\tt \footnotesize find(BOOK2 isAt ?location)}, etc.). The
@@
 execution (the desires and the intentions) is central, and aims at selecting at
-each step the best subgoal. It becomes then an intention that the robot commits
+each step the best sub-goal. It becomes then an intention that the robot commits
 to.
@@
 our design leads to a good \emph{observability} (knowledge flows are explicit
-and easy to capture since they are centralized) as well as high modularity
+and easy to capture since they are centralised) as well as high modularity
 (modules communicate through a defined and unified API).
@@
 three sources. {\it A priori} knowledge is stored in an ontology (the {\sc
-OpenRobots} ontology, discussed below) and loaded at startup. This static source covers the
+OpenRobots} ontology, discussed below) and loaded at start-up. This static source covers the
 \emph{common-sense} knowledge of the robot, and optionally some
@@
 the concept of \emph{computables} to lazily evaluate/acquire symbolic facts when
-needed, we have an \emph{explicit} approach where we greedly compute and assert
+needed, we have an \emph{explicit} approach where we greedily compute and assert
 symbolic statements (like spatial relations between objects, see
 Section~\ref{sect|sit-ass}). This design choice trades scalability for explicit
-reasoning: at any time, the full belief state is explicitely stated, and
+reasoning: at any time, the full belief state is explicitly stated, and
 provides the reasoner with the largest possible inference domain. This is of
@@
 Section~\ref{sect|ctrl}), where, by definition, we do not know beforehand what
-we are ``looking for'', and we need instead greedly infer as much as possible
+we are ``looking for'', and we need instead greedily infer as much as possible
 to build a belief state as comprehensive as possible.
@@
 to the context of service robotics in an human-like environment. For performance
-reasons as well as clarity, we do not attempt to proactively extend the
+reasons as well as clarity, we do not attempt to pro-actively extend the
 conceptual coverage of the field. We instead opportunistically extend the
@@
 the knowledge model, and hence implemented directly within the {\sc Oro} server.
-We present here three of them: \emph{reasoning}, \emph{theory of mind} modeling
+We present here three of them: \emph{reasoning}, \emph{theory of mind} modelling
 and our (naive) approach to \emph{memory management}.
@@
 Our knowledge base implements such a mechanism: when the robots recognises that
-a new agent has been created in the knowledge base, it initializes a new,
+a new agent has been created in the knowledge base, it initialises a new,
 independent, ontology for this agent. All the ontologies that are created share
@@
 	\item \concept{isOn}: computes if an object $O_1$ is on another object $O_2$ by
-	evaluating the center of mass of $O_1$ according to the bounding box of $O_2$.
+	evaluating the centre of mass of $O_1$ according to the bounding box of $O_2$.
 
@@
 Figure~\ref{fig::reach-ex} explains how addressee-dependent reachability is
-computed, based on Generalized Inverse Kinematics and collision detection.
+computed, based on Generalised Inverse Kinematics and collision detection.
 
@@
 \emph{pick}), or ``a hand holding an object is over a container'' (precursor for
-a \emph{throw}). \emph{Spark} recognizes a handful \fxfatal{which ones?} of these
+a \emph{throw}). \emph{Spark} recognises a handful \fxfatal{which ones?} of these
 primitives, that are relied on by the robot to track the state transitions in
@@
 
-\subsection{Human-aware task planning}
+\subsection{Human-Aware Task Planning}
 \label{hatp}
@@
 
-\paragraph{Agents and action streams}
+\subsubsection{Agents and Action Streams}
 
@@
 
-We also use shared plans to verbalize the sequence of actions and to explain
+We also use shared plans to verbalise the sequence of actions and to explain
 the human partner how a task may be shared~\cite{warnier2012when}.
@@
 
-\paragraph{Action costs and social rules}
+\subsubsection{Action Costs and Social Rules}
 
@@
 
-\caption{An alternative plan where the \emph{wasted time} social rule is used to optimize the
+\caption{An alternative plan where the \emph{wasted time} social rule is used to optimise the
 total duration of the task. }
@@
 
-\paragraph{Several levels of cooperation} 
-By tuning its costs
-and adapting its social rules, HATP can be used to compute various
-alternative plans. These plans can be categorized into several levels
-of cooperation
+\subsubsection{Commitment Levels}
+
+By tuning its costs and adapting its social rules, HATP can be used to compute
+various alternative plans. These plans can be categorised into several levels of
+cooperation
 
@@
 
-\subsubsection{Action execution and monitoring}\label{sec:action}
+\subsubsection{Action Execution and Monitoring}\label{sec:action}
 
@@
 
-\paragraph{Human aware motion, placement and manipulation planning}
+\paragraph{Human-Aware Motion, Placement and Manipulation Planning}
 Manipulation and navigation actions listed in
@@
 
-\section{Experimental outcomes}
+\section{Experimental Support}
 \label{sec:expe}
 
-Our architecture has been deployed and tested in a number of experiments on
+Our architecture has been deployed and tested in a number of studies on
 several robotic platforms. Table~\ref{table|experiences} lists the most
@@
 \begin{tabular}{llll}
- \bf{Experiment} & Focus & \emph{Notes} & Reference \\
+ \bf{Study} & Focus & \emph{Notes} & Reference \\
 \hline
@@
 
-Because only objects marked with tags are currently recognized (in our typical
+Because only objects marked with tags are currently recognised (in our typical
 lab experiments, it may be a dozen of them), the robot grounding mechanisms work
@@
 The cognitive model that the robot builds for the agent it interacts with is
-still simple and mostly focused on geometric features (who sees what? What are
-our relative positions? etc.). Extending this knowledge with more subtle
+still simple and mostly focused on geometric features (\emph{Who sees what? What are
+our relative positions?} etc.). Extending this knowledge with more subtle
 perceptions (emotional state for instance) remains to be done.
@@
 subtlety for a robot than autonomous independent task achievement. HATP, by
-co-planning for the robot and the human, and modeling social rules as planning
+co-planning for the robot and the human, and modelling social rules as planning
 parameters, is a first step in this direction.
@@
 collaboration abilities. Several theories dealing with
-collaboration~\cite{Cohen1991,Grosz1996,Clark1996} emphasize that
+collaboration~\cite{Cohen1991,Grosz1996,Clark1996} emphasise that
 collaborative tasks have specific requirements compared to individual
 ones, \eg, since the robot and the person share a common goal, they
-have to agree on the manner to realize it, they must show their
+have to agree on the manner to realise it, they must show their
 commitment to the goal during execution, etc. Several robotic systems
@@
 questions like \emph{Is your object small or large?} Based on the knowledge it
-has acquired, the robot is able to minimize the number of questions required to
+has acquired, the robot is able to minimise the number of questions required to
 find the object.
@@
 
-\section*{Acknowledgments}
+\section*{Acknowledgements}
 

### 2014-05-06 paper.tex
Rewrite the experiment section
- let's be honest: {experiments->studies}
- rephrase the Tom and Jerry study to be more interesting

@@
 \subsubsection{Reasoning}
+\label{reasoning}
 
@@
 
-\begin{table}
+\begin{table}[ht!]
 \begin{center}
-\begin{tabular}{p{8cm}}
+\begin{tabular}{p{0.8\columnwidth}}
 \hline
@@
 
-\section{Experimental Support}
+\section{Support Studies}
 \label{sec:expe}
@@
 several robotic platforms. Table~\ref{table|experiences} lists the most
-significant ones, with their main focuses and reference publications for the
-interested reader.
+significant ones, with their main focuses and reference publications.
 
@@
 \end{center}
-\caption{Main experiments conducted with our cognitive architecture.}
-\label{table|experiences}
+\caption{Main studies conducted with our cognitive architecture.}
+\label{table|experiments}
 \end{table*}
 
-The two following sections present one such experiment (the {\it Interactive
-Grounding I} scenario), and the synopsis of another one (the {\it Cleaning the
-table} scenario).
-
-The first one is focused on knowledge representation and verbal interaction:
-the human asks for help to find and pack objects. In this experiment, robot
-actions are limited to simple motions (like head tracking or predefined
-pick-and-place).
-
-The second one involves the {\sc Shary} execution controller and the HATP symbolic
-task planner besides {\sc Oro} and {\sc Dialogs}. In this scenario, the human and the
-robot try to cooperatively remove objects from a table. behaviours and motions
-are fully planned.
-
-\subsection{\emph{Moving to London} scenario}
+We briefly introduce two of them in order to illustrate in a practical way
+different aspects of the architecture that were previously presented.  The first
+one is focused on knowledge representation and verbal interaction: the human
+asks for help to find and pack objects ({\it Interactive Grounding I} in
+Table~\ref{table|experiments}). The second one ({\it Cleaning the table})
+involves the {\sc Shary} execution controller and the HATP symbolic task planner
+besides {\sc Oro} and {\sc Dialogs}. In this scenario, the human and the robot
+try to cooperatively remove objects from a table, and behaviours and motions are
+fully planned and executed.
+
+\subsection{Interactive Grounding}
 \label{moving-london}
 
-This first experiment is based on the following daily life situation: Tom and
-Jerry are moving to London, so they are packing things in boxes. The scenario
-takes places in the living-room, where Jido (our robot) is observing while they
-move things here and there. To assess the reasoning abilities of the robot they
-ask Jido for information. Ideally, the robot should
-also perform actions when required (\eg hand an object when asking ``give
-me...'').
-
-Perception of objects is done through a tag-based system and humans are
-detected through motion capture. The robot knowledge base is pre-loaded with
-the \emph{{\sc Oro} Commonsense Ontology}.  We next describe two
+This first study is based on the following scenario: Tom and Jerry are moving to
+London, so they are packing things in boxes. \emph{Jido}, a dual-arm
+manipulator, is observing while they move things here and there
+(Figure~\ref{fig|vpt}).  They ask questions to Jido to know where certain
+objects are located (\emph{Where is the white tape? What is behind the toolbox?}
+etc.). This study focuses on multi-modal ,interactive grounding only, and
+the robot does not perform any action.
+
+Objects are perceived solely through 2D fiducial markers sticked on them, and humans are
+tracked through motion capture. The robot knowledge base is initialised with
+the {\sc Oro} commonsense ontology.  We next describe two
 situations where we can follow the internal robot's reasoning and the
@@
 
-\paragraph{Implicit disambiguation through visual perspective taking}
+\paragraph{Implicit Disambiguation Through Visual Perspective Taking}
 
@@
   \centering
-  \includegraphics[width=0.9\linewidth]{pt.jpg}
-\caption{Asking the robot to hand over a tape on the table.}
+  \subfigure [Interactive grounding in a cluttered environment.]{
   \label{fig|vpt}
+  \includegraphics[width=0.48\linewidth]{pt.jpg}
+}%
+\subfigure [Disambiguation through pointing.]{
+  \label{fig|pointing}
+  \includegraphics[width=0.48\linewidth]{inTheBox2.jpg}
+}
 \end{figure}
@@
 approaches the table and asks Jido to hand him the video tape: ``Jido, can
-you give me the video tape''. The \textsc{Dialogs} module queries the ontology to
+you give me the video tape''. The \textsc{Dialogs} module processes this
+sentence, and queries the ontology to
 identify the object the human is referring to: \stmt{?obj type VideoTape}.
@@
 
-However, only one is visible for Tom (the one on the
-table). Thus, although there is an ambiguity from the robot's perspective
-(since it can see both video tapes), based on the perspective of its human
-partner it infers that Tom is referring to the video tape on the table, and not
-the one inside the box which is not visible for him.
-
-Since only one object remains (\ie the referent that \concept{VideoTape} stands
-for is not ambiguous anymore), the robot infers that the human refers to it and
-eventually execute the command, \ie give it to the human. Alternatively, the
-robot first asks the human for confirmation (``was \concept{videoTape1} the
-object you had in mind?'') before proceeding with the action.
-Table~\ref{table|ptbeliefs} lists the robot's beliefs about itself and its human
-partner involved in this situation.
-
-\begin{table}[ht!]
-\begin{center}
-    \subfigure{
-\begin{tabular}{l}
-\hline
-Robot's beliefs about itself (\emph{robot's model}):\\
-\hline
-  \hspace{0.7cm}\stmt{videoTape1 type VideoTape}\\
-  \hspace{0.7cm}\stmt{videoTape1 isOn table}\\
-  \hspace{0.7cm}\stmt{videoTape1 isVisible true}\\
-  \hspace{0.7cm}\stmt{videoTape2 type VideoTape}\\
-  \hspace{0.7cm}\stmt{videoTape2 isIn cardBoardBox}\\
-  \hspace{0.7cm}\stmt{videoTape2 isVisible true}\\
-\hline
-\end{tabular}
-}\hspace{1em}%
-    \subfigure{
-\begin{tabular}{l}
-\hline
-Robot's beliefs about Tom (\emph{Tom's model}):\\
-\hline
-  \hspace{0.7cm}\stmt{videoTape1 type VideoTape}\\
-  \hspace{0.7cm}\stmt{videoTape1 isOn table}\\
-  \hspace{0.7cm}\stmt{videoTape1 isVisible true}\\
-  \hspace{0.7cm}\stmt{videoTape2 type VideoTape}\\
-  \hspace{0.7cm}\stmt{videoTape2 isIn cardBoardBox}\\
-  \hspace{0.7cm}\stmt{videoTape2 isVisible false}\\
- \hline
-\end{tabular}
-}
-\end{center}
-\caption{Robot's beliefs about itself and its human partner.}
-\label{table|ptbeliefs}
-\end{table}
-
-\paragraph{Explicit disambiguation through verbal interaction and gestures}
-
-\begin{figure}[!ht]
-  \centering
-  \includegraphics[width=0.9\linewidth]{inTheBox2.jpg}
-\caption{Jerry asks Jido for the content of the box by pointing at it.}
-  \label{fig|pointing}
-\end{figure}
-
-In this situation, Jerry enters the living room without knowing where Tom had
+However, only one is visible for Tom (the one on the table). Although there is
+an ambiguity from the robot's perspective, the human referred to the tape using
+the definite article \emph{the}: this is interpreted by the natural language
+processor as the human referring to a precise object, in that case, the visible
+one.\footnote{Other heuristics are available to the {\sc Dialogs} module: for
+instance, if a tape had been recently mentioned in the dialogue, this instance
+would have been selected instead as the referent.}
+
+%Since only one object remains (\ie the referent that \concept{VideoTape} stands
+%for is not ambiguous any more), the robot infers that the human refers to it and
+%eventually execute the command, \ie give it to the human. Alternatively, the
+%robot first asks the human for confirmation (``was \concept{videoTape1} the
+%object you had in mind?'') before proceeding with the action.
+%Table~\ref{table|ptbeliefs} lists the robot's beliefs about itself and its human
+%partner involved in this situation.
+%
+%\begin{table}[ht!]
+%\begin{center}
+%    \subfigure{
+%\begin{tabular}{l}
+%\hline
+%Robot's beliefs about itself (\emph{robot's model}):\\
+%\hline
+%  \hspace{0.7cm}\stmt{videoTape1 type VideoTape}\\
+%  \hspace{0.7cm}\stmt{videoTape1 isOn table}\\
+%  \hspace{0.7cm}\stmt{videoTape1 isVisible true}\\
+%  \hspace{0.7cm}\stmt{videoTape2 type VideoTape}\\
+%  \hspace{0.7cm}\stmt{videoTape2 isIn cardBoardBox}\\
+%  \hspace{0.7cm}\stmt{videoTape2 isVisible true}\\
+%\hline
+%\end{tabular}
+%}\hspace{1em}%
+%    \subfigure{
+%\begin{tabular}{l}
+%\hline
+%Robot's beliefs about Tom (\emph{Tom's model}):\\
+%\hline
+%  \hspace{0.7cm}\stmt{videoTape1 type VideoTape}\\
+%  \hspace{0.7cm}\stmt{videoTape1 isOn table}\\
+%  \hspace{0.7cm}\stmt{videoTape1 isVisible true}\\
+%  \hspace{0.7cm}\stmt{videoTape2 type VideoTape}\\
+%  \hspace{0.7cm}\stmt{videoTape2 isIn cardBoardBox}\\
+%  \hspace{0.7cm}\stmt{videoTape2 isVisible false}\\
+% \hline
+%\end{tabular}
+%}
+%\end{center}
+%\caption{Robot's beliefs about itself and its human partner.}
+%\label{table|ptbeliefs}
+%\end{table}
+%
+\paragraph{Explicit Disambiguation Through Verbal Interaction and Gestures}
+
+In this second situation, Jerry enters the living room without knowing where Tom had
 placed the video tapes (Figure~\ref{fig|pointing}). So he first asks Jido:
@@
 figure out which box Jerry is talking about. Similar to the previous situation,
-there are two available boxes:
+two boxes are visible:
 
@@
 
-However both are visible and the cognitive ambiguity resolution cannot be
-applied. The only option is to ask Jerry which box he is referring to: ``Which
-box, the toolbox or the cardboard box?'' Jerry could answer the
-question. Instead, he decides to point at it while indicating: ``This box''
-(Figure~\ref{fig|pointing}). The robot's perception identifies the {\tt
+However both are also visible to the human and the previous ambiguity resolution
+procedure can not be applied. The robot generates a question (using the
+\emph{Discrimination} algorithm that we introduced at section~\ref{reasoning})
+and asks Jerry which box he is referring to: ``Which box, the toolbox or the
+cardboard box?'' Jerry can answer the question, but he instead decides to point
+at it: ``This box'' (Figure~\ref{fig|pointing}). {\sc Spark} identifies the {\tt
 cardBoardBox} as being pointed at and looked at by the human and updates the
-ontology with this new information using a rule available in the commonsense
-ontology ({\tt \textbf{pointsAt}(?ag, ?obj) $\land$ \textbf{looksAt}(?ag, ?obj) $\to$
-\textbf{focusesOn}(?ag, ?obj)}). The \textsc{Dialogs} module is then able to merge both
+ontology with this new information using a rule available in the common-sense
+ontology \stmt{pointsAt(?ag, ?obj) $\land$ looksAt(?ag, ?obj) $\to$
+focusesOn(?ag, ?obj)}. The \textsc{Dialogs} module is then able to merge both
 sources of information, verbal (``this'') and deictic to distinguish the box
 Jerry refers to:
+
 \begin{center} 
@@
 Finally, the \textsc{Dialogs} queries the ontology about the content of the box
-and the question can be answered: ``Jido-E''. Note that the object's label is
-used instead of its ID. This way we enhance interaction using familiar names
-given by the users:
-
-\begin{center}
-\stmt{?obj isIn cardBoardBox}\\
-\hspace{0.7cm}$\Rightarrow$ \concept{?obj = videoTape2}\\
-\end{center}
-
-At this point Jerry wants to know where the other tape is, and that is exactly
-what he asks Jido: ``And where is the other tape?''. In this situation, the
-\textsc{Dialogs} module is able to interpret that Jerry is not referring to the
-video which they were just talking about, but to the other one:
+and the question can be answered: ``Wall-E'' (the label of the object, stored in the
+ontology, is used to interact with the user).
+
+%\begin{center}
+%\stmt{?obj isIn cardBoardBox}\\
+%\hspace{0.7cm}$\Rightarrow$ \concept{?obj = videoTape2}\\
+%\end{center}
+
+At this point Jerry wants to know where is the other tape: ``And where is the
+other tape?''. \textsc{Dialogs} processes this sentence using the
+\concept{differentFrom} OWL predicate:
 
@@
 
-Since there is only one possible ``other'' video (there are only two videos in
-the scene), it can directly answer Jerry: ``The other tape is on the table and
-next to the toolbox.''
+Since there is only one possible ``other'' video, no specific disambiguation is
+required, and the robot verbalises the result of the query to the knowledge
+base: ``The other tape is on the table and next to the toolbox.''
 

### 2014-05-06 paper.tex
Improve discussion introduction, reshuffle its organisation

@@
 
-\section{Discussion: when Artificial Intelligence enables Human-Robot
+\section{Discussion: When Artificial Intelligence Enables Human-Robot
 Interaction}
@@
 
-This last section summarises our main contributions and tries to clarify the ``take home'' message: what are the challenges
+This last section tries to clarify the ``take home'' message: what are the challenges
 that human-robot interaction brings to artificial intelligence, and what is the
@@
 
-\subsection{The Key Issues}
-
-We would like first to restate the main challenges to artificial intelligence
-raised by human-robot interaction.
-
-\fxfatal{TDB}
-
-\subsubsection{Embodied Cognition}
+We first restate the challenge of \emph{embodied cognition} and list what we
+seen as the main decisional issues that need to be tackled to push forward
+human-robot interaction. We then reflect on the importance of knowledge handling
+in modern robotics and put our choice of the RDF formalism into perspective.
+Finally, we briefly re-discuss two facets of HRI: the importance of
+\emph{modelling the humans}, and the difficulty of \emph{acting in presence of
+humans}.
+
+\subsection{Embodied Cognition}
 
@@
 
+\subsection{The Key Decisional Issues}
+
+How can we restate the main challenges, in particular in term of decision
+making, that human-robot interaction brings to artificial intelligence?
+
+\fxfatal{TDB}
+
 \subsection{Knowledge and Robotics}
 \label{krs-discussion}
+
+\subsubsection{Knowledge and Architecture}
+
+As a whole, the components that we have introduced so far build a
+knowledge-oriented architecture: Knowledge is explicitly stored in one central
+and consistent repository of facts, accessible for all modules. It relies on a
+\emph{strict formalism} (OWL statements), with a \emph{well defined vocabulary}
+(stated in the common-sense ontology). These first two points lead to a
+\emph{loosely-coupled architecture} where modules can be removed or replaced
+easily by other ones as long as they share the same semantics (modules are
+defined by the knowledge they produce).
+
+Also, we adopt a symbolic, reactive, event-driven approach to robot control. By
+managing events at the same level as the reasoner, we take full advantage of the
+inference abilities of {\sc Oro} to trigger events whose \texttt{true}
+conditions can be inferred.
+
+And finally, this architecture allows for the combination of very different
+knowledge modalities in a single homogeneous environment, bringing mutual
+benefits to components. For instance, the dialogue processing module can
+perfectly run without any geometric perception, but its disambiguation routines
+can transparently benefit from it when available (since richer symbolic
+descriptions of objects are then available).
+
+Those items are not new \emph{per se}. It is however interesting to underline
+the shift of focus this implies during the design and integration phases of
+robots: components of our deliberative layer are defined and glued together by
+the knowledge they produce and consume. Human-robot interaction, because it
+supposes operations at \emph{human-level} and in environments with complex
+semantics, acts here as a motivational force.
 
@@
 the robot.
-
-\subsubsection{Knowledge and Architecture}
-
-As a whole, the components that we have introduced so far build a
-knowledge-oriented architecture: Knowledge is explicitly stored in one central
-and consistent repository of facts, accessible for all modules. It relies on a
-\emph{strict formalism} (OWL statements), with a \emph{well defined vocabulary}
-(stated in the common-sense ontology). These first two points lead to a
-\emph{loosely-coupled architecture} where modules can be removed or replaced
-easily by other ones as long as they share the same semantics (modules are
-defined by the knowledge they produce).
-
-Also, we adopt a symbolic, reactive, event-driven approach to robot control. By
-managing events at the same level as the reasoner, we take full advantage of the
-inference abilities of {\sc Oro} to trigger events whose \texttt{true}
-conditions can be inferred.
-
-And finally, this architecture allows for the combination of very different
-knowledge modalities in a single homogeneous environment, bringing mutual
-benefits to components. For instance, the dialogue processing module can
-perfectly run without any geometric perception, but its disambiguation routines
-can transparently benefit from it when available (since richer symbolic
-descriptions of objects are then available).
-
-Those items are not new \emph{per se}. It is however interesting to underline
-the shift of focus this implies during the design and integration phases of
-robots: components of our deliberative layer are defined and glued together by
-the knowledge they produce and consume. Human-robot interaction, because it
-supposes operations at \emph{human-level} and in environments with complex
-semantics, acts here as a motivational force.
 

### 2014-05-06 paper.tex
Tikzified the 'manip_run' figure
While here, indicate on the figure the flow of the experiment

@@
   \centering
-    \includegraphics[width=1.0\textwidth]{manip_run.pdf} \\
+
+\renewcommand{\stmt}[1]{{\footnotesize \tt  #1}}
+\resizebox{\textwidth}{!}{%
+\begin{tikzpicture}[
+        >=latex,
+        box/.style={draw,rectangle, dotted, minimum width=#1, minimum height=3.8cm},
+        box/.default={4.5cm},
+        every edge/.style={draw, ultra thick, ->},
+        every node/.style={align=center},
+        robot/.style={fill=RoyalBlue!50},
+        plan/.style={draw,
+                     thick,  
+                     circle, 
+                     font=\sf,
+                     align=center,
+                     fill=BurntOrange!50, 
+                     minimum size=1cm, 
+                     inner sep=0.1cm}]
+
+        \coordinate (figbottom) at (-0.5, -28.5);
+
+        \path (-0.5,0) edge (figbottom) node[sloped, above left, rotate=90] {time};
+
+        \node at (2,0) (percept) {\bf Perception};
+            \node[below=0.5 of percept.south west, anchor=mid] {camera};
+            \node[below=0.5 of percept.south east, anchor=mid] {3D model};
+        \node[right=4 of percept, minimum width=2.5cm] (kb) {\bf Knowledge};
+            \node[below=0.5 of kb.south west, anchor=mid east] (kbr) {robot};
+            \node[below=0.5 of kb.south east, anchor=mid west] (kbh) {human};
+            \draw[dotted] (kb.south) to (figbottom -| kb);
+        \node[right=4 of kb] (plan) {\bf Plans};
+            \node[below=0.5 of plan.south west, anchor=mid east] (probot) {robot};
+            \node[below=0.5 of plan.south east, anchor=mid west] (phuman) {human};
+        \node[right=5 of plan] (action) {\bf Actions};
+            \node[below=0.5 of action.south west, anchor=mid east] (arobot) {robot};
+            \node[below=0.5 of action.south east, anchor=west] (ahuman) {human\\(monitoring)};
+
+        \draw[dashed] (-0.6,-1.2) --(24, -1.2);
+
+        \node[anchor=east] at (-0.5, -3) (t1) {$t_1$};
+        \node[anchor=east, below=6 of t1] (t2) {$t_2$};
+        \node[anchor=east, below=6 of t2] (t3) {$t_3$};
+        \node[anchor=east, below=4 of t3] (t4) {$t_4$};
+        \node[anchor=east, below=5 of t4] (t5) {$t_5$};
+
+        %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+        %%% PERCEPTIONS
+        %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+        \node at (t1 -| percept) (cam1) {\includegraphics[height=2cm]{manip_run_cam1.png}};
+        \node at (t2 -| percept) (cam2) {\includegraphics[height=2cm]{manip_run_cam2.png}};
+        \node at (t3 -| percept) (cam3) {\includegraphics[height=2cm]{manip_run_cam3.png}};
+        \node at (t4 -| percept) (cam4) {\includegraphics[height=2cm]{manip_run_cam4.png}};
+        \node at (t5 -| percept) (cam5) {\includegraphics[height=2cm]{manip_run_cam5.png}};
+
+        %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+        %%% KNOWLEDGE
+        %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+        \node[align=left] at (cam1 -| kbr) (kb1) {\stmt{TAPE isReachable true}\\
+                                                        \stmt{TAPE isVisible true} \\
+                                                        \stmt{TAPE isOn TABLE}};
+
+        \node[align=left] at (cam1 -| kbh) {\stmt{TAPE isReachable false}\\
+                                                        \stmt{TAPE isVisible true}};
+
+        \node[align=left] at (cam2 -| kbr) (kb2) {\stmt{ROBOT hasInHand TAPE}};
+
+        \node[align=left] at (cam3 -| kbr) {\stmt{TAPE isReachable true}\\
+                                            \stmt{TAPE isVisible true} \\
+                                            \stmt{TAPE isOn TABLE}};
+
+        \node[align=left ] at (cam3 -| kbh) (kb3) {\stmt{TAPE isReachable true}\\
+                                             \stmt{TAPE isVisible true}};
+
+        \node[align=left] at (cam4 -| kbh) (kb4) {\stmt{HUMAN hasInHand TAPE}};
+
+        \node[align=left] at (cam5 -| kbr)  {\stmt{TAPE isIn BIN}};
+        \node[align=left] at (cam5 -| kbh) (kb5) {\stmt{TAPE isIn BIN}};
+
+        %%%%%%%%%%%%%%%% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+        %%% PLANS
+        %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+        \node[below=1 of plan] (incoming) {\bf \Large Incoming plan \\ \it Clean the table!};
+
+        \node[anchor=north, plan, robot] at (cam1.south -| probot) (pr1) {\action{TAKE}{GREY\_TAPE}{TABLE}};
+        \node[anchor=north, plan, robot] at (cam2.south -| probot)  (pr2) {\action{PUTRV}{GREY\_TAPE}{TABLE}} edge[<-] (pr1);
+        \node[anchor=north, plan] at (cam3.south -| phuman) (ph1) {\action{TAKE}{GREY\_TAPE}{TABLE}} edge[<-] (pr2);
+        \node[anchor=north, plan]  at (cam4.south -| phuman) (ph2) {\action{PUT}{GREY\_TAPE}{BIN}} edge[<-] (ph1);
+
+        \node[anchor=north] at (cam5.south -| plan) (done) {\bf \Large Goal completed};
+
+        %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+        %%% ACTIONS
+        %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+        \node at (pr1 -| arobot) (ep1) {\it evaluate preconditions};
+
+        \node[below=0.1 of ep1] (mhp1) {%
+            \begin{tikzpicture}
+                \node (title) {\bf motion planning};
+                \node[below=0.1 of title.south west, label=below:{\tt PICK\_GOTO}] (mapg) {\includegraphics{MHP_ARM_PICK_GOTO}};
+                \node[right=of mapg, label=below:{\tt TAKE\_TO\_FREE}] (mattf) {\includegraphics{MHP_ARM_TAKE_TO_FREE}} edge[<-] (mapg);
+            \end{tikzpicture}
+        };
+        \node[below=0.1 of mhp1] (me1) {\bf motion execution};
+        \node[below=0.1 of me1] (ae1) {\it assess effects};
+
+        %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+        \node at (pr2 -| arobot) (ep2) {\it evaluate preconditions};
+
+        \node[below=0.1 of ep2] (mhp2) {%
+            \begin{tikzpicture}
+                \node (title) {\bf motion planning};
+                \node[below=0.1 of title, label=below:{\tt ESCAPE}] (maeo) {\includegraphics{MHP_ARM_ESCAPE_OBJECT}};
+                \node[left=of maeo, label=below:{\tt PLACE\_FROM\_FREE}] (mapff) {\includegraphics{MHP_ARM_PLACE_FROM_FREE}} edge (maeo);
+                \node[right=of maeo, label=below:{\tt TO\_FREE}] (maf) {\includegraphics{MHP_ARM_FREE}} edge[<-] (maeo);
+            \end{tikzpicture}
+        };
+        \node[below=0.1 of mhp2] (me2) {\bf motion execution};
+        \node[below=0.1 of me2] (ae2) {\it assess effects};
+
+        %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+        \node[anchor=north] at (ph1 -| ahuman) (wait1) {%
+            \begin{tikzpicture}
+                \node (title) {\it wait for pick\\ {\tt TAPE} from {\tt TABLE}};
+                \node[below=0.1 of title] {\includegraphics{wait_for_pick}};
+            \end{tikzpicture}
+        };
+
+        \node[anchor=north] at (ph2 -| ahuman) (wait2) {%
+            \begin{tikzpicture}
+                \node (title) {\it wait for put\\ {\tt TAPE} to {\tt BIN}};
+                \node[below=0.1 of title] {\includegraphics{wait_for_throw}};
+            \end{tikzpicture}
+        };
+
+        %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+        %%% FLOW
+        %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+        \draw[dotted, ->, in=45, out=180] (incoming.west) to (kb1);
+        \draw[dotted, ->, bend right] (kb1) to (pr1);
+        \draw[dotted, ->, bend left] (pr1) to (ep1);
+        \draw[dotted, ->, in=45, out=180] (ae1.west) to (kb2);
+        \draw[dotted, ->, bend right] (kb2) to (pr2);
+        \draw[dotted, ->, bend left] (pr2) to (ep2);
+        \draw[dotted, ->, in=45, out=180] (ae2.west) to (kb3);
+        \draw[dotted, ->, bend right] (kb3) to (ph1);
+        \draw[dotted, ->, bend left] (ph1) to (wait1);
+        \draw[dotted, ->, in=-20, out=180] (wait1.west) to (kb4.east);
+        \draw[dotted, ->, bend right] (kb4) to (ph2);
+        \draw[dotted, ->, bend left] (ph2) to (wait2);
+        \draw[dotted, ->, in=20, out=180] (wait2.west) to (kb5);
+        \draw[dotted, ->, bend left] (kb5.east) to (done.north);
+
+ \end{tikzpicture}
+ }
  \caption {Sequence of events of a simplified version of the \emph{Cleaning the table} scenario.}

### 2014-05-06 paper.tex
Rewrite the 'Clean the table' study

@@
 
-\begin{tabular}{llll}
- \bf{Study} & Focus & \emph{Notes} & Reference \\
+\begin{tabular}{lll}
+ \bf{Study} & Focus & Reference \\
 \hline
-{\it Point \& Learn} (2010) & Interactive knowledge acquisition & & \cite{Lemaignan2010} \\
-{\it Spy Game} (2010) & Interactive object discrimination & & \cite{Ros2010b} \\
-{\it Interactive Grounding I} (2011) & Multi-modal interaction, perspective taking & \emph{immobile robot} & \cite{lemaignan2011what} \\
-{\it Roboscopie} (2011) & Theater, reflection on the future of HRI & \emph{semi-autonomous} & \cite{lemaignan2012roboscopie} \\
-{\it Cleaning the table} (2011) & Whole stack integration & & \cite{Alami2011} \\
-{\it I'm in your shoes} (2012) & False beliefs & & \cite{warnier2012when} \\
-{\it Give me this} (2012) & Natural joint object manipulation & & \cite{gharbi2013natural} \\
-{\it Interactive Grounding II} (2012) & Multi-modal interaction, perspective taking & & \cite{lemaignan2013talking} \\
+{\it Point \& Learn} (2010) & Interactive knowledge acquisition & \cite{Lemaignan2010} \\
+{\it Spy Game} (2010) & Interactive object discrimination & \cite{Ros2010b} \\
+{\it Interactive Grounding I} (2011) & Multi-modal interaction, perspective taking & \cite{lemaignan2011what} \\
+{\it Roboscopie} (2011) & Theater, reflection on the future of HRI & \cite{lemaignan2012roboscopie} \\
+{\it Cleaning the table} (2011) & Whole stack integration & \cite{Alami2011} \\
+{\it I'm in your shoes} (2012) & False beliefs & \cite{warnier2012when} \\
+{\it Give me this} (2012) & Natural joint object manipulation & \cite{gharbi2013natural} \\
+{\it Interactive Grounding II} (2012) & Multi-modal interaction, perspective taking & \cite{lemaignan2013talking} \\
 \hline
@@
 
-\subsection{\emph{Cleaning the table} scenario}
-
-This experiment involved a more complex decisional layer where the {\sc Oro} server
-was used in conjunction with the HATP symbolic task planner and the {\sc Shary}
-execution controller (the complete software architecture we deployed for this
-experiment is the one pictured in Figure~\ref{fig|archi},
-page~\pageref{fig|archi}).
-
-In this scenario (Figure~\ref{fig|cleantable-video}), a human and a robot
-cooperate to remove objects from a table. The robot produces symbolic plans for
-both itself and the human (Figure~\ref{plan-etat2})
-that allow the robot to (verbally) share the task with the human (like ``I take
-the green box and I put it in the trashbin, you take the black video tape and
-you throw it in the trashbin''). Plans are created based on perceived
-visibility and reachability of the objects, and the robot also monitors the
-human activities to track the advancement of the whole plan.
-
-\begin{figure}
+\subsection{\emph{Cleaning the table} study}
+
+\begin{figure}[ht!]
     \centering
-    \includegraphics[width=0.7\columnwidth]{cleantable.jpg}
-
-    \caption{Snapshot of the film of the ``Clean the Table'' scenario. The
-    physical situation, the {\sc Spark} model, and the current step of the plan can
-    be seen.}
+    \includegraphics[width=0.6\columnwidth]{cleantable.jpg}
+
+    \caption{Face-to-face setting of the \emph{Clean the Table} scenario. The
+    physical situation, the {\sc Spark} model, and the current step of the plan
+    are visible on the picture.}
 
@@
 
-\begin{figure*}[thpb]
-  \centering
-\resizebox{0.9\textwidth}{!}{%
-\begin{tikzpicture}[
-    >=latex,
-    robot/.style={fill=RoyalBlue!50},
-    every edge/.style={<-, draw, ultra thick},
-    every node/.style={ draw, 
-                        thick,  
-                        circle, 
-                        font=\sf,
-                        align=center,
-                        node distance=2cm,
-                        fill=BurntOrange!50, 
-                        minimum size=2cm, 
-                        inner sep=0.1cm}]
-
-    \node[font=\rm\large, draw=none, fill=none, rotate=90] at (-2, 0) {Human\\actions};
-    \node[font=\rm\large, draw=none, fill=none, rotate=90] at (-2, -3.5) {Robot\\actions};
-
-    \node (h1) {\action{TAKE}{WALLE\_TAPE}{TABLE}};
-    \node[right=of h1] (h2) {\action{THROW}{WALLE\_TAPE}{TRASH\_1 }} edge (h1);
-
-    \node[robot, below=1.5 of h1] (r1) {\action{TAKE}{GREY\_TAPE}{TABLE }};
-    \node[robot, right=of r1] (r2) {\action{PUTRV}{GREY\_TAPE}{TABLE}} edge (r1);
-    \node[robot, right=of r2] (r3) {\action{TAKE}{LOTR\_TAPE}{TABLE }} edge (r2);
-    \node[robot, right=of r3] (r4) {\action{PUTRV}{LOTR\_TAPE}{TABLE }} edge (r3);
-
-    \node[right=of h2] (h3) {\action{TAKE}{GREY\_TAPE}{TABLE}} edge (h2) edge (r2);
-    \node[right=of h3] (h4) {\action{THROW}{GREY\_TAPE}{TRASH\_1}} edge (h3);
-    \node[right=of h4] (h5) {\action{TAKE}{LOTR\_TAPE}{TABLE}} edge (h4) edge (r4);
-    \node[right=of h5] (h6) {\action{THROW}{LOTR\_TAPE}{TRASH\_1}} edge (h5);
-
-\end{tikzpicture}
-}
-
-  \caption {A plan produced by HATP to execute the high-level order ``clean the
-  table''. Two streams of actions are generated: for the human (top) and for the
-  robot (bottom). Synchronization points ensure the coordination.}
-
-  \label{plan-etat2}
-\end{figure*}
-
-Figure~\ref{fig|cleantable-timeline} presents a sample of the whole task and
-shows how the different components produce and use symbolic knowledge.
-
-This sequence of events depicts a run with only one tape on the table. The tape is
-reachable by the robot only, while the human is the one who can reach the
-trashbin to throw objects.
+This study demonstrates a richer decision-making process. The {\sc Oro} server
+is used in conjunction with the HATP symbolic task planner and the {\sc Shary}
+execution controller to produce and execute a shared plan.
+
+%In this scenario (Figure~\ref{fig|cleantable-video}), a human and a robot
+%cooperate to remove objects from a table. The robot produces symbolic plans for
+%both itself and the human (Figure~\ref{plan-etat2})
+%that allow the robot to (verbally) share the task with the human (like ``I take
+%the green box and I put it in the trashbin, you take the black video tape and
+%you throw it in the trashbin''). Plans are created based on perceived
+%visibility and reachability of the objects, and the robot also monitors the
+%human activities to track the advancement of the whole plan.
+
+%
+%\begin{figure*}[thpb]
+%  \centering
+%\resizebox{0.9\textwidth}{!}{%
+%\begin{tikzpicture}[
+%    >=latex,
+%    robot/.style={fill=RoyalBlue!50},
+%    every edge/.style={<-, draw, ultra thick},
+%    every node/.style={ draw, 
+%                        thick,  
+%                        circle, 
+%                        font=\sf,
+%                        align=center,
+%                        node distance=2cm,
+%                        fill=BurntOrange!50, 
+%                        minimum size=2cm, 
+%                        inner sep=0.1cm}]
+%
+%    \node[font=\rm\large, draw=none, fill=none, rotate=90] at (-2, 0) {Human\\actions};
+%    \node[font=\rm\large, draw=none, fill=none, rotate=90] at (-2, -3.5) {Robot\\actions};
+%
+%
+%    \node (h1) {\action{TAKE}{WALLE\_TAPE}{TABLE}};
+%    \node[right=of h1] (h2) {\action{THROW}{WALLE\_TAPE}{TRASH\_1 }} edge (h1);
+%
+%    \node[robot, below=1.5 of h1] (r1) {\action{TAKE}{GREY\_TAPE}{TABLE }};
+%    \node[robot, right=of r1] (r2) {\action{PUTRV}{GREY\_TAPE}{TABLE}} edge (r1);
+%    \node[robot, right=of r2] (r3) {\action{TAKE}{LOTR\_TAPE}{TABLE }} edge (r2);
+%    \node[robot, right=of r3] (r4) {\action{PUTRV}{LOTR\_TAPE}{TABLE }} edge (r3);
+%
+%    \node[right=of h2] (h3) {\action{TAKE}{GREY\_TAPE}{TABLE}} edge (h2) edge (r2);
+%    \node[right=of h3] (h4) {\action{THROW}{GREY\_TAPE}{TRASH\_1}} edge (h3);
+%    \node[right=of h4] (h5) {\action{TAKE}{LOTR\_TAPE}{TABLE}} edge (h4) edge (r4);
+%    \node[right=of h5] (h6) {\action{THROW}{LOTR\_TAPE}{TRASH\_1}} edge (h5);
+%
+%
+%\end{tikzpicture}
+%}
+%
+%  \caption {A plan produced by HATP to execute the high-level order ``clean the
+%  table''. Two streams of actions are generated: for the human (top) and for the
+%  robot (bottom). Synchronization points ensure the coordination.}
+%
+%  \label{plan-etat2}
+%\end{figure*}
+%
+Figure~\ref{fig|cleantable-timeline} presents a practical sample of the whole
+task. It depicts a run with a single tape on a table. The tape is
+reachable by the robot only, while the bin where objects are supposed to be
+placed is reachable by the human only: the robot needs to come up with a shared
+plan that involves joint actions.
+
+The goal is first received by the execution controller (after processing of the
+user request by the {\sc Dialogs} module, not shown on the figure). At
+$t_1$, the tape is computed by the robot as being reachable by the robot only
+(columns \emph{Perception} and \emph{Knowledge}), and the execution controller
+invokes the task planner, which produces a joint plan (column \emph{Plan}).
+
+The first task ({\tt TAKE(GREY\_TAPE, TABLE)}) is instantiated by calling the 3D
+motion planner (column \emph{Actions}, left). The motion planner returns two atomic
+actions ({\tt PICK\_GOTO} followed by {\tt TAKE\_TO\_FREE}) that the controller
+executes (by first reaching the object, grasping it and bringing it back to a
+free position).  Once the robot's perception allows the controller to ensure
+that the task's post-conditions are fulfilled (at $t_2$, by satisfying the
+statement \stmt{ROBOT hasInHand TAPE}), the next task is started (placing the
+tape so that it becomes reachable by the human).
+
+At $t_3$, the tape is now reachable by the human, and the next tasks (taking the
+tape and placing it in the bin) have to be performed by the human: the robot
+instructs the user to do so (not pictured on the figure) and monitors the
+actions of the human to detect when the tasks' post-conditions are satisfied
+(column \emph{Actions}, right). When these post-conditions are fulfilled, the
+goal is considered to be completed.
+
+This simple example illustrates how the symbolic facts are produced from the
+situation assessment, and, in parallel, used by the execution controller to
+assess the overall progress of the plan.
 
@@
             \draw[dotted] (kb.south) to (figbottom -| kb);
-        \node[right=4 of kb] (plan) {\bf Plans};
+        \node[right=4 of kb] (plan) {\bf Plan};
             \node[below=0.5 of plan.south west, anchor=mid east] (probot) {robot};
@@
 
-        \node at (pr1 -| arobot) (ep1) {\it evaluate preconditions};
+        \node at (pr1 -| arobot) (ep1) {\it evaluate pre-conditions};
 
@@
         \node[below=0.1 of mhp1] (me1) {\bf motion execution};
-        \node[below=0.1 of me1] (ae1) {\it assess effects};
+        \node[below=0.1 of me1] (ae1) {\it assess post-conditions};
 
         %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-        \node at (pr2 -| arobot) (ep2) {\it evaluate preconditions};
+        \node at (pr2 -| arobot) (ep2) {\it evaluate pre-conditions};
 
@@
         \node[below=0.1 of mhp2] (me2) {\bf motion execution};
-        \node[below=0.1 of me2] (ae2) {\it assess effects};
+        \node[below=0.1 of me2] (ae2) {\it assess post-conditions};
 
@@
  }
- \caption {Sequence of events of a simplified version of the \emph{Cleaning the table} scenario.}
+ \caption {Chronology of the \emph{Cleaning the table} scenario. The light
+     arrows indicate the general flow. This figure
+     represents a simplified version: only one object, {\tt TAPE}, must
+     be removed from the table and placed in the bin {\tt BIN}.  }
   \label{fig|cleantable-timeline}
@@
 
-The plan produced in that case by HATP is straightforward and shown in the
-third row. It consists of four successive actions involving the robot and the
-human. Robot grasps the tape and then places it on the table at a position
-where it is visible and reachable for the human. The human is then asked to pick
-the tape and throw it in the trashbin.
+%The plan produced in that case by HATP is straightforward and shown in the
+%third row. It consists of four successive actions involving the robot and the
+%human. Robot grasps the tape and then places it on the table at a position
+%where it is visible and reachable for the human. The human is then asked to pick
+%the tape and throw it in the trashbin.
 

### 2014-05-06 paper.tex
Mention we use OpenGL occlusion checks to compute visibility

@@
 \item \emph{Sees}: An important ability to know about an agent is to predict
-\emph{What can it see?}, \ie what is within its field of view (FOV). Being able
-to compute this information, the robot can reuse it for instance to infers
-which object a human is searching for (the one that is not currently visible,
-otherwise the user would not be searching for it).  In
-figure~\ref{fig|spark-oro}\emph{c} the field of view of a person is
-illustrated with a grey cone (broader one). While he is able to see the two
-small boxes on the table in front of him, the big box on his right is out of
-his FOV, and therefore, he is not able to see it.
+    \emph{What can it see?}, \ie what is within its field of view (FOV). Being
+    able to compute this information (by dynamically placing an OpenGL camera at
+    the location of the eyes and running occlusion checks from it), the robot
+    can reuse it for instance to infers which object a human is searching for
+    (the one that is not currently visible, otherwise the user would not be
+    searching for it).  In figure~\ref{fig|spark-oro}\emph{c} the field of view
+    of a person is illustrated with a grey cone (broader one). While he is able
+    to see the two small boxes on the table in front of him, the big box on his
+    right is out of his FOV, and therefore, he is not able to see it.
 
@@
     narrower field of view, the field of attention (FOA).
-Figure~\ref{fig::sparkRepresentations}\emph{c}
-shows the field of attention of a person with a green cone (narrower one). In
-this example only the grey box satisfies the \concept{looksAt} relation.
+    Figure~\ref{fig::sparkRepresentations}\emph{c} shows the field of attention
+    of a person with a green cone (narrower one). In this example only the grey
+    box satisfies the \concept{looksAt} relation.
 

### 2014-05-06 rebuttal.tex
Address more comments in the rebutta letter

@@
 
-\emph{1. The novelty of this paper with respect to other papers by the same
-authors, including one in IJSR 2011, has been questioned. The authors should
-convincingly explain what are the novel contributions and the added value of
-this work compared to the previous publications.}
+\begin{itemize}
 
-We address this question with our reponse to reviewer 3, below.
-
-\emph{2. The form and depth of the presentation are below the standards of an
+    \item \emph{The form and depth of the presentation are below the standards of an
 AIJ article.  The reviewers made many suggestions to make the presentation
@@
 are not discussed here).
+
+\item \emph{The novelty of this paper with respect to other papers by the same
+authors, including one in IJSR 2011, has been questioned. The authors should
+convincingly explain what are the novel contributions and the added value of
+this work compared to the previous publications.}
+
+We address this question with our reponse to reviewer 3, below.
+
+\end{itemize}
 
@@
 
+Reviewer 2 raised several concerns regarding the overall article organisation
+and lack of focus: \emph{``The overall impression is that the paper tries to
+cove a too brad set of topics. As a result the takeaway message for the user is
+too vague.''}
+
+S/he recommended to modify the section 3 to either provide more informations
+\emph{``to understand the problems and the technical solutions''} or to focus on
+one particular cognitive skill; asked for \emph{``more substantial validation
+of the specific problem(s) chosen in section 3''} in section 4, and suggested
+several modifications to the discussion.
+
+We have attempted to address these concerns in the following way:
+
 \begin{itemize}
+    \item section 3 underwent a large rewrite, in particular to provide
+        technical insights, as well as a better overall picture of the role and
+        importance of these skills. We have also added a subsection on the
+        reasoning processes that take place in the background.
+    \item the section 4 (previously ``experimental outcomes'') has been
+        rewritten to make clear that we only present here \emph{case studies}
+        that are meant to illustrate and clarify how the whole architecture
+        works ``in the real world''. Besides, more details have been provided to the
+        \emph{Clean the Table} study (as well as a redesigned figure) to help
+        the user to follow the execution flow.
+    \item the discussion section has been entirely reorganised, following the
+        Reviewer recommendation. The key decisional issues have been explicitly
+        stated, and overlaps between subsections have been addressed.
+\end{itemize}
+
+We address hereafter the other specific points raised by the reviewer:
+
+\begin{itemize}
+
+    \item \emph{pg. 7 (Section 3.1) ``From an artificial intelligence
+            perspective, Spark takes care of merging sensing modalities, it
+            computes and grounds perceptions into symbolic knowledge (situation
+            assessment), it is responsible temporal interpretation of situations
+            and performs spatial perspective-taking.'' given on this
+        characterization, Spark should be discussed in greater detail.} We have
+        significantly developed the presentation of {\sc Spark}, in
+        particular to exhaustively cover what and how symbolic relations are
+        computed.
+
     \item \emph{pg 10 (section 3.2.1)

### 2014-05-07 paper.tex
Proof-reading of sections 1,2,3 and 4

@@
 to an object); the robot must be able to take part to joint actions, both
-proactively (by planning and proposing resulting plans to the human) and
+pro-actively (by planning and proposing resulting plans to the human) and
 reactively; the robot must be able to move and act in a safe, efficient and
@@
     (B)}. The robot must adapt on-line its behaviours by merging computed plans
-    with reactive control. The robot explicitly reasons on the fact that it is
-    (or not) observed by the human. Reasoning and planning take place at
-    symbolic {\it (C)} as well as geometric level and take into account agents
-    beliefs, perspectives and capabilities {\it (D)} as estimated by the robot.}
+    {\it (C)} with reactive control. The robot explicitly reasons on the fact
+    that it is (or not) observed by the human. Reasoning and planning take place
+    at symbolic as well as geometric level and take into account agents beliefs,
+    perspectives and capabilities {\it (D)} as estimated by the robot.}
 
@@
 three challenges structure the whole field of human-robot interaction research,
-and can be analysed in terms of the cognitive skills that they require.
+and they can be analysed in terms of the cognitive skills that they require.
 
@@
 
-The robot controller then decides on an action to execute next and who has to
-perform it, from the human, the robot or both in case of a joint action, and
-eventually controls or monitors its execution. The operation continues until the
-goal is achieved, is declared unachievable or is abandoned by the human.
+The robot controller (with the help of a task planner) decides what action to
+execute next, and who should perform it from the human or the robot (or both in
+case of a joint action), and finally controls or monitors its execution. The
+operation continues until the goal is achieved, is declared unachievable or is
+abandoned by the human.
 
 \begin{inparaenum}
+
 This translates into several decisional, planning, representation skills that
@@
 with respect to state of the world and the task at hand, \item to build and
-maintain, and iteratively refine shared (human-robot) plans, \item to
-instantiate and execute the actions it has to perform, and conversely,
-to monitor those achieved by its human partner.
+iteratively refine shared (human-robot) plans, \item to instantiate and execute
+the actions it has to perform, and conversely, to monitor those achieved by
+its human partner.
+
 \end{inparaenum}
@@
 Such abilities must be designed and implemented in generic ways and
-provide several levels of parametrization such as they adapt to
+must provide several levels of parametrization such as they adapt to
 various environments, different tasks and variable levels of engagement of the
-robot, ranging from teammate behaviour to assistant or proactive helper.
+robot, ranging from teammate behaviour to assistant or pro-active helper.
+
+These are the challenges that we will discuss in this article.
 
@@
 
-The next section presents the overall architecture and the knowledge model we
-have developed for our robots. Section~\ref{sec:impl} gives some insights on
-each of the main components of the architecture, and try to highlight their
-significance for artificial intelligence. Section~\ref{sec:expe} presents two
-experiments that illustrate in a concrete way what can be currently achieved
-with our robots.  Section~\ref{sect|conclusion} finally summarises our main
-contributions and relate them to other recent researches in the field.
+The next section introduces the architecture as a whole, as well as the
+knowledge model that we have developed for our robots. Then,
+section~\ref{sec:impl} gives some insights on each of the main components of the
+architecture, and try to highlight their significance for artificial
+intelligence. Section~\ref{sec:expe} presents two studies that illustrate in a
+practical way what can be currently achieved with our robots.
+Section~\ref{sect|conclusion} finally restate our main contributions and discuss
+the key challenges that human-robot interaction brings to artificial
+intelligence.
 
@@
 
-\tikzset{subpart/.style={draw, font=\scriptsize, fill opacity=0.5, text opacity=1, fill=black!50}}
+\tikzset{subpart/.style={draw, font=\scriptsize, fill opacity=0.5, text opacity=1, fill=white!50}}
 
@@
 
-\caption{Overview of the architecture. The deliberative layer (on the top),
-composed of six main modules, interacts with the low-level sensori-motor
-layer (on the bottom). Knowledge is centrally managed in an active
-\emph{semantic blackboard}, pictured above with a thick border. The links
-between components depicted on the figure underline the central role of the
-knowledge base: most of the datastreams are actually symbolic statements exchanged
-through the semantic blackboard.}
+\caption{Overview of the architecture. A deliberative layer, composed of six
+    main modules, interacts with a low-level sensori-motor layer. Knowledge is
+    centrally managed in an active \emph{semantic blackboard}, pictured above
+    with a thick border. The links between components depicted on the figure
+    underline the central role of the knowledge base: most of the datastreams are
+    actually symbolic statements exchanged through this semantic blackboard.}
 
@@
 
-\subsection{Building a human-aware deliberative layer}
+\subsection{Building a Human-Aware Deliberative Layer}
 
@@
 blackboard}, connects most of the modules together: the \emph{geometric
-reasoning} module (called {\sc Spark}) produces symbolic assertions (like
+reasoning} module ({\sc Spark}) produces symbolic assertions (like
 \stmt{BOOK1 isOn TABLE}) describing the state and dynamics of the robot's
 environment. These logical statements are stored in the knowledge base, and
-reused by the language processing module ({\sc Dialogs}), the symbolic task
+queried back by the language processing module ({\sc Dialogs}), the symbolic task
 planner (HATP) and the execution controller. The output of the language
-processing module, or the activities started by the robot controller are
+processing module and the activities started by the robot controller are
 likewise stored as symbolic statements.
@@
 For instance, when processing a sentence like ``give me another book'', the {\sc
-Dialogs} module sends at some point a request to the knowledge base that looks
-like {\tt \footnotesize find(?book type Book, ?book differentFrom
-BOOK1)}\footnote{We present a similar but complete example at
-section~\ref{moving-london}.} and may write back assertions like \stmt{HUMAN
-desires GIVE\_ACTION45, GIVE\_ACTION45 actsOn BOOK2} to {\sc Oro}. The HATP
-planner then uses the knowledge base to initialise the planning domains with
-similar requests ({\tt \footnotesize find(BOOK2 isAt ?location)}, etc.). The
-execution controller typically monitor conditions (by subscribing to events
-like: {\tt \footnotesize onNewMatch(HUMAN desires ?goal)}) and stores what the
-robot is currently doing (\stmt{myself currentlyPerforms GIVE\_ACTION45}).
+Dialogs} module queries the knowledge base: {\tt \footnotesize find(?book type
+Book, ?book differentFrom BOOK1)}\footnote{We present study with a complete
+example at section~\ref{moving-london}.}, and write back assertions like
+\stmt{HUMAN desires GIVE\_ACTION45, GIVE\_ACTION45 actsOn BOOK2} to {\sc
+Oro}. The HATP planner then uses the knowledge base to initialise the
+planning domains with similar requests ({\tt \footnotesize find(BOOK2 isAt
+?location)}, etc.), and the execution controller typically monitor
+conditions (by subscribing to events like: {\tt \footnotesize
+onNewMatch(HUMAN desires ?goal)}) and stores what the robot is currently
+doing (\stmt{myself currentlyPerforms GIVE\_ACTION45}).
 
@@
 component does not purely act as an alternative sensing modality that would be
-fed to a distinct deliberative component: on the contrary, it actively queries
+fed to a distinct deliberative component. On the contrary, it actively queries
 the knowledge base to interpret and disambiguate natural language, and in that
@@
 As for any cognitive system, this fundamental interaction between knowledge and
-actions is central to our approach, and involves the dialogue component that
-acquires \emph{desires} from the other agents, the planner and the execution
-controller that may first decide to take or not into account an incoming desire
-as a \emph{goal}, and then generates and manages \emph{intentions} from these
-goals with the help of a symbolic task planner.
-
-Other activities are however conducted in parallel, without being explicitly
-triggered by \emph{desires} in the BDI sense, like situation assessment and
-monitoring or other forms of dialogue (including performative dialogue that can
-possibly change the internal state of the robot, but does not lead directly to
-the creation of \emph{desires}, like assertion of new facts or question
-answering).
+actions is central to our approach as well, and typically involves the dialogue
+module to acquire \emph{desires} from the other agents, and the planner and the
+execution controller that to first decide to take or not into account an
+incoming desire as a \emph{goal}, and then to generate and manage
+\emph{intentions} from these goals through the symbolic task planner.
+
+Building upon a BDI architecture, other activities are conducted in parallel,
+without being explicitly triggered by \emph{desires} in the BDI sense, like
+situation assessment and monitoring or other forms of dialogue (including
+performative dialogue that can possibly change the internal state of the robot,
+but does not lead directly to the creation of \emph{desires}, like assertion of
+new facts or question answering).
 
@@
 Oro} server~\cite{Lemaignan2010}, Figure~\ref{fig|spark-oro} top) which stores
-knowledge as it is produced by each of the deliberative components (the
+knowledge as it is produced by each of the other deliberative components (the
 clients). It exposes a {\tt json}-based RPC API~\cite{lemaignan2012kbapi} to
-query the knowledge base. The clients of the {\sc Oro} server are in charge of
-managing themselves the knowledge (when to add, when to update, when to
-retract knowledge): no meta-semantics is carried over that would let the server
-manage these dynamics itself.\footnote{There is one exception: so-called
-\emph{memory profiles} let the server discard facts after a specific period
-of time. We present this experimental feature at section~\ref{memory}.} We
-represent knowledge as RDF triples in the OWL sub-language. Each time
-triples are added or removed from the knowledge base, a Description Logics
-reasoner ({\sc Pellet}\footnote{\tt http://clarkparsia.com/pellet/})
-classifies the whole ontology and inserts all possible inferred triples.
+query the knowledge base.  We represent knowledge as RDF triples in the OWL
+sub-language. Each time triples are added or removed from the knowledge base, a
+Description Logics reasoner ({\sc Pellet}\footnote{\tt
+http://clarkparsia.com/pellet/}) classifies the whole ontology and inserts all
+possible inferred triples.  The clients of the {\sc Oro} server are in charge of
+managing themselves the knowledge (when to add, when to update, when to retract
+knowledge): no meta-semantics is carried over that would let the server manage
+these dynamics itself.\footnote{There is one exception: so-called \emph{memory
+profiles} let the server discard facts after a specific period of time. We
+present this experimental feature at section~\ref{memory}.}
 
@@
 departs from other approaches like the CAST model~\cite{Hawes2007} where
-knowledge is represented as a diffuse resource, or the CRAM/KnowRob
+knowledge is represented as a diffuse, pervasive resource, or the CRAM/KnowRob
 architecture~\cite{Beetz2010} where the knowledge base is an active hub that
@@
 and easy to capture since they are centralised) as well as high modularity
-(modules communicate through a defined and unified API).
+(modules communicate through an explicit and unified API).
 
@@
 Section~\ref{sect|ctrl}), where, by definition, we do not know beforehand what
-we are ``looking for'', and we need instead greedily infer as much as possible
+we are ``looking for'', and we need instead to greedily infer as much as possible
 to build a belief state as comprehensive as possible.
@@
 community. Then, because it is a well established project with numerous links to
-other on-line databases (like Wikipedia or WordNet), the reuse of important {\sc
+other on-line databases (like Wikipedia or WordNet), the reuse of key {\sc
 OpenCyc} concepts ensures to a certain extend that the knowledge stored by the
@@
 human is not considered as an object. In {\sc Cyc}, an object has a more precise
-definition: it is something \emph{partially tangible}. That includes obviously
+definition: it is something \emph{partially tangible}. This includes obviously
 the humans, and actually many other entities that would not be commonly said to
@@
 well-defined and standard semantics to exchange informations between artificial
-systems.
+systems in a non-ambiguous way.
 
@@
 
-\concept{Bottle} $\equiv$ \concept{Container} {\bf and} \concept{Tableware}
-{\bf that} (\concept{hasShape} {\bf value} \concept{cylinderShape} {\bf and}
-\concept{hasCharacteristicDimension} {\bf only} \concept{\em int[>= 0.1, <=
-0.3]})
+$$
+\text{\tt Bottle} \equiv \text{\tt Container} \wedge 
+                         \text{\tt Tableware} \wedge
+                         \text{\tt hasShape} \in \text{\tt CylinderShape}^{\mathcal{I}} \wedge
+                         \text{\tt hasMainDimension} \in [0.1, 0.3]
+$$
 
@@
 
-\emph{Symbol grounding} relates the knowledge model to the perception and
-actuation capabilities of the robot, and our architecture adopts here a
-bidirectional approach.  The different components that we have described so far
+\emph{Symbol grounding} connects hence the knowledge model to the perception and
+actuation capabilities of the robot, and our architecture adopts here also a
+bidirectional approach.  The different components that we have mentioned so far
 provide a \textit{bottom-up} grounding process: geometric reasoning and dialogue
@@
 While the previous section provided an overview of our cognitive architecture
-as well as its associated knowledge model, this section discusses each of its main
+as well as its associated knowledge model, this section discusses its main
 ``building blocks''. Each of these components may have connections to several
@@
 classes that an individual belongs to). In case of logical inconsistency, the
-reasoner can also provide \emph{explanations} that are important for debugging.
-
-Besides those, {\sc Oro} server implements several algorithms to identify
+reasoner can also provide \emph{explanations} that are used in particular for
+debugging.
+
+Besides those, {\sc Oro} server implements several algorithms (presented
+in~\cite{Ros2010b}) to identify
 similarities and differences between concepts (classes or
-instances)~\cite{Ros2010b}: the \emph{Common Ancestors} algorithm, useful to
+instances): the \emph{Common Ancestors} algorithm, useful to
 determine the most specific class(es) that include a given set of individuals;
@@
 properties of individuals, either statically asserted in the common-sense
-ontology, acquired by the robot through perception or proactive questioning of
+ontology, acquired by the robot through perception or pro-active questioning of
 the human partner, or derived from other reasoning algorithms like the
@@
 the same common-sense knowledge, but rely on each agent's perspective for the
-actual instantiation: the robot (geometrically) computes that the book is in its
-own field of view, but not in the human one. The robot knowledge contains the
+actual instantiation: if the robot (geometrically) computes that the book is in its
+own field of view, but not in the human one, the robot knowledge contains the
 fact \stmt{book isVisible true} while the human model contains \stmt{book
@@
 Memory has been studied at length in the cognitive psychology and
-neuropsychology communities: very briefly, the idea of \emph{short-term} and
-\emph{long-term} memory is due to Atkinson and Shiffrin~\cite{Atkinson1968} ,
-Anderson~\cite{Anderson1976} splits memory into \emph{declarative} (explicit)
-and \emph{procedural} (implicit) memories, Tulving~\cite{Tulving1985} organises
+neuro-psychology communities: to give the main pointers, the idea of \emph{short-term} and
+\emph{long-term} memory is due to Atkinson and Shiffrin~\cite{Atkinson1968};
+Anderson~\cite{Anderson1976} proposes to split memory into \emph{declarative} (explicit)
+and \emph{procedural} (implicit) memories; Tulving~\cite{Tulving1985} organises
 the concepts of \emph{procedural}, \emph{semantic} and \emph{episodic} memories
 into a hierarchy. Short-term memory is eventually refined with the concept of
-\emph{working memory} by Baddeley~\cite{Baddeley2010}
+\emph{working memory} by Baddeley~\cite{Baddeley2010}.
+In the field of cognitive architectures, the {\sc Soar}
+architecture~\cite{Lehman2006} is one of those that tries to reproduce a
+human-like memory organisation. The GLAIR cognitive
+architecture~\cite{Shapiro2009} also has a concept of long term/short term and
+episodic/semantic memories.
 
@@
 \emph{memory profile} is optionally attached to them.  Three such profiles are
-predefined: {\tt short term}, {\tt episodic} and {\tt long term}. They are
+predefined: {\it short term}, {\it episodic} and {\it long term}. They are
 currently attached to different lifetime for the statements (respectively 10
 seconds, 5 minutes and no time limit). After this duration, the statements are
-automatically removed from the knowledge base.
+automatically retracted.
 
@@
 refer to the semantics of the statements (that is expected to be related to an
-event) and not to a specific life duration.
+event) and not to a specific lifespan.
 
 We rely however on this mechanism in certain cases: for instance, some modules
-like the natural language processor use the {\tt short term} memory profile to
+like the natural language processor use the {\it short term} memory profile to
 mark for a few seconds important concepts that are currently manipulated by the
@@
 this feature to trace certain knowledge-related processes.
-
-In the larger field of cognitive architectures, the {\sc Soar}
-architecture~\cite{Lehman2006} is one of those that tries to reproduce a
-human-like memory organisation. The GLAIR cognitive
-architecture~\cite{Shapiro2009} also has a concept of long term/short term and
-episodic/semantic memories.
 
@@
 Numerous approaches exist, like amodal \emph{proxies}~\cite{Jacobsson2008},
-grounded amodal representations~\cite{Mavridis2006}, semantic maps
-(\cite{Nuechter2008, Galindo2008,Blodow2011}) or affordance-based planning and
-object classification~\cite{Lorken2008, Varadarajan2011}.
+grounded amodal representations~\cite{Mavridis2006}, semantic
+maps~\cite{Nuechter2008, Galindo2008,Blodow2011} or affordance-based planning
+and object classification~\cite{Lorken2008, Varadarajan2011}.
 
@@
 We rely on a dedicated geometric and temporal reasoning module called {\sc
-Spark} (\emph{SPAtial Reasoning \& Knowledge}~\cite{Sisbot2011}). It is a
+Spark} (\emph{SPAtial Reasoning \& Knowledge}, \cite{Sisbot2011}). It is a
 situation assessment reasoner that generates symbolic knowledge from the
@@
 modalities and as bridge with the symbolic layer. This geometric model is
-continuously updated at run-time by the robot based on its sensors (in our
-experiments, objects are identified and localised through 2D fiducial markers, while
-humans are tracked with Kinect-like devices, optionally assisted by motion
+updated at run-time by the robot based on its sensors (in most of
+our studies, objects are identified and localised through 2D fiducial markers,
+while humans are tracked with Kinect-like devices, optionally assisted by motion
 capture to accurately track the head motion, which is required to compute what
@@
 
-From an artificial intelligence perspective, {\sc Spark} takes care of merging
-sensing modalities, it computes and grounds perceptions into symbolic
-knowledge (situation assessment), it is responsible for temporal interpretation of
-situations and performs spatial perspective-taking.
-
-{\sc Spark} is designed to run and continuously update the knowledge base (in our
-experiments, it runs at about 10Hz). At each steps, it re-computes spatial
+{\sc Spark} is designed to run and continuously update the knowledge base (in
+our experiments, it runs at about 10Hz). At each steps, it re-computes spatial
 relationships and affordances for the whole scene, and send the delta (new
 relations and relations that do not hold anymore) to the knowledge base. This
-approach may raise scalability concerns (we did not however hit such issues in
-our constrained scenarios), but greatly simplifies the management of the
-\emph{dynamics} of the knowledge (\emph{When do I discard outdated knowledge?
-When do I update it?}). Since it is equivalent to a reset of the reasoner
-domain, it also hides issues linked to non-monotonic reasoning
+approach may raise scalability concerns (we did not however hit major
+performance issues in our constrained scenarios), but greatly simplifies the
+management of the \emph{dynamics} of the knowledge (\emph{When do I discard
+outdated knowledge?  When do I update it?}). Since it is equivalent to a reset
+of the reasoner domain, it also hides issues linked to non-monotonic reasoning
 (see~\cite{McCarthy2007} for a discussion on that question).
@@
 
-\subsubsection{Building an Agent-aware Symbolic Model of the Environment}
+\subsubsection{Building an Agent-Aware Symbolic Model of the Environment}
 \label{sect|situ}
 
-\paragraph{On Perspective Taking} Visual perspective taking refers to the
+\paragraph{On Perspective Taking} \emph{Visual} perspective taking refers to the
 ability for visually perceiving the environment from other's point of view.
@@
 perception of one person differs from the other one. In developmental
-psychology, one typical example consists of two similar objects in a room (\eg.
+psychology, one typical example consists of two similar objects in a room (\eg
 two balls) where both are visible for the child, but only one is visible for
@@
 
-Spatial perspective taking refers to the qualitative spatial location of
+\emph{Spatial} perspective taking refers to the qualitative spatial location of
 objects (or agents) with respect to a frame (\eg \emph{the keys on my left}).
@@
 
-In {\sc Spark}, we use two types of the frames of reference: \emph{egocentric}
+In {\sc Spark}, we use two types of the frames of reference: \emph{ego-centric}
 (from the robot perspective) and \emph{allo-centric}, \ie addressee-centred
@@
 
-\paragraph{Agent-Independent Locations}
-
-We can refer to object locations with respect to other objects in the
-environment, such as \emph{above, next to, in}, etc. {\sc Spark} computes
-three main relations based on the bounding box and centre of mass of the
-objects (Figure~\ref{fig|spark-oro}\emph{a}):
-
-\begin{itemize}
-	\item \concept{isOn}: computes if an object $O_1$ is on another object $O_2$ by
-	evaluating the centre of mass of $O_1$ according to the bounding box of $O_2$.
-
-	\item \concept{isIn}: evaluates if an object $O_1$ is inside another object
-	$O_2$ based on their bounding boxes $BB_{O_1}$ and $BB_{O_2}$.
-
-	\item \concept{isNextTo}: indicates whether an object $O_1$ is next to another
-	object $O_2$. We cannot use a simple distance threshold to determine if two
-	objects are next to each other since the relation is highly dependent on the
-	dimensions of the objects. For instance, the maximum distance between large
-	objects (\eg two houses) to consider them as being next to each other is much
-	larger than the maximum distance we would consider for two small objects (\eg
-	two bottles). Thus, the relation between the dimensions and the distances of
-	the objects are taken into account.
-
-\end{itemize}
+{\sc Spark} computes three main agent-independent relations based on the
+bounding box and centre of mass of the objects
+(Figure~\ref{fig|spark-oro}\emph{a}): \concept{isOn} holds when an object $O_1$
+is on another object $O_2$, and is computed by evaluating the centre of mass of
+$O_1$ according to the bounding box of $O_2$.  \concept{isIn} evaluates if an
+object $O_1$ is inside another object $O_2$ based on their bounding boxes
+$BB_{O_1}$ and $BB_{O_2}$.  \concept{isNextTo} indicates whether an object $O_1$
+is next to another object $O_2$. We cannot use a simple distance threshold to
+determine if two objects are next to each other since the relation is highly
+dependent on the dimensions of the objects. For instance, the maximum distance
+between large objects (\eg two houses) to consider them as being next to each
+other is much larger than the maximum distance we would consider for two small
+objects (\eg two bottles). Thus, the relation between the dimensions and the
+distances of the objects are taken into account.
 
@@
 
-\paragraph{Agent-Dependent Placements}
-
-While in previous section we listed several \emph{absolute} location predicate,
-many topological relations are directly dependent from the observation point.
-
-The predicate \concept{hasRelativePosition} represents spatial locations
+Many topological relations are conversely dependent from the observation point.
+The predicate \concept{hasRelativePosition} represents such spatial relations
 between agents and objects that are agent dependent.  We compute these spatial
@@
 description of the relative positioning of objects in the environment.
+Table~\ref{facts} summarises all the symbolic spatial relations computed by {\sc
+Spark}.
 
@@
 in general to fully ground the human-robot interaction, and a model of the
-current capabilities of the agents surrounding the robot is also required.
-
-There are a number of common properties for a robot and a human related to
-their capabilities in a given situation: they can both reach, grasp, look at,
-point at, etc.: we group them in the \emph{Agent} category, defined as entities
-that can act in the environment and manipulate it.
+current capabilities of the agents interacting with the robot is also required.
 
@@
     narrower field of view, the field of attention (FOA).
-    Figure~\ref{fig::sparkRepresentations}\emph{c} shows the field of attention
+    Figure~\ref{fig|spark-oro}\emph{c} shows the field of attention
     of a person with a green cone (narrower one). In this example only the grey
@@
 
-\item \emph{Points At}: verifies whether an object is pointed at by an agent.
+\item \emph{Points At} holds when an object is pointed at by an agent.
     This relation is computed by placing a virtual camera on the hand, aligned
@@
 
-If a larger object occludes a smaller one while an agent is pointing at them, the
-outcome of the evaluation will result only in one relation, \ie \stmt{agent\_01
-pointsAt object\_01} since the small one is not visible to the agent.  On the
-contrary, if the small object is in front of the big one, then both objects
-will satisfy the relation, which may generate an ambiguity (which object the
-agent refers to?) that is let to be solved by other discrimination algorithms.
-
 To make recognition more robust, these three first capabilities are filtered
@@
 \item \emph{Reachable}: it allows the robot to estimate the agent's capability
-to reach an object, which is fundamental for task planning. For example, if the
-user asks the robot to give him/her an object, the robot must compute a transfer
-point where the user is able to get the object afterwards. 
-Figure~\ref{fig::reach-ex} explains how addressee-dependent reachability is
-computed, based on Generalised Inverse Kinematics and collision detection.
+    to reach an object, which is fundamental for task planning. For example, if
+    the user asks the robot to give him/her an object, the robot must compute a
+    transfer point where the user is able to get the object afterwards.
+    Reachability is computed for each agents based on Generalised Inverse
+    Kinematics and collision detection.
 
@@
 Table~\ref{facts} lists all the symbolic facts that are computed by {\sc Spark},
-along with the type of the statements' subjects and objects.
+along with the admissible classes for the subjects and objects of the statements.
 
@@
 		 \concept{Agent}  & \concept{sees}  &  \concept{SpatialThing}  &    \\ 
-		 \concept{SpatialThing}  & \concept{isInFieldOfView}  &  \concept{xsd:boolean}  & via inference: \par \tiny \stmt{myself sees *} $\Leftrightarrow$ \stmt{* isInFieldOfView true} \\ 
+		 \concept{SpatialThing}  & \concept{isInFieldOfView}  & \concept{xsd:boolean}  & \par \tiny \concept{myself sees *} $\Leftrightarrow$ \concept{* isInFieldOfView true} \\ 
 		 \concept{Agent}  & \concept{pointsAt} $\equiv$ \concept{cyc:pointingToward}  & \concept{SpatialThing} \\ 
-		 \concept{Agent}  & \concept{focusesOn}  &  \concept{SpatialThing}  &  via inference: \par \tiny \concept{looksAt} $\wedge$ \concept{pointsAt} $\Rightarrow$ \concept{focusesOn} \\
+		 \concept{Agent}  & \concept{focusesOn}  &  \concept{SpatialThing}  &  \par \tiny \concept{looksAt} $\wedge$ \concept{pointsAt} $\Rightarrow$ \concept{focusesOn} \\
 		\concept{Agent} & \concept{seesWithHeadMovement} &  \concept{SpatialThing} \\
@@
 
-\fxfatal{Rephrase}
+\fxfatal{Rephrase or remove}
 It is sometimes difficult or even impossible to see and/or track an object in
@@
 Monitoring human activity is crucial for maintaining a coherent state of the
-world. Full human action and activity monitoring is a task that requires
+world. While full human action and activity recognition is a task that requires
 knowledge and reasoning both on high level facts like goals, intentions and
-plans, as well as bottom-up data from agent and object motions. Simple temporal
+plans, as well as bottom-up data from agent and object motions, simple temporal
 and geometric reasoning on human hand trajectories and potential objects
-placements can provide some useful clues for high level human monitoring
+placements can provide clues for high level human monitoring
 processes. We call this temporal and geometric reasoning \emph{primitive action
@@
 \emph{pick}), or ``a hand holding an object is over a container'' (precursor for
-a \emph{throw}). \emph{Spark} recognises a handful \fxfatal{which ones?} of these
+a \emph{throw}). {\sc Spark} recognises a handful \fxfatal{which ones?} of these
 primitives, that are relied on by the robot to track the state transitions in
-the human plans (as explained below, at section~\ref{hatp}).
-
-%\subsubsection{Ego-centric and Allo-centric Frames}
-%
-%{\sc Spark} also supports \emph{perspective-taking}: spatial relations between entities can
-%be computed from different viewpoints, which let the robot build for each
-%agents it interacts with a different, \emph{perspective-aware} symbolic model
-%of the environment. These models are separately stored in the knowledge base.
-%
-%This allows us to deal with ambiguities that arise when one speaker refers to
-%an object within a reference system (or changes the reference system, \ie
-%switches perspective) without making the reference frame
-%explicit~\cite{Breazeal2006, Ros2010}.
-%
-%As a result, the robot stores models of the environment either in the
-%\emph{ego-centric} reference frame (from the robot perspective) or in 
-%\emph{allo-centric} frames (addressee-centred).
+the human plans.
 
@@
 The main weakness of our current implementation relates to the way semantics are
-conveyed: because we use 2D markers for artifacts, we skip most of the object
-extraction and recognition challenges. Each object has a unique identifier,
+conveyed: because we use 2D markers for artifacts, we avoid issues related to object
+segmentation and recognition. Each object has a unique identifier,
 which enable us to load a proper CAD model suitable for 3D spatial reasoning
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Multi-modal Communication}
+\subsection{Multi-Modal Communication}
 \label{sect|com}
 
-\subsubsection{Natural language grounding}
-
-Natural language processing is one of the fields of human-robot interaction
-for which the introduction of the semantic layer has been most beneficial.
-In~\cite{Lemaignan2011a} we detail the techniques and the tool called
-\texttt{dialogs} that we have developed for natural English language parsing and
-grounding, along with verbalisation and (minimalist) dialogue management.
+\subsubsection{Natural Language Grounding}
+
+In our experience, natural language processing is one of the fields of
+human-robot interaction for which the introduction of the semantic layer has
+been most beneficial.  In~\cite{Lemaignan2011a} we detail the techniques and the
+tool called {\sc Dialogs} that we have developed for natural English language
+parsing and grounding, along with verbalisation and (minimalist) dialogue
+management.
 
@@
 
-    \caption{Processing an order: this simple example (no ambiguity), taken
-    from~\cite{Lemaignan2011a}, shows the thematic roles computed by 
-    {\sc Dialogs} for the order ``Give me the book''.}
+    \caption{Processing of a simple, non-ambiguous order, taken
+        from~\cite{Lemaignan2011a}. Thematic roles ({\tt performedBy}, {\tt
+        actsOnObject}, {\tt receivedBy}) are automatically extracted by 
+    {\sc Dialogs} from the order ``Give me the book.''.}
 
@@
 
-\subsubsection{Multi-modal communication}
+\subsubsection{Multi-Modal Communication}
 
@@
 First, particular steps of the grounding process explicitly check for the
-presence and value of specific facts: for instance, when several instances
-match a category (the human says ``give me the bottle'' and the robot knows
-about three bottles), the module may decide (it actually depends on the
-quantifier preceding the class) to discard some of them based on their
-\emph{visibility} for the speaker (implicit communication context built on the
-human posture).
+presence and value of specific facts: for instance, when several instances match
+a category (the human says ``give me the bottle'' and the robot knows about
+three bottles), the module may decide (it actually depends on the quantifier
+preceding the class, see the example at section~\ref{moving-london}) to discard
+some of them based on their \emph{visibility} for the speaker (implicit
+communication context built on the human posture).
 
@@
 generate unambiguous references to discriminate between similar objects: ``do
-you mean the bottle that is next to the glass?'' (we kindly refer the interested
-readers to~\cite{Ros2010b} for a detailed account of our approach to interactive
-concept discrimination, along with the related algorithms).
+you mean the bottle that is next to the glass?'' (\cite{Ros2010b} provides a
+detailed account of our approach to interactive concept discrimination, along
+with the related algorithms).
 
@@
 Our execution controllers rely on symbolic task planning to convert long-term
-desires into a succession of atomic actions. We use in our architecture the
-HATP planner (for Human Aware Task Planner~\cite{Alili2008, Alili2009}).  HATP
-is a refinement of a Hierarchical Task Network (HTN). As an HTN, it performs an
+desires into a succession of atomic actions. We use in our architecture the HATP
+planner (for Human Aware Task Planner~\cite{Alili2008, Alili2009}).  HATP is a
+refinement of a Hierarchical Task Network (HTN). As an HTN, it performs an
 iterative task decomposition into sub-tasks until reaching atomic actions. The
@@
 task and represents the procedural knowledge of the robot. It is stored outside
-of the central knowledge base, using a specific formalism.
+of the central knowledge base, using a specific formalism (see the related
+discussion at the end of this section).
 
@@
 
-HATP originality resides in its ability to produce plans for the robot's actions
-\emph{as well as for the other participants} (humans or robots), so-called
-\emph{shared plans}. This effectively enriches the robot's interaction
-capabilities by allowing for a form of \emph{human behaviour prediction}, which
-is used in turn by the robot execution controller to monitor the engagement of
-the human partner during joint tasks.
+The originality of HATP resides in its ability to produce plans for the robot's
+actions \emph{as well as} for the other participants (humans or robots), that we
+call \emph{shared plans}. This effectively enriches the interaction capabilities
+of the robot by allowing for a form of human behaviour \emph{prediction}, which
+is used by the robot execution controller to monitor the engagement of the human
+partner during joint tasks.
 
@@
 
-    \node[right=9cm of h2] (h3) {\action{TAKE}{BOTTLE}{TABLE}} edge (h2) edge (r4);
+    \node at (r5 |- h1) (h3) {\action{TAKE}{BOTTLE}{TABLE}} edge (h2) edge (r4);
     \node[right=of h3] (h4) {\action{PUT}{BOTTLE}{BIN1}} edge (h3);
@@
 }
-\caption{A plan produced by HATP with two action streams (here, human actions
-    are at the top while robot actions are at the bottom). The arrow between the
-    two streams represents a synchronization point ({\sf PUTRV} stands for {\it Put it so it is both
+
+\caption{A plan produced by HATP with two action streams (human actions at the
+top, robot actions at the bottom). The arrow between the two streams represents
+a synchronization point ({\sf PUTRV} stands for {\it Put it so it is both
 reachable and visible}).}
@@
 
-In addition to these costs, HATP also takes into
-account a set of social rules.  Social rules are constraints aiming at leading
-the plan construction towards the best plan according to some human preferences.
-The social rules we have defined so far deal with:
-
-\begin{itemize}
-\item undesirable state: to avoid a state in which the human could
-  feel uncomfortable;
-\item undesirable sequence: to eliminate sequences of actions that can
-  be misinterpreted by the human;
-\item effort balancing: to adjust the work effort of the agents;
-\item wasted time: used to avoid long delays between the actions of
-  the human partner;
-\item intricate links: to limit dependencies between the actions of
-  two or more agents.
-\end{itemize}
+In addition to these costs, HATP also takes into account a set of social rules.
+Social rules are heuristics aiming at leading the plan construction towards the
+best plan according to some human preferences.  The social rules we have defined
+so far deal with \emph{undesirable states} to avoid a state in which the human
+could feel uncomfortable; \emph{undesirable sequences} to eliminate sequences of
+actions that can be misinterpreted by the human; \emph{effort balancing} to
+adjust the work effort of each of the agents; \emph{no wasted time} to avoid
+long delays between the actions of the human partner; and \emph{no intricate
+links} to limit dependencies between the actions of two or more agents.
 
@@
 
-\caption{An alternative plan where the \emph{wasted time} social rule is used to optimise the
+\caption{An alternative plan where the \emph{no wasted time} social rule is used to optimise the
 total duration of the task. }
@@
 Figure~\ref{plan_hatp2} illustrates an alternative plan to the previous 
-one (Figure~\ref{plan_hatp1}) if the wasted time social rule is used.
-The obtained shared plan is the best plan according to a global evaluation of
-these multiple criteria.
-
-\subsubsection{Commitment Levels}
-
-By tuning its costs and adapting its social rules, HATP can be used to compute
-various alternative plans. These plans can be categorised into several levels of
-cooperation
-
-\begin{itemize}
-\item helping the human to achieve his goal by acting for him
-\item sharing concrete resources by handing some objects
-\item collaboration of the robot and the human by coordinating their
-  actions towards a human-robot joint goal.
-\end{itemize}
+one (Figure~\ref{plan_hatp1}) where the \emph{no wasted time} social rule is
+applied. The resulting shared plan is a better plan according to a global evaluation of
+the costs and rules.
+
+%\subsubsection{Commitment Levels}
+%
+%By tuning its costs and adapting its social rules, HATP can be used to compute
+%various alternative plans. These plans can be categorised into several levels of
+%cooperation
+%
+%\begin{itemize}
+%\item helping the human to achieve his goal by acting for him
+%\item sharing concrete resources by handing some objects
+%\item collaboration of the robot and the human by coordinating their
+%  actions towards a human-robot joint goal.
+%\end{itemize}
+%
 
@@
 specific format, outside of the central declarative knowledge repository, and
-the plans are directly communicated to the robot controller. Thus, like other
-cognitive architectures, we have independent declarative and procedural
-knowledge stores.
+the plans are directly communicated to the robot controller. This leads to
+independent declarative and procedural knowledge stores.
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Robot control}
+\subsection{Robot Execution Control}
 \label{sect|ctrl}
 
-While parts of the architecture ({\sc Spark}, {\sc Oro}) have been deployed with external
-execution controllers (like {\sc Cram}~\cite{Beetz2010} or the BERT
+While parts of the architecture ({\sc Spark}, {\sc Oro}) have been deployed with
+external execution controllers (like {\sc Cram}~\cite{Beetz2010} or the BERT
 platform~\cite{Lallee2010b}, as reported in~\cite{Lemaignan2010}), we have also
@@
 in Figure~\ref{fig|archi}. {\sc Shary}~\cite{clodic2008shary} is the main one,
-written in the PRS~\cite{Ingrand1996} language. We also use the Python-based
-{\sc pyRobots} collection of tools that provide a large set of high-level
-actions and an event-based architecture well suited for prototyping.
-They both rely on deep integration with the knowledge base that serves as the
-primary source of semantics for decision making.
-
-The novelty of our contribution here is the interleaving of a semantic-aware
-architecture designed for autonomy with human-aware decisional components. To
-our knowledge, no other robotic architecture covers in depth those two
-perspectives.
-
-The Ke Jia project~\cite{Chen2010} is maybe one of the closest
-approach. It integrates on a mobile platform a knowledge representation language
-with natural language processing, task planning and motion planning. Default and
-non-monotonic reasoning has been especially researched within the Ke Jia project
-for symbolic task planning~\cite{Ji2011} and underspecified natural language
-processing. The Ke Jia robot has been demonstrated in several tasks involving human-robot
-interaction with natural language. These tasks include a task with multiple
-\emph{pick \& carry} that are globally optimised, naive physics reasoning via
-taught rules or more complex scenarios with the robot delivering drinks, taking
-into account changing and mutually exclusive preferences of users.
-
-Also notable, Beetz et al.~\cite{Beetz2010} proposes a cognitive architecture
-called \textsc{CRAM} (Cognitive Robot Abstract Machine) that deeply integrates
-with \textsc{KnowRob}~\cite{Tenorth2009a}, a knowledge processing framework
-based on Prolog. Its underlying storage is based on an OWL ontology, derived
-from \textsc{OpenCyc}. \textsc{CRAM} and \textsc{KnowRob} have been demonstrated
-on several real-world scenarios, where natural language recipes extracted from
-the Internet had to be translated into plans and executed in a kitchen
-environment, perceived and rebuilt on-line by the robots. This architecture has
-however not been deployed in many scenarios involving human-robot interactions.
-
-\subsubsection{Desires and experiences}
+written in the \emph{Procedural Reasoning System} language~\cite{Ingrand1996}.
+We have also developed the Python-based {\sc pyRobots} collection of tools to
+provide a large set of high-level actions and an event-based architecture well
+suited for prototyping.  They both rely on deep integration with the knowledge
+base that serves as the primary source of semantics for decision making.
+
+%The novelty of our contribution here is the interleaving of a semantic-aware
+%architecture designed for autonomy with human-aware decisional components. To
+%our knowledge, no other robotic architecture covers in depth those two
+%perspectives.
+%
+%The Ke Jia project~\cite{Chen2010} is maybe one of the closest
+%approach. It integrates on a mobile platform a knowledge representation language
+%with natural language processing, task planning and motion planning. Default and
+%non-monotonic reasoning has been especially researched within the Ke Jia project
+%for symbolic task planning~\cite{Ji2011} and underspecified natural language
+%processing. The Ke Jia robot has been demonstrated in several tasks involving human-robot
+%interaction with natural language. These tasks include a task with multiple
+%\emph{pick \& carry} that are globally optimised, naive physics reasoning via
+%taught rules or more complex scenarios with the robot delivering drinks, taking
+%into account changing and mutually exclusive preferences of users.
+%
+%Also notable, Beetz et al.~\cite{Beetz2010} proposes a cognitive architecture
+%called \textsc{CRAM} (Cognitive Robot Abstract Machine) that deeply integrates
+%with \textsc{KnowRob}~\cite{Tenorth2009a}, a knowledge processing framework
+%based on Prolog. Its underlying storage is based on an OWL ontology, derived
+%from \textsc{OpenCyc}. \textsc{CRAM} and \textsc{KnowRob} have been demonstrated
+%on several real-world scenarios, where natural language recipes extracted from
+%the Internet had to be translated into plans and executed in a kitchen
+%environment, perceived and rebuilt on-line by the robots. This architecture has
+%however not been deployed in many scenarios involving human-robot interactions.
+%
+
+\subsubsection{Desires and Experiences}
 
@@
 
-\emph{Desires} are typically human orders (``Give me that book''). The nature
-of the desired action (to pick, give, look, bring, show...), along with the
-action parametrization (what is acted on? who should perform the action? etc.)
-are extracted from the knowledge base, and either passed to a task planner
-(presented in the previous section) or executed if the procedure is directly
+\emph{Desires} are typically human orders (``Give me that book''). The nature of
+the desired action (to pick, give, look, bring, show...), along with the action
+parametrization (thematic roles) are extracted from the knowledge base, and
+either passed to the task planner or executed if the procedure is directly
 available.
@@
 parametrization to lower the effort the human needs to provide for the following
-joint manipulation tasks.\footnote{Note that this very example has been
+joint manipulation tasks.\footnote{Note that this specific example has been
 implemented as a proof-of-concept. A broader framework that would support
@@
 
-\subsubsection{Event-driven control}
+\subsubsection{Event-Driven Control}
 
@@
 subscribe to an event by passing through an event pattern (in its simplest
-form, a partial statement like \stmt{? type Elephant}) and a callback.  Each
-time a new instance of elephant appears in the knowledge base, the callback is
+form, a partial statement like \stmt{? type Book}) and a callback.  Each
+time a new instance of a book appears in the knowledge base, the callback is
 triggered.
@@
 expressiveness: for instance, by subscribing to the event \stmt{human1 desires
-?action, ?action type Look, ?action hasGoal myself}, we could trigger a
+?action, ?action type Take, ?action actsOnObject ?obj, ?obj type Toy}, we could trigger a
 behaviour when the human expresses (through dialogue, gestures...) that he
-wants to look at the robot itself.
+wants to somehow get a toy.
 
@@
 Note also that we take advantage of the reasoning capabilities of the system:
-for example, the goal of the action (\stmt{action hasGoal myself}) may not be
+for example, the type of the object (\stmt{?obj type Toy}) may not be
 explicitly asserted, but inferred by the reasoner based on other assertions.
@@
 
-Action execution and monitoring is the central activity of the robot executive
-controller, and involves interaction with most of the components presented on
-Figure~\ref{fig|archi}.
-
-Like most robotic architecture, actions are split into \emph{atomic actions}
-and more complex \emph{tasks}. Tasks are created either statically or dynamically.
-The {\sc pyRobots} controller, for instance, combines the actions listed in
+Like most robotic architectures, actions are split into \emph{atomic actions}
+that are combined into \emph{tasks}. Tasks are created either statically or dynamically.
+The {\sc pyRobots} controller, for instance, statically combines the actions listed in
 table~\ref{table|pyrobots_actions} to implement the high-level tasks it exposes: {\tt
@@
 {\tt pickObject}, {\tt bringObject}, {\tt putObject}, {\tt hideObject}.
-
 Our other controller, {\sc Shary}, relies on the symbolic task planner to
-generate at run-time suitable sets of atomic actions.
+dynamically generate, at run-time, suitable sets of atomic actions.
 
@@
 \end{center}
-\caption{Main {\sc pyRobots} \emph{actions}, sorted by categories. These
+\caption{Main {\sc pyRobots} actions, sorted by categories. These
 actions are combined at run-time into higher-level \emph{tasks}.}
@@
 
-\paragraph{Human-Aware Motion, Placement and Manipulation Planning}
-Manipulation and navigation actions listed in
-table~\ref{table|pyrobots_actions} rely on a collection of dedicated 3D motion
-planner to generate movements that are aware of surrounding humans. Robot
-placement and end-effectors trajectories are computed on-line to allow object
-manipulation that take into account task specific constraints and human
-postures, abilities and preferences. \cite{Sisbot2008, Mainprice2011,
-Pandey2011} detail these techniques.
-
-\paragraph{Action Execution and Monitoring Robot Controller}
-Based on context and on the shared plan elaborated by HATP for a given goal,
-the robot controller decides to execute an action or to ask its human
-partner to do it.  Actions feasibility by the human or the robot are
-regularly reconsidered based on the reachability / visibility
-computation mechanisms.
-
-Robot action execution is based on simple automatons  that translate
-symbolic planning atomic actions into sequences of planned arm motions
-and gripper commands to execute according to current state
-of the 3-tuple (gripper, object, furniture). We have three states
-according to whether the object is in gripper and if it is in gripper
-whether it is on furniture.  These states are directly obtained from
-the updated symbolic state of the world in the ontology.
-
-For plan action monitoring, primitive actions recognition is used. A primitive
-action detection is interpreted as action success if it is the expected one and
-failure otherwise. The robot also reacts to the absence of activity.
+Manipulation and navigation actions listed in table~\ref{table|pyrobots_actions}
+rely ona dedicated 3D motion planner ({\sc Mhp}, depicted in
+Figure~\ref{fig|archi}): robot placement and end-effectors trajectories are
+computed on-line to allow object manipulation that take into account task
+specific constraints and human postures, abilities and preferences.
+\cite{Sisbot2008, Mainprice2011, Pandey2011} detail these techniques.
+
+%\paragraph{Action Execution and Monitoring Robot Controller}
+%Based on context and on the shared plan elaborated by HATP for a given goal,
+%the robot controller decides to execute an action or to ask its human
+%partner to do it.  Actions feasibility by the human or the robot are
+%regularly reconsidered based on the reachability / visibility
+%computation mechanisms.
+%
+%Robot action execution is based on simple automatons  that translate
+%symbolic planning atomic actions into sequences of planned arm motions
+%and gripper commands to execute according to current state
+%of the 3-tuple (gripper, object, furniture). We have three states
+%according to whether the object is in gripper and if it is in gripper
+%whether it is on furniture.  These states are directly obtained from
+%the updated symbolic state of the world in the ontology.
+%
+%For plan action monitoring, primitive actions recognition is used. A primitive
+%action detection is interpreted as action success if it is the expected one and
+%failure otherwise. The robot also reacts to the absence of activity.
+%
 
@@
 Our architecture has been deployed and tested in a number of studies on
-several robotic platforms. Table~\ref{table|experiences} lists the most
+several robotic platforms. Table~\ref{table|experiments} lists the most
 significant ones, with their main focuses and reference publications.
@@
 We briefly introduce two of them in order to illustrate in a practical way
-different aspects of the architecture that were previously presented.  The first
-one is focused on knowledge representation and verbal interaction: the human
-asks for help to find and pack objects ({\it Interactive Grounding I} in
-Table~\ref{table|experiments}). The second one ({\it Cleaning the table})
-involves the {\sc Shary} execution controller and the HATP symbolic task planner
-besides {\sc Oro} and {\sc Dialogs}. In this scenario, the human and the robot
-try to cooperatively remove objects from a table, and behaviours and motions are
-fully planned and executed.
+different aspects of the architecture.  The first one is focused on knowledge
+representation and verbal interaction: the human asks for help to find and pack
+objects ({\it Interactive Grounding I} in Table~\ref{table|experiments}). The
+second one ({\it Cleaning the table}) involves the {\sc Shary} execution
+controller and the HATP symbolic task planner. In this scenario, the human and
+the robot try to cooperatively remove objects from a table, and behaviours and
+motions are fully planned and then executed.
 
@@
 This first study is based on the following scenario: Tom and Jerry are moving to
-London, so they are packing things in boxes. \emph{Jido}, a dual-arm
+London, so they are packing things in boxes. \emph{Jido}, a mobile 
 manipulator, is observing while they move things here and there
 (Figure~\ref{fig|vpt}).  They ask questions to Jido to know where certain
-objects are located (\emph{Where is the white tape? What is behind the toolbox?}
-etc.). This study focuses on multi-modal ,interactive grounding only, and
-the robot does not perform any action.
+objects are located. This study focuses on multi-modal, interactive grounding only, and
+the robot does not actually perform any action besides simple head movements.
 
@@
 situations where we can follow the internal robot's reasoning and the
-interaction with the users.
+interaction with the user.
 
@@
 sentence, and queries the ontology to
-identify the object the human is referring to: \stmt{?obj type VideoTape}.
+identify the object the human is referring to: \concept{find(?obj type
+VideoTape)}.
 
@@
 inside the cardboard box. Thus, the knowledge base returns both: $\Rightarrow$
-\concept{?obj = [videoTape1, videoTape2]}.
+\concept{?obj = [BLACK\_TAPE, WHITE\_TAPE]}.
 
@@
 %eventually execute the command, \ie give it to the human. Alternatively, the
-%robot first asks the human for confirmation (``was \concept{videoTape1} the
+%robot first asks the human for confirmation (``was \concept{BLACK\_TAPE} the
 %object you had in mind?'') before proceeding with the action.
@@
 %\hline
-%  \hspace{0.7cm}\stmt{videoTape1 type VideoTape}\\
-%  \hspace{0.7cm}\stmt{videoTape1 isOn table}\\
-%  \hspace{0.7cm}\stmt{videoTape1 isVisible true}\\
-%  \hspace{0.7cm}\stmt{videoTape2 type VideoTape}\\
-%  \hspace{0.7cm}\stmt{videoTape2 isIn cardBoardBox}\\
-%  \hspace{0.7cm}\stmt{videoTape2 isVisible true}\\
+%  \hspace{0.7cm}\stmt{BLACK\_TAPE type VideoTape}\\
+%  \hspace{0.7cm}\stmt{BLACK\_TAPE isOn table}\\
+%  \hspace{0.7cm}\stmt{BLACK\_TAPE isVisible true}\\
+%  \hspace{0.7cm}\stmt{WHITE\_TAPE type VideoTape}\\
+%  \hspace{0.7cm}\stmt{WHITE\_TAPE isIn cardBoardBox}\\
+%  \hspace{0.7cm}\stmt{WHITE\_TAPE isVisible true}\\
 %\hline
@@
 %\hline
-%  \hspace{0.7cm}\stmt{videoTape1 type VideoTape}\\
-%  \hspace{0.7cm}\stmt{videoTape1 isOn table}\\
-%  \hspace{0.7cm}\stmt{videoTape1 isVisible true}\\
-%  \hspace{0.7cm}\stmt{videoTape2 type VideoTape}\\
-%  \hspace{0.7cm}\stmt{videoTape2 isIn cardBoardBox}\\
-%  \hspace{0.7cm}\stmt{videoTape2 isVisible false}\\
+%  \hspace{0.7cm}\stmt{BLACK\_TAPE type VideoTape}\\
+%  \hspace{0.7cm}\stmt{BLACK\_TAPE isOn table}\\
+%  \hspace{0.7cm}\stmt{BLACK\_TAPE isVisible true}\\
+%  \hspace{0.7cm}\stmt{WHITE\_TAPE type VideoTape}\\
+%  \hspace{0.7cm}\stmt{WHITE\_TAPE isIn cardBoardBox}\\
+%  \hspace{0.7cm}\stmt{WHITE\_TAPE isVisible false}\\
 % \hline
@@
 figure out which box Jerry is talking about. Similar to the previous situation,
-two boxes are visible:
-
-\begin{center}
-\begin{tabular}{l}
-\stmt{?obj type box}\\
-\hspace{0.7cm}$\Rightarrow$ {\tt ?obj = [cardBoardBox, toolbox]}
-\end{tabular}
-\end{center}
-
-However both are also visible to the human and the previous ambiguity resolution
+two boxes are visible: \concept{find(?obj type Box)} $\Rightarrow$ \concept{?obj = [cardBoardBox, toolbox]}
+
+However both are visible to the human and the previous ambiguity resolution
 procedure can not be applied. The robot generates a question (using the
-\emph{Discrimination} algorithm that we introduced at section~\ref{reasoning})
+\emph{Discrimination} algorithm that we mentioned at section~\ref{reasoning})
 and asks Jerry which box he is referring to: ``Which box, the toolbox or the
@@
 %\stmt{?obj isIn cardBoardBox}\\
-%\hspace{0.7cm}$\Rightarrow$ \concept{?obj = videoTape2}\\
+%\hspace{0.7cm}$\Rightarrow$ \concept{?obj = WHITE\_TAPE}\\
 %\end{center}
@@
 \stmt{?obj type VideoTape}\\
-\stmt{?obj differentFrom videoTape2}\\
-\hspace{0.7cm}$\Rightarrow$ \concept{?obj = [videoTape1]}
+\stmt{?obj differentFrom WHITE\_TAPE}\\
+\hspace{0.7cm}$\Rightarrow$ \concept{?obj = [BLACK\_TAPE]}
 \end{tabular}
@@
 
-Since there is only one possible ``other'' video, no specific disambiguation is
+Since there is only one possible ``other'' videotape, no specific disambiguation is
 required, and the robot verbalises the result of the query to the knowledge
-base: ``The other tape is on the table and next to the toolbox.''
-
-\begin{center}
-\begin{tabular}{l}
-\stmt{videoTape1 isOn table}\\
-\stmt{videoTape1 isNextTo toolbox}
-\end{tabular}
-\end{center}
+base \stmt{BLACK\_TAPE isOn table, BLACK\_TAPE isNextTo toolbox} to ``The other
+tape is on the table and next to the toolbox.''
 
@@
 
-This study demonstrates a richer decision-making process. The {\sc Oro} server
-is used in conjunction with the HATP symbolic task planner and the {\sc Shary}
-execution controller to produce and execute a shared plan.
+This study (Figure~\ref{fig|cleantable-video}) exhibits a richer decision-making
+process. The {\sc Oro} server is used in conjunction with the HATP symbolic task
+planner and the {\sc Shary} execution controller to produce and execute a shared
+plan.
 
@@
 %
+
 Figure~\ref{fig|cleantable-timeline} presents a practical sample of the whole
@@
 reachable by the robot only, while the bin where objects are supposed to be
-placed is reachable by the human only: the robot needs to come up with a shared
-plan that involves joint actions.
+placed at the end is reachable by the human only: the robot needs to come up
+with a shared plan that involves joint actions.
 
 The goal is first received by the execution controller (after processing of the
-user request by the {\sc Dialogs} module, not shown on the figure). At
-$t_1$, the tape is computed by the robot as being reachable by the robot only
-(columns \emph{Perception} and \emph{Knowledge}), and the execution controller
-invokes the task planner, which produces a joint plan (column \emph{Plan}).
-
-The first task ({\tt TAKE(GREY\_TAPE, TABLE)}) is instantiated by calling the 3D
-motion planner (column \emph{Actions}, left). The motion planner returns two atomic
-actions ({\tt PICK\_GOTO} followed by {\tt TAKE\_TO\_FREE}) that the controller
-executes (by first reaching the object, grasping it and bringing it back to a
-free position).  Once the robot's perception allows the controller to ensure
-that the task's post-conditions are fulfilled (at $t_2$, by satisfying the
-statement \stmt{ROBOT hasInHand TAPE}), the next task is started (placing the
+user request by the {\sc Dialogs} module, not shown on the figure). At $t_1$ on
+Figure~\ref{fig|cleantable-timeline}, the tape is computed by the robot as being
+reachable by the robot only (columns \emph{Perception} and \emph{Knowledge}),
+and the execution controller invokes the task planner, which produces a joint
+plan (column \emph{Plan}) to move the tape so that the human can pick it and
+drop it into the bin.
+
+The first task ({\tt TAKE(GREY\_TAPE, TABLE)}) is instantiated by first checking
+that the task pre-conditions hold, and then calling the 3D motion planner
+(column \emph{Actions}, left). The motion planner returns two atomic actions
+({\tt PICK\_GOTO} followed by {\tt TAKE\_TO\_FREE}) that the controller executes
+(by first reaching the object, grasping it and bringing it back to a free
+position).  The robot's perception monitors the evolution of the scene until the
+task's post-conditions are verified (at $t_2$, by satisfying the statement
+\stmt{ROBOT hasInHand TAPE}), and the next task is then started (placing the
 tape so that it becomes reachable by the human).
@@
 tape and placing it in the bin) have to be performed by the human: the robot
-instructs the user to do so (not pictured on the figure) and monitors the
-actions of the human to detect when the tasks' post-conditions are satisfied
-(column \emph{Actions}, right). When these post-conditions are fulfilled, the
-goal is considered to be completed.
+instructs the user to do so (verbal interaction not pictured on the figure) and
+monitors the actions of the human to detect when the tasks' post-conditions are
+satisfied (column \emph{Actions}, right). When these post-conditions are
+fulfilled, the goal is considered to be completed.
 
@@
  }
- \caption {Chronology of the \emph{Cleaning the table} scenario. The light
-     arrows indicate the general flow. This figure
-     represents a simplified version: only one object, {\tt TAPE}, must
-     be removed from the table and placed in the bin {\tt BIN}.  }
+
+ \caption {Chronology of the \emph{Cleaning the table} scenario, presented here in a simplified
+ version: a single object, {\tt TAPE}, must be removed from the table and
+ dropped in the bin {\tt BIN}. The light arrows sketch the global execution
+ flow.}
+
   \label{fig|cleantable-timeline}

### 2014-05-07 paper.tex
Typo: Incoming {plan->goal}
Proof-reading section 5 + added TBD notes there

@@
 
-        \node[below=1 of plan] (incoming) {\bf \Large Incoming plan \\ \it Clean the table!};
+        \node[below=1 of plan] (incoming) {\bf \Large Incoming goal \\ \it Clean the table!};
 
@@
 
-This last section tries to clarify the ``take home'' message: what are the challenges
+This last section tries to clarify the ``takeaway'' message: what are the challenges
 that human-robot interaction brings to artificial intelligence, and what is the
@@
 
-We first restate the challenge of \emph{embodied cognition} and list what we
-seen as the main decisional issues that need to be tackled to push forward
-human-robot interaction. We then reflect on the importance of knowledge handling
-in modern robotics and put our choice of the RDF formalism into perspective.
-Finally, we briefly re-discuss two facets of HRI: the importance of
-\emph{modelling the humans}, and the difficulty of \emph{acting in presence of
-humans}.
+We first restate the challenge of \emph{embodied cognition} in general and list
+what we see as the main decisional issues that need to be tackled to push
+forward human-robot interaction. We then reflect on the importance of knowledge
+handling in robotic architectures that deal with complex semantics and put into
+perspective our choice of the RDF formalism.  Finally, we briefly re-discuss two
+facets of HRI: the importance of \emph{modelling the humans}, and the difficulty
+of \emph{acting in presence of humans}.
 
@@
 
-Robotics is traditionally regarded as the prototypical instance of \emph{embodied}
-artificial intelligence, and this dimension is especially prevalent in
-human-robot interaction, where agents have to share a joint physical
-environment.
-
-It results in a tight coupling between the symbolic and the geometric realms:
+Robotics is traditionally regarded as the prototypical instance of
+\emph{embodied} artificial intelligence, and this dimension is especially
+prevalent in human-robot interaction, where agents have to share and collaborate
+in a joint physical environment.
+
+This leads to a tight coupling between the symbolic and the geometric realms:
 while AI at its origins was mostly a matter of symbolic models, it has been
@@
 
-The challenge of symbol grounding is tightly linked to this issue. It
+The challenge of symbol grounding is also tightly linked to this issue. It
 corresponds to the identification or creation, and then, maintenance of a link
-between the symbol (the syntactic form of knowledge the computer will
-manipulate) and its semantics, \ie its meaning, anchored in the world (the
+between the symbol (the syntactic form of knowledge that the computer
+manipulates) and its semantics, \ie its meaning, anchored in the world (the
 relations between the symbol, the referent of the symbol, and mediating minds
-is classically referred as the \emph{semantic triangle}, and has been
+is classically referred as the \emph{semantic triangle} and has been
 extensively studied in linguistics). The issue of grounding is well known in
@@
 
-Grounding is implemented at different levels in our architecture. The main source
-of grounded symbolic knowledge is the situation assessment module, {\sc Spark}.
-Not only it builds symbolic facts from spatial reasoning, but it also relies on
-temporal reasoning to track the world state and build explanations to interpret
-unexpected perceptions (we mentioned above the example of a disappearing
-object).
+As we have seen, grounding is implemented at different levels in our
+architecture. The main source of grounded symbolic knowledge is the situation
+assessment module, {\sc Spark}.  Not only it builds symbolic facts from spatial
+reasoning, but it also relies on (limited) temporal reasoning to track the world
+state and build explanations to interpret unexpected perceptions (like a
+suddenly disappearing object).
 
@@
 it already knows). Non verbal interaction (like gestures) is also partially
-grounded. When the user points at something, it can say ``give me this'' and
-the robot grounds \emph{this} to the pointed artifact.
+grounded: We have presented how pointing, for instance, is used by the robot to
+ground \emph{``this''} or \emph{``that''} to the pointed artifact.
 
@@
 the task. It remains to be seen if systematic marking of manipulated objects
-(for instance with RFID tags) is a realistic path. It may be the case if the
+(for instance with RFID tags) is a realistic path.\footnote{It may be the case if the
 robot is expected to be used in a well-defined, slowly changing environment.
 Real-time recognition of large libraries of objects is another path which may
-be viable soon~\cite{Dean2013Fast}.
+be viable soon~\cite{Dean2013Fast}.}
+
+We see however that we have only scratched the surface of a generic grounding
+capability. Context, cultural background, ``naive physics'' knowledge, emotional
+state of the human are some of the numerous other determinants that need to be
+accounted for when grounding human-robot interaction.
 
@@
 and consistent repository of facts, accessible for all modules. It relies on a
-\emph{strict formalism} (OWL statements), with a \emph{well defined vocabulary}
+strict formalism (OWL statements), with a well defined vocabulary
 (stated in the common-sense ontology). These first two points lead to a
@@
 inference abilities of {\sc Oro} to trigger events whose \texttt{true}
-conditions can be inferred.
+conditions can be (possibly indirectly) inferred.
 
@@
 benefits to components. For instance, the dialogue processing module can
-perfectly run without any geometric perception, but its disambiguation routines
+run without any geometric perception, but its disambiguation routines
 can transparently benefit from it when available (since richer symbolic
@@
 
-Knowledge flows in our architecture through a \emph{semantic blackboard}, the
-{\sc Oro} server. We rely on Description Logics (OWL) to represent and
-manipulate facts.
+The {\sc Oro} server relies on Description Logics (OWL) to represent and
+manipulate knowledge.
 
@@
 KnowRob}~\cite{Tenorth2009a}) that mix RDF with more expressive logic languages
-like {\sc Prolog} with other limitations, like closed-world reasoning. The
-classification performance is another issue: from our experience, with an
+like {\sc Prolog} with other limitations, like closed-world reasoning.
+Classification performance is another issue: from our experience, with an
 ontology sized for a standard experiment (about 100 classes and 200 instances),
@@
 
-Alternatives like \emph{Answer Set Programming} have also been
+While alternatives like \emph{Answer Set Programming} have been
 successfully investigated in robotics~\cite{Chen2010,Erdem2012}, in particular
-to deal with non-monotonic reasoning. However we did not actually hit any brick wall
+to deal with non-monotonic reasoning, we did not actually hit any brick wall
 while working with OWL ontologies. We may reconsider this choice at a later
@@
 conceptually and technically easy to add support for independent knowledge
-models, one per agent the robot interacts with --see section~\ref{sect|tom}).
+models, one per agent the robot interacts with).
 
@@
 
+\subsubsection{Limits of Disambiguation at Semantic Level}
+
+Interaction with humans implies the ability to deal with semantics: semantics of
+verbal interaction, semantics of gestures, etc.  As a consequence, it also
+implies to deal with semantic disambiguation.
+
+One prototypical example of semantic disambiguation has been given in
+\cite{Ros2010b} with the child game \emph{spygame}: two players are facing
+each other with a set of random objects in-between, one player mentally choose
+one object, and the other player has to guess the object by asking closed
+questions like \emph{Is your object small or large?} Based on the knowledge it
+has acquired, the robot is able to minimise the number of questions required to
+find the object.
+
+When playing this kind of game, however, the issue arises that the robot has no
+way to select which knowledge about the object is relevant in the interaction
+context. For instance, the knowledge base may store facts like \stmt{obj1 type
+ActiveConcept} (which internally means that this concept was mentioned in a
+discussion in the last few seconds): this information is not a relevant
+property of \concept{obj1} when trying to disambiguate concepts with humans.
+This distinction between \emph{internal knowledge} (meaningful to
+the system only) and \emph{common knowledge} (whose meaning is understood by
+all the interactors) has not been properly dealt with in our architecture.
+
+Besides, even knowledge that belongs to the \emph{common knowledge} may not be
+appropriate in a given interaction context. For instance, the system may
+compute that at a given instant the human is looking at the object: \stmt{human
+looksAt obj1}. This property makes sense to both parties, but in the context of
+the \emph{spygame}, we would like to mainly use immanent properties, not
+volatile like a gaze. More research is required to identify relevant
+interaction contexts and knowledge classes attached to them.
+
 \subsection{Bringing the Humans into the Equation}
@@
 
-Perspective taking is this human ability which allows one to put him/herself in
-another person's point of view. Studied in psychology
-literature~\cite{Flavell1992,Tversky1999}, this ability is crucial when
-interacting with people by allowing one to reason on others' understanding of
-the world in terms of visual perception, spatial descriptions, affordances and
-beliefs, etc.  In the last years these notions have been gradually employed in
-HRI.~\cite{Breazeal2006} presents a learning algorithm that takes into account
-information about a teacher's visual perspective in order to learn a task.
-~\cite{Johnson2005} apply visual perspective taking for action recognition
-between two robots.~\cite{Trafton2005} use both visual and spatial perspective
-taking to find out the referent indicated by a human partner.
+\fxfatal{Re-organise, rephrase, complement}
 
@@
 
-HRI offers new challenges to planning
-by adding a complementary constraint: the synthesis of plans that are not only
-feasible but legible (in terms of their intention) and compliant with human
-preferences. This constraint is often ill-defined, dynamic, context-dependent.
-
-Indeed, we posit that human-robot collaborative task achievement calls for more
-subtlety for a robot than autonomous independent task achievement. HATP, by
-co-planning for the robot and the human, and modelling social rules as planning
-parameters, is a first step in this direction.
+HRI offers new challenges to planning by adding a supplementary constraint: the
+synthesis of plans that are not only feasible but legible (in terms of their
+intention) and compliant with human preferences. This constraint is often
+ill-defined, dynamic and context-dependent.  HATP, by co-planning for the robot and
+the human, and modelling social rules as planning parameters, is a first step in
+this direction.
+
+\fxfatal{What are the next steps?}
 
 \paragraph{Joint actions with humans}\label{sec:soa}
+\fxfatal{too many citation -> synthesis needed!}
 The human presence brings new requirements for robot's abilities both
@@
 
-\paragraph{Limits of Disambiguation at Semantic Level}
-
-Interaction with humans implies the ability to deal with semantics: semantics of
-verbal interaction, semantics of gestures, etc.  As a consequence, it also
-implies to deal with semantic disambiguation.
-
-One prototypical example of semantic disambiguation has been given in
-\cite{Ros2010b} with the child game \emph{spygame}: two players are facing
-each other with a set of random objects in-between, one player mentally choose
-one object, and the other player has to guess the object by asking closed
-questions like \emph{Is your object small or large?} Based on the knowledge it
-has acquired, the robot is able to minimise the number of questions required to
-find the object.
-
-When playing this kind of game, however, the issue arises that the robot has no
-way to select which knowledge about the object is relevant in the interaction
-context. For instance, the knowledge base may store facts like \stmt{obj1 type
-ActiveConcept} (which internally means that this concept was mentioned in a
-discussion in the last few seconds): this information is not a relevant
-property of \concept{obj1} when trying to disambiguate concepts with humans.
-This distinction between \emph{internal knowledge} (meaningful to
-the system only) and \emph{common knowledge} (whose meaning is understood by
-all the interactors) has not been properly dealt with in our architecture.
-
-Besides, even knowledge that belongs to the \emph{common knowledge} may not be
-appropriate in a given interaction context. For instance, the system may
-compute that at a given instant the human is looking at the object: \stmt{human
-looksAt obj1}. This property makes sense to both parties, but in the context of
-the \emph{spygame}, we would like to mainly use immanent properties, not
-volatile like a gaze. More research is required to identify relevant
-interaction contexts and knowledge classes attached to them.
-
 \subsection{The Next Steps}
@@
 Contexts are currently often limited to the current spatial and temporal
-situation. Some projects model offer the possibility to jump in the past or to
+situation. Some models offer the possibility to jump in the past or to
 switch to another agent's perspective, but in current approaches, selecting a

### 2014-05-16 paper.tex
Rachid's comments (typos/style...)

@@
 
-Human-Robot interaction is an area filled with challenges for artificial
+Human-Robot interaction is an area full of challenges for artificial
 intelligence: dynamic, partially unknown environments that are not originally
@@
 that requires fine, low-latency control, representation and management of
-several mental models, good situation assessment skills...the list goes on.
+several mental models, pertinent situation assessment skills...the list goes on.
 
@@
 exploitation of each agent (human and robot) knowledge in separate cognitive
-models, natural multi-modal communication, human-aware task planning, and human
+models, natural multi-modal communication, ``human-aware'' task planning, and human
 and robot interleaved plan achievement.
@@
 lays at the crossroad of several domains of AI and requires to tackle them in a
-holistic manner: Modelling humans and human cognition; acquiring, representing,
+holistic manner: modelling humans and human cognition; acquiring, representing,
 manipulating in a tractable way abstract knowledge at the human level; reasoning
@@
 interactive object manipulation, fetch and carry tasks and other similar
-domestic tasks by taking into account, at every stages, the intentions, beliefs,
+domestic tasks by taking into account, at every stage, the intentions, beliefs,
 perspectives, skills of the human partner.  Namely, the robot must be able to
@@
 
-\emph{Communication}, \emph{Joint Actions}, \emph{Human-Aware Execution}: these
-three challenges structure the whole field of human-robot interaction research,
-and they can be analysed in terms of the cognitive skills that they require.
+These three challenges, \emph{Communication}, \emph{Joint Actions},
+\emph{Human-Aware Execution}, structure the whole field of human-robot
+interaction research, and they can be analysed in terms of the cognitive skills
+that they require.
 
@@
 operation continues until the goal is achieved, is declared unachievable or is
-abandoned by the human.
+abandoned by the human~\cite{Klein2004}.
 
@@
 
-Such abilities must be designed and implemented in generic ways and
+Such abilities must be designed and implemented in a generic way and
 must provide several levels of parametrization such as they adapt to
@@
 section~\ref{sec:impl} gives some insights on each of the main components of the
-architecture, and try to highlight their significance for artificial
+architecture, and tries to highlight their significance for artificial
 intelligence. Section~\ref{sec:expe} presents two studies that illustrate in a
@@
   \path (spark.100) edge [->, bend right] node[label] {symbolic \\ facts} (oro);
+  \path (spark.5) edge [->, bend right] node[label] {environment\\model} (mhp);
   \path (shary) edge [<->, bend left] node[label] {events, \\ world model and \\ agents beliefs} (oro);
@@
   \path (lowlevel) edge [->] (spark);
-  \path (lowlevel.east) edge [<-, bend right=80, looseness=1.2] (shary.east);
+  \path (lowlevel.east) edge [<-, bend right=80, looseness=1.2] node[label] {atomic\\actions} (shary.east);
 
@@
 In particular, dealing with the rich semantics of natural interaction with
-humans has been one of our major focus. To this end, we have been deploying
+humans has been one major focus. To this end, we have been deploying
 over the last years explicit knowledge representation and manipulation in the
@@
 
-To some extend, this architecture moves away from classical layered approaches
+To some extent, this architecture moves away from classical layered approaches
 found in robotics~\cite{Gat1998three, Volpe2001CLARAty, Goldberg2002}.
@@
 the knowledge base to interpret and disambiguate natural language, and in that
-regards, it is fully integrated to the deliberative process itself (we detail
+regards, it is fully integrated to the deliberative process itself (we explain
 natural language processing at section~\ref{sect|com}).
@@
 manipulated). The second part of the knowledge is acquired at run-time from
-perception, interaction and planning. The next sections detail these processes.
-Lastly, the third source of symbolic statements is the reasoner itself.
+perception, interaction and planning. The next sections go into details of these
+processes.  Lastly, the third source of symbolic statements is the reasoner
+itself.
 
@@
     \node [node distance=1, below right=of thing] (tthing) {temporal thing} edge (thing);
-        \node [below=of tthing] (tte) {thing with temporal extend} edge (tthing);
+        \node [below=of tthing] (tte) {thing with temporal extent} edge (tthing);
             \node [below right=of tte] {time interval} edge (tte);
@@
 other on-line databases (like Wikipedia or WordNet), the reuse of key {\sc
-OpenCyc} concepts ensures to a certain extend that the knowledge stored by the
+OpenCyc} concepts ensures to a certain extent that the knowledge stored by the
 robot can be shared or extended with well-defined semantics. A good example is
-the concept of \emph{Object}: In everyday conversation, an object is a
+the concept of \emph{Object}: in everyday conversation, an object is a
 relatively small physical thing, that can be typically manipulated. Normally, a
@@
 The set of actions that the robot can interpret typically corresponds to the
-planing domain in use: the robot can only process the \concept{GoTo} action if
+planning domain in use: the robot can only process the \concept{GoTo} action if
 the corresponding task description with its pre- and post-conditions is
@@
 link that the robot has to established between its percepts and its symbol must
-\emph{model} to a large extend the human representations in order to effectively
+\emph{model} to a large extent the human representations in order to effectively
 support communication.
@@
 actuation capabilities of the robot, and our architecture adopts here also a
-bidirectional approach.  The different components that we have mentioned so far
-provide a \textit{bottom-up} grounding process: geometric reasoning and dialogue
-processing modules constantly build and push new symbolic contents about the
-world to the knowledge base where it becomes accessible to decisional layers. In
-parallel, the knowledge base relies on reasoning in a \textit{top-down} way to
-produce new facts, \ie new beliefs that may lead to actual decision making and
-physical actions.
+bidirectional approach: \emph{bottom-up} and \emph{top-down}.  The different
+components that we have mentioned so far provide a bottom-up grounding process:
+geometric reasoning and dialogue processing modules constantly build and push
+new symbolic contents about the world to the knowledge base where it becomes
+accessible to decisional layers. In parallel, the knowledge base relies on
+reasoning in a top-down way to produce new facts, \ie new beliefs that may lead
+to actual decision making and physical actions.
 
@@
 \begin{inparaenum} In our context, we call \emph{cognitive skills} the
-deliberative behaviours that are \item \emph{stateful} both in terms of
-\emph{time} (keeping track of previous states is required for the component
-to perform adequately) and \emph{modalities} (for instance, \emph{situated}
-dialogue requires interleaving different communication modalities), \item
-manipulate \emph{explicit semantics}, typically by the mean of symbolic
-reasoning, \item operate at the \emph{human-level}, \ie are legible to the
-humans, typically by acting at similar levels of abstraction.\end{inparaenum}
+deliberative behaviours that are \item \emph{stateful} (keeping track of
+previous states is typically required for the component to perform adequately),
+\item \emph{amodal} in that the skill is not inherently bound to a specific
+perception of actuation modality, \item manipulate \emph{explicit semantics},
+typically by the mean of symbolic reasoning, \item operate at the
+    \emph{human-level}, \ie are legible to the humans, typically by acting at
+similar levels of abstraction.\end{inparaenum}
 
@@
 
-\subsubsection{Reasoning}
+\subsubsection{Symbolic Reasoning}
 \label{reasoning}
@@
 mark for a few seconds important concepts that are currently manipulated by the
-robot as \emph{active concepts}: If a human asks the robot ``Give
+robot as \emph{active concepts}: if a human asks the robot ``Give
 me all red objects'', the human, the \concept{Give} action, and each red
@@
 memory (which can be considered, in this case, to be a working memory). We use
-this feature to trace certain knowledge-related processes.
+this feature to trace certain knowledge-related processes. On the other hand,
+our perception layer does not make use of this mechanism: as described in the
+next section, the environment model of the robot is continuously updated and the
+derived symbolic knowledge is therefore transient.
 
@@
         box/.style={draw,rectangle, dotted, minimum width=#1, minimum height=3.8cm},
-        box/.default={4.5cm},
+        box/.default={4cm},
         every edge/.style={draw, ultra thick},
@@
 
-    \node[draw=none] at (0,-1) {\includegraphics[width=0.8\columnwidth]{spark-oro-notext.pdf}};
-
-    \node[box=16cm, minimum height=5.7cm, solid, thick, rounded corners] at (0,1.8) (oro) {};
+    \node[draw=none] at (0,-1.5) {\includegraphics[width=1.\columnwidth]{spark-oro-notext.pdf}};
+
+    \node[box=16cm, minimum height=5.3cm, solid, thick, rounded corners] at (0,1.8) (oro) {};
     \node at (0,4.2) {\sc \Large oro-server};
@@
 
-    \node at (-3,-0.6) {{\it OWL store:} OpenJena};
-    \node at (3,-0.6) {{\it Reasoner:} Pellet};
+    \node at (-3,-0.4) {{\it OWL store:} OpenJena};
+    \node at (3,-0.4) {{\it Reasoner:} Pellet};
 
@@
 
-    \node[box=16cm, minimum height=3.5cm, dotted, thick, rounded corners] at (0,-3.75) (spark) {};
+    \node[box=18cm, minimum height=4.5cm, dotted, thick, rounded corners] at (0,-4.2) (spark) {};
     \path (spark.north) edge[->] node[align=left,anchor=west] {Symbolic\\facts} (oro.south);
 
-    \node[anchor=north] at (-5.5,-4.5) {a) Agent-independent\\locations};
-    \node[anchor=north] at (-1.8,-4.5) {b) Agent-dependent\\placements};
-    \node[anchor=north] at (1.5,-4.5) {c) Visibility};
-    \node[anchor=north] at (5.3,-4.5) {d) Reachability};
+    \node[anchor=north] at (-6.5,-5.3) {a) Agent-independent\\locations};
+    \node[anchor=north] at (-2.1,-5.3) {b) Agent-dependent\\placements};
+    \node[anchor=north] at (1.6,-5.3) {c) Visibility};
+    \node[anchor=north] at (6.5,-5.3) {d) Reachability};
 
@@
 
-{\sc Spark} is an \emph{amodal}~\cite{Mavridis2006} geometric model of the
+{\sc Spark} embeds an \emph{amodal}~\cite{Mavridis2006} geometric model of the
 environment that serves both as basis for the fusion of the perception
-modalities and as bridge with the symbolic layer. This geometric model is
-updated at run-time by the robot based on its sensors (in most of
-our studies, objects are identified and localised through 2D fiducial markers,
-while humans are tracked with Kinect-like devices, optionally assisted by motion
-capture to accurately track the head motion, which is required to compute what
-the human is looking at).
+modalities and as bridge with the symbolic layer. This geometric model is made
+of CAD 3D models of the objects, furnitures and robots, and full body, rigged
+models of the humans, and it is updated at run-time by the robot based on its
+sensors (in most of our studies, objects are identified and localised through 2D
+fiducial markers, while humans are tracked with Kinect-like devices, optionally
+assisted by motion capture to accurately track the head motion, which is
+required to compute what the human is looking at).
 
@@
 {\sc Spark} is designed to run and continuously update the knowledge base (in
-our experiments, it runs at about 10Hz). At each steps, it re-computes spatial
+our experiments, it runs at about 10Hz). At each step, it re-computes spatial
 relationships and affordances for the whole scene, and send the delta (new
@@
 
-\item \emph{Sees}: An important ability to know about an agent is to predict
-    \emph{What can it see?}, \ie what is within its field of view (FOV). Being
-    able to compute this information (by dynamically placing an OpenGL camera at
-    the location of the eyes and running occlusion checks from it), the robot
-    can reuse it for instance to infers which object a human is searching for
-    (the one that is not currently visible, otherwise the user would not be
-    searching for it).  In figure~\ref{fig|spark-oro}\emph{c} the field of view
-    of a person is illustrated with a grey cone (broader one). While he is able
-    to see the two small boxes on the table in front of him, the big box on his
-    right is out of his FOV, and therefore, he is not able to see it.
+\item \emph{Sees}: this relation describes what the agent can see, \ie what is
+    within its field of view (FOV). In our current implementation, this
+    affordance is computed by dynamically placing an OpenGL camera at the
+    location of the eyes and running occlusion checks from it.  In
+    figure~\ref{fig|spark-oro}\emph{c} the field of view of a person is
+    illustrated with a grey cone (broader one). While he is able to see the two
+    small boxes on the table in front of him, the big box on his right is out of
+    his FOV, and therefore, he is not able to see it.
 
@@
 symbolic states of objects (whether the object is on a furniture, in an agent
-hand, in a container, etc.): According to the robot perception of what has
+hand, in a container, etc.): according to the robot perception of what has
 happened since the object was last seen, the robot builds a belief of the
@@
 segmentation and recognition. Each object has a unique identifier,
-which enable us to load a proper CAD model suitable for 3D spatial reasoning
+which enables us to load a proper CAD model suitable for 3D spatial reasoning
 and prevent recognition mistakes in the knowledge base. While {\sc Spark}
@@
 instance) is not properly addressed in our stack. Temporal reasoning is used
-only locally, and does not allow for tracking of long or global events.
+only locally, and does not allow for tracking of long sequences or global events.
 
@@
 human-robot interaction for which the introduction of the semantic layer has
-been most beneficial.  In~\cite{Lemaignan2011a} we detail the techniques and the
+been most beneficial.  In~\cite{Lemaignan2011a} we explain the techniques and the
 tool called {\sc Dialogs} that we have developed for natural English language
@@
 interprets common temporal and place adverbs (like \emph{above} or
-\emph{tomorrow}) and translates to a certain extend \emph{states} (``I'm
+\emph{tomorrow}) and translates to a certain extent \emph{states} (``I'm
 tired'') into \emph{experiences} (\stmt{HUMAN experiences state\_1, state\_1
@@
 
-\subsubsection{Multi-Modal Communication}
+\subsubsection{Multi-Modality}
 
@@
 behaviour when the human expresses (through dialogue, gestures...) that he
-wants to somehow get a toy.
+wants a toy.
 
@@
 Manipulation and navigation actions listed in table~\ref{table|pyrobots_actions}
-rely ona dedicated 3D motion planner ({\sc Mhp}, depicted in
+rely on a dedicated 3D motion planner ({\sc Mhp}, depicted in
 Figure~\ref{fig|archi}): robot placement and end-effectors trajectories are
@@
 specific constraints and human postures, abilities and preferences.
-\cite{Sisbot2008, Mainprice2011, Pandey2011} detail these techniques.
+We kindly refer the interested reader to~\cite{Sisbot2008, Mainprice2011,
+Pandey2011} for details of these techniques.
 
@@
 relies on its relation to the physical world (so-called \emph{embodied
-cognition}). Varela~\cite{Varela1992} is one of the main discoverer of these
+cognition}). Varela~\cite{Varela1992} is one of the main discoverers of these
 mechanisms, and coined the concept of \emph{enactivism} as the theoretical
@@
 it already knows). Non verbal interaction (like gestures) is also partially
-grounded: We have presented how pointing, for instance, is used by the robot to
+grounded: we have presented how pointing, for instance, is used by the robot to
 ground \emph{``this''} or \emph{``that''} to the pointed artifact.
@@
 As a whole, the components that we have introduced so far build a
-knowledge-oriented architecture: Knowledge is explicitly stored in one central
+knowledge-oriented architecture: knowledge is explicitly stored in one central
 and consistent repository of facts, accessible for all modules. It relies on a
@@
 And finally, this architecture allows for the combination of very different
-knowledge modalities in a single homogeneous environment, bringing mutual
+knowledge sources in a single homogeneous environment, bringing mutual
 benefits to components. For instance, the dialogue processing module can
-run without any geometric perception, but its disambiguation routines
+run without any situation assessment, but its disambiguation routines
 can transparently benefit from it when available (since richer symbolic

### 2014-05-18 paper.tex
Contribution Rachid - HATP

@@
 desires into a succession of atomic actions. We use in our architecture the HATP
-planner (for Human Aware Task Planner~\cite{Alili2008, Alili2009}).  HATP is a
-refinement of a Hierarchical Task Network (HTN). As an HTN, it performs an
-iterative task decomposition into sub-tasks until reaching atomic actions. The
-\emph{planning domain} defines the set of methods describing how to decompose a
-task and represents the procedural knowledge of the robot. It is stored outside
-of the central knowledge base, using a specific formalism (see the related
-discussion at the end of this section).
+planner (for Human Aware Task Planner~\cite{Alili2008,
+Alili2009,Lallement2014}).
+
+HATP (Hierarchical Agent- based Task Planner) planning framework extends the
+traditional HTN (Hierarchical Task Network) planning domain representation and
+semantics by making them more suitable to elaborate plans which involve humans
+and robots acting together toward a joint goal. HATP is used by the robot to
+elaborate human-robot \emph{shared plans} which are then used to anticipate
+human action, to propose to human to act or even to ask help from the human.
+
+The HATP \emph{planning domain} defines a set of methods describing how to
+decompose a task and represents the procedural knowledge of the robot as well as
+its knowledge about the actions that the human partner is able to achieve. It is
+stored outside of the central knowledge base, using a specific formalism (see
+the related discussion at the end of this section).
 
@@
 actions \emph{as well as} for the other participants (humans or robots), that we
-call \emph{shared plans}. This effectively enriches the interaction capabilities
+call \emph{shared plans}.
+
+An important feature of HATP is that it treats agents as ``first-class
+entities'' in the domain representation language. It can therefore
+distinguish between the different agents in the domain as well as
+between agents and the other entities such as tables and chairs. This
+facilitates a post-processing step in HATP that splits the final
+solution (sequence of actions) into two (or more if there are several
+humans) synchronised solution
+streams, one for the robot and one for the human, so that the streams may be executed in
+parallel and synchronised when necessary.
+
+This effectively enriches the interaction capabilities
 of the robot by allowing for a form of human behaviour \emph{prediction}, which
@@
 
-In addition to these costs, HATP also takes into account a set of social rules.
-Social rules are heuristics aiming at leading the plan construction towards the
-best plan according to some human preferences.  The social rules we have defined
-so far deal with \emph{undesirable states} to avoid a state in which the human
-could feel uncomfortable; \emph{undesirable sequences} to eliminate sequences of
-actions that can be misinterpreted by the human; \emph{effort balancing} to
-adjust the work effort of each of the agents; \emph{no wasted time} to avoid
-long delays between the actions of the human partner; and \emph{no intricate
-links} to limit dependencies between the actions of two or more agents.
+To address some of those challenges HATP includes mechanisms called
+"social tules" to filter plans so as to keep only those suitable for
+HRI. To this end, HATP allows the specification of the following
+filtering criteria:
+
+\emph{Wasted time}: Avoids plans where the human mentioned in a plan spends a lot
+of its time waiting.
+
+\emph{Effort balancing:} Avoids plans where efforts are
+not fairly distributed among the agents mentioned in a plan.  It is
+indeed sometimes beneficial to balance efforts between the human and
+the robot.
+
+\emph{Control of intricacy:} Avoids plans with too many interdependencies
+between the actions of agents mentioned in the plan, as a problem with
+executing just one of those actions could invalidate the entire
+plan. Also intricate human-robot activity may cause discomfort since
+the human will find himself repeatedly in a situation where his is
+waiting for the robot to act
+
+\emph{Undesirable sequences:} Avoids plans that violate specific user-defined
+sequences (for instance sequences which can be misinterpreted by the human).
+
+Combining some of the above criteria could help yield the
+following interesting \emph{shared plans}: (1) the human ends up doing a
+number of the tasks, but yet the overall effort is significantly lower than what the robot puts
+to do a lower number of effort-intensive tasks; (2) avoiding, when
+possible, having the human wait for the robot several times, which
+essentially prevents the streams from having too many causal links
+between them.
+
+ In the current implementation, the so-called social rules are
+implemented by looking through all the plans produced and filtering
+out the ones that do not meet the requirements specified. In the
+future we intend to study algorithms that do such filtering online,
+rather than after primitive solutions are found.
 

### 2014-05-18 paper.tex
Contribution Rachid - Execution control

@@
 
-%The novelty of our contribution here is the interleaving of a semantic-aware
-%architecture designed for autonomy with human-aware decisional components. To
-%our knowledge, no other robotic architecture covers in depth those two
-%perspectives.
-%
-%The Ke Jia project~\cite{Chen2010} is maybe one of the closest
-%approach. It integrates on a mobile platform a knowledge representation language
-%with natural language processing, task planning and motion planning. Default and
-%non-monotonic reasoning has been especially researched within the Ke Jia project
-%for symbolic task planning~\cite{Ji2011} and underspecified natural language
-%processing. The Ke Jia robot has been demonstrated in several tasks involving human-robot
-%interaction with natural language. These tasks include a task with multiple
-%\emph{pick \& carry} that are globally optimised, naive physics reasoning via
-%taught rules or more complex scenarios with the robot delivering drinks, taking
-%into account changing and mutually exclusive preferences of users.
-%
-%Also notable, Beetz et al.~\cite{Beetz2010} proposes a cognitive architecture
-%called \textsc{CRAM} (Cognitive Robot Abstract Machine) that deeply integrates
-%with \textsc{KnowRob}~\cite{Tenorth2009a}, a knowledge processing framework
-%based on Prolog. Its underlying storage is based on an OWL ontology, derived
-%from \textsc{OpenCyc}. \textsc{CRAM} and \textsc{KnowRob} have been demonstrated
-%on several real-world scenarios, where natural language recipes extracted from
-%the Internet had to be translated into plans and executed in a kitchen
-%environment, perceived and rebuilt on-line by the robots. This architecture has
-%however not been deployed in many scenarios involving human-robot interactions.
-%
+One of the main role of {\sc Shary} is to control the elaboration and execution
+of shared plans~\cite{Grosz1996,Clark1996,Kemp2007}. This means essentially
+contex-based refinement and robot actions execution as well as monitoring of
+those achieved by its human partner~\cite{Breazeal2003}. The challenge is to
+build such abilities in a generic way, and to provide several levels of
+parametrization allowing to adapt to various environments, and various levels of
+involvement of the robot ranging from teammate behavior to assistant or
+proactive helper.  Depending on this, the robot controller invokes the various
+human-aware planners and will react to events detected at ORO level.
+
+SHARY’S originality, as a supervision system, lies in its ability to take into
+account not only the task achievement but also communication and monitoring
+needed to support interactive task achievement ~\cite{Rich1997,Sidner2005} in a
+flexible way. SHARY allows to link action execution to more or less elaborated
+“communication policies” in order to produce signals towards the human and to
+react to human action.
+
+A communication act is the central notion in the formalism. It plays the role of
+transition condition. Consequently, a communication act represents an
+information exchange between the two agents. This exchange can be realized
+through dialog or by an expressive motion or a combination of the two. It
+enables each agent to communicate his belief about the task to be realized in
+order to share mutual knowledge and to synchronyze their activities.  This is
+done through real-time task-based situation assessment achieved by the
+combination of SPARK monitoring processes and ORO inference mechanisms
+\cite{fiore2014}.
 

### 2014-05-18 paper.tex
Contribution Rachid - Key decisional issues

@@
 
-How can we restate the main challenges, in particular in term of decision
-making, that human-robot interaction brings to artificial intelligence?
-
-\fxfatal{TDB}
+Human-robot interaction and more particularly one of it major subfields
+which is human-robot collaboration provides a rich and renewed context
+and challenge for the decisional abilities of a robot.
+Indeed, the context where a robot has to achieve a task together with
+humans gives rise to the following questions, each of them being a
+challenge of is own for the robot decisional capabilities  what, who,
+when; where and how ?
+
+\emph{What to do next,} at different levels of abstractions and taking into
+account several not only the current state but also the a long term
+goal, is the basic or standard question for an intelligent robot. Its
+is enriched here by complementary considerations dealing with the
+physical and mental state of its human partner, the set of possible
+actions are enriched by communication robot acts.
+
+\emph{Who should act now} is of key importance and need also to be
+decided upon by the robot. Indeed, it is sometimes better - one could
+say ``more intelligent'' - for the robot to wait and leave ist human
+partner act. Turn-taking opens various decisional challenges.
+
+\emph{When to perform a given action} is enriched here by the
+human-context since it has take into account the human, his needs, his
+rhythms and his mental state. While performing its share of the task
+the robot will have to produce signals towards the human and to
+respond to signals produced by him at the proper pace.
+
+\emph{Where to perform an action} plays an important role in presence of
+humans since here again, the choice is not trivial and might need
+elaborated decision. Robot here has to take into account effort
+sharing, visibility of its action by the human, disturbance or
+discomfort  induced by its action.
+
+\emph{How to perform an action} plays a key role when it comes to a
+robot which has to interact with humans. Very often, there are
+numerous ways to perform an action or to achieve a goal. Deciding which
+one to perform is a non trivial decision problem. One way to take into
+account this aspect is to develop cost-based planners which tend to
+search for plans that satisfy an acceptable "cost" in terms, for
+instance, of ``acceptability'' or ``legibility''.
+
+It is also clear that these different questions are not independent
+and should often be considered in one decision step. The human-aware task and
+motion planners, which we have built, are instances of systems which
+have been specially designed to deal with this intricacy.
+
+In one word, the quality of the decision will be measured here in
+terms of ``pertinence'' of robot behaviour. This notion covefrs
+several aspects as dicussed above, including efficiency.
 

### 2014-05-18 paper.tex
Contribution Rachid - Bringing human into equation

@@
 
-\subsubsection{Modelling the Humans}
-
-\fxfatal{Re-organise, rephrase, complement}
+One essential and incredibly complex question is the consideration of
+humans in the system, how they are modeled and how these models are
+used by the robot to perceive and interpret their motion and action,
+to synthesize its behaviour in order to act on the environment and to
+interact with its human partners.
+
+We claim that it is possible and useful to devise and use models of
+the humans, how they act, their needs and preferences. This can be
+done in very limited, but still interesting and useful, contexts such
+as collaborative human-robot objects manipulation, fetch-and-carry and
+associated activities in a domestic or professional environment.
+
+\subsubsection{Estimate human mental state}
 
@@
 The cognitive model that the robot builds for the agent it interacts with is
-still simple and mostly focused on geometric features (\emph{Who sees what? What are
-our relative positions?} etc.). Extending this knowledge with more subtle
+still simple and mostly focused on geometric features and affordances (\emph{Who sees what? What are
+our relative positions? reachability estimation}  etc.). Extending this knowledge with more subtle
 perceptions (emotional state for instance) remains to be done.
 
-We should also mention that modelling of human behaviours also appear in our
-architecture at the level of task and motion planning with a set of
-\emph{social rules}, including proxemics, action legibility or time
-optimization.
+\subsubsection{Motion and action execution in close vicinity to the human}
+
+The human model we use embeds also human preferences and associated
+constraints that could be taken into account when synthesizing robot motion
+or elaborating robot plans.
+
+This includes proxemics (human-robot distance) and associated issues
+(visibility) but also legibility and acceptability criterias
+expressed in terms of \emph{social rules} that the elaborated plans should
+satisfy.
 

### 2014-05-19 paper.tex
Rephrase, re-organize discussion, taking into account Rachid's contributions

--- 2014-05-18 paper.tex
@@
 independent declarative and procedural knowledge stores.
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 parametrization allowing to adapt to various environments, and various levels of
-involvement of the robot ranging from teammate behavior to assistant or
+involvement of the robot ranging from teammate behaviour to assistant or
 proactive helper.  Depending on this, the robot controller invokes the various
@@
 
-This last section tries to clarify the ``takeaway'' message: what are the challenges
-that human-robot interaction brings to artificial intelligence, and what is the
-current state-of-the-art in this field?
-
-We first restate the challenge of \emph{embodied cognition} in general and list
-what we see as the main decisional issues that need to be tackled to push
-forward human-robot interaction. We then reflect on the importance of knowledge
-handling in robotic architectures that deal with complex semantics and put into
-perspective our choice of the RDF formalism.  Finally, we briefly re-discuss two
-facets of HRI: the importance of \emph{modelling the humans}, and the difficulty
-of \emph{acting in presence of humans}.
+This last section tries to clarify the ``takeaway'' message: what are the
+challenges that human-robot interaction brings to artificial intelligence?  We
+first restate the challenge of \emph{embodied cognition} in general and
+summarise what we see as the main decisional issues that need to be tackled to
+push forward human-robot interaction. Finally, we reflect on the importance of
+knowledge handling in robotic architectures that deal with complex semantics and
+put into perspective our choice of the RDF formalism.
 
@@
 mechanisms, and coined the concept of \emph{enactivism} as the theoretical
-framework that study the links between cognition, embodiment and actions.
+framework that study the links between cognition, embodiment and actions.  This
+has since been thoroughly studied in robotics and artificial intelligence
+(Pfeifer and Bongard~\cite{pfeifer2007body} is one of the reference).
 
@@
 architecture. The main source of grounded symbolic knowledge is the situation
-assessment module, {\sc Spark}.  Not only it builds symbolic facts from spatial
-reasoning, but it also relies on (limited) temporal reasoning to track the world
-state and build explanations to interpret unexpected perceptions (like a
-suddenly disappearing object).
-
-Because {\sc Spark} is also able to track humans, which enables in turn
-perspective-aware spatial reasoning, it produces grounded symbolic knowledge
-for interacting agents as well. This is a typical \emph{embodied} cognitive
-skill.
+assessment module, {\sc Spark}. It builds symbolic facts from spatial reasoning,
+and also relies on (limited) temporal reasoning to track the world state and
+build explanations to interpret unexpected perceptions (like an object that
+suddendly disappears). Because {\sc Spark} also tracks humans, which enables
+perspective-aware spatial reasoning, it can produce as well grounded symbolic
+knowledge for the agents it interacts with. This is a typical \emph{embodied}
+cognitive skill.
 
 Grounding also occurs during verbal and non-verbal communication. The {\sc
-Dialogs} module grounds new concepts introduced by the speaker by asking questions
-until it can attach the new concept to concepts already present in the
-knowledge pool (typically, if I ask the robot ``bring me a platypus'', the
-robot asks ``what is a platypus''. I may answer ``a mammal'' and the robot will
-ask ``what is a mammal?'' and so on until it anchors the new concepts to ones
-it already knows). Non verbal interaction (like gestures) is also partially
-grounded: we have presented how pointing, for instance, is used by the robot to
-ground \emph{``this''} or \emph{``that''} to the pointed artifact.
-
-Because only objects marked with tags are currently recognised (in our typical
-lab experiments, it may be a dozen of them), the robot grounding mechanisms work
-in a small-sized closed world. This assumption obviously greatly facilitates
-the task. It remains to be seen if systematic marking of manipulated objects
-(for instance with RFID tags) is a realistic path.\footnote{It may be the case if the
-robot is expected to be used in a well-defined, slowly changing environment.
-Real-time recognition of large libraries of objects is another path which may
-be viable soon~\cite{Dean2013Fast}.}
-
-We see however that we have only scratched the surface of a generic grounding
-capability. Context, cultural background, ``naive physics'' knowledge, emotional
-state of the human are some of the numerous other determinants that need to be
-accounted for when grounding human-robot interaction.
+Dialogs} module grounds new concepts introduced by the speaker by asking
+questions until it can attach the new concept to concepts already present in the
+knowledge pool (typically, if I ask the robot ``bring me a platypus'', the robot
+asks ``what is a platypus''. I may answer ``a mammal'' and the robot will ask
+``what is a mammal?'' and so on until it anchors the new concepts to ones it
+already knows). Embodied interactions (like gestures) are also taken into
+account at this level: we have presented how pointing, for instance, is used by
+the robot to ground \emph{``this''} or \emph{``that''} to the pointed artifact.
+
+Note that because only objects marked with 2D fiducial markers are currently
+recognised (in our typical lab experiments, it may be a dozen of them), the
+robot's grounding mechanisms only work in a small-sized closed world. This
+assumption greatly facilitates the task, and we see that we have only scratched
+the surface of a generic grounding capability. Context, cultural background,
+``naive physics'' knowledge, emotional state of the human are some of the
+numerous other determinants that need to be accounted for when grounding
+human-robot interaction.
 
@@
 
-How can we restate the main challenges, in particular in term of decision
-making, that human-robot interaction brings to artificial intelligence?
-
-\fxfatal{TDB}
+With this perspective on \emph{embodied cognition} in mind, it seems appropriate
+to restate the key challenges brought by human-robot interaction to artificial
+intelligence.
+
+\subsubsection{The W-questions}
+
+Our context, where a robot has to achieve a task together with
+humans, raises issues to be tackled by the robots' decisional components, and
+that we can summarises as ``W-questions'': \emph{What}, \emph{Who}, \emph{When},
+\emph{Where} and \emph{How}?
+
+\emph{What to do next}, at different levels of abstractions and while taking
+into account not only the current state but also the long term goals, is the
+basic question for an intelligent robot. Its is made more complex here by having
+to deal with the partially observable physical and mental state of the human
+partner, and by the extended set of possible actions.
+
+\emph{Who should act now} is of key importance and also needs to be
+decided upon by the robot. It is sometimes expected as an intelligent behaviour
+for the robot to wait and let its human partner act instead. Correct management
+of \emph{turn-taking} leads to various decisional challenges.
+
+\emph{When to perform a given action} is enriched here by the human-context
+since it has to take into account the human, his/her needs, his/her rhythms and
+pace, and his/her mental state. While performing its share of the task, the
+robot has to produce signals directed towards the human and to respond to
+signals produced back at the proper pace.
+
+\emph{Where to perform an action} plays an important role as well: the choice is
+not trivial and might need elaborated decision. The robot is expected to take
+into account effort sharing, visibility of its action by the human, disturbance
+or discomfort induced by its action.
+
+\emph{How to perform an action}, finally, needs to be reflected upon by the
+robot: several options to perform an action or to achieve a goal are often
+available, and selecting one is a non-trivial decision problem. Cost-based
+planners are one possible approach: they search for plans that satisfy an
+acceptable cost in terms of \emph{acceptability} or \emph{legibility}, for
+instance.
+
+These five questions should not be considered independently from each others and
+often require, on the contrary, to be dealt with in a single decision step. The
+human-aware task and motion planners which we have built, are instances of
+systems which have been designed to deal with such intricate decision issues.
+
+\subsubsection{Putting the Humans into Equations}
+
+The correlate of these five \emph{W-questions} is the issue the \emph{human
+models}: taking appropriate decisions with and in presence of humans require
+appropriate models of the human. Models that tell the robot what the human
+\emph{can do}, \emph{would like to do}, \emph{knows}, \emph{could infer}, etc.
+
+While the task of describing all the (dynamic!) human models that are useful to
+robots is immense (if doable at all), we claim that it is possible to devise and
+use such models in limited, but still interesting and useful, contexts such as
+collaborative human-robot objects manipulation, fetch-and-carry and associated
+activities in a domestic or professional environment.
+
+In our implementation, perspective taking, for instance, is tightly connected to
+the symbolic knowledge models, and since our knowledge base allows for storage
+of one knowledge model per agent, we have been able to endow the robot with a
+simple theory of mind (as explained in section~\ref{sect|tom}): we explicitly
+model what the robot knows about its partners in a symbolic way. This knowledge
+is then re-used in different places, to correctly interpret what the human says,
+or to plan tasks that are actually doable for the human.
+
+The cognitive model that the robot builds for the agents it interacts with
+remains today simple and mostly focused on geometric features and affordances
+(\emph{who sees what? what are our relative positions? what is reachable to
+whom?}  etc.). Extending this knowledge with more subtle perceptions (emotional
+state for instance) remains to be done.
+
+Motion and action execution also requires human models, and the one we use
+embeds human preferences and physical constraints that need to be accounted for
+when synthesizing robot motion or elaborating robot plans. This includes
+proxemics (human-robot distance) and associated issues (visibility) but also
+legibility and acceptability criterias expressed in terms of \emph{social rules}
+that the elaborated plans should satisfy.
+
+%
+%\paragraph{Joint actions with humans}\label{sec:soa}
+%\fxfatal{too many citation -> synthesis needed!}
+%
+%The human presence brings new requirements for robot's abilities both
+%at the functional and at the deliberative levels~\cite{Klein2004}. The
+%topics involve motion~\cite{Kulic2007,Berg2004,Madhava2006},
+%navigation~\cite{Althaus2004,Sisbot2007}, manipulation~\cite{Kemp2007}
+%in presence of humans as well as perception of human
+%activities~\cite{Breazeal2001,Burger2008}. Also, when
+%interacting with humans, robots need to incorporate communication and
+%collaboration abilities. Several theories dealing with
+%collaboration~\cite{Cohen1991,Grosz1996,Clark1996} emphasise that
+%collaborative tasks have specific requirements compared to individual
+%ones, \eg, since the robot and the person share a common goal, they
+%have to agree on the manner to realise it, they must show their
+%commitment to the goal during execution, etc. Several robotic systems
+%have already been built based on these
+%theories~\cite{Rich1997,Sidner2005,Breazeal2003} and they
+%all have shown benefits of this approach. They have also shown how
+%difficult it is to manage turn-taking between communication partners
+%and to interleave task realization and communication in a generic
+%way. Finally, today only few
+%systems~\cite{Fong2006,Breazeal2003,Sisbot2008} take humans into
+%account at all levels.
 
@@
 \label{krs-discussion}
+
+As thorougly presented in this article, we have built the decisional
+capabilities of our robots around this idea of explicit knowledge manipulation.
+We finally briefely comment on this design choice, our implementation based on RDF,
+and the limitations that we perceive.
 
@@
 
-\subsection{Bringing the Humans into the Equation}
-
-\subsubsection{Modelling the Humans}
-
-\fxfatal{Re-organise, rephrase, complement}
-
-In our implementation, perspective taking is tightly connected to the symbolic
-knowledge models, and since our knowledge base allows for storage of one
-knowledge model per agent, we have been able to endow the robot with a simple
-theory of mind (as explained in section~\ref{sect|tom}): we explicitly model
-what the robot knows about its partners in a symbolic way. This knowledge is
-then re-used in different places, to correctly interpret what the human says,
-or to plan tasks that are actually doable for the human.
-
-The cognitive model that the robot builds for the agent it interacts with is
-still simple and mostly focused on geometric features (\emph{Who sees what? What are
-our relative positions?} etc.). Extending this knowledge with more subtle
-perceptions (emotional state for instance) remains to be done.
-
-We should also mention that modelling of human behaviours also appear in our
-architecture at the level of task and motion planning with a set of
-\emph{social rules}, including proxemics, action legibility or time
-optimization.
-
-\subsubsection{Acting in Presence of Humans}
-
-HRI offers new challenges to planning by adding a supplementary constraint: the
-synthesis of plans that are not only feasible but legible (in terms of their
-intention) and compliant with human preferences. This constraint is often
-ill-defined, dynamic and context-dependent.  HATP, by co-planning for the robot and
-the human, and modelling social rules as planning parameters, is a first step in
-this direction.
-
-\fxfatal{What are the next steps?}
-
-\paragraph{Joint actions with humans}\label{sec:soa}
-\fxfatal{too many citation -> synthesis needed!}
-The human presence brings new requirements for robot's abilities both
-at the functional and at the deliberative levels~\cite{Klein2004}. The
-topics involve motion~\cite{Kulic2007,Berg2004,Madhava2006},
-navigation~\cite{Althaus2004,Sisbot2007}, manipulation~\cite{Kemp2007}
-in presence of humans as well as perception of human
-activities~\cite{Breazeal2001,Burger2008}. Also, when
-interacting with humans, robots need to incorporate communication and
-collaboration abilities. Several theories dealing with
-collaboration~\cite{Cohen1991,Grosz1996,Clark1996} emphasise that
-collaborative tasks have specific requirements compared to individual
-ones, \eg, since the robot and the person share a common goal, they
-have to agree on the manner to realise it, they must show their
-commitment to the goal during execution, etc. Several robotic systems
-have already been built based on these
-theories~\cite{Rich1997,Sidner2005,Breazeal2003} and they
-all have shown benefits of this approach. They have also shown how
-difficult it is to manage turn-taking between communication partners
-and to interleave task realization and communication in a generic
-way. Finally, today only few
-systems~\cite{Fong2006,Breazeal2003,Sisbot2008} take humans into
-account at all levels.
-
 \subsection{The Next Steps}

### 2014-05-19 paper.tex
Rephrase, integrate HATP section, based on Rachid's contribution

@@
 desires into a succession of atomic actions. We use in our architecture the HATP
-planner (for Human Aware Task Planner~\cite{Alili2008,
-Alili2009,Lallement2014}).
-
-HATP (Hierarchical Agent- based Task Planner) planning framework extends the
+planner (\emph{Human Aware Task Planner})~\cite{Alili2008,
+Alili2009,Lallement2014}.
+
+The HATP planning framework extends the
 traditional HTN (Hierarchical Task Network) planning domain representation and
@@
 streams, one for the robot and one for the human, so that the streams may be executed in
-parallel and synchronised when necessary.
+parallel and synchronised when necessary (Figure~\ref{plan_hatp1}).
 
@@
 
-The planner also generates synchronisation points between the agents.
-Figure~\ref{plan_hatp1} illustrates such a plan with two streams.  Here, the
-robot needs to wait for the success of the human agent's {\tt PUT} action
-(estimated through the fulfilment of the action post-conditions, typically
-$\neg$\stmt{GREY\_TAPE isOn TABLE}) to be able to start its own action.
+The planner also generates synchronisation points between the agents.  In the
+plan depicted on Figure~\ref{plan_hatp1}, the robot needs to wait for the
+success of the human agent's {\tt PUT} action (estimated through the fulfilment
+of the action post-conditions, typically $\neg$\stmt{GREY\_TAPE isOn TABLE}) to
+be able to start its own action.
 
@@
 \caption{A plan produced by HATP with two action streams (human actions at the
-top, robot actions at the bottom). The arrow between the two streams represents
-a synchronization point ({\sf PUTRV} stands for {\it Put it so it is both
-reachable and visible}).}
+top, robot actions at the bottom, {\sf PUTRV} stands here for {\it Put it so it is both
+reachable and visible}). The arrow between the two streams represents
+a synchronization point.}
 
@@
 
-To address some of those challenges HATP includes mechanisms called
-"social tules" to filter plans so as to keep only those suitable for
-HRI. To this end, HATP allows the specification of the following
-filtering criteria:
-
-\emph{Wasted time}: Avoids plans where the human mentioned in a plan spends a lot
-of its time waiting.
-
-\emph{Effort balancing:} Avoids plans where efforts are
-not fairly distributed among the agents mentioned in a plan.  It is
+HATP includes mechanisms called \emph{social rules} to filter plans so as to
+keep only those considered as suitable for human-robot interaction. The planner
+allows the specification of the following filtering criteria:
+
+\emph{Wasted time}: avoids plans where the human spends a lot
+of its time being idle,
+
+\emph{Effort balancing}: avoids plans where efforts are
+not fairly distributed among the agents taking part to the plan.  It is
 indeed sometimes beneficial to balance efforts between the human and
-the robot.
-
-\emph{Control of intricacy:} Avoids plans with too many interdependencies
+the robot,
+
+\emph{Control of intricacy}: avoids plans with too many interdependencies
 between the actions of agents mentioned in the plan, as a problem with
@@
 the human will find himself repeatedly in a situation where his is
-waiting for the robot to act
-
-\emph{Undesirable sequences:} Avoids plans that violate specific user-defined
+waiting for the robot to act,
+
+\emph{Undesirable sequences}: avoids plans that violate specific user-defined
 sequences (for instance sequences which can be misinterpreted by the human).
 
-Combining some of the above criteria could help yield the
-following interesting \emph{shared plans}: (1) the human ends up doing a
-number of the tasks, but yet the overall effort is significantly lower than what the robot puts
-to do a lower number of effort-intensive tasks; (2) avoiding, when
-possible, having the human wait for the robot several times, which
-essentially prevents the streams from having too many causal links
-between them.
-
- In the current implementation, the so-called social rules are
+Combining the above criteria, we typically yield shared plans with desirable
+interaction features like the human still being engaged in a number of tasks
+while its overall level of required efforts remains low, or avoiding having the
+human to wait for the robot (by essentially preventing the action streams from
+having too many causal links between them).
+
+Figure~\ref{plan_hatp2} illustrates such a socially-optimised plan where the
+\emph{no wasted time} social rule is applied. The resulting shared plan is
+considered better than the one depicted on Figure~\ref{plan_hatp1} according to
+a global evaluation of the costs and rules.
+
+In the current implementation, the social rules are effectively
 implemented by looking through all the plans produced and filtering
-out the ones that do not meet the requirements specified. In the
+out the ones that do not meet the desired requirements. In the
 future we intend to study algorithms that do such filtering online,
-rather than after primitive solutions are found.
+rather than after initial solutions are found.
 
@@
 \end{figure}
-
-Figure~\ref{plan_hatp2} illustrates an alternative plan to the previous 
-one (Figure~\ref{plan_hatp1}) where the \emph{no wasted time} social rule is
-applied. The resulting shared plan is a better plan according to a global evaluation of
-the costs and rules.
 

### 2014-05-19 paper.tex
Rephrase, integrate SHARY section, based on Rachid's contribution

@@
 provide a large set of high-level actions and an event-based architecture well
-suited for prototyping.  They both rely on deep integration with the knowledge
-base that serves as the primary source of semantics for decision making.
+suited for prototyping.  They both rely on extensive integration with the knowledge
+base, that serves as the primary source of semantics for decision making process.
 
@@
 of shared plans~\cite{Grosz1996,Clark1996,Kemp2007}. This means essentially
-contex-based refinement and robot actions execution as well as monitoring of
+context-based refinement and robot actions execution, as well as monitoring of
 those achieved by its human partner~\cite{Breazeal2003}. The challenge is to
 build such abilities in a generic way, and to provide several levels of
-parametrization allowing to adapt to various environments, and various levels of
+parametrisation allowing to adapt to various environments, and various levels of
 involvement of the robot ranging from teammate behaviour to assistant or
 proactive helper.  Depending on this, the robot controller invokes the various
-human-aware planners and will react to events detected at ORO level.
-
-SHARY’S originality, as a supervision system, lies in its ability to take into
-account not only the task achievement but also communication and monitoring
-needed to support interactive task achievement ~\cite{Rich1997,Sidner2005} in a
-flexible way. SHARY allows to link action execution to more or less elaborated
-“communication policies” in order to produce signals towards the human and to
-react to human action.
-
-A communication act is the central notion in the formalism. It plays the role of
-transition condition. Consequently, a communication act represents an
-information exchange between the two agents. This exchange can be realized
-through dialog or by an expressive motion or a combination of the two. It
-enables each agent to communicate his belief about the task to be realized in
-order to share mutual knowledge and to synchronyze their activities.  This is
-done through real-time task-based situation assessment achieved by the
-combination of SPARK monitoring processes and ORO inference mechanisms
-\cite{fiore2014}.
+human-aware planners and will react to events detected at {\sc Oro} level (as
+described below).
+
+{\sc Shary}'s originality, as an execution control system, lies in its ability
+to take into account not only the task achievement but also communication and
+monitoring needed to support interactive task
+achievement~\cite{Rich1997,Sidner2005} in a flexible way. {\sc Shary} allows to
+link action execution \emph{communication policies} in order to produce signals
+towards the human and to react to human actions. The \emph{communication act} is
+the central concept in this formalism. It plays the role of transition
+condition, and, as such, represents an information exchange between the two
+agents. This exchange can be realised through dialogue, by an expressive motion
+or a combination of the two. It enables each agent to communicate their beliefs
+about the task to be achieved, in order to share mutual knowledge and to
+synchronise their activities.  This is done through real-time task-based
+situation assessment achieved by the combination of {\sc Shary} monitoring
+processes and {\sc Oro} inference mechanisms~\cite{fiore2014}.
 

### 2014-05-19 paper.tex
Finished rebuttal letter + last fixes in the paper

@@
 
-\usepackage[draft, nomargin, marginclue, footnote]{fixme}
+\usepackage[nomargin, marginclue, footnote]{fixme}
 
@@
 
-We do not claim to address here the whole issue. This article attempts however to organise
-it into a coherent challenge for artificial intelligence, and to explain and
-illustrate some of the paths that we have investigated on our robots and that
-result in a deliberative, knowledge-oriented, architecture for human-robot
-interaction.
+We do not claim to address here the issue as a whole. This article attempts
+however to organise it into a coherent challenge for artificial intelligence,
+and to explain and illustrate some of the paths that we have investigated on our
+robots, that result in a set of deliberative, knowledge-oriented, software
+components designed for human-robot interaction.
 
@@
 
-Numerous approaches exist, like amodal \emph{proxies}~\cite{Jacobsson2008},
-grounded amodal representations~\cite{Mavridis2006}, semantic
+Numerous approaches exist, like amodal (in the sense of modality-independent)
+\emph{proxies}~\cite{Jacobsson2008}, grounded amodal
+representations~\cite{Mavridis2006}, semantic
 maps~\cite{Nuechter2008, Galindo2008,Blodow2011} or affordance-based planning
@@
 
-\fxfatal{Rephrase or remove}
 It is sometimes difficult or even impossible to see and/or track an object in
@@
 symbolic states of objects (whether the object is on a furniture, in an agent
-hand, in a container, etc.): according to the robot perception of what has
+hand, in a container, etc.), and according to the robot perception of what has
 happened since the object was last seen, the robot builds a belief of the
@@
 and geometric reasoning on human hand trajectories and potential objects
-placements can provide clues for high level human monitoring
-processes. We call this temporal and geometric reasoning \emph{primitive action
-recognition}. Those primitive actions are assessed through monitoring situations
-like ``an empty hand is close to an object on a table'' (precursor for a
-\emph{pick}), or ``a hand holding an object is over a container'' (precursor for
-a \emph{throw}). {\sc Spark} recognises a handful \fxfatal{which ones?} of these
-primitives, that are relied on by the robot to track the state transitions in
-the human plans.
+placements can provide clues for high level human monitoring processes. We call
+this temporal and geometric reasoning \emph{primitive action recognition}. Those
+primitive actions are assessed through monitoring situations like ``an empty
+hand is close to an object on a table'' (precursor for a \emph{pick}), or ``a
+hand holding an object is over a container'' (precursor for a \emph{throw}).
+{\sc Spark} recognises a core set of such primitives that are relied upon by the
+robot to track the state transitions in the human plans.
 
@@
 human-robot dialogue (the planner vocabulary contains concepts like
-\texttt{give}, \texttt{table}, \texttt{is on}...). Hence only a few ontology
-rules have been required to map both the knowledge extracted from the situation
-assessment and the statements originated from the verbal interaction to the
-planner domain.
-
-Also, we would like to mention that, after some research (see appendix B
-of~\cite{Lemaignan2012a} for a detailed discussion)  we have decided to
-represent neither the planning domain nor the resulting plans in the knowledge
-base: the planning domain (with task pre- and postconditions) is stored in a
-specific format, outside of the central declarative knowledge repository, and
-the plans are directly communicated to the robot controller. This leads to
-independent declarative and procedural knowledge stores.
+\texttt{give}, \texttt{table}, \texttt{is on}...). This leads to a natural
+mapping (low ``impedance mismatch'') between the knowledge extracted from the situation
+assessment or the dialogue, and the planner.
+
+Also, we would like to mention that, after some research we have decided against
+representing the planning domain (\ie, the set of tasks with their pre- and
+post-conditions) in the knowledge base itself due to expressiveness issues (see
+appendix B of~\cite{Lemaignan2012a} for a detailed discussion). This effectively
+leads to independent declarative and procedural knowledge stores.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2016-06-10 paper.tex
New abstract

@@
 
-Human-Robot interaction is an area full of challenges for artificial
-intelligence: dynamic, partially unknown environments that are not originally
-designed for autonomous machines; a large variety of situations and objects to
-deal with, with possibly complex semantics; physical interactions with humans
-that requires fine, low-latency control, representation and management of
-several mental models, pertinent situation assessment skills...the list goes on.
-
-This article sheds light on some key decisional issues that are to be tackled
-for a cognitive robot to share space and tasks with a human, and present our
-take on these challenges. We adopt a constructive approach based on the
-identification and the effective implementation of individual and collaborative
-skills. These cognitive abilities cover geometric reasoning and situation
-assessment mainly based on perspective-taking and affordances, management and
-exploitation of each agent (human and robot) knowledge in separate cognitive
-models, natural multi-modal communication, ``human-aware'' task planning, and human
-and robot interleaved plan achievement.
-
-We present our design choices, the articulations between the diverse
-deliberative components of the robot, experimental results, and eventually
-discuss the strengths and weaknesses of our approach. It appears that
-\emph{explicit} knowledge management, both symbolic and geometric, proves to be
-key as it pushes for a different, more \emph{semantic} way to address the
-decision-making issue in human-robot interactions.
+Human-Robot Interaction challenges Artificial Intelligence in many regards:
+dynamic, partially unknown environments that were not originally designed for
+robots; a broad variety of situations with rich semantics to understand and
+handle; physical interactions with humans that requires fine, low-latency yet
+socially acceptable control strategies; natural and multi-modal communication
+which mandates common-sense knowledge and the representation of possibly
+divergent mental models. This article attempts to characterize these challenges
+and evidences the key decisional issues that need to be addressed for a
+cognitive robot to successfully share space and tasks with a human.
+
+We identify in particular the needed individual and collaborative cognitive
+skills: geometric reasoning and situation assessment based on perspective-taking
+and affordance analysis; acquisition and representation of knowledge models for
+multiple interactants (humans and robots); natural multi-modal dialogue;
+human-aware task planning; and human-robot joint plan achievement.  The article
+discusses each of these abilities (including reference implementations) and show
+how they combine in a coherent and original cognitive architecture for
+human-robot interaction. Supported by experimental results, we eventually show
+how explicit knowledge management, both symbolic and geometric, proves to be
+instrumental to richer and more natural human-robot interactions by pushing for
+pervasive, human-level semantics within the robot's deliberative system.
 

### 2016-06-13 paper.tex
Polish abstract

@@
 
-We identify in particular the needed individual and collaborative cognitive
+We identify the needed individual and collaborative cognitive
 skills: geometric reasoning and situation assessment based on perspective-taking
@@
 multiple interactants (humans and robots); natural multi-modal dialogue;
-human-aware task planning; and human-robot joint plan achievement.  The article
-discusses each of these abilities (including reference implementations) and show
+human-aware task planning; human-robot joint plan achievement.  The article
+discusses each of these abilities, presents reference implementations, and show
 how they combine in a coherent and original cognitive architecture for

### 2016-06-13 paper.tex
New tentative title
Update affiliations

@@
 \title{\LARGE \bf
-Human-Robot Interaction: Tackling the AI Challenges
+Artifical Cognition for Social Human-Robot Interaction:\\ An Implementation
 }
 
-\author{Séverin Lemaignan, Mathieu Warnier, E. Akin Sisbot, Rachid Alami}
+\author{Séverin Lemaignan$^{1,2}$, Mathieu Warnier$^1$, E. Akin Sisbot$^1$,
+Rachid Alami$^1$}
 
 \address{
-CNRS, LAAS, 7 avenue du Colonel Roche, F-31400 Toulouse, France\\
-Univ de Toulouse, LAAS, F-31400 Toulouse, France\\
-{\tt firstname.lastname@laas.fr}
+$^1$CNRS, LAAS, 7 avenue du Colonel Roche, F-31400 Toulouse, France\\
+Univ de Toulouse, LAAS, Toulouse, France\\
+{\tt firstname.surname@laas.fr}
+}
+
+\address{
+$^2$Centre for Robotics and Neural Systems\\
+Plymouth University, Plymouth, United Kingdom\\
+{\tt firstname.surname@plymouth.ac.uk}
 }
@@
 
-This work has been supported by EU FP7 ``SAPHARI'' under grant agreement no.
-ICT-287513.
+This work has been partially supported by EU FP7 ``SAPHARI'' (grant ICT-287513),
+and the EU H2020 Marie Sklodowska-Curie Actions project DoRoThy (grant 657227).
 

### 2016-06-13 paper.tex
Fixing language 1

@@
 \newcommand{\concept}[1]{{\small \texttt{#1}}}
-\newcommand{\stmt}[1]{{\footnotesize \tt $\langle$ #1\relax$\rangle$}}
-
-\newcommand{\ie}{{\textit{i.e.\ }}}
+\newcommand{\stmt}[1]{{\footnotesize\tt$\langle$#1\relax$\rangle$}}
+
+\usepackage{xspace}
+\newcommand{\ie}{{i.e.\xspace}}
 \newcommand{\cf}{{\textit{cf\ }}}
-\newcommand{\eg}{{\textit{e.g.\ }}}
-\newcommand{\al}{{\textit{et al.\ }}}
+\newcommand{\eg}{{e.g.\xspace}}
+\newcommand{\al}{{et al.\xspace}}
 
@@
 illustrates this context: the two agents share a common space and exchange
-information through multiple modalities, and the robot is expected to achieve
+information through multiple modalities (we specifically consider verbal
+communication, deictic gestures and social gaze), and the robot is expected to achieve
 interactive object manipulation, fetch and carry tasks and other similar
-domestic tasks by taking into account, at every stage, the intentions, beliefs,
+chores by taking into account, at every stage, the intentions, beliefs,
 perspectives, skills of the human partner.  Namely, the robot must be able to
@@
 \caption{The robot reasons and acts in domestic interaction scenarios.  The
-    sources of information are multi-modal dialogue {\it (A)}, and
+    sources of information are multi-modal dialogue {\it (A)} and
     perspective-aware monitoring of the environment and human activity {\it
@@
 
-These three challenges, \emph{Communication}, \emph{Joint Actions},
-\emph{Human-Aware Execution}, structure the whole field of human-robot
-interaction research, and they can be analysed in terms of the cognitive skills
-that they require.
-
-A \emph{joint action}, for instance, builds from:
+These three challenges, \emph{communication}, \emph{joint actions},
+\emph{human-aware execution}, structure the research in human-robot
+interaction. They can be understood in terms of cognitive skills that they
+mandate. \emph{Joint action}, for instance, builds from:
 
@@
 The robot controller (with the help of a task planner) decides what action to
-execute next, and who should perform it from the human or the robot (or both in
+execute next, and who should perform it from the robot or the human (or both in
 case of a joint action), and finally controls or monitors its execution. The
@@
 need to be available to the robot. It must be able \item to represent and
-manipulate symbolic belief states, \item to acquire and maintain them up-to-date
-with respect to state of the world and the task at hand, \item to build and
+manipulate symbolic belief states, \item to acquire and keep them up-to-date
+with respect to the state of the world and the task at hand, \item to build and
 iteratively refine shared (human-robot) plans, \item to instantiate and execute
@@
 
-Such abilities must be designed and implemented in a generic way and
-must provide several levels of parametrization such as they adapt to
+Besides, such abilities should be designed and implemented in a yask-independent
+manner, and should provide sufficient levels of parametrization such as they adapt to
 various environments, different tasks and variable levels of engagement of the
@@
 
-\subsection{Article organization}
-
-The remaining of the article discusses the robotic architecture we have built to
+\subsection{Article overview}
+
+The remaining of the article discusses the robotic architecture that we have built to
 tackle the AI challenges of human-robot interaction, and how its relates to
@@
 knowledge model that we have developed for our robots. Then,
-section~\ref{sec:impl} gives some insights on each of the main components of the
-architecture, and tries to highlight their significance for artificial
+Section~\ref{sec:impl} details each of the main components of the
+architecture, and highlights their significance for artificial
 intelligence. Section~\ref{sec:expe} presents two studies that illustrate in a
@@
   %%% DIALOGS
-  \node at (-6, -2) [skill, fill=BrickRed!50] (dialogs) {{\sc Dialogs} \\ Dialogue processing};
+  \node at (-6, -2) [skill, fill=BrickRed!50] (dialogs) {{\sc Dialogs} --
+      Natural\\ language processing};
 
@@
       \begin{tikzpicture}
-          \node at (0,0) (geom) {{\sc Spark} -- Geometric \& Temporal Reasoning};
+          \node at (0,0) (geom) {{\sc Spark} -- Geometric \& temporal situation assessment};
         \node [subpart, below=0.2 of geom.south west, anchor=north west] (world-update) {Sensors fusion};
@@
   %%% MHP
-    \node at (9,0)[skill, fill=BurntOrange!50] (mhp) {{\sc mhp} -- Human-aware \\ Motion and Manipulation \\ Planning};
+    \node at (9,0)[skill, fill=BurntOrange!50] (mhp) {{\sc mhp} -- Human-aware \\ motion and manipulation \\ planning};
 
@@
       \begin{tikzpicture}
-        \node at (0,0) (exec) {Execution Controller};
+          \node at (0,0) (exec) {{\sc Shary | pyRobots} -- Execution controllers};
         \node [subpart, below=0.2 of exec.south west, anchor=north west] (plans) {Goal \& Plans \\ management};
@@
   %%% Relations between components
-  \path (shary.340) edge [<->, bend left] node[label] {motion plan \\ requests} (mhp);
+  \path (shary.340) edge [->, bend left] node[label] {motion plan \\ requests} (mhp);
   \path (shary.west) edge [<->, bend right] node[label] {shared \\ plans} (hatp);
-  \path (hatp) edge [<->, bend right] node[label] {world model and \\ agents beliefs} (oro.170);
+  \path (hatp) edge [<-, bend right] node[label] {world model and \\ agents beliefs} (oro.170);
   \path (dialogs) edge [<->, bend left] node[label] {natural language \\ grounding} (oro.190);
@@
     with a thick border. The links between components depicted on the figure
-    underline the central role of the knowledge base: most of the datastreams are
+    underline the central role of the knowledge base: many of the datastreams are
     actually symbolic statements exchanged through this semantic blackboard.}
@@
 
-Connecting multiple independent software modules in one coherent robotic
+Bridging multiple independent software modules in one coherent robotic
 architecture is not only a technical challenge, but also a design and
-architectural challenge.
-
-In particular, dealing with the rich semantics of natural interaction with
-humans has been one major focus. To this end, we have been deploying
-over the last years explicit knowledge representation and manipulation in the
-deliberative layer of the robots as a {\it lingua franca} between our
-components.
+architectural challenge. In particular, properly managing the rich semantics of
+human-level natural interaction raises a range of issues. Our basic assumption
+and guiding principle is that human-level interaction is easier to achieve if
+the robot itself relies internally on human-level semantics.  We implement this
+principle by extensively relying on explicit knowledge representation and
+manipulation: software components communicate with each other using first-order
+logic statements organized into ontologies and whose semantics are close to the
+ones manipulated by humans.
 
 Figure~\ref{fig|archi} gives an overview of our architecture. An active
-knowledge base ({\sc Oro}), conveniently thought as a \emph{semantic
-blackboard}, connects most of the modules together: the \emph{geometric
-reasoning} module ({\sc Spark}) produces symbolic assertions (like
+knowledge base ({\sc Oro}), conveniently thought as a semantic
+blackboard, connects most of the modules together: the geometric
+reasoning module ({\sc Spark}) produces symbolic assertions (like
 \stmt{BOOK1 isOn TABLE}) describing the state and dynamics of the robot's
@@
 Dialogs} module queries the knowledge base: {\tt \footnotesize find(?book type
-Book, ?book differentFrom BOOK1)}\footnote{We present study with a complete
+Book, ?book differentFrom BOOK1)}\footnote{We present a complete
 example at section~\ref{moving-london}.}, and write back assertions like
@@
 Oro}. The HATP planner then uses the knowledge base to initialise the
-planning domains with similar requests ({\tt \footnotesize find(BOOK2 isAt
-?location)}, etc.), and the execution controller typically monitor
+planning domains with similar requests (like {\tt \footnotesize find(BOOK2 isAt
+?location)}), and the execution controller typically monitor
 conditions (by subscribing to events like: {\tt \footnotesize
@@
 
-To some extent, this architecture moves away from classical layered approaches
-found in robotics~\cite{Gat1998three, Volpe2001CLARAty, Goldberg2002}.
-Interactions between components at the deliberative level are mostly
-bidirectional and we do not introduce layers of abstraction amongst software
-components\footnote{We do have lower-level modules to execute actions or manage
-sensors, but all cognition-related modules live in the same global deliberative
-layer.}. Dialogue processing, for instance, illustrates this structure. This
-component does not purely act as an alternative sensing modality that would be
-fed to a distinct deliberative component. On the contrary, it actively queries
-the knowledge base to interpret and disambiguate natural language, and in that
-regards, it is fully integrated to the deliberative process itself (we explain
-natural language processing at section~\ref{sect|com}).
+This architecture moves away from classical layered approaches found in
+robotics~\cite{Gat1998three, Volpe2001CLARAty, Goldberg2002}. Contrary to
+layered architectures, interactions between components at the deliberative level
+are here essentially bidirectional and we do not introduce layers of abstraction
+amongst these deliberative components\footnote{We do have lower-level modules to
+execute actions or manage sensors, but all cognition-related modules live in the
+same global deliberative space.}. The natural language processing component
+illustrates this structure: instead of being an independent input modality whose
+outputs would be unidirectionally fed to ``higher'' decisional components, it
+lives in the deliberative space at the same level as other deliberative
+components, and uses the knowledge base in a bidirectional manner, to interpret,
+disambiguate natural language, and eventually store newly produced
+interpretations (natural language processing is discussed in
+Section~\ref{sect|com}).
 
@@
 module to acquire \emph{desires} from the other agents, and the planner and the
-execution controller that to first decide to take or not into account an
+execution controller to first decide to take into account (or not) an
 incoming desire as a \emph{goal}, and then to generate and manage
@@
 
-Building upon a BDI architecture, other activities are conducted in parallel,
-without being explicitly triggered by \emph{desires} in the BDI sense, like
-situation assessment and monitoring or other forms of dialogue (including
-performative dialogue that can possibly change the internal state of the robot,
-but does not lead directly to the creation of \emph{desires}, like assertion of
-new facts or question answering).
+We extend upon BDI architectures by running other background deliberative task,
+without them being explicitly triggered by \emph{desires} in the BDI sense.
+The main ones include situation assessment, action monitoring and processing of
+non-imperative speech (including performative dialogue that can possibly change
+the internal state of the robot, but does not lead directly to the creation of
+\emph{desires}, like assertion of new facts or question answering).
 
@@
 knowledge as it is produced by each of the other deliberative components (the
-clients). It exposes a {\tt json}-based RPC API~\cite{lemaignan2012kbapi} to
-query the knowledge base.  We represent knowledge as RDF triples in the OWL
+clients). It exposes a {\tt json}-based RPC API to
+query the knowledge base~\cite{lemaignan2012kbapi}.  We represent knowledge as RDF triples in the OWL
 sub-language. Each time triples are added or removed from the knowledge base, a
@@
 collaborative human-robot objects manipulation, fetch-and-carry and associated
-activities in a domestic or professional environment.
+activities in home or work environment.
 

### 2016-06-13 paper.tex
Fixing language 2

@@
 knowledge as it is produced by each of the other deliberative components (the
-clients). It exposes a {\tt json}-based RPC API to
-query the knowledge base~\cite{lemaignan2012kbapi}.  We represent knowledge as RDF triples in the OWL
-sub-language. Each time triples are added or removed from the knowledge base, a
-Description Logics reasoner ({\sc Pellet}\footnote{\tt
-http://clarkparsia.com/pellet/}) classifies the whole ontology and inserts all
-possible inferred triples.  The clients of the {\sc Oro} server are in charge of
-managing themselves the knowledge (when to add, when to update, when to retract
-knowledge): no meta-semantics is carried over that would let the server manage
-these dynamics itself.\footnote{There is one exception: so-called \emph{memory
-profiles} let the server discard facts after a specific period of time. We
-present this experimental feature at section~\ref{memory}.}
+clients). It exposes a {\tt json}-based RPC API to query the knowledge
+base~\cite{lemaignan2012kbapi}.  We represent knowledge as RDF triples in the
+OWL sub-language~\footnote{\url{http://www.w3.org/TR/owl2-overview/}}. Every
+time triples are added or removed from the knowledge base, a Description Logics
+reasoner ({\sc Pellet}\footnote{\url{https://github.com/complexible/pellet}})
+classifies the whole ontology and inserts all possible inferred triples.  The
+clients of the {\sc Oro} server are in charge of managing themselves the
+knowledge (when to add, when to update, when to retract knowledge): no
+meta-semantics are carried over that would let the server manage these dynamics
+itself.\footnote{With one exception: so-called \emph{memory profiles} let
+the server automatically discard facts (\ie forget about them) after a specific
+period of time. We present this feature at section~\ref{memory}.}
 
@@
 passive component to the rest of the system -- even though it actually actively
-processes the knowledge pool in the background, for instance with the reasoner)
+processes the knowledge pool in the background to perform inferences)
 departs from other approaches like the CAST model~\cite{Hawes2007} where
@@
 
-At run-time, the knowledge available to the robot at a given time comes from
-three sources. {\it A priori} knowledge is stored in an ontology (the {\sc
-OpenRobots} ontology, discussed below) and loaded at start-up. This static source covers the
+At run-time, the knowledge available to the robot comes from three sources. {\it
+A priori} knowledge is stored in an ontology (the {\sc OpenRobots} ontology,
+discussed hereafter) and loaded at start-up. This static source covers the
 \emph{common-sense} knowledge of the robot, and optionally some
@@
 perception, interaction and planning. The next sections go into details of these
-processes.  Lastly, the third source of symbolic statements is the reasoner
-itself.
-
-Contrary to other projects like KnowRob~\cite{Tenorth2009a} that relies on
+processes. Lastly, the third source of symbolic statements comes from the
+inferences produced by the reasoner.
+
+Contrary to similar projects like KnowRob~\cite{Tenorth2009a} that relies on
 the concept of \emph{computables} to lazily evaluate/acquire symbolic facts when
@@
 
-\paragraph{The OpenRobots Ontology} The {\sc Oro} common-sense ontology has been
-designed from two requirements: covering our experimental needs and conforming
-as much as possible to the {\sc OpenCyc} upper ontology.
+\subsubsection{RDF as a Formalism for Semantics}
+
+The {\sc Oro} server relies on Description Logics (OWL) to represent and
+manipulate knowledge. Relying on RDF triples and Description Logics has advantages such as good
+understanding of its trades-off, thanks to being widespread in the semantic Web
+community, the availability of mature libraries to manipulate the ontology,
+interoperability with several major on-line knowledge bases ({\sc
+OpenCyc}, {\sc WordNet}, {\sc DBPedia} or {\sc RoboEarth}~\cite{Waibel2011} are
+some examples), open-world reasoning, and the formal guarantee of decidability
+(it is always possible to classify a Description Logics ontology).
+
+It also has restrictions, both basic (the suitability of
+Description Logics when reasoning on --typically non-monotonic-- commonsense
+knowledge has been questioned) and practical: RDF triples imply binary
+predicates, which constrains the expressiveness of the system or leads to
+inconvenient reifications. Alternatives have been proposed (like {\sc
+KnowRob}~\cite{Tenorth2009a}) that interleave RDF with more expressive logic languages
+like {\sc Prolog} with however other limitations, like closed-world reasoning.
+
+Classification performance is another issue: in our experience, an
+ontology sized for a typical study (about 100 classes and 200 instances),
+classification takes around 100ms, which leads to lags during
+interactions.  Besides, the performances are difficult to predict: the insertion
+of simple statements may indirectly change abruptly the logical
+complexity of the whole knowledge model and lead to notable degradation of
+classification time.
+
+This knowledge model also largely excludes representation of continuous
+phenomena (like time) or uncertain phenomena. When required (for instance for
+action recognition), these are managed by dedicated components (like {\sc
+Spark}), and are not exposed at the semantic level.
+
+While alternatives like \emph{Answer Set Programming} have been successfully
+investigated in robotics~\cite{Chen2010,Erdem2012}, in particular to deal with
+non-monotonic reasoning, we did not actually encounter major reasoning issues
+while working with OWL ontologies. We may reconsider this choice at a later
+stage, but until now it has proven an effective framework to quickly explore
+implementations of new cognitive abilities (for instance, it has been
+conceptually and technically easy to add support for independent knowledge
+models, one per agent the robot interacts with).
+
+Anecdotally, and because ontologies and RDF statements remain conceptually simple
+(compared to full logical languages like {\sc Prolog}), their adoption has also
+effectively helped to grow awareness amongst colleagues on the significance of
+the ``semantic level'' when developing new components for the robot.
+
+\subsubsection{The OpenRobots Ontology}
+
+As previously said, the logical statements exchanged between the deliberative
+components are organized within the {\sc Oro} knowledge base into an ontology.
+The conceptualization (\ie the system of concepts or \emph{TBox}) of this
+ontology is generally statically asserted, while the instantiation (the
+\emph{ABox}) of the ontology is generally dynamically asserted at run-time, by
+the other components.
+
+The \emph{OpenRobots} common-sense ontology represents the statically asserted
+part of the ontology. It has been designed from two requirements: beining
+practical (\ie covering our experimental needs) and conforming as much as
+possible to existing standards (in our case, the {\sc OpenCyc} upper
+ontology~\cite{lenat1990cyc}).
 
@@
 other on-line databases (like Wikipedia or WordNet), the reuse of key {\sc
-OpenCyc} concepts ensures to a certain extent that the knowledge stored by the
-robot can be shared or extended with well-defined semantics. A good example is
-the concept of \emph{Object}: in everyday conversation, an object is a
-relatively small physical thing, that can be typically manipulated. Normally, a
-human is not considered as an object. In {\sc Cyc}, an object has a more precise
-definition: it is something \emph{partially tangible}. This includes obviously
+OpenCyc} concepts ensures that the knowledge stored by the robot can be shared
+or extended with well-defined semantics. The concept of \emph{Object} represents
+a typical case of ambiguous meaning: in everyday conversation, an object is a
+relatively small physical thing, that can be typically manipulated. A human is
+not usually considered as an object. {\sc OpenCyc} however precisely defines an
+object as anything at least \emph{partially tangible}. This includes obviously
 the humans, and actually many other entities that would not be commonly said to
@@
 in the {\sc Oro} model between \emph{temporal} and \emph{spatial} entities (formally,
-$(\text{TemporalThing} \sqcap \text{SpatialThing})^{\mathcal{I}} = \emptyset$, with
+$(\text{\tt TemporalThing} \sqcap \text{\tt SpatialThing})^{\mathcal{I}} = \emptyset$, with
 $\mathcal{I}$ the \emph{interpretation} of our model).
@@
 (\ie the current robot configuration and the actions required by the scenario).
-The set of actions that the robot can interpret typically corresponds to the
-planning domain in use: the robot can only process the \concept{GoTo} action if
-the corresponding task description with its pre- and post-conditions is
-available to the symbolic task planner.
-
-The tree in Figure~\ref{fig|upper_tbox}\footnote{This subset of the ontology is
-indeed a tree. This has however not to be the case in general, and, as a matter
-of fact, the TBox of the whole {\sc Oro} common-sense ontology does not form a tree}
-is not equally developed in each directions at lower levels. The descendants of
-\concept{PartiallyTangibleThing} (\ie what we commonly call \emph{objects}), for
-instance, are pictured in Figure~\ref{fig|tangible_things_tbox}. This excerpt
-from the ontology makes the bottom-up design process clear: only few types of
-\emph{partially tangible} things appear, and those are only subclasses relevant
-to the context of service robotics in an human-like environment. For performance
-reasons as well as clarity, we do not attempt to pro-actively extend the
-conceptual coverage of the field. We instead opportunistically extend the
-common-sense knowledge when required by experiments.
+The set of actions that the robot can interpret usually closely resemble the
+planning domain in use (\ie the set of tasks known to the symbolic task planner,
+with their corresponding pre- and post-conditions).
+
+The tree\footnote{While this subset of the ontology is a tree, it does not
+generally have to be the case. In particular, the concept system (the
+\emph{TBox}) of the {\sc Oro} common-sense ontology does not form a tree.} in
+Figure~\ref{fig|upper_tbox} is not equally developed in every directions. For
+example, the subclasses of \concept{PartiallyTangibleThing} (\ie what we
+commonly call \emph{objects}) are shown in
+Figure~\ref{fig|tangible_things_tbox}. Our `bottom-up'' design process of the
+ontology is clearly apparent on this figure: only the subclasses relevant to the
+context of service robotics in an human-like environment are explicitly stated.
+For performance reasons as well as clarity, we decided against an extended
+conceptual coverage of what ``partially tangible things'' might be. We instead
+opportunistically extend the common-sense knowledge when required by studies.
 
@@
 }
-    \caption{TBox of the specialisations of \concept{PartiallyTangibleThing}.}
+    \caption{Subclasses of \concept{PartiallyTangibleThing} explicitly stated in the
+    OpenRobots Ontology.}
     \label{fig|tangible_things_tbox}
@@
 \concept{Bottle} as found in the {\sc Oro} ontology is a typical example:
-
 $$
@@
 If a human informs the robot that a given object is indeed a bottle, the robot
-can then derive a few more informations on this object. And, conversely, if the
-human affirms that a car is a bottle, the robot may question this assertion
-because of the inconsistent size (the {\sc Dialogs} module relies on such
-logical consistency checks when processing natural language inputs, both to
-ensure that the verbal input has been correctly acquired and to verify that what
-the human says does actually make sense).
-
-We will come back and discuss in section~\ref{krs-discussion} the strengths and
-weaknesses of this knowledge framework.
+derives more informations on the object. Conversely, if the human affirms that
+``a car is a bottle'', the reasoner may detect logical contradictions (like
+inconsistent sizes) and reject the assertion. The {\sc Dialogs} module relies on
+such logical consistency checks when processing natural language inputs, both to
+ensure that the verbal input has been correctly acquired and parsed, and also to
+verify that what the human says is logically consistent.
+
+We discuss the strengths and
+weaknesses of this knowledge framework in Section~\ref{krs-discussion}.
 
@@
 
-\subsubsection{RDF as a Formalism for Semantics}
-
-The {\sc Oro} server relies on Description Logics (OWL) to represent and
-manipulate knowledge.
-
-Relying on RDF triples and Description Logics has advantages such as good
-understanding of its trades-off, thanks to being widespread in the semantic Web
-community, the availability of mature libraries to manipulate the ontology,
-interoperability with several major on-line knowledge bases ({\sc
-OpenCyc}, {\sc WordNet}, {\sc DBPedia} or {\sc RoboEarth}~\cite{Waibel2011} are
-some examples), open-world reasoning, and the formal guarantee of decidability
-(it is always possible to classify a Description Logics ontology).
-
-It also has notable limitations, both fundamental (the suitability of
-Description Logics when reasoning on --typically non-monotonic-- commonsense
-knowledge is questionable) and practical: RDF triples imply only binary
-predicates, which constrains the expressiveness of the system or leads to
-cumbersome reifications. Alternatives exist (like {\sc
-KnowRob}~\cite{Tenorth2009a}) that mix RDF with more expressive logic languages
-like {\sc Prolog} with other limitations, like closed-world reasoning.
-Classification performance is another issue: from our experience, with an
-ontology sized for a standard experiment (about 100 classes and 200 instances),
-classification typically takes about 100ms, which becomes problematic during
-interactions.  Besides, the performances are difficult to predict, since a
-seemingly inoffensive new statement may indirectly change abruptly the logical
-complexity of the whole knowledge model and lead to notable degradation of
-classification time.
-
-This knowledge model also largely excludes representation of continuous
-phenomena (like time) or uncertain phenomena. When required (for instance for
-action recognition), these are managed by dedicated components (like {\sc
-Spark}), and are not exposed at the semantic level.
-
-While alternatives like \emph{Answer Set Programming} have been
-successfully investigated in robotics~\cite{Chen2010,Erdem2012}, in particular
-to deal with non-monotonic reasoning, we did not actually hit any brick wall
-while working with OWL ontologies. We may reconsider this choice at a later
-stage, but until now it has proven an effective framework to quickly explore
-implementations of new cognitive abilities (for instance, it has been
-conceptually and technically easy to add support for independent knowledge
-models, one per agent the robot interacts with).
-
-Besides, because ontologies and RDF statements are relatively simple concepts
-to grasp, it also effectively helped to grow awareness amongst colleagues on
-the significance of the ``semantic level'' when developing new components for
-the robot.
-
 \subsubsection{Limits of Disambiguation at Semantic Level}

### 2016-06-14 paper.tex
Fixing language 3

@@
 \newcommand{\eg}{{e.g.\xspace}}
-\newcommand{\al}{{et al.\xspace}}
+\newcommand{\etal}{{et al.\xspace}}
 
@@
 action recognition), these are managed by dedicated components (like {\sc
-Spark}), and are not exposed at the semantic level.
-
-While alternatives like \emph{Answer Set Programming} have been successfully
-investigated in robotics~\cite{Chen2010,Erdem2012}, in particular to deal with
-non-monotonic reasoning, we did not actually encounter major reasoning issues
-while working with OWL ontologies. We may reconsider this choice at a later
-stage, but until now it has proven an effective framework to quickly explore
-implementations of new cognitive abilities (for instance, it has been
-conceptually and technically easy to add support for independent knowledge
-models, one per agent the robot interacts with).
-
-Anecdotally, and because ontologies and RDF statements remain conceptually simple
-(compared to full logical languages like {\sc Prolog}), their adoption has also
-effectively helped to grow awareness amongst colleagues on the significance of
-the ``semantic level'' when developing new components for the robot.
+Spark}, discussed at Section~\ref{sect|sit-ass}), and are not exposed at
+the semantic level.
+
+Alternative formalisms have been successfully investigated in robotics to
+address some of these restrictions. \emph{Answer Set Programming} has been
+used~\cite{Chen2010,Erdem2012} for instance to better supports non-monotonic
+reasoning. We have also pointed in~\cite{lemaignan2015mutual} how modal logics like
+the epistemic logics could be relevant to the particular field of social
+human-robot interaction as they would allow for convenient representation of
+alternative mental models. First-order logic and OWL ontologies have however
+proved so far a simple, effective and sufficient symbolic framework for our
+experimental applications.
+
+Anecdotally, and because ontologies and RDF statements remain conceptually
+simple (compared to full logical languages like {\sc Prolog} or modal logics),
+their adoption has also effectively helped to grow awareness amongst colleagues
+on the significance of the ``semantic level'' when developing new components
+for the robot.
 
@@
 ``a car is a bottle'', the reasoner may detect logical contradictions (like
-inconsistent sizes) and reject the assertion. The {\sc Dialogs} module relies on
-such logical consistency checks when processing natural language inputs, both to
-ensure that the verbal input has been correctly acquired and parsed, and also to
-verify that what the human says is logically consistent.
-
-We discuss the strengths and
-weaknesses of this knowledge framework in Section~\ref{krs-discussion}.
+inconsistent sizes) and reject the assertion. The {\sc Dialogs} module relies
+on such logical consistency checks when processing natural language inputs,
+both to ensure that the verbal input has been correctly acquired and parsed,
+and also to verify that what the human says is logically consistent.  We
+further discuss the strengths and weaknesses of this knowledge framework in
+Section~\ref{krs-discussion}.
 
@@
 skill, in particular in the human-robot interaction context: in that case, the
-link that the robot has to established between its percepts and its symbol must
-\emph{model} to a large extent the human representations in order to effectively
+link that the robot has to establish between percepts and symbols must
+map as well as possible to the human representations in order to effectively
 support communication.
@@
 
-While the previous section provided an overview of our cognitive architecture
-as well as its associated knowledge model, this section discusses its main
-``building blocks''. Each of these components may have connections to several
-others, and we invite the reader to refer to Figure~\ref{fig|archi} as a guide
-if needed.
-
-\begin{inparaenum} In our context, we call \emph{cognitive skills} the
-deliberative behaviours that are \item \emph{stateful} (keeping track of
-previous states is typically required for the component to perform adequately),
+The previous section has provided an overview of our cognitive architecture
+as well as the associated knowledge model. We discuss in this section its
+building blocks. Each of these components have connections to several
+others, and Figure~\ref{fig|archi} can be referred to as a guide.
+
+\begin{inparaenum} We call \emph{cognitive skills} the
+deliberative behaviours that are \item \emph{stateful}, \ie keeping track of
+previous states is typically needed for the component to perform adequately;
 \item \emph{amodal} in that the skill is not inherently bound to a specific
-perception of actuation modality, \item manipulate \emph{explicit semantics},
-typically by the mean of symbolic reasoning, \item operate at the
+perception or actuation modality; \item manipulate \emph{explicit semantics},
+typically by the mean of symbolic reasoning; \item operate at the
     \emph{human-level}, \ie are legible to the humans, typically by acting at
@@
 
-We start by discussing first the main \emph{internal} cognitive capabilities,
-implemented in the knowledge base itself, and then discuss successively the
-situation assessment module {\sc Spark}, the dialogue processor {\sc Dialogs},
-the symbolic task planner HATP, and finally the main features of our execution
-controllers.
+We present first the main \emph{internal} cognitive capabilities, implemented in
+the {\sc Oro} knowledge base itself, and then discuss successively the situation
+assessment module {\sc Spark}, the dialogue processor {\sc Dialogs}, the
+symbolic task planner HATP, and finally the main features of our execution
+controllers {\sc Shary} and {\sc pyRobots}. Note that greater details on each of
+these modules can be found in their respective publications (the corresponding
+references are provided hereafter).
 
@@
 As mentioned in the previous section, we use the Pellet open-source reasoner to
-reason on the knowledge base. This enables several standard inference services:
-\emph{consistency checking}, \emph{concept satisfiability},
-\emph{classification} and \emph{realisation} (computation of the most specific
-classes that an individual belongs to). In case of logical inconsistency, the
-reasoner can also provide \emph{explanations} that are used in particular for
-debugging.
-
-Besides those, {\sc Oro} server implements several algorithms (presented
+reason on the knowledge base. It supports several standard inference processes:
+consistency checking, concept satisfiability, classification and realisation
+(computation of the most specific classes that a concept belongs to). In
+case of logical inconsistency, the reasoner can also provide explanations (we
+currently only use them for debugging purposes).
+
+Besides, {\sc Oro} server implements several algorithms (presented
 in~\cite{Ros2010b}) to identify
@@
 agent, possibly including knowledge that contradicts the subject's own model: for
-example, a book can be at the same time \emph{visible} for myself, and \emph{not
-visible} for you. Children develop this skill, which is essential to understand others' perspectives during
+example, a book can be at the same time \emph{visible} for agent A, and \emph{not
+visible} for agent B. Children develop this skill, which is essential to understand others' perspectives during
 interactions, around the age of three.
 
-From a robotics point of view, it supposes the ability to build, store and
+From the point of view of robotics, it supposes the ability to build, store and
 retrieve separate models of the beliefs of the agents the robot interacts with.
 Our knowledge base implements such a mechanism: when the robots recognises that
-a new agent has been created in the knowledge base, it initialises a new,
-independent, ontology for this agent. All the ontologies that are created share
+a new agent has appeared, it initialises a new
+independent ontology for this agent. All the ontologies that are created share
 the same common-sense knowledge, but rely on each agent's perspective for the
@@
 One classical application of this cognitive skill is the so-called
-\emph{False-Belief} experiment (also known as the \emph{Sally and Ann}
-experiment)~\cite{Leslie2000}: a child is asked to watch a scene where two
-people, A and B, manipulate objects. Then A leaves and B hides away one
+\emph{False-Belief} experiment (also known as the \emph{Sally and Anne}
+experiment, introduced by~\cite{wimmer1983beliefs} from an original
+experimental setting by~\cite{baron1985does}): a child is asked to watch a scene where two
+people, A and B, manipulate objects. At some point, A leaves and B hides away one
 object. When A comes back, we ask the child ``where do you think A will
@@
 in the knowledge base, we have been able to replicate this experience with
-our robots~\cite{warnier2012when}.
+our robots~\cite{warnier2012when}, in a manner similar to Breazeal
+\etal\cite{breazeal2009embodied}.
 
@@
 Memory has been studied at length in the cognitive psychology and
-neuro-psychology communities: to give the main pointers, the idea of \emph{short-term} and
-\emph{long-term} memory is due to Atkinson and Shiffrin~\cite{Atkinson1968};
-Anderson~\cite{Anderson1976} proposes to split memory into \emph{declarative} (explicit)
-and \emph{procedural} (implicit) memories; Tulving~\cite{Tulving1985} organises
-the concepts of \emph{procedural}, \emph{semantic} and \emph{episodic} memories
-into a hierarchy. Short-term memory is eventually refined with the concept of
-\emph{working memory} by Baddeley~\cite{Baddeley2010}.
-In the field of cognitive architectures, the {\sc Soar}
-architecture~\cite{Lehman2006} is one of those that tries to reproduce a
-human-like memory organisation. The GLAIR cognitive
-architecture~\cite{Shapiro2009} also has a concept of long term/short term and
-episodic/semantic memories.
+neuro-psychology communities: the idea of \emph{short-term} and \emph{long-term}
+memory is due to Atkinson and Shiffrin~\cite{Atkinson1968};
+Anderson~\cite{Anderson1976} proposes to split memory into \emph{declarative}
+(explicit) and \emph{procedural} (implicit) memories; Tulving~\cite{Tulving1985}
+organises the concepts of \emph{procedural}, \emph{semantic} and \emph{episodic}
+memories into a hierarchy. Short-term memory is eventually refined with the
+concept of \emph{working memory} by Baddeley~\cite{Baddeley2010}.  In the field
+of cognitive architectures, the {\sc Soar} architecture~\cite{Lehman2006} is one
+of those that tries to reproduce a human-like memory organisation. The GLAIR
+cognitive architecture~\cite{Shapiro2009} also has a concept of long term/short
+term and episodic/semantic memories.
 
@@
 forgetting facts after a variable amount of \emph{time}, it actually covers
-more mechanisms that are relevant to robotics, like selective remembering
-triggered by a specific context or reinforcement learning.
+more mechanisms that are relevant to robotics, like priming
+(concept pre-activation triggered by a specific
+context~\cite{baxter2012modelling}) or reinforcement learning.
 
@@
 families.  When new statements are inserted in the knowledge base, a
-\emph{memory profile} is optionally attached to them.  Three such profiles are
+\emph{memory profile} is attached to them.  Three such profiles are
 predefined: {\it short term}, {\it episodic} and {\it long term}. They are
@@
 seconds, 5 minutes and no time limit). After this duration, the statements are
-automatically retracted.
+automatically removed.
 
@@
 
-We rely however on this mechanism in certain cases: for instance, some modules
+We rely however on this mechanism in certain cases: some modules
 like the natural language processor use the {\it short term} memory profile to
-mark for a few seconds important concepts that are currently manipulated by the
+mark concepts that are currently manipulated by the
 robot as \emph{active concepts}: if a human asks the robot ``Give
@@
 memory (which can be considered, in this case, to be a working memory). We use
-this feature to trace certain knowledge-related processes. On the other hand,
+this feature during dialogue disambiguation for instance. On the other hand,
 our perception layer does not make use of this mechanism: as described in the
 next section, the environment model of the robot is continuously updated and the
-derived symbolic knowledge is therefore transient.
+derived symbolic knowledge is therefore transient: it lasts only as long as the
+environment remains in the same state.
 
@@
 
-    \node[anchor=north] at (-6.5,-5.3) {a) Agent-independent\\locations};
-    \node[anchor=north] at (-2.1,-5.3) {b) Agent-dependent\\placements};
-    \node[anchor=north] at (1.6,-5.3) {c) Visibility};
+    \node[anchor=north] at (-6.5,-5.3) {a) Agent-independent\\(allocentric)
+        spatial relations};
+    \node[anchor=north] at (-2.1,-5.3) {b) Agent-dependent\\(egocentric) spatial
+        relations};
+    \node[anchor=north] at (1.6,-5.3) {c) Visibility/Pointing};
     \node[anchor=north] at (6.5,-5.3) {d) Reachability};
@@
     part). {\sc Spark} computes symbolic relationships between objects and
-    agents, and feeds them to the knowledge base.}
+    agents, and exports them to the knowledge base.}
 
@@
 cognitive skill that a robot exhibits when it assesses the nature and content of its
-surroundings and efficiently monitors its evolution.
+surroundings and monitors its evolution.
 
@@
 We rely on a dedicated geometric and temporal reasoning module called {\sc
-Spark} (\emph{SPAtial Reasoning \& Knowledge}, \cite{Sisbot2011}). It is a
-situation assessment reasoner that generates symbolic knowledge from the
-geometry of the environment with respect to relations between objects,
-robots and humans (Figures~\ref{fig|spark-oro}
-and~\ref{fig:sparkSubfigures}), also taking into account the different
-perspective that each agent has on the environment.
-
-{\sc Spark} embeds an \emph{amodal}~\cite{Mavridis2006} geometric model of the
-environment that serves both as basis for the fusion of the perception
-modalities and as bridge with the symbolic layer. This geometric model is made
-of CAD 3D models of the objects, furnitures and robots, and full body, rigged
-models of the humans, and it is updated at run-time by the robot based on its
-sensors (in most of our studies, objects are identified and localised through 2D
-fiducial markers, while humans are tracked with Kinect-like devices, optionally
-assisted by motion capture to accurately track the head motion, which is
-required to compute what the human is looking at).
+Spark} (\emph{SPAtial Reasoning \& Knowledge}, presented in~\cite{Sisbot2011}).
+It is a situation assessment reasoner that generates symbolic knowledge from the
+geometry of the environment with respect to relations between objects, robots
+and humans (Figures~\ref{fig|spark-oro} and~\ref{fig:sparkSubfigures}), also
+taking into account the different perspective that each agent has on the
+environment.  {\sc Spark} embeds an \emph{amodal} (as defined by Mavridis and
+Roy in~\cite{Mavridis2006}: the different perceptual modalities are abstracted
+away into a blended spatial model) geometric model of the environment that
+serves both as basis for the fusion of the perception modalities and as bridge
+with the symbolic layer. This geometric model is built from 3D CAD models of the
+objects, furnitures and robots, and full body, rigged models of humans
+(Figure~\ref{fig:sparkScreenshot}).  It is updated at run-time by the robot's
+sensors (usually, a combination of vision-based tracking of 2D fiducial markers
+to identify and localize objects, and Kinect-based skeleton tracking of humans,
+optionally assisted by motion capture to accurately track the head motion, which
+is required to compute what the human is looking at).
 
@@
        \subfigure[Physical setting]{%
-%           \label{fig:wuweiPhoto}
+           \label{fig:wuweiPhoto}
            \includegraphics[width=0.5\textwidth]{etat2-P1010769_brightened-v2.jpg}
@@
        \subfigure[Corresponding 3D model view]{%
-%          \label{fig:sparkScreenshot}
+          \label{fig:sparkScreenshot}
           \includegraphics[width=0.43\textwidth]{etat2_photo.png}

### 2016-06-14 paper.tex
Fixing language 4

@@
 
-%Our approach, implemented in the {\sc Spark} module, is a grounded amodal
-%representation of the robot environment, where different sensing modalities
-%(vision-based fiducial markers tracking, RGBD human tracking, motion capture)
-%are merged and where amodal spatial reasoning takes place (for instance,
-%explanation generation for unexpected perceptions like a disappearing object).
-
-{\sc Spark} is designed to run and continuously update the knowledge base (in
-our experiments, it runs at about 10Hz). At each step, it re-computes spatial
-relationships and affordances for the whole scene, and send the delta (new
+{\sc Spark} runs and continuously updates the knowledge base at about 10Hz. At each step, it re-computes spatial
+relations and affordances for the whole scene, and send the delta (new
 relations and relations that do not hold anymore) to the knowledge base. This
 approach may raise scalability concerns (we did not however hit major
-performance issues in our constrained scenarios), but greatly simplifies the
+performance issues in our constrained scenarios), but simplifies the
 management of the \emph{dynamics} of the knowledge (\emph{When do I discard
@@
 
-   \caption{Toy scenario involving videotapes that have to be manipulated, with
-       other objects used as supports and containers.  After identification and
-       localisation of the set of objects and acquisition of the position and
-       posture of the human partner, the robot is able to compute that two tapes
-       only are reachable by itself: the black and grey (in the 3D model) tapes.
-       The third tape and the pink box are only reachable by the human. This
-       situation leads to the symbolic model presented in
-       Table~\ref{table|beliefsfig7}.  }%
+   \caption{Test setup involving videotapes that are manipulated, with other
+       objects acting as supports or containers. After identification and
+       localisation of the set of objects (using fiducial markers) and
+       acquisition of the position and posture of the human partner (using
+       skeleton tracking), the robot is able to compute that two tapes only are
+       reachable by itself: the black and grey (in the 3D model) tapes.  The
+       third tape and the pink box are only reachable by the human. This
+       physical situation is transformed by {\sc Spark} into the symbolic model
+       presented in Table~\ref{table|beliefsfig7}.  }%
 
@@
 \hline
-  \hspace{0.7cm}\stmt{PINK\_BOX isReachable false}\\
+\hspace{0.7cm}\stmt{PINK\_BOX isReachable {\bf false}}\\
   \hspace{0.7cm}\stmt{WHITE\_TAPE isReachable {\bf false}}\\
@@
 \hline
-  \hspace{0.7cm}\stmt{PINK\_BOX isReachable false}\\
+  \hspace{0.7cm}\stmt{PINK\_BOX isReachable {\bf true}}\\
   \hspace{0.7cm}\stmt{WHITE\_TAPE isReachable {\bf true}}\\
@@
 
-\paragraph{On Perspective Taking} \emph{Visual} perspective taking refers to the
+\paragraph{Perspective Taking} \emph{Visual} perspective taking refers to the
 ability for visually perceiving the environment from other's point of view.
@@
 the one visible from the adult point of view), without asking~\cite{Moll2006}.
-
-Besides, in order to compute a visual perspective, the actual visibility alone
-is not enough. We include not only what the other person sees in a given
-moment, but also what he \emph{can} see with a minimal effort (moving the eyes
-or the head). To model the potential visibility of an object we compute a
-visibility ratio while turning the head of the agent model towards the object.
 
@@
 location (\eg. ``go north'') no ambiguity arises.
-
-In {\sc Spark}, we use two types of the frames of reference: \emph{ego-centric}
-(from the robot perspective) and \emph{allo-centric}, \ie addressee-centred
-(from the human perspective).
+In {\sc Spark}, agent-dependent spatial relations are computed from the frame of
+reference of each agent.
 
@@
 (like \emph{on}, \emph{next to}...) instead of precise, absolute positions.
-These type of descriptors have been studied in the context of language
-grounding \cite{O'Keefe1999,Matuszek2010,Regier2001,Kelleher2006,Blisard2005}.
-In {\sc Spark} we focus on agent-independent symbolic locations and agent-dependent,
-relative locations.
+These type of descriptors have been studied in the context of language grounding
+\cite{O'Keefe1999,Matuszek2010,Regier2001,Kelleher2006,Blisard2005}.  {\sc
+Spark} distinguishes between agent-independent symbolic locations (allocentric
+spatial relations) and agent-dependent, relative locations (egocentric spatial
+relations).
 
@@
 $BB_{O_1}$ and $BB_{O_2}$.  \concept{isNextTo} indicates whether an object $O_1$
-is next to another object $O_2$. We cannot use a simple distance threshold to
+is next to another object $O_2$. We do not use a simple distance threshold to
 determine if two objects are next to each other since the relation is highly
@@
 objects (\eg two bottles). Thus, the relation between the dimensions and the
-distances of the objects are taken into account.
-
-{\sc Spark} also compute symbolic facts related to agent independent world dynamics.
-The predicate \concept{isMoving} states, for each tracked entity, whether it is
-currently moving or not.
-
-Many topological relations are conversely dependent from the observation point.
-The predicate \concept{hasRelativePosition} represents such spatial relations
-between agents and objects that are agent dependent.  We compute these spatial
-locations by dividing the space around the referent (an agent) into $n$ regions
-based on arbitrary angle values relative to the referent orientation
-(Figure~\ref{fig|spark-oro}\emph{b}).  For
-example, for $n = 4$ we would have the space divided into \emph{front, left,
-right} and \emph{back}. Additionally, two proximity values, \emph{near} and
-\emph{far}, are also considered. The number of regions and proximity values can
-be chosen depending on the context where the interaction takes place.
+distances of the objects are taken into account.  {\sc Spark} also compute
+symbolic facts related to agent independent world dynamics.  The predicate
+\concept{isMoving} states, for each tracked entity, whether it is currently
+moving or not.
+
+Many other topological relations are dependent from the observation point
+(egocentric perspective).  The predicate \concept{hasRelativePosition}
+represents such spatial relations between agents and objects that are agent
+dependent.  We compute these spatial locations by dividing the space around the
+referent (an agent) into $n$ regions based on arbitrary angle values relative to
+the referent orientation (Figure~\ref{fig|spark-oro}\emph{b}).  For example, for
+$n = 4$ we would have the space divided into \emph{front, left, right} and
+\emph{back}. Additionally, two proximity values, \emph{near} and \emph{far}, are
+also considered. The number of regions and proximity values can be chosen
+depending on the context where the interaction takes place.
 
@@
     figure~\ref{fig|spark-oro}\emph{c} the field of view of a person is
-    illustrated with a grey cone (broader one). While he is able to see the two
+    illustrated with a grey cone (the wider one). While he is able to see the two
     small boxes on the table in front of him, the big box on his right is out of
     his FOV, and therefore, he is not able to see it.
+
+    Besides, {\sc Spark} also computes the \concept{seesWithHeadMovement}
+    relation by simulating a small left-right rotation of the head. It
+    represents what an agent \emph{could} see with a minimal effort.
 
@@
     Figure~\ref{fig|spark-oro}\emph{c} shows the field of attention
-    of a person with a green cone (narrower one). In this example only the grey
+    of a person with a green cone (the narrower one). In this example only the grey
     box satisfies the \concept{looksAt} relation.
@@
     This relation is computed by placing a virtual camera on the hand, aligned
-    with the forearm. \concept{pointsAt} is particularly useful during
-    interaction when one of the agents is referring to an object saying ``this''
-    or ``that'' while pointing at it.
-
-To make recognition more robust, these three first capabilities are filtered
+    with the forearm. \concept{pointsAt} is typically used during dialogue
+    grounding, for instance when one of the agents is referring to an object
+    saying ``this'' or ``that'' while pointing at it.
+\end{itemize}
+
+To ensure a sufficiently stable recognition, these three capabilities are post-filtered
 with an hysteresis function at the geometric level.
 
-\item \emph{Reachable}: it allows the robot to estimate the agent's capability
-    to reach an object, which is fundamental for task planning. For example, if
-    the user asks the robot to give him/her an object, the robot must compute a
-    transfer point where the user is able to get the object afterwards.
-    Reachability is computed for each agents based on Generalised Inverse
-    Kinematics and collision detection.
+\begin{itemize}
+\item \emph{Reachable} enables the robot to estimate the agent's ability to
+    reach for an object, which is instrumental for effective social task
+    planning. For example, if the user asks the robot to give him/her an object,
+    the robot must compute a transfer point where the user is able to get the
+    object afterwards.  Reachability is computed for each agents based on
+    Generalised Inverse Kinematics and collision detection.
 
@@
 
-Table~\ref{facts} lists all the symbolic facts that are computed by {\sc Spark},
-along with the admissible classes for the subjects and objects of the statements.
+Table~\ref{facts} also lists these abilities, along with the admissible classes
+for the subjects and objects of the statements.
 
@@
     \centering
-    \begin{tabular}{p{1.5cm}lp{2cm}p{3.7cm}}
+    \begin{tabular}{p{1.5cm}lp{2cm}p{5cm}}
     \textbf{Subject} & \textbf{Predicate} & \textbf{Object} & \emph{Notes} \\ 
@@
 		 \concept{Agent}  & \concept{pointsAt} $\equiv$ \concept{cyc:pointingToward}  & \concept{SpatialThing} \\ 
-		 \concept{Agent}  & \concept{focusesOn}  &  \concept{SpatialThing}  &  \par \tiny \concept{looksAt} $\wedge$ \concept{pointsAt} $\Rightarrow$ \concept{focusesOn} \\
+		 \concept{Agent}  & \concept{focusesOn}  &  \concept{SpatialThing}  &  \par \tiny \concept{looksAt} $\wedge$ \concept{pointsAt} $\Leftrightarrow$ \concept{focusesOn} \\
 		\concept{Agent} & \concept{seesWithHeadMovement} &  \concept{SpatialThing} \\
-		\concept{Agent} & \concept{reaches} &  \concept{Object} \\
+        \concept{Agent} & \concept{canReach} &  \concept{Object} & \\ 
+        \concept{Object} & \concept{isReachable} &  \concept{xsd:boolean} & \par \tiny \concept{myself canReach *} $\Leftrightarrow$ \concept{* isReachable true} \\
 
@@
 
-It is sometimes difficult or even impossible to see and/or track an object in
-certain states. This happens, for instance, when the object is hidden in a
-container, when the robot holds it in its gripper, and more generally in any
-state in which it is hidden by something else. {\sc Spark} models the possible
-symbolic states of objects (whether the object is on a furniture, in an agent
-hand, in a container, etc.), and according to the robot perception of what has
-happened since the object was last seen, the robot builds a belief of the
-current possible state with associated probability model.
+Occlusions, be they purposeful (when an agent hides an object behind or inside
+another one) or not, commonly lead to difficulties in perceiving and tracking
+objects in certain states. To mitigate this issue, {\sc Spark} also models the
+possible symbolic states of objects (whether the object is on a furniture, in an
+agent hand, in a container, etc.) with an associated probability distribution
+based on the robot's perception of what has happened since the object was last
+seen.  This probabilistic model of the environment is not currently integrated
+with the knowledge base, and is instead directly queried by the execution
+controllers when necessary.
 
@@
 
-Monitoring human activity is crucial for maintaining a coherent state of the
-world. While full human action and activity recognition is a task that requires
-knowledge and reasoning both on high level facts like goals, intentions and
-plans, as well as bottom-up data from agent and object motions, simple temporal
-and geometric reasoning on human hand trajectories and potential objects
-placements can provide clues for high level human monitoring processes. We call
-this temporal and geometric reasoning \emph{primitive action recognition}. Those
-primitive actions are assessed through monitoring situations like ``an empty
-hand is close to an object on a table'' (precursor for a \emph{pick}), or ``a
-hand holding an object is over a container'' (precursor for a \emph{throw}).
-{\sc Spark} recognises a core set of such primitives that are relied upon by the
-robot to track the state transitions in the human plans.
+Monitoring human activity is needed by the execution controllers to track the
+engagement of the human and the progress of his/her actions. Full human action
+and activity recognition is a task that requires knowledge and reasoning both on
+high level facts like goals, intentions and plans, as well as bottom-up data
+from agent and object motions. {\Spark} implements a set of simple temporal and
+geometric heuristics on human hand trajectories and possible objects placements
+to recognize simple atomic actions. Those primitive actions are assessed through
+monitoring situations like ``an empty hand is close to an object on a table''
+(precursor for a \emph{pick}), or ``a hand holding an object is over a
+container'' (precursor for a \emph{throw}).  {\sc Spark} recognises a set of
+such primitives that are relied upon by the execution controller to track the
+state transitions in the human plans.
 
@@
 ({\tt PICK\_GOTO} followed by {\tt TAKE\_TO\_FREE}) that the controller executes
-(by first reaching the object, grasping it and bringing it back to a free
+(by first reaching for the object, grasping it and bringing it back to a free
 position).  The robot's perception monitors the evolution of the scene until the

### 2016-06-14 paper.tex
Fixing language 5

@@
 high level facts like goals, intentions and plans, as well as bottom-up data
-from agent and object motions. {\Spark} implements a set of simple temporal and
+from agent and object motions. {\sc Spark} implements a set of simple temporal and
 geometric heuristics on human hand trajectories and possible objects placements
@@
 
-The main weakness of our current implementation relates to the way semantics are
-conveyed: because we use 2D markers for artifacts, we avoid issues related to object
-segmentation and recognition. Each object has a unique identifier,
-which enables us to load a proper CAD model suitable for 3D spatial reasoning
-and prevent recognition mistakes in the knowledge base. While {\sc Spark}
-algorithms themselves are ignorant of the input sources, and would work equally
-well with a full object recognition stack, we do not have tackled it yet.
+In its current form, our situation assessment module makes two strong
+assumptions: the objects are known in advance (hence, we can rely on proper 3D
+CAD model for spatial reasoning) and the robot benefits of an almost perfect
+perception, made possible by the use of fiducial markers.  Each object receives
+a unique tag which enables us to accurately localize it in 3D and prevent
+recognition ambiguities that would be then reflected in the knowledge base.
+While {\sc Spark} algorithms themselves are ignorant of the input sources, and
+would work equally well with a full object recognition stack, we are yet to
+investigate this research area.
 
 Besides, temporal reasoning (essential for accurate action recognition for
-instance) is not properly addressed in our stack. Temporal reasoning is used
+instance) is not properly addressed in our architecture. Temporal reasoning is used
 only locally, and does not allow for tracking of long sequences or global events.
@@
 
-In our experience, natural language processing is one of the fields of
-human-robot interaction for which the introduction of the semantic layer has
-been most beneficial.  In~\cite{Lemaignan2011a} we explain the techniques and the
-tool called {\sc Dialogs} that we have developed for natural English language
-parsing and grounding, along with verbalisation and (minimalist) dialogue
-management.
-
-Natural language input (in experiments, we rely on an Android-based interface,
-with Google speech recognition) is parsed into a grammatical structure, and
-atoms of each sentence are resolved with the help of the ontology to ground
-concepts like objects (\ie when a user says ``pick the can'', it resolves to which
-instance of \emph{can} the user is referring to) and actions. Sentences are
-sorted into questions, desires and statements, and processed accordingly.
-Figure~\ref{dialogs|ex} gives a example of the processing of a simple order.
+Natural language is a basic interaction modality that we use in our system both
+as input (processing of the human speech) and as output (verbalization of the
+robot intents). Natural language processing is facilitated by our architecture
+mostly processing informations with semantics that are already close to the
+human-level. This section presents the main features of our speech processor,
+{\sc Dialogs}, that include semantic and multi-modal grounding, and interactive
+disambiguation. Algorithmic and implementation details are provided
+in~\cite{Lemaignan2011a}.
+
+We acquire natural speech input from the human participants through a custom
+Android-based interface that relies on the Google speech recognition API for
+speech-to-text and relays the textual transcript to the robot. It is parsed into
+a grammatical structure (\emph{Part of Speech} tagging), and the atoms of each
+sentence are resolved with the help
+of the ontology to ground concepts like objects (\ie when a user says ``pick the
+can'', it resolves to which instance of \emph{can} the user is referring to) and
+actions.  Sentences are classified into questions, desires and statements, and
+processed accordingly.  Figure~\ref{dialogs|ex} gives a example of the
+processing of a simple, non-ambiguous command.
 
@@
     \centering
-	\begin{tabular}{l|l}
-	\emph{Initial human knowledge} &
-	\emph{Human input}\\
-
+    \begin{tabular}{l}
+        \emph{Initial human knowledge} \\
         \hline
-
-    	\stmt{book\_1 type Book} &
-	``Give me the book.'' \\
-
-    	\stmt{book\_1 hasColor blue} & \\
-	\vspace{0.5em}\\
+        \stmt{book\_1 type Book} \\
+        \stmt{human\_1 rdf:type Human} \\
+        ~\\
+        ~\\
+        ~\\
+    \end{tabular}
+    \begin{tabular}{l}
+        \emph{Input}\\
+
         \hline
 
-	\emph{Generated partial statements} &
+        \concept{human\_1} says:\\
+        ``Give me the book'' \\
+        ~\\
+        ~\\
+        ~\\
+
+    \end{tabular}
+    \begin{tabular}{l}
+
+        \emph{Generated query} \\
+        \hline
+        \concept{find(?obj type Book)} \\ 
+        \hspace{0.2cm}$\Rightarrow$ \concept{?obj = book\_1} \\
+        ~\\
+        ~\\
+        ~\\
+    \end{tabular}
+    \begin{tabular}{l}
+
         \emph{Newly created statements}\\
         \hline
-    	\stmt{?obj type Book} & 
-	\stmt{human desires sit\_1} \\
-
-	\hspace{0.2cm}$\Rightarrow$ \concept{?obj = book\_1}
-    	& \stmt{sit\_1 type Give} \\
-    	& \stmt{sit\_1 performedBy myself} \\
-    	& \stmt{sit\_1 actsOnObject book\_01} \\
-    	& \stmt{sit\_1 receivedBy human} \\
+        \stmt{human\_1 desires sit\_1} \\
+         \stmt{sit\_1 type Give} \\
+         \stmt{sit\_1 performedBy myself} \\
+         \stmt{sit\_1 actsOnObject book\_01} \\
+         \stmt{sit\_1 receivedBy human} \\
     \end{tabular}
 
-    \caption{Processing of a simple, non-ambiguous order, taken
+    \caption{Processing of a simple, non-ambiguous command, taken
         from~\cite{Lemaignan2011a}. Thematic roles ({\tt performedBy}, {\tt
         actsOnObject}, {\tt receivedBy}) are automatically extracted by 
-    {\sc Dialogs} from the order ``Give me the book.''.}
+    {\sc Dialogs} from the imperative sentence ``Give me the book.''.}
 
@@
 more information), anaphora resolution (``give \emph{it} to me'') based on
-dialogue history. It also permits knowledge extension by learning new semantic
+dialogue history. It also supports knowledge extension by learning new semantic
 structures (for instance, a sentence like ``learn that cats are animals'' is
@@
 after checking for possible contradictions with existing knowledge),
-interprets common temporal and place adverbs (like \emph{above} or
-\emph{tomorrow}) and translates to a certain extent \emph{states} (``I'm
-tired'') into \emph{experiences} (\stmt{HUMAN experiences state\_1, state\_1
-hasFeature tired}). We kindly refer the readers to~\cite{Lemaignan2011a} for a
-complete discussion of these capabilities and the related algorithms.
+interprets common temporal and spatial adverbs (like \emph{above} or
+\emph{tomorrow}) and translates simple expressions of internal state (``I'm
+tired'') into \emph{experiences} (\stmt{human\_1 experiences state\_1, state\_1
+hasFeature tired}, see also Section~\ref{sect:desires}). A full account of the
+{\sc Dialogs} features and the corresponding algorithms is available
+in~\cite{Lemaignan2011a}.
 
@@
 
-Because all components rely on the same RDF formalism to format their outputs,
-the different communication modalities (\emph{explicit} like verbal, deictic or
-based on gaze, or \emph{implicit} like postures) are presented in a homogeneous
-way, as statements in the knowledge base. The dialogue grounding process makes
-use of them at two distinct levels to provide seamless multi-modal interaction.
-
-First, particular steps of the grounding process explicitly check for the
+Because the components of our architecture rely on the same RDF formalism to
+represent their outputs, the different communication modalities (\emph{explicit}
+like verbal, deictic or based on gaze, or \emph{implicit} like the position of
+the human) are presented in a homogeneous way, as statements in the knowledge
+base. The dialogue grounding process makes use of them at two distinct levels to
+provide multi-modal concept grounding.
+
+First, specific steps of the grounding process explicitly check for the
 presence and value of specific facts: for instance, when several instances match
@@
 some of them based on their \emph{visibility} for the speaker (implicit
-communication context built on the human posture).
+communication context taking into account the human position).
 
@@
 the cost of a simpler interaction: if the human says ``this'' without the robot
-tracking what the human points at, no \stmt{HUMAN pointsAt ...} fact is
-available in the knowledge base, and the robot fallbacks on the anaphora
-resolution step alone.
+tracking what the human points at, no \stmt{human\_1 pointsAt ...} fact would
+possibly be available in the knowledge base, and the robot fallbacks on the
+anaphora resolution step alone.
 
@@
 
-The HATP \emph{planning domain} defines a set of methods describing how to
+The HATP planning domain defines a set of methods describing how to
 decompose a task and represents the procedural knowledge of the robot as well as
@@
 The originality of HATP resides in its ability to produce plans for the robot's
-actions \emph{as well as} for the other participants (humans or robots), that we
+actions as well as for the other participants (humans or robots), that we
 call \emph{shared plans}.
 
-An important feature of HATP is that it treats agents as ``first-class
+HATP treats agents as ``first-class
 entities'' in the domain representation language. It can therefore
@@
 between agents and the other entities such as tables and chairs. This
-facilitates a post-processing step in HATP that splits the final
+enables a post-processing step that splits the final
 solution (sequence of actions) into two (or more if there are several
@@
 
-This effectively enriches the interaction capabilities
-of the robot by allowing for a form of human behaviour \emph{prediction}, which
-is used by the robot execution controller to monitor the engagement of the human
-partner during joint tasks.
-
-The planner also generates synchronisation points between the agents.  In the
-plan depicted on Figure~\ref{plan_hatp1}, the robot needs to wait for the
-success of the human agent's {\tt PUT} action (estimated through the fulfilment
-of the action post-conditions, typically $\neg$\stmt{GREY\_TAPE isOn TABLE}) to
-be able to start its own action.
+This effectively enriches the interaction capabilities of the robot by providing
+the system with what is in essence a prediction of the human behaviour. This
+prediction is used by the robot execution controller to monitor the engagement
+of the human partner during joint tasks.
+
+The planner also generates synchronisation points between the agents. For
+instance, in the plan depicted on Figure~\ref{plan_hatp1}, the human needs to
+wait for the success of the robot's action {\tt PUTRV}. The robot monitors the
+success of its own action (by checking for the fulfilment of the action
+post-conditions; in this particular case \stmt{HUMAN sees BOTTLE} and
+\stmt{HUMAN canReach BOTTLE}) to estimate what and when the human is likely to
+perform his next action (here, \concept{TAKE(BOTTLE, TABLE)}). This, in turn,
+allows the robot to monitor the human engagement and progress in the shared
+plan.
 
@@
 
-\caption{A plan produced by HATP with two action streams (human actions at the
-top, robot actions at the bottom, {\sf PUTRV} stands here for {\it Put it so it is both
-reachable and visible}). The arrow between the two streams represents
-a synchronization point.}
+\caption{An example of plan produced by HATP for a task consisting in
+    cooperatively moving objects into their associated bins. Two action streams
+    are generated (human actions at the top, robot actions at the bottom, {\sf
+    PUTRV} stands here for {\it Put it so it is both reachable and visible}).
+    The arrow between the two streams represents a synchronization point.}
 
@@
 
-A cost and a duration function is associated to each action.  The duration
+A cost and a duration function are associated to each action.  The duration
 function provides a duration interval for the action achievement and is used, in
@@
 Figure~\ref{plan_hatp2} illustrates such a socially-optimised plan where the
-\emph{no wasted time} social rule is applied. The resulting shared plan is
-considered better than the one depicted on Figure~\ref{plan_hatp1} according to
-a global evaluation of the costs and rules.
+\emph{no wasted time} social rule is applied: compared to the plan depicted in
+Figure~\ref{plan_hatp1}, the robot first moves the bottle so that the human can
+immediately take it and put it into the bin, thus removing the human idle time.
+The resulting shared plan is considered better than the one depicted on
+Figure~\ref{plan_hatp1} according to a global evaluation of the costs and rules.
 
@@
 
-\caption{An alternative plan where the \emph{no wasted time} social rule is used to optimise the
+    \caption{An alternative plan for the task presented in
+    Figure~\ref{plan_hatp1} where the \emph{no wasted time} social rule is used to optimise the
 total duration of the task. }
@@
 
-Because HATP is a generic symbolic task planner, we have been able to design a
-planning domain at a semantic level which is close to the one used in the
-human-robot dialogue (the planner vocabulary contains concepts like
-\texttt{give}, \texttt{table}, \texttt{is on}...). This leads to a natural
-mapping (low ``impedance mismatch'') between the knowledge extracted from the situation
-assessment or the dialogue, and the planner.
+As HATP is a generic symbolic task planner and do not enforce any abstraction
+level for the planning domain, we were able to design a domain built from
+top-level tasks whose semantics are close to the one used in the human-robot
+dialogue: the planner domain effectively contains concepts like \texttt{give},
+\texttt{table}, \texttt{is on}. This leads to an effective mapping between the
+knowledge extracted from the situation assessment or the dialogue, and the
+planner.
 
@@
 \subsubsection{Desires and Experiences}
+\label{sect:desires}
 

### 2016-06-14 paper.tex
Fixing language 6

@@
         \stmt{book\_1 type Book} \\
-        \stmt{human\_1 rdf:type Human} \\
+        \stmt{human\_1 type Human} \\
         ~\\
@@
 
-The HATP planning framework extends the
-traditional HTN (Hierarchical Task Network) planning domain representation and
-semantics by making them more suitable to elaborate plans which involve humans
-and robots acting together toward a joint goal. HATP is used by the robot to
-elaborate human-robot \emph{shared plans} which are then used to anticipate
-human action, to propose to human to act or even to ask help from the human.
+The HATP planning framework extends the traditional HTN (Hierarchical Task
+Network) planning domain representation and semantics by making them more
+suitable to elaborate plans which involve humans and robots acting together
+toward a joint goal. HATP is used by the robot to elaborate human-robot
+\emph{shared plans}~\cite{Grosz1996,Clark1996,Kemp2007} which are then used to
+anticipate human action, to propose to human to act or even to ask help from
+the human.
 
@@
 written in the \emph{Procedural Reasoning System} language~\cite{Ingrand1996}.
-We have also developed the Python-based {\sc pyRobots} collection of tools to
-provide a large set of high-level actions and an event-based architecture well
+We have also developed the Python-based {\sc pyRobots}~\cite{lemaignan2015pyrobots} that
+provides a large set of high-level actions and an event-based architecture well
 suited for prototyping. They both rely on extensive integration with the knowledge
-base, that serves as the primary source of semantics for decision making process.
-
-One of the main role of {\sc Shary} is to control the elaboration and execution
-of shared plans~\cite{Grosz1996,Clark1996,Kemp2007}. This means essentially
-context-based refinement and robot actions execution, as well as monitoring of
-those achieved by its human partner~\cite{Breazeal2003}. The challenge is to
-build such abilities in a generic way, and to provide several levels of
-parametrisation allowing to adapt to various environments, and various levels of
-involvement of the robot ranging from teammate behaviour to assistant or
-proactive helper.  Depending on this, the robot controller invokes the various
-human-aware planners and will react to events detected at {\sc Oro} level (as
-described below).
+base, that serves as the primary source of semantics for the decision making process.
+
+One of the main role of {\sc Shary} is to control the production and execution
+of shared plans. This means essentially context-based refinement and robot
+actions execution, as well as monitoring of those achieved by its human
+partner. One of the key design goal is to build such abilities in a generic
+way, and to provide several levels of parametrisation allowing to adapt to
+various environments, and various levels of involvement of the robot ranging
+from teammate behaviour to assistant or proactive helper. Based on this,
+the robot controller invokes the adequate human-aware planners and react to
+events triggered by the {\sc Oro} knowledge base (as described below).
 
 {\sc Shary}'s originality, as an execution control system, lies in its ability
-to take into account not only the task achievement but also communication and
-monitoring needed to support interactive task
-achievement~\cite{Rich1997,Sidner2005} in a flexible way. {\sc Shary} allows to
-link action execution \emph{communication policies} in order to produce signals
-towards the human and to react to human actions. The \emph{communication act} is
-the central concept in this formalism. It plays the role of transition
-condition, and, as such, represents an information exchange between the two
-agents. This exchange can be realised through dialogue, by an expressive motion
-or a combination of the two. It enables each agent to communicate their beliefs
-about the task to be achieved, in order to share mutual knowledge and to
-synchronise their activities.  This is done through real-time task-based
-situation assessment achieved by the combination of {\sc Shary} monitoring
-processes and {\sc Oro} inference mechanisms~\cite{fiore2014}.
+to take into account not only the task achievement but also the communication
+and monitoring needed to support interactive task
+achievement~\cite{Rich1997,Sidner2005} in a flexible manner. {\sc Shary} allows
+to bind action execution to \emph{communication policies} in order to produce
+signals towards the human and to react to human actions. The
+\emph{communication act} is the central concept in this formalism. It
+represents an information exchange between the two agents and plays the role of
+a transition condition. This exchange can be realised through dialogue, by an
+expressive motion or a combination of the two. It enables each agent to
+communicate their beliefs about the task to be achieved, in order to share
+mutual knowledge and to synchronise their activities.  This is done through
+real-time task-based situation assessment achieved by the combination of {\sc
+Shary} monitoring processes and {\sc Oro} inference
+mechanisms~\cite{fiore2014}.
+
+\subsubsection{Event-Driven Control}
+
+The {\sc Oro} server supports two paradigms to access its content: RPC-style
+queries (based on the standard SPARQL language) or events. A module can
+subscribe to an event by passing through an event pattern (in its simplest
+form, a partial statement like \stmt{? type Book}) and a callback.  Each
+time a new instance of a book appears in the knowledge base, the callback is
+triggered.
+
+This allows us to write reactive robot controllers with a high level of
+expressiveness: for instance, by subscribing to the event \stmt{human1 desires
+?action, ?action type Take, ?action actsOnObject ?obj, ?obj type Toy}, we could trigger a
+behaviour when the human expresses (through dialogue, gestures...) that he
+wants a toy.
+
+The robot controller designer does not need to directly care about how this
+\emph{desire} is produced (this is delegated to perception modules), he can
+focus on the semantic of the desire.
+
+Note also that we take advantage of the reasoning capabilities of the system:
+for example, the type of the object (\stmt{?obj type Toy}) may not be
+explicitly asserted, but inferred by the reasoner based on other assertions.
 
@@
 
-We split the interaction situations stemming from the situation assessment and
-communication components in two categories: \emph{desires} (related to
+We divide the interaction situations perceived from the situation assessment and
+the communication components into two categories: \emph{desires} (related to
 \emph{performative acts} in Austin's classification of speech
@@
 
-\emph{Desires} are typically human orders (``Give me that book''). The nature of
-the desired action (to pick, give, look, bring, show...), along with the action
+\emph{Desires} are typically human commands (``Give me that book''). The nature of
+the desired action (to pick, to give, to look, to bring, to show...), along with the action
 parametrization (thematic roles) are extracted from the knowledge base, and
@@
 implemented as a proof-of-concept. A broader framework that would support
-action alteration based on the user's experienced states remains to be worked on.}
-
-\subsubsection{Event-Driven Control}
-
-The {\sc Oro} server proposes two paradigms to access its content: RPC-style
-queries (based on the standard SPARQL language) or events. A module can
-subscribe to an event by passing through an event pattern (in its simplest
-form, a partial statement like \stmt{? type Book}) and a callback.  Each
-time a new instance of a book appears in the knowledge base, the callback is
-triggered.
-
-This allows us to write reactive robot controllers with a high level of
-expressiveness: for instance, by subscribing to the event \stmt{human1 desires
-?action, ?action type Take, ?action actsOnObject ?obj, ?obj type Toy}, we could trigger a
-behaviour when the human expresses (through dialogue, gestures...) that he
-wants a toy.
-
-The robot controller designer does not need to directly care about how this
-\emph{desire} is produced (this is delegated to perception modules), he can
-focus on the semantic of the desire.
-
-Note also that we take advantage of the reasoning capabilities of the system:
-for example, the type of the object (\stmt{?obj type Toy}) may not be
-explicitly asserted, but inferred by the reasoner based on other assertions.
+action alteration based on the user's experienced states remains to be investigated.}
 
@@
 Our other controller, {\sc Shary}, relies on the symbolic task planner to
-dynamically generate, at run-time, suitable sets of atomic actions.
+dynamically generate, at run-time, suitable sequence of atomic actions.
 
@@
 
-Manipulation and navigation actions listed in table~\ref{table|pyrobots_actions}
-rely on a dedicated 3D motion planner ({\sc Mhp}, depicted in
+The symbolic manipulation and navigation actions listed in
+Table~\ref{table|pyrobots_actions} are eventually executed by the execution
+controller with the help of a dedicated 3D motion planner (called {\sc Mhp},
 Figure~\ref{fig|archi}): robot placement and end-effectors trajectories are
 computed on-line to allow object manipulation that take into account task
-specific constraints and human postures, abilities and preferences.
-We kindly refer the interested reader to~\cite{Sisbot2008, Mainprice2011,
-Pandey2011} for details of these techniques.
+specific constraints and human postures, abilities and preferences.  We kindly
+refer the interested reader to~\cite{Sisbot2008, Mainprice2011, Pandey2011} for
+details of these techniques.
 
@@
 {\it Interactive Grounding I} (2011) & Multi-modal interaction, perspective taking & \cite{lemaignan2011what} \\
-{\it Roboscopie} (2011) & Theater, reflection on the future of HRI & \cite{lemaignan2012roboscopie} \\
-{\it Cleaning the table} (2011) & Whole stack integration & \cite{Alami2011} \\
+{\it Roboscopie} (2011) & Human-Robot theatre performance & \cite{lemaignan2012roboscopie} \\
+{\it Cleaning the table} (2011) & Complete architecture integration & \cite{Alami2011} \\
 {\it I'm in your shoes} (2012) & False beliefs & \cite{warnier2012when} \\
@@
 
-We briefly introduce two of them in order to illustrate in a practical way
+We present here elements of two of them in order to illustrate in a practical way
 different aspects of the architecture.  The first one is focused on knowledge
@@
 
-This first study is based on the following scenario: Tom and Jerry are moving to
-London, so they are packing things in boxes. \emph{Jido}, a mobile 
-manipulator, is observing while they move things here and there
-(Figure~\ref{fig|vpt}).  They ask questions to Jido to know where certain
-objects are located. This study focuses on multi-modal, interactive grounding only, and
-the robot does not actually perform any action besides simple head movements.
+This first study is based on a ``home move'' backstory: two users are moving
+their belongings to a different home, and need the help of a robot to pack.
+\emph{Jido}, a single-arm mobile manipulator, is observing while they carry
+over boxes (Figure~\ref{fig|vpt}), and answers questions pertaining to the
+location of specific objects.  This study focuses on multi-modal, interactive
+grounding only, and the robot does not actually perform any action besides
+verbal interaction and simple head movements.
 
@@
 
-Tom enters the room while carrying a big box (Figure~\ref{fig|vpt}). He
+User A enters the room while carrying a large box (Figure~\ref{fig|vpt}). He
 approaches the table and asks Jido to hand him the video tape: ``Jido, can
@@
 
-There are two video tapes in the scene: one on the table, and another one
-inside the cardboard box. Thus, the knowledge base returns both: $\Rightarrow$
+Two video tapes are visible to the robot: one on the table, and another one
+inside the cardboard box. Thus, the knowledge base returns both: 
 \concept{?obj = [BLACK\_TAPE, WHITE\_TAPE]}.
 
-However, only one is visible for Tom (the one on the table). Although there is
+However, only one is visible for User A (the one on the table). Although there is
 an ambiguity from the robot's perspective, the human referred to the tape using
@@
 %  \hspace{0.7cm}\stmt{WHITE\_TAPE type VideoTape}\\
-%  \hspace{0.7cm}\stmt{WHITE\_TAPE isIn cardBoardBox}\\
+%  \hspace{0.7cm}\stmt{WHITE\_TAPE isIn CARDBOARD\_BOX}\\
 %  \hspace{0.7cm}\stmt{WHITE\_TAPE isVisible true}\\
@@
 %\hline
-%Robot's beliefs about Tom (\emph{Tom's model}):\\
+%Robot's beliefs about User A (\emph{User A's model}):\\
 %\hline
@@
 %  \hspace{0.7cm}\stmt{WHITE\_TAPE type VideoTape}\\
-%  \hspace{0.7cm}\stmt{WHITE\_TAPE isIn cardBoardBox}\\
+%  \hspace{0.7cm}\stmt{WHITE\_TAPE isIn CARDBOARD\_BOX}\\
 %  \hspace{0.7cm}\stmt{WHITE\_TAPE isVisible false}\\
@@
 
-In this second situation, Jerry enters the living room without knowing where Tom had
-placed the video tapes (Figure~\ref{fig|pointing}). So he first asks Jido:
-``What's in the box?''. Before the robot can answer the question it has to
-figure out which box Jerry is talking about. Similar to the previous situation,
-two boxes are visible: \concept{find(?obj type Box)} $\Rightarrow$ \concept{?obj = [cardBoardBox, toolbox]}
+In this second situation, User B enters the living room without knowing where
+User A had placed the video tapes (Figure~\ref{fig|pointing}). So he first asks
+Jido: ``What's in the box?''. The robot first need to ground the word ``box''.
+Similar to the previous situation, two boxes are visible: \concept{find(?obj
+type Box)} $\Rightarrow$ \concept{?obj = [CARDBOARD\_BOX, TOOLBOX]}
 
@@
 procedure can not be applied. The robot generates a question (using the
-\emph{Discrimination} algorithm that we mentioned at section~\ref{reasoning})
-and asks Jerry which box he is referring to: ``Which box, the toolbox or the
-cardboard box?'' Jerry can answer the question, but he instead decides to point
+\emph{Discrimination} algorithm, Section~\ref{reasoning})
+and asks User B which box he is referring to by verbalizing the following question: ``Which box, the toolbox or the
+cardboard box?'' User B can answer the question, but he instead decides to point
 at it: ``This box'' (Figure~\ref{fig|pointing}). {\sc Spark} identifies the {\tt
-cardBoardBox} as being pointed at and looked at by the human and updates the
-ontology with this new information using a rule available in the common-sense
+CARDBOARD\_BOX} as being pointed at, as well as looked at, by the human and updates the
+ontology with this new information. The reasonner applies a rule available in the common-sense
 ontology \stmt{pointsAt(?ag, ?obj) $\land$ looksAt(?ag, ?obj) $\to$
-focusesOn(?ag, ?obj)}. The \textsc{Dialogs} module is then able to merge both
-sources of information, verbal (``this'') and deictic to distinguish the box
-Jerry refers to:
+focusesOn(?ag, ?obj)}. The \textsc{Dialogs} module merges both
+sources of information, verbal (``this'') and deictic, by issuing a specific
+query to the knowledge base that eventually lifts the ambiguity:
 
@@
     \begin{tabular}{l} 
-        \stmt{Jerry pointsAt carboardBox}\\ 
-        \stmt{Jerry looksAt carboardBox}\\ 
-        $\to$ \stmt{Jerry focusesAt carboardBox}\\ 
-        \hspace{0.7cm}$\Rightarrow$ {\tt ?obj = [cardBoardBox]}
+        \concept{find(?obj type Box; USER\_B focusesOn ?obj)}\\ 
+        \hspace{0.7cm}$\Rightarrow$ {\tt ?obj = [CARDBOARD\_BOX]}
     \end{tabular} 
@@
 
-Finally, the \textsc{Dialogs} queries the ontology about the content of the box
-and the question can be answered: ``Wall-E'' (the label of the object, stored in the
-ontology, is used to interact with the user).
+Finally, \textsc{Dialogs} queries the ontology about the content of the box and
+the question can be answered: ``Wall-E'' (the label of the object is statically
+asserted in the ontology).
 
 %\begin{center}
-%\stmt{?obj isIn cardBoardBox}\\
+%\stmt{?obj isIn CARDBOARD\_BOX}\\
 %\hspace{0.7cm}$\Rightarrow$ \concept{?obj = WHITE\_TAPE}\\
@@
 
-At this point Jerry wants to know where is the other tape: ``And where is the
+At this point User B wants to know where is the other tape: ``And where is the
 other tape?''. \textsc{Dialogs} processes this sentence using the
@@
 
-Since there is only one possible ``other'' videotape, no specific disambiguation is
-required, and the robot verbalises the result of the query to the knowledge
-base \stmt{BLACK\_TAPE isOn table, BLACK\_TAPE isNextTo toolbox} to ``The other
-tape is on the table and next to the toolbox.''
+Since there is only one possible ``other'' videotape, no specific
+disambiguation is required. The referent is uniquely identified and {\sc
+Dialogs} can then query for its location. The robot finally verbalises the
+result \stmt{BLACK\_TAPE isOn table, BLACK\_TAPE isNextTo TOOLBOX} into ``The
+other tape is on the table and next to the toolbox.''
 
@@
 
-    \caption{Face-to-face setting of the \emph{Clean the Table} scenario. The
+    \caption{The face-to-face setup of the \emph{Clean the Table} study. The
     physical situation, the {\sc Spark} model, and the current step of the plan
@@
 
-Figure~\ref{fig|cleantable-timeline} presents a practical sample of the whole
+Figure~\ref{fig|cleantable-timeline} walks through a simplifed version of the whole
 task. It depicts a run with a single tape on a table. The tape is
-reachable by the robot only, while the bin where objects are supposed to be
-placed at the end is reachable by the human only: the robot needs to come up
-with a shared plan that involves joint actions.
+reachable by the robot only, while the bin (where the objects are supposed to be
+eventually moved to) is reachable by the human only: the robot needs to come up
+with a shared plan that involves a joint action.
 
@@
 
-The first task ({\tt TAKE(GREY\_TAPE, TABLE)}) is instantiated by first checking
-that the task pre-conditions hold, and then calling the 3D motion planner
-(column \emph{Actions}, left). The motion planner returns two atomic actions
-({\tt PICK\_GOTO} followed by {\tt TAKE\_TO\_FREE}) that the controller executes
-(by first reaching for the object, grasping it and bringing it back to a free
-position).  The robot's perception monitors the evolution of the scene until the
-task's post-conditions are verified (at $t_2$, by satisfying the statement
-\stmt{ROBOT hasInHand TAPE}), and the next task is then started (placing the
-tape so that it becomes reachable by the human).
+The first task (\concept{TAKE(GREY\_TAPE, TABLE)}) is instantiated by checking
+that the task pre-conditions hold (in particular, \stmt{GREY\_TAPE isOn TABLE}
+must be true), and calling the 3D motion planner (column \emph{Actions},
+left). The motion planner returns two atomic actions ({\tt PICK\_GOTO} followed
+by {\tt TAKE\_TO\_FREE}) that the controller executes (by first reaching for
+the object, grasping it and bringing it back to a free position).  The robot's
+perception monitors the evolution of the scene until the task's post-conditions
+are verified (at $t_2$, by satisfying the statement \stmt{ROBOT hasInHand
+TAPE}), and the next task is then started (placing the tape so that it becomes
+reachable by the human).
 

### 2016-06-15 paper.tex
Polish on the timeline figure

@@
 
+        \fill[gray!10!white] (4.6,.5) rectangle (figbottom);
         \node at (2,0) (percept) {\bf Perception};
@@
             \node[below=0.5 of percept.south east, anchor=mid] {3D model};
+
         \node[right=4 of percept, minimum width=2.5cm] (kb) {\bf Knowledge};
@@
             \node[below=0.5 of kb.south east, anchor=mid west] (kbh) {human};
-            \draw[dotted] (kb.south) to (figbottom -| kb);
+            \draw[dotted] (kbr) to (figbottom -| kbr);
+            \draw[dotted] (kbh) to (figbottom -| kbh);
+
+        \fill[gray!10!white] (11.5,.5) rectangle (16.5,-28.5);
         \node[right=4 of kb] (plan) {\bf Plan};
@@
             \node[below=0.5 of plan.south east, anchor=mid west] (phuman) {human};
+            \draw[dotted] (probot) to (figbottom -| probot);
+            \draw[dotted] (phuman) to (figbottom -| phuman);
+
         \node[right=5 of plan] (action) {\bf Actions};
@@
             \node[below=0.5 of action.south east, anchor=west] (ahuman) {human\\(monitoring)};
+            \draw[dotted] (arobot) to (figbottom -| arobot);
+            \draw[dotted] (ahuman) to (figbottom -| ahuman);
 
@@
 
-        \node[anchor=east] at (-0.5, -3) (t1) {$t_1$};
-        \node[anchor=east, below=6 of t1] (t2) {$t_2$};
-        \node[anchor=east, below=6 of t2] (t3) {$t_3$};
-        \node[anchor=east, below=4 of t3] (t4) {$t_4$};
-        \node[anchor=east, below=5 of t4] (t5) {$t_5$};
+        \node[anchor=east] at (-0.5, -3) (t1) {\Large $t_1$};
+        \node[anchor=east, below=6 of t1] (t2) {\Large $t_2$};
+        \node[anchor=east, below=6 of t2] (t3) {\Large $t_3$};
+        \node[anchor=east, below=4 of t3] (t4) {\Large $t_4$};
+        \node[anchor=east, below=5 of t4] (t5) {\Large $t_5$};
 
@@
 
-        \node[align=left] at (cam1 -| kbr) (kb1) {\stmt{TAPE isReachable true}\\
+        \node[fill=white,align=left] at (cam1 -| kbr) (kb1) {\stmt{TAPE isVisible true}\\
+                                                        \stmt{TAPE isReachable true}\\
+                                                        \stmt{TAPE isOn TABLE}\\
+                                                        \stmt{BIN isVisible true}\\
+                                                        \stmt{BIN isReachable false}};
+
+        \node[fill=white,align=left] at (cam1 -| kbh) {\stmt{TAPE isVisible true}\\
+                                                        \stmt{TAPE isReachable false}\\
+                                                        \stmt{TAPE isOn TABLE}\\
+                                                        \stmt{BIN isVisible true}\\
+                                                        \stmt{BIN isReachable true}};
+
+        \node[fill=white,align=left] at (cam2 -| kbr) (kb2) {\stmt{ROBOT hasInHand TAPE}};
+
+        \node[fill=white,align=left] at (cam3 -| kbr) {\stmt{TAPE isReachable true}\\
                                             \stmt{TAPE isVisible true} \\
@@
 
-        \node[align=left] at (cam1 -| kbh) {\stmt{TAPE isReachable false}\\
+        \node[fill=white,align=left ] at (cam3 -| kbh) (kb3) {\stmt{TAPE isReachable true}\\
                                              \stmt{TAPE isVisible true}};
 
-        \node[align=left] at (cam2 -| kbr) (kb2) {\stmt{ROBOT hasInHand TAPE}};
-
-        \node[align=left] at (cam3 -| kbr) {\stmt{TAPE isReachable true}\\
-                                            \stmt{TAPE isVisible true} \\
-                                            \stmt{TAPE isOn TABLE}};
-
-        \node[align=left ] at (cam3 -| kbh) (kb3) {\stmt{TAPE isReachable true}\\
-                                             \stmt{TAPE isVisible true}};
-
-        \node[align=left] at (cam4 -| kbh) (kb4) {\stmt{HUMAN hasInHand TAPE}};
-
-        \node[align=left] at (cam5 -| kbr)  {\stmt{TAPE isIn BIN}};
-        \node[align=left] at (cam5 -| kbh) (kb5) {\stmt{TAPE isIn BIN}};
+        \node[fill=white,align=left] at (cam4 -| kbh) (kb4) {\stmt{HUMAN hasInHand TAPE}};
+
+        \node[fill=white,align=left] at (cam5 -| kbr)  {\stmt{TAPE isIn BIN}};
+        \node[fill=white,align=left] at (cam5 -| kbh) (kb5) {\stmt{TAPE isIn BIN}};
 
@@
 
-        \node[below=1 of plan] (incoming) {\bf \Large Incoming goal \\ \it Clean the table!};
-
-        \node[anchor=north, plan, robot] at (cam1.south -| probot) (pr1) {\action{TAKE}{GREY\_TAPE}{TABLE}};
-        \node[anchor=north, plan, robot] at (cam2.south -| probot)  (pr2) {\action{PUTRV}{GREY\_TAPE}{TABLE}} edge[<-] (pr1);
-        \node[anchor=north, plan] at (cam3.south -| phuman) (ph1) {\action{TAKE}{GREY\_TAPE}{TABLE}} edge[<-] (pr2);
-        \node[anchor=north, plan]  at (cam4.south -| phuman) (ph2) {\action{PUT}{GREY\_TAPE}{BIN}} edge[<-] (ph1);
-
-        \node[anchor=north] at (cam5.south -| plan) (done) {\bf \Large Goal completed};
+        \node[fill=gray!10!white,below=1 of plan] (incoming) {\bf \Large Incoming goal \\ \it Clean the table!};
+
+        \node[anchor=north, plan, robot] at (cam1.south -| probot) (pr1) {\action{TAKE}{TAPE}{TABLE}};
+        \node[anchor=north, plan, robot] at (cam2.south -| probot)  (pr2) {\action{PUTRV}{TAPE}{TABLE}} edge[<-] (pr1);
+        \node[anchor=north, plan] at (cam3.south -| phuman) (ph1) {\action{TAKE}{TAPE}{TABLE}} edge[<-] (pr2);
+        \node[anchor=north, plan]  at (cam4.south -| phuman) (ph2) {\action{PUT}{TAPE}{BIN}} edge[<-] (ph1);
+
+        \node[fill=gray!10!white,anchor=north] at (cam5.south -| plan) (done) {\bf \Large Goal completed};
 
@@
 
-        \node at (pr1 -| arobot) (ep1) {\it evaluate pre-conditions};
+        \node[fill=white] at (pr1 -| arobot) (ep1) {\it evaluate pre-conditions};
 
@@
             \begin{tikzpicture}
-                \node (title) {\bf motion planning};
+                \node[fill=white] (title) {\bf motion planning};
                 \node[below=0.1 of title.south west, label=below:{\tt PICK\_GOTO}] (mapg) {\includegraphics{MHP_ARM_PICK_GOTO}};
-                \node[right=of mapg, label=below:{\tt TAKE\_TO\_FREE}] (mattf) {\includegraphics{MHP_ARM_TAKE_TO_FREE}} edge[<-] (mapg);
+                \node[fill=white,right=of mapg, label=below:{\tt TAKE\_TO\_FREE}] (mattf) {\includegraphics{MHP_ARM_TAKE_TO_FREE}} edge[<-] (mapg);
             \end{tikzpicture}
         };
-        \node[below=0.1 of mhp1] (me1) {\bf motion execution};
-        \node[below=0.1 of me1] (ae1) {\it assess post-conditions};
+        \node[fill=white,below=0.1 of mhp1] (me1) {\bf motion execution};
+        \node[fill=white,below=0.1 of me1] (ae1) {\it assess post-conditions};
 
         %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-        \node at (pr2 -| arobot) (ep2) {\it evaluate pre-conditions};
+        \node[fill=white] at (pr2 -| arobot) (ep2) {\it evaluate pre-conditions};
 
@@
             \begin{tikzpicture}
-                \node (title) {\bf motion planning};
-                \node[below=0.1 of title, label=below:{\tt ESCAPE}] (maeo) {\includegraphics{MHP_ARM_ESCAPE_OBJECT}};
+                \node[fill=white] (title) {\bf motion planning};
+                \node[fill=white,below=0.1 of title, label=below:{\tt ESCAPE}] (maeo) {\includegraphics{MHP_ARM_ESCAPE_OBJECT}};
                 \node[left=of maeo, label=below:{\tt PLACE\_FROM\_FREE}] (mapff) {\includegraphics{MHP_ARM_PLACE_FROM_FREE}} edge (maeo);
-                \node[right=of maeo, label=below:{\tt TO\_FREE}] (maf) {\includegraphics{MHP_ARM_FREE}} edge[<-] (maeo);
+                \node[fill=white,right=of maeo, label=below:{\tt TO\_FREE}] (maf) {\includegraphics{MHP_ARM_FREE}} edge[<-] (maeo);
             \end{tikzpicture}
         };
-        \node[below=0.1 of mhp2] (me2) {\bf motion execution};
-        \node[below=0.1 of me2] (ae2) {\it assess post-conditions};
+        \node[fill=white,below=0.1 of mhp2] (me2) {\bf motion execution};
+        \node[fill=white,below=0.1 of me2] (ae2) {\it assess post-conditions};
 
@@
             \begin{tikzpicture}
-                \node (title) {\it wait for pick\\ {\tt TAPE} from {\tt TABLE}};
+                \node[fill=white] (title) {\it wait for pick\\ {\tt TAPE} \it from {\tt TABLE}};
                 \node[below=0.1 of title] {\includegraphics{wait_for_pick}};
@@
             \begin{tikzpicture}
-                \node (title) {\it wait for put\\ {\tt TAPE} to {\tt BIN}};
+                \node[fill=white] (title) {\it wait for put\\ {\tt TAPE} \it into {\tt BIN}};
                 \node[below=0.1 of title] {\includegraphics{wait_for_throw}};
@@
 
- \caption {Chronology of the \emph{Cleaning the table} scenario, presented here in a simplified
+ \caption {Timeline of the \emph{Cleaning the table} study, presented here in a simplified
  version: a single object, {\tt TAPE}, must be removed from the table and

### 2016-06-16 paper.tex
Even more polish on the timeline figure
Added a summary of the paper

@@
 
-        \path (-0.5,0) edge (figbottom) node[sloped, above left, rotate=90] {time};
-
         \fill[gray!10!white] (4.6,.5) rectangle (figbottom);
+
+        \path (-0.5,0) edge (figbottom) node[sloped, above left, rotate=90] {\large\bf time};
+
         \node at (2,0) (percept) {\bf Perception};
@@
 
-        \fill[gray!10!white] (11.5,.5) rectangle (16.5,-28.5);
-        \node[right=4 of kb] (plan) {\bf Plan};
+        \fill[gray!10!white] (12,.5) rectangle (17,-28.5);
+        \node[right=4.5 of kb] (plan) {\bf Plan};
             \node[below=0.5 of plan.south west, anchor=mid east] (probot) {robot};
@@
 
+\section{Conclusion}
+\label{sect|conclusion}
+
+\subsection{Summary of the Contribution}
+
+We have presented in this paper a large deliberative architecture designed for
+social robots. While most of its sub-components have been independently
+presented in other publications, we offer here for the first time a perspective
+on the overall integration of these components into a coherent and consistent
+system for social human-robot interaction. We have first exposed our underlying
+knowledge model based on Description Logics~\cite{Lemaignan2010} and some of the
+resulting reasoning capabilities pertaining to disambiguation~\cite{Ros2010b}
+and mental modelling~\cite{warnier2012when} that are shown to effectively
+scaffold interaction using human-level semantics and cognitive skills. We have
+then presented our approach to symbol grounding, build on an amodal situation
+assessment environment~\cite{Sisbot2011} that supports perspective
+taking~\cite{Marin2008,Ros2010}. We
+combine it further with an advanced natural
+language processor~\cite{Lemaignan2011a} to provide complete multi-modal interactive
+communication.  The paper also covers our symbolic social task
+planner~\cite{Alili2008, Alili2009, Lallement2014}: it generates predictive
+plans of the human actions that enable the system to plan for joint human-robot
+tasks. It can also make use of social heuristics to optimize plans for social
+acceptability. We briefly mention our human-aware motion and manipulation
+planner~\cite{Sisbot2008,Mainprice2011,Pandey2011}, and finally present two
+execution controller, the PRS-based {\sc Shary}~\cite{clodic2008shary,fiore2014} and the
+event-driven {\sc pyRobots}~\cite{lemaignan2015pyrobots}.
+
+The integration of these components in a consistent, working and observable
+system is supported by the design of our software interfaces: most of the
+datastreams exchanged by our components uses high-level semantics in the form of
+first-order logic statements. The two reported support studies further evidence
+how our use of explicit, high-level semantics provide a low-impedance,
+human-level mean of building a complex deliberative
+architecture for socially interactive robots~\cite{Lemaignan2013}.
+
+\subsection{A Deliberative Architecture for Social Robots}
+
+Rachid -- TDB: what are the other 'deliberative architecture for socially interactive
+robots' and where/how ours is better?
+
+Discussion on deliberative architectures != cognitive architecture: a \emph{cognitive
+architecture} is usually understood as an artificial yet principled model of (human)
+cognition, while we propose a practical approach to consistently combine a large
+set of independent cognitive processes by bridging them through high-level semantics.
+
 \subsection{The Next Steps}

### 2016-06-16 paper.tex
Fixing language 7

@@
 knowledge model that we have developed for our robots. Then,
-Section~\ref{sec:impl} details each of the main components of the
-architecture, and highlights their significance for artificial
-intelligence. Section~\ref{sec:expe} presents two studies that illustrate in a
-practical way what can be currently achieved with our robots.
-Section~\ref{sect|conclusion} finally restate our main contributions and discuss
-the key challenges that human-robot interaction brings to artificial
-intelligence.
+Section~\ref{sec:impl} details each of the main components of the architecture,
+and highlights their significance for artificial intelligence.
+Section~\ref{sec:expe} presents two studies that illustrate in a practical way
+what can be currently achieved with our robots.  The
+Sections~\ref{sect|discussion} and~\ref{sect|conclusion} finally restate our
+main contributions and discuss the key challenges that human-robot interaction
+brings to artificial intelligence.
 
@@
 
-\subsection{\emph{Cleaning the table} study}
+\subsection{\emph{Cleaning the table} Study}
 
@@
 Interaction}
-\label{sect|conclusion}
-
-This last section tries to clarify the ``takeaway'' message: what are the
-challenges that human-robot interaction brings to artificial intelligence?  We
-first restate the challenge of \emph{embodied cognition} in general and
-summarise what we see as the main decisional issues that need to be tackled to
-push forward human-robot interaction. Finally, we reflect on the importance of
-knowledge handling in robotic architectures that deal with complex semantics and
-put into perspective our choice of the RDF formalism.
+\label{sect|discussion}
+
+The previous sections provide a perspective on a complete deliberative
+architecture for social robots, including an implementation supported
+``real-world'' experimental deployments. This section synthesises next what we
+see as the main challenges that human-robot interaction brings to artificial
+intelligence.  We discuss first how embodied cognition is a very real challenge
+in human-robot interaction; we rephrase then the requirements of joint actions
+in terms of five questions; we discuss the importance of building and
+maintaining a multi-level model of the human; and we finally reflect on the
+importance of explicit knowledge management in robotic architectures that deal
+with human-level semantics and state in that respect the current limits of our
+logic framework.
 
@@
 \emph{embodied} artificial intelligence, and this dimension is especially
-prevalent in human-robot interaction, where agents have to share and collaborate
-in a joint physical environment.
-
-This leads to a tight coupling between the symbolic and the geometric realms:
-while AI at its origins was mostly a matter of symbolic models, it has been
-since recognised that not only the mind is not a purely abstract system,
-disconnected from the physical world, but even more, cognition fundamentally
-relies on its relation to the physical world (so-called \emph{embodied
-cognition}). Varela~\cite{Varela1992} is one of the main discoverers of these
-mechanisms, and coined the concept of \emph{enactivism} as the theoretical
-framework that study the links between cognition, embodiment and actions.  This
-has since been thoroughly studied in robotics and artificial intelligence
-(Pfeifer and Bongard~\cite{pfeifer2007body} is one of the reference).
+prevalent in human-robot interaction, where agents have to share and
+collaborate in a joint physical environment.  This leads to a tight coupling
+between the symbolic and the geometric realms: while AI at its origins was
+mostly concerned with symbolic models, it has been since recognised that not
+only the mind is not a purely abstract system, disconnected from the physical
+world, but even more, cognition fundamentally relies on its relation to the
+physical world (so-called \emph{embodied cognition}). Varela~\cite{Varela1992}
+is one of the main discoverers of these mechanisms, and coined the concept of
+\emph{enactivism} as the theoretical framework that study the links between
+cognition, embodiment and actions.  This has since been thoroughly studied in
+robotics and artificial intelligence (Pfeifer and
+Bongard~\cite{pfeifer2007body} is one of the reference).
 
@@
 
-\subsection{The Key Decisional Issues}
-
-With this perspective on \emph{embodied cognition} in mind, it seems appropriate
-to restate the key challenges brought by human-robot interaction to artificial
-intelligence.
-
-\subsubsection{The W-questions}
+\subsection{The W-questions of Joint Action}
 
@@
 
-\subsubsection{Putting the Humans into Equations}
+\subsection{Putting the Humans into Equations}
 
@@
 
-As thorougly presented in this article, we have built the decisional
+As thoroughly presented in this article, we have built the decisional
 capabilities of our robots around this idea of explicit knowledge manipulation.
-We finally briefely comment on this design choice, our implementation based on RDF,
-and the limitations that we perceive.
+We finally briefly comment on this design choice and the limitations that we
+have observed.
 
@@
 
-One prototypical example of semantic disambiguation has been given in
-\cite{Ros2010b} with the child game \emph{spygame}: two players are facing
+We have studied a prototypical example of semantic disambiguation in
+\cite{Ros2010b} with the children's ``spygame'': two players are facing
 each other with a set of random objects in-between, one player mentally choose
@@
 implement it in the case of a personal robot assistant essentially
-devoted to fetch-and-carry, as well as for interactive manipulation
+devoted interactive manipulation
 tasks and associated activities.
@@
 
-Another broader direction to head deals with \emph{contexts representation}.
+Another direction to head to deals with \emph{contexts representation}.
 Contexts are currently often limited to the current spatial and temporal
-situation. Some models offer the possibility to jump in the past or to
-switch to another agent's perspective, but in current approaches, selecting a
-context always basically consists in retrieving a set of beliefs corresponding
-to a situation, and temporarily replacing the current beliefs by those other
-ones. This misses the fact that at a given moment, not one but many contexts
-co-exist at different scales. We do not want to retrieve one monolithic set of
-beliefs, but instead carefully craft a context from several \emph{atomic}
-contexts. Techniques for representation of overlapping pools of knowledge
-largely remain to be developed, as well as efficient algorithms to retrieve (or
-discard) such context-related pools of knowledge. This is a challenge not only
-for robotics, but more generally for artificial intelligence.
-
-The ability to explicitly manage contexts and context switches would endow the
-robot with a cognitive capability similar to what is known as
-\emph{context-dependent memory} in cognitive psychology. This is also related
-to Tulving's \emph{autonoetic consciousness}~\cite{Tulving1985a}: the ability
-to reflect upon its own past or future experiences.
-
-Much remain to be done to this regard, starting with a formal analysis of what
-are the relevant contexts for our robots.
+situation. Some models offer the possibility to jump in the past or to switch to
+another agent's perspective, but in current approaches, selecting a context
+always basically consists in retrieving a set of beliefs corresponding to a
+situation, and temporarily replacing the current beliefs by those other ones.
+This misses the fact that at a given moment, not one but many contexts co-exist
+at different scales. We do not want to retrieve one monolithic set of beliefs,
+but instead carefully craft a context from several \emph{atomic} contexts.
+Techniques for representation of overlapping pools of knowledge largely remain
+to be developed, as well as efficient algorithms to retrieve (or discard) such
+context-related pools of knowledge. This is a challenge not only for robotics,
+but more generally for artificial intelligence.  The ability to explicitly
+manage contexts and context switches would endow the robot with a cognitive
+capability similar to what is known as \emph{context-dependent memory} in
+cognitive psychology. This is also related to Tulving's \emph{autonoetic
+consciousness}~\cite{Tulving1985a}: the ability to reflect upon its own past or
+future experiences.  Much remain to be done to this regard, starting with a
+formal analysis of what are the relevant contexts for our robots.
 

### 2016-06-16 paper.tex
Fixing language 8

@@
 between the symbol (the syntactic form of knowledge that the computer
-manipulates) and its semantics, \ie its meaning, anchored in the world (the
+manipulates) and its semantics (its meaning, anchored in the world). The
 relations between the symbol, the referent of the symbol, and mediating minds
 is classically referred as the \emph{semantic triangle} and has been
-extensively studied in linguistics). The issue of grounding is well known in
+extensively studied in linguistics. The issue of grounding is well known in
 cognitive science and is summarised by Harnard~\cite{Harnad1990} by this
@@
 questions until it can attach the new concept to concepts already present in the
-knowledge pool (typically, if I ask the robot ``bring me a platypus'', the robot
-asks ``what is a platypus''. I may answer ``a mammal'' and the robot will ask
-``what is a mammal?'' and so on until it anchors the new concepts to ones it
-already knows). Embodied interactions (like gestures) are also taken into
-account at this level: we have presented how pointing, for instance, is used by
-the robot to ground \emph{``this''} or \emph{``that''} to the pointed artifact.
-
-Note that because only objects marked with 2D fiducial markers are currently
-recognised (in our typical lab experiments, it may be a dozen of them), the
-robot's grounding mechanisms only work in a small-sized closed world. This
-assumption greatly facilitates the task, and we see that we have only scratched
-the surface of a generic grounding capability. Context, cultural background,
-``naive physics'' knowledge, emotional state of the human are some of the
-numerous other determinants that need to be accounted for when grounding
-human-robot interaction.
+knowledge pool (if one asks the robot ``bring me a margarita'', the robot may
+initially asks ``what is a margarita?''. The user would answer ``a cocktail''
+and the robot would continue the grounding -- ``what is a cocktail?'' -- until
+it anchors the new concept to ones it already knows). Embodied interactions
+(like gestures) are also taken into account at this level: we have presented how
+pointing, for instance, is used by the robot to ground \emph{``this''} or
+\emph{``that''} to the pointed artifact.
+
+Note that, because only objects marked with 2D fiducial markers are currently
+recognised (typically, about ten of them are simultaneously used in a given
+experiment), our grounding mechanisms have only been exercised in small-sized
+closed world. This simplifies the task, and we can not claim that our approach
+provides a generic grounding capability. Beyond the perception of geometric
+entities, context, cultural background, ``naive physics'' knowledge, emotional
+state of the human are some of the numerous other determinants that need to be
+accounted for when grounding human-robot interaction.
 
@@
 collaborative human-robot objects manipulation, fetch-and-carry and associated
-activities in home or work environment.
-
-In our implementation, perspective taking, for instance, is tightly connected to
+activities in home or work environments.
+
+In our architecture, perspective taking, for instance, is tightly connected to
 the symbolic knowledge models, and since our knowledge base allows for storage
@@
 (\emph{who sees what? what are our relative positions? what is reachable to
-whom?}  etc.). Extending this knowledge with more subtle perceptions (emotional
-state for instance) remains to be done.
+whom?}). Extending this knowledge with more subtle perceptions (emotional
+state for instance) remains to be explored beyond simple examples like the
+processing of explicit verbal statements like ``I'm tired!''
+(Section~\ref{sect:desires}).
 
@@
 
-\subsection{Knowledge and Robotics}
+\subsection{Explicit Knowledge for Social Robotics}
 \label{krs-discussion}
@@
 capabilities of our robots around this idea of explicit knowledge manipulation.
-We finally briefly comment on this design choice and the limitations that we
-have observed.
-
-\subsubsection{Knowledge and Architecture}
-
-As a whole, the components that we have introduced so far build a
-knowledge-oriented architecture: knowledge is explicitly stored in one central
-and consistent repository of facts, accessible for all modules. It relies on a
-strict formalism (OWL statements), with a well defined vocabulary
-(stated in the common-sense ontology). These first two points lead to a
-\emph{loosely-coupled architecture} where modules can be removed or replaced
-easily by other ones as long as they share the same semantics (modules are
-defined by the knowledge they produce).
-
-Also, we adopt a symbolic, reactive, event-driven approach to robot control. By
-managing events at the same level as the reasoner, we take full advantage of the
-inference abilities of {\sc Oro} to trigger events whose \texttt{true}
-conditions can be (possibly indirectly) inferred.
-
-And finally, this architecture allows for the combination of very different
-knowledge sources in a single homogeneous environment, bringing mutual
+We finally briefly comment on this design choice and the limitations of our
+approach.
+
+\subsubsection{Explicit Knowledge in our Architecture}
+
+The components that we have presented so far build a \emph{knowledge-oriented}
+architecture: knowledge is explicitly stored in one central and consistent
+repository of facts, accessible for all modules. It relies on a strict formalism
+(OWL statements), with a well defined vocabulary (stated in the common-sense
+ontology). These first two points lead to a loosely-coupled architecture
+where modules can be removed or replaced easily by other ones as long as they
+share the same semantics: modules are defined by the knowledge they produce or
+consume.
+
+Also, we adopt a symbolic, hybrid (reactive and planning-based), event-driven
+approach to robot control. By managing events at the same level as the reasoner,
+we take full advantage of the inference abilities of {\sc Oro} to trigger events
+whose \texttt{true} conditions can be (possibly indirectly) inferred using
+human-level semantics.
+
+And finally, this architecture allows for the combination of different
+knowledge sources in a uniform model, bringing mutual
 benefits to components. For instance, the dialogue processing module can
@@
 
-Those items are not new \emph{per se}. It is however interesting to underline
-the shift of focus this implies during the design and integration phases of
-robots: components of our deliberative layer are defined and glued together by
-the knowledge they produce and consume. Human-robot interaction, because it
+Those characteristics are not novel \emph{per se}. We want however underline the
+shift of focus brought by this approach during the design and integration phases
+of robots: components of our deliberative layer are defined and bound together
+by the knowledge they produce and consume. Human-robot interaction, because it
 supposes operations at \emph{human-level} and in environments with complex

### 2016-06-16 rebuttal.tex

@@
 
-\section{General Comments}
+Dear Editors, Dear Reviewers,
 
-\begin{itemize}
+\vspace{2em}
 
-    \item \emph{The form and depth of the presentation are below the standards of an
-AIJ article.  The reviewers made many suggestions to make the presentation
-better structured, more detailed, and more focused.}
-
-We would like first to thank the reviewers for their thoughtful review of our
+We would like first to thank the reviewers for their thorough review of our
 work. The manuscript underwent a major rewrite, impacting both the overall style
-(which should now be much more pleasant to read) and the whole organisation of
+(which should now be of much higher quality) and the whole organisation of
 the paper (following the reviewers' recommendations). We also dedicated time to
@@
 
-We address hereafter each of the specific concerns raised by the reviewers (note
-that minor typos/style issues have been fixed directly in the manuscript, and
-are not discussed here).
+The novelty of this contribution has been questioned by one of the reviewers.
+We hopefully address this concern hereafter, along with the other comments of
+Reviewer 3.
 
-\item \emph{The novelty of this paper with respect to other papers by the same
-authors, including one in IJSR 2011, has been questioned. The authors should
-convincingly explain what are the novel contributions and the added value of
-this work compared to the previous publications.}
-
-We address this question with our reponse to reviewer 3, below.
-
-\end{itemize}
+We have as well attempted to address individually each of the other concerns
+raised by the reviewers in this document (note that minor typos/style issues
+have been fixed directly in the manuscript, and are not discussed here).
 
@@
 to maintain a quasi-global world belief?} This is a design choice that trades
-off scalability for complete domain explicitation. We have clarified and
+off scalability for an explicit domain. We have clarified and
 justified this design choice in section 2.2 (\emph{Knowledge model}). Also, in
@@
 that can typically not be computed ``on demand'' because they are observable
-only in specific situations.
+only at specific times.
 
@@
         algorithms are detailed in \cite{Ros2010b} (Best Paper Award at
-        ROMAN2010). We have added a reference to this paper.
+        ROMAN2010). We now briefly mention them in the paper (p.7), and we have added a
+        reference to this paper.
+
+    \item \emph{p11 l 42-44: These concepts like "feeling uncomfortable" or "could be
+      misinterpreted" are very complex and consist of multiple facets. Please
+        explain which of these facets are supported by the system and how this
+          information (and the classifier determining this information) is
+          derived.
+            I would expect the real difficulty to be how to determine these
+            aspects in
+              a nontrivial way, rather than their combination by considering
+              them as "yet
+          another cost function".}
+
+          Section 3.4.2 attempts to better explain these techniques, based on
+          \emph{a posteriori} filtering. The application of these rules is
+          currently static: the execution controller calls the planner with
+          pre-defined parameters. As noted by the reviewer, the online
+          adaptation of these parameters, while technically straightforward,
+          implies a level of interaction management that we have yet to
+          explore.
 
@@
 once it has changed are crucial. This should be explained in more detail.}
-The dynamics of the knowledge are managed by the clients (ie, the deliberative
-components like the situation assessment module, the dialogue processing module,
-etc.) of the knowledge base themselves. In particular, the situation assessment
-module recomputes at every step the whole set of spatial relations, and update
-accordingly the knowledge base. This has been clarified in both section 2.2
-(\emph{Knowledge model}) and section 3.2 (\emph{Situation Assessment, Human
-Awareness, Perspective Taking})
+The dynamics of the knowledge are managed by the clients  of the knowledge base
+themselves (ie, the deliberative components like the situation assessment
+module, the dialogue processing module, etc.). In particular, the situation
+assessment module recomputes at every step the whole set of spatial relations,
+and update accordingly the knowledge base. This has been clarified in both
+section 2.2 (\emph{Knowledge model}) and section 3.2 (\emph{Situation
+Assessment, Human Awareness, Perspective Taking})
+
+    \item \emph{The paragraphs in Section 3.4.3 appear to be rather
+        disconnected, some effort should be made to turn this into a readable
+        text and to explain design choices like the 3-tuple representation or the
+        gripper states that just drop out of the blue.}
+
+        This section (now 3.5) has been reworked, and should now be
+        self-explanatory.
 
@@
         \emph{code names} for the experiments' names. These have been reworded.
+        Besides, we have replaced the terminology \emph{Experiments} by
+        \emph{Case studies} to better reflect the nature of these results.
 
@@
         reference to KnowRob TBOX, that originated in a discussion with KnowRob's author who told
-        us then that KnowRob could not update at run-time its TBox. Since we were
-        not able to find litterature confirming this fact, it is likely better
-        not to mention it.
+        us then that KnowRob could not update at run-time its TBox. We were
+        however        not able to find literature confirming this fact, and we
+        have removed this reference.
 
@@
         the user to follow the execution flow.
-    \item the discussion section has been entirely reorganised, following the
-        Reviewer recommendation. The key decisional issues have been explicitly
+    \item the discussion section has been entirely reorganised and refocused, following the
+        Reviewer recommendations. The key decisional issues have been explicitly
         stated, and overlaps between subsections have been addressed.
@@
 
+    \item \emph{introduction "This article attempts however to organize them
+            (i.e. the AI challenges
+            in HRI) into a coherent challenge for artificial intelligence, and
+            also to make explicit 
+            some of the paths that have been investigated on our robots and that
+            result in a deliberative, 
+            knowledge-oriented, architecture for human-robot interaction." This
+            suggests that the
+            architecture plays a central role, but the contribution in terms of
+            architecture is not 
+        the most interesting one.}
+
+        We have accordingly rephrased this paragraph to make clear that our
+        contribution focuses on a set of deliberative components that interact
+        to support human-robot interaction.
+
+    \item \emph{pge 4 (section 2)
+            "execution controller that may decide to take an incoming desire
+            into account to create
+            its own internal goals." The role of the execution controller
+            remains unclear (discussed
+        also in subsection 3.4.3).}
+
+            Section 2.2 (Knowledge Model) has been largely rephrased, and
+            hopefully clarifies the role of the execution controller.
+
+    \item \emph{"Aligning the upper part of the ontology on OpenCyc ..." In my
+            view the requirement of
+            compatibility with other ontologies is not so critical. What would
+            be really nice are
+            ontologies (or techniques for extracting form general resources)
+            that capture
+        the common sense knowledge that is essential for robot operation.}
+
+            We agree with the reviewer. While, until now, our focus has not been on
+            merging external sources of common-sense knowledge with the robot's
+            knowledge model, we believe that by initially relying on de-facto
+            standards like OpenCYC ensures to some extent that we will be able
+            to do so in the coming years.
+
     \item \emph{pg. 7 (Section 3.1) ``From an artificial intelligence
@@
             and performs spatial perspective-taking.'' given on this
-        characterization, Spark should be discussed in greater detail.} We have
+        characterization, Spark should be discussed in greater detail.}
+
+        We have
         significantly developed the presentation of {\sc Spark}, in
@@
             nice, but are such rules compatible with the existing knowledge?
-            Adding them as triples may work only in absence of contradictions} The
-            various natural language grounding processes that are made possible
-            within this semantic
-            architecture would require a relatively long discussion to be adequately
-            explained. They are besides already presented in~\cite{Lemaignan2011a}.
-            We have added an explicit reference to this previous publication.
+            Adding them as triples may work only in absence of contradictions}
+
+            The various natural language grounding processes that are made
+            possible within this semantic architecture would require a
+            relatively long discussion to be adequately explained. They are
+            besides already presented in~\cite{Lemaignan2011a}.  We have added
+            an explicit reference to this previous publication.
 
             To specifically answer the reviewer question on how possible
-            contradictions are managed, we indeed check that new statements proposed
-            by the human do not lead to contradictions before inserting them.
+            contradictions are managed, we indeed check that, \emph{when
+            proposed by humans}, new statements
+            do not lead to contradictions before inserting
+            them.
 
@@
             at the cost of a simpler interaction."
-        Possibly true, but unjustified claim.} An example has been added to
+        Possibly true, but unjustified claim.}
+
+        An example has been added to
         clarify this fall-back mechanism.
+
+    \item \emph{pg. 10-11 section 3.3
+            HATP planner is another interesting contribution that could be
+            further detailed for the AIJ reader. 
+            It is unclear why for planning a representation "outside the central
+            knowledge base" is needed. What is then the relationship between the
+            knowledge in the two distinct KBs?
+            See also the sentence "Hence only a few ontology rules have been
+            required to map both the  knowledge extracted from the situation assessment and the statements
+            originated from the verbal  interaction to the planner domain." ( pg. 12)}
+
+            The section on HATP (section 3.3) has been rewritten and extended to
+            better reflect the capabilities of this planner. References to
+            recent work on this tool have been added as well.  The paragraph
+            that discuss the issue of procedural vs declarative knowledge has
+            been rephrased as well.
 
@@
             discrimination transparently benefit from added modalities." Concept
-            remains unclear.} The paragraph has been rephrased and the example
+            remains unclear.}
+
+            The paragraph has been rephrased and the example
             updated to clarify the concepts.
@@
             future perspectives. In that case it would be nice to read which AI
-            techniques could help tackling the issue.} The presentation of
+            techniques could help tackling the issue.}
+
+            The presentation of
             ``internal cognitive skills'' has been extended, refined and integrated
-            to section 3. Specifically, several litterature references have been
-            added to the discussion on memory, both on the theoritical side and on
+            to section 3. Specifically, several literature references have been
+            added to the discussion on memory, both on the theoretical side and on
             existing implementations.
@@
             understandable by the robot at run-time'' This is very difficult to
-            understand. Provide more details and, possibly, an example} The paragraph
+            understand. Provide more details and, possibly, an example}
+
+            The paragraph
             has been rephrased and illustrated to be easier to understand.
@@
 to appear in the proceedings of the \emph{2011 International Symposium on
-Robotics Research}, but has yet to be actually pusblished.  While this article
+Robotics Research}, but has yet to be actually published.  While this article
 indeed presents our architecture in a way similar to our proposed contribution,
@@
 
+\bibliographystyle{abbrv}
+\bibliography{laas_hri}
+
 \end{document}

### 2016-06-16 paper.tex
Implemented comments from reviewer 1

@@
 
-Human-robot interaction is a challenge for artificial intelligence. This field
+Human-robot interaction is a challenge for Artificial Intelligence. This field
 lays at the crossroad of several domains of AI and requires to tackle them in a
@@
 We do not claim to address here the issue as a whole. This article attempts
-however to organise it into a coherent challenge for artificial intelligence,
+however to organise it into a coherent challenge for Artificial Intelligence,
 and to explain and illustrate some of the paths that we have investigated on our
@@
 perspectives, skills of the human partner.  Namely, the robot must be able to
-recognise, understand and participate to communication situations, both explicit
+recognise, understand and participate in communication situations, both explicit
 (\eg the human addresses verbally the robot) and implicit (\eg the human points
-to an object); the robot must be able to take part to joint actions, both
+to an object); the robot must be able to take part in joint actions, both
 pro-actively (by planning and proposing resulting plans to the human) and
@@
 
-\subsection{Article overview}
+\subsection{Article Overview}
 
@@
 tackle the AI challenges of human-robot interaction, and how its relates to
-other approaches. We propose to organise this discussion in four sections.
+other approaches. We organise this discussion in four sections.
 
@@
 Section~\ref{sec:impl} details each of the main components of the architecture,
-and highlights their significance for artificial intelligence.
+and highlights their significance for Artificial Intelligence.
 Section~\ref{sec:expe} presents two studies that illustrate in a practical way
@@
 main contributions and discuss the key challenges that human-robot interaction
-brings to artificial intelligence.
+brings to Artificial Intelligence.
 
@@
 be objects (the Earth for instance). Thus the importance of relying on
-well-defined and standard semantics to exchange informations between artificial
+well-defined and standard semantics to exchange information between artificial
 systems in a non-ambiguous way.
@@
 If a human informs the robot that a given object is indeed a bottle, the robot
-derives more informations on the object. Conversely, if the human affirms that
+derives more information on the object. Conversely, if the human affirms that
 ``a car is a bottle'', the reasoner may detect logical contradictions (like
@@
 
-\subsubsection{Working memory}
+\subsubsection{Working Memory}
 \label{memory}
@@
 of cognitive architectures, the {\sc Soar} architecture~\cite{Lehman2006} is one
-of those that tries to reproduce a human-like memory organisation. The GLAIR
+of those that try to reproduce a human-like memory organisation. The GLAIR
 cognitive architecture~\cite{Shapiro2009} also has a concept of long term/short
@@
 
-It is worth emphasising that if memory is commonly associated to the process of
+It is worth emphasising that while memory is commonly associated with the process of
 forgetting facts after a variable amount of \emph{time}, it actually covers
@@
 inserting statements such as \stmt{human type ActiveConcept} in the short-term
-memory (which can be considered, in this case, to be a working memory). We use
-this feature during dialogue disambiguation for instance. On the other hand,
-our perception layer does not make use of this mechanism: as described in the
-next section, the environment model of the robot is continuously updated and the
-derived symbolic knowledge is therefore transient: it lasts only as long as the
-environment remains in the same state.
+memory (which can be considered, in this case, to be a working memory).
+Likewise, recently seen or updated geometric entities are flagged as
+\concept{ActiveConcept}. We use this feature during dialogue disambiguation for
+instance. On the other hand, our perception layer does not make use of this
+mechanism: as described in the next section, the environment model of the robot
+is continuously updated and the derived symbolic knowledge is therefore
+transient: it lasts only as long as the environment remains in the same state.
 
@@
 \paragraph{Perspective Taking} \emph{Visual} perspective taking refers to the
-ability for visually perceiving the environment from other's point of view.
+ability for visually perceiving the environment from another person's point of view.
 This ability allows us to identify the objects in situations where the visual
@@
 other is much larger than the maximum distance we would consider for two small
-objects (\eg two bottles). Thus, the relation between the dimensions and the
-distances of the objects are taken into account.  {\sc Spark} also compute
-symbolic facts related to agent independent world dynamics.  The predicate
-\concept{isMoving} states, for each tracked entity, whether it is currently
-moving or not.
+objects (\eg two bottles). Thus, the distance threshold is scaled with the
+objects' size. {\sc Spark} also compute symbolic facts related to agent
+independent world dynamics.  The predicate \concept{isMoving} states, for each
+tracked entity, whether it is currently moving or not.
 
@@
     location of the eyes and running occlusion checks from it.  In
-    figure~\ref{fig|spark-oro}\emph{c} the field of view of a person is
+    Figure~\ref{fig|spark-oro}\emph{c} the field of view of a person is
     illustrated with a grey cone (the wider one). While he is able to see the two
@@
 and activity recognition is a task that requires knowledge and reasoning both on
-high level facts like goals, intentions and plans, as well as bottom-up data
+high-level facts like goals, intentions and plans, as well as bottom-up data
 from agent and object motions. {\sc Spark} implements a set of simple temporal and
@@
 robot intents). Natural language processing is facilitated by our architecture
-mostly processing informations with semantics that are already close to the
+mostly processing information with semantics that are already close to the
 human-level. This section presents the main features of our speech processor,
@@
 actions.  Sentences are classified into questions, desires and statements, and
-processed accordingly.  Figure~\ref{dialogs|ex} gives a example of the
+processed accordingly.  Figure~\ref{dialogs|ex} gives an example of the
 processing of a simple, non-ambiguous command.
@@
 tracking what the human points at, no \stmt{human\_1 pointsAt ...} fact would
-possibly be available in the knowledge base, and the robot fallbacks on the
+possibly be available in the knowledge base, and the robot falls back to the
 anaphora resolution step alone.
@@
 function provides a duration interval for the action achievement and is used, in
-one hand, to schedule the different streams and, in the other hand, as an
+one hand, to schedule the different streams and, on the other hand, as an
 additional cost function.
@@
 ``real-world'' experimental deployments. This section synthesises next what we
-see as the main challenges that human-robot interaction brings to artificial
-intelligence.  We discuss first how embodied cognition is a very real challenge
+see as the main challenges that human-robot interaction brings to Artificial
+Intelligence.  We discuss first how embodied cognition is a very real challenge
 in human-robot interaction; we rephrase then the requirements of joint actions
@@
 manipulates) and its semantics (its meaning, anchored in the world). The
-relations between the symbol, the referent of the symbol, and mediating minds
+relation between the symbol, the referent of the symbol, and mediating minds
 is classically referred as the \emph{semantic triangle} and has been
@@
 
-\emph{When to perform a given action} is enriched here by the human-context
+\emph{When to perform a given action} is enriched here by the human context
 since it has to take into account the human, his/her needs, his/her rhythms and
@@
 The correlate of these five \emph{W-questions} is the issue the \emph{human
-models}: taking appropriate decisions with and in presence of humans require
-appropriate models of the human. Models that tell the robot what the human
-\emph{can do}, \emph{would like to do}, \emph{knows}, \emph{could infer}, etc.
-
-While the task of describing all the (dynamic!) human models that are useful to
+models}: taking appropriate decisions with and in presence of humans requires
+appropriate models of the human: what the human \emph{can do}, \emph{would like
+to do}, \emph{knows}, \emph{could infer}, etc.
+
+While the task of describing all the (dynamic) human models that are useful to
 robots is immense (if doable at all), we claim that it is possible to devise and
@@
 proxemics (human-robot distance) and associated issues (visibility) but also
-legibility and acceptability criterias expressed in terms of \emph{social rules}
+legibility and acceptability criteria expressed in terms of \emph{social rules}
 that the elaborated plans should satisfy.
@@
 Human-Robot Interaction is and is going to remain a challenging field for
-artificial intelligence. We hope that this contribution helps with clarifying
+Artificial Intelligence. We hope that this contribution helps with clarifying
 some of these challenges.

### 2016-06-16 paper.tex
Taken into account most of the reviewer 2 comments

@@
 
-\usepackage[nomargin, marginclue, footnote]{fixme}
+\usepackage[draft,nomargin, marginclue, footnote]{fixme}
 
@@
 
-Human-robot interaction is a challenge for Artificial Intelligence. This field
-lays at the crossroad of several domains of AI and requires to tackle them in a
-holistic manner: modelling humans and human cognition; acquiring, representing,
-manipulating in a tractable way abstract knowledge at the human level; reasoning
-on this knowledge to make decisions; and eventually instantiating those
-decisions into physical actions both legible to and in coordination with humans.
-Many AI techniques are invited, from visual processing to symbolic reasoning,
-from task planning to \emph{theory of mind} building, from reactive control to
-action recognition and learning.
+Human-Robot Interaction (HRI) represents a challenge for Artificial Intelligence
+(AI). It lays at the crossroad of many subdomains of AI and in effect, it calls
+for their integration: modelling humans and human cognition; acquiring,
+representing, manipulating in a tractable way abstract knowledge at the human
+level; reasoning on this knowledge to make decisions; eventually instantiating
+those decisions into physical actions both legible to and in coordination with
+humans.  Many AI techniques are mandated, from visual processing to symbolic
+reasoning, from task planning to \emph{theory of mind} building, from reactive
+control to action recognition and learning.
 
@@
 The robot controller (with the help of a task planner) decides what action to
-execute next, and who should perform it from the robot or the human (or both in
-case of a joint action), and finally controls or monitors its execution. The
+execute next, and who should perform it, from the robot or the human (or both in
+case of a joint action). It finally controls and monitors its execution. The
 operation continues until the goal is achieved, is declared unachievable or is
@@
 Besides, such abilities should be designed and implemented in a yask-independent
-manner, and should provide sufficient levels of parametrization such as they adapt to
+manner, and should provide sufficient levels of parametrization, so that they adapt to
 various environments, different tasks and variable levels of engagement of the
@@
 
-The remaining of the article discusses the robotic architecture that we have built to
-tackle the AI challenges of human-robot interaction, and how its relates to
-other approaches. We organise this discussion in four sections.
+The remaining of the article discusses the robotic architecture that we have
+built to approach social and autonomous human-robot interaction, and how it
+relates to other approaches. We organise this discussion in four sections.
 
@@
 knowledge (when to add, when to update, when to retract knowledge): no
-meta-semantics are carried over that would let the server manage these dynamics
-itself.\footnote{With one exception: so-called \emph{memory profiles} let
-the server automatically discard facts (\ie forget about them) after a specific
+meta-semantics are carried over that would let the server manage itself these
+dynamics.\footnote{With one exception: so-called \emph{memory profiles} let the
+server automatically discard facts (\ie forget about them) after a specific
 period of time. We present this feature at section~\ref{memory}.}
@@
 
-Contrary to similar projects like KnowRob~\cite{Tenorth2009a} that relies on
-the concept of \emph{computables} to lazily evaluate/acquire symbolic facts when
+Contrary to similar projects like KnowRob~\cite{Tenorth2009a} that relies on the
+concept of \emph{computables} to lazily evaluate/acquire symbolic facts when
 needed, we have an \emph{explicit} approach where we greedily compute and assert
 symbolic statements (like spatial relations between objects, see
-Section~\ref{sect|sit-ass}). This design choice trades scalability for explicit
-reasoning: at any time, the full belief state is explicitly stated, and
-provides the reasoner with the largest possible inference domain. This is of
-special importance for an \emph{event-based} architecture like ours (see
+Section~\ref{sect|sit-ass}). For instance, whenever a client queries KnowRob to
+know if \stmt{object1 isOn object2}, KnowRob calls a geometric reasonner to
+evaluate if this specific spatial relation holds at that point in time. With our
+approach, spatial relations are instead computed and asserted {\it a priori}.
+This design choice trades scalability for explicit reasoning: at any time, the
+full belief state is explicited (which may be computationally expensive), and
+therefore provides the reasoner with the largest possible inference domain. This
+is of special importance for an \emph{event-based} architecture like ours (see
 Section~\ref{sect|ctrl}), where, by definition, we do not know beforehand what
-we are ``looking for'', and we need instead to greedily infer as much as possible
-to build a belief state as comprehensive as possible.
+we are ``looking for'', and we need instead to greedily infer as much as
+possible to build a belief state as comprehensive as possible.
 
@@
 the humans, and actually many other entities that would not be commonly said to
-be objects (the Earth for instance). Thus the importance of relying on
-well-defined and standard semantics to exchange information between artificial
-systems in a non-ambiguous way.
+be objects (the Earth for instance).\footnote{In such cases of discrepancy
+between {\sc OpenCyc} concepts and human terminology, we manually label the {\sc
+OpenCyc} concepts with appropriate human names: for instance, the OpenRobots
+Ontology associates the label ``\emph{object}'' to the concept
+\concept{cyc:Artifact} instead of the concept \concept{cyc:PartiallyTangible}.
+These labels are used in priority during the grounding of verbal interactions.}
+Thus the importance of relying on well-defined and standard semantics to
+exchange information between artificial systems in a non-ambiguous way.
 
@@
 \text{\tt Bottle} \equiv \text{\tt Container} \wedge 
-                         \text{\tt Tableware} \wedge
+                         \text{\tt GraspableObject} \wedge
                          \text{\tt hasShape} \in \text{\tt CylinderShape}^{\mathcal{I}} \wedge
@@
 \emph{Symbol grounding} connects hence the knowledge model to the perception and
-actuation capabilities of the robot, and our architecture adopts here also a
-bidirectional approach: \emph{bottom-up} and \emph{top-down}.  The different
-components that we have mentioned so far provide a bottom-up grounding process:
-geometric reasoning and dialogue processing modules constantly build and push
-new symbolic contents about the world to the knowledge base where it becomes
-accessible to decisional layers. In parallel, the knowledge base relies on
-reasoning in a top-down way to produce new facts, \ie new beliefs that may lead
-to actual decision making and physical actions.
+actuation capabilities of the robot. The different components that we have
+mentioned so far provide a bottom-up grounding process: geometric reasoning and
+dialogue processing modules constantly build and push new symbolic contents
+about the world to the knowledge base where it becomes accessible to decisional
+layers. We detail in the next sections this process.
 
@@
 
-{\sc Spark} runs and continuously updates the knowledge base at about 10Hz. At each step, it re-computes spatial
-relations and affordances for the whole scene, and send the delta (new
-relations and relations that do not hold anymore) to the knowledge base. This
-approach may raise scalability concerns (we did not however hit major
-performance issues in our constrained scenarios), but simplifies the
-management of the \emph{dynamics} of the knowledge (\emph{When do I discard
-outdated knowledge?  When do I update it?}). Since it is equivalent to a reset
-of the reasoner domain, it also hides issues linked to non-monotonic reasoning
+{\sc Spark} runs and continuously updates the knowledge base at about 10Hz. At
+each step, it re-computes spatial relations and affordances for the whole scene,
+and send the delta (new relations and relations that do not hold anymore) to the
+knowledge base. This approach may raise scalability concerns (we did not however
+hit major performance issues in our constrained scenarios) as well as prevent
+reasoning on the situation history, but simplifies the management of the
+dynamics of the knowledge (\emph{When do I discard outdated knowledge?  When do
+I update it?}). Since it is equivalent to a reset of the reasoner domain, it
+also essentially nullifies issues linked to non-monotonic reasoning
 (see~\cite{McCarthy2007} for a discussion on that question).
@@
 Humans commonly refer to the positions of objects with symbolic descriptors
-(like \emph{on}, \emph{next to}...) instead of precise, absolute positions.
-These type of descriptors have been studied in the context of language grounding
+(like \emph{on}, \emph{next to}...) instead of precise, absolute positions
+(qualitative spatial reasoning). These type of descriptors have been studied in the context of language grounding
 \cite{O'Keefe1999,Matuszek2010,Regier2001,Kelleher2006,Blisard2005}.  {\sc
@@
 
-{\sc Spark} computes three main agent-independent relations based on the
-bounding box and centre of mass of the objects
+{\sc Spark} computes three main agent-independent relations~\cite{Sisbot2011}
+based on the bounding box and centre of mass of the objects
 (Figure~\ref{fig|spark-oro}\emph{a}): \concept{isOn} holds when an object $O_1$
@@
 container'' (precursor for a \emph{throw}).  {\sc Spark} recognises a set of
-such primitives that are relied upon by the execution controller to track the
-state transitions in the human plans.
+such primitives. When combined with the other geometric computations and a predictive
+plan of the human actions (see Section~\ref{hatp}), the execution controller can
+track the fulfilment of the pre- and post-conditions of the predicted human
+actions. As a result, the robot can monitor the engagement of the human and the
+overall progress of the human-robot shared plan.
 
@@
 a grammatical structure (\emph{Part of Speech} tagging), and the atoms of each
-sentence are resolved with the help
-of the ontology to ground concepts like objects (\ie when a user says ``pick the
-can'', it resolves to which instance of \emph{can} the user is referring to) and
-actions.  Sentences are classified into questions, desires and statements, and
-processed accordingly.  Figure~\ref{dialogs|ex} gives an example of the
-processing of a simple, non-ambiguous command.
+sentence are resolved with the help of the ontology to ground concepts like
+objects (\ie when a user says ``pick the can'', it resolves to which instance of
+\emph{can} the user is referring to) and actions.  Figure~\ref{dialogs|ex} gives
+an example of the processing of a simple, non-ambiguous command, while the first
+study (Section~\ref{moving-london}) walks through more complex examples.
+Heuristics (like the presence of a question mark or the use of imperative mood)
+are used to classify the sentences into questions, desires or statements, and
+{\sc Dialogs} processes them accordingly (by answering questions or
+updating the knowledge base).
 
@@
 
-        \emph{Generated query} \\
+        \emph{Generated query to ontology} \\
         \hline
@@
          \stmt{sit\_1 actsOnObject book\_01} \\
-         \stmt{sit\_1 receivedBy human} \\
+         \stmt{sit\_1 receivedBy human\_1} \\
     \end{tabular}
@@
         actsOnObject}, {\tt receivedBy}) are automatically extracted by 
-    {\sc Dialogs} from the imperative sentence ``Give me the book.''.}
+    {\sc Dialogs} from the imperative sentence ``Give me the book.'' The new
+    resulting statements (right column) are added to the knowledge base, and
+    may eventually trigger an event in the execution controller (see
+    Section~\ref{events}).}
 
@@
 generate unambiguous references to discriminate between similar objects: ``do
-you mean the bottle that is next to the glass?'' (\cite{Ros2010b} provides a
-detailed account of our approach to interactive concept discrimination, along
-with the related algorithms).
+you mean the bottle that is next to the glass?''. \cite{Ros2010b} provides a
+detailed account of our approach to interactive concept clarification and
+discrimination, along with the related algorithms.
 
@@
 \emph{shared plans}~\cite{Grosz1996,Clark1996,Kemp2007} which are then used to
-anticipate human action, to propose to human to act or even to ask help from
-the human.
+anticipate human action, to suggest a course of action to humans, or even to ask
+help from the human.
 
@@
 keep only those considered as suitable for human-robot interaction. The planner
-allows the specification of the following filtering criteria:
+allows for the specification of the following filtering criteria:
 
@@
 
-\emph{Control of intricacy}: avoids plans with too many interdependencies
+\emph{Simplicity}: avoids plans with too many interdependencies
 between the actions of agents mentioned in the plan, as a problem with
@@
 
-One of the main role of {\sc Shary} is to control the production and execution
+One of the main roles of {\sc Shary} is to control the production and execution
 of shared plans. This means essentially context-based refinement and robot
@@
 Shary} monitoring processes and {\sc Oro} inference
-mechanisms~\cite{fiore2014}.
+mechanisms~\cite{fiore2014}. \fixme{Rachid, reviewer 2 asks for more details
+about this process -> M. Fiore work?}
 
 \subsubsection{Event-Driven Control}
+\label{events}
 
@@
 expressiveness: for instance, by subscribing to the event \stmt{human1 desires
-?action, ?action type Take, ?action actsOnObject ?obj, ?obj type Toy}, we could trigger a
-behaviour when the human expresses (through dialogue, gestures...) that he
-wants a toy.
+?action, ?action type Give, ?action actsOnObject ?obj, ?obj type Book}, we could
+trigger a behaviour when the human expresses (through dialogue, gestures...)
+that he wants the robot to give her a book.
 
@@
 \subsection{\emph{Cleaning the table} Study}
+\label{sec:cleantable}
 
@@
 
-This study (Figure~\ref{fig|cleantable-video}) exhibits a richer decision-making
-process. The {\sc Oro} server is used in conjunction with the HATP symbolic task
-planner and the {\sc Shary} execution controller to produce and execute a shared
-plan.
+This second study demonstrates a richer decision-making process where the {\sc
+Oro} server is used in conjunction with the HATP symbolic task planner and the
+{\sc Shary} execution controller to produce and execute a shared plan. The task
+consists in cooperatively cleaning a table by moving objects into their target
+bins (Figure~\ref{fig|cleantable-video}).
 
@@
 is one of the main discoverers of these mechanisms, and coined the concept of
-\emph{enactivism} as the theoretical framework that study the links between
+\emph{enactivism} as the theoretical framework that studies the links between
 cognition, embodiment and actions.  This has since been thoroughly studied in
@@
 extensively studied in linguistics. The issue of grounding is well known in
-cognitive science and is summarised by Harnard~\cite{Harnad1990} by this
+cognitive science and is summarised by Harnad~\cite{Harnad1990} by this
 question: ``how the semantic interpretation of a formal symbol system can be
@@
 humans, raises issues to be tackled by the robots' decisional components, and
-that we can summarises as ``W-questions'': \emph{What}, \emph{Who}, \emph{When},
+that we can summarise as ``W-questions'': \emph{What}, \emph{Who}, \emph{When},
 \emph{Where} and \emph{How}?
@@
 into account not only the current state but also the long term goals, is the
-basic question for an intelligent robot. Its is made more complex here by having
+basic question for an intelligent robot. It is made more complex here by having
 to deal with the partially observable physical and mental state of the human
@@
 systems which have been designed to deal with such intricate decision issues.
+
+\fixme{Rachid: the reviewer 2 asks for a better connection between these general
+questions and the specific approaches that we take}
 
@@
 
-Rachid -- TDB: what are the other 'deliberative architecture for socially interactive
-robots' and where/how ours is better?
-
-Discussion on deliberative architectures != cognitive architecture: a \emph{cognitive
+\fixme{Rachid -- TDB: what are the other 'deliberative architecture for socially interactive
+robots' and where/how ours is better?}
+
+\fixme{Discussion on deliberative architectures != cognitive architecture}: a \emph{cognitive
 architecture} is usually understood as an artificial yet principled model of (human)

### 2016-06-16 paper.tex
Typo

@@
 
-Besides, such abilities should be designed and implemented in a yask-independent
+Besides, such abilities should be designed and implemented in a task-independent
 manner, and should provide sufficient levels of parametrization, so that they adapt to

### 2016-06-17 paper.tex
James' proof-reading of pages 13-14

@@
 \usepackage{xspace}
-\newcommand{\ie}{{i.e.\xspace}}
+\newcommand{\ie}{i.e.\xspace}
 \newcommand{\cf}{{\textit{cf\ }}}
-\newcommand{\eg}{{e.g.\xspace}}
-\newcommand{\etal}{{et al.\xspace}}
+\newcommand{\eg}{e.g.\xspace}
+\newcommand{\etal}{et al.\xspace}
 
@@
     \caption{List of statements describing agent-independent spatial
-        relationships between objects (top), agent-dependent placements (middle)
-        and attentional states and abilities of agents (bottom).
-        ``$\rightarrow$'' indicates sub-properties. When existing, the
-        equivalent predicate in the {\sc OpenCyc} standard (prefix \concept{cyc:})
-        is specified. Note that some relationships are not computed by {\sc Spark}, but
-        are instead inferred by the reasoner.}
+        relationships between objects (top), agent-dependent placements
+        (middle), and attentional states and abilities of agents (bottom).
+        ``$\rightarrow$'' indicates sub-properties. Where existing, the
+        equivalent predicate in the {\sc OpenCyc} standard (prefix
+        \concept{cyc:}) is specified. Note that some relationships are not
+        computed by {\sc Spark}, but are instead inferred by the reasoner.}
 
@@
 
-Besides, temporal reasoning (essential for accurate action recognition for
+Additionally, temporal reasoning (essential for accurate action recognition for
 instance) is not properly addressed in our architecture. Temporal reasoning is used
@@
 Natural language is a basic interaction modality that we use in our system both
-as input (processing of the human speech) and as output (verbalization of the
-robot intents). Natural language processing is facilitated by our architecture
-mostly processing information with semantics that are already close to the
+as an input (processing of the human speech) and as an output (verbalization of the
+robot intentions). Natural language processing is facilitated by our architecture
+that processes information with semantics that are already close to a
 human-level. This section presents the main features of our speech processor,
@@
 We acquire natural speech input from the human participants through a custom
-Android-based interface that relies on the Google speech recognition API for
-speech-to-text and relays the textual transcript to the robot. It is parsed into
+Android-based interface. The interface relies on the Google speech recognition API for
+speech-to-text and relays the textual transcript to the robot. The text is parsed into
 a grammatical structure (\emph{Part of Speech} tagging), and the atoms of each
 sentence are resolved with the help of the ontology to ground concepts like
-objects (\ie when a user says ``pick the can'', it resolves to which instance of
+objects (\ie when a user says ``pick up the can'', it resolves to which instance of
 \emph{can} the user is referring to) and actions.  Figure~\ref{dialogs|ex} gives
-an example of the processing of a simple, non-ambiguous command, while the first
+an example of the processing of a simple, non-ambiguous command. The first
 study (Section~\ref{moving-london}) walks through more complex examples.
-Heuristics (like the presence of a question mark or the use of imperative mood)
-are used to classify the sentences into questions, desires or statements, and
-{\sc Dialogs} processes them accordingly (by answering questions or
-updating the knowledge base).
+Heuristics, like the presence of a question mark or the use of imperative mood,
+are used to classify the sentences into questions, desires or statements.
+{\sc Dialogs} processes these accordingly by answering questions or
+updating the knowledge base.
 
@@
         actsOnObject}, {\tt receivedBy}) are automatically extracted by 
-    {\sc Dialogs} from the imperative sentence ``Give me the book.'' The new
+    {\sc Dialogs} from the imperative sentence ``Give me the book.'' The 
     resulting statements (right column) are added to the knowledge base, and
@@
 The system supports quantification (``give me \{a | the | some | all | any |
-...\} can''), thematic roles (action-specific predicates that qualify the
+$n$\} can''), thematic roles (action-specific predicates that qualify the
 actions), interactive disambiguation (the robot asks questions when it needs
-more information), anaphora resolution (``give \emph{it} to me'') based on
+more information), and anaphora resolution (``give \emph{it} to me'') based on
 dialogue history. It also supports knowledge extension by learning new semantic
-structures (for instance, a sentence like ``learn that cats are animals'' is
+structures. For instance, a sentence like ``learn that cats are animals'' is
 converted into \stmt{Cat subClassOf Animal} and added to the knowledge base
-after checking for possible contradictions with existing knowledge),
-interprets common temporal and spatial adverbs (like \emph{above} or
-\emph{tomorrow}) and translates simple expressions of internal state (``I'm
-tired'') into \emph{experiences} (\stmt{human\_1 experiences state\_1, state\_1
-hasFeature tired}, see also Section~\ref{sect:desires}). A full account of the
-{\sc Dialogs} features and the corresponding algorithms is available
-in~\cite{Lemaignan2011a}.
+after checking for possible contradictions with existing knowledge. {\sc
+Dialogs} finally interprets common temporal and spatial adverbs (like
+\emph{above} or \emph{tomorrow}) and translates simple expressions of internal
+state  into \emph{experiences} (for instance, ``I'm tired'' is processed into
+\stmt{human\_1 experiences state\_1, state\_1 hasFeature tired}, see also
+Section~\ref{sect:desires}). A full account of the {\sc Dialogs} features and
+the corresponding algorithms is available in~\cite{Lemaignan2011a}.
 
@@
 Because the components of our architecture rely on the same RDF formalism to
-represent their outputs, the different communication modalities (\emph{explicit}
-like verbal, deictic or based on gaze, or \emph{implicit} like the position of
-the human) are presented in a homogeneous way, as statements in the knowledge
-base. The dialogue grounding process makes use of them at two distinct levels to
-provide multi-modal concept grounding.
-
-First, specific steps of the grounding process explicitly check for the
-presence and value of specific facts: for instance, when several instances match
-a category (the human says ``give me the bottle'' and the robot knows about
-three bottles), the module may decide (it actually depends on the quantifier
-preceding the class, see the example at section~\ref{moving-london}) to discard
-some of them based on their \emph{visibility} for the speaker (implicit
-communication context taking into account the human position).
-
-Another example, when the human says ``this'', the robot checks if the human is
-currently pointing at some object. In that case, \emph{this} is replaced by the
-object focused on. Otherwise, the robot performs a lookup in the dialogue history
-to find a previous concept that the user could refer to (anaphora resolution).
-
-Note that, while the system benefits from the complementary modalities, they are
-not all required. The dialogue system can run with the verbal modality alone, at
-the cost of a simpler interaction: if the human says ``this'' without the robot
+represent their outputs, the different communication modalities are presented in
+a homogeneous way, as symbolic statements in the knowledge base. This applies
+both to \emph{explicit} modalities (verbal communication, deictic gestures,
+social gaze), and \emph{implicit} modalities (like the body position of the
+human). The dialogue grounding process makes use of them at two distinct levels
+to provide multi-modal concept grounding.
+
+First, specific steps of the grounding process explicitly check for the presence
+and value of certain facts. For instance, when several instances match a
+category (the human says ``give me the bottle'' and the robot knows about three
+bottles), the module may decide to discard some of the candidates based on their
+\emph{visibility} for the speaker (implicit communication context taking into
+account the human position). In this particular case, the heuristic is selected
+by {\sc Dialogs} based on the quantifier preceding the class (``give me
+\underline{the} bottle''). The first study (Section~\ref{moving-london})
+illustrates the details of this process.
+
+As another example, when the human says ``this'', the robot checks if the human
+is currently pointing at an object. In that case, \emph{this} is replaced by the
+object focused on. Otherwise, the robot performs anaphora resolution by looking
+up in the dialogue history to find a previous concept that the user could refer
+to.
+
+Note that while the system benefits from complementary modalities, they are not
+all required. The system can run with the verbal modality alone, at the cost of
+a simpler interaction. For example, if the human says ``this'' without the robot
 tracking what the human points at, no \stmt{human\_1 pointsAt ...} fact would
-possibly be available in the knowledge base, and the robot falls back to the
+possibly be available in the knowledge base, and the robot falls back on the
 anaphora resolution step alone.
@@
 The second level of integration of multi-modality is implicit. By continuously
-computing symbolic properties from the geometric model, richer descriptions and
-hence discrimination possibilities are available. For instance, the robot may
-compute that one bottle is next to a glass while another one stands alone, and
-these symbolic descriptions are transparently re-used in a dialogue context to
-generate unambiguous references to discriminate between similar objects: ``do
-you mean the bottle that is next to the glass?''. \cite{Ros2010b} provides a
-detailed account of our approach to interactive concept clarification and
-discrimination, along with the related algorithms.
+computing symbolic properties from the geometric model, richer symbolic
+descriptions are available to the system to verbalize or discriminate entities.
+For instance, the robot may compute that one bottle is next to a glass, while
+another one stands alone. These symbolic descriptions are transparently
+re-used in a dialogue context to generate unambiguous references to discriminate
+between similar objects: ``do you mean the bottle that is next to the glass?''.
+The physical context of the interaction is used as an implicit communication
+modality by {\sc Dialogs}. \cite{Ros2010b} provides a detailed account of our
+approach toward interactive concept clarification and discrimination, along with the
+related algorithms.
 
@@
 Our execution controllers rely on symbolic task planning to convert long-term
-desires into a succession of atomic actions. We use in our architecture the HATP
+desires into a succession of atomic actions. Our architecture uses the HATP
 planner (\emph{Human Aware Task Planner})~\cite{Alili2008,
@@
 
-The HATP planning framework extends the traditional HTN (Hierarchical Task
-Network) planning domain representation and semantics by making them more
-suitable to elaborate plans which involve humans and robots acting together
-toward a joint goal. HATP is used by the robot to elaborate human-robot
+The HATP planning framework extends the traditional Hierarchical Task
+Network (HTN) planning domain representation and semantics by making them more
+suitable to produce plans which involve humans and robots acting together
+toward a joint goal. HATP is used by the robot to produce human-robot
 \emph{shared plans}~\cite{Grosz1996,Clark1996,Kemp2007} which are then used to
@@
 %\paragraph{Action Execution and Monitoring Robot Controller}
-%Based on context and on the shared plan elaborated by HATP for a given goal,
+%Based on context and on the shared plan produced by HATP for a given goal,
 %the robot controller decides to execute an action or to ask its human
@@
 embeds human preferences and physical constraints that need to be accounted for
-when synthesizing robot motion or elaborating robot plans. This includes
+when synthesizing robot motion or producing robot plans. This includes
 proxemics (human-robot distance) and associated issues (visibility) but also
 legibility and acceptability criteria expressed in terms of \emph{social rules}
-that the elaborated plans should satisfy.
+that the produced plans should satisfy.
 

### 2016-06-19 paper.tex
Update on Overleaf.

@@
 which mandates common-sense knowledge and the representation of possibly
-divergent mental models. This article attempts to characterize these challenges
-and evidences the key decisional issues that need to be addressed for a
+divergent mental models. This article is an attempt to characterize these challenges
+and to exhibit a set of key decisional issues that need to be addressed for a
 cognitive robot to successfully share space and tasks with a human.
@@
 multiple interactants (humans and robots); natural multi-modal dialogue;
-human-aware task planning; human-robot joint plan achievement.  The article
-discusses each of these abilities, presents reference implementations, and show
+human-aware task planning; human-robot joint task achievement.  The article
+discusses each of these abilities, presents a working implementation, and shows
 how they combine in a coherent and original cognitive architecture for

### 2016-06-20 paper.tex
Minor changes in the abstract + typo in title

@@
 \title{\LARGE \bf
-Artifical Cognition for Social Human-Robot Interaction:\\ An Implementation
+Artificial Cognition for Social Human-Robot Interaction:\\ An Implementation
 }
@@
 robots; a broad variety of situations with rich semantics to understand and
-handle; physical interactions with humans that requires fine, low-latency yet
+interpret; physical interactions with humans that requires fine, low-latency yet
 socially acceptable control strategies; natural and multi-modal communication
@@
 discusses each of these abilities, presents a working implementation, and shows
-how they combine in a coherent and original cognitive architecture for
+how they combine in a coherent and original deliberative architecture for
 human-robot interaction. Supported by experimental results, we eventually show

### 2016-06-20 paper.tex
Update on Overleaf.
Clarify ontology caption following reviewer 2 comment

@@
 \address{
-$^1$CNRS, LAAS, 7 avenue du Colonel Roche, F-31400 Toulouse, France\\
-Univ de Toulouse, LAAS, Toulouse, France\\
+$^1$LAAS-CNRS, Univ. de Toulouse, CNRS\\
+7 avenue du Colonel Roche, F-31400 Toulouse, France\\
 {\tt firstname.surname@laas.fr}
@@
 
-\caption{The upper part of the {\sc Oro} common-sense TBox. All these concepts
-    belong to the {\sc OpenCyc} namespace.}
+    \caption{The upper part of the {\sc Oro} common-sense conceptualization
+    (TBox). These concepts are shared with the {\sc OpenCyc} upper-ontology.
+    They relate to each other through `is-a' subsumption relations.}
 

### 2016-06-20 rebuttal2.tex
Added a rebuttal for the 2nd round of revisions

\documentclass{article}
\usepackage[utf8]{inputenc}

\begin{document}

Dear Editors,

\vspace{2em}

Thank you for coming back to us, we are glad that the special issue is still
going on.

We have hopefully addressed the remaining concerns raised during the second
round of reviews.

\subsection*{Language}

Your first and most stringent comment relates to the language. We have carefully
proof-read the manuscript with the help of a native English speaker.
Specifically:

\begin{itemize}
    \item a special care has been given to rephrase colloquial expressions and
        non-academic phrasing,
    \item the abstract has been entierly rewritten,
    \item all the specific language changes pointed out by the editors and
        reviewers have been integrated.
\end{itemize}

To facilitate the review process, we attach an annotated version of the
manuscript where all these changes have been highlighted.

\subsection*{Relation to the state of the art}

We initially made the choice to discuss the existing litterature in each of the
``cognitive skills'' sections, for readability reasons. We however acknowledge
that a higher-level discussion of our work with respect to the state of the art
of human-robot interaction systems would be useful.

To this end, we have split Section 5 into a discussion (Section 5) and a
conclusion (Section 6) that states clearly our contribution (including
references to all our previous publications -- Section 6.1), followed by a
comparative discussion with the related approaches (Section 6.2).

\subsection*{Reviewer Comments}

Reviewer 2's comments have all been addressed in the manuscript.
We highlight hereafter the main changes.

\begin{itemize}
    \item examples have been added to Section 2.2 to make it easier to follow,
    \item we have clarified how the system manages discrepancies between OpenCyc
        terminology and human natural language by explicit manual labelling
        (Section 2.2.2)
    \item We have precised that our approach of transient geometric situation assessment
        prevents reasoning on the situation history (Section 3.2)
    \item We have clarified how the {\tt nextTo} spatial relation is computed.
    \item We have clarified how primitive action recognition is used in
        combination with situation assessment and task planning to monitor the
        human activities (Section 3.2.3)
    \item The implicit aspect of multi-modality during dialogue processing has
        been made clearer (Section 3.3.2)
\end{itemize}

Finally, as suggested by Reviewer 3, we have changed the title of the article
from ``Human-Robot Interaction: Tackling the AI Challenges'' to the more
specific ``Artificial Cognition for Social Human-Robot Interaction: An
Implementation''.

\bibliographystyle{abbrv}
\bibliography{laas_hri}

\end{document}

### 2016-06-29 paper.tex
Update on Overleaf.

@@
 and affordance analysis; acquisition and representation of knowledge models for
-multiple interactants (humans and robots); natural multi-modal dialogue;
+multiple agents (humans and robots, with their specificities) natural multi-modal dialogue;
 human-aware task planning; human-robot joint task achievement.  The article
@@
 
-We focus on a specific class of interactions: collaborative task achievement
+The proposed scheme is specially designed for  interaction, and more particularly cooperative task achievement by one robot and one or several humans - called here the robot human partners.
+It is applicable to domains such as assistant or service robots in domestic and public places as well as to teammate robots in the factory of the future.
+
+We focus on a specific class of interactions: human-robot collaborative task achievement
 supported by multi-modal and situated communication. Figure~\ref{fig:hri-dec}
-illustrates this context: the two agents share a common space and exchange
+illustrates this context: the human and the robot share a common space and exchange
 information through multiple modalities (we specifically consider verbal
@@
 chores by taking into account, at every stage, the intentions, beliefs,
-perspectives, skills of the human partner.  Namely, the robot must be able to
+perspectives, skills of its human partner.  Namely, the robot must be able to
 recognise, understand and participate in communication situations, both explicit
@@
 
-These three challenges, \emph{communication}, \emph{joint actions},
+These three challenges, \emph{communication}, \emph{joint action},
 \emph{human-aware execution}, structure the research in human-robot
@@
     \item a belief state that includes {\it a priori} common-sense knowledge and
-        mental models of each of the interactors.
+        mental models of each of the agents involved (the robot and its human partners).
 
@@
 
-\subsection{Article Overview}
-
+\subsection{Contribution and Article Overview}
+
+The main contributions concern the architecture of the decisional layer of a robot designed to share 
+the space and the task with humans and to act and interact in a way that facilitates human action and decision.
+
+We will show in the sequel that:
+\begin{itemize}
+
+\item such architecture allows to achieve advanced grounding in a relatively complex environment 
+involving one or several humans and a robot
+
+\item  it is able, thanks to the combination of perspective taking and affordances computation 
+and with the help of an inference engine, to provide pertinent knowledge for situated dialog
+
+\item  this architecture provides a framework and and a set of mechanisms that allow 
+the robot to reason about the mental state of its human partners
+
+\item  the same set of affordances and induced facts together with an explicit reasoning
+on humans and robot abilities in a given context,  allow to synthesize relevant human-robot shared plans
+
+\item  it opens the possibility and provides a propitious context for human-aware planning 
+at symbolic as well as at geometric level
+
+\end{itemize}
+
+Another important aspect of such an architecture is that it has been fully implemented and 
+demonstrated on several use cases. It is also noteworthy that it proved to be a pertinent 
+framework  for new contributions about human-robot joint  action as well as for multi-disciplinary studies.
+
+While several contributions in the literature provide very interesting insights and contributions 
+on one aspect or another, we are not aware of a fully implemented architecture that effectively
+combines in a coherent way all the issues mentioned above. Besides, this has been done with an effort
+to make explicit all the knowledge involved.
+
+We do not claim that this is the sole architecture that can be designed for such problems nor that it
+covers all the issues. Our point here is that it is a fully implemented architecture that allowed us to
+identify, refine and combine a number of key cognitive abilities that we believe are necessary. Indeed, 
+we show here below that it provided us a mean to develop and explore a number of key features for HRI 
+in terms of models and algorithms and their articulation in an integrated decisional system for a 
+cognitive and interactive robot.
+
+\vspace{0.5cm}
 The remaining of the article discusses the robotic architecture that we have
-built to approach social and autonomous human-robot interaction, and how it
+built to address social and autonomous human-robot interaction, and how it
 relates to other approaches. We organise this discussion in four sections.
@@
 knowledge model that we have developed for our robots. Then,
-Section~\ref{sec:impl} details each of the main components of the architecture,
-and highlights their significance for Artificial Intelligence.
+Section~\ref{sec:impl} provides further details on each of the main components of the architecture.
 Section~\ref{sec:expe} presents two studies that illustrate in a practical way
 what can be currently achieved with our robots.  The
-Sections~\ref{sect|discussion} and~\ref{sect|conclusion} finally restate our
+Sections~\ref{sect|discussion} and~\ref{sect|conclusion} finally summarize our
 main contributions and discuss the key challenges that human-robot interaction
@@
 
-Bridging multiple independent software modules in one coherent robotic
+Articulating multiple independent software modules in one coherent robotic
 architecture is not only a technical challenge, but also a design and
@@
 knowledge base ({\sc Oro}), conveniently thought as a semantic
-blackboard, connects most of the modules together: the geometric
-reasoning module ({\sc Spark}) produces symbolic assertions (like
-\stmt{BOOK1 isOn TABLE}) describing the state and dynamics of the robot's
-environment. These logical statements are stored in the knowledge base, and
-queried back by the language processing module ({\sc Dialogs}), the symbolic task
+blackboard, connects most of the modules: the geometric
+reasoning module ({\sc Spark}) produces ar relatively high frequency symbolic assertions (like
+\stmt{BOOK1 isOn TABLE}) describing the state of the robot environment and its evolution over time. 
+These logical statements are stored in the knowledge base, and
+queried back, when necessary, by the language processing module ({\sc Dialogs}), the symbolic task
 planner (HATP) and the execution controller. The output of the language
-processing module and the activities started by the robot controller are
+processing module and the activities managed by the robot controller are
 likewise stored as symbolic statements.
@@
             \node [left=of stt] {furniture} edge (stt);
-            \node [below right=of stt] (ea) {embodied agent} edge (stt);
+            \node [below right=of stt] (ea) {cited agent} edge (stt);
                 \node [below=of ea] {human} edge (ea);
@@
 %\bibliographystyle{abbrv}
-\bibliography{laas_hri}
+\bibliography{laas_hri,other_citations}
 

### 2016-06-29 paper.tex
Polished Rachid's suggestions

@@
 and affordance analysis; acquisition and representation of knowledge models for
-multiple agents (humans and robots, with their specificities) natural multi-modal dialogue;
+multiple agents (humans and robots, with their specificities); natural multi-modal dialogue;
 human-aware task planning; human-robot joint task achievement. The article
@@
 
-The proposed scheme is specially designed for  interaction, and more particularly cooperative task achievement by one robot and one or several humans - called here the robot human partners.
+%The proposed scheme is specially designed for  interaction, and more particularly cooperative task achievement by one robot and one or several humans - called here the robot human partners.
 It is applicable to domains such as assistant or service robots in domestic and public places as well as to teammate robots in the factory of the future.
@@
 
-We will show in the sequel that:
+We present a model of cognitive integration for service robots that:
+
 \begin{itemize}
 
-\item such architecture allows to achieve advanced grounding in a relatively complex environment 
-involving one or several humans and a robot
-
-\item  it is able, thanks to the combination of perspective taking and affordances computation 
-and with the help of an inference engine, to provide pertinent knowledge for situated dialog
-
-\item  this architecture provides a framework and and a set of mechanisms that allow 
-the robot to reason about the mental state of its human partners
-
-\item  the same set of affordances and induced facts together with an explicit reasoning
-on humans and robot abilities in a given context,  allow to synthesize relevant human-robot shared plans
-
-\item  it opens the possibility and provides a propitious context for human-aware planning 
-at symbolic as well as at geometric level
+    \item achieves advanced grounding in complex real environments involving one or
+        several humans and a robot;
+
+    \item supports a distributed computation of symbolic knowledge for
+        situated dialogue, thanks to the combination of perspective taking,
+        affordances computation and logical inference;
+
+    \item provides generic mechanisms for the robot to reason about the mental state
+        of its human partners;
+
+    \item reuses the same set of affordances and inferences, together with
+        explicit contextual reasoning on humans and robot abilities, to generate
+        human-robot shared plans;
+
+    \item exposes a principled approach to integrate a set of complex cognitive
+        components in an explicit, semantics-oriented and loosely-coupled
+        fashion.
 
@@
 
-Another important aspect of such an architecture is that it has been fully implemented and 
-demonstrated on several use cases. It is also noteworthy that it proved to be a pertinent 
-framework  for new contributions about human-robot joint  action as well as for multi-disciplinary studies.
-
-While several contributions in the literature provide very interesting insights and contributions 
-on one aspect or another, we are not aware of a fully implemented architecture that effectively
-combines in a coherent way all the issues mentioned above. Besides, this has been done with an effort
-to make explicit all the knowledge involved.
-
-We do not claim that this is the sole architecture that can be designed for such problems nor that it
-covers all the issues. Our point here is that it is a fully implemented architecture that allowed us to
-identify, refine and combine a number of key cognitive abilities that we believe are necessary. Indeed, 
-we show here below that it provided us a mean to develop and explore a number of key features for HRI 
-in terms of models and algorithms and their articulation in an integrated decisional system for a 
-cognitive and interactive robot.
+Importantly, this architecture is fully implemented and we demonstrate it on
+several robotic platforms and in several actual interaction scenarios. It proved
+to be an effective framework for novel contributions about human-robot joint
+action, as well as for multi-disciplinary studies.
 
@@
 blackboard, connects most of the modules: the geometric
-reasoning module ({\sc Spark}) produces ar relatively high frequency symbolic assertions (like
+reasoning module ({\sc Spark}) produces at relatively high frequency symbolic assertions (like
 \stmt{BOOK1 isOn TABLE}) describing the state of the robot environment and its evolution over time. 
@@
             \node [left=of stt] {furniture} edge (stt);
-            \node [below right=of stt] (ea) {cited agent} edge (stt);
+            \node [below right=of stt] (ea) {embodied agent} edge (stt);
                 \node [below=of ea] {human} edge (ea);
@@
 
-\subsection{Summary of the Contribution}
+\subsection{A Deliberative Architecture for Social Robots}
 
@@
 
-\subsection{A Deliberative Architecture for Social Robots}
-
 \fixme{Rachid -- TDB: what are the other 'deliberative architecture for socially interactive
 robots' and where/how ours is better?}
+
+While several contributions in the literature provide insights and contributions 
+on one aspect or another, we are not aware of a fully implemented architecture that effectively
+combines in a coherent manner all the points mentioned above.
+We do not claim that this represents the only architecture that can be designed
+to address such problems, nor that our approach fully addresses each and every
+of these issues. Our point here is that it is a fully implemented architecture that allowed us to
+identify, refine and combine a number of key cognitive abilities that we believe are necessary. Indeed, 
+we show here below that it provided us a mean to develop and explore a number of key features for HRI 
+in terms of models and algorithms and their articulation in an integrated decisional system for a 
+cognitive and interactive robot.
 

### 2016-06-30 paper.tex
Added Aurelie as co-author
Update on Overleaf.

@@
 \author{Séverin Lemaignan$^{1,2}$, Mathieu Warnier$^1$, E. Akin Sisbot$^1$,
-Rachid Alami$^1$}
+    Aurélie Clodic$^1$, Rachid Alami$^1$}
 
@@
 
-We focus on a specific class of interactions: human-robot collaborative task achievement
+We focus on a specific class of interactions: human-robot collaborative task achievement \cite{alami2013human} 
 supported by multi-modal and situated communication. Figure~\ref{fig:hri-dec}
@@
 This translates into several decisional, planning, representation skills that
-need to be available to the robot. It must be able \item to represent and
-manipulate symbolic belief states, \item to acquire and keep them up-to-date
-with respect to the state of the world and the task at hand, \item to build and
+need to be available to the robot \cite{alami2011robot}.
+It must be able \item to represent and manipulate symbolic belief states, \item to acquire and keep them up-to-date with respect to the state of the world and the task at hand, \item to build and
 iteratively refine shared (human-robot) plans, \item to instantiate and execute
@@
 Importantly, this architecture is fully implemented and we demonstrate it on
-several robotic platforms and in several actual interaction scenarios. It proved
+several robotic platforms and in several actual interaction scenarios. It also proved
 to be an effective framework for novel contributions about human-robot joint
-action, as well as for multi-disciplinary studies.
+action \cite{fiore2016planning,devin2016implemented,milliez2016using}, as well as for multi-disciplinary studies \cite{dautenhahn2006may,koay2007exploratory,Ros2010b,dehais2011physiological,ferreira2015users,GharbiROMAN2015} .
 
@@
 Book, ?book differentFrom BOOK1)}\footnote{We present a complete
-example at section~\ref{moving-london}.}, and write back assertions like
+example in section~\ref{moving-london}.}, and write back assertions like
 \stmt{HUMAN desires GIVE\_ACTION45, GIVE\_ACTION45 actsOn BOOK2} to {\sc
@@
 server automatically discard facts (\ie forget about them) after a specific
-period of time. We present this feature at section~\ref{memory}.}
+period of time. We present this feature in section~\ref{memory}.}
 
@@
 action recognition), these are managed by dedicated components (like {\sc
-Spark}, discussed at Section~\ref{sect|sit-ass}), and are not exposed at
+Spark}, discussed in Section~\ref{sect|sit-ass}), and are not exposed at
 the semantic level.
@@
 
-Anecdotally, and because ontologies and RDF statements remain conceptually
+Incidentally, and because ontologies and RDF statements remain conceptually
 simple (compared to full logical languages like {\sc Prolog} or modal logics),
@@
 context of service robotics in an human-like environment are explicitly stated.
-For performance reasons as well as clarity, we decided against an extended
+For performance reasons as well as clarity, we have decided against an extended
 conceptual coverage of what ``partially tangible things'' might be. We instead
@@
 about the world to the knowledge base where it becomes accessible to decisional
-layers. We detail in the next sections this process.
+layers. We detail this process in the next sections.
 
@@
 play a key role in the process of \emph{interactive grounding} of the semantics
-of concepts (we discuss this process at section~\ref{sect|com}). Clarification
+of concepts (we discuss this process in section~\ref{sect|com}). Clarification
 and discrimination algorithms are based on what we call \emph{descriptors}, \ie
@@
 
-From the point of view of robotics, it supposes the ability to build, store and
-retrieve separate models of the beliefs of the agents the robot interacts with.
-Our knowledge base implements such a mechanism: when the robots recognises that
-a new agent has appeared, it initialises a new
-independent ontology for this agent. All the ontologies that are created share
-the same common-sense knowledge, but rely on each agent's perspective for the
-actual instantiation: if the robot (geometrically) computes that the book is in its
+From the point of view of interactive robotics, it supposes the robot ability to build, store and
+retrieve separate models of the beliefs of the humans it interacts with.
+Our knowledge base implements such a mechanism: when the robot detects that
+a new human has appeared, it initialises a new
+independent ontology for this human agent. All the ontologies that are created share
+the same common-sense knowledge, but rely on the estimation of the robot of each agent's perspective for their actual instantiation: if the robot (geometrically) computes that the book is in its
 own field of view, but not in the human one, the robot knowledge contains the
@@
 beliefs} on the world (A still thinks the object is at its original
-position since he did not see B hiding it). Using separate knowledge models
+position since he did not see B hiding it in a new place). Using separate knowledge models
 in the knowledge base, we have been able to replicate this experience with
@@
 
-   \caption{Test setup involving videotapes that are manipulated, with other
+   \caption{Test setup involving videotapes boxes that are manipulated, with other
        objects acting as supports or containers. After identification and
@@
        reachable by itself: the black and grey (in the 3D model) tapes.  The
-       third tape and the pink box are only reachable by the human. This
-       physical situation is transformed by {\sc Spark} into the symbolic model
+       third tape and the pink container box are only reachable by the human. This
+       physical situation is transformed by {\sc Spark} into the set of facts 
        presented in Table~\ref{table|beliefsfig7}.  }%
@@
 child is able to correctly identify which ball the adult is referring to (\ie
-the one visible from the adult point of view), without asking~\cite{Moll2006}.
+the one visible from the adult point of view), without asking~\cite{Moll2006}. This ia typically an ability our robot has thanks to the proposed architecture.
 
@@
     reach for an object, which is instrumental for effective social task
-    planning. For example, if the user asks the robot to give him/her an object,
-    the robot must compute a transfer point where the user is able to get the
-    object afterwards.  Reachability is computed for each agents based on
+    planning. For example, if the human asks the robot to give her/him an object,
+    the robot must compute a transfer point where she/he s able to get the
+    object afterwards.  Reachability is computed for each agent (human or robot) based on
     Generalised Inverse Kinematics and collision detection.
@@
 Building such a robotic architecture is the work of many hands, and we would
-like to acknowledge here the worthwhile contributions of Samir Alili, Aurélie
-Clodic, Raquel Ros Espinoza, Mamoun Gharbi, Julien Guitton, Matthieu Herrb, Jim
+like to acknowledge here the worthwhile contributions of Samir Alili, Raquel Ros Espinoza, Mamoun Gharbi, Julien Guitton, Matthieu Herrb, Jim
 Mainprice and Amit Kumar Pandey.

### 2016-06-30 paper.tex
Update on Overleaf.
Proof-reading of Rachid's changes

@@
 
-We focus on a specific class of interactions: human-robot collaborative task achievement \cite{alami2013human} 
+We focus on a specific class of interactions: human-robot collaborative task
+achievement~\cite{alami2013human} 
 supported by multi-modal and situated communication. Figure~\ref{fig:hri-dec}
@@
 This translates into several decisional, planning, representation skills that
-need to be available to the robot \cite{alami2011robot}.
-It must be able \item to represent and manipulate symbolic belief states, \item to acquire and keep them up-to-date with respect to the state of the world and the task at hand, \item to build and
-iteratively refine shared (human-robot) plans, \item to instantiate and execute
-the actions it has to perform, and conversely, to monitor those achieved by
-its human partner.
+need to be available to the robot~\cite{alami2011robot}. It must be able \item
+to represent and manipulate symbolic belief states, \item to acquire and keep
+them up-to-date with respect to the state of the world and the task at hand,
+\item to build and iteratively refine shared (human-robot) plans, \item to
+instantiate and execute the actions it has to perform, and conversely, to
+monitor those achieved by its human partner.
 
@@
 to be an effective framework for novel contributions about human-robot joint
-action \cite{fiore2016planning,devin2016implemented,milliez2016using}, as well as for multi-disciplinary studies \cite{dautenhahn2006may,koay2007exploratory,Ros2010b,dehais2011physiological,ferreira2015users,GharbiROMAN2015} .
+action~\cite{fiore2016planning,devin2016implemented,milliez2016using}, as well
+as for multi-disciplinary studies~\cite{dautenhahn2006may,koay2007exploratory,Ros2010b,dehais2011physiological,ferreira2015users,GharbiROMAN2015}.
 
@@
 child is able to correctly identify which ball the adult is referring to (\ie
-the one visible from the adult point of view), without asking~\cite{Moll2006}. This ia typically an ability our robot has thanks to the proposed architecture.
+the one visible from the adult point of view), without asking~\cite{Moll2006}.
+Our architecture endows the robot with such a cognitive skill.
 
@@
     reach for an object, which is instrumental for effective social task
-    planning. For example, if the human asks the robot to give her/him an object,
-    the robot must compute a transfer point where she/he s able to get the
+    planning. For example, if the human asks the robot to give her an object,
+    the robot must compute a transfer point where she is able to get the
     object afterwards.  Reachability is computed for each agent (human or robot) based on
-    Generalised Inverse Kinematics and collision detection.
+    Generalised Inverse Kinematics and collision detection. More than that, our robot is able 
+    to compute an estimate of the effort the
 
@@
 Building such a robotic architecture is the work of many hands, and we would
-like to acknowledge here the worthwhile contributions of Samir Alili, Raquel Ros Espinoza, Mamoun Gharbi, Julien Guitton, Matthieu Herrb, Jim
-Mainprice and Amit Kumar Pandey.
+like to acknowledge here the worthwhile contributions of Samir Alili, Raquel Ros
+Espinoza, Mamoun Gharbi, Julien Guitton, Matthieu Herrb, Jim Mainprice and Amit
+Kumar Pandey.
 

### 2016-06-30 paper.tex
Update on Overleaf.

@@
     Generalised Inverse Kinematics and collision detection. More than that, our robot is able 
-    to compute an estimate of the effort the
+    to compute an estimate of the effort needed by an agent to reach an object\cite{pandey2013affordance}.
 
@@
 Monitoring human activity is needed by the execution controllers to track the
-engagement of the human and the progress of his/her actions. Full human action
+engagement of the human and the progress of his/her actions and also to synchtronize seamlessly
+its own actions with the human actions. Full human action
 and activity recognition is a task that requires knowledge and reasoning both on
 high-level facts like goals, intentions and plans, as well as bottom-up data
-from agent and object motions. {\sc Spark} implements a set of simple temporal and
+from human and object motions. {\sc Spark} implements a set of simple temporal and
 geometric heuristics on human hand trajectories and possible objects placements

### 2016-06-30 paper.tex
Added a couple of citations + rephrase theory of mind section

@@
 time triples are added or removed from the knowledge base, a Description Logics
-reasoner ({\sc Pellet}\footnote{\url{https://github.com/complexible/pellet}})
+reasoner ({\sc Pellet}~\cite{sirin2007pellet})
 classifies the whole ontology and inserts all possible inferred triples.  The
@@
 visible} for agent B. Children develop this skill, which is essential to understand others' perspectives during
-interactions, around the age of three.
-
-From the point of view of interactive robotics, it supposes the robot ability to build, store and
-retrieve separate models of the beliefs of the humans it interacts with.
-Our knowledge base implements such a mechanism: when the robot detects that
-a new human has appeared, it initialises a new
-independent ontology for this human agent. All the ontologies that are created share
-the same common-sense knowledge, but rely on the estimation of the robot of each agent's perspective for their actual instantiation: if the robot (geometrically) computes that the book is in its
-own field of view, but not in the human one, the robot knowledge contains the
-fact \stmt{book isVisible true} while the human model contains \stmt{book
-isVisible false} (this computation, called \emph{perspective taking}, is
-discussed in the next section).
+interactions, around the age of three~\cite{perner2012infants}.
+
+From the point of view of interactive robotics, it supposes the robot ability to
+build, store and retrieve separate models of the beliefs of the humans it
+interacts with.  Our knowledge base implements such a mechanism: when the robot
+detects that a new human has appeared, it initialises a new independent
+knowledge model (an ontology) for this human agent. All the ontologies that are
+created share the same common-sense knowledge, but rely on the estimation of the
+robot of each agent's perspective for their actual instantiation. For example,
+the robot may (geometrically) compute that the book is in its own field of view,
+but not in the human one (the detail of this computation, called
+\emph{perspective taking}, is discussed in the next section). The robot updates
+accordingly the two knowledge models it maintains: the \concept{robot} model is
+updated with the fact \stmt{book isVisible true}, while the \concept{human}
+model is updated with \stmt{book isVisible false}. These two logical statements
+are simultaneously asserted, yet contradict when taken together. Since the two
+knowledge models are realized as two independent ontologies, the contradiction
+does not actually occur and both the models remain logically consistent.
 

### 2016-06-30 paper.tex
{his/her -> her}

@@
 object within a reference system (or changes the reference system, \ie switches
-perspective) without informing his/her partner about it~\cite{Breazeal2006,
+perspective) without informing her partner about it~\cite{Breazeal2006,
 Ros2010}. For example, the speaker could ask for the ``keys on the left''.
@@
     Generalised Inverse Kinematics and collision detection. More than that, our robot is able 
-    to compute an estimate of the effort needed by an agent to reach an object\cite{pandey2013affordance}.
+    to compute an estimate of the effort needed by an agent to reach an object~\cite{pandey2013affordance}.
 
@@
 Monitoring human activity is needed by the execution controllers to track the
-engagement of the human and the progress of his/her actions and also to synchtronize seamlessly
+engagement of the human and the progress of her actions and also to synchtronize seamlessly
 its own actions with the human actions. Full human action
@@
 Additionally, temporal reasoning (essential for accurate action recognition for
-instance) is not properly addressed in our architecture. Temporal reasoning is used
+instance) is not properly addressed in the current version of our system. Temporal reasoning is used
 only locally, and does not allow for tracking of long sequences or global events.
@@
 \emph{When to perform a given action} is enriched here by the human context
-since it has to take into account the human, his/her needs, his/her rhythms and
-pace, and his/her mental state. While performing its share of the task, the
+since it has to take into account the human, her needs, her rhythms and
+pace, and her mental state. While performing its share of the task, the
 robot has to produce signals directed towards the human and to respond to

### 2016-06-30 paper.tex
Update on Overleaf.
Work on conclusion

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Multi-Modal Communication}
+\subsection{Multi-Modal Communication and Situated Dialogue}
 \label{sect|com}
@@
 as an input (processing of the human speech) and as an output (verbalization of the
-robot intentions). Natural language processing is facilitated by our architecture
+robot intentions as well human-robot shared plans). Natural language processing is facilitated by our architecture
 that processes information with semantics that are already close to a
@@
 
-\subsubsection{Multi-Modality}
+\subsubsection{Dealing with Multi-Modality}
 
@@
 
-Our execution controllers rely on symbolic task planning to convert long-term
-desires into a succession of atomic actions. Our architecture uses the HATP
+Whenever necessary, the execution controllers rely on symbolic task planning to convert long-term
+desires into a set of partially ordered elementary actions. This is the role of the HATP
 planner (\emph{Human Aware Task Planner})~\cite{Alili2008,
@@
 anticipate human action, to suggest a course of action to humans, or even to ask
-help from the human.
+help from the human if needed.
 
 The HATP planning domain defines a set of methods describing how to
-decompose a task and represents the procedural knowledge of the robot as well as
+incrementally decompose a task and to allocate subtasks and actions to the robot and/or the human depending on the context. This represents the procedural knowledge of the robot as well as
 its knowledge about the actions that the human partner is able to achieve. It is
@@
 
+We briefly discuss here below how HATP incrementally builds and synchronises streams of actions concerning each or the agents (humans and robot) involved in a task, and how it promotes plans that satisfy humans needs and preferences as well as comfort and legibility issues.
+
 \subsubsection{Agents and Action Streams}
 
-The originality of HATP resides in its ability to produce plans for the robot's
-actions as well as for the other participants (humans or robots), that we
-call \emph{shared plans}.
+The originality of HATP resides in its ability to produce \emph{shared plans} which might involve 
+the robot as well as the other participants (humans or robots).
 
@@
 prediction is used by the robot execution controller to monitor the engagement
-of the human partner during joint tasks.
-
-The planner also generates synchronisation points between the agents. For
+of the human partner during plan achievement.
+
+The planner also generates causal links and synchronisation points between the robot and his human partners. For
 instance, in the plan depicted on Figure~\ref{plan_hatp1}, the human needs to
@@
 perform his next action (here, \concept{TAKE(BOTTLE, TABLE)}). This, in turn,
-allows the robot to monitor the human engagement and progress in the shared
-plan.
+allows the robot to monitor the human engagement and the progress in the shared
+plan execution.
 
 We also use shared plans to verbalise the sequence of actions and to explain
-the human partner how a task may be shared~\cite{warnier2012when}.
+the human partner how a task may be shared~\cite{warnier2012when,fiore2016planning,milliez2016using}.
 
@@
     are generated (human actions at the top, robot actions at the bottom, {\sf
-    PUTRV} stands here for {\it Put it so it is both reachable and visible}).
-    The arrow between the two streams represents a synchronization point.}
+    PUTRV} stands here for {\it Put it so it is both Reachable and Visible}).
+    The arrow between the two streams represents a synchronization point between 
+    the robot and the human based on a causal link.}
 
@@
 one hand, to schedule the different streams and, on the other hand, as an
-additional cost function.
+additional cost function. Cost here is not used here to estimate only standard time and energy consumption but
+to estimate a number factors that measure the satisfaction of the human in terms of acceptability and legibility of robot behavior.
 
@@
 presented in other publications, we offer here for the first time a perspective
-on the overall integration of these components into a coherent and consistent
+on the model of integration of these components into a coherent and consistent
 system for social human-robot interaction. We have first exposed our underlying
@@
 assessment environment~\cite{Sisbot2011} that supports perspective
-taking~\cite{Marin2008,Ros2010}. We
-combine it further with an advanced natural
-language processor~\cite{Lemaignan2011a} to provide complete multi-modal interactive
-communication.  The paper also covers our symbolic social task
+taking~\cite{Marin2008,Ros2010}. We combine it further with a situated natural
+language processor~\cite{Lemaignan2011a} to provide complete multi-modal
+interactive communication.  The paper also covers our symbolic social task
 planner~\cite{Alili2008, Alili2009, Lallement2014}: it generates predictive
@@
 planner~\cite{Sisbot2008,Mainprice2011,Pandey2011}, and finally present two
-execution controller, the PRS-based {\sc Shary}~\cite{clodic2008shary,fiore2014} and the
-event-driven {\sc pyRobots}~\cite{lemaignan2015pyrobots}.
+execution controller, the PRS-based {\sc Shary}~\cite{clodic2008shary,fiore2014}
+and the event-driven {\sc pyRobots}~\cite{lemaignan2015pyrobots}.
 
 The integration of these components in a consistent, working and observable
-system is supported by the design of our software interfaces: most of the
-datastreams exchanged by our components uses high-level semantics in the form of
-first-order logic statements. The two reported support studies further evidence
-how our use of explicit, high-level semantics provide a low-impedance,
-human-level mean of building a complex deliberative
-architecture for socially interactive robots~\cite{Lemaignan2013}.
-
-\fixme{Rachid -- TDB: what are the other 'deliberative architecture for socially interactive
-robots' and where/how ours is better?}
-
-While several contributions in the literature provide insights and contributions 
-on one aspect or another, we are not aware of a fully implemented architecture that effectively
-combines in a coherent manner all the points mentioned above.
-We do not claim that this represents the only architecture that can be designed
-to address such problems, nor that our approach fully addresses each and every
-of these issues. Our point here is that it is a fully implemented architecture that allowed us to
-identify, refine and combine a number of key cognitive abilities that we believe are necessary. Indeed, 
-we show here below that it provided us a mean to develop and explore a number of key features for HRI 
-in terms of models and algorithms and their articulation in an integrated decisional system for a 
-cognitive and interactive robot.
-
-\fixme{Discussion on deliberative architectures != cognitive architecture}: a \emph{cognitive
-architecture} is usually understood as an artificial yet principled model of (human)
-cognition, while we propose a practical approach to consistently combine a large
-set of independent cognitive processes by bridging them through high-level semantics.
+system builds upon the particular design of the interfaces between the
+cognitive components: the information streams use high-level semantics,
+represented as first-order logic statements. In that sense, our deliberative
+architecture is similar to projects like {\sc cram}~\cite{Beetz2010} or PEIS
+Ecology~\cite{saffiotti2005peis, daoutis2012cooperative}, with however a
+stronger emphasize on the representation and interaction with humans. We
+however distinguish ourselves from research on \emph{cognitive architectures}:
+cognitive architectures are usually understood as an artificial yet principled
+model of (human) cognition. While some have been deployed on autonomous robots,
+like HAMMER~\cite{demiris2006hierarchical} or ACT-R/E~\cite{trafton2013act},
+most are primarily concerned with the modeling of human cognition and are less
+focused on practical deployment on socially interactive robots.
+
+In that sense, our contribution in terms of architecture is a practical one:
+our integration model enables to consistently combine a large set of
+technically independent yet cognitively interdependent cognitive processes. We
+bridge them through explicit, human-level semantics, and we show that this
+results in a fully implemented system, effectively deployed on several
+platforms and in several real interaction scenarios.
+
+Besides, this work introduces several new contributions related to the
+representation and the management of humans in an autonomous robotic system.
+Specifically, we mentioned in the introduction the following four points:
+
+\begin{itemize}
+
+	\item our system achieves multi-modal and interactive grounding in complex
+		real environments involving one or several humans and a robot;
+
+	\item it supports a distributed computation of symbolic knowledge for
+		situated dialogue, thanks to the combination of perspective taking,
+		affordances computation and logical inference;
+
+	\item it provides generic mechanisms for the robot to reason about the mental state
+		of its human partners;
+
+	\item and, by reusing the computed affordances and inference, it generates,
+		monitors and takes part to human-robot shared plans;
+
+\end{itemize}
+
+While several contributions in the literature provide insights and
+contributions on one aspect or another (references are in the corresponding
+subsections), we are not aware of a fully implemented architecture that
+effectively combines in a coherent manner all these points. Our approach
+allowed us to identify, refine and combine a number of key cognitive abilities
+that we believe are needed for successful and autonomous social human-robot
+interaction. The novelty and relevance of this contribution to HRI is
+underlined by the range of multi-disciplinary collaborations and studies that
+have been made possible by our
+architecture~\cite{dautenhahn2006may,koay2007exploratory,Ros2010b,dehais2011physiological,ferreira2015users,GharbiROMAN2015}.
 

### 2016-06-30 paper.tex
Update on Overleaf.

--- 2016-06-30 paper.tex
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Multi-Modal Communication}
+\subsection{Multi-Modal Communication and Situated Dialog}
 \label{sect|com}
@@
 one hand, to schedule the different streams and, on the other hand, as an
-additional cost function.
+additional cost function. Cost here is not used here to estimate only standard time and energy consumption but
+to estimate a number factors that measure the satisfaction of the human in terms of acceptabiliof human based not
 

### 2016-06-30 paper.tex
Update on Overleaf.

@@
 additional cost function. Cost here is not used here to estimate only standard time and energy consumption but
-to estimate a number factors that measure the satisfaction of the human in terms of acceptabiliof human based not
+to estimate a number factors that measure the satisfaction of the human in terms of acceptability and legibility of robot of human based not
 

### 2016-06-30 paper.tex
Update on Overleaf.

--- 2016-06-30 paper.tex
@@
 additional cost function. Cost here is not used here to estimate only standard time and energy consumption but
-to estimate a number factors that measure the satisfaction of the human in terms of acceptabiliof human based not
+to estimate a number factors that measure the satisfaction of the human in terms of acceptability and legibility of robot behavior.
 

### 2016-06-30 paper.tex
Update on Overleaf.

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Multi-Modal Communication and Situated Dialog}
+\subsection{Multi-Modal Communication and Situated Dialogue}
 \label{sect|com}
@@
 
-HATP includes mechanisms called \emph{social rules} to filter plans so as to
-keep only those considered as suitable for human-robot interaction. The planner
+HATP includes mechanisms called \emph{social rules} to promote plans that are
+considered as suitable for human-robot interaction. The planner
 allows for the specification of the following filtering criteria:
@@
 
-\emph{Effort balancing}: avoids plans where efforts are
+\emph{Effort balancing and human desires}: avoids plans where efforts are
 not fairly distributed among the agents taking part to the plan.  It is

### 2016-06-30 paper.tex
Update on Overleaf.

@@
 
-\emph{Effort balancing and human desires}: avoids plans where efforts are
-not fairly distributed among the agents taking part to the plan.  It is
+\emph{Effort balancing with respect of human desires}: avoids plans where efforts are
+not distributed among the agents taking part to the plan following the human preferences.  It is
 indeed sometimes beneficial to balance efforts between the human and
-the robot,
-
-\emph{Simplicity}: avoids plans with too many interdependencies
+the robot. Sometimes the human wants to do more or, on the contrary, prefers to leave the 
+robot do all what it can.
+
+\emph{Simplicity}: promotes plans that limit  too many interdependencies
 between the actions of agents mentioned in the plan, as a problem with

### 2016-06-30 paper.tex
Mention KeJia in the conclusion

@@
 knowledge representation systems, like {\sc KnowRob}~\cite{Tenorth2009a} or PEIS
-K\&R~\cite{Daoutis2009}) has multiple advantages. First the design of this part
-of the ontology is generally difficult: it pertains to abstract concepts whose
-mutual relations comes to philosophical debates. The upper taxonomy of {\sc
-OpenCyc} represents a relative consensus, at least within the semantic Web
-community. Then, because it is a well established project with numerous links to
-other on-line databases (like Wikipedia or WordNet), the reuse of key {\sc
-OpenCyc} concepts ensures that the knowledge stored by the robot can be shared
-or extended with well-defined semantics. The concept of \emph{Object} represents
-a typical case of ambiguous meaning: in everyday conversation, an object is a
-relatively small physical thing, that can be typically manipulated. A human is
-not usually considered as an object. {\sc OpenCyc} however precisely defines an
-object as anything at least \emph{partially tangible}. This includes obviously
-the humans, and actually many other entities that would not be commonly said to
-be objects (the Earth for instance).\footnote{In such cases of discrepancy
-between {\sc OpenCyc} concepts and human terminology, we manually label the {\sc
-OpenCyc} concepts with appropriate human names: for instance, the OpenRobots
-Ontology associates the label ``\emph{object}'' to the concept
+K\&R~\cite{daoutis2012cooperative}) has multiple advantages. First the design of
+this part of the ontology is generally difficult: it pertains to abstract
+concepts whose mutual relations comes to philosophical debates. The upper
+taxonomy of {\sc OpenCyc} represents a relative consensus, at least within the
+semantic Web community. Then, because it is a well established project with
+numerous links to other on-line databases (like Wikipedia or WordNet), the reuse
+of key {\sc OpenCyc} concepts ensures that the knowledge stored by the robot can
+be shared or extended with well-defined semantics. The concept of \emph{Object}
+represents a typical case of ambiguous meaning: in everyday conversation, an
+object is a relatively small physical thing, that can be typically manipulated.
+A human is not usually considered as an object. {\sc OpenCyc} however precisely
+defines an object as anything at least \emph{partially tangible}. This includes
+obviously the humans, and actually many other entities that would not be
+commonly said to be objects (the Earth for instance).\footnote{In such cases of
+discrepancy between {\sc OpenCyc} concepts and human terminology, we manually
+label the {\sc OpenCyc} concepts with appropriate human names: for instance, the
+OpenRobots Ontology associates the label ``\emph{object}'' to the concept
 \concept{cyc:Artifact} instead of the concept \concept{cyc:PartiallyTangible}.
@@
 
-\emph{Simplicity}: promotes plans that limit  too many interdependencies
+\emph{Simplicity}: promotes plans that limit as much as possible the interdependencies
 between the actions of agents mentioned in the plan, as a problem with
@@
 In the current implementation, the social rules are effectively
-implemented by looking through all the plans produced and filtering
-out the ones that do not meet the desired requirements. In the
-future we intend to study algorithms that do such filtering online,
-rather than after initial solutions are found.
+implemented as filters and that are applied whenever a plan is produced.
+In the future we intend to extend the plan-search algorithm and integrate the social rules 
+in the process itself.
 
@@
 
-As HATP is a generic symbolic task planner and do not enforce any abstraction
+As HATP is a generic symbolic task planner and does not enforce any abstraction
 level for the planning domain, we were able to design a domain built from
@@
 in Figure~\ref{fig|archi}. {\sc Shary}~\cite{clodic2008shary} is the main one,
-written in the \emph{Procedural Reasoning System} language~\cite{Ingrand1996}.
+written in the \emph{Procedural Reasoning System (PRS)} language~\cite{Ingrand1996}.
 We have also developed the Python-based {\sc pyRobots}~\cite{lemaignan2015pyrobots} that
@@
 actions execution, as well as monitoring of those achieved by its human
-partner. One of the key design goal is to build such abilities in a generic
+partner. One of the key design goals is to build such abilities in a generic
 way, and to provide several levels of parametrisation allowing to adapt to
@@
 to take into account not only the task achievement but also the communication
-and monitoring needed to support interactive task
+and monitoring acts needed to support human-robot interactive task
 achievement~\cite{Rich1997,Sidner2005} in a flexible manner. {\sc Shary} allows
@@
 \emph{desire} is produced (this is delegated to perception modules), he can
-focus on the semantic of the desire.
+focus on the semantics of the desire.
 
@@
 
-\subsubsection{Desires and Experiences}
+\subsubsection{Complying with Human Desires and Experiences}
 \label{sect:desires}
@@
 controller and the HATP symbolic task planner. In this scenario, the human and
-the robot try to cooperatively remove objects from a table, and behaviours and
+the robot are in a situation when they need to cooperatively remove objects from a table, and behaviours and
 motions are fully planned and then executed.
@@
 The integration of these components in a consistent, working and observable
-system builds upon the particular design of the interfaces between the
-cognitive components: the information streams use high-level semantics,
-represented as first-order logic statements. In that sense, our deliberative
-architecture is similar to projects like {\sc cram}~\cite{Beetz2010} or PEIS
-Ecology~\cite{saffiotti2005peis, daoutis2012cooperative}, with however a
-stronger emphasize on the representation and interaction with humans. We
-however distinguish ourselves from research on \emph{cognitive architectures}:
-cognitive architectures are usually understood as an artificial yet principled
-model of (human) cognition. While some have been deployed on autonomous robots,
-like HAMMER~\cite{demiris2006hierarchical} or ACT-R/E~\cite{trafton2013act},
-most are primarily concerned with the modeling of human cognition and are less
-focused on practical deployment on socially interactive robots.
+system builds upon the particular design of the interfaces between the cognitive
+components: the information streams use high-level semantics, represented as
+first-order logic statements. In that sense, our deliberative architecture is
+similar to projects like {\sc cram}~\cite{Beetz2010}, KeJia~\cite{Chen2010} or
+PEIS Ecology~\cite{saffiotti2005peis, daoutis2012cooperative}, with however a
+stronger emphasize on the representation and interaction with humans. We however
+distinguish ourselves from research on \emph{cognitive architectures}: cognitive
+architectures are usually understood as an artificial yet principled model of
+(human) cognition. While some have been deployed on autonomous robots, like
+HAMMER~\cite{demiris2006hierarchical} or ACT-R/E~\cite{trafton2013act}, most are
+primarily concerned with the modeling of human cognition and are less focused on
+practical deployment on socially interactive robots.
 

### 2016-06-30 paper.tex
Update on Overleaf.

@@
 knowledge representation systems, like {\sc KnowRob}~\cite{Tenorth2009a} or PEIS
-K\&R~\cite{daoutis2012cooperative}) has multiple advantages. First the design of
-this part of the ontology is generally difficult: it pertains to abstract
-concepts whose mutual relations comes to philosophical debates. The upper
-taxonomy of {\sc OpenCyc} represents a relative consensus, at least within the
-semantic Web community. Then, because it is a well established project with
-numerous links to other on-line databases (like Wikipedia or WordNet), the reuse
-of key {\sc OpenCyc} concepts ensures that the knowledge stored by the robot can
-be shared or extended with well-defined semantics. The concept of \emph{Object}
-represents a typical case of ambiguous meaning: in everyday conversation, an
-object is a relatively small physical thing, that can be typically manipulated.
-A human is not usually considered as an object. {\sc OpenCyc} however precisely
-defines an object as anything at least \emph{partially tangible}. This includes
-obviously the humans, and actually many other entities that would not be
-commonly said to be objects (the Earth for instance).\footnote{In such cases of
-discrepancy between {\sc OpenCyc} concepts and human terminology, we manually
-label the {\sc OpenCyc} concepts with appropriate human names: for instance, the
-OpenRobots Ontology associates the label ``\emph{object}'' to the concept
+K\&R~\cite{Daoutis2009}) has multiple advantages. First the design of this part
+of the ontology is generally difficult: it pertains to abstract concepts whose
+mutual relations comes to philosophical debates. The upper taxonomy of {\sc
+OpenCyc} represents a relative consensus, at least within the semantic Web
+community. Then, because it is a well established project with numerous links to
+other on-line databases (like Wikipedia or WordNet), the reuse of key {\sc
+OpenCyc} concepts ensures that the knowledge stored by the robot can be shared
+or extended with well-defined semantics. The concept of \emph{Object} represents
+a typical case of ambiguous meaning: in everyday conversation, an object is a
+relatively small physical thing, that can be typically manipulated. A human is
+not usually considered as an object. {\sc OpenCyc} however precisely defines an
+object as anything at least \emph{partially tangible}. This includes obviously
+the humans, and actually many other entities that would not be commonly said to
+be objects (the Earth for instance).\footnote{In such cases of discrepancy
+between {\sc OpenCyc} concepts and human terminology, we manually label the {\sc
+OpenCyc} concepts with appropriate human names: for instance, the OpenRobots
+Ontology associates the label ``\emph{object}'' to the concept
 \concept{cyc:Artifact} instead of the concept \concept{cyc:PartiallyTangible}.
@@
 The integration of these components in a consistent, working and observable
-system builds upon the particular design of the interfaces between the cognitive
-components: the information streams use high-level semantics, represented as
-first-order logic statements. In that sense, our deliberative architecture is
-similar to projects like {\sc cram}~\cite{Beetz2010}, KeJia~\cite{Chen2010} or
-PEIS Ecology~\cite{saffiotti2005peis, daoutis2012cooperative}, with however a
-stronger emphasize on the representation and interaction with humans. We however
-distinguish ourselves from research on \emph{cognitive architectures}: cognitive
-architectures are usually understood as an artificial yet principled model of
-(human) cognition. While some have been deployed on autonomous robots, like
-HAMMER~\cite{demiris2006hierarchical} or ACT-R/E~\cite{trafton2013act}, most are
-primarily concerned with the modeling of human cognition and are less focused on
-practical deployment on socially interactive robots.
+system builds upon the particular design of the interfaces between the
+cognitive components: the information streams use high-level semantics,
+represented as first-order logic statements. In that sense, our deliberative
+architecture is similar to projects like {\sc cram}~\cite{Beetz2010} or PEIS
+Ecology~\cite{saffiotti2005peis, daoutis2012cooperative}, with however a
+stronger emphasize on the representation and interaction with humans. We
+however distinguish ourselves from research on \emph{cognitive architectures}:
+cognitive architectures are usually understood as an artificial yet principled
+model of (human) cognition. While some have been deployed on autonomous robots,
+like HAMMER~\cite{demiris2006hierarchical} or ACT-R/E~\cite{trafton2013act},
+most are primarily concerned with the modeling of human cognition and are less
+focused on practical deployment on socially interactive robots.
 

### 2016-06-30 paper.tex
Update on Overleaf.

--- 2016-06-30 paper.tex
@@
 controller and the HATP symbolic task planner. In this scenario, the human and
-the robot are in a situation when they need to cooperatively remove objects from a table, and behaviours and
+the robot are in a context when they need to cooperatively remove objects from a table. Robot behaviours and
 motions are fully planned and then executed.
@@
 \emph{Jido}, a single-arm mobile manipulator, is observing while they carry
-over boxes (Figure~\ref{fig|vpt}), and answers questions pertaining to the
+over boxes (Figure~\ref{fig|vpt}), and answers questions concerning the
 location of specific objects.  This study focuses on multi-modal, interactive
-grounding only, and the robot does not actually perform any action besides
+grounding only. The robot essentially observes, builds and maintains knowledge about its human
+partners perspectives and affordances and does not actually perform any action besides
 verbal interaction and simple head movements.
@@
 In this second situation, User B enters the living room without knowing where
-User A had placed the video tapes (Figure~\ref{fig|pointing}). So he first asks
-Jido: ``What's in the box?''. The robot first need to ground the word ``box''.
+User A has placed the video tapes (Figure~\ref{fig|pointing}). So he first asks
+Jido: ``What's in the box?''. The robot first needs to ground the word ``box''.
 Similar to the previous situation, two boxes are visible: \concept{find(?obj
@@
 Oro} server is used in conjunction with the HATP symbolic task planner and the
-{\sc Shary} execution controller to produce and execute a shared plan. The task
+{\sc Shary} execution controller to synthesise and execute a shared plan. The task
 consists in cooperatively cleaning a table by moving objects into their target
@@
 see as the main challenges that human-robot interaction brings to Artificial
-Intelligence.  We discuss first how embodied cognition is a very real challenge
+Intelligence.  We discuss first how embodied cognition is a an essential challenge
 in human-robot interaction; we rephrase then the requirements of joint actions
@@
 made intrinsic to the system?''. This issue has a major practical importance in
-robotic: for a robot to be both endowed with a symbolic representational and
+robotics: for a robot to be both endowed with a symbolic representational and
 reasoning system, and able to \emph{act} in the physical world, it must ground

### 2016-06-30 paper.tex
Commented out the section on 'Object Position Hypotheses' as it was never really properly finished

@@
 
-\paragraph{Hypotheses on Objects States and Positions}
-
-Occlusions, be they purposeful (when an agent hides an object behind or inside
-another one) or not, commonly lead to difficulties in perceiving and tracking
-objects in certain states. To mitigate this issue, {\sc Spark} also models the
-possible symbolic states of objects (whether the object is on a furniture, in an
-agent hand, in a container, etc.) with an associated probability distribution
-based on the robot's perception of what has happened since the object was last
-seen.  This probabilistic model of the environment is not currently integrated
-with the knowledge base, and is instead directly queried by the execution
-controllers when necessary.
+%\paragraph{Hypotheses on Objects States and Positions}
+%
+%Occlusions, be they purposeful (when an agent hides an object behind or inside
+%another one) or not, commonly lead to difficulties in perceiving and tracking
+%objects in certain states. To mitigate this issue, {\sc Spark} also models the
+%possible symbolic states of objects (whether the object is on a furniture, in an
+%agent hand, in a container, etc.) with an associated probability distribution
+%based on the robot's perception of what has happened since the object was last
+%seen.  This probabilistic model of the environment is not currently integrated
+%with the knowledge base, and is instead directly queried by the execution
+%controllers when necessary.
 

### 2016-06-30 rebuttal2.tex
WIP on the rebuttal

@@
 
-Thank you for coming back to us, we are glad that the special issue is still
-going on.
-
-We have hopefully addressed the remaining concerns raised during the second
-round of reviews.
+Thank you for coming back to us, we are certainly glad that the special issue is
+still going on.  We have hopefully addressed the remaining concerns raised
+during the second round of reviews.  Besides, we have updated many bibliography
+references (and added new ones) to reflect the continous progress of the field
+since the initial submission in 2013.
 
@@
 To facilitate the review process, we attach an annotated version of the
-manuscript where all these changes have been highlighted.
+manuscript where all the changes since the last submission have been highlighted.
 
-\subsection*{Relation to the state of the art}
+\subsection*{Clarifiying the contribution}
 
-We initially made the choice to discuss the existing litterature in each of the
-``cognitive skills'' sections, for readability reasons. We however acknowledge
-that a higher-level discussion of our work with respect to the state of the art
-of human-robot interaction systems would be useful.
+We initially made the choice to discuss the existing litterature in relation to
+our contribution in each of the ``cognitive skills'' sections, for readability
+reasons. We however acknowledge that a higher-level discussion of our work with
+respect to the state of the art of human-robot interaction systems would be
+helpful.
 
-To this end, we have split Section 5 into a discussion (Section 5) and a
-conclusion (Section 6) that states clearly our contribution (including
-references to all our previous publications -- Section 6.1), followed by a
-comparative discussion with the related approaches (Section 6.2).
+To this end, we have:
+
+\begin{itemize}
+    \item added a summary of the key contribution at the end of the introduction
+        (Section 1.2)
+    \item split Section 5 into a discussion (Section 5) and a
+conclusion (Section 6) that clearly states our contribution (including, per your
+request, references to all our previous publications -- Section 6.1) and relate
+it to the start of the art.
+\end{itemize}
 
@@
 
-Reviewer 2's comments have all been addressed in the manuscript.
+Reviewer 2 suggested that the discussion carried in section 5.3 "Knowledge and
+Robotics" was out-of-scope for this article. After a light re-organization,
+we have eventually decided to maintain it as we believe it is helpful in
+pointing some of the limits of our approach.
+
+The other comments from Reviewer 2 have all been addressed in the manuscript.
 We highlight hereafter the main changes.
@@
 \begin{itemize}
-    \item examples have been added to Section 2.2 to make it easier to follow,
+    \item the re-organization of section 5 into section 5 'Discussion' and
+        section 6 'Conclusion' hopefully addresses the concern that the last part of the
+        paper was relatively disconnected from the core of the article;
+    \item examples have been added to Section 2.2 to make it easier to follow;
     \item we have clarified how the system manages discrepancies between OpenCyc
         terminology and human natural language by explicit manual labelling
-        (Section 2.2.2)
+        (Section 2.2.2);
     \item We have precised that our approach of transient geometric situation assessment
-        prevents reasoning on the situation history (Section 3.2)
-    \item We have clarified how the {\tt nextTo} spatial relation is computed.
+        prevents reasoning on the situation history (Section 3.2);
+    \item We have clarified how the {\tt nextTo} spatial relation is computed;
+    \item We have removed the paragraph on state probabilities (Section 3.2.2)
+        as it referred to an unfinished research proposal;
     \item We have clarified how primitive action recognition is used in
         combination with situation assessment and task planning to monitor the
-        human activities (Section 3.2.3)
+        human activities (Section 3.2.3);
     \item The implicit aspect of multi-modality during dialogue processing has
-        been made clearer (Section 3.3.2)
+        been made clearer (Section 3.3.2).
 \end{itemize}

### 2016-06-30 paper.tex
Typos

@@
 which mandates common-sense knowledge and the representation of possibly
-divergent mental models. This article is an attempt to characterize these challenges
+divergent mental models. This article is an attempt to characterise these challenges
 and to exhibit a set of key decisional issues that need to be addressed for a
@@
 what can be currently achieved with our robots.  The
-Sections~\ref{sect|discussion} and~\ref{sect|conclusion} finally summarize our
+Sections~\ref{sect|discussion} and~\ref{sect|conclusion} finally summarise our
 main contributions and discuss the key challenges that human-robot interaction
@@
     with a thick border. The links between components depicted on the figure
-    underline the central role of the knowledge base: many of the datastreams are
+    underline the central role of the knowledge base: many of the data streams are
     actually symbolic statements exchanged through this semantic blackboard.}
@@
 manipulation: software components communicate with each other using first-order
-logic statements organized into ontologies and whose semantics are close to the
+logic statements organised into ontologies and whose semantics are close to the
 ones manipulated by humans.
@@
 Section~\ref{sect|sit-ass}). For instance, whenever a client queries KnowRob to
-know if \stmt{object1 isOn object2}, KnowRob calls a geometric reasonner to
+know if \stmt{object1 isOn object2}, KnowRob calls a geometric reasoner to
 evaluate if this specific spatial relation holds at that point in time. With our
@@
 This design choice trades scalability for explicit reasoning: at any time, the
-full belief state is explicited (which may be computationally expensive), and
+full belief state is made explicit (which may be computationally expensive), and
 therefore provides the reasoner with the largest possible inference domain. This
@@
 As previously said, the logical statements exchanged between the deliberative
-components are organized within the {\sc Oro} knowledge base into an ontology.
+components are organised within the {\sc Oro} knowledge base into an ontology.
 The conceptualization (\ie the system of concepts or \emph{TBox}) of this
@@
 The \emph{OpenRobots} common-sense ontology represents the statically asserted
-part of the ontology. It has been designed from two requirements: beining
+part of the ontology. It has been designed from two requirements: being
 practical (\ie covering our experimental needs) and conforming as much as
@@
 are simultaneously asserted, yet contradict when taken together. Since the two
-knowledge models are realized as two independent ontologies, the contradiction
+knowledge models are realised as two independent ontologies, the contradiction
 does not actually occur and both the models remain logically consistent.
@@
 sensors (usually, a combination of vision-based tracking of 2D fiducial markers
-to identify and localize objects, and Kinect-based skeleton tracking of humans,
+to identify and localise objects, and Kinect-based skeleton tracking of humans,
 optionally assisted by motion capture to accurately track the head motion, which
@@
 Monitoring human activity is needed by the execution controllers to track the
-engagement of the human and the progress of her actions and also to synchtronize seamlessly
+engagement of the human and the progress of her actions and also to synchronise seamlessly
 its own actions with the human actions. Full human action
@@
 geometric heuristics on human hand trajectories and possible objects placements
-to recognize simple atomic actions. Those primitive actions are assessed through
+to recognise simple atomic actions. Those primitive actions are assessed through
 monitoring situations like ``an empty hand is close to an object on a table''
@@
 perception, made possible by the use of fiducial markers.  Each object receives
-a unique tag which enables us to accurately localize it in 3D and prevent
+a unique tag which enables us to accurately localise it in 3D and prevent
 recognition ambiguities that would be then reflected in the knowledge base.
@@
 computing symbolic properties from the geometric model, richer symbolic
-descriptions are available to the system to verbalize or discriminate entities.
+descriptions are available to the system to verbalise or discriminate entities.
 For instance, the robot may compute that one bottle is next to a glass, while
@@
 additional cost function. Cost here is not used here to estimate only standard time and energy consumption but
-to estimate a number factors that measure the satisfaction of the human in terms of acceptability and legibility of robot behavior.
+to estimate a number factors that measure the satisfaction of the human in terms
+of acceptability and legibility of robot behaviour.
 
@@
 
-Objects are perceived solely through 2D fiducial markers sticked on them, and humans are
+Objects are perceived through 2D fiducial markers attached to them, and humans are
 tracked through motion capture. The robot knowledge base is initialised with
@@
 CARDBOARD\_BOX} as being pointed at, as well as looked at, by the human and updates the
-ontology with this new information. The reasonner applies a rule available in the common-sense
+ontology with this new information. The reasoner applies a rule available in the common-sense
 ontology \stmt{pointsAt(?ag, ?obj) $\land$ looksAt(?ag, ?obj) $\to$
@@
 Oro} server is used in conjunction with the HATP symbolic task planner and the
-{\sc Shary} execution controller to synthesise and execute a shared plan. The task
+{\sc Shary} execution controller to produce and execute a shared plan. The task
 consists in cooperatively cleaning a table by moving objects into their target
@@
 
-Figure~\ref{fig|cleantable-timeline} walks through a simplifed version of the whole
+Figure~\ref{fig|cleantable-timeline} walks through a simplified version of the whole
 task. It depicts a run with a single tape on a table. The tape is
@@
 see as the main challenges that human-robot interaction brings to Artificial
-Intelligence.  We discuss first how embodied cognition is a an essential challenge
+Intelligence.  We discuss first how embodied cognition is an essential challenge
 in human-robot interaction; we rephrase then the requirements of joint actions
@@
 build explanations to interpret unexpected perceptions (like an object that
-suddendly disappears). Because {\sc Spark} also tracks humans, which enables
+suddenly disappears). Because {\sc Spark} also tracks humans, which enables
 perspective-aware spatial reasoning, it can produce as well grounded symbolic
@@
 embeds human preferences and physical constraints that need to be accounted for
-when synthesizing robot motion or producing robot plans. This includes
+when synthesising robot motion or producing robot plans. This includes
 proxemics (human-robot distance) and associated issues (visibility) but also
@@
 plans of the human actions that enable the system to plan for joint human-robot
-tasks. It can also make use of social heuristics to optimize plans for social
+tasks. It can also make use of social heuristics to optimise plans for social
 acceptability. We briefly mention our human-aware motion and manipulation
@@
 PEIS Ecology~\cite{saffiotti2005peis, daoutis2012cooperative}, with however a
-stronger emphasize on the representation and interaction with humans. We however
+stronger emphasise on the representation and interaction with humans. We however
 distinguish ourselves from research on \emph{cognitive architectures}: cognitive

### 2016-06-30 paper.tex
Polish the abstract

@@
 \title{\LARGE \bf
-Artificial Cognition for Social Human-Robot Interaction:\\ An Implementation
+Artificial Cognition for Social Human-Robot Interaction:\\ An
+    Implementation
 }
@@
 
-We identify the needed individual and collaborative cognitive
-skills: geometric reasoning and situation assessment based on perspective-taking
-and affordance analysis; acquisition and representation of knowledge models for
-multiple agents (humans and robots, with their specificities); natural multi-modal dialogue;
-human-aware task planning; human-robot joint task achievement. The article
-discusses each of these abilities, presents a working implementation, and shows
-how they combine in a coherent and original deliberative architecture for
-human-robot interaction. Supported by experimental results, we eventually show
-how explicit knowledge management, both symbolic and geometric, proves to be
-instrumental to richer and more natural human-robot interactions by pushing for
-pervasive, human-level semantics within the robot's deliberative system.
+We identify the needed individual and collaborative cognitive skills: geometric
+reasoning and situation assessment based on perspective-taking and
+affordance analysis; acquisition and representation of knowledge models for
+multiple agents (humans and robots, with their specificities); situated,
+natural and multi-modal dialogue; human-aware task planning; human-robot
+joint task achievement. The article discusses each of these abilities,
+presents working implementations, and shows how they combine in a coherent
+and original deliberative architecture for human-robot interaction.
+Supported by experimental results, we eventually show how explicit knowledge
+management, both symbolic and geometric, proves to be instrumental to richer
+and more natural human-robot interactions by pushing for pervasive,
+human-level semantics within the robot's deliberative system.
 
@@
 
-The robot controller (with the help of a task planner) decides what action to
+The robot controller \cite{clodic2014key} (with the help of a task planner) decides what action to
 execute next, and who should perform it, from the robot or the human (or both in
-case of a joint action). It finally controls and monitors its execution. The
+case of a collaborative action such as a handover \cite{Mainprice2012,waldhart2015planning}), how it should achieved and what signals should
+be sensed and/or produced by the robot to facilitate human-robot joint action
+\cite{sebanz2006joint,knoblich20113,pacherie2012phenomenology,vesper2010minimal}.
+It finally controls and monitors its execution. The
 operation continues until the goal is achieved, is declared unachievable or is
@@
 symbolic task planner HATP, and finally the main features of our execution
-controllers {\sc Shary} and {\sc pyRobots}. Note that greater details on each of
+controllers {\sc Shary}\cite{clodic2008shary} and {\sc pyRobots}. Note that greater details on each of
 these modules can be found in their respective publications (the corresponding
@@
 and monitoring acts needed to support human-robot interactive task
-achievement~\cite{Rich1997,Sidner2005} in a flexible manner. {\sc Shary} allows
+achievement~\cite{Rich1997,Sidner2005,sebanz2006joint} in a flexible manner. {\sc Shary} allows
 to bind action execution to \emph{communication policies} in order to produce
-signals towards the human and to react to human actions. The
+multi-modal signals towards the human and to react to human actions and signals. The
 \emph{communication act} is the central concept in this formalism. It
 represents an information exchange between the two agents and plays the role of
-a transition condition. This exchange can be realised through dialogue, by an
+a transition condition. This exchange can be realised through speech, through an
 expressive motion or a combination of the two. It enables each agent to
@@
 Shary} monitoring processes and {\sc Oro} inference
-mechanisms~\cite{fiore2014}. \fixme{Rachid, reviewer 2 asks for more details
-about this process -> M. Fiore work?}
+mechanisms.
+
+Two recent developments have been added that improve robot adaptation to human action at the action and at the plan level.
+In~\cite{fiore2014}, the system is able to estimate its human intentions in collaborative tasks and to choose appropriate actions, using a set of POMDP models.
+In~\cite{devin2016implemented}, a framework is proposed which allows the robot to estimate its human partner mental state not only about the environment but also about the state of goals, plans and actions and to take them into account when executing human-robot shared plans.
 
@@
 
-\subsubsection{Action Execution and Monitoring}\label{sec:action}
+\subsubsection{Action Refinement, Execution and Monitoring}\label{sec:action}
 
@@
 The symbolic manipulation and navigation actions listed in
-Table~\ref{table|pyrobots_actions} are eventually executed by the execution
-controller with the help of a dedicated 3D motion planner (called {\sc Mhp},
-Figure~\ref{fig|archi}): robot placement and end-effectors trajectories are
-computed on-line to allow object manipulation that take into account task
-specific constraints and human postures, abilities and preferences.  We kindly
-refer the interested reader to~\cite{Sisbot2008, Mainprice2011, Pandey2011} for
-details of these techniques.
+Table~\ref{table|pyrobots_actions} are eventually further refined and executed by the execution
+controller with the help of a  set of dedicated human-aware geometric motion
+planning functions provided by a component (called {\sc Mhp},
+Figure~\ref{fig|archi}).
+
+These functions are designed to plan navigation~\cite{Sisbot2007} and
+manipulation~\cite{Mainprice2011,sisbot2012human} paths not only safe
+but also comfortable and “socially acceptable” by reasoning explicitly
+on human’s kinematics, vision field, posture and preferences
+\cite{Madhava2006,kruse2013human,Rios-MartinezSL15}.  Besides, when necessary, they compute pertinent robot
+and objects placements relative to the human\cite{Pandey2011}
+
+Also, when an action needs motion of the human and the robot, it is
+possible to plan for both of them in order to allow the robot to take
+initiative by computing automatically where the interaction might
+preferably take place\cite{Mainprice2012,waldhart2015planning}, 
+thus decreasing the cognitive weight of interaction on human
+side. This is somehow the geometric counterpart of HATP.
 

### 2016-06-30 paper.tex
Update on Overleaf.

@@
 \title{\LARGE \bf
-Artificial Cognition for Social Human-Robot Interaction:\\ An
-    Implementation
+Artificial Cognition for Social Human-Robot Interaction:\\ An Implementation
 }
@@
 
-We identify the needed individual and collaborative cognitive skills: geometric
-reasoning and situation assessment based on perspective-taking and
-affordance analysis; acquisition and representation of knowledge models for
-multiple agents (humans and robots, with their specificities); situated,
-natural and multi-modal dialogue; human-aware task planning; human-robot
-joint task achievement. The article discusses each of these abilities,
-presents working implementations, and shows how they combine in a coherent
-and original deliberative architecture for human-robot interaction.
-Supported by experimental results, we eventually show how explicit knowledge
-management, both symbolic and geometric, proves to be instrumental to richer
-and more natural human-robot interactions by pushing for pervasive,
-human-level semantics within the robot's deliberative system.
+We identify the needed individual and collaborative cognitive
+skills: geometric reasoning and situation assessment based on perspective-taking
+and affordance analysis; acquisition and representation of knowledge models for
+multiple agents (humans and robots, with their specificities); natural multi-modal dialogue;
+human-aware task planning; human-robot joint task achievement. The article
+discusses each of these abilities, presents a working implementation, and shows
+how they combine in a coherent and original deliberative architecture for
+human-robot interaction. Supported by experimental results, we eventually show
+how explicit knowledge management, both symbolic and geometric, proves to be
+instrumental to richer and more natural human-robot interactions by pushing for
+pervasive, human-level semantics within the robot's deliberative system.
 

### 2016-06-30 paper.tex
Update on Overleaf.

--- 2016-06-30 paper.tex
@@
 
-We also use shared plans to verbalise the sequence of actions and to explain
-the human partner how a task may be shared~\cite{warnier2012when,fiore2016planning,milliez2016using}.
+The hierarchical structure of the shared plans obtained through the HTN refinement process provide a good basis for plan verbalization allowing the robot, when asked, to explain to its human partner how a task can be  shared~\cite{fiore2016planning,milliez2016using}.
 
@@
 \emph{Wasted time}: avoids plans where the human spends a lot
-of its time being idle,
+of her/his time being idle,
 

### 2016-06-30 paper.tex
Proof-read Rachid's latest changes

@@
 
-The robot controller \cite{clodic2014key} (with the help of a task planner) decides what action to
-execute next, and who should perform it, from the robot or the human (or both in
-case of a collaborative action such as a handover \cite{Mainprice2012,waldhart2015planning}), how it should achieved and what signals should
-be sensed and/or produced by the robot to facilitate human-robot joint action
-\cite{sebanz2006joint,knoblich20113,pacherie2012phenomenology,vesper2010minimal}.
+The robot controller (with the help of a task planner) decides what action to
+execute next~\cite{clodic2014key}, and who should perform it, from the robot or the human (or both in
+case of a collaborative action such as a handover~\cite{Mainprice2012,waldhart2015planning}), how it should achieved and what signals should
+be sensed and/or produced by the robot to facilitate human-robot joint action~\cite{sebanz2006joint,knoblich20113,pacherie2012phenomenology,vesper2010minimal}.
 It finally controls and monitors its execution. The
@@
 symbolic task planner HATP, and finally the main features of our execution
-controllers {\sc Shary}\cite{clodic2008shary} and {\sc pyRobots}. Note that greater details on each of
+controllers {\sc Shary} and {\sc pyRobots}. Note that greater details on each of
 these modules can be found in their respective publications (the corresponding
@@
 
-{\sc Shary}'s originality, as an execution control system, lies in its ability
-to take into account not only the task achievement but also the communication
+Execution control systems for social robots are expected 
+to account not only for the task achievement but also for the communication
 and monitoring acts needed to support human-robot interactive task
-achievement~\cite{Rich1997,Sidner2005,sebanz2006joint} in a flexible manner. {\sc Shary} allows
+achievement~\cite{Rich1997,Sidner2005,sebanz2006joint}. {\sc Shary} supports
+such mechanisms. It allows
 to bind action execution to \emph{communication policies} in order to produce
@@
 
-Two recent developments have been added that improve robot adaptation to human action at the action and at the plan level.
-In~\cite{fiore2014}, the system is able to estimate its human intentions in collaborative tasks and to choose appropriate actions, using a set of POMDP models.
-In~\cite{devin2016implemented}, a framework is proposed which allows the robot to estimate its human partner mental state not only about the environment but also about the state of goals, plans and actions and to take them into account when executing human-robot shared plans.
+Two recent developments have further improved the robot's ability to adapt to
+human action, both at the action level and at the plan level.
+In~\cite{fiore2014}, an extension is proposed for the
+system to estimate the intentions of the human during collaborative tasks using
+POMDP models. The controller adjusts then appropriately the chosen set of actions.
+In~\cite{devin2016implemented}, the controller is complemented with a framework
+that allows the robot to estimate the mental state of the human partner, not
+only related to the environment but also related to the state of goals, plans
+and actions. They are then accounted for by the execution controller.
 
@@
 controller with the help of a set of dedicated human-aware geometric motion
-planning functions provided by a component (called {\sc Mhp},
-Figure~\ref{fig|archi}).
+planning functions provided by a component called {\sc Mhp} (see Figure~\ref{fig|archi}).
 
 These functions are designed to plan navigation~\cite{Sisbot2007} and
-manipulation~\cite{Mainprice2011,sisbot2012human} paths not only safe
-but also comfortable and “socially acceptable” by reasoning explicitly
-on human’s kinematics, vision field, posture and preferences
-\cite{Madhava2006,kruse2013human,Rios-MartinezSL15}.  Besides, when necessary, they compute pertinent robot
-and objects placements relative to the human\cite{Pandey2011}
-
-Also, when an action needs motion of the human and the robot, it is
-possible to plan for both of them in order to allow the robot to take
-initiative by computing automatically where the interaction might
-preferably take place\cite{Mainprice2012,waldhart2015planning}, 
-thus decreasing the cognitive weight of interaction on human
-side. This is somehow the geometric counterpart of HATP.
+manipulation~\cite{Mainprice2011,sisbot2012human} paths not only safe but also
+comfortable and socially acceptable by reasoning explicitly on human’s
+kinematics, vision field, posture and
+preferences~\cite{Madhava2006,kruse2013human,Rios-MartinezSL15}. They also
+provide routines to compute spatial placements for robot and objects that obey
+constraints related to the interaction, like optimal mutual reachability
+or optimal visibility~\cite{Pandey2011}
+
+Finally, when an action requires the motion of both the human and the robot,
+{\sc Mhp} can plan for both of them in order for the robot to take
+the lead by automatically computing where the interaction might
+preferably take place~\cite{Mainprice2012,waldhart2015planning}.
+This can effectively smoothen the interaction by off-loading a part of the
+cognitive load of the interaction from the human to the robot.
+As such, this is the geometric counterpart of the HATP symbolic task
+planner.
 
@@
 acceptability. We briefly mention our human-aware motion and manipulation
-planner~\cite{Sisbot2008,Mainprice2011,Pandey2011}, and finally present two
-execution controller, the PRS-based {\sc Shary}~\cite{clodic2008shary,fiore2014}
+planner~\cite{Sisbot2007,Sisbot2008,Mainprice2011,Pandey2011,sisbot2012human,kruse2013human}, and finally present two
+execution controller, the PRS-based {\sc
+Shary}~\cite{clodic2008shary,fiore2014,clodic2014key}
 and the event-driven {\sc pyRobots}~\cite{lemaignan2015pyrobots}.

### 2016-06-30 paper.tex
Proof-read Rachid's latest changes

@@
 
-The hierarchical structure of the shared plans obtained through the HTN refinement process provide a good basis for plan verbalization allowing the robot, when asked, to explain to its human partner how a task can be  shared~\cite{fiore2016planning,milliez2016using}.
+The hierarchical structure of the shared plans obtained through the HTN
+refinement process provide a good basis for plan verbalization. Upon request, it allows the
+robot to explain to its human partner how a task can be
+shared~\cite{fiore2016planning,milliez2016using}.
 
@@
 \emph{Wasted time}: avoids plans where the human spends a lot
-of her/his time being idle,
+of her time being idle,
 

### 2016-06-30 paper.tex
Update on Overleaf.

@@
 Figure~\ref{plan_hatp1}, the robot first moves the bottle so that the human can
-immediately take it and put it into the bin, thus removing the human idle time.
+immediately take it and put it into the bin, thus reducing the human idle time.
 The resulting shared plan is considered better than the one depicted on
@@
 situations where divergent beliefs appear between the human and the robot.
-Preliminary results have been published in~\cite{warnier2012when}.
+Some preliminary results have been presented 
+in~\cite{warnier2012when,milliez2014framework}  where we consider
+divergent beliefs about the state of the world, and
+in~\cite{devin2016implemented} where the robot is able to deal with
+divergent beliefs related to the state of the task.
 

### 2016-07-01 paper.tex
Update on Overleaf.

@@
 developed dedicated robot controllers which integrate the whole stack introduced
-in Figure~\ref{fig|archi}. {\sc Shary}~\cite{clodic2008shary} is the main one,
+in Figure~\ref{fig|archi}. {\sc Shary}~\cite{clodic2009} is the main one,
 written in the \emph{Procedural Reasoning System (PRS)} language~\cite{Ingrand1996}.
@@
 achievement~\cite{Rich1997,Sidner2005,sebanz2006joint}. {\sc Shary} supports
-such mechanisms. It allows
+such mechanisms~\cite{Clodic2007smc}. It allows
 to bind action execution to \emph{communication policies} in order to produce
@@
 a transition condition. This exchange can be realised through speech, through an
-expressive motion or a combination of the two. It enables each agent to
+expressive motion or a combination of the two. It enables the robot as well as the human to
 communicate their beliefs about the task to be achieved, in order to share
@@
 mechanisms.
+In Shary, each action description contains not only the way to execute it and monitor it, but also the way to initiate it (e.g.
+in the case of a shared action, the way both agents need to commit to the task), to suspend (or cancel or stop) it (and for each case the way this transition has to be shared with the human partner if any).
+This is supported by a description of the beliefs of each involved agent concerning the action achievement process and a set of monitoring processes allowing to update it
 
@@
 
-\subsubsection{Action Refinement, Execution and Monitoring}\label{sec:action}
+\subsubsection{Human-Aware Action Refinement, Execution and Monitoring}\label{sec:action}
 
@@
 
-These five questions should not be considered independently from each others and
+These five questions should not be considered independently from each other and
 often require, on the contrary, to be dealt with in a single decision step. The
@@
 
-\fixme{Rachid: the reviewer 2 asks for a better connection between these general
-questions and the specific approaches that we take}
+They are in fact taken into account in most, if not in all, the
+components and at different levels of abstraction: at the shared plan
+level as well as at action refinement and execution level.
+
+Repeatedly, the choice of the next action, of who has to act (the human, the robot
+or both of them) and when to perform it is made by {\sc Shary} with
+the help of {\sc HATP}, based on perspective-taking and
+situation assessment ({\sc Spark} and {\sc Oro} as well as on
+estimation of the human mental state ({\sc Shary} with the help of
+{\sc Oro}).
+
+Where and how to perform an action is made by a full set of on-line
+planning functions provided by {\sc Mhp} that computes not only
+trajectories but also the pertinent placements and postures based not
+only on geometric information but also estimation of the human current
+mental state and preferences ({\sc Oro}).
 
@@
 architecture: knowledge is explicitly stored in one central and consistent
-repository of facts, accessible for all modules. It relies on a strict formalism
+repository of facts, accessible to all modules. It relies on a strict formalism
 (OWL statements), with a well defined vocabulary (stated in the common-sense
@@
 execution controller, the PRS-based {\sc
-Shary}~\cite{clodic2008shary,fiore2014,clodic2014key}
+Shary}~\cite{clodic2009,fiore2014,clodic2014key}
 and the event-driven {\sc pyRobots}~\cite{lemaignan2015pyrobots}.

### 2016-07-01 paper.tex
Update on Overleaf.

@@
   \path (lowlevel) edge [->] (spark);
-  \path (lowlevel.east) edge [<-, bend right=80, looseness=1.2] node[label] {atomic\\actions} (shary.east);
+  \path (lowlevel.east) edge [<-, bend right=80, looseness=1.2] node[label] {elementary\\actions} (shary.east);
 
@@
 geometric heuristics on human hand trajectories and possible objects placements
-to recognise simple atomic actions. Those primitive actions are assessed through
+to recognise simple elemntary actions. Those primitive actions are assessed through
 monitoring situations like ``an empty hand is close to an object on a table''
@@
 
-Like most robotic architectures, actions are split into \emph{atomic actions}
+Like most robotic architectures, actions are split into \emph{elementary actions}
 that are combined into \emph{tasks}. Tasks are created either statically or dynamically.
@@
 Our other controller, {\sc Shary}, relies on the symbolic task planner to
-dynamically generate, at run-time, suitable sequence of atomic actions.
+dynamically generate, at run-time, suitable sequence of elementary actions.
 
@@
 must be true), and calling the 3D motion planner (column \emph{Actions},
-left). The motion planner returns two atomic actions ({\tt PICK\_GOTO} followed
+left). The motion planner returns two elementary actions ({\tt PICK\_GOTO} followed
 by {\tt TAKE\_TO\_FREE}) that the controller executes (by first reaching for
@@
 \emph{embodied} artificial intelligence, and this dimension is especially
-prevalent in human-robot interaction, where agents have to share and
+prevalent in human-robot interaction, where the robot has to share and
 collaborate in a joint physical environment.  This leads to a tight coupling
@@
 
-We have presented in this paper a large deliberative architecture designed for
+We have presented in this paper an instance of a complete deliberative architecture designed for
 social robots. While most of its sub-components have been independently
@@
 Artificial Intelligence. We hope that this contribution helps with clarifying
-some of these challenges.
+of these challenges and making them concrete decisional problems.
 
@@
 
-This work has been partially supported by EU FP7 ``SAPHARI'' (grant ICT-287513),
-and the EU H2020 Marie Sklodowska-Curie Actions project DoRoThy (grant 657227).
+This work has been partially supported by EU FP7 "CHRIS" (grant 215805) and ``SAPHARI'' (grant ICT-287513) projects, and the EU H2020 Marie Sklodowska-Curie Actions project DoRoThy (grant 657227).
 

### 2016-07-01 paper.tex
Typos

@@
 geometric heuristics on human hand trajectories and possible objects placements
-to recognise simple elemntary actions. Those primitive actions are assessed through
+to recognise simple elementary actions. Those primitive actions are assessed through
 monitoring situations like ``an empty hand is close to an object on a table''
@@
 Artificial Intelligence. We hope that this contribution helps with clarifying
-of these challenges and making them concrete decisional problems.
+these challenges and making them concrete decisional problems.
 
@@
 
-This work has been partially supported by EU FP7 "CHRIS" (grant 215805) and ``SAPHARI'' (grant ICT-287513) projects, and the EU H2020 Marie Sklodowska-Curie Actions project DoRoThy (grant 657227).
+This work has been partially supported by EU FP7 CHRIS (grant 215805) and
+SAPHARI (grant ICT-287513) projects, and the EU H2020 Marie Sklodowska-Curie
+Actions project DoRoThy (grant 657227).
 

### 2016-07-01 paper.tex
Proof-reading Aurelie + Rachid

@@
 real-time task-based situation assessment achieved by the combination of {\sc
-Shary} monitoring processes and {\sc Oro} inference
-mechanisms.
-In Shary, each action description contains not only the way to execute it and monitor it, but also the way to initiate it (e.g.
-in the case of a shared action, the way both agents need to commit to the task), to suspend (or cancel or stop) it (and for each case the way this transition has to be shared with the human partner if any).
-This is supported by a description of the beliefs of each involved agent concerning the action achievement process and a set of monitoring processes allowing to update it
+Shary} monitoring processes and {\sc Oro} inference mechanisms. In {\sc Shary},
+each action description contains not only how to execute it and monitor it,
+but also how to initiate it (\eg in the case of a shared action, how
+both agents need to commit to the task) and how to suspend (or cancel) it. In
+each of these cases, the action description also makes explicit how the
+transition has to be communicated to the human partner if necessary. This is
+supported by a description of the beliefs of each of the agents
+regarding the action achievement process, as well as a set of so-called
+\emph{monitoring processes} that update them.
 
@@
 often require, on the contrary, to be dealt with in a single decision step. The
-human-aware task and motion planners which we have built, are instances of
+human-aware task and motion planners which we have built are instances of
 systems which have been designed to deal with such intricate decision issues.
 
-They are in fact taken into account in most, if not in all, the
-components and at different levels of abstraction: at the shared plan
-level as well as at action refinement and execution level.
-
-Repeatedly, the choice of the next action, of who has to act (the human, the robot
-or both of them) and when to perform it is made by {\sc Shary} with
-the help of {\sc HATP}, based on perspective-taking and
-situation assessment ({\sc Spark} and {\sc Oro} as well as on
-estimation of the human mental state ({\sc Shary} with the help of
-{\sc Oro}).
-
-Where and how to perform an action is made by a full set of on-line
-planning functions provided by {\sc Mhp} that computes not only
-trajectories but also the pertinent placements and postures based not
-only on geometric information but also estimation of the human current
-mental state and preferences ({\sc Oro}).
+They are in fact considered in each of the components, at different levels of
+abstraction (from the abstract shared plan level to the action refinement and
+execution level).  At every interaction step, the choice of the next action, of
+who has to act (the human, the robot or both of them) and when to perform it is
+made by {\sc Shary} with the help of HATP, based on perspective-taking and
+situation assessment ({\sc Spark} and {\sc Oro} as well as on estimation of the
+human mental state ({\sc Shary} with the help of {\sc Oro}). The actual
+realization of the action is eventually supported by a full set of on-line
+planning functions provided by {\sc Mhp} that computes not only trajectories but
+also the pertinent placements and postures based not only on geometric
+information but also estimation of the human current mental state and
+preferences ({\sc Oro}).
 

### 2016-07-03 paper.tex
Update on Overleaf.

@@
 
-This architecture moves away from classical layered approaches found in
-robotics~\cite{Gat1998three, Volpe2001CLARAty, Goldberg2002}. Contrary to
-layered architectures, interactions between components at the deliberative level
-are here essentially bidirectional and we do not introduce layers of abstraction
+Our architecture has not been designed to replicate or provide a
+plausible model of human cognition.  The main design principles where
+to nurture the decisional components of the robot with models of human
+behaviour and human preferences in order to develop a pertinent
+cognition for a robot that is able to serve and interact fluently with
+humans. In this sense, it aims at the same objectives as\cite{Fong2005,trafton2013act}.
+
+This architecture is based on the same principles as a number of known robotic
+layered architectures~\cite{Alami98,Gat1998three,Volpe2001CLARAty, Goldberg2002}.
+
+However, the main point for us was to refine and study in detail the
+``internal'' architecture of the deliberative level.
+The interactions between components at the deliberative level
+are essentially bidirectional and we do not introduce layers of abstraction
 amongst these deliberative components\footnote{We do have lower-level modules to
@@
 Section~\ref{sect|com}). 
-
-Our architecture however relates to \emph{Beliefs, Desires, Intentions} (BDI)
+Another example is the intricate relation between high-level symbolic
+planning and geometric planning to deal with affordances and human preferences.
+
+Our architecture relates to \emph{Beliefs, Desires, Intentions} (BDI)
 architectures. As put by Woolridge~\cite{Woolridge1999}, BDI architectures are

### 2016-07-04 paper.tex
Second round of proof-reading 1

--- 2016-07-01 paper.tex
@@
 
-We identify the needed individual and collaborative cognitive skills: geometric
+We identify first the needed individual and collaborative cognitive skills: geometric
 reasoning and situation assessment based on perspective-taking and
@@
     \item a physical environment, estimated through the robot's exteroceptive
-        sensing capabilities, and complemented by inferences drawn from previous
+        sensing capabilities, and augmented by inferences drawn from previous
         observations,
@@
 
-The main contributions concern the architecture of the decisional layer of a robot designed to share 
-the space and the task with humans and to act and interact in a way that facilitates human action and decision.
-
-We present a model of cognitive integration for service robots that:
+Our main contributions focus on the architecture of the decisional layer of
+social robots. Specifically, the deliberative architecture of a robot designed
+to share space and tasks with humans, and to act and interact in a way that
+supports the human's own actions and decisions. We present hereafter a model of cognitive
+integration for service robots that:
 
 \begin{itemize}
-
-    \item achieves advanced grounding in complex real environments involving one or
+    \item exposes a principled approach to integrate a set of complex cognitive
+        components in an explicit, semantics-oriented and yet loosely-coupled
+        fashion;
+
+    \item achieves multi-modal and interactive symbol
+        grounding~\cite{Harnad1990} in complex, real-world environments involving one or
         several humans and a robot;
 
-    \item supports a distributed computation of symbolic knowledge for
-        situated dialogue, thanks to the combination of perspective taking,
-        affordances computation and logical inference;
+    \item distributes the computation of symbolic knowledge by combining perspective taking,
+        affordances computation, situated dialogue and logical inference;
 
@@
         explicit contextual reasoning on humans and robot abilities, to generate
-        human-robot shared plans;
-
-    \item exposes a principled approach to integrate a set of complex cognitive
-        components in an explicit, semantics-oriented and loosely-coupled
-        fashion.
+        human-robot shared plans.
 
@@
 
-Importantly, this architecture is fully implemented and we demonstrate it on
-several robotic platforms and in several actual interaction scenarios. It also proved
-to be an effective framework for novel contributions about human-robot joint
+This architecture is fully implemented and we demonstrate it on several robotic
+platforms and in several interaction scenarios. It eventually proves to be an
+effective framework for novel contributions about human-robot joint
 action~\cite{fiore2016planning,devin2016implemented,milliez2016using}, as well
-as for multi-disciplinary studies~\cite{dautenhahn2006may,koay2007exploratory,Ros2010b,dehais2011physiological,ferreira2015users,GharbiROMAN2015}.
-
-\vspace{0.5cm}
-The remaining of the article discusses the robotic architecture that we have
-built to address social and autonomous human-robot interaction, and how it
-relates to other approaches. We organise this discussion in four sections.
-
-The next section introduces the architecture as a whole, as well as the
-knowledge model that we have developed for our robots. Then,
-Section~\ref{sec:impl} provides further details on each of the main components of the architecture.
-Section~\ref{sec:expe} presents two studies that illustrate in a practical way
-what can be currently achieved with our robots.  The
-Sections~\ref{sect|discussion} and~\ref{sect|conclusion} finally summarise our
-main contributions and discuss the key challenges that human-robot interaction
-brings to Artificial Intelligence.
+as for multi-disciplinary
+studies~\cite{dautenhahn2006may,koay2007exploratory,Ros2010b,dehais2011physiological,ferreira2015users,GharbiROMAN2015}.
+
+\vspace{0.5cm} The remaining of the article details this robotic architecture.
+We organise this discussion in five sections. The next section introduces the
+architecture as a whole, as well as the knowledge model that we have developed
+for our robots. Section~\ref{sec:impl} discusses each of the cognitive
+components of the architecture.  Section~\ref{sec:expe} presents two studies
+that illustrate in a practical way what can be currently achieved with our
+robots.  The Sections~\ref{sect|discussion} and~\ref{sect|conclusion} finally
+summarise our main contributions and restate the key challenges that human-robot
+interaction brings to Artificial Intelligence.
 
@@
 architectural challenge. In particular, properly managing the rich semantics of
-human-level natural interaction raises a range of issues. Our basic assumption
+natural interactions with humans raises a range of issues. Our basic assumption
 and guiding principle is that human-level interaction is easier to achieve if
@@
 blackboard, connects most of the modules: the geometric
-reasoning module ({\sc Spark}) produces at relatively high frequency symbolic assertions (like
-\stmt{BOOK1 isOn TABLE}) describing the state of the robot environment and its evolution over time. 
-These logical statements are stored in the knowledge base, and
-queried back, when necessary, by the language processing module ({\sc Dialogs}), the symbolic task
-planner (HATP) and the execution controller. The output of the language
-processing module and the activities managed by the robot controller are
-likewise stored as symbolic statements.
-
-For instance, when processing a sentence like ``give me another book'', the {\sc
-Dialogs} module queries the knowledge base: {\tt \footnotesize find(?book type
-Book, ?book differentFrom BOOK1)}\footnote{We present a complete
-example in section~\ref{moving-london}.}, and write back assertions like
-\stmt{HUMAN desires GIVE\_ACTION45, GIVE\_ACTION45 actsOn BOOK2} to {\sc
-Oro}. The HATP planner then uses the knowledge base to initialise the
-planning domains with similar requests (like {\tt \footnotesize find(BOOK2 isAt
-?location)}), and the execution controller typically monitor
-conditions (by subscribing to events like: {\tt \footnotesize
-onNewMatch(HUMAN desires ?goal)}) and stores what the robot is currently
-doing (\stmt{myself currentlyPerforms GIVE\_ACTION45}).
-
-This architecture moves away from classical layered approaches found in
-robotics~\cite{Gat1998three, Volpe2001CLARAty, Goldberg2002}. Contrary to
-layered architectures, interactions between components at the deliberative level
-are here essentially bidirectional and we do not introduce layers of abstraction
-amongst these deliberative components\footnote{We do have lower-level modules to
-execute actions or manage sensors, but all cognition-related modules live in the
-same global deliberative space.}. The natural language processing component
-illustrates this structure: instead of being an independent input modality whose
-outputs would be unidirectionally fed to ``higher'' decisional components, it
-lives in the deliberative space at the same level as other deliberative
-components, and uses the knowledge base in a bidirectional manner, to interpret,
-disambiguate natural language, and eventually store newly produced
-interpretations (natural language processing is discussed in
-Section~\ref{sect|com}).
-
-Our architecture however relates to \emph{Beliefs, Desires, Intentions} (BDI)
+reasoning module ({\sc Spark}) produces at relatively high frequency symbolic
+assertions describing the state of the robot environment and its evolution over
+time.  These logical statements are stored in the knowledge base, and queried
+back, when necessary, by the language processing module ({\sc Dialogs}), the
+symbolic task planner (HATP) and the execution controller ({\sc Shary} or {\sc
+pyRobots}). The output of the language processing module and the activities
+managed by the robot controller are stored back as symbolic statements as well.
+
+For instance, a book laying on a furniture might be picked up by {\sc Spark} and
+represented in symbolic terms as \stmt{BOOK1 type Book, BOOK1 isOn TABLE}. These
+symbolic statements are stored in the knowledge base {\sc Oro} and made
+available to the other cognitive modules. Later, the robot might process a
+sentence like ``give me another book''. The {\sc Dialogs} module would then query the
+knowledge base: {\tt \footnotesize find(?obj type Book, ?obj differentFrom
+BOOK1)},
+and write back assertions like \stmt{HUMAN desires GIVE\_ACTION45,
+GIVE\_ACTION45 actsOn BOOK2} to {\sc Oro}. This would in turn trigger the
+execution controller {\sc Shary} to prepare to act. It would first call the HATP
+planner. The planner uses the knowledge base to initialise the planning domain (\eg
+{\tt \footnotesize find(BOOK2 isAt ?location)}), and returns a full symbolic
+plan to the execution controller. Finally, the controller would execute the plan and monitor
+its achievement, both for itself and for the human. We present complete examples
+of similar interactions in Section~\ref{sec:expe}.
+
+Our architecture has not been designed to replicate or provide a
+plausible model of human cognition, and in this sense, we distinguish ourselves
+from research on \emph{cognitive architectures}. Instead, our main design
+principle is to nurture the decisional components of the robot with models of human
+behaviour and human preferences in order to develop an effective artifical
+cognition for a robot that is able to serve and interact seamlessly with
+humans. In this sense, it shares its objectives with~\cite{Fong2005,trafton2013act}.
+
+At high level, this architecture relies on the same principles as a number of
+known robotic layered architectures~\cite{Alami98,Gat1998three,Volpe2001CLARAty,
+Goldberg2002}. However, the main point for us was to refine and study in detail
+the internals of the deliberative layer. In our model, we suggest that the interactions
+between components within this deliberative level have to be essentially
+bidirectional. We also suggest not to introduce any sub-layers of abstraction amongst these deliberative
+components\footnote{We must clarify that we do have lower-level modules to execute actions or manage
+sensors, but all cognition-related modules live in the same global deliberative
+space.}. The natural language processing component illustrates this structure:
+instead of being an independent input modality whose outputs would be
+unidirectionally fed to ``higher'' decisional components, it lives in the
+deliberative space at the same level as other deliberative components, and make
+use of the knowledge base in a bidirectional manner, to interpret, disambiguate natural
+language, and eventually store newly produced interpretations (natural language
+processing is discussed in Section~\ref{sect|com}). Another example is the
+intricate relation between high-level symbolic planning and geometric planning
+to deal with affordances and human preferences.
+
+Our architecture relates to \emph{Beliefs, Desires, Intentions} (BDI)
 architectures. As put by Woolridge~\cite{Woolridge1999}, BDI architectures are
@@
 each step the best sub-goal. It becomes then an intention that the robot commits
-to.
-
-As for any cognitive system, this fundamental interaction between knowledge and
+to. As for any cognitive system, this fundamental interaction between knowledge and
 actions is central to our approach as well, and typically involves the dialogue
@@
 
-We extend upon BDI architectures by running other background deliberative task,
+We extend upon BDI architectures by running other background deliberative tasks,
 without them being explicitly triggered by \emph{desires} in the BDI sense.
@@
 clients of the {\sc Oro} server are in charge of managing themselves the
-knowledge (when to add, when to update, when to retract knowledge): no
-meta-semantics are carried over that would let the server manage itself these
+knowledge (when to add, when to update, when to retract knowledge) as no
+meta-semantics are carried over that let the server manage itself these
 dynamics.\footnote{With one exception: so-called \emph{memory profiles} let the
@@
 A priori} knowledge is stored in an ontology (the {\sc OpenRobots} ontology,
-discussed hereafter) and loaded at start-up. This static source covers the
-\emph{common-sense} knowledge of the robot, and optionally some
+discussed hereafter) and is loaded at start-up. This static source implements the
+\emph{common-sense} knowledge of the robot, and might optionally include 
 scenario-specific knowledge (for instance, about objects that are to be
@@
 perception, interaction and planning. The next sections go into details of these
-processes. Lastly, the third source of symbolic statements comes from the
+processes. The third source of symbolic statements comes from the
 inferences produced by the reasoner.
@@
 Section~\ref{sect|sit-ass}). For instance, whenever a client queries KnowRob to
-know if \stmt{object1 isOn object2}, KnowRob calls a geometric reasoner to
+know if \stmt{OBJECT1 isOn OBJECT2}, KnowRob calls a geometric reasoner to
 evaluate if this specific spatial relation holds at that point in time. With our
-approach, spatial relations are instead computed and asserted {\it a priori}.
+approach, spatial relations are instead computed and asserted {\it a priori} by
+a dedicated process that continuously runs in the background.
 This design choice trades scalability for explicit reasoning: at any time, the
-full belief state is made explicit (which may be computationally expensive), and
+full belief state is made explicit, and
 therefore provides the reasoner with the largest possible inference domain. This
 is of special importance for an \emph{event-based} architecture like ours (see
-Section~\ref{sect|ctrl}), where, by definition, we do not know beforehand what
-we are ``looking for'', and we need instead to greedily infer as much as
-possible to build a belief state as comprehensive as possible.
+Section~\ref{sect|ctrl}), where an explicit and comprehensive belief state is
+required to not miss events.
 
@@
 understanding of its trades-off, thanks to being widespread in the semantic Web
-community, the availability of mature libraries to manipulate the ontology,
+community; the availability of mature libraries to manipulate the ontology;
 interoperability with several major on-line knowledge bases ({\sc
-OpenCyc}, {\sc WordNet}, {\sc DBPedia} or {\sc RoboEarth}~\cite{Waibel2011} are
-some examples), open-world reasoning, and the formal guarantee of decidability
+OpenCyc}, {\sc WordNet}, {\sc DBPedia} or {\sc RoboEarth}~\cite{Waibel2011} for
+examples); open-world reasoning (which enables us to represent that some facts
+might be unknown to the robot); and the formal guarantee of decidability
 (it is always possible to classify a Description Logics ontology).
@@
 accordingly the two knowledge models it maintains: the \concept{robot} model is
-updated with the fact \stmt{book isVisible true}, while the \concept{human}
-model is updated with \stmt{book isVisible false}. These two logical statements
+updated with the fact \stmt{BOOK isVisible true}, while the \concept{human}
+model is updated with \stmt{BOOK isVisible false}. These two logical statements
 are simultaneously asserted, yet contradict when taken together. Since the two
@@
 objects that are found are successively marked as \emph{active concepts} by
-inserting statements such as \stmt{human type ActiveConcept} in the short-term
+inserting statements such as \stmt{HUMAN type ActiveConcept} in the short-term
 memory (which can be considered, in this case, to be a working memory).
@@
         \hline
-        \stmt{book\_1 type Book} \\
-        \stmt{human\_1 type Human} \\
+        \stmt{BOOK1 type Book} \\
+        \stmt{HUMAN1 type Human} \\
         ~\\
@@
 
-        \concept{human\_1} says:\\
+        \concept{HUMAN1} says:\\
         ``Give me the book'' \\
@@
         \concept{find(?obj type Book)} \\ 
-        \hspace{0.2cm}$\Rightarrow$ \concept{?obj = book\_1} \\
+        \hspace{0.2cm}$\Rightarrow$ \concept{?obj = BOOK1} \\
         ~\\
@@
         \hline
-        \stmt{human\_1 desires sit\_1} \\
-         \stmt{sit\_1 type Give} \\
-         \stmt{sit\_1 performedBy myself} \\
-         \stmt{sit\_1 actsOnObject book\_01} \\
-         \stmt{sit\_1 receivedBy human\_1} \\
+        \stmt{HUMAN1 desires SIT1} \\
+         \stmt{SIT1 type Give} \\
+         \stmt{SIT1 performedBy myself} \\
+         \stmt{SIT1 actsOnObject BOOK1} \\
+         \stmt{SIT1 receivedBy HUMAN1} \\
     \end{tabular}
@@
 state  into \emph{experiences} (for instance, ``I'm tired'' is processed into
-\stmt{human\_1 experiences state\_1, state\_1 hasFeature tired}, see also
+\stmt{HUMAN1 experiences STATE1, STATE1 hasFeature tired}, see also
 Section~\ref{sect:desires}). A full account of the {\sc Dialogs} features and
@@
 a simpler interaction. For example, if the human says ``this'' without the robot
-tracking what the human points at, no \stmt{human\_1 pointsAt ...} fact would
+tracking what the human points at, no \stmt{HUMAN1 pointsAt ...} fact would
 possibly be available in the knowledge base, and the robot falls back on the
@@
 This allows us to write reactive robot controllers with a high level of
-expressiveness: for instance, by subscribing to the event \stmt{human1 desires
-?action, ?action type Give, ?action actsOnObject ?obj, ?obj type Book}, we could
+expressiveness: for instance, by subscribing to the event \stmt{HUMAN1 desires
+?action, ?action type Give, ?action actsOnObjecT ?obj, ?obj type Book}, we could
 trigger a behaviour when the human expresses (through dialogue, gestures...)
@@
 Note also that we take advantage of the reasoning capabilities of the system:
-for example, the type of the object (\stmt{?obj type Toy}) may not be
+for example, the type of the object (\stmt{?obj type Book}) may not be
 explicitly asserted, but inferred by the reasoner based on other assertions.
@@
 way to select which knowledge about the object is relevant in the interaction
-context. For instance, the knowledge base may store facts like \stmt{obj1 type
+context. For instance, the knowledge base may store facts like \stmt{OBJ1 type
 ActiveConcept} (which internally means that this concept was mentioned in a
@@
 appropriate in a given interaction context. For instance, the system may
-compute that at a given instant the human is looking at the object: \stmt{human
+compute that at a given instant the human is looking at the object: \stmt{HUMAN
 looksAt obj1}. This property makes sense to both parties, but in the context of

### 2016-07-04 paper.tex
Second round of proof-reading 2

@@
 ontology sized for a typical study (about 100 classes and 200 instances),
-classification takes around 100ms, which leads to lags during
+classification takes around 100ms, which may introduce perceptible delays during
 interactions.  Besides, the performances are difficult to predict: the insertion
-of simple statements may indirectly change abruptly the logical
-complexity of the whole knowledge model and lead to notable degradation of
+of seemingly simple statements may change abruptly the logical
+complexity of the knowledge model and lead to a noticeable degradation of
 classification time.
@@
 phenomena (like time) or uncertain phenomena. When required (for instance for
-action recognition), these are managed by dedicated components (like {\sc
-Spark}, discussed in Section~\ref{sect|sit-ass}), and are not exposed at
-the semantic level.
+action recognition), these are managed within dedicated components (like {\sc
+Spark}, discussed in Section~\ref{sect|sit-ass}), and are not represented in the
+knowledge base.
 
 Alternative formalisms have been successfully investigated in robotics to
-address some of these restrictions. \emph{Answer Set Programming} has been
-used~\cite{Chen2010,Erdem2012} for instance to better supports non-monotonic
+address some of these restrictions. Besides the Prolog/OWL combination relied
+upon by {\sc KnowRob}, \emph{Answer Set Programming} has been
+used in robotics~\cite{Chen2010,Erdem2012} for instance to better supports non-monotonic
 reasoning. We have also pointed in~\cite{lemaignan2015mutual} how modal logics like
 the epistemic logics could be relevant to the particular field of social
-human-robot interaction as they would allow for convenient representation of
+human-robot interaction as they allow for the representation of
 alternative mental models. First-order logic and OWL ontologies have however
@@
 
-As previously said, the logical statements exchanged between the deliberative
+As previously mentioned, the logical statements exchanged between the deliberative
 components are organised within the {\sc Oro} knowledge base into an ontology.
 The conceptualization (\ie the system of concepts or \emph{TBox}) of this
-ontology is generally statically asserted, while the instantiation (the
+ontology is statically asserted\footnote{It can however be altered at runtime,
+see the ``Cat'' example in Section~\ref{sec:nlp}.}, while the instantiation (the
 \emph{ABox}) of the ontology is generally dynamically asserted at run-time, by
-the other components.
+the other cognitive components.
 
@@
 practical (\ie covering our experimental needs) and conforming as much as
-possible to existing standards (in our case, the {\sc OpenCyc} upper
+possible to existing standards (specifically, the {\sc OpenCyc} upper
 ontology~\cite{lenat1990cyc}).
@@
 the choices of concepts to model, \emph{top-down} regarding the upper part of
-the taxonomy. This upper part of the ontology is pictured on
+the conceptualization. This upper part of the ontology is pictured on
 Figure~\ref{fig|upper_tbox}. All the classes visible in this figure belong to
@@
 be shared or extended with well-defined semantics. The concept of \emph{Object}
-represents a typical case of ambiguous meaning: in everyday conversation, an
-object is a relatively small physical thing, that can be typically manipulated.
+is a good example as it represents a typical case of ambiguous meaning: in everyday conversation, an
+object is a relatively small physical thing, that can be usually manipulated.
 A human is not usually considered as an object. {\sc OpenCyc} however precisely
@@
 obviously the humans, and actually many other entities that would not be
-commonly said to be objects (the Earth for instance).\footnote{In such cases of
-discrepancy between {\sc OpenCyc} concepts and human terminology, we manually
-label the {\sc OpenCyc} concepts with appropriate human names: for instance, the
-OpenRobots Ontology associates the label ``\emph{object}'' to the concept
-\concept{cyc:Artifact} instead of the concept \concept{cyc:PartiallyTangible}.
-These labels are used in priority during the grounding of verbal interactions.}
-Thus the importance of relying on well-defined and standard semantics to
-exchange information between artificial systems in a non-ambiguous way.
+commonly said to be objects (the Earth for instance).
+By relying on well-defined and standard semantics to
+exchange information between artificial systems, we avoid semantic ambiguities.
+As the robot interacts with humans, we must however address the discrepancy
+between {\sc OpenCyc} concepts and human terminology. In these situations, we
+manually label the {\sc OpenCyc} concepts with appropriate human names: for
+instance, the OpenRobots Ontology associates the label ``\emph{object}'' to the
+concept \concept{cyc:Artifact} instead of the concept
+\concept{cyc:PartiallyTangible}.  These labels are used in priority during the
+grounding of verbal interactions.
 
@@
 
-The class \concept{PurposefulAction} is the superset of all the actions that are
+On this same figure, the class \concept{PurposefulAction} represents the superset of all the actions that are
 purposefully performed by the robot (or another agent). Actual actions (\ie
@@
 Figure~\ref{fig|tangible_things_tbox}. Our `bottom-up'' design process of the
-ontology is clearly apparent on this figure: only the subclasses relevant to the
-context of service robotics in an human-like environment are explicitly stated.
+ontology is apparent on this figure: only the subclasses relevant to the
+context of service robotics in an human-like environment are asserted:
 For performance reasons as well as clarity, we have decided against an extended
@@
 If a human informs the robot that a given object is indeed a bottle, the robot
-derives more information on the object. Conversely, if the human affirms that
+can consequently derive more information on the object. Conversely, if the human affirms that
 ``a car is a bottle'', the reasoner may detect logical contradictions (like
@@
 both to ensure that the verbal input has been correctly acquired and parsed,
-and also to verify that what the human says is logically consistent.  We
-further discuss the strengths and weaknesses of this knowledge framework in
+and also to verify that what the human says is logically consistent.
+
+We further discuss the strengths and weaknesses of this knowledge framework in
 Section~\ref{krs-discussion}.
@@
 and reasoned about~\cite{Harnad1990}. This represents an important cognitive
-skill, in particular in the human-robot interaction context: in that case, the
+skill, in particular in the human-robot interaction context: in this situation, the
 link that the robot has to establish between percepts and symbols must
@@
 actuation capabilities of the robot. The different components that we have
-mentioned so far provide a bottom-up grounding process: geometric reasoning and
+mentioned so far exhibit grounding mechanisms: geometric reasoning and
 dialogue processing modules constantly build and push new symbolic contents
-about the world to the knowledge base where it becomes accessible to decisional
-layers. We detail this process in the next sections.
+about the world to the knowledge base. We detail this process in the next sections.
 
@@
 
-The previous section has provided an overview of our cognitive architecture
-as well as the associated knowledge model. We discuss in this section its
-building blocks. Each of these components have connections to several
-others, and Figure~\ref{fig|archi} can be referred to as a guide.
+The previous section has introduced the integration model of our architecture,
+as well as the associated knowledge model. We discuss in this section each of
+its building blocks. They are pictured in Figure~\ref{fig|archi}, along with
+their connections to the others components.
 
@@
 deliberative behaviours that are \item \emph{stateful}, \ie keeping track of
-previous states is typically needed for the component to perform adequately;
+previous states is typically needed for the component to perform appropriately;
 \item \emph{amodal} in that the skill is not inherently bound to a specific
@@
 
-We present first the main \emph{internal} cognitive capabilities, implemented in
+We present first the main internal cognitive capabilities, implemented in
 the {\sc Oro} knowledge base itself, and then discuss successively the situation
@@
 
-As mentioned in the previous section, we use the Pellet open-source reasoner to
-reason on the knowledge base. It supports several standard inference processes:
+As mentioned in the previous section, we use the {\sc Pellet} open-source reasoner to
+reason on the knowledge base. It supports several standard inference mechanisms:
 consistency checking, concept satisfiability, classification and realisation
@@
 build, store and retrieve separate models of the beliefs of the humans it
-interacts with.  Our knowledge base implements such a mechanism: when the robot
+interacts with.  Our knowledge base implements such a
+mechanism~\cite{Lemaignan2010}: when the robot
 detects that a new human has appeared, it initialises a new independent
@@
 robot of each agent's perspective for their actual instantiation. For example,
-the robot may (geometrically) compute that the book is in its own field of view,
+the robot can (geometrically) compute that a given book is in its own field of view,
 but not in the human one (the detail of this computation, called
@@
 are simultaneously asserted, yet contradict when taken together. Since the two
-knowledge models are realised as two independent ontologies, the contradiction
-does not actually occur and both the models remain logically consistent.
+knowledge models are however implemented as two independent ontologies, the contradiction
+does not actually appear and both the models remain logically consistent.
 
@@
 beliefs} on the world (A still thinks the object is at its original
-position since he did not see B hiding it in a new place). Using separate knowledge models
+position since he did not see B hiding it in a new place). Relying on these separate knowledge models
 in the knowledge base, we have been able to replicate this experience with
@@
 robot as \emph{active concepts}: if a human asks the robot ``Give
-me all red objects'', the human, the \concept{Give} action, and each red
-objects that are found are successively marked as \emph{active concepts} by
+me all red objects'', the human, the \concept{Give} action, and every red
+objects that are found are marked as active concepts by
 inserting statements such as \stmt{HUMAN type ActiveConcept} in the short-term
@@
 Likewise, recently seen or updated geometric entities are flagged as
-\concept{ActiveConcept}. We use this feature during dialogue disambiguation for
-instance. On the other hand, our perception layer does not make use of this
-mechanism: as described in the next section, the environment model of the robot
+\concept{ActiveConcept}. We use this feature during dialogue disambiguation to
+access concepts recently referred to. On the other hand, our perception layer does not make use of this
+mechanism. As described in the next section, the environment model of the robot
 is continuously updated and the derived symbolic knowledge is therefore
@@
 \centering
-\resizebox{0.8\textwidth}{!}{%
+\resizebox{\textwidth}{!}{%
 \begin{tikzpicture}[
@@
 
-    \node[box=16cm, minimum height=5.3cm, solid, thick, rounded corners] at (0,1.8) (oro) {};
-    \node at (0,4.2) {\sc \Large oro-server};
-
-    \node[anchor=south] at (-4.6,2.4) {Description Logics \\ (Open World assumption)};
-    \node[box=3.7cm] at (-4.6,1.9) (owl) {};
-    \node[anchor=south] at (-1,2.4) {Multiple simultaneous \\ symbolic models};
-    \node[box=3.3cm] at (-1,1.9) (multi) {};
-    \node[anchor=south] at (2.1,2.4) {Bio-inspired \\ memory};
-    \node[box=2.7cm] at (2.1,1.9) (mem) {};
-    \node[anchor=south] at (5,2.4) {Discrimination \& \\ categorization};
-    \node[box=3cm] at (5,1.9) (disc) {};
-
-    \node at (-3,-0.4) {{\it OWL store:} OpenJena};
-    \node at (3,-0.4) {{\it Reasoner:} Pellet};
-
-    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-    \node[box=18cm, minimum height=4.5cm, dotted, thick, rounded corners] at (0,-4.2) (spark) {};
-    \path (spark.north) edge[->] node[align=left,anchor=west] {Symbolic\\facts} (oro.south);
-
-    \node[anchor=north] at (-6.5,-5.3) {a) Agent-independent\\(allocentric)
+    \node[anchor=north] at (-6.5,-3.3) {a) Agent-independent\\(allocentric)
         spatial relations};
-    \node[anchor=north] at (-2.1,-5.3) {b) Agent-dependent\\(egocentric) spatial
+    \node[anchor=north] at (-2.1,-3.3) {b) Agent-dependent\\(egocentric) spatial
         relations};
-    \node[anchor=north] at (1.6,-5.3) {c) Visibility/Pointing};
-    \node[anchor=north] at (6.5,-5.3) {d) Reachability};
+    \node[anchor=north] at (1.6,-3.3) {c) Visibility/Pointing};
+    \node[anchor=north] at (6.5,-3.3) {d) Reachability};
 
@@
 }
-    \caption{Functional overview of knowledge base ({\sc Oro} server, top part)
-    and the geometric situation assessment module (\emph{{\sc Spark}}, bottom
-    part). {\sc Spark} computes symbolic relationships between objects and
-    agents, and exports them to the knowledge base.}
+    \caption{Functional overview of the geometric situation assessment module
+    {\sc Spark}. {\sc Spark} computes symbolic relationships between objects
+    and agents, and exports them to the knowledge base.  }
 
@@
 Spark} (\emph{SPAtial Reasoning \& Knowledge}, presented in~\cite{Sisbot2011}).
-It is a situation assessment reasoner that generates symbolic knowledge from the
+It acts as a situation assessment reasoner that generates symbolic knowledge from the
 geometry of the environment with respect to relations between objects, robots
@@
 knowledge base. This approach may raise scalability concerns (we did not however
-hit major performance issues in our constrained scenarios) as well as prevent
+observed performance issues in our constrained scenarios, involving typically about 10 objects
+and two agents) as well as prevent
 reasoning on the situation history, but simplifies the management of the
@@
 I update it?}). Since it is equivalent to a reset of the reasoner domain, it
-also essentially nullifies issues linked to non-monotonic reasoning
-(see~\cite{McCarthy2007} for a discussion on that question).
+also essentially nullifies issues linked to non-monotonic reasoning in the
+knowledge base (see~\cite{McCarthy2007} for a discussion on that question).
 
@@
 ability for visually perceiving the environment from another person's point of view.
-This ability allows us to identify the objects in situations where the visual
-perception of one person differs from the other one. In developmental
-psychology, one typical example consists of two similar objects in a room (\eg
-two balls) where both are visible for the child, but only one is visible for
-the adult. Thus, when the adult asks the child to hand over ``the ball'', the
+This ability allows us to properly handle and represent situations where the visual
+perception of one person differs from the other one. A typical example found in developmental
+psychology consists of two similar objects in a room (\eg
+two balls) that are both visible for a child, but only one is visible to the
+adult. When the adult requests the child to hand over ``the ball'', the
 child is able to correctly identify which ball the adult is referring to (\ie
@@
 \subsubsection{Natural Language Grounding}
+\label{sec:nlp}
 

### 2016-07-04 paper.tex
Second round of proof-reading 3

@@
 (like \emph{on}, \emph{next to}...) instead of precise, absolute positions
-(qualitative spatial reasoning). These type of descriptors have been studied in the context of language grounding
+(qualitative spatial reasoning). These type of descriptors have been extensively studied in the context of language grounding
 \cite{O'Keefe1999,Matuszek2010,Regier2001,Kelleher2006,Blisard2005}.  {\sc
@@
 
-{\sc Spark} computes three main agent-independent relations~\cite{Sisbot2011}
+{\sc Spark} computes three main agent-independent relations
 based on the bounding box and centre of mass of the objects
-(Figure~\ref{fig|spark-oro}\emph{a}): \concept{isOn} holds when an object $O_1$
+(Figure~\ref{fig|spark-oro}\emph{a},~\cite{Sisbot2011}): \concept{isOn} holds when an object $O_1$
 is on another object $O_2$, and is computed by evaluating the centre of mass of
@@
 $BB_{O_1}$ and $BB_{O_2}$.  \concept{isNextTo} indicates whether an object $O_1$
-is next to another object $O_2$. We do not use a simple distance threshold to
+is next to another object $O_2$. Note that we do not use a simple distance threshold to
 determine if two objects are next to each other since the relation is highly
@@
 objects (\eg two bottles). Thus, the distance threshold is scaled with the
-objects' size. {\sc Spark} also compute symbolic facts related to agent
+objects' size. Finally, {\sc Spark} also computes symbolic facts related to agent
 independent world dynamics.  The predicate \concept{isMoving} states, for each
@@
 (egocentric perspective).  The predicate \concept{hasRelativePosition}
-represents such spatial relations between agents and objects that are agent
-dependent.  We compute these spatial locations by dividing the space around the
+represents the superset of such spatial relations between agents and objects
+that are agent-dependent. We compute these spatial relations by dividing the space around the
 referent (an agent) into $n$ regions based on arbitrary angle values relative to
@@
 depending on the context where the interaction takes place.
-
 Through perspective taking, {\sc Spark} computes for each agent a symbolic
 description of the relative positioning of objects in the environment.
+
 Table~\ref{facts} summarises all the symbolic spatial relations computed by {\sc
@@
     Figure~\ref{fig|spark-oro}\emph{c} the field of view of a person is
-    illustrated with a grey cone (the wider one). While he is able to see the two
-    small boxes on the table in front of him, the big box on his right is out of
-    his FOV, and therefore, he is not able to see it.
+    illustrated with a grey cone (the wider one). While she is able to see the two
+    small boxes on the table in front of her, the big box on her right is out of
+    her FOV, and therefore, she is not able to see it.
 
@@
 
-To ensure a sufficiently stable recognition, these three capabilities are post-filtered
-with an hysteresis function at the geometric level.
+We apply an hysteresis filter at the geometric level to ensure a sufficiently
+stable recognition of these three capabilities.
 
@@
     object afterwards.  Reachability is computed for each agent (human or robot) based on
-    Generalised Inverse Kinematics and collision detection. More than that, our robot is able 
+    Generalised Inverse Kinematics and collision detection. Besides, the robot is able 
     to compute an estimate of the effort needed by an agent to reach an object~\cite{pandey2013affordance}.
@@
 Monitoring human activity is needed by the execution controllers to track the
-engagement of the human and the progress of her actions and also to synchronise seamlessly
+engagement of the human and the progress of their actions. It is also needed to synchronise seamlessly
 its own actions with the human actions. Full human action
@@
 geometric heuristics on human hand trajectories and possible objects placements
-to recognise simple elementary actions. Those primitive actions are assessed through
+to recognise simple elementary actions. Those primitive actions are assessed by
 monitoring situations like ``an empty hand is close to an object on a table''
 (precursor for a \emph{pick}), or ``a hand holding an object is over a
-container'' (precursor for a \emph{throw}).  {\sc Spark} recognises a set of
+container'' (precursor for a \emph{put}).  {\sc Spark} recognises a set of
 such primitives. When combined with the other geometric computations and a predictive
@@
 track the fulfilment of the pre- and post-conditions of the predicted human
-actions. As a result, the robot can monitor the engagement of the human and the
+actions. The robot relies on these to monitor the engagement of the human and the
 overall progress of the human-robot shared plan.
@@
 
-In its current form, our situation assessment module makes two strong
+In its current form, our situation assessment module makes two 
 assumptions: the objects are known in advance (hence, we can rely on proper 3D
-CAD model for spatial reasoning) and the robot benefits of an almost perfect
+CAD model for spatial reasoning) and the robot benefits of an nearly perfect
 perception, made possible by the use of fiducial markers.  Each object receives
-a unique tag which enables us to accurately localise it in 3D and prevent
-recognition ambiguities that would be then reflected in the knowledge base.
-While {\sc Spark} algorithms themselves are ignorant of the input sources, and
-would work equally well with a full object recognition stack, we are yet to
-investigate this research area.
+a unique tag which enables an accurate localisation in 3D and prevents
+recognition ambiguities that would be otherwise reflected in the knowledge base.
+While {\sc Spark} algorithms do not concern themselves with the nature of the input sources, and
+would work equally well with a full object recognition stack, we did not
+investigate this research area so far.
 
 Additionally, temporal reasoning (essential for accurate action recognition for
-instance) is not properly addressed in the current version of our system. Temporal reasoning is used
+instance) is not generally addressed in the current state of our system. Temporal reasoning is used
 only locally, and does not allow for tracking of long sequences or global events.
@@
 as an input (processing of the human speech) and as an output (verbalization of the
-robot intentions as well human-robot shared plans). Natural language processing is facilitated by our architecture
-that processes information with semantics that are already close to a
-human-level. This section presents the main features of our speech processor,
+robot intentions, as well as human-robot shared plans). Natural language
+processing is facilitated as our architecture
+manipulates semantics that are close to the
+human level. This section presents the main features of our speech processor,
 {\sc Dialogs}, that include semantic and multi-modal grounding, and interactive
-disambiguation. Algorithmic and implementation details are provided
+disambiguation. Algorithms and implementation details are provided
 in~\cite{Lemaignan2011a}.
@@
 Android-based interface. The interface relies on the Google speech recognition API for
-speech-to-text and relays the textual transcript to the robot. The text is parsed into
-a grammatical structure (\emph{Part of Speech} tagging), and the atoms of each
-sentence are resolved with the help of the ontology to ground concepts like
+speech-to-text (ASR) and relays the textual transcript to the robot. The text is parsed into
+a grammatical structure (\emph{Part of Speech} tagging) by a custom
+heuristics-based parser. The resulting atoms are then resolved with the help of
+the knowledge base to ground concepts like
 objects (\ie when a user says ``pick up the can'', it resolves to which instance of
-\emph{can} the user is referring to) and actions.  Figure~\ref{dialogs|ex} gives
+\concept{Can} the user is referring to) and actions.  Figure~\ref{dialogs|ex} gives
 an example of the processing of a simple, non-ambiguous command. The first
@@
     \begin{tabular}{l}
-        \emph{Initial human knowledge} \\
+        \emph{Initial human model} \\
         \hline
@@
 
-        \emph{Newly created statements}\\
+        \emph{Statements added to robot model}\\
         \hline
-        \stmt{HUMAN1 desires SIT1} \\
-         \stmt{SIT1 type Give} \\
-         \stmt{SIT1 performedBy myself} \\
-         \stmt{SIT1 actsOnObject BOOK1} \\
-         \stmt{SIT1 receivedBy HUMAN1} \\
+        \stmt{HUMAN1 desires SITUATION1} \\
+         \stmt{SITUATION1 type Give} \\
+         \stmt{SITUATION1 performedBy myself} \\
+         \stmt{SITUATION1 actsOnObject BOOK1} \\
+         \stmt{SITUATION1 receivedBy HUMAN1} \\
     \end{tabular}
@@
 more information), and anaphora resolution (``give \emph{it} to me'') based on
-dialogue history. It also supports knowledge extension by learning new semantic
+the dialogue history and the working memory. It also supports knowledge extension by learning new semantic
 structures. For instance, a sentence like ``learn that cats are animals'' is
@@
 
-\subsubsection{Dealing with Multi-Modality}
-
-Because the components of our architecture rely on the same RDF formalism to
+\subsubsection{Dialogue and Multi-Modality}
+
+Because all of the components of our architecture rely on the same RDF formalism to
 represent their outputs, the different communication modalities are presented in
@@
 both to \emph{explicit} modalities (verbal communication, deictic gestures,
-social gaze), and \emph{implicit} modalities (like the body position of the
+gaze focus), and \emph{implicit} modalities (like the body position of the
 human). The dialogue grounding process makes use of them at two distinct levels
@@
 a simpler interaction. For example, if the human says ``this'' without the robot
-tracking what the human points at, no \stmt{HUMAN1 pointsAt ...} fact would
-possibly be available in the knowledge base, and the robot falls back on the
+tracking what the human points at, no \stmt{HUMAN1 pointsAt ...} fact is
+possibly available in the knowledge base, and the robot falls back on the
 anaphora resolution step alone.
@@
 modality by {\sc Dialogs}. \cite{Ros2010b} provides a detailed account of our
-approach toward interactive concept clarification and discrimination, along with the
+approach to interactive concept clarification and discrimination, along with the
 related algorithms.
@@
 \emph{shared plans}~\cite{Grosz1996,Clark1996,Kemp2007} which are then used to
-anticipate human action, to suggest a course of action to humans, or even to ask
+anticipate human action, to suggest a course of action to humans, or possibly to ask
 help from the human if needed.
@@
 
-We briefly discuss here below how HATP incrementally builds and synchronises streams of actions concerning each or the agents (humans and robot) involved in a task, and how it promotes plans that satisfy humans needs and preferences as well as comfort and legibility issues.
+We discuss next how HATP incrementally builds and synchronises streams of
+actions for each or the agents (humans and robot) involved in a task, and how it
+promotes plans that satisfy humans needs and preferences as well as comfort and
+legibility.
 
@@
 The originality of HATP resides in its ability to produce \emph{shared plans} which might involve 
-the robot as well as the other participants (humans or robots).
+the robot as well as the other participants.
 
@@
 allows the robot to monitor the human engagement and the progress in the shared
-plan execution.
+plan execution. A complete example is presented in
+Section~\ref{sec:cleaning-table}.
 
@@
 
-A cost and a duration function are associated to each action.  The duration
-function provides a duration interval for the action achievement and is used, in
-one hand, to schedule the different streams and, on the other hand, as an
-additional cost function. Cost here is not used here to estimate only standard time and energy consumption but
-to estimate a number factors that measure the satisfaction of the human in terms
-of acceptability and legibility of robot behaviour.
+A duration and a cost function are associated to each action. The duration
+function provides a duration interval for the action achievement. It serves both
+to schedule the different streams and as an additional cost function. Besides
+time and energy consumption, the cost function integrates factors that measure
+the satisfaction of the human in terms of acceptability and legibility of the
+resulting robot behaviour.
 
 HATP includes mechanisms called \emph{social rules} to promote plans that are
-considered as suitable for human-robot interaction. The planner
-allows for the specification of the following filtering criteria:
+considered as suitable for human-robot interaction. The following constraints
+can be set:
 
 \emph{Wasted time}: avoids plans where the human spends a lot
-of her time being idle,
+of her time being idle;
 
 \emph{Effort balancing with respect of human desires}: avoids plans where efforts are
-not distributed among the agents taking part to the plan following the human preferences.  It is
+not distributed among the agents taking part to the plan, respective of the human preferences.  It is
 indeed sometimes beneficial to balance efforts between the human and
-the robot. Sometimes the human wants to do more or, on the contrary, prefers to leave the 
-robot do all what it can.
+the robot. Sometimes the human wants to do more, or on the contrary, prefers to leave the 
+robot do most of the work;
 
 \emph{Simplicity}: promotes plans that limit as much as possible the interdependencies
-between the actions of agents mentioned in the plan, as a problem with
-executing just one of those actions could invalidate the entire
-plan. Also intricate human-robot activity may cause discomfort since
-the human will find himself repeatedly in a situation where his is
-waiting for the robot to act,
+between the actions of agents involved in the plan, as an issue during the
+execution of one of those actions would put the entire plan at risk. Also intricate human-robot activity may cause discomfort since
+the human will find herself repeatedly in a situation where she is
+waiting for the robot to act;
 
@@
 
-Combining the above criteria, we typically yield shared plans with desirable
-interaction features like the human still being engaged in a number of tasks
-while its overall level of required efforts remains low, or avoiding having the
-human to wait for the robot (by essentially preventing the action streams from
-having too many causal links between them).
+Combining the above criteria, we yield shared plans with desirable
+interaction features like having the human engaged in a number of tasks
+while her overall level of efforts remains low, or avoiding having the
+human to wait for the robot by preventing the action streams from
+having too many causal links between them.
 
@@
 immediately take it and put it into the bin, thus reducing the human idle time.
-The resulting shared plan is considered better than the one depicted on
-Figure~\ref{plan_hatp1} according to a global evaluation of the costs and rules.
-
-In the current implementation, the social rules are effectively
-implemented as filters and that are applied whenever a plan is produced.
-In the future we intend to extend the plan-search algorithm and integrate the social rules 
-in the process itself.
+
+In the current implementation, the social rules are effectively implemented as
+filters applied to the set of all possible plans computed by the planner.  In
+the future, we intend to extend the plan-search algorithm and integrate the
+social rules in the process itself.
 
@@
 As HATP is a generic symbolic task planner and does not enforce any abstraction
-level for the planning domain, we were able to design a domain built from
+level for the planning domain, we have designed a planning domain made of
 top-level tasks whose semantics are close to the one used in the human-robot
-dialogue: the planner domain effectively contains concepts like \texttt{give},
-\texttt{table}, \texttt{is on}. This leads to an effective mapping between the
+dialogue: the planner domain effectively contains concepts like \texttt{Give},
+\texttt{table}, \texttt{isOn}. This leads to an effective mapping between the
 knowledge extracted from the situation assessment or the dialogue, and the
@@
 
-Also, we would like to mention that, after some research we have decided against
-representing the planning domain (\ie, the set of tasks with their pre- and
-post-conditions) in the knowledge base itself due to expressiveness issues (see
-appendix B of~\cite{Lemaignan2012a} for a detailed discussion). This effectively
-leads to independent declarative and procedural knowledge stores.
+Due to expressiveness issues, we do not represent the planning domain (\ie, the set of tasks with their pre- and
+post-conditions) in the knowledge base directly (see
+appendix B of~\cite{Lemaignan2012a} for the full rationale). This effectively
+leads to independent declarative (the {\sc Oro} knowledge base) and procedural
+(the planner) knowledge stores.
 
@@
 platform~\cite{Lallee2010b}, as reported in~\cite{Lemaignan2010}), we have also
-developed dedicated robot controllers which integrate the whole stack introduced
+developed dedicated robot controllers which integrate into the deliberative
+architecture introduced
 in Figure~\ref{fig|archi}. {\sc Shary}~\cite{clodic2009} is the main one,
@@
 One of the main roles of {\sc Shary} is to control the production and execution
-of shared plans. This means essentially context-based refinement and robot
-actions execution, as well as monitoring of those achieved by its human
+of shared plans. This means essentially context-based refinement and execution
+of actions, as well as monitoring of those achieved by its human
 partner. One of the key design goals is to build such abilities in a generic
 way, and to provide several levels of parametrisation allowing to adapt to
-various environments, and various levels of involvement of the robot ranging
+various environments and various levels of involvement of the robot, ranging
 from teammate behaviour to assistant or proactive helper. Based on this,
@@
 and monitoring acts needed to support human-robot interactive task
-achievement~\cite{Rich1997,Sidner2005,sebanz2006joint}. {\sc Shary} supports
-such mechanisms~\cite{Clodic2007smc}. It allows
+achievement~\cite{Rich1997,Sidner2005,sebanz2006joint}. {\sc Shary} supports as
+well
+such a mechanism~\cite{Clodic2007smc}. It allows
 to bind action execution to \emph{communication policies} in order to produce
@@
 queries (based on the standard SPARQL language) or events. A module can
-subscribe to an event by passing through an event pattern (in its simplest
+subscribe to an event by registering an event pattern (in its simplest
 form, a partial statement like \stmt{? type Book}) and a callback.  Each
@@
 expressiveness: for instance, by subscribing to the event \stmt{HUMAN1 desires
-?action, ?action type Give, ?action actsOnObjecT ?obj, ?obj type Book}, we could
-trigger a behaviour when the human expresses (through dialogue, gestures...)
-that he wants the robot to give her a book.
+?action, ?action type Give, ?action actsOnObject ?obj, ?obj type Book}, we
+effectively trigger a behaviour when the human expresses (through dialogue,
+gestures...) that he wants the robot to give her a book.
 
@@
 \emph{desire} is produced (this is delegated to perception modules), he can
-focus on the semantics of the desire.
-
-Note also that we take advantage of the reasoning capabilities of the system:
+instead focus on the semantics of the desire.
+
+Note that we transparently take advantage of the reasoning capabilities of the
+system to trigger event to trigger events:
 for example, the type of the object (\stmt{?obj type Book}) may not be
@@
 the desired action (to pick, to give, to look, to bring, to show...), along with the action
-parametrization (thematic roles) are extracted from the knowledge base, and
-either passed to the task planner or executed if the procedure is directly
-available.
+parametrization (thematic roles) are fetched by the execution controller in the knowledge base, and
+are either sent as goals to the task planner, or executed if the elementary
+action is directly available.
 
@@
 
-\subsubsection{Human-Aware Action Refinement, Execution and Monitoring}\label{sec:action}
-
-Like most robotic architectures, actions are split into \emph{elementary actions}
-that are combined into \emph{tasks}. Tasks are created either statically or dynamically.
-The {\sc pyRobots} controller, for instance, statically combines the actions listed in
-table~\ref{table|pyrobots_actions} to implement the high-level tasks it exposes: {\tt
-lookAt}, {\tt moveTo}, {\tt getFromHuman}, {\tt showObject}, {\tt giveObject},
-{\tt pickObject}, {\tt bringObject}, {\tt putObject}, {\tt hideObject}.
-Our other controller, {\sc Shary}, relies on the symbolic task planner to
-dynamically generate, at run-time, suitable sequence of elementary actions.
-
-\begin{table}[ht!]
-\begin{center}
-\begin{tabular}{p{0.8\columnwidth}}
-\hline
-    {\bf Manipulation} \\
-     {\tt attachobject}, {\tt basicgive}, {\tt basictake}, {\tt close\_gripper}, {\tt configure\_grippers}, {\tt grab\_gripper}, {\tt handover}, {\tt hide}, {\tt open\_gripper}, {\tt pick}, {\tt put}, {\tt put\_accessible}, {\tt release\_gripper}, {\tt show}, {\tt take} \\
-\hline
-    {\bf Gaze control} \\
-     {\tt glance\_to}, {\tt look\_at}, {\tt sweep}, {\tt switch\_active\_stereo\_pair}, {\tt track}, {\tt cancel\_track} \\
-\hline
-    {\bf Navigation} \\
-     {\tt carry}, {\tt follow}, {\tt cancel\_follow}, {\tt goto}, {\tt moveclose}, {\tt waypoints} \\
-\hline
-    {\bf Local navigation} \\
-     {\tt dock}, {\tt rotate}, {\tt translate} \\
-\hline
-    {\bf Posture configuration} \\
-     {\tt extractpose}, {\tt idle}, {\tt manipose}, {\tt movearm}, {\tt rest}, {\tt setpose}, {\tt settorso}, {\tt tuckedpose} \\
-\hline
-\end{tabular}
-\end{center}
-\caption{Main {\sc pyRobots} actions, sorted by categories. These
-actions are combined at run-time into higher-level \emph{tasks}.}
-
-\label{table|pyrobots_actions}
-\end{table}
-
-The symbolic manipulation and navigation actions listed in
-Table~\ref{table|pyrobots_actions} are eventually further refined and executed by the execution
+\subsubsection{Human-Aware Action Refinement}\label{sec:action}
+
+%Like most robotic architectures, actions are split into \emph{elementary actions}
+%that are combined into \emph{tasks}. Tasks are created either statically or dynamically.
+%The {\sc pyRobots} controller, for instance, statically combines the actions listed in
+%table~\ref{table|pyrobots_actions} to implement the high-level tasks it exposes: {\tt
+%lookAt}, {\tt moveTo}, {\tt getFromHuman}, {\tt showObject}, {\tt giveObject},
+%{\tt pickObject}, {\tt bringObject}, {\tt putObject}, {\tt hideObject}.
+%Our other controller, {\sc Shary}, relies on the symbolic task planner to
+%dynamically generate, at run-time, suitable sequence of elementary actions.
+%
+%\begin{table}[ht!]
+%\begin{center}
+%\begin{tabular}{p{0.8\columnwidth}}
+%\hline
+%    {\bf Manipulation} \\
+%     {\tt attachobject}, {\tt basicgive}, {\tt basictake}, {\tt close\_gripper}, {\tt configure\_grippers}, {\tt grab\_gripper}, {\tt handover}, {\tt hide}, {\tt open\_gripper}, {\tt pick}, {\tt put}, {\tt put\_accessible}, {\tt release\_gripper}, {\tt show}, {\tt take} \\
+%\hline
+%    {\bf Gaze control} \\
+%     {\tt glance\_to}, {\tt look\_at}, {\tt sweep}, {\tt switch\_active\_stereo\_pair}, {\tt track}, {\tt cancel\_track} \\
+%\hline
+%    {\bf Navigation} \\
+%     {\tt carry}, {\tt follow}, {\tt cancel\_follow}, {\tt goto}, {\tt moveclose}, {\tt waypoints} \\
+%\hline
+%    {\bf Local navigation} \\
+%     {\tt dock}, {\tt rotate}, {\tt translate} \\
+%\hline
+%    {\bf Posture configuration} \\
+%     {\tt extractpose}, {\tt idle}, {\tt manipose}, {\tt movearm}, {\tt rest}, {\tt setpose}, {\tt settorso}, {\tt tuckedpose} \\
+%\hline
+%\end{tabular}
+%\end{center}
+%\caption{Main {\sc pyRobots} actions, sorted by categories. These
+%actions are combined at run-time into higher-level \emph{tasks}.}
+%
+%\label{table|pyrobots_actions}
+%\end{table}
+%
+
+Before being executed, elementary actions are further refined by the execution
 controller with the help of a set of dedicated human-aware geometric motion
-planning functions provided by a component called {\sc Mhp} (see Figure~\ref{fig|archi}).
+planning functions provided by a component called {\sc Mhp} (see architecture in Figure~\ref{fig|archi}).
 
@@
 constraints related to the interaction, like optimal mutual reachability
-or optimal visibility~\cite{Pandey2011}
+or optimal visibility~\cite{Pandey2011}.
 
@@
 As such, this is the geometric counterpart of the HATP symbolic task
-planner.
+planner: it is able to make use of a set of social rules to adapt geometric plans to
+social interactions.
 
@@
 We present here elements of two of them in order to illustrate in a practical way
-different aspects of the architecture.  The first one is focused on knowledge
+the different aspects of the architecture.  The first one is focused on knowledge
 representation and verbal interaction: the human asks for help to find and pack
@@
 controller and the HATP symbolic task planner. In this scenario, the human and
-the robot are in a context when they need to cooperatively remove objects from a table. Robot behaviours and
+the robot need to cooperatively remove objects from a table. Robot behaviours and
 motions are fully planned and then executed.
@@
 \emph{Jido}, a single-arm mobile manipulator, is observing while they carry
-over boxes (Figure~\ref{fig|vpt}), and answers questions concerning the
+over boxes (Figure~\ref{fig:interactive-grounding}), and answers questions concerning the
 location of specific objects.  This study focuses on multi-modal, interactive
-grounding only. The robot essentially observes, builds and maintains knowledge about its human
-partners perspectives and affordances and does not actually perform any action besides
+grounding only: the robot observes, builds and maintains knowledge about its human
+partners perspectives and affordances but does not actually perform any physical action besides
 verbal interaction and simple head movements.
@@
 
-\paragraph{Implicit Disambiguation Through Visual Perspective Taking}
+\paragraph{Implicit disambiguation through visual perspective taking}
 
@@
 }
+    \caption{A scenario involving multi-modal, interactive grounding: the humans
+    can refer to invisible or ambiguous objects that the robots anchor to
+    physical objects through multi-modal interactions with the user.}
+    \label{fig:interactive-grounding}
 \end{figure}
@@
 User A enters the room while carrying a large box (Figure~\ref{fig|vpt}). He
-approaches the table and asks Jido to hand him the video tape: ``Jido, can
+approaches the table and asks Jido to hand him over a video tape: ``Jido, can
 you give me the video tape''. The \textsc{Dialogs} module processes this
@@
 Two video tapes are visible to the robot: one on the table, and another one
-inside the cardboard box. Thus, the knowledge base returns both: 
+inside the cardboard box. The knowledge base returns both: 
 \concept{?obj = [BLACK\_TAPE, WHITE\_TAPE]}.
 
-However, only one is visible for User A (the one on the table). Although there is
-an ambiguity from the robot's perspective, the human referred to the tape using
-the definite article \emph{the}: this is interpreted by the natural language
-processor as the human referring to a precise object, in that case, the visible
-one.\footnote{Other heuristics are available to the {\sc Dialogs} module: for
+However, only one is visible to User A (the one on the table). Although there is
+an ambiguity from the robot's perspective, the human referred to the video tape using
+the definite quantifier \emph{the}: this is interpreted by the natural language
+processor as the human referring to a known object, \ie the one visible in the
+human's knowledge model.\footnote{Other heuristics are available to the {\sc Dialogs} module: for
 instance, if a tape had been recently mentioned in the dialogue, this instance
@@
 %
-\paragraph{Explicit Disambiguation Through Verbal Interaction and Gestures}
-
-In this second situation, User B enters the living room without knowing where
-User A has placed the video tapes (Figure~\ref{fig|pointing}). So he first asks
+\paragraph{Explicit disambiguation through verbal interaction and gestures}
+\label{sec:cleaning-table}
+
+In the second situation, User B enters the room without knowing where
+User A had moved the video tapes (Figure~\ref{fig|pointing}). He first asks
 Jido: ``What's in the box?''. The robot first needs to ground the word ``box''.
@@
 and asks User B which box he is referring to by verbalizing the following question: ``Which box, the toolbox or the
-cardboard box?'' User B can answer the question, but he instead decides to point
+cardboard box?'' User B could answer the question, but he instead decides to point
 at it: ``This box'' (Figure~\ref{fig|pointing}). {\sc Spark} identifies the {\tt
@@
 focusesOn(?ag, ?obj)}. The \textsc{Dialogs} module merges both
-sources of information, verbal (``this'') and deictic, by issuing a specific
+sources of information, verbal (``this'') and deictic, by issuing a 
 query to the knowledge base that eventually lifts the ambiguity:
@@
 Finally, \textsc{Dialogs} queries the ontology about the content of the box and
-the question can be answered: ``Wall-E'' (the label of the object is statically
-asserted in the ontology).
+the question can be answered: ``Wall-E'' (the label of the object had been statically
+asserted in the ontology at start-up).
 
@@
 At this point User B wants to know where is the other tape: ``And where is the
-other tape?''. \textsc{Dialogs} processes this sentence using the
+other tape?''. \textsc{Dialogs} is able to convert ``the other tape'' into a new
+query using the
 \concept{differentFrom} OWL predicate:
@@
 disambiguation is required. The referent is uniquely identified and {\sc
-Dialogs} can then query for its location. The robot finally verbalises the
+Dialogs} can then query for its location: \concept{find(BLACK\_TAPE isAt ?loc)}. The robot finally verbalises the
 result \stmt{BLACK\_TAPE isOn table, BLACK\_TAPE isNextTo TOOLBOX} into ``The
@@
 
-\subsection{\emph{Cleaning the table} Study}
+\subsection{Collaborative Task Planning}
 \label{sec:cleantable}
@@
 Figure~\ref{fig|cleantable-timeline} walks through a simplified version of the whole
-task. It depicts a run with a single tape on a table. The tape is
+task. It depicts a run with a single video tape on a table. The video tape is
 reachable by the robot only, while the bin (where the objects are supposed to be
@@
 user request by the {\sc Dialogs} module, not shown on the figure). At $t_1$ on
-Figure~\ref{fig|cleantable-timeline}, the tape is computed by the robot as being
+Figure~\ref{fig|cleantable-timeline}, the video tape is computed by the robot as being
 reachable by the robot only (columns \emph{Perception} and \emph{Knowledge}),
@@
 are verified (at $t_2$, by satisfying the statement \stmt{ROBOT hasInHand
-TAPE}), and the next task is then started (placing the tape so that it becomes
-reachable by the human).
-
-At $t_3$, the tape is now reachable by the human, and the next tasks (taking the
+TAPE}), and the next task is then started (placing the tape so that it is
+reachable to the human).
+
+At $t_3$, the video tape is now reachable to the human, and the next tasks (taking the
 tape and placing it in the bin) have to be performed by the human: the robot
-instructs the user to do so (verbal interaction not pictured on the figure) and
+instructs the user to do so (verbal interaction not represented on the figure) and
 monitors the actions of the human to detect when the tasks' post-conditions are
 satisfied (column \emph{Actions}, right). When these post-conditions are
-fulfilled, the goal is considered to be completed.
-
-This simple example illustrates how the symbolic facts are produced from the
-situation assessment, and, in parallel, used by the execution controller to
+fulfilled, the goal is considered to be achieved.
+
+This example illustrates how the symbolic facts are produced by the
+situation assessment module {\sc Spark}, and, in parallel, used by the execution controller to
 assess the overall progress of the plan.

### 2016-07-04 paper.tex
Second round of proof-reading 4

@@
 The previous sections provide a perspective on a complete deliberative
-architecture for social robots, including an implementation supported
-``real-world'' experimental deployments. This section synthesises next what we
+architecture for social robots, including its implementation, supported
+by experimental deployments. This section synthesises what we
 see as the main challenges that human-robot interaction brings to Artificial
-Intelligence.  We discuss first how embodied cognition is an essential challenge
+Intelligence.  We first discuss how embodied cognition is an essential challenge
 in human-robot interaction; we rephrase then the requirements of joint actions
@@
 physical world (so-called \emph{embodied cognition}). Varela~\cite{Varela1992}
-is one of the main discoverers of these mechanisms, and coined the concept of
+is one of the main discoverer of these mechanisms, and coined the concept of
 \emph{enactivism} as the theoretical framework that studies the links between
@@
 and the robot would continue the grounding -- ``what is a cocktail?'' -- until
-it anchors the new concept to ones it already knows). Embodied interactions
+it anchors the new concept to ones it already knows like \concept{Drink}). Embodied interactions
 (like gestures) are also taken into account at this level: we have presented how
@@
 closed world. This simplifies the task, and we can not claim that our approach
-provides a generic grounding capability. Beyond the perception of geometric
-entities, context, cultural background, ``naive physics'' knowledge, emotional
-state of the human are some of the numerous other determinants that need to be
-accounted for when grounding human-robot interaction.
+provides a generic grounding capability. Context, cultural background, ``naive physics'' knowledge, emotional
+state of the human are some of the numerous determinants beyond the perception of geometric
+entities. They would need to be accounted for when grounding human-robot interaction.
 
@@
 Our context, where a robot has to achieve a task together with
-humans, raises issues to be tackled by the robots' decisional components, and
-that we can summarise as ``W-questions'': \emph{What}, \emph{Who}, \emph{When},
+humans, raises another set of specific issues to be tackled by the robot's
+decisional components. We summarise them as the ``W-questions'': \emph{What}, \emph{Who}, \emph{When},
 \emph{Where} and \emph{How}?
@@
 
-\emph{When to perform a given action} is enriched here by the human context
-since it has to take into account the human, her needs, her rhythms and
+\emph{When to perform a given action} is made more complex by the presence of
+humans. The robot has to take into account her needs, her rhythms and
 pace, and her mental state. While performing its share of the task, the
 robot has to produce signals directed towards the human and to respond to
-signals produced back at the proper pace.
+signals produced back at the proper pace to ensure collaboration.
 
@@
 available, and selecting one is a non-trivial decision problem. Cost-based
-planners are one possible approach: they search for plans that satisfy an
-acceptable cost in terms of \emph{acceptability} or \emph{legibility}, for
-instance.
+planners augmented with social rules are one possible approach: they search for plans that satisfy an
+acceptable cost in terms of \emph{acceptability} or \emph{legibility} as well.
 
@@
 made by {\sc Shary} with the help of HATP, based on perspective-taking and
-situation assessment ({\sc Spark} and {\sc Oro} as well as on estimation of the
+situation assessment ({\sc Spark} and {\sc Oro}), as well as on the estimation of the
 human mental state ({\sc Shary} with the help of {\sc Oro}). The actual
@@
 capabilities of our robots around this idea of explicit knowledge manipulation.
-We finally briefly comment on this design choice and the limitations of our
-approach.
-
-\subsubsection{Explicit Knowledge in our Architecture}
+
+\paragraph{Explicit knowledge in our architecture}
 
@@
 ontology). These first two points lead to a loosely-coupled architecture
-where modules can be removed or replaced easily by other ones as long as they
+where modules can be removed or replaced by other ones as long as they
 share the same semantics: modules are defined by the knowledge they produce or
@@
 
-Those characteristics are not novel \emph{per se}. We want however underline the
+We want to underline the
 shift of focus brought by this approach during the design and integration phases
@@
 by the knowledge they produce and consume. Human-robot interaction, because it
-supposes operations at \emph{human-level} and in environments with complex
+supposes operations at human level and in environments with complex
 semantics, acts here as a motivational force.
 
-\subsubsection{Limits of Disambiguation at Semantic Level}
+\paragraph{Limits of disambiguation at semantic level}
 
@@
 discussion in the last few seconds): this information is not a relevant
-property of \concept{obj1} when trying to disambiguate concepts with humans.
+property of \concept{OBJ1} when trying to disambiguate concepts with humans.
 This distinction between \emph{internal knowledge} (meaningful to
 the system only) and \emph{common knowledge} (whose meaning is understood by
-all the interactors) has not been properly dealt with in our architecture.
+all the agents) has not been properly dealt with in our architecture.
 
@@
 compute that at a given instant the human is looking at the object: \stmt{HUMAN
-looksAt obj1}. This property makes sense to both parties, but in the context of
+looksAt OBJ1}. This property makes sense to both parties, but in the context of
 the \emph{spygame}, we would like to mainly use immanent properties, not
@@
 PEIS Ecology~\cite{saffiotti2005peis, daoutis2012cooperative}, with however a
-stronger emphasise on the representation and interaction with humans. We however
+stronger emphasise on the specificities of the interaction with humans.
+Importantly, we
 distinguish ourselves from research on \emph{cognitive architectures}: cognitive
@@
 primarily concerned with the modeling of human cognition and are less focused on
-practical deployment on socially interactive robots.
-
+the effective deployment on socially interactive robots.
 In that sense, our contribution in terms of architecture is a practical one:
@@
 
-Besides, this work introduces several new contributions related to the
+This work introduces several new contributions related to the
 representation and the management of humans in an autonomous robotic system.
@@
 subsections), we are not aware of a fully implemented architecture that
-effectively combines in a coherent manner all these points. Our approach
-allowed us to identify, refine and combine a number of key cognitive abilities
-that we believe are needed for successful and autonomous social human-robot
-interaction. The novelty and relevance of this contribution to HRI is
-underlined by the range of multi-disciplinary collaborations and studies that
+effectively combines in a coherent manner all these points. The novelty and relevance of this contribution to HRI is
+further underlined by the range of multi-disciplinary collaborations and studies that
 have been made possible by our
@@
 shared plan'' and ``good/acceptable robot behaviour'' in this context. There are
-obviously large avenues for learning and adaptation in this context.
-
-Another direction to head to deals with \emph{contexts representation}.
+large avenues for learning and adaptation in this context.
+
+Another direction to head to deals with context representation.
 Contexts are currently often limited to the current spatial and temporal
-situation. Some models offer the possibility to jump in the past or to switch to
-another agent's perspective, but in current approaches, selecting a context
-always basically consists in retrieving a set of beliefs corresponding to a
+situation. Some of our models offer the possibility to jump in the past or to switch to
+another agent's perspective, but in our current approach, selecting a context
+essentially consists in retrieving a set of beliefs corresponding to a
 situation, and temporarily replacing the current beliefs by those other ones.
@@
 but instead carefully craft a context from several \emph{atomic} contexts.
-Techniques for representation of overlapping pools of knowledge largely remain
+Techniques for representation of overlapping ``pools'' of knowledge largely remain
 to be developed, as well as efficient algorithms to retrieve (or discard) such
@@
 
-Human-Robot Interaction is and is going to remain a challenging field for
+Human-Robot Interaction is and will remain a challenging field for
 Artificial Intelligence. We hope that this contribution helps with clarifying

### 2016-07-04 rebuttal2.tex
Finish the rebuttal

@@
 \usepackage[utf8]{inputenc}
+\usepackage{pdfpages}
 
@@
 during the second round of reviews.  Besides, we have updated many bibliography
-references (and added new ones) to reflect the continous progress of the field
+references (and added new ones) to reflect the continuous progress of the field
 since the initial submission in 2013.
@@
         non-academic phrasing,
-    \item the abstract has been entierly rewritten,
+    \item the abstract has been entirely rewritten,
     \item all the specific language changes pointed out by the editors and
@@
 
-To facilitate the review process, we attach an annotated version of the
+To facilitate the review process, we attach hereafter an annotated version of the
 manuscript where all the changes since the last submission have been highlighted.
 
-\subsection*{Clarifiying the contribution}
+\subsection*{Clarifying the contribution}
 
-We initially made the choice to discuss the existing litterature in relation to
+We initially made the choice to discuss the existing literature in relation to
 our contribution in each of the ``cognitive skills'' sections, for readability
@@
     \item added a summary of the key contribution at the end of the introduction
-        (Section 1.2)
+        (Section 1.2);
+    \item rephrased Section 2.1 to better position our contribution;
     \item split Section 5 into a discussion (Section 5) and a
@@
 
-\subsection*{Reviewer Comments}
+\subsection*{Specific Reviewer Comments}
 
@@
     \item The implicit aspect of multi-modality during dialogue processing has
-        been made clearer (Section 3.3.2).
+        been made clearer (Section 3.3.2);
+    \item The list of 'W- questions' is now better connected to our
+        contribution.
 \end{itemize}
@@
 
+\vspace{1em}
+We hope that this set of revisions addresses adequately the remaining concerns,
+and we certainly remain available if further clarifications are required.
+
+\vspace{1em}
+Best regards,
+
+The authors
 \bibliographystyle{abbrv}
@@
 
+\includepdf[frame=true,pages=-]{diff}
+
 \end{document}
